### 2006-06-30 WhitePapers/ParallelRenderingSystems/paper.tex
Update website to announce subversion import, add necessary documentation files to build website, change Makefile to build website in source tree

\NeedsTeXFormat{LaTeX2e}
\documentclass[10pt,a4paper]{scrartcl}

\usepackage{tabularx,graphicx,setspace}

\ifx\pdfoutput\undefined
  % We're not running pdftex
  % european (better) fonts -- does not look good with pdflatex
  \usepackage[T1]{fontenc}
  \newcommand{\href}[2]{#2\\{\hspace*{5mm}\scriptsize <#1>}\\}
\else
  \pdfcompresslevel=9
  \def\pdfBorderAttrs{/Border [0 0 0] } % No border around Links
  \usepackage{hyperref}
\fi

\title{An Analysis of Parallel Rendering Systems}
\author{Stefan Eilemann\thanks{eilemann@gmail.com}\\[\medskipamount]
%  EyeScale Software Sarl
}
\date{}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
% suppress  single floating lines on top (widow) and bottom(club)
%  10000 is infinity
%  tradeoff: maybe underfull vboxes
\clubpenalty=10000
\widowpenalty=10000

\oddsidemargin .5cm
\evensidemargin .5cm
\topmargin 1cm
\setlength{\textwidth}{15cm}
\setlength{\textheight}{20cm}
\begin{document}

% paragraph spacing -- should be rubber length
%\setlength{\parskip}{1.5ex plus 0.5ex minus 0.5ex}
% paragraph indention for the first line
%\setlength{\parindent}{0em}

\maketitle
\thispagestyle{empty}
% \begin{figure}[ht]
% \centering
% \includegraphics[width=5em]{logo.pdf}
% \end{figure}
\vfill
\abstract{
This White Paper analyzes and classifies the different approaches taken
to parallel rendering. The purpose is to clarify common
misconceptions and false expectations of the capabilities of the
different classes of parallel rendering software.

We examine the rendering pipeline of the typical visualization
application and identify the typical bottlenecks found in this
pipeline. The methods used for parallel rendering are classified in
three fundamental approaches and then analyzed with respect to their
influence on this rendering pipeline, which leads to conclusions about
possible performance gains and the necessary porting effort.

We advocate the need for a generic, open parallel rendering framework to
build scalable graphics software. The inherent limitations of transparent
multi-pipe software are justified, and a full software stack for
graphics clusters is outlined.
}
\vfill
\begin{tabularx}{.85\columnwidth}{|l|l|X|}
\hline
\bf Version & \bf Date & \bf Changes \\
\hline
0.5         & May 19, 2006   & Initial Version\\
\hline
\end{tabularx}
\clearpage
\pagenumbering{arabic}

\section{Introduction}

The purpose of this white paper is to argue for the need for a middleware to
build truely scalable, interactive visualization applications. In
contrast to high-performance computing (HPC), the high-performance
visualization (HPV) community has not generally accepted the need to
parallelize applications for performance. The belief that scalabity can be
achieved transparently by some magic software is still wide-spread, and
contributes to the slow adoption of graphics clusters and virtual reality
installations due to the application performance and (un-)availability.

% Definitions

\subsection{Motivation}
% Introduce the topic
Application developers are faced with making their applications
aware of multiple graphics cards for two reasons -- either the
application has to run on advanced display systems, or the rendering has
to be accelerated by using parallel rendering on multiple graphics cards.

\subsubsection{Multi-View Rendering}
The first motivation is often to run the application on advanced display
systems: workstations with multiple monitors, walls build out of
multiple displays or projectors as well as immersive environments. Such
installations need more than one graphics card to drive all the display
systems. Therefore the rendering has to be distributed across a
number of cards and often across multiple systems. The outputs of the
displays has to be synchronized spatially and time-wise to provide one
coherent image. In immersive environments, stereo rendering is a requirement.

\begin{figure}[ht]
\centering
\includegraphics[width=0.9\columnwidth]{images/sp_to_mp.pdf}
\caption{Some advanced display systems}
\label{FIG_parallel}
\end{figure}

Applications which scale their display size often have to scale
their rendering performance, because the increased display size exposes
previously undiscovered bottlenecks in the rendering pipeline. The higher
display resolution increases the needed pixel fill rate. Due to the increased
image fidelity, data of a higher resolution is used -- which increases
the requirements on the whole rendering pipeline.

\subsubsection{Scalable Rendering}
The second motivation is to accelerate the rendering by aggregating
multiple resources, because the application hits limitations of a single
graphics system.

\begin{figure}[ht]
\centering
\includegraphics[width=0.45\columnwidth]{images/pipeline.pdf}
\caption{The basic rendering pipeline}
\label{FIG_pipeline}
\end{figure}

Figure~\ref{FIG_pipeline} shows a simplified rendering pipeline and the
resources used by during rendering. The typical performance bottlenecks
in such a pipeline are:
\begin{description}
\item[Fill Rate] The amount of rendered pixels can exceed the processing
  capabilities of the graphics card's rasterization pipeline, that
  is, the amount of pixels the card can render in a given time. Volume
  rendering and complex fragment shaders easily limit the application's
  performance due to a fill rate bottleneck.
\item[Geometry Transform] Similar to the fill rate bottleneck, but
  less common, are bottlenecks in the GPU's vertex processing
  capabilities, that is, the amount of triangles a card can
  process. Complex geometry and procedural computations in the vertex
  shader can cause this bottleneck.
\item[GPU Memory] Certain applications, for example volume rendering and
  terrain visualization, easily exceed the amount of memory available on
  the graphics card. For example, a volume of size $1024^3$ has a size
  of 1~GB if it is black and white, and 4~GB if it is colored.
\item[Bus Bandwidth] Dynamic data, such as time-varying data sets and
  interactive modifications of the model, has high requirements on
  the system's IO capabilities. The data can not be cached on the GPU's
  memory but has to be transferred to the card from main memory (or
  disk) during rendering. This is a common bottleneck of interactive
  applications and leads to a rendering performance below the GPU's
  capabilities.
\item[CPU Performance] Similar to the bus bandwidth, the CPU can become
  the bottleneck when traversing the database to generate the rendering
  commands. Tesselation, visibility computations and other
  interpretation of the application data for rendering cause this bottleneck.
\item[Main Memory] The actual model database often holds additional
  application-specific information, which is not used for
  rendering. Additionally, the amount of data displayed is drastically reduced
  during rendering to allow interactive framerates. This leads to
  application scenarios where the required main memory size exceeds the
  capabilities of a single system.
\item[IO Bandwidth] Data sets which exceed the main memory size are
  visualized by roaming through the data, that is, a subset of the whole
  data set is loaded into main memory to be processed and rendered. As
  the user roams through the data, new pieces of it are loaded from
  storage, which can be a bottleneck.
\end{description}

\subsubsection{Parallelism in Graphics Hardware}
Graphics cards address some bottlenecks through hardware
parallelism. GPU's have multiple vertex pipelines (geometry
transform) and multiple fragment pipelines (fill rate) which
accelerate the fill rate and geometry transform rate in hardware,
transparently to the application. Due to physical limitations, the
number of vertex and fragment shading pipelines on a single GPU is limited.

Recent implementations of multi-GPU systems (nVidia SLI, ATI CrossFire)
provide additional parallelism. The individual graphic processors
produce a partial image which is then composited using the hardware
support. The individual GPU's can sit on the same card, or on separate
cards with a special connection between them.

In such a setup, the graphics library dispatches the rendering commands to all
rendering units, a process which is transparent to the application. The overall
pixel fill rate is increased, while other potential bottlenecks are not
addressed, since all commands are sent to all cards and each card needs to
have a copy of all the data (textures, display lists, etc.). A special
case is the alternate frame rendering (AFR) mode,
where each card renders full, alternating frames. This mode also scales
geometry transform and bus bandwidth, while increasing the latency
between input and the rendered output.

Since the application is unmodified, all rendering commands are still
submitted by one thread, which has to be able to saturate the additional
graphics cards to make use of the additional hardware. Distributed scene
graphs and parallel rendering frameworks can in theory parallelize the
rendering for such hardware, once the hardware vendors provide the
necessary API's to program the hardware.

\section{Transparent Multi-Pipe Rendering Software}
% Technical Overview
Transparent rendering software replaces the system's rendering library,
e.g. OpenGL or Direct3D, with a new 'intercept' library. The
application's output is taken at the rendering command level, that is,
the entry points of the rendering library are replaced by new
functions. The new functions package the rendering command stream,
analyze it and send it to a number of rendering processes, potentially
over the network to other systems. This approach is typically used for
the applications which want to use multiple graphics cards to scale the
display size. A limitation of this approach is that unmodified
applications may not be able to run on non-planar display systems,
because important data is not rendered due to the application's view
frustum culling.

% Bottlenecks
Transparent solutions can only increase the rendering performance if the
main bottleneck is on the graphics card (geometry transform or fill rate). The
processing stages to produce the rendering commands are untouched and
remain single-threaded. Potential CPU bottlenecks in the rendering
thread are amplified, since packaging and sending of the rendering
commands to other processes is more time-consuming than writing the
commands directly to a GPU. The bandwidth and latency to the rendering
nodes is worse than the direct local connection to the graphics card. In
reality, this approach often scales the display size at the expense of
performance, since applications are rarely limited by the GPU
processing speed.

% Possible Optimisations
Analyzing the command stream allows some optimizations. The transparent
library can determine the visiblity of rendering commands and textures for each
rendering unit, and only send the data to the appropriate processes.
This can reduce the rendering load of the individual graphics
cards. However, this command stream filtering is only benefitial for the
performance if it can be performed faster than the actual rendering --
which is normally only the case for static chunks of data, such as display
lists and large textures. % mention OpenGL state tracking?

% Parallel Extensions
Some parallel rendering frameworks support parallelization of the
application, that is, the application is modified and renders using
multiple threads. This support is often rudimentary and does not address other
problems, such as configuration and data distribution. Furthermore, by
working on the low-level graphics command stream, the attainable
performance is limited up-front. From an application developers point of
view, writing a parallel application using a transparent framework is at
least the same effort as using a parallel rendering framework. For the
purpose of this white paper, such an extension to a transparent software
can be considered a parallel rendering framework.

\section{Distributed Scene Graphs}
The distributed scene graph approach requires that the application uses
the given scene graph to describe the model database. The rendering of
that scene graph is parallelized by the scene graph's implementation, to
a great extend transparently to the application. The scene graph is
replicated and kept up-to-date on all rendering nodes. During rendering,
the individual instances are traversed in parallel to generate the
rendering commands, which are directly send to the graphics hardware.

From the applications point of view, this approach is similar to the
transparent solutions, in that the rendering is parallelized. The big
difference is that a distributed scene graph approaches the problem at a
much higher level, which allows significant optimisations and leads to
better performance. In contrast to transparent rendering software, the
traversal to generate the rendering commands is parallelized, and the
commands are send --in parallel-- directly to the graphics card. The
scene graph frees the application developer of the task to parallelize
the rendering.

\section{Parallel Rendering Frameworks}
Parallel rendering frameworks help application developers to parallelize
their application. They do not impose a scene graph or other rendering
libraries on the application. By solving the common problems of
multi-pipe rendering, the application developer can focus on
application-specific problems. Parallel rendering frameworks address
synchronization, task decomposition and load-balancing, data transport
as well as the composition of the rendering result. In many ways, they
are  comparable to HPC libraries such as MPI and PVM, while being
focused on interactive applications and parallel rendering.

Parallel rendering frameworks provide an execution environment for any
application, regardless of the used rendering software. Refactoring the
application, other then strictly needed for multi-pipe rendering, is not
necessary. The application provides entry points for its rendering
functions. Depending on the current configuration, the framework creates
the necessary rendering threads which are initialised with a rendering
context by the application's init functions. When the application
request a new frame of rendering, the framework calls the necessary
rendering callbacks in the correct context and order.

\section{Conclusion}

From a technical point of view, choosing the right approach to
parallel rendering depends on the requirements and bottlenecks of the
application. Figure~\ref{FIG_parallelism} gives an overview how well the
different bottlenecks can be addressed with the various solutions.

\begin{figure}[ht]
\centering
\includegraphics[width=0.45\columnwidth]{images/parallelism.pdf}
\caption{Scalability of the different parallel rendering approaches}
\label{FIG_parallelism}
\end{figure}

% transparent rendering
The least intrusive approach of using a transparent software is often explored
first, which frequently does not meet the expectations and
requirements. Performance gains with transparent solutions are 
possible with benchmarks or well-behaved applications, but
real-world software often does not run faster, slowdowns are commonly
observed. Transparent rendering software often supports only a subset of
the features and extensions of the rendering API, in particular anything
which requires a roundtrip to the graphics card is unsupported or
slow. On the other hand, a transparent solution is a viable way to run a
certain set of applications on graphics clusters. For legacy
applications, which cannot be modified, it is the only possibility to make
them multi-pipe ready. The low porting effort is a strong benefit for
transparent rendering software.

% distributed scene graphs
Distributed scene graphs are the ideal and obvious solution if the
application already uses such a scene graph, or is planning on using
one in the near future. The distributed scene graph, especially when
build upon a parallel rendering framework, can deliver optimal
performance while requiring little to no application changes. Porting an
existing application to a distributed scene graph is often impractical
due to the high porting cost.

% parallel rendering frameworks
Parallel rendering frameworks are the common middleware for any
multi-pipe application. Distributed scene graphs and transparent
rendering software can use it as the base for their
implementation. Applications not suitable for these approaches
can easily implement their parallel rendering on such a framework. The
porting effort is reduced to the unavoidable refactoring needed to
separate the rendering from the core application in order to to make it
distributable. A parallel rendering framework addresses common problems
encountered when doing such a port and follows the natural programming
model for any multi-pipe application. For high-performance applications,
a parallel rendering framework is the only solution short of a
completely custom implementation.

Parallel rendering frameworks can and should be the foundation of
transparent solutions and distributed scene graphs. These approaches
then become 'applications' of the parallel rendering framework to
address a certain subset of parallel rendering applications - namely
performance-uncritical and legacy applications, as well as applications
already written using a specific scene graph.

\begin{figure}[ht]
\centering
\includegraphics[width=0.45\columnwidth]{images/layers.pdf}
\caption{Ideal software layering for a visualization system}
\label{FIG_parallel}
\end{figure}

% short conclusion.
The three approaches to parallel rendering complement each other and
serve different application needs. Transparent solutions are a quick way
to run interactive applications on graphic clusters, but they do not
provide the necessary performance for many applications. Likewise,
distributed scene graphs are often too invasive since they require the
applications to use a certain format --the scene graph-- to decribe their
data.

\appendix
\section{Current Parallel Rendering Software}
This appendix gives a short and incomplete overview of the existing
software solutions for parallel rendering.

\subsection{Transparent Multi-Pipe Rendering Software}
\subsubsection{Chromium}
Chromium is a Open Source solution originally developed by the
Stanford University. Some graphics cluster vendors, most notably
Tungsten Graphics, support and extend Chromium on their hardware
platforms. Chromium provides functionality for the synchronization
of the rendering commands when using multiple, parallel application
instances.

\subsubsection{VGP Software Solutions}
The Virtual Graphics Platform sold by ModViz, Inc. is a proprietary
software solution similar to Chromium. The VGP Integration API allows to
augment the OpenGL command stream with additional information to improve
the rendering performance.

\subsubsection{OpenGL Multipipe}
OpenGL Multipipe is a transparent software solution delivered with SGI's
multi-pipe machines. It is not sold separately. For building parallel
applications, SGI offered OpenGL Multipipe SDK.

\subsection{Distributed Scene Graphs}
\subsubsection{TGS Open Inventor Cluster Edition}
The Open Inventor Cluster Edition is an extension to the Open Inventor
scene graph to support parallel, distributed rendering on graphics
clusters. It does support both multi-view and scalable rendering.

\subsection{Parallel Rendering Frameworks}
\subsubsection{OpenGL Multipipe SDK}
OpenGL Multipipe SDK from Silicon Graphics, Inc. is a proprietary
framework for the development of scalable graphics applications. It does
not support distributed rendering.

\subsubsection{CAVELib}
CAVELib from VRCO Inc. is a proprietary, parallel rendering framework
which supports parallel multi-view rendering.

\subsubsection{Equalizer}
Equalizer is an open source parallel rendering framework under development
by the University of Z\"urich and others. Equalizer will
support multi-view and scalable rendering, together with a resource
management system and a transparent rendering layer.

\vfill
\begin{spacing}{.5}{\footnotesize
OpenGL, Open Inventor and OpenGL Multipipe is a trademark of Silicon
Graphics, Inc. SLI is a trademark of nVidia. CrossFire is a trademark of
ATI. Virtual Graphics Platform is a trademark of ModViz, Inc. CAVELib is
a trademark of VRCO Inc. All other products named are trademarks of
their respective owners.
}\end{spacing}
\end{document}

### 2006-07-05 WhitePapers/ProjectEqualizer/paper.tex
Add project description document to web site, rewrite overview page, other minor updates to the web site

\NeedsTeXFormat{LaTeX2e}
\documentclass[10pt,a4paper]{scrartcl}

\usepackage{tabularx,graphicx,setspace}

\ifx\pdfoutput\undefined
  % We're not running pdftex
  % european (better) fonts -- does not look good with pdflatex
  \usepackage[T1]{fontenc}
  \newcommand{\href}[2]{#2\\{\hspace*{5mm}\scriptsize <#1>}\\}
\else
  \pdfcompresslevel=9
  \def\pdfBorderAttrs{/Border [0 0 0] } % No border around Links
  \usepackage{hyperref}
\fi

\title{Project Equalizer}
\author{Stefan Eilemann\thanks{eilemann@gmail.com}\\[\medskipamount]
%  EyeScale Software Sarl
}
\date{\htmladdnormallink{http://www.equalizergraphics.com/}
  {http://www.equalizergraphics.com/}}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
% suppress  single floating lines on top (widow) and bottom(club)
%  10000 is infinity
%  tradeoff: maybe underfull vboxes
\clubpenalty=10000
\widowpenalty=10000

\oddsidemargin .5cm
\evensidemargin .5cm
\topmargin 1cm
\setlength{\textwidth}{15cm}
\setlength{\textheight}{20cm}
\begin{document}

% paragraph spacing -- should be rubber length
%\setlength{\parskip}{1.5ex plus 0.5ex minus 0.5ex}
% paragraph indention for the first line
%\setlength{\parindent}{0em}

\maketitle
\thispagestyle{empty}
% \begin{figure}[ht]
% \centering
% \includegraphics[width=5em]{logo.pdf}
% \end{figure}
\vfill
\abstract{
% what
Equalizer is a project to develop software to simplify the creation of scalable
graphics applications and to improve the usability of multipipe
visualization systems. The main components of Equalizer are a resource
server, which controls the visualization system's configuration, a
client side library for the development of scalable graphics software, a
transparent software layer to execute unmodified applications alongside
with scalable applications, as well as remote visualization
capabilities. Equalizer is designed to support all kinds of
visualization systems ranging from laptops, workstations, multipipe
shared memory systems to graphics clusters.

% why
The purpose of Equalizer is to build a foundation for high performance
visualization by providing a common base for all kinds of applications
used in multipipe environments, in particular for parallel graphics
software.

% how/when
The open development approach of Equalizer maximizes the benefits for
hardware vendors, software developers, research institutions and end
users. Interested parties are strongly encouraged to contact us. We are
looking for contributions or founding for several components of Equalizer.
}
\vfill
{\center\begin{tabularx}{.88\textwidth}{|l|l|X|}
\hline
\bf Version & \bf Date & \bf Changes \\
\hline
0.1         & July 5, 2006   & Initial Draft\\
\hline
\multicolumn{3}{c}{\small Latest version at 
  \htmladdnormallink{http://www.equalizergraphics.com/documents/ProjectEqualizer.pdf}
  {http://www.equalizergraphics.com/documents/ProjectEqualizer.pdf}}\\
\end{tabularx}}
\clearpage
\pagenumbering{arabic}

\section{Overview}
As outlined in \cite{Analysis}, there are three fundamentally different
approaches to parallel rendering. Transparent solutions offer an easy
solution for applications which cannot or need not to be modified, but
are limited in performance and compatibility. Distributed
scene graphs offer better performance, but require the application
to use a certain scene graph. The third approach, a parallel rendering
framework offers a generic API for the development of parallel, scalable
graphics applications.

Today we have satisfying, though often proprietary, solutions for the
transparent approach\footnote{e.g., Chromium from Stanford University,
VGP from ModViz, SVN from IBM}. Distributed scene graphs are beginning
to emerge\footnote{e.g., ScaleViz from Mercury Computer Systems} as
well. There is currently no general programming framework to create
parallel graphics applications. Furthermore, all existing solutions are
'islands', that is, they are configured separately and do possibly
interfere with each other when running simultanously on the same
system. Equalizer addresses these issues.

Unlike the HPC community, visualization software developers do not
widely address performance bottlenecks by parallelizing
their applications. One of the reasons for the lack of parallel graphics
applications is the lack of a standard programming interface, which
implements specific knowledge needed to create scalable visualization
applications. Equalizer provides a programming framework for high
performance visualization.

\section{Project Description}

Equalizer is a project to create new software to facilitate the
development and usage of software for multipipe
graphics systems by addressing the shortcomings of the current
software situation. It adds the two missing components, a
parallel programming framework and a resource management system, and
integrates existing solutions, such as a transparent layer, scene graphs
and remote visualization capabilities.

Equalizer addresses all visualization systems. An Equalizer application
can run unmodified on a laptop, a multipipe workstation, a graphics
cluster or a multipipe shared memory visualization server.

\subsection{Development Model}

The Equalizer development is conducted in an open way. It is licensed
under the LGPL open source license. Contributing parties, such as
hardware vendors, software developers and end users, have the following
benefits from contributing to the project:

\begin{description}
\item[Lower Development Costs] The common functionality of any multipipe
  application is outsourced to the Equalizer framework, thus freeing the
  software developers to reimplement basic algorithms. The expertise for
  high performance visualization is partly embedded into Equalizer, and
  has not to be developed inhouse. The open development approach ensures
  that Equalizer delivers the functionality most needed by the community.
\item[Higher Functionality] Additional features, which would not be
  cost-effective to implement in the application, are provided by the
  Equalizer framework. New features will become available over time,
  and will often be usable by the application with no or minimal code
  changes.
\item[Ease of Use] The central resource management of Equalizer requires
  only a one-time setup, thus minimizing the impact for end users. Once
  a visualization system is set up, unmodified and scalable OpenGL
  applications can be run without any modification. Simple
  configurations, especially shared memory systems, will be configured
  automatically.
\item[Reduced Development Risk] The open source license of Equalizer is
  the guarantee for software developers that the product will be
  available indefinitely. Critical features or bugs, in case they are
  not addressed by the maintainers, can still be addressed internally.
\end{description}

\subsection{Programming Framework}

The programming framework is the main contribution of the
Equalizer project. It addresses common problems when creating parallel
graphics applications, and thus facilitates the development of scalable
multipipe rendering software. It addresses the process creation and
synchronization, rendering task distribution, data transport, and the
composition of the rendering results. The software developer plugs the
application's rendering code into the framework. The Equalizer server
deploys this rendering code according to the system's configuration and
load. The programming framework is the client library to the Equalizer
server, as well as the header files for this library.

\subsection{Resource Management System}

The core of the resource management system is the Equalizer server,
which is managing the system's configuration. It is configured to know the
system resources and schedules the execution of rendering tasks on these
resources. Equalizer applications provide a render client\footnote{The
  render client can be the same executable as the application.}, which
contains the application's rendering code. This render client is
instanciated on the render nodes by the Equalizer server. The
transparent layer and distributed scene graphs are technically an
Equalizer application.

For standalone usage on simple machines, the Equalizer client library
will be able to automatically configure and use a builtin server, to
facilitate the deployment on workstations. Naturally, certain
optimisations, such as the per system loadbalancing is lost without the
use of a dedicated server.

\subsection{Integration}
The following components will be based on existing open source software, which
will be integrated with Equalizer. This integration enables a seamless
execution and end user experience. On one hand, by relying on the
Equalizer server for configuration, an visualization systems has only
one set of configuration files, instead of disjunct configurations for
each of the tools. On the other hand, the central resource management improves
the performance and interoperability between applications running on
the same system.

\subsubsection{Transparent Layer}
The transparent layer enables the execution of unmodified applications
on the cluster. Applications which do not run at the required
performance can then be addressed individually by porting them to
directly use the Equalizer programming framework. This transparent layer
will most likely be built using the Chromium\cite{Humphreys02}
framework. For usability reasons, a virtualization solution for the 2D
user interface, such as Xdmx X11 proxy server, will be integrated with
the Equalizer core and the transparent layer.

\subsubsection{Remote Visualization}
Remote visualization provides the capability of executing the
application on a powerful, central visualization system. The integration
with Equalizer allows for seamless integration of remote visualization,
and enables optimisations due to the knowledge gathered by the
framework during rendering.

\subsubsection{Scene Graphs}
The integration with scene graphs enables existing applications
using such a scene graph to quickly be able to take advantadge of
multipipe systems.

Two different approaches might be taken. In the first approach, each render node
holds a copy of the scene graph, and applies the changes made by the
application. In the second approach, the application sends high-level
rendering commands to the nodes, which together with a caching mechanism
can provide excellent performance.

\section{Milestones}

Much of the development of Equalizer is driven by the actual need of the
users, which are at the moment the software developers.

By the end of 2006, the programming framework will be developed to a
usable state for the development of commercial multipipe applications by
an experienced OpenGL programmer. This includes some scalability modes
to show the benefits of scalable rendering on commodity graphics clusters.

The transparent layer will be developed in 2007. The founding for this
effort is currently investigated. Interested parties are strongly
encouraged to contact us with proposals.

\subsection{Programming Framework}
The core part of the Equalizer project is the programming framework and
resource server. Naturally, these two components are the first major
milestone. Currently, they are usable to start the development of
parallel graphics applications. Both the server and the client library
miss major functionality, particularly decomposition modes.

\subsection{Transparent Layer}
The transparent layer is an important component, since it lowers the
barrier of using Equalizer and graphics clusters significantly. It will
be the next milestone to be addressed.

\subsection{Integration}
The integration of the other components will be driven by the actual
demand.

Remote visualization is an easy target, since it requires only the
integration of suitable compression for the image transport. Everything
else should be easily configurable using the standard Equalizer
capabilities.

The development of a distributed scene graph is relatively independent
from the Equalizer core development and might take the form of a
seperate project.

%\begin{figure}[ht]
%\centering
%\includegraphics[width=0.9\columnwidth]{images/sp_to_mp.pdf}
%\caption{Some advanced display systems}
%\label{FIG_parallel}
%\end{figure}

\bibliographystyle{abbrvurl}
\bibliography{paper}
\end{document}

### 2006-09-19 WhitePapers/ParallelRenderingSystems/paper.tex
Add project description document to web site, rewrite overview page, other minor updates to the web site
Add overview presentation, tweak layering figure in overview presentation and project white paper

@@
 % paragraph spacing -- should be rubber length
-%\setlength{\parskip}{1.5ex plus 0.5ex minus 0.5ex}
+%\setlength{\parskip}{\baselineskip}
 % paragraph indention for the first line
@@
 \vfill
-\begin{tabularx}{.85\columnwidth}{|l|l|X|}
+{\center\begin{tabularx}{.88\textwidth}{|l|l|X|}
 \hline
@@
 \hline
+0.5.2       & September 19, 2006   & Tweaked Figure~\ref{FIG_parallel}\\
+0.5.1       & July 5, 2006         & Added link to latest version\\
 0.5         & May 19, 2006         & Initial Version\\
 \hline
-\end{tabularx}
+\multicolumn{3}{c}{\footnotesize Latest version at 
+  \htmladdnormallink{http://www.equalizergraphics.com/documents/ParallelRenderingSystems.pdf}
+  {http://www.equalizergraphics.com/documents/ParallelRenderingSystems.pdf}}\\
+\end{tabularx}}
 \clearpage

### 2006-10-05 WhitePapers/ProjectEqualizer/paper.tex
minor documentation (project status) update

@@
 \hline
-0.1         & July 5, 2006   & Initial Draft\\
+0.2         & October 5, 2006 & update status\\
+0.1         & July 5, 2006    & initial draft\\
 \hline
@@
 
-By the end of 2006, the programming framework will be developed to a
-usable state for the development of commercial multipipe applications by
-an experienced OpenGL programmer. This includes some scalability modes
-to show the benefits of scalable rendering on commodity graphics clusters.
+Currently, the programming framework is usable for the development of
+commercial multipipe applications by an experienced OpenGL
+programmer. By the end of 2006, we will have implemented the core
+scalable rendering engine to show the benefits of scalable rendering on
+commodity graphics clusters.
 
@@
 milestone. Currently, they are usable to start the development of
-parallel graphics applications. Both the server and the client library
-miss major functionality, particularly decomposition modes.
+parallel graphics applications. Right now Equalizer supports multipipe
+rendering for display walls and immersive virtual reality installations,
+including stereo and head tracking.
 

### 2007-01-04 WhitePapers/ProjectEqualizer/paper.tex
website: Update project description and feature list

@@
 server, which controls the visualization system's configuration, a
-client side library for the development of scalable graphics software, a
+client-side library for the development of scalable graphics software, a
 transparent software layer to execute unmodified applications alongside
@@
 \hline
+0.3         & January 4, 2007 & update status wrt 0.2 release, some rewording\\
 0.2         & October 5, 2006 & update status\\
@@
 scene graphs offer better performance, but require the application
-to use a certain scene graph. The third approach, a parallel rendering
-framework offers a generic API for the development of parallel, scalable
-graphics applications.
+to use a certain scene graph and are therefore relatively invasive. The
+third approach, a parallel rendering framework offers a minimally
+invasive approach for the development of parallel, scalable graphics
+applications.
 
@@
 VGP from ModViz, SVN from IBM}. Distributed scene graphs are beginning
-to emerge\footnote{e.g., ScaleViz from Mercury Computer Systems} as
+to emerge\footnote{e.g., OpenSG, ScaleViz from Mercury Computer Systems} as
 well. There is currently no general programming framework to create
@@
 The Equalizer development is conducted in an open way. It is licensed
-under the LGPL open source license. Contributing parties, such as
+under the LGPL open source license, allowing the creation of commercial
+and open source applications using the framework, while ensuring a
+healthy community around the project. Contributing parties, such as
 hardware vendors, software developers and end users, have the following
@@
   the guarantee for software developers that the product will be
-  available indefinitely. Critical features or bugs, in case they are
+  available 'forever'. Critical features or bugs, in the event that they are
   not addressed by the maintainers, can still be addressed internally.
@@
 graphics applications, and thus facilitates the development of scalable
-multipipe rendering software. It addresses the process creation and
-synchronization, rendering task distribution, data transport, and the
-composition of the rendering results. The software developer plugs the
-application's rendering code into the framework. The Equalizer server
-deploys this rendering code according to the system's configuration and
-load. The programming framework is the client library to the Equalizer
-server, as well as the header files for this library.
+multipipe rendering software. Equalizer provides functionality in the
+area of process creation and synchronization, rendering task
+distribution, data transport, as well as the composition of the
+rendering results. The software developer plugs the application's
+rendering code into the framework. The Equalizer server deploys this
+rendering code according to the system's configuration and load. The
+programming framework is the client library to the Equalizer server, as
+well as the header files for this library.
 
@@
 Much of the development of Equalizer is driven by the actual need of the
-users, which are at the moment the software developers.
+users, which are at the moment the software developers. A current
+feature list can be found in \cite{Features}.
 
@@
 commercial multipipe applications by an experienced OpenGL
-programmer. By the end of 2006, we will have implemented the core
-scalable rendering engine to show the benefits of scalable rendering on
-commodity graphics clusters.
+programmer. Version 0.2 added support for scalable rendering to show the
+benefits of scalable rendering on commodity graphics clusters and shared
+memory systems.
 
-The transparent layer will be developed in 2007. The founding for this
-effort is currently investigated. Interested parties are strongly
-encouraged to contact us with proposals.
+For 2007, the main areas we plan to address are the extension of the
+resource server for multi-application usage, general reliability and
+performance improvements, a Windows port, as well as the integration of
+a transparent layer. The founding for this effort is currently
+investigated. Interested parties are strongly encouraged to contact us
+with proposals.
 
@@
 rendering for display walls and immersive virtual reality installations,
-including stereo and head tracking.
+including stereo and head tracking, as well as a wide range of scalable
+rendering features.
 

### 2007-01-11 WhitePapers/ParallelRenderingSystems/paper.tex
website: update api page, update parallel rendering systems WP

@@
 \hline
+0.5.3       & January 11, 2007     & Updated Figure~\ref{FIG_multipipe}\\
 0.5.2       & September 19, 2006   & Tweaked Figure~\ref{FIG_parallel}\\
@@
 \caption{Some advanced display systems}
-\label{FIG_parallel}
+\label{FIG_multipipe}
 \end{figure}

### 2007-01-11 WhitePapers/ParallelRenderingSystems/paper.tex
Rework parallel rendering WP

--- 2006-09-19 WhitePapers/ParallelRenderingSystems/paper.tex
@@
   This White Paper analyzes and classifies the different approaches taken
-to parallel rendering. The purpose is to clarify common
+  to parallel, interactive rendering. The purpose is to clarify common
   misconceptions and false expectations of the capabilities of the
@@
   We advocate the need for a generic, open parallel rendering framework to
-build scalable graphics software. The inherent limitations of transparent
-multi-pipe software are justified, and a full software stack for
-graphics clusters is outlined.
+  build scalable graphics software. The drawbacks of other existing
+  solutions are outlined, and a full software stack for graphics clusters
+  is proposed. The Equalizer project\footnote{
+    \htmladdnormallink{http://www.equalizergraphics.com}
+    {http://www.equalizergraphics.com/}} aims to provide such an
+  implementation, combining existing software packages with newly
+  developed middleware.
 }
+
 \vfill
@@
 \hline
+0.6         & January 11, 2007     & Minor rework of the whole paper\\
 0.5.2       & September 19, 2006   & Tweaked Figure~\ref{FIG_parallel}\\
@@
 contributes to the slow adoption of graphics clusters and virtual reality
-installations due to the application performance and (un-)availability.
+installations due to bad application performance and unavailability.
 
@@
 % Introduce the topic
-Application developers are faced with making their applications
+Application developers are typically faced with making their applications
 aware of multiple graphics cards for two reasons -- either the
-application has to run on advanced display systems, or the rendering has
-to be accelerated by using parallel rendering on multiple graphics cards.
+application has to run on multipipe display systems, or the rendering has
+to be accelerated by using parallel rendering on multiple graphics
+cards. The two reasons are not mutually exclusive.
 
 \subsubsection{Multi-View Rendering}
-The first motivation is often to run the application on advanced display
-systems: workstations with multiple monitors, walls build out of
-multiple displays or projectors as well as immersive environments. Such
-installations need more than one graphics card to drive all the display
-systems. Therefore the rendering has to be distributed across a
-number of cards and often across multiple systems. The outputs of the
-displays has to be synchronized spatially and time-wise to provide one
-coherent image. In immersive environments, stereo rendering is a requirement.
+The first motivation is often to run the application on advanced,
+multipipe display systems: workstations with multiple monitors, walls
+build out of multiple screens or projectors as well as immersive
+environments. Such installations need more than one graphics card to
+drive all the display systems. Therefore the rendering has to be
+distributed across a number of cards, and often across multiple
+systems. The outputs of the displays has to be synchronized spatially
+and time-wise to provide one coherent image. In immersive environments,
+active or passive stereo rendering is a requirement.
 
@@
 multiple resources, because the application hits limitations of a single
-graphics system.
+graphics system. Figure~\ref{FIG_pipeline} shows a simplified rendering
+pipeline and the resources used by during rendering. The typical
+performance bottlenecks in such a pipeline are:
 
-\begin{figure}[ht]
+\begin{figure}[htb]
 \centering
@@
 
-Figure~\ref{FIG_pipeline} shows a simplified rendering pipeline and the
-resources used by during rendering. The typical performance bottlenecks
-in such a pipeline are:
 \begin{description}
@@
 
-\section{Transparent Multi-Pipe Rendering Software}
+\section{Transparent Multipipe Rendering Software}
 % Technical Overview
@@
 % Parallel Extensions
-Some parallel rendering frameworks support parallelization of the
+Some transparent rendering packages support parallelization of the
 application, that is, the application is modified and renders using
@@
 From the applications point of view, this approach is similar to the
-transparent solutions, in that the rendering is parallelized. The big
-difference is that a distributed scene graph approaches the problem at a
-much higher level, which allows significant optimisations and leads to
-better performance. In contrast to transparent rendering software, the
-traversal to generate the rendering commands is parallelized, and the
-commands are send --in parallel-- directly to the graphics card. The
-scene graph frees the application developer of the task to parallelize
-the rendering.
+transparent solutions, in that the rendering is parallelized by some
+'magic' software. The big difference is that a distributed scene graph
+approaches the problem at a much higher level, which allows significant
+optimizations and leads to better performance. In contrast to
+transparent rendering software, the traversal to generate the rendering
+commands is parallelized, and the commands are send --in parallel--
+directly to the graphics card. The scene graph frees the application
+developer of the task to parallelize the rendering.
 
@@
 libraries on the application. By solving the common problems of
-multi-pipe rendering, the application developer can focus on
+any multiple application, the application developer can focus on
 application-specific problems. Parallel rendering frameworks address
@@
 application, regardless of the used rendering software. Refactoring the
-application, other then strictly needed for multi-pipe rendering, is not
-necessary. The application provides entry points for its rendering
+application, other then necessary for multipipe rendering, is not
+needed. The application provides entry points for its rendering
 functions. Depending on the current configuration, the framework creates
-the necessary rendering threads which are initialised with a rendering
+the necessary rendering threads which are initialized with a rendering
 context by the application's init functions. When the application
@@
 applications, which cannot be modified, it is the only possibility to make
-them multi-pipe ready. The low porting effort is a strong benefit for
+them multipipe-ready. The low porting effort is a strong benefit for
 transparent rendering software.
@@
 Parallel rendering frameworks are the common middleware for any
-multi-pipe application. Distributed scene graphs and transparent
+multipipe application. Distributed scene graphs and transparent
 rendering software can use it as the base for their
@@
 encountered when doing such a port and follows the natural programming
-model for any multi-pipe application. For high-performance applications,
+model for any multipipe application. For high-performance applications,
 a parallel rendering framework is the only solution short of a
@@
 performance-uncritical and legacy applications, as well as applications
-already written using a specific scene graph.
+already written using a specific scene graph. Figure~\ref{FIG_parallel}
+illustrates the proposed software stack.
 
@@
 \includegraphics[width=0.45\columnwidth]{images/layers.pdf}
-\caption{Ideal software layering for a visualization system}
+\caption{Proposed software stack for a visualization system}
 \label{FIG_parallel}
@@
 distributed scene graphs are often too invasive since they require the
-applications to use a certain format --the scene graph-- to decribe their
-data.
+applications to use a certain format and API --the scene graph-- to
+decribe their data.
 
@@
 
-\subsection{Transparent Multi-Pipe Rendering Software}
+\subsection{Transparent Multipipe Rendering Software}
 \subsubsection{Chromium}
@@
 OpenGL Multipipe is a transparent software solution delivered with SGI's
-multi-pipe machines. It is not sold separately. For building parallel
+multipipe machines. It is not sold separately. For building parallel
 applications, SGI offered OpenGL Multipipe SDK.
@@
 
+\subsubsection{OpenSG}
+OpenSG is an open source scenegraph and provides various functionality
+for parallel rendering, including network data distribution and scalable
+rendering modes.
+
 \subsection{Parallel Rendering Frameworks}
@@
 framework for the development of scalable graphics applications. It does
-not support distributed rendering.
+not support distributed rendering, and is not actively developed anymore.
 
@@
 Equalizer is an open source parallel rendering framework under development
-by the University of Z\"urich and others. Equalizer will
-support multi-view and scalable rendering, together with a resource
-management system and a transparent rendering layer.
+by the University of Z\"urich and others. Equalizer supports multi-view
+and scalable rendering, and a resource management system and a
+transparent rendering layer is planned.
 

### 2007-03-13 WhitePapers/ProjectEqualizer/paper.tex
[ ] May break build [ ] Breaks existing applications [ ] Bugfix [ ] New Feature [ ] Cleanup [ ] Optimization [x] Documentation
Description: Update use cases web page

@@
   % what
-Equalizer is a project to develop software to simplify the creation of scalable
-graphics applications and to improve the usability of multipipe
-visualization systems. The main components of Equalizer are a resource
-server, which controls the visualization system's configuration, a
-client-side library for the development of scalable graphics software, a
-transparent software layer to execute unmodified applications alongside
-with scalable applications, as well as remote visualization
-capabilities. Equalizer is designed to support all kinds of
-visualization systems ranging from laptops, workstations, multipipe
+  Equalizer is a project to develop software to simplify the creation of
+  scalable graphics applications and to improve the usability of
+  multipipe visualization systems. The main components of Equalizer are
+  a resource server, which controls the visualization system's
+  configuration, a client-side library for the development of scalable
+  graphics software, a transparent software layer to execute unmodified
+  applications alongside with scalable applications, as well as remote
+  visualization capabilities. Equalizer is designed to support all kinds
+  of visualization systems ranging from laptops, workstations, multipipe
   shared memory systems to graphics clusters.
@@
   hardware vendors, software developers, research institutions and end
-users. Interested parties are strongly encouraged to contact us. We are
-looking for contributions or founding for several components of Equalizer.
+  users. Interested parties are strongly encouraged to contact us. We
+  are looking for contributions or founding for several components of
+  Equalizer. 
 }
-\vfill
-{\center\begin{tabularx}{.88\textwidth}{|l|l|X|}
+\vfill {\center\begin{tabularx}{.88\textwidth}{|l|l|X|}
     \hline
@@
     0.1         & July 5, 2006    & initial draft\\
-\hline
-\multicolumn{3}{c}{\small Latest version at 
+    \hline \multicolumn{3}{c}{\small Latest version at
       \htmladdnormallink{http://www.equalizergraphics.com/documents/ProjectEqualizer.pdf}
@@
 solution for applications which cannot or need not to be modified, but
-are limited in performance and compatibility. Distributed
-scene graphs offer better performance, but require the application
-to use a certain scene graph and are therefore relatively invasive. The
-third approach, a parallel rendering framework offers a minimally
-invasive approach for the development of parallel, scalable graphics
-applications.
+are limited in performance and compatibility. Distributed scene graphs
+offer better performance, but require the application to use a certain
+scene graph and are therefore relatively invasive. The third approach, a
+parallel rendering framework offers a minimally invasive approach for
+the development of parallel, scalable graphics applications.
 
 Today we have satisfying, though often proprietary, solutions for the
-transparent approach\footnote{e.g., Chromium from Stanford University,
-VGP from ModViz, SVN from IBM}. Distributed scene graphs are beginning
-to emerge\footnote{e.g., OpenSG, ScaleViz from Mercury Computer Systems} as
-well. There is currently no general programming framework to create
-parallel graphics applications. Furthermore, all existing solutions are
-'islands', that is, they are configured separately and do possibly
-interfere with each other when running simultanously on the same
-system. Equalizer addresses these issues.
+transparent approach\footnote{Chromium from Tungsten Graphcis/Stanford
+  University, VGP from ModViz, SVN from IBM}. Distributed scene graphs
+are beginning to emerge\footnote{e.g., OpenSG, ScaleViz from Mercury
+  Computer Systems} as well. There is currently no general programming
+framework to create parallel graphics applications. Furthermore, all
+existing solutions are 'islands', that is, they are configured
+separately and do possibly interfere with each other when running
+simultanously on the same system. The goal of Equalizer is to address
+these issues.
 
-Unlike the HPC community, visualization software developers do not
-widely address performance bottlenecks by parallelizing
-their applications. One of the reasons for the lack of parallel graphics
+Unlike the HPC community, visualization software developers do not yet
+widely address performance bottlenecks by parallelizing their
+applications. One of the reasons for the lack of parallel graphics
 applications is the lack of a standard programming interface, which
 implements specific knowledge needed to create scalable visualization
-applications. Equalizer provides a programming framework for high
+applications. Equalizer provides such a programming framework for high
 performance visualization.
@@
 Equalizer is a project to create new software to facilitate the
-development and usage of software for multipipe
-graphics systems by addressing the shortcomings of the current
-software situation. It adds the two missing components, a
-parallel programming framework and a resource management system, and
-integrates existing solutions, such as a transparent layer, scene graphs
-and remote visualization capabilities.
+development and usage of software for multipipe graphics systems by
+addressing the shortcomings of the current software situation. It adds
+the two missing components, a parallel programming framework and a
+resource management system, and will integrate existing solutions, such
+as a transparent layer, scene graphs and remote visualization
+capabilities.
 
-Equalizer addresses all visualization systems. An Equalizer application
-can run unmodified on a laptop, a multipipe workstation, a graphics
-cluster or a multipipe shared memory visualization server.
+Equalizer addresses all types of visualization systems. An Equalizer
+application can run unmodified on a laptop, a multipipe workstation, a
+graphics cluster or a multipipe shared memory visualization server.
 
@@
   application is outsourced to the Equalizer framework, thus freeing the
-  software developers to reimplement basic algorithms. The expertise for
-  high performance visualization is partly embedded into Equalizer, and
-  has not to be developed inhouse. The open development approach ensures
-  that Equalizer delivers the functionality most needed by the community.
+  software developers to reimplement basic parallel rendering
+  algorithms. The expertise for high performance visualization is partly
+  embedded into Equalizer, and has not to be developed in house. The
+  open development approach ensures that Equalizer delivers the
+  functionality most needed by the community.
 \item[Higher Functionality] Additional features, which would not be
   cost-effective to implement in the application, are provided by the
-  Equalizer framework. New features will become available over time,
-  and will often be usable by the application with no or minimal code
+  Equalizer framework. New features will become available over time, and
+  will often be usable by the application with no or minimal code
   changes.
@@
   the guarantee for software developers that the product will be
-  available 'forever'. Critical features or bugs, in the event that they are
-  not addressed by the maintainers, can still be addressed internally.
+  available 'forever'. Critical features or bugs, in the event that they
+  are not addressed by the maintainers, can still be addressed
+  internally.
 \end{description}
@@
 
-The programming framework is the main contribution of the
-Equalizer project. It addresses common problems when creating parallel
-graphics applications, and thus facilitates the development of scalable
-multipipe rendering software. Equalizer provides functionality in the
-area of process creation and synchronization, rendering task
-distribution, data transport, as well as the composition of the
-rendering results. The software developer plugs the application's
-rendering code into the framework. The Equalizer server deploys this
-rendering code according to the system's configuration and load. The
-programming framework is the client library to the Equalizer server, as
-well as the header files for this library.
+The programming framework is the main contribution of the Equalizer
+project. It addresses common problems when creating parallel graphics
+applications, and thus facilitates the development of scalable multipipe
+rendering software. Equalizer provides functionality in the area of
+process creation and synchronization, rendering task distribution, data
+transport, as well as the composition of the rendering results. The
+software developer plugs the application's rendering code into the
+framework. The Equalizer server deploys this rendering code according to
+the system's configuration and load. The programming framework is the
+client library to the Equalizer server, as well as the header files for
+this library.
 
@@
 The core of the resource management system is the Equalizer server,
-which is managing the system's configuration. It is configured to know the
-system resources and schedules the execution of rendering tasks on these
-resources. Equalizer applications provide a render client\footnote{The
-  render client can be the same executable as the application.}, which
-contains the application's rendering code. This render client is
-instanciated on the render nodes by the Equalizer server. The
-transparent layer and distributed scene graphs are technically an
-Equalizer application.
+which is managing the system's configuration. It is configured to know
+the system resources and schedules the execution of rendering tasks on
+these resources. Equalizer applications provide a render
+client\footnote{The render client can be the same executable as the
+  application.}, which contains the application's rendering code. This
+render client is instanciated on the render nodes by the Equalizer
+server. The transparent layer and distributed scene graphs are
+technically an Equalizer application.
 
 For standalone usage on simple machines, the Equalizer client library
-will be able to automatically configure and use a builtin server, to
+will be able to automatically configure and use a built-in server, to
 facilitate the deployment on workstations. Naturally, certain
-optimisations, such as the per system loadbalancing is lost without the
-use of a dedicated server.
+optimisations, such as the per-system loadbalancing is lost without the
+use of a dedicated, system-wide server.
 
@@
 execution and end user experience. On one hand, by relying on the
-Equalizer server for configuration, an visualization systems has only
+Equalizer server for configuration, a visualization systems has only
 one set of configuration files, instead of disjunct configurations for
@@
 with Equalizer allows for seamless integration of remote visualization,
-and enables optimisations due to the knowledge gathered by the
-framework during rendering.
+and enables optimisations due to the knowledge gathered by the framework
+during rendering.
 
 \subsubsection{Scene Graphs}
-The integration with scene graphs enables existing applications
-using such a scene graph to quickly be able to take advantadge of
-multipipe systems.
+The integration with scene graphs enables existing applications using
+such a scene graph to quickly be able to take advantadge of parallel
+rendering on multipipe systems.
 
-Two different approaches might be taken. In the first approach, each render node
-holds a copy of the scene graph, and applies the changes made by the
-application. In the second approach, the application sends high-level
-rendering commands to the nodes, which together with a caching mechanism
-can provide excellent performance.
+Two different approaches might be taken. In the first approach, each
+render node holds a copy of the scene graph, and applies the changes
+made by the application. In the second approach, the application sends
+high-level rendering commands to the nodes, which together with a
+caching mechanism can provide excellent performance.
 
@@
 commercial multipipe applications by an experienced OpenGL
-programmer. Version 0.2 added support for scalable rendering to show the
-benefits of scalable rendering on commodity graphics clusters and shared
-memory systems.
+programmer. Version 0.2 added support for scalable rendering, showing
+the benefits of scalable rendering on commodity graphics clusters and
+shared memory systems.
 

### 2007-08-19 Developer/ProgrammingGuide/paper.tex
Start Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\NeedsTeXFormat{LaTeX2e}
\documentclass[10pt,a4]{scrartcl}

\usepackage{tabularx,graphicx,a4}

\ifx\pdfoutput\undefined
  % We're not running pdftex
  % european (better) fonts -- does not look good with pdflatex
  \usepackage[T1]{fontenc}
  \newcommand{\href}[2]{#2\\{\hspace*{5mm}\scriptsize <#1>}\\}
\else
  \pdfcompresslevel=9
  \def\pdfBorderAttrs{/Border [0 0 0] } % No border around Links
  \usepackage{hyperref}
\fi

\title{Equalizer Programming Guide}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.1, \today
}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}

% suppress  single floating lines on top (widow) and bottom(club)
%  10000 is infinity
%  tradeoff: maybe underfull vboxes
\clubpenalty=10000
\widowpenalty=10000

\begin{document}

\maketitle
\vfill

\thispagestyle{empty}
\begin{figure}[ht]
  \centering
  \includegraphics[width=4cm]{logo.pdf}
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill

%\abstract{
%  Equalizer is a project to develop software to simplify the creation of
%  scalable graphics applications and to improve the usability of
%  multipipe visualization systems.
%}

\vfill {\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline \multicolumn{3}{c}{\small
      \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
      {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}\\
  \end{tabularx}}

\clearpage
\tableofcontents
\thispagestyle{empty}

\clearpage
\pagenumbering{arabic}

\section{Introduction}

Equalizer provides a framework for the development of parallel OpenGL
applications. Equalizer-based applications can run a single
shared-memory system with multiple graphics cards (GPU's) or on a
distributed graphics cluster. This Programming Guide introduces the
programming interface using the \textsf{eqPly} example shipped with Equalizer.

Any questions related to Equalizer programming and this Programming
Guide should be directed to the \textsf{eq-dev} mailing
list\footnote{see
  \htmladdnormallink{http://www.equalizergraphics.com/lists.html}
  {http://www.equalizergraphics.com/lists.html}}.

\section{Getting Started}

\subsection{Compiling and running \textsf{eqPly}}

A prerequisite for this Programming Guide is a working \textsf{eqPly}
example. The Quickstart
Guide\footnote{\htmladdnormallink{http://www.equalizergraphics.com/documents/EqualizerGuide.html}{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
explains how to run it. \textsf{eqPly} can also be executed without a
server, which simplifies the development cycle. In this case it will be
configured to use one window.

\subsection{Equalizer Processes}

\subsubsection{The Server}

An Equalizer server is responsible for managing one visualization
system\footnote{a shared memory system or graphics cluster}. Currently
it is only useful for running one application at a time, but it will be
extended to support multiple applications concurrently and efficiently
on one system. The server controls and potentially launches the
application's rendering clients.

\subsubsection{The Application}

The application connects to a server, which chooses a configuration for
the application. It provides a render client, to be launched by the
server. The application reacts on events and controls the rendering.

\subsubsection{The Render Client}

The render client implements the rendering part of an application. It is
passive, and receives all its rendering tasks from the server. The tasks
are executed by calling the appropriate task methods (see
\ref{ssTaskMethods}).

The application might be a rendering client, in which case it can also
contribute to the rendering. It can choose not to implement any render
client-related code, in which case it is reduced to be the application's
'master' process without any OpenGL windows.

The rendering client can be the same executable as the application, as
is the case with \textsf{eqPly}. Real-world applications often implement
a separate, light-weight rendering client.

\section{The Programming Interface}

Equalizer uses a C++ programming interface. The API is minimally
invasive, that is, Equalizer imposes only the minimal, natural execution
framework upon the application. It does not impose a scene graph or does
interfere in any way with the application's rendering code.

\subsection{\label{ssTaskMethods}Task Methods}

The application subclasses Equalizer objects and overrides virtual
functions to implement certain functionality, e.g., the OpenGL rendering
code in \textsf{eq::Channel::frameDraw}. These task methods are in
concept similar to C function callbacks.

\section{The Resource Tree}

The rendering resources are represented in a hierarchical tree structure
which corresponds to the physical and logical resources found in a 3D
rendering environment.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\columnwidth]{desk.pdf}{\caption{\label{fConfig}An
      example configuration}}
\end{figure}

\fig{fConfig} shows one example configuration, running on a single
machine (node) using one graphics card (pipe) with one window to render
to two output channels connected to a L-shaped projection table. The
compound description is only used by the server to compute the rendering
tasks. The application is not aware of compounds, and does not need to
concern itself with the parallel rendering logics of a configuration.

For testing and development purposes it is possible to use multiple
instances for one resource, e.g., to run multiple render client nodes on
one computer. For deployment one node and pipe should be used for each
computer and graphics card, respectively.

\subsection{Configuration}

The root of the resource tree is the \textsf{eq::Config}, which
represents the current configuration of the application. It currently
only holds the local node, not all nodes of the configuration.

\subsection{Node}

An \textsf{eq::Node} is the representation of a single computer in the
system. It is one operating system process of the render client. All
node task methods are executed from the main application thread.

\subsection{Pipe}

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU). In
the current implementation it is also one operating system thread,
unless the pipe's thread hint is set to \textsf{false}. All pipe and
child window and channel task methods are executed from the pipe thread
for threaded pipes or from the main application thread for non-threaded
pipes\footnote{see also
  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/nonthreaded.html}{http://www.equalizergraphics.com/documents/design/nonthreaded.html}
  for non-threaded rendering}.

Further versions of Equalizer might introduce threaded windows, where
all window-related task methods are executed in a separate operating
system thread.

\subsection{Window}

An \textsf{eq::Window} is an drawable and OpenGL context. The drawable
can be an on-screen window or an off-screen PBuffer or
FBO\footnote{off-screen drawables are not yet implemented, but can be
  created by the application and used with Equalizer}.

\subsection{Channel}

The \textsf{eq::Channel} is the abstraction of an OpenGL viewport within
its parent window. It is the entity executing the actual rendering.

\bibliographystyle{abbrvurl}
\bibliography{paper}
\end{document}

### 2007-08-21 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-\usepackage{tabularx,graphicx,a4}
+\usepackage{tabularx,graphicx,a4,listings}
 
@@
   \textbf{INCOMPLETE}\\[\medskipamount]
-  Version 0.1, \today
+  Version 0.2, \today
 }
@@
 \vfill
+\lstset{language=C++}
 
@@
 
-\vfill {\center\begin{tabularx}{\textwidth}{|l|l|X|}
+\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
     \hline
@@
     \hline
+    0.2         & Aug 20, 2007 & main function described\\
     0.1         & Aug 19, 2007 & outlined the basic concepts\\
@@
 The application subclasses Equalizer objects and overrides virtual
-functions to implement certain functionality, e.g., the OpenGL rendering
-code in \textsf{eq::Channel::frameDraw}. These task methods are in
-concept similar to C function callbacks.
+functions to implement certain functionality, e.g., the application's
+OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
+methods are in concept similar to C function callbacks. The
+\textsf{eqPly} section will discuss the most important task methods. A
+full list of all task methods can be found on the
+website\footnote{\htmladdnormallink{http://www.equalizergraphics.com/documents/design/taskMethods.html}{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.
 
-\section{The Resource Tree}
+\subsection{The Resource Tree}
 
@@
 
-\subsection{Configuration}
+\subsubsection{Configuration}
 
@@
 
-\subsection{Node}
+\subsubsection{Node}
 
@@
 
-\subsection{Pipe}
+\subsubsection{Pipe}
 
@@
 pipes\footnote{see also
-  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/nonthreaded.html}{http://www.equalizergraphics.com/documents/design/nonthreaded.html}
-  for non-threaded rendering}.
+  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/nonthreaded.html}{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.
 
@@
 
-\subsection{Window}
+\subsubsection{Window}
 
@@
 
-\subsection{Channel}
+\subsubsection{Channel}
 
@@
 
-\bibliographystyle{abbrvurl}
-\bibliography{paper}
+\subsection{Resource Usage}
+
+How the rendering resources are to be used is configured using a
+compound tree. Each compound has a channel, which it uses to execute the
+rendering tasks. The rendering tasks are computed by the server and send
+to the render clients. At no point the application or render clients
+have or need knowledge of compounds. The configuration of compounds is
+not in the scope of this document\footnote{see
+  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/compounds.html}{http://www.equalizergraphics.com/documents/design/compounds.html}}.
+
+\section{The \textsf{eqPly} polygonal renderer}
+
+The \textsf{eqPly} example is shipped with the Equalizer distribution
+and serves as a simple reference implementation of an Equalizer-based
+application. Its focus is not on rendering features or visual quality.
+It serves as a test bed for most of the Equalizer features.
+
+In this section the source code of \textsf{eqPly} is discussed in
+detail, and relevant design decision and remarks are raised.
+
+All classes in the example are in the \textsf{eqPly} namespace to avoid
+type name ambiguities, in particular for the \textsf{Window} class.
+
+\subsection{The main Function}
+
+The main function starts off with parsing the command line into the
+\textsf{LocalInitData} data structure, which in part will be distributed
+to all render client nodes. For actual command line parsing is done by
+the \textsf{LocalInitData} class and will be discussed there:
+
+{\small\begin{lstlisting}
+int main( int argc, char** argv )
+{
+    // 1. parse arguments
+    eqPly::LocalInitData initData;
+    initData.parseArguments( argc, argv );
+\end{lstlisting}}
+
+The second step is to initialize the Equalizer library. The
+initialization function of Equalizer also parses the command line, which
+is used to set certain default values based on Equalizer-specific
+options\footnote{Equalizer-specific options always start with --eq-},
+e.g., the default server location. Furthermore, a node factory is
+provided:
+
+{\small\begin{lstlisting}
+    // 2. Equalizer initialization
+    NodeFactory nodeFactory;
+    if( !eq::init( argc, argv, &nodeFactory ))
+    {
+        EQERROR << "Equalizer init failed" << endl;
+        return EXIT_FAILURE;
+    }
+\end{lstlisting}}
+
+The node factory is used by Equalizer to create the object instances for
+the rendering entities. Each of the classes inherits from the same type
+provided by Equalizer in the \textsf{eq} namespace. The provided
+\textsf{eq::NodeFactory} base class instantiates a 'plain' Equalizer
+object, thus making it possible to selectively subclass individual
+entity types. For each rendering resource used in the configuration, one
+C++ object will be created:
+
+{\small\begin{lstlisting}
+class NodeFactory : public eq::NodeFactory
+{
+public:
+    virtual eq::Config*  createConfig()  { return new eqPly::Config; }
+    virtual eq::Node*    createNode()    { return new eqPly::Node; }
+    virtual eq::Pipe*    createPipe()    { return new eqPly::Pipe; }
+    virtual eq::Window*  createWindow()  { return new eqPly::Window; }
+    virtual eq::Channel* createChannel() { return new eqPly::Channel; }
+};
+\end{lstlisting}}
+
+The third step is to create an instance of the application and to
+initialize it locally. The application is an \textsf{eq::Client}, which
+is an \textsf{eqNet::Node}. The underlying network distribution in
+Equalizer is a peer-to-peer network structure of
+\textsf{eqNet::Node}s. The application programmer rarely is aware of the
+classes in the \textsf{eqNet} namespace, but both the
+\textsf{eq::Client} and the server are \textsf{eqNet::Node}s. The local
+initialization of nodes creates a local listening socket, so that the
+node, and therefore the \textsf{eq::Client} can communicate over the
+network with other nodes, such as the server and the rendering clients.
+
+{\small\begin{lstlisting}
+    // 3. initialization of local client node
+    RefPtr< eqPly::Application > client = new eqPly::Application( initData );
+    if( !client->initLocal( argc, argv ))
+    {
+        EQERROR << "Can't init client" << endl;
+        eq::exit();
+        return EXIT_FAILURE;
+    }
+\end{lstlisting}}
+
+Finally everything is set up to run the \textsf{eqPly} application:
+
+{\small\begin{lstlisting}
+    // 4. run client
+    const int ret = client->run();
+\end{lstlisting}}
+
+After it has finished, the application and Eqply is deinitialized and
+the \textsf{main} function returns:
+
+{\small\begin{lstlisting}
+    // 5. cleanup and exit
+    client->exitLocal();
+    client = 0;
+
+    eq::exit();
+    return ret;
+}
+\end{lstlisting}}
+
+\subsection{Application}
+
+In the \textbf{eqPly} case, the application is also the render client. It has
+three run-time behaviours:
+
+\begin{enumerate}
+  \item \textbf{Application}: The executable started by the user, which
+    is the controlling entity in the rendering session.
+  \item \textbf{Auto-launched Render Client}: The typical render client,
+    started by the server. In this mode, \textsf{initLocal} never
+    returns and the process is terminated by the server.
+  \item \textbf{Resident Render Client}: Manually pre-started render
+    client, listening on a specified port for server commands.
+\end{enumerate}
+
 \end{document}

### 2007-08-26 Developer/ProgrammingGuide/paper.tex
Update Programming Guide, fix minor issue in website footer, add link to previous polls.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   \textbf{INCOMPLETE}\\[\medskipamount]
-  Version 0.2, \today
+  Version 0.3, \today
 }
@@
 \begin{figure}[ht]
-  \centering
-  \includegraphics[width=4cm]{logo.pdf}
+  \hfill
+  \includegraphics[width=4cm]{logo.pdf}\hfill
+  \includegraphics[width=4cm]{vmml.pdf}\hfill\vspace{-1em}\\
 %  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
@@
     \hline
+    0.3         & Aug 26, 2007 & application and render client code described\\
     0.2         & Aug 20, 2007 & main function described\\
@@
   \centering
-  \includegraphics[width=\columnwidth]{desk.pdf}{\caption{\label{fConfig}An
+  \includegraphics[width=\columnwidth]{cave.pdf}{\caption{\label{fConfig}An
       example configuration}}
@@
 
-\fig{fConfig} shows one example configuration, running on a single
-machine (node) using one graphics card (pipe) with one window to render
-to two output channels connected to a L-shaped projection table. The
-compound description is only used by the server to compute the rendering
+\fig{fConfig} shows one example configuration for a four-side
+CAVE\texttrademark, running on two machines (node) using three graphics
+cards (pipe) with one window each to render to the four output channels
+connected to the projectors for each of the walls. The compound
+description is only used by the server to compute the rendering
 tasks. The application is not aware of compounds, and does not need to
@@
 e.g., the default server location. Furthermore, a node factory is
-provided:
+provided. The \textsf{EQERROR} macro, and its counterparts
+\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
+debugging outputs with various logging levels:
 
@@
     is the controlling entity in the rendering session.
-  \item \textbf{Auto-launched Render Client}: The typical render client,
-    started by the server. In this mode, \textsf{initLocal} never
-    returns and the process is terminated by the server.
-  \item \textbf{Resident Render Client}: Manually pre-started render
-    client, listening on a specified port for server commands.
+  \item \textbf{Auto-launched render client}: The typical render client,
+    started by the server. The server starts the executable with special
+    parameters, which cause \textsf{Client::initLocal} to never
+    return. During exit, the server terminates the process.
+  \item \textbf{Resident render client}: Manually pre-started render
+    client, listening on a specified port for server commands. This mode
+    is selected using the command-line option \textsf{--eq-client} and
+    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \htmladdnormallink{http://www.equalizergraphics.com/documents/design/residentNodes.html}{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
 \end{enumerate}
 
+\subsubsection{Main Loop}
+
+The application's main loop starts by connecting the application to an
+Equalizer server. The command line parameter \textsf{--eq-server}
+explicitly specifies a server address. If no server was specified,
+\textsf{Client::connectServer} try first to connect to a server on the
+local machine using the default port 4242. If that fails, it will create
+a server running within the applications process with a default
+1-channel configuration\footnote{see
+  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/standalone.html}{http://www.equalizergraphics.com/documents/design/standalone.html}}:
+
+{\small\begin{lstlisting}
+int Application::run()
+{
+    // 1. connect to server
+    RefPtr<eq::Server> server = new eq::Server;
+    if( !connectServer( server ))
+    {
+        EQERROR << "Can't open server" << endl;
+        return EXIT_FAILURE;
+    }
+\end{lstlisting}}
+
+The second step is to ask the server for a configuration. The
+\textsf{ConfigParams} are a placeholder for later implementations to
+provide additional hints and information to the server for choosing a
+configuration. The configuration is created using
+\textsf{NodeFactory::createConfig}. Therefore it is of type
+\textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
+making q cast necessary:
+
+{\small\begin{lstlisting}
+    // 2. choose config
+    eq::ConfigParams configParams;
+    Config* config = static_cast<Config*>(server->chooseConfig( configParams ));
+
+    if( !config )
+    {
+        EQERROR << "No matching config on server" << endl;
+        disconnectServer( server );
+        return EXIT_FAILURE;
+    }
+\end{lstlisting}}
+
+Finally it is time to initialize the configuration. For statistics, the
+time for this operation is measures and printed. During initialization
+the server launches and connects all render client nodes, and calls the
+appropriate initialization task methods, as explained in later
+sections. \textsf{Config::init} does return after all nodes, pipes,
+windows and channels are initialized. It returns \textsf{true} only if
+all init task methods were successful. The \textsf{EQLOG} macro allows
+topic-specific logging. The numeric topic values are specified in the
+respective \textsf{log.h} header files:
+
+{\small\begin{lstlisting}
+    // 3. init config
+    eqBase::Clock clock;
+
+    config->setInitData( _initData );
+    if( !config->init( ))
+    {
+        EQERROR << "Config initialization failed: " 
+                << config->getErrorMessage() << endl;
+        server->releaseConfig( config );
+        disconnectServer( server );
+        return EXIT_FAILURE;
+    }
+
+    EQLOG( eq::LOG_CUSTOM ) << "Config init took " << clock.getTimef() << " ms"
+                            << endl;
+\end{lstlisting}}
+
+When the configuration was successfully initialized, the actual main
+loop is executed. The main loop runs until the user exits the
+configuration or a maximum number of frames, specified as a command-line
+argument, has been rendered. The latter is useful for benchmarks. The
+\textsf{Clock} is reused for measuring the overall performance. A new
+frame is started using \textsf{Config::startFrame} and a frame is
+finished using \textsf{Config::endFrame}.
+
+When the frame is started, the server computes all rendering tasks and
+sends them to the appropriate render client nodes. The render client
+nodes dispatch the tasks to the correct node or pipe thread, were they
+are executed in the order they arrive.
+
+The frame finished is \textsf{latency} frames behind the frame
+started. The latency is specified in the configuration file, and allows
+several outstanding frames whose tasks are already queued in the node
+and pipe threads for execution. This allows overlapped execution and
+minimizes idle times.
+
+When the main loop is finished, \textsf{Config::finishAllFrames} catches
+up with the latency. It returns after all outstanding frames have been
+rendered:
+
+{\small\begin{lstlisting}
+    // 4. run main loop
+    uint32_t maxFrames = _initData.getMaxFrames();
+
+    clock.reset();
+    while( config->isRunning( ) && maxFrames-- )
+    {
+        config->startFrame();
+        // config->renderData(...);
+        config->finishFrame();
+    }
+    const uint32_t frame = config->finishAllFrames();
+    const float    time  = clock.getTimef();
+    EQLOG( eq::LOG_CUSTOM ) << "Rendering took " << time << " ms (" << frame
+                            << " frames @ " << ( frame / time * 1000.f)
+                            << " FPS)" << endl;
+\end{lstlisting}}
+
+The remainder of the application code cleans up in the reverse order of
+the initialization. The config is exited, released and the connection to
+the server is closed:
+
+{\small\begin{lstlisting}
+    // 5. exit config
+    clock.reset();
+    config->exit();
+    EQLOG( eq::LOG_CUSTOM ) << "Exit took " << clock.getTimef() << " ms" <<endl;
+
+    // 6. cleanup and exit
+    server->releaseConfig( config );
+    if( !disconnectServer( server ))
+        EQERROR << "Client::disconnectServer failed" << endl;
+    server = 0;
+    return EXIT_SUCCESS;
+}
+\end{lstlisting}}
+
+\subsubsection{Render Clients}
+
+In the second and third case, when the executable is used as a render
+client, \textsf{Client::initLocal} never returns. In order to keep the
+client resident, the \textsf{eqPly} example overrides the client loop to
+keep it running beyond one configuration run:
+
+{\small\begin{lstlisting}
+bool Application::clientLoop()
+{
+    if( !_initData.isResident( )) // execute only one config run
+        return eq::Client::clientLoop();
+
+    // else execute client loops 'forever'
+    while( true ) // TODO: implement SIGHUP handler to exit?
+    {
+        if( !eq::Client::clientLoop( ))
+            return false;
+        EQINFO << "One configuration run successfully executed" << endl;
+    }
+    return true;
+}
+\end{lstlisting}}
+
+\if 0
+{\small\begin{lstlisting}
+\end{lstlisting}}
+\fi
+
 \end{document}

### 2007-08-31 Developer/ProgrammingGuide/paper.tex
New features: - divide Config::init into startInit and finishInit - more functionality in monitor primitive
Cleanups:
- replace error-prone _pendingRequestID synchronization during config
  init and exit with state monitoring

Bugfixes:
- Don't use alloca during big package send
- Correctly report config exit results

Documentation:
- Update Programming Guide
- Update data transmission API

[ ] May break build
[ ] Breaks existing applications
[x] Bugfix
[x] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-\usepackage{tabularx,graphicx,a4,listings}
+\usepackage{tabularx,graphicx,a4,listings,wrapfig,subfigure}
 
@@
   \textbf{INCOMPLETE}\\[\medskipamount]
-  Version 0.3, \today
+  Version 0.4, \today
 }
@@
     \hline
-    0.3         & Aug 26, 2007 & application and render client code described\\
-    0.2         & Aug 20, 2007 & main function described\\
+    0.4         & Aug 31, 2007 & added distributed objects\\
+    0.3         & Aug 26, 2007 & added application and render client\\
+    0.2         & Aug 20, 2007 & added main function\\
     0.1         & Aug 19, 2007 & outlined the basic concepts\\
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\columnwidth]{cave.pdf}{\caption{\label{fConfig}An
+\begin{wrapfigure}{r}{.5\textwidth}
+  \includegraphics[width=.5\textwidth]{cave.pdf}{\caption{\label{fConfig}An
       example configuration}}
-\end{figure}
+\end{wrapfigure}
 
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
 int main( int argc, char** argv )
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
     // 2. Equalizer initialization
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
 class NodeFactory : public eq::NodeFactory
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
     // 3. initialization of local client node
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
     // 4. run client
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
     // 5. cleanup and exit
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
 int Application::run()
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
     // 2. choose config
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
     // 3. init config
@@
 frame is started using \textsf{Config::startFrame} and a frame is
-finished using \textsf{Config::endFrame}.
+finished using \textsf{Config::finishFrame}.
 
@@
 
-The frame finished is \textsf{latency} frames behind the frame
-started. The latency is specified in the configuration file, and allows
-several outstanding frames whose tasks are already queued in the node
-and pipe threads for execution. This allows overlapped execution and
-minimizes idle times.
+\begin{wrapfigure}{r}{.5\textwidth}
+  \subfigure[]{\label{fSync}\includegraphics[width=.22\textwidth]{sync}}\hfil
+  \subfigure[]{\label{fAsync}\includegraphics[width=.22\textwidth]{async}}%
+  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
+\end{wrapfigure}
+\textsf{Config::finishFrame} synchronizes on the completion of the frame
+\textsf{current - latency}. The latency is specified in the
+configuration file, and allows several outstanding frames for which the
+tasks are already queued in the node and pipe threads for
+execution. This enables overlapped execution and minimizes idle
+times. The first \textsf{latency Config::finishFrame} return
+immediately, since they have no frame to synchronize
+upon. Figure~\ref{fSyncAsync} shows the execution of (hypothetical)
+rendering tasks without latency~(\ref{fSync}) and with a latency of
+one~(\ref{fAsync}).
 
@@
 up with the latency. It returns after all outstanding frames have been
-rendered:
+rendered, and is needed here to provide an accurate measurement of the
+framerate:
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
     // 4. run main loop
@@
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
     // 5. exit config
@@
 In the second and third case, when the executable is used as a render
-client, \textsf{Client::initLocal} never returns. In order to keep the
-client resident, the \textsf{eqPly} example overrides the client loop to
-keep it running beyond one configuration run:
+client, \textsf{Client::initLocal} never returns. Therefore the
+application's main loop is never executed. In order to keep the client
+resident, the \textsf{eqPly} example overrides the client loop to keep
+it running beyond one configuration run:
 
-{\small\begin{lstlisting}
+{\footnotesize\begin{lstlisting}
 bool Application::clientLoop()
@@
 
+\subsection{Distributed Objects}
+
+Equalizer provides distributed objects which help implementing data
+distribution in a graphics cluster. The master version of a distributed
+object is registered with a \textsf{eqNet::Session}, which assigns a
+session-unique identifier to the object. Other nodes can map their
+instance of the object to this identifier, thus synchronizing the
+object's data with the remotely registered object.
+
+Distributed object are created by subclassing from
+\textsf{eqNet::Object}. Distributed objects can be static (immutable) or
+dynamic. Dynamic objects are versioned.
+
+The \textsf{eqPly} example has a static distributed object to provide
+initial data to all rendering nodes, as well as a versioned object to
+provide frame-specific data, such as the camera position, to the
+rendering methods.
+
+\subsubsection{InitData - a Static Distributed Object}
+
+The \textsf{InitData} class holds a couple of parameters needed during
+initialization. These parameters never change during one configuration
+run, and are therefore static.
+
+A static distributed object either has to provide a pointer and size to
+its data using \textsf{setInstanceData} or it has to implement
+\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
+approach can be used if all distributed member variables can be outlayed
+in one contiguous block of memory. The second approach is used
+otherwise.
+
+The \textsf{InitData} class contains a string of variable
+length. Therefore it uses the second approach of manually serializing
+and deserializing its data. The serialization is done in
+\textsf{getInstanceData}. The serialized data is cached using a
+(non-distributed) member variable, which is cleared by each setter of
+distributed data. Serializing the data is a simple matter of allocating
+a large enough memory buffer, and copying the data into the buffer:
+
+{\footnotesize\begin{lstlisting}
+const void* InitData::getInstanceData( uint64_t* size )
+{
+    *size = sizeof( uint32_t ) + sizeof( eq::WindowSystem ) +
+            _filename.length() + 1;
+    if( _instanceData )
+        return _instanceData;
+
+    _instanceData = new char[ *size ];
+
+    reinterpret_cast< uint32_t* >( _instanceData )[0] =  _frameDataID;
+    reinterpret_cast< uint32_t* >( _instanceData )[1] =  _windowSystem;
+
+    const char* string = _filename.c_str();
+    memcpy( _instanceData + sizeof( uint64_t ), string, _filename.length()+1 );
+
+    return _instanceData;
+}
+\end{lstlisting}}
+
+After the data is no longer needed, \textsf{releaseInstanceData} is
+called by Equalizer to allow the object to free the data. Since the data
+is cached for further usage, the release method is not overwritten.
+
+The memory buffer returned by \textsf{getInstanceData} is transferred to
+the remote node during object mapping and passed to
+\textsf{applyInstanceData} for deserialization. The deserialization
+routine reads the data back into its own member variables:
+
+{\footnotesize\begin{lstlisting}
+void InitData::applyInstanceData( const void* data, const uint64_t size )
+{
+    EQASSERT( size > ( sizeof( _frameDataID ) + sizeof( _windowSystem )));
+
+    _frameDataID  = reinterpret_cast< const uint32_t* >( data )[0];
+    _windowSystem = reinterpret_cast< const eq::WindowSystem* >( data )[1];
+    _filename     = static_cast<const char*>( data ) + sizeof( uint64_t );
+
+    EQASSERT( _frameDataID != EQ_ID_INVALID );
+
+    EQINFO << "New InitData instance" << endl;
+}
+\end{lstlisting}}
+
+\subsubsection{FrameData - a Versioned Distributed Object}
+
+Versioned objects have to override \textsf{isStatic} to return false, in
+order for Equalizer to know that they should be versioned. Right now,
+only the master instance of the object is writable, that is,
+\textsf{eqNet::Object::commit} can be called to generate a new
+version.
+
+Upon \textsf{commit} the delta data from the previous version is sent to
+all mapped slave instances. The data is queued on the remote node, and
+is applied when the application calls \textsf{sync} to synchronize the
+object to a new version. The \textsf{sync} method might block if a
+version has not been committed or is still in transmission.
+
+In addition to the instance data (de)serialization methods needed to map
+an object, versioned objects may implement \textsf{pack} and
+\textsf{unpack} to serialize or deserialize the changes since the last
+version.
+
+If the delta data happens to be layed out contiguously in memory,
+\textsf{setDeltaData} might be used. The default implementation of
+\textsf{pack} and \textsf{unpack} (de)serialize the delta data or the
+instance data if no delta data has been specified.
+
+The frame data is layed out in one anonymous structure in
+memory. It also does not track changes since it is relatively small in
+size and changes frequently. Therefore, for the instance and delta
+date are the same and set in the constructor:
+
+{\footnotesize\begin{lstlisting}
+        FrameData()
+            {
+                reset();
+                setInstanceData( &data, sizeof( Data ));
+                EQINFO << "New FrameData " << std::endl;
+            }
+\end{lstlisting}}
+
 \if 0
-{\small\begin{lstlisting}
+\subsection{Config}
+
+\subsubsection{Config Initialization}
+
+The \textsf{eqPly::Config} class holds
+{\footnotesize\begin{lstlisting}
+bool Config::init()
+{
+    // init distributed objects
+    _frameData.data.color = _initData.useColor();
+    registerObject( &_frameData );
+    _initData.setFrameDataID( _frameData.getID( ));
+
+    registerObject( &_initData );
+\end{lstlisting}}
+\fi
+
+\if 0
+{\footnotesize\begin{lstlisting}
 \end{lstlisting}}

### 2007-09-03 Developer/ProgrammingGuide/paper.tex
Update Programming Guide, disable data transmission API, some example cleanups
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-\title{Equalizer Programming Guide}
+\title{Equalizer Programming Guide\\
+  {\footnotesize\mdseries
+    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
+    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
+}
 \author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
@@
   \textbf{INCOMPLETE}\\[\medskipamount]
-  Version 0.4, \today
+  Version 0.5, \today
 }
@@
   \hfill
-  \includegraphics[width=4cm]{logo.pdf}\hfill
-  \includegraphics[width=4cm]{vmml.pdf}\hfill\vspace{-1em}\\
+  \includegraphics[width=4cm]{images/logo.pdf}\hfill
+  \includegraphics[width=4cm]{images/vmml.pdf}\hfill\vspace{-1em}\\
 %  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
@@
     \hline
+    0.5         & Sep 3, 2007  & added config and partly pipe\\
     0.4         & Aug 31, 2007 & added distributed objects\\
@@
     0.1         & Aug 19, 2007 & outlined the basic concepts\\
-    \hline \multicolumn{3}{c}{\small
-      \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
-      {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}\\
+    \hline
   \end{tabularx}}
@@
 \subsection{Equalizer Processes}
+
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/processes.pdf}
+  {\caption{\small\label{fProcesses}Equalizer Processes}}
+\end{wrapfigure}
+All functionality of Equalizer is accessed through the Equalizer client
+library, which implements all functionality discussed in this document.
 
@@
 
+Methods called by the application always have the form
+\textsf{verb[Noun]}, whereas methods called by Equalizer (`Task
+Methods') have the form \textsf{nounVerb}.
+
 \subsection{\label{ssTaskMethods}Task Methods}
@@
 \textsf{eqPly} section will discuss the most important task methods. A
-full list of all task methods can be found on the
+full list can be found on the
 website\footnote{\htmladdnormallink{http://www.equalizergraphics.com/documents/design/taskMethods.html}{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.
@@
 
-\begin{wrapfigure}{r}{.5\textwidth}
-  \includegraphics[width=.5\textwidth]{cave.pdf}{\caption{\label{fConfig}An
-      example configuration}}
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/cave.pdf}
+  {\caption{\small\label{fConfig}An example configuration}}
 \end{wrapfigure}
@@
 system. It is one operating system process of the render client. All
-node task methods are executed from the main application thread.
+node task methods are executed from the main application thread. Each
+configuration might also use an application node, in which case the
+application process is also used for rendering.
 
@@
     }
-\end{lstlisting}}
+\end{lstlisting}}%>>
 
@@
                             << endl;
-\end{lstlisting}}
+\end{lstlisting}}%>>
 
@@
 
-\begin{wrapfigure}{r}{.5\textwidth}
-  \subfigure[]{\label{fSync}\includegraphics[width=.22\textwidth]{sync}}\hfil
-  \subfigure[]{\label{fAsync}\includegraphics[width=.22\textwidth]{async}}%
-  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
+\begin{wrapfigure}{r}{.4\textwidth}
+  \subfigure[]{\label{fSync}
+    \includegraphics[width=.18\textwidth]{images/sync}}\hfil
+  \subfigure[]{\label{fAsync}
+    \includegraphics[width=.18\textwidth]{images/async}}%
+  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
 \end{wrapfigure}
@@
                             << " FPS)" << endl;
-\end{lstlisting}}
+\end{lstlisting}}%>>
 
@@
 }
-\end{lstlisting}}
+\end{lstlisting}}%>>
 
@@
             }
-\end{lstlisting}}
+\end{lstlisting}}%>>
 
-\if 0
 \subsection{Config}
 
-\subsubsection{Config Initialization}
+The configuration is driving the applications rendering, that is, it is
+responsible for updating the data based on received events, requesting
+new frames to be rendered and to provide the render clients with the
+necessary data.
 
-The \textsf{eqPly::Config} class holds
+\subsubsection{Config Initialization and Exit}
+
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/configInit.pdf}
+  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
+\end{wrapfigure}
+The config initialization happens in parallel, that is, all config
+initialization tasks are transmitted by the server at once and their
+execution is synchronized later.
+
+The tasks are executed by the node and pipe threads in parallel. The
+parent's initialization methods are always executed before any child
+initialization method. This is necessary to allow a speedy startup of
+the configuration on large-scale graphics clusters. On the other hand,
+it means that initialization functions are called even if the parent's
+initialization has failed.
+
+The \textsf{eqPly::Config} class holds the master versions of the
+initialization and frame data. Both objects are registered with the
+\textsf{eq::Config}, which is the \textsf{eqNet::Session} used for
+rendering. Equalizer takes care of the session setup and exit in
+\textsf{Client::choose\-Config} and \textsf{Client::releaseConfig},
+respectively.
+
+The frame data is registered first, since its identifier is transmitted
+using the initialization data, which is registered afterwards. The
+identifier of the initialization data is transmitted to the render
+client nodes using the \textsf{initID} parameter of
+\textsf{eq::Config::init}. Equalizer will pass this identifier to all
+\textsf{configInit} calls of the respective objects:
+
 {\footnotesize\begin{lstlisting}
@@
     registerObject( &_initData );
+
+    // init config
+    _running = eq::Config::init( _initData.getID( ));
+    if( !_running )
+        return false;
 \end{lstlisting}}
-\fi
+
+If the configuration was initialized correctly, the configuration tries
+to set up a tracking device for head tracking. Equalizer does not
+provide extensive support for tracking device, as this is an orthogonal
+problem to parallel rendering, and has been solved already by a number
+of implementations\footnote{VRCO trackd, OpenVPRN, etc.}. The example
+code in \textsf{eqPly} is one reference implementation for the
+integration of such a tracking library:
+
+{\footnotesize\begin{lstlisting}
+    // init tracker
+    if( !_initData.getTrackerPort().empty( ))
+    {
+        if( !_tracker.init( _initData.getTrackerPort() ))
+            EQWARN << "Failed to initialise tracker" << endl;
+        else
+        {
+            // Set up position of tracking system in world space
+            // Note: this depends on the physical installation.
+            vmml::Matrix4f m( vmml::Matrix4f::IDENTITY );
+            m.scale( 1.f, 1.f, -1.f );
+            //m.x = .5;
+            _tracker.setWorldToEmitter( m );
+
+            m = vmml::Matrix4f::IDENTITY;
+            m.rotateZ( -M_PI_2 );
+            _tracker.setSensorToObject( m );
+            EQLOG( eq::LOG_CUSTOM ) << "Tracker initialised" << endl;
+        }
+    }
+
+    return true;
+}
+\end{lstlisting}}%>>
+
+The exit of the configuration stops the render clients by calling
+\textsf{eq::Config::exit}, and then deregisters the initialization and
+frame data objects with the session:
+
+{\footnotesize\begin{lstlisting}
+bool Config::exit()
+{
+    _running = false;
+    const bool ret = eq::Config::exit();
+
+    _initData.setFrameDataID( EQ_ID_INVALID );
+    deregisterObject( &_initData );
+    deregisterObject( &_frameData );
+
+    return ret;
+}
+\end{lstlisting}}
+
+\subsubsection{Frame Control}
+
+The rendering frames are issued by the application. The \textsf{Config}
+only overrides \textsf{startFrame} in order to update the its data
+before forwarding the start frame request to the \textsf{eq::Config}.
+
+If a tracker is used, the current head position and orientation is
+retrieved and given to Equalizer, which uses the head matrix together
+with the wall or projection description to compute the view
+frustra\footnote{see
+  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/immersive.html}{http://www.equalizergraphics.com/documents/design/immersive.html}}.
+
+The camera position is updated and the frame data is commited, which
+generates a new version. This version is passed to the rendering
+callbacks and will be used to synchronize the frame data to the state
+belonging to the current frame:
+
+{\footnotesize\begin{lstlisting}
+uint32_t Config::startFrame()
+{
+    // update head position
+    if( _tracker.isRunning() )
+    {
+        _tracker.update();
+        const vmml::Matrix4f& headMatrix = _tracker.getMatrix();
+        setHeadMatrix( headMatrix );
+    }
+
+    // update database
+    _frameData.data.rotation.preRotateX( -0.001f * _spinX );
+    _frameData.data.rotation.preRotateY( -0.001f * _spinY );
+    const uint32_t version = _frameData.commit();
+
+    return eq::Config::startFrame( version );
+}
+\end{lstlisting}}
+
+\subsubsection{Event Handling}
+
+Events are send by the render clients to the application using
+{eq::Config::sendEvent}. At the end of the frame,
+\textsf{Config::finishFrame} calls \textsf{Config::handleEvents} to do
+the event handling. The default implementation processes all pending
+events by calling \textsf{Config::handleEvent} for each of them.
+
+For event-driven execution, the application can override
+\textsf{Config::handleEvents} to blockingly receive events using
+\textsf{Config::nextEvent} until a new frame has to be rendered.
+
+The \textsf{eqPly} example continuously renders new frames. It
+implements \textsf{Config::hand\-le\-Event} to provide the various reactions
+to user input:
+
+{\footnotesize\begin{lstlisting}
+bool Config::handleEvent( const eq::ConfigEvent* event )
+{
+    switch( event->type )
+    {
+        case eq::ConfigEvent::WINDOW_CLOSE:
+            _running = false;
+            return true;
+
+        [...]
+
+        default:
+            break;
+    }
+    return eq::Config::handleEvent( event );
+}
+\end{lstlisting}}
+
+\subsection{Node}
+
+Foreach active render client, one \textsf{eq::Node} instance is
+created on the appropriate machine. Nodes are only instantiated on their
+render client processes, i.e., each process should have only one
+instance of the \textsf{eq::Node} class. The application process might
+also have a node class, which is handled in exactly the same way as the
+render client nodes.
+
+During node initialization the initialization data is mapped to
+a local instance using the passed identifier from
+\textsf{Config::init}. The model is loaded based on the filename in the
+initialization data. No pipe, window or channel tasks methods are
+executed before \textsf{Node::configInit} has returned.
+
+{\footnotesize\begin{lstlisting}
+bool Node::configInit( const uint32_t initID )
+{
+    eq::Config* config = getConfig();
+    const bool  mapped = config->mapObject( &_initData, initID );
+    EQASSERT( mapped );
+
+    const string& filename = _initData.getFilename();
+    EQINFO << "Loading model " << filename << endl;
+
+    _model = PlyFileIO::read( filename.c_str( ));
+    if( !_model)
+        EQWARN << "Can't load model: " << filename << endl;
+
+    return eq::Node::configInit( initID );
+}
+\end{lstlisting}}%>>
+
+The node config exit deletes the loaded model and unmaps the
+initialization data:
+
+{\footnotesize\begin{lstlisting}
+bool Node::configExit()
+{
+    delete _model;
+    _model = NULL;
+
+    eq::Config* config = getConfig();
+    config->unmapObject( &_initData );
+
+    return eq::Node::configExit();
+}
+\end{lstlisting}}
+
+\subsubsection{Frame Control}
+
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/mainloop.pdf}
+  {\caption{\small\label{fFrameSync}Synchronization of frame tasks}}
+\end{wrapfigure}
+The application has extended control over the task synchronization
+during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
+\textsf{frameStart} task methods of the various entities. The entity
+unlock all its children by calling \textsf{startFrame}, e.g., the
+\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
+to unlock the pipe threads. Note that certain \textsf{startFrame} calls
+are currently empty, as the synchronization is implicit due to the
+threading model.
+
+Likewise, \textsf{Config::finishFrame} causes the invokation of the
+\textsf{frameFinish} task methods. These task methods unlock their
+parents by calling \textsf{finishFrame}.
+
+The explicit synchronization of child or parent resources allows the
+application to optimize the processing, by doing certain, independent
+operations when the child or parent resources are already unlocked.
+
+\fig{fFrameSync} outlines the synchronization for the application, node
+and pipe classes. The window and channel synchronization is similar and
+omitted for simplicity. The \textsf{eqPly} example does not override
+\textsf{Node::frameStart} or \textsf{frameFinish}, but it is absolutely
+vital for the execution that \textsf{Node::startFrame} or
+\textsf{Node::finishFrame} are called, respectively. The default
+implementation of the node task methods does take care of that.
+
+\subsection{Pipe}
+
+All task methods for a pipe and its children are executed in a separate
+thread. This approach optimizes usage of the GPU resource, since all
+tasks are executed serially and do not compete for the usage, i.e.,
+context switches are minimized. Later versions of Equalizer might
+introduce threaded windows to allow the parallel and independent
+execution of rendering tasks on a single pipe.
+
+\subsubsection{Initialization and Exit}
+
+Pipe threads are not explicitely synchronized with each other, that is,
+pipes might be rendering different frames at one given time. Therefore
+frame-specific data has to be allocated for each pipe thread, which in
+the \textsf{eqPly} example is the frame data. The frame data is a member
+variable of the \textsf{eqPly::Pipe}, and is mapped to the identifier
+provided by the initialization data. The initialization in
+\textsf{eq::Pipe} does GPU-specific initialization, e.g., the display
+connection is opened when glX/X11 is used:
+
+{\footnotesize\begin{lstlisting}
+bool Pipe::configInit( const uint32_t initID )
+{
+    const Node*     node        = static_cast<Node*>( getNode( ));
+    const InitData& initData    = node->getInitData();
+    const uint32_t  frameDataID = initData.getFrameDataID();
+    eq::Config*     config      = getConfig();
+
+    const bool mapped = config->mapObject( &_frameData, frameDataID );
+    EQASSERT( mapped );
+
+    return eq::Pipe::configInit( initID );
+}
+\end{lstlisting}}
+
+The config exit is again symmetric to the config initialization. The
+frame data is unmapped and GPU-specific data is de-initialized by
+\textsf{eq::Config::exit}:
+
+{\footnotesize\begin{lstlisting}
+bool Pipe::configExit()
+{
+    eq::Config* config = getConfig();
+    config->unmapObject( &_frameData );
+
+    return eq::Pipe::configExit();
+}
+\end{lstlisting}}
+
+\subsubsection{Window System}
+
+Equalizer supports multiple window system interfaces, at the moment
+glX/X11, WGL and AGL/Carbon. Some operating systems, and therefore some
+Equalizer versions, support multiple window systems\footnote{see
+  \htmladdnormallink{http://www.equalizergraphics.com/compatibility.html}{http://www.equalizergraphics.com/compatibility.html}}.
+
+Each pipe might use a different window system for rendering, which is
+determined before \textsf{Pipe::configInit} by
+\textsf{Pipe::selectWindowSystem}. The default implementation of
+\textsf{selectWindowSystem} loops over all window systems and returns
+the first supported window system, determined using
+\textsf{supportsWindowSystem}.
+
+The \textsf{eqPly} examples allows selecting the window system using a
+command line option. Therefore the implementation of
+\textsf{selectWindowSystem} is overwritten, and return the specified
+window system, if it is valid:
+
+{\footnotesize\begin{lstlisting}
+eq::WindowSystem Pipe::selectWindowSystem() const
+{
+    const Node*            node     = static_cast<Node*>( getNode( ));
+    const InitData&        initData = node->getInitData();
+    const eq::WindowSystem ws       = initData.getWindowSystem();
+
+    if( ws == eq::WINDOW_SYSTEM_NONE )
+        return eq::Pipe::selectWindowSystem();
+    if( !supportsWindowSystem( ws ))
+    {
+        EQWARN << "Window system " << ws 
+               << " not supported, using default window system" << endl;
+        return eq::Pipe::selectWindowSystem();
+    }
+
+    return ws;
+}
+\end{lstlisting}}%>>
+
+\subsubsection{Frame Control}
+
+TBD
+
+{\footnotesize\begin{lstlisting}
+void Pipe::frameStart( const uint32_t frameID, const uint32_t frameNumber )
+{
+    _frameData.sync( frameID );
+    startFrame( frameNumber );
+}
+\end{lstlisting}}
 

### 2007-09-14 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   \textbf{INCOMPLETE}\\[\medskipamount]
-  Version 0.5, \today
+  Version 0.6, \today
 }
@@
 \newcommand{\fig}[1]{Figure~\ref{#1}}
+\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}
 
@@
     \hline
+    0.6         & Sep 14, 2007 & added pipe, window and object manager\\
     0.5         & Sep 3,  2007 & added config and partly pipe\\
@@
 
-Equalizer provides a framework for the development of parallel OpenGL
+Equalizer provides a framework for the development of parallel
+OpenGL\texttrademark\ 
 applications. Equalizer-based applications can run a single
@@
 distributed graphics cluster. This Programming Guide introduces the
-programming interface using the \textsf{eqPly} example shipped with Equalizer.
+programming interface using the \textsf{eqPly} example shipped with
+Equalizer as a guideline.
 
@@
 Guide should be directed to the \textsf{eq-dev} mailing
-list\footnote{see
-  \htmladdnormallink{http://www.equalizergraphics.com/lists.html}
-  {http://www.equalizergraphics.com/lists.html}}.
+list\footnote{see \link{http://www.equalizergraphics.com/lists.html}}.
+
+Equalizer is the next evolution of a generic parallel programming
+interface for visualization applications using OpenGL. Existing
+solutions, such as SGI's OpenGL Multipipe\texttrademark\ SDK, VRCO's
+Cavelib\texttrademark\ and VRJuggler, implement a subset of concepts
+similar to Equalizer. In other areas, e.g., tracking device support,
+they provide more functionality.
+
+Equalizer implements the minimum necessary layer to build parallel and
+scalable OpenGL applications. The programmer structures the application
+so that the OpenGL rendering can be executed in parallel, potentially
+across multiple processes for cluster-based execution. Equalizer
+provides the domain-specific parallel rendering know-how and abstracts
+threading, synchronization, windowing and event handling for
+portability. It is a `GLUT on steroids', providing parallel and
+distributed execution, scalable rendering features and fully
+customizable event handling.
 
@@
 example. The Quickstart
-Guide\footnote{\htmladdnormallink{http://www.equalizergraphics.com/documents/EqualizerGuide.html}{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
+Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
 explains how to run it. \textsf{eqPly} can also be executed without a
@@
 
-\subsubsection{The Server}
+\subsubsection{Server}
 
@@
 
-\subsubsection{The Application}
+\subsubsection{Application}
 
@@
 
-\subsubsection{The Render Client}
+\subsubsection{Render Clients}
 
@@
 full list can be found on the
-website\footnote{\htmladdnormallink{http://www.equalizergraphics.com/documents/design/taskMethods.html}{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.
+website\footnote{\link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.
 
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/cave.pdf}
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/cave.pdf}
   {\caption{\small\label{fConfig}An example configuration}}
-\end{wrapfigure}
+\end{figure}
 
@@
 pipes\footnote{see also
-  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/nonthreaded.html}{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.
+  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.
 
@@
 not in the scope of this document\footnote{see
-  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/compounds.html}{http://www.equalizergraphics.com/documents/design/compounds.html}}.
+  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.
 
@@
     is selected using the command-line option \textsf{--eq-client} and
-    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \htmladdnormallink{http://www.equalizergraphics.com/documents/design/residentNodes.html}{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
+    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
 \end{enumerate}
@@
 1-channel configuration\footnote{see
-  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/standalone.html}{http://www.equalizergraphics.com/documents/design/standalone.html}}:
+  \link{http://www.equalizergraphics.com/documents/design/standalone.html}}.
 
@@
 
-\subsubsection{Config Initialization and Exit}
+\subsubsection{Initialization and Exit}
 
@@
 problem to parallel rendering, and has been solved already by a number
-of implementations\footnote{VRCO trackd, OpenVPRN, etc.}. The example
+of implementations\footnote{VRCO Trackd, VRPN, etc.}. The example
 code in \textsf{eqPly} is one reference implementation for the
@@
 frustra\footnote{see
-  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/immersive.html}{http://www.equalizergraphics.com/documents/design/immersive.html}}.
+  \link{http://www.equalizergraphics.com/documents/design/immersive.html}}.
 
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/mainloop.pdf}
-  {\caption{\small\label{fFrameSync}Synchronization of frame tasks}}
-\end{wrapfigure}
 The application has extended control over the task synchronization
@@
 \textsf{frameStart} task methods of the various entities. The entity
-unlock all its children by calling \textsf{startFrame}, e.g., the
+unlock all its children by calling \textsf{startFrame}, e.g.,
 \textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
-to unlock the pipe threads. Note that certain \textsf{startFrame} calls
-are currently empty, as the synchronization is implicit due to the
-threading model.
+to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
+e.g., \textsf{Window::startFrame}, are currently empty since the
+synchronization is implicit due to the execution within the thread.
 
@@
 operations when the child or parent resources are already unlocked.
+
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/mainloop.pdf}
+  {\caption{\small\label{fFrameSync}Synchronization of frame tasks}}
+\end{figure}
 
@@
 Equalizer versions, support multiple window systems\footnote{see
-  \htmladdnormallink{http://www.equalizergraphics.com/compatibility.html}{http://www.equalizergraphics.com/compatibility.html}}.
+  \link{http://www.equalizergraphics.com/compatibility.html}}.
 
@@
 
-TBD
+All task methods for a given frame of the pipe, window and channel
+entities belonging to the thread are executed in one block, starting
+with \textsf{Pipe::frameStart} and finished by
+\textsf{Pipe::finishFrame}. The frame start callback is therefore the
+natural place to update all frame-specific data to the version belonging
+to the frame.
+
+In \textsf{eqPly}, the version of the only frame-specific object
+\textsf{FrameData} is passed as the per-frame id from
+\textsf{Config::startFrame} to the frame task methods. The pipe uses
+this version to update its instance of the frame data to the current
+version, and then unlocks its child entities by calling
+\textsf{startFrame}:
 
@@
 
+\subsection{Window}
+
+The Equalizer window holds an OpenGL drawable and rendering
+context. Subclassed windows should maintain all data specific to the
+OpenGL context. The Equalizer window creation routines share the OpenGL
+context with the first window of the pipe, this allowing the reuse of
+OpenGL objects, e.g., display lists and textures.
+
+\subsubsection{Initialization and Exit}
+
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
+  {\caption{\small\label{fWindowInit}Window Initialization Flow Chart}}
+\end{wrapfigure}
+
+The initialization sequence uses multiple, overrideable task
+methods. The main task method \textsf{configInit} executes a `child'
+task method to create the drawable and context. The child task method
+depends on the pipe's window system. The default implementations
+\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
+creates a visible window using OS-specific methods. If the OpenGL
+drawable and context was created successfully, \textsf{configInit} calls
+\textsf{configInitGL}, which performs the generic OpenGL state
+initialization. The default implementation sets up some typical OpenGL
+state, e.g., it enables the depth test.
+
+\fig{fWindowInit} shows a flow chart of the window initialization. The
+colored functions are task methods and can be replaced by
+application-specific implementations.
+
+The window-system specific initialization takes into account various
+attributes set in the configuration file. Attributes include the size of
+various frame buffer attachments (color, alpha, depth, stencil) as well
+as other framebuffer properties, such as quad-buffered stereo,
+doublebuffering, fullscreen mode and window decorations. Some of the
+attributes, such as stereo, doublebuffer and stencil can be set to
+\textsf{eq::AUTO}, in which case the default implementation tries to use
+them and gradually backs off upon failure.
+
+The \textsf{eqPly} window initialization function first calls
+\textsf{eq::Window::configInit} to use the generic window setup. If that
+was successful, it initializes an object manager and an overlay logo:
+
+{\footnotesize\begin{lstlisting}
+bool Window::configInit( const uint32_t initID )
+{
+    if( !eq::Window::configInit( initID ))
+        return false;
+
+    eq::Pipe*  pipe        = getPipe();
+    Window*    firstWindow = static_cast< Window* >( pipe->getWindow( 0 ));
+
+    EQASSERT( !_objects );
+
+    if( firstWindow == this )
+    {
+        _objects = new ObjectManager( getGLFunctions( ));
+        _loadLogo();
+    }
+    else
+    {
+        _objects     = firstWindow->_objects;
+        _logoTexture = firstWindow->_logoTexture;
+        _logoSize    = firstWindow->_logoSize;
+    }
+
+    if( !_objects ) // happens if first window failed to initialize
+        return false;
+
+    return true;
+}
+\end{lstlisting}}
+
+The object manager is used to create OpenGL objects in a multipipe,
+multi-threaded execution environment. It is described in detail
+below. It is used in conjunction with a reference pointer here, since it
+is potentially `owned' by multiple windows at the same time.
+
+The logo texture is loaded from the file system and bound to a texture
+ID used later by the channel for rendering. The details of this code are
+omitted here, since they are pretty straight-forward and not
+Equalizer-specific.
+
+The window exit deallocates all OpenGL objects, if the object manager is
+about to be disposed. The object manager does not delete the object in
+its destructor, since it might now longer have a current OpenGL
+context. In any case, \textsf{eq::Window::configExit} is called to
+destroy the drawable and context:
+
+{\footnotesize\begin{lstlisting}
+bool Window::configExit()
+{
+    if( _objects.isValid() && _objects->getRefCount() == 1 )
+        _objects->deleteAll();
+
+    _objects = 0;
+    return eq::Window::configExit();
+}
+\end{lstlisting}}
+
+\subsubsection{Object Manager}
+
+The object manager is --strictly speaking-- not part of the window. It
+is discussed here since the \textsf{eqPly} window uses an object
+manager.
+
+The object manager (OM) is a utility class and can be used to manage
+OpenGL objects across shared contexts. Typically one OM is used for each
+set of shared contexts, which normally spawns all contexts of a single
+GPU\footnote{\link{http://www.equalizergraphics.com/documents/design/objectManager.html}}.
+
+The OM is a template class. The template type is the key type, by which
+objects can be identified. The same key is used by all contexts to get
+the OpenGL name of an object. In \textsf{eqPly}, a key of type
+\textsf{const void *} is used. The rendering code uses the address of
+the data item to be rendered to obtain the associated OpenGL object.
+
+The usage of objects is reference counted. If an application releases
+the objects properly, they are automatically de-allocated. It is also
+possible to manually manage de-allocation of objects, which is often the
+more convenient use case.
+
+Currently handling for display lists, textures and buffers is
+implemented. For each object, the following functions are available:
+
+\begin{description}
+\item[supportsObjects()]: returns true if the usage for this particular
+  type of objects is supported. For basic objects, which are supported
+  on all OpenGL implementations, this function is not implemented.
+\item[getObject( key )]: returns the object associated with the given
+  key, or FAILED. Increases the reference count of existing objects.
+\item[newObject( key )]: allocates a new object for the given
+  key. Returns FAILED if the object already exists or if the allocation
+  failed. Sets the reference count of a newly created object to one.
+\item[obtainObject( key )]: convenience function which gets or obtains
+  the object associated with the given key. Returns FAILED only if the
+  object allocation failed.
+\item[releaseObject( key \textbar\ name )]: decreases the reference count and
+  deletes the object if the reference count reaches zero.
+\item[deleteObject( key \textbar\ name )]: manually deletes the object. To be
+  used if reference counting is not used.
+\end{description}
+
+\subsection{Channel}
+
+The channel is the heart of the application in that it contains the
+actual rendering code. The channel is used to perform various rendering
+operations.
+
 \if 0

### 2007-09-21 Developer/ProgrammingGuide/paper.tex
Update Programming Guide and related illustrations.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   \textbf{INCOMPLETE}\\[\medskipamount]
-  Version 0.6, \today
+  Version 0.7, \today
 }
@@
 
+\clearpage
+\tableofcontents
+\thispagestyle{empty}
 \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
     \hline
-    \bf Version & \bf Date     & \bf Changes \\
+    \bf Rev & \bf Date     & \bf Changes \\
     \hline
+    0.7     & Sep 21, 2007 & started channel and image compositing,
+                             added event handling, moved to new eqPly
+                             code base\\
     0.6     & Sep 14, 2007 & added pipe, window and object manager\\
@@
   \end{tabularx}}
-
+\thispagestyle{empty}
 \clearpage
-\tableofcontents
-\thispagestyle{empty}
-
-\clearpage
+
 \pagenumbering{arabic}
@@
 is used to set certain default values based on Equalizer-specific
-options\footnote{Equalizer-specific options always start with --eq-},
+options\footnote{Equalizer-specific options always start with -\,-eq-},
 e.g., the default server location. Furthermore, a node factory is
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
+\begin{wrapfigure}{r}{.6\textwidth}
   \subfigure[]{\label{fSync}
-    \includegraphics[width=.18\textwidth]{images/sync}}\hfil
+    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
   \subfigure[]{\label{fAsync}
-    \includegraphics[width=.18\textwidth]{images/async}}%
+    \includegraphics[width=.25\textwidth]{images/async}}%
   {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
@@
 immediately, since they have no frame to synchronize
-upon. Figure~\ref{fSyncAsync} shows the execution of (hypothetical)
-rendering tasks without latency~(\ref{fSync}) and with a latency of
-one~(\ref{fAsync}).
+upon. \fig{fSyncAsync} shows the execution of (hypothetical)
+rendering tasks without latency~(\fig{fSync}) and with a latency of
+one~(\fig{fAsync}).
 
@@
 {
-    *size = sizeof( uint32_t ) + sizeof( eq::WindowSystem ) +
-            _filename.length() + 1;
+    *size = sizeof( StaticData ) + _filename.length() + 1;
     if( _instanceData )
@@
 
-    reinterpret_cast< uint32_t* >( _instanceData )[0] =  _frameDataID;
-    reinterpret_cast< uint32_t* >( _instanceData )[1] =  _windowSystem;
+    reinterpret_cast< StaticData* >( _instanceData )[0] = _data;
 
     const char* string = _filename.c_str();
-    memcpy( _instanceData + sizeof( uint64_t ), string, _filename.length()+1 );
+    memcpy( _instanceData + sizeof( StaticData ), string, _filename.length()+1);
 
@@
 called by Equalizer to allow the object to free the data. Since the data
-is cached for further usage, the release method is not overwritten.
+is cached for further usage, the release method is not overwritten in
+\textsf{InitData}.
 
@@
 {
-    EQASSERT( size > ( sizeof( _frameDataID ) + sizeof( _windowSystem )));
-
-    _frameDataID  = reinterpret_cast< const uint32_t* >( data )[0];
-    _windowSystem = reinterpret_cast< const eq::WindowSystem* >( data )[1];
-    _filename     = static_cast<const char*>( data ) + sizeof( uint64_t );
-
-    EQASSERT( _frameDataID != EQ_ID_INVALID );
-
+    _data     = reinterpret_cast< const StaticData* >( data )[0];
+    _filename = static_cast<const char*>( data ) + sizeof( StaticData );
+
+    EQASSERT( _data.frameDataID != EQ_ID_INVALID );
     EQINFO << "New InitData instance" << endl;
@@
 Events are send by the render clients to the application using
-{eq::Config::sendEvent}. At the end of the frame,
+\textsf{eq::Config::sendEvent}. At the end of the frame,
 \textsf{Config::finishFrame} calls \textsf{Config::handleEvents} to do
@@
 implements \textsf{Config::hand\-le\-Event} to provide the various reactions
-to user input:
+to user input, most importantly camera updates based on mouse
+events. The camera position has to be handled correctly with respect to
+latency, and is therefore saved in the frame data:
 
@@
         [...]
+
+        case eq::ConfigEvent::POINTER_MOTION:
+            if( event->pointerMotion.buttons == eq::PTR_BUTTON_NONE )
+                return true;
+
+            if( event->pointerMotion.buttons == eq::PTR_BUTTON1 )
+            {
+                _spinX = 0;
+                _spinY = 0;
+
+                _frameData.data.rotation.preRotateX( 
+                    -0.005f * event->pointerMotion.dx );
+                _frameData.data.rotation.preRotateY(
+                    -0.005f * event->pointerMotion.dy );
+            }
+            else if( event->pointerMotion.buttons == eq::PTR_BUTTON2 ||
+                     event->pointerMotion.buttons == ( eq::PTR_BUTTON1 |
+                                                       eq::PTR_BUTTON3 ))
+            {
+                _frameData.data.translation.z +=
+                    .005f * event->pointerMotion.dy;
+            }
+            else if( event->pointerMotion.buttons == eq::PTR_BUTTON3 )
+            {
+                _frameData.data.translation.x += 
+                    .0005f * event->pointerMotion.dx;
+                _frameData.data.translation.y -= 
+                    .0005f * event->pointerMotion.dy;
+            }
+            return true;
 
@@
 
-    _model = PlyFileIO::read( filename.c_str( ));
-    if( !_model)
+    _model = new Model();
+    if ( !_model->readFromFile( filename.c_str() ) )
+    {
         EQWARN << "Can't load model: " << filename << endl;
+        delete _model;
+        _model = 0;
+    }
 
@@
   \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
-  {\caption{\small\label{fWindowInit}Window Initialization Flow Chart}}
+  {\caption{\small\label{fWindowInit}Window Initialization}}
 \end{wrapfigure}
@@
 \textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
-creates a visible window using OS-specific methods. If the OpenGL
+create an on-screen window using OS-specific methods. If the OpenGL
 drawable and context was created successfully, \textsf{configInit} calls
@@
 \textsf{eq::Window::configInit} to use the generic window setup. If that
-was successful, it initializes an object manager and an overlay logo:
+was successful, it initializes a state object and an overlay logo:
 
@@
 
-    EQASSERT( !_objects );
+    EQASSERT( !_state );
 
@@
     {
-        _objects = new ObjectManager( getGLFunctions( ));
+        _state = new VertexBufferState( getGLFunctions( ));
         _loadLogo();
+
+        const Node*     node     = static_cast< const Node* >( getNode( ));
+        const InitData& initData = node->getInitData();
+
+        if( initData.useVBOs() )
+        {
+            const eq::GLFunctions* glFunctions = getGLFunctions();
+            // Check if all VBO funcs available, else leave DISPLAY_LIST_MODE on
+            if( glFunctions->hasGenBuffers() && glFunctions->hasBindBuffer() &&
+                glFunctions->hasBufferData() && glFunctions->hasDeleteBuffers())
+            {
+                _state->setRenderMode( mesh::BUFFER_OBJECT_MODE );
+                EQINFO << "VBO rendering enabled" << endl;
             }
             else
-    {
-        _objects     = firstWindow->_objects;
+                EQWARN << "VBO function pointers missing, using display lists" 
+                       << endl;
+        }
+    }
+    else
+    {
+        _state       = firstWindow->_state;
         _logoTexture = firstWindow->_logoTexture;
@@
 
-    if( !_objects ) // happens if first window failed to initialize
+    if( !_state ) // happens if first window failed to initialize
         return false;
@@
 }
-\end{lstlisting}}
-
-The object manager is used to create OpenGL objects in a multipipe,
-multi-threaded execution environment. It is described in detail
-below. It is used in conjunction with a reference pointer here, since it
-is potentially `owned' by multiple windows at the same time.
+\end{lstlisting}}%>>
+
+The state object is used to handle the creation of OpenGL objects in a
+multipipe, multi-threaded execution environment. It uses an object
+manager, which is described in detail below. It is used in conjunction
+with a reference pointer here, since it is potentially `owned' by
+multiple windows at the same time.
 
@@
 
-The window exit deallocates all OpenGL objects, if the object manager is
+The window exit deallocates all OpenGL objects if the state object is
 about to be disposed. The object manager does not delete the object in
-its destructor, since it might now longer have a current OpenGL
-context. In any case, \textsf{eq::Window::configExit} is called to
+its destructor, since it does not know if an OpenGL context is still
+current. In any case, \textsf{eq::Window::configExit} is called to
 destroy the drawable and context:
@@
 {
-    if( _objects.isValid() && _objects->getRefCount() == 1 )
-        _objects->deleteAll();
-
-    _objects = 0;
+    if( _state.isValid() && _state->getRefCount() == 1 )
+        _state->deleteAll();
+
+    _state = 0;
     return eq::Window::configExit();
@@
 The object manager is --strictly speaking-- not part of the window. It
-is discussed here since the \textsf{eqPly} window uses an object
-manager.
+is discussed here since the \textsf{eqPly} window uses an object manager.
+
+The state object in \textsf{eqPly} gathers all rendering state, which
+includes an object manager for object state handling. This state object
+can easily be replaced to use the rendering code in a stand-alone
+application, e.g., for an GLUT-based renderer.
 
@@
 
+\subsubsection{Initialization and Exit}
+
+During channel initialization, the near and far planes are set to
+reasonable values to contain the whole model. During rendering, the near
+and far planes are adjusted to the current model position:
+
+{\footnotesize\begin{lstlisting}
+bool Channel::configInit( const uint32_t initID )
+{
+    setNearFar( 0.1f, 10.0f );
+    return true;
+}
+\end{lstlisting}}
+
+\subsubsection{Rendering}
+
+The central rendering routine is \textsf{Channel::frameDraw}. This
+routine contains the application's OpenGL rendering code, which is being
+rendered using the rendering context provided by Equalizer.
+
+In \textsf{eqPly}, the OpenGL context is first set up using various
+\textsf{apply} convenience methods from the base Equalizer channel
+class. Each of the \textsf{apply} methods uses the corresponding
+\textsf{get} method(s) and then calls the approriate OpenGL
+function(s). It is also possible to just query the values from Equalizer
+using the \textsf{get} methods, and use them to set up the OpenGL state
+appropriatly, for example by passing the parameters to the renderer used
+by the application.
+
+For example, the implementation for \textsf{eq::Channel::applyBuffer}
+does set up the correct rendering buffer and color mask, which depends
+on the current eye pass and possible anaglyphic stereo parameters:
+
+{\footnotesize\begin{lstlisting}
+void eq::Channel::applyBuffer()
+{
+    glReadBuffer( getReadBuffer( ));
+    glDrawBuffer( getDrawBuffer( ));
+
+    const ColorMask& colorMask = getDrawBufferMask();
+    glColorMask( colorMask.red, colorMask.green, colorMask.blue, true );
+}
+\end{lstlisting}}
+
+The following contextual information has to be used in order to render
+the view expected by Equalizer:
+\begin{description}
+\item[Buffer]: The OpenGL read and draw buffer as well as color mask.
+  These parameters are influenced by the current eye pass, eye
+  seperation and anaglyphic stereo settings.
+\item[Viewport]: The two-dimensional pixel viewport restricting the
+  rendering area within the channel. For correct operations, both
+  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
+  viewport is influenced by the destination channel's viewport
+  definition and compound viewports set for sort-first/2D decompositions.
+\item[Frustum]: Frustum parameters as defined by
+  \textsf{glFrustum}. The frustum is influenced by the destination
+  channel's view definition, compound viewports, head matrix and the
+  current eye pass.
+\item[Head Transformation:] A transformation matrix positioning the
+  frustum. This is typically an identity matrix and is used for off-axis
+  frustra in immersive rendering.
+\item[Range]: A one-dimensional range with the interval [0..1]. This
+  parameter is optional and should be used by the application to render
+  only the appropriate subset of its data.
+\end{description}
+
+Consequently, the first part of the \textsf{frameDraw} method in
+\textsf{eqPly} look as follows:
+
+{\footnotesize\begin{lstlisting}
+void Channel::frameDraw( const uint32_t frameID )
+{
+    applyBuffer();
+    applyViewport();
+
+    glMatrixMode( GL_PROJECTION );
+    glLoadIdentity();
+
+    applyFrustum();
+
+    glMatrixMode( GL_MODELVIEW );
+    glLoadIdentity();
+    applyHeadTransform();
+\end{lstlisting}}
+
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/DB.pdf}
+  {\caption{\small\label{fDB}Final view of an DB compound}}
+\end{wrapfigure}
+
+After the basic view setup, a directional light is set up, and the model
+is positioned using the camera parameters from the frame data. The
+camera parameters are transported using the the frame data to ensure
+that all channels render the frame from the same position.
+
+Furthermore, a color is set in case the model does not contain color
+information, or the color information is not to be used. In sort-last
+rendering, \textsf{eqPly} renders the individual parts with different
+colors to illustrate the algorithm, as shown in \fig{fDB}. The Equalizer
+channel provides a method to obtain a random, but unique color for all
+channels in the configuration:
+
+{\footnotesize\begin{lstlisting}
+    glLightfv( GL_LIGHT0, GL_POSITION, lightpos );
+
+    const Pipe*      pipe      = static_cast<Pipe*>( getPipe( ));
+    const FrameData& frameData = pipe->getFrameData();
+
+    glTranslatef( frameData.data.translation.x,
+                  frameData.data.translation.y,
+                  frameData.data.translation.z );
+    glMultMatrixf( frameData.data.rotation.ml );
+
+    Node*            node  = (Node*)getNode();
+    const Model*     model = node->getModel();
+    const eq::Range& range = getRange();
+
+    if( !range.isFull( )) // Color DB-patches
+    {
+        const vmml::Vector3ub color = getUniqueColor();
+        glColor3ub( color.r, color.g, color.b );
+    }
+    else if( !frameData.data.color || (model && !model->hasColors( )) )
+    {
+        glColor3f( 1.0f, 1.0f, 1.0f );
+    }
+
+\end{lstlisting}}
+
+Finally the model is rendered, if it could be loaded during
+initialization. Otherwise a quad is drawn in its place:
+
+{\footnotesize\begin{lstlisting}
+    if( model )
+    {
+        _drawModel( model );
+    }
+    else
+    {
+        glColor3f( 1.f, 1.f, 0.f );
+        glNormal3f( 0.f, -1.f, 0.f );
+        glBegin( GL_TRIANGLE_STRIP );
+        glVertex3f(  .25f, 0.f,  .25f );
+        glVertex3f(  .25f, 0.f, -.25f );
+        glVertex3f( -.25f, 0.f,  .25f );
+        glVertex3f( -.25f, 0.f, -.25f );
+        glEnd();
+    }
+}
+\end{lstlisting}}
+
 \if 0
 {\footnotesize\begin{lstlisting}
+void Channel::_drawModel( const Model* model )
+{
+    Window*                  window    = static_cast<Window*>( getWindow() );
+    mesh::VertexBufferState& state     = window->getState();
+
+    const Pipe*              pipe      = static_cast<Pipe*>( getPipe( ));
+    const FrameData&         frameData = pipe->getFrameData();
+
+    const eq::Range&         range     = getRange();
+    vmml::FrustumCullerf     culler;
+
+    state.setColors( frameData.data.color && 
+                     range.isFull() && 
+                     model->hasColors() );
+    _initFrustum( culler, model->getBoundingSphere( ));
+
+    model->beginRendering( state );
+
+\end{lstlisting}}
+
+{\footnotesize\begin{lstlisting}
+\end{lstlisting}}
+
+The model data is organized a in 3-dimensional kD-tree\footnote{See also
+  \link{http://en.wikipedia.org/wiki/Kd-tree}} for efficient view
+frustum culling. During preprocessing, each node of the tree gets a
+database range assigned.
+
+{\footnotesize\begin{lstlisting}
 \end{lstlisting}}
@@
 
+\section{Advanced Features}
+
+This section discusses some features of interest which are not covered
+by the \textsf{eqPly} section. Where possible, code examples from the
+Equalizer distribution are used.
+
+\subsection{Event Handling}
+
+Event handling is one of the areas which requires a lot of
+flexibility. On one hand, the implementation differs slightly for each
+window system due to conceptual differences in the OS-specific
+implementation. On the other hand, each application and widget set has
+its own model on how events are to be handled. Therefore, event handling
+in Equalizer is customizable in any stage of the processing\footnote{\link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.
+
+The default implementation aims to provide a quick and easy
+implementation of event handling, while allowing all necessary
+customizations. It gathers all events in the application main thread, so
+that the developer only has to implement \textsf{Config::processEvent}
+to update its data based on the pre-processed, generic events.
+
+\subsubsection{Threading}
+
+Where possible, events are received and processed by a separate per-node
+event thread to allow asynchronous event handling. Currently an event
+thread is only used by the X11/glX window system. WGL receives and
+processes the events from the pipe threads which created the
+windows. AGL receives the events from application or node main
+thread. Whenever the term \textbf{event thread} is used, it refers to
+the thread receiving the event, i.e., a per-node thread for glX, the pipe
+thread for WGL and the main thread for AGL.
+
+\subsubsection{Initialization and Exit}
+
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/eventInit.pdf}
+  {\caption{\small\label{fEventInit}Event Handling Initialization}}
+\end{wrapfigure}
+
+During window and pipe initialization the event handling is set up. For
+both entities, \textsf{initEvent\-Handler} is called to register the pipe
+or window with an event handler. This method may be overwritten to use a
+custom event handler, or to not install an event handler to disable
+event handling in Equalizer. Likewise, \textsf{exitEventHandler} is
+called to de-initialize event handling.
+
+An event handler consists of two parts: the generic base class providing
+the interface and some generic functions, and the window-system-specific
+part providiing the actual implementation.
+
+Pipe event handling is only used for glX, where one \textsf{Display}
+connection is created to subscribe to window events. Event handling is
+initialized whenever a new, window-system-specific pipe or window handle
+is set. First, \textsf{exitEventHandler} is called to de-initialize
+event handling for the old handle (if set), and then
+\textsf{initEvent\-Handler} is called for the new handle. AGL and glX use
+an event handler singleton, whereas WGL uses one event handler per
+window.
+
+\subsubsection{Message Pump}
+
+WGL and AGL require the application to manually receive and dispatch
+(`pump') events. On WGL, this has to happen on each thread with windows,
+whereas on AGL it has to happen only on the main thread. By default,
+Equalizer pumps these events automatically for the application.
+
+The methods \textsf{Client::useMessagePump} and
+\textsf{Pipe::useMessagePump} can be overriden to return \textsf{false}
+to disable this behaviour for their respective threads. On non-threaded
+pipes, \textsf{Pipe::useMessagePump} is not called.
+
+\subsubsection{Event Data Flow}
+
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/eventHandling.pdf}
+  {\caption{\small\label{fEventProcessing}Event Processing}}
+\end{wrapfigure}
+
+Events are received by an event handler. The event handler finds the
+\textsf{eq::Window} associated to the event. It then creates a generic
+\textsf{WindowEvent}, which holds important event data in a format
+independent of the window system. The original event is attached to the
+generic window event.
+
+The event handler then passes the window event to
+\textsf{Window::processEvent}, which is responsible for either handling
+the event locally, or for translating it into a generic
+\textsf{ConfigEvent}. The config events are send to the application
+thread using \textsf{Config::sendEvent}. If the event was processed, the
+function has to return \textsf{true}. If \textsf{false} is returned, the
+event will be passed to a previously installed, window-system-specific
+event handling function. The default implementation of
+\textsf{Window::processEvent} passes most events on to the application.
+
+Events sent using \textsf{Config::sendEvent} are queued in the
+application thread. After a frame has been finished,
+\textsf{Config::finishFrame} calls \textsf{Config::processEvents}. The
+default implementation of this method provides non-blocking event
+processing, that is, it calls \textsf{Config::processEvent} for each
+queued event. By overriding this function, event-driven execution can be
+implemented.
+
+\subsubsection{Custom Events in eqPixelBench}
+
+The \textsf{eqPixelBench} example is a benchmark program to measure the
+pixel transfer rates from and to the framebuffer of all channels within
+a configuration. It uses custom config events to send the gathered data
+to the application. It is much simpler than the \textsf{eqPly} example
+since it does not provide any useful rendering or user interaction.
+
+The rendering routine of \textsf{eqPixelBench} in
+\textsf{Channel::frameDraw} loops through a number of pixel formats and
+types. For each of them, it measures the time to readback and assemble a
+full-channel image. The format, type, size and time is recorded in a
+config event, which is sent to the application.
+
+The \textsf{ConfigEvent} derives from the \textsf{eq::ConfigEvent}
+structure and has the following definition:
+
+{\footnotesize\begin{lstlisting}
+struct ConfigEvent : public eq::ConfigEvent
+{
+public:
+    enum Type
+    {
+        READBACK = eq::ConfigEvent::USER,
+        ASSEMBLE
+    };
+
+    ConfigEvent()
+        {
+            size = sizeof( ConfigEvent );
+        }
+
+    // channel name is in user event data
+    char           formatType[64];
+    vmml::Vector2i area;
+    float          msec;
+};
+\end{lstlisting}}
+
+The \textsf{Config::sendEvent} method transmits a
+\textsf{eq::ConfigEvent} or derived class to the application. The
+ConfigEvent has to be a C-type structure, and its \textsf{size}
+member has to be set to the full size of the event to be transmitted.
+Each event has a type, by which the config processing function can
+identify it.
+
+User-defined types start at \textsf{eq::ConfigEvent::USER}, and the
+member variable \textsf{ConfigEvent::user} can be used to store
+\textsf{EQ\_USER\_EVENT\_SIZE}\footnote{currently 32 bytes} bytes. In
+this space, the channel's name is stored. Additional variables are used
+to transport the pixel format and type, the size and the time it took
+for rendering.
+
+On the application end, \textsf{Config::handleEvent} uses the ostream
+operator for the derived config event to output these events in a nicely
+formatted way:
+
+{\footnotesize\begin{lstlisting}
+std::ostream& operator << ( std::ostream& os, const ConfigEvent* event );
+...
+bool Config::handleEvent( const eq::ConfigEvent* event )
+{
+    switch( event->type )
+    {
+        case ConfigEvent::READBACK:
+        case ConfigEvent::ASSEMBLE:
+            cout << static_cast< const ConfigEvent* >( event ) << endl;
+            return true;
+
+        default:
+            return eq::Config::handleEvent( event );
+    }
+}
+\end{lstlisting}}%>>
+
+\subsection{Image Compositing for Scalable Rendering}
+
+Two task methods are responsible for collecting and compositing the
+result image during scalable rendering, when multiple channels are
+rendering for a single view. The source channels producing an
+\textsf{outputFrame} use \textsf{Channel::frameReadback} to read the
+pixel data from the frame buffer. The channels receiving one or multiple
+\textsf{inputFrame} use \textsf{Channel::frameAssemb\-le} to assemble
+the frames into the framebuffer. Equalizer takes care of the network
+transport of frame buffer data between nodes, if needed.
+
+\subsubsection{Frame, FrameData and Images}
+
+A \textsf{eq::Frame} references a \textsf{eq::Fra\-me\-Data}. The frame data
+is the object connecting output with input frames. It is a holder for
+images and is used to signal the availability of the pixel data. An
+\textsf{eq::Image} holds a two-dimensional snapshot of the framebuffer
+and can contain color and/or depth information.
+
+Readback and assemble operations on frames and images are designed to be
+asynchronous. They have a start and finish method for readback and
+assemble to allow the initiation and synchronization of the operation.
+Currently, only synchronous readback and assembly using
+\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
+respective start method of the image.
+
+\if 0
+
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/assembly.pdf}
+  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
+\end{wrapfigure}
+
+The offset of input and output frames characterizes the position of the
+frame data with respect to the framebuffer, that is, the
+\textbf{window's} lower-left corner. For output frames this is the
+position of the channel in 
+\fi
+
+\subsubsection{Custom Assembly in eVolve}
+
+\subsection{Head Tracking}
+\if 0
+{\footnotesize\begin{lstlisting}
+\end{lstlisting}}
+\fi
+
 \end{document}

### 2007-09-26 Developer/eqPly/chapters/abstract.tex
Added LaTeX sources of semester thesis about the new eqPly in /doc/Developer/eqPly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

% Copyright (c) 2007, Tobias Wolf <twolf@access.unizh.ch>
% All rights reserved.

%
% Abstrakt Datei
%

\clearpage
\null
\vfil

\begin{center}
\sffamily
\textbf{Zusammenfassung}
\end{center}
\normalfont
Equalizer ist ein Open Source Framework für verteiltes, paralleles Rendering. 
Es setzt auf OpenGL auf, einer plattformunabhängigen Programmierschnittstelle 
für Computergrafik. Im Rahmen dieser Semesterarbeit wird auf der Basis von 
Equalizer und OpenGL ein Programm zur effizienten Darstellung von polygonalen
Daten entwickelt.

In einem ersten Schritt wird eine effiziente Datenstruktur für die Verwaltung
der polygonalen Daten umgesetzt. In einem zweiten Schritt wird ein optisch
ansprechendes Rendering für die Darstellung dieser Daten implementiert. Das
Ergebnis dieser Arbeit ist eine neue Beispielapplikation für Equalizer.

\vfil

\begin{center}
\sffamily
\textbf{Abstract}
\end{center}
\normalfont
Equalizer is an open source framework for distributed, parallel rendering.
It is built upon OpenGL, a platform independant application programming 
interface for computer graphics. In the course of this semester thesis an
application for the efficient presentation of polygonal data is developed, 
based upon Equalizer and OpenGL.

In a first step, an efficient data structure for the storage of the polygonal
data is developed. In a second step, a visually appealing rendering for the 
presentation of that data is implemented. The result of this thesis is a new
example application for Equalizer.

\vfil
\clearpage

%
% EOF
%

### 2007-09-26 Developer/eqPly/chapters/appendix.tex
Added LaTeX sources of semester thesis about the new eqPly in /doc/Developer/eqPly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

% Copyright (c) 2007, Tobias Wolf <twolf@access.unizh.ch>
% All rights reserved.

%
% Kapiteldatei
%

\appendix
\chapter{Inhalt der CD-ROM}

Die der Semesterarbeit beiliegende CD-ROM besitzt den folgenden Inhalt:

\begin{table}[hb]
\centering
\begin{tabular}{ll}
\textbf{Pfad} & \textbf{Beschreibung} \\
zusfg.pdf & Zusammenfassung der Arbeit in Deutsch \\
abstract.pdf & Zusammenfassung der Arbeit in Englisch \\
semester.pdf & Elektronische Version der Arbeit \\
latex & LaTeX Quellen der Arbeit \\
eq & Equalizer SVN Snapshot per 30. September 2007 \\
\end{tabular}
\caption{Inhalt der CD-ROM}
\label{tbl:cd-contents}
\end{table}

Obwohl die CD-ROM einen kompletten Equalizer SVN Snapshot enthält, wird
dringend empfohlen bei Bedarf eine aktuelle Version aus dem Repository
auszuchecken \cite{equalizer:downloads}. Sowohl Equalizer als auch eqPly 
befinden sich noch in aktiver Entwicklung.

Der Source Code der ursprünglichen Beispielapplikation befindet sich im
Verzeichnis \emph{eq/src/examples/eqPly/}. Das Ergebnis dieser Arbeit, der
Source Code der neuen Applikation, liegt in \emph{eq/src/examples/eqPlyNew/}.

Um die Applikationen auszuführen ist ein vollständiges Kompilieren von 
Equalizer notwendig. Unter Mac OS X und Linux reicht es aus, dazu in das 
\emph{eq/src} Verzeichnis zu wechseln und \emph{make} auszuführen. Für 
Windows enthält \emph{eq/src/VS2005} eine Visual Studio 2005 Solution. 
Die notwendigen Voraussetzungen und Hinweise für das Kompilieren unter 
den verschiedenen Betriebssystemen finden sich in den \emph{README} Dateien
sowie der \emph{FAQ} im Verzeichnis \emph{eq/src/}.

%
% EOF
%

### 2007-09-26 Developer/eqPly/chapters/application.tex
Added LaTeX sources of semester thesis about the new eqPly in /doc/Developer/eqPly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

% Copyright (c) 2007, Tobias Wolf <twolf@access.unizh.ch>
% All rights reserved.

%
% Kapiteldatei
%

\chapter{Applikationsarchitektur}
\label{application}

In diesem Kapitel wird die generelle Architektur der Applikation diskutiert.
Da grosse Teile der Architektur durch das Equalizer Framework beeinflusst
werden, sollen zu Beginn des Kapitels die wichtigsten Konzepte aus Equalizer 
kurz vorgestellt werden. Im Anschluss daran wird dann auf die Applikation 
selbst eingegangen.

\section{Equalizer}

In dieser Sektion sollen einige Aspekte von Equalizer kurz beleuchtet werden, 
die einen Einfluss auf die Architektur und das Design der Applikation haben. 
Dies sind zum einen eine handvoll Equalizer Klassen, und die zur Verfügung
gestellten Dekompositionsmodi zum anderen.

\subsection{Klassen}
\label{eq-classes}

Equalizer abstrahiert die verschiedenen Systemresourcen in mehreren Klassen. 
Im folgenden sollen die für das Verständnis der Beispielapplikation 
notwendigen Klassen kurz vorgestellt werden. Tabelle \ref{tbl:eq-classes}
stellt die beschriebenen Klassen gegen Ende des Abschnittes zusammengefasst 
dar.

\begin{figure}[ht]
\centering
\includegraphics[width=0.92\textwidth]{figures/sampleConfig.png}
\caption[Eine Equalizer Beispielkonfiguration]
{Eine Equalizer Beispielkonfiguration (Quelle: \cite{equalizer:main})}
\label{fig:sample-config}
\end{figure}

An oberster Stelle der Resourcenverwaltung steht eine Konfiguration, welche
von der Klasse \emph{eq::Config} repräsentiert wird. Die Konfiguration enthält 
zum einen eine Beschreibung der zu verwendenden physikalischen und logischen
Resourcen, zum anderen eine Definition wie die verschiedenen Arbeitsschritte 
im Detail auf die einzelnen Resourcen zu verteilen sind.

An nächster Stelle stehen Knoten der Klasse \emph{eq::Node}, welche einzelne
Computer innerhalb des Rendering Clusters symbolisieren. Jeder solche Knoten 
besitzt eine oder mehrere Grafikkarten, die von der Klasse \emph{eq::Pipe} 
repräsentiert werden. \emph{Pipes} enthalten ihrerseits wiederum Fenster der 
Klasse \emph{eq::Window}, welche einzelne OpenGL \emph{Drawables} und 
\emph{Contexts} verwalten. Jedes Fenster kann abschliessendend noch mehrere 
\emph{Viewports} besitzen, die durch die Klasse \emph{eq::Channel} 
repräsentiert werden. Resourcen auf der selben Pipe teilen sich vorhandene 
OpenGL Objekte wie beispielsweise Display Listen, Texturen und VBOs.

\glossary{name={Pipe},description={Grafikkarte.}}
\glossary{name={Drawable},description={OpenGL Zeichenbereich.}}
\glossary{name={Context},description={OpenGL Kontext.}}
\glossary{name={Viewport},description={Sichtbereich innerhalb eines Fensters.}}

Abbildung \ref{fig:sample-config} stellt ein Beispiel einer solchen
Konfiguration dar. Es handelt sich um eine CAVE\texttrademark (Cave Automatic 
Virtual Environment), mit zwei Computern (Nodes), drei Grafikkarten (Pipes) 
und zugehörigen Fenstern (Windows), die zusammen vier verschiedene Kanäle 
(Channels) rendern, die den Projektionen auf die vier Wände entsprechen.

Neben den oben genannten Resourcenklassen spielt noch eine weitere Klasse
eine wichtige Rolle für das Verständnis der Applikation. Da mehrere Kopien 
der Applikation im Cluster laufen, müssen sich diese gemeinsam genutzte 
Daten auf irgend eine Weise teilen können. Equalizer stellt dafür verteilte 
Objekte der Klasse \emph{eqNet::Object} bereit, die sofern benötigt auch 
Versionierung unterstützen.

Das Equalizer Framework sieht vor, dass Applikationen eigene Subklassen von
den genannten Basisklassen ableiten. Innerhalb der verschiedenen Klassen
stehen eine Reihe von vordefinierten \emph{Task} Methoden zur Verfügung, 
welche von der Applikation wiederum nach Bedarf überschrieben werden können 
um die gewünschte Funktionalität zu erreichen. Eine detaillierte Beschreibung 
der Equalizer Task Methoden findet sich in \cite{equalizer:tasks}.

\glossary{name={Task},description={Aufgabe.}}

\begin{table}[hb]
\centering
\begin{tabular}{ll}
\textbf{Equalizer Klasse} & \textbf{Bedeutung} \\
eq::Config & Beschreibung der vorhandenen Resourcen und deren Verwendung \\
eq::Node & Einzelner Computer innerhalb des Rendering Clusters \\
eq::Pipe & Grafikkarte(n) des jeweiligen Knotens \\
eq::Window & OpenGL Drawables auf der entsprechenden Pipe\\
eq::Channel & Viewports innerhalb des zugehörigen Fensters \\
eqNet::Object & Verteilte Objekte mit Unterstützung für Versionierung\\
\end{tabular}
\caption{Wichtige Equalizer Klassen und deren Bedeutung}
\label{tbl:eq-classes}
\end{table}

\subsection{Dekompositionsmodi}
\label{eq-decomposition}

Zum Abschluss dieser Sektion soll noch ein Blick auf die verschiedenen
Dekompositionsmodi geworfen werden die Equalizer unterstützt, da diese 
ebenfalls einen gewissen Einfluss auf das Applikationsdesign haben. Zur
Übersicht sind diese in Tabelle \ref{tbl:eq-decomposition} gegen Ende
wieder gemeinsam aufgeführt.

\textbf{2D}:~~Im 2D Modus wird der zweidimensionale Bildschirmraum in 
mehrere Kacheln aufgeteilt (siehe Abbildung \ref{fig:eq-2d}). Verschiedene 
Rendereinheiten bearbeiten nur ihren eigenen Teil, und die Teilergebnisse 
werden am Ende zu einem Gesamtbild zusammengefügt. Damit jede Rendereinheit 
effizient arbeitet, sollte die Applikation View Frustum Culling einsetzen 
und die Datenstruktur entsprechende Unterstützung bieten.

\begin{figure}[ht]
\centering
\includegraphics[width=0.90\textwidth]{figures/2D.png}
\caption[Equalizer 2D Dekompositionsmodus]
{Equalizer 2D Dekompositionsmodus (Quelle: \cite{equalizer:main})}
\label{fig:eq-2d}
\end{figure}

\textbf{DB}:~~Im DB Modus wird die Datenstruktur aufgeteilt. Jede Rendereinheit
ist für einen bestimmten Anteil der polygonalen Daten zuständig und rendert nur
diese (siehe Abbildung \ref{fig:eq-db}). Die Zusammensetzung erfolgt dann mit 
Hilfe der Tiefenpufferinformation. Damit dieser Modus funktioniert muss die 
Datenstruktur zusätzliche Informationen enthalten, die Teilbereiche auf
eindimensionale Intervalle abbilden.

\begin{figure}[ht]
\centering
\includegraphics[width=0.90\textwidth]{figures/DB.png}
\caption[Equalizer DB Dekompositionsmodus]
{Equalizer DB Dekompositionsmodus (Quelle: \cite{equalizer:main})}
\label{fig:eq-db}
\end{figure}

\textbf{Eye}:~~Der Eye Modus wird im Stereo Rendering verwendet, verschiedene
Rendereinheiten sind für die Ansicht welche jeweils ein bestimmtes Auge sieht
verantwortlich. Dieser Modus hat keinen Einfluss auf das Applikationsdesign.

Neben diesen einfachen Dekompositionen unterstützt Equalizer auch beliebige
Kombinationen derselben. Dies hat jedoch auf die Applikation keinen weiteren
Einfluss und wird darum hier nicht weiter erläutert.

\begin{table}[hb]
\centering
\begin{tabular}{ll}
\textbf{Modus} & \textbf{Auswirkungen} \\
2D & Applikation und Datenstruktur sollten Frustum Culling unterstützen \\
DB & Datenstruktur sollte auf eindimensionale Intervalle abbildbar sein \\
Eye & Keine besonderen Auswirkungen \\
\end{tabular}
\caption{Equalizer Dekompositionsmodi und deren Auswirkung auf die Applikation}
\label{tbl:eq-decomposition}
\end{table}

\section{eqPly}

Nachdem nun die grundlegenden Aspekte von Equalizer behandelt wurden, welche 
Einfluss auf Applikationsarchitektur und -design nehmen, wird in dieser Sektion
nun die Applikation selbst betrachtet. Sie trägt den Namen \textbf{eqPly}.

Aufgrund der relativ klar vorgegebenen Struktur für Equalizer Programme weicht 
die Architektur der neuen Applikation kaum von der des alten Beispielsprogrammes
ab. Abbildung \ref{fig:eq-classes} am Ende der Sektion zeigt die Zusammenhänge 
der Klassen sowohl untereinander als auch in Relation zu Equalizer. Da der 
Aufbau von \emph{eqPly} in \cite{equalizer:guide} schon ausführlich behandelt 
wird, beschränken sich die folgenden Ausführungen auf ein Minimum. Die Klassen 
betreffend Datenstruktur und Rendering (\emph{mesh::Vertex*}) werden gesondert 
in Kapitel \ref{data-structure} und \ref{rendering} behandelt.

\subsection{Datenklassen}

Die Datenklassen sind verteilte Objekte, und somit von \emph{eqNet::Object}
abgeleitet. \emph{eqPly::InitData} kapselt den Knoten übergreifenden, aber
statisch bleibenden Zustand der Applikation. Hauptsächlich sind dies der Pfad 
zum Modell, ein Schalter ob der neue VBO Rendermodus verwendet werden soll, 
und die ID des zu verwendenden \emph{Frame} Datenobjektes.

\glossary{name={Frame},description={Einzelbild einer Bildfolge.}}

\emph{eqPly::LocalInitData} ist von \emph{eqPly::InitData} abgeleitet und 
enthält den statischen, lokalen Knoten spezifischen Zustand, bespielsweise den 
Port zur Kommunikation mit einem eventuell vorhandenen \emph{Head Tracker}, 
oder ob der Knoten beim Rendering Farben des Modells nutzen oder ignorieren 
soll. Die Klasse ist auch zuständig für das Parsen der Kommandozeilenparameter.

\glossary{name={Head Tracker},description={Gerät zur Verfolgung der 
Kopfbewegung mit Hilfe externer Referenzpunkte.}}

\emph{eqPly::FrameData} ist ein versioniertes, verteiltes Objekt und enhält 
dynamische Daten die sich von einem Frame zum anderen verändern können. Dies
sind in \emph{eqPly} vor allem die Rotations- und Translationsmatrizen für
die Kameraposition, die sich sowohl durch Benutzerinteraktion als auch durch
kontinuierliche Animation fortlaufend ändert.

\subsection{Abgeleitete Klassen}

Wie im Abschnitt \ref{eq-classes} bereits beschrieben, leiten Equalizer 
Applikationen eine Reihe von Basisklassen ab und überschreiben darin die Task 
Methoden um ihre gewünschte Funktionsweise zu implementieren. Genau dies 
geschieht in den Klassen \emph{eqPly::Config, eqPly::Node, eqPly::Pipe, 
eqPly::Window} und \emph{eqPly::Channel}.

\emph{eqPly::Config} ist die Steuerung der Applikation. Die Klasse verwaltet 
die Primärkopien der \emph{eqPly::InitData} und \emph{eqPly::FrameData} Objekte
und behandelt die Registrierung dieser verteilten Objekte. Sie initialisiert 
und kommuniziert mit einem eventuell vorhandenen Head Tracker, ist für die 
Ereignisbehandlung zuständig und aktualisiert sämtliche Frame relevanten Daten 
zu Beginn eines neuen Frames. Dazu gehört insbesonders die neue Position der 
Kamera.

Die Aufgaben von \emph{eqPly::Node} sind das \emph{Mapping} der verteilten 
statischen \emph{eqPly::InitData} auf eine lokale Kopie sowie das Laden des 
Polygonmodells aus der in \emph{eqPly::InitData} übermittelten Datei.

\glossary{name={Mapping},description={Abbildung oder Zuordnung.}}

Da nicht von allen Rendereinheiten zur gleichen Zeit derselbe Frame gerendert
werden muss, unterhält \emph{eqPly::Pipe} ein Mapping des dynamischen 
\emph{eqPly::FrameData} Objekts. Zu Beginn eines neuen Frames wird die ID des 
zu rendernden Frames übermittelt, und die Klasse synchronisiert ihre Kopie der
\emph{eqPly::FrameData} mit der gewünschten Version.

\emph{eqPly::Window} enthält den \emph{VertexBufferState}, welcher den 
\emph{State} für das Rendering der Datenstruktur sowie die auf der selben Pipe 
gemeinsam genutzten OpenGL Objekte verwaltet. \emph{eqPly::Window} ist auch 
zuständig für das Laden des Equalizer Logos, die Initialiserung der OpenGL 
\emph{Extensions} sowie der Auswahl des zu verwendenden Rendermodus. Auf die 
State Klasse und OpenGL Extensions wird in Kapitel \ref{rendering} noch näher 
eingegangen.

\glossary{name={State},description={Zustand.}}
\glossary{name={Extensions},description={Erweiterungen von OpenGL, welche den 
Einzug in den OpenGL Hauptstandard noch nicht geschafft haben.}}

\emph{eqPly::Channel} leistet die Hauptarbeit der Applikation. Die Klasse ist
zuständig für das Zeichnen des Logos, die Berechnung des Frustums, für die 
Anwendung der verschiedenen Modell- und Projektionstransformationen sowie das
View Frustom Culling und Rendering des Modells.

\subsection{Hilfsklassen}
\label{helper-classes}

Zwei weitere Module sind Bestandteil von \emph{eqPly}. Zum einen handelt es
sich um die Dateien \emph{ply.h} und \emph{plyfile.cpp}, welche die Bibliothek
für das Laden der PLY Dateien darstellen. Da sie ursprünglich in reinem C 
geschrieben wurde, ist die Bibliothek weder objektorientiert noch besonders
einfach zu verwenden.

Zum anderen sind da noch die Dateien \emph{tracker.h} und \emph{tracker.cpp}.
Dabei handelt es sich um eine Klasse zur Kommunikation mit einem Head Tracker, 
der von \emph{eqPly} bei Vorhandensein verwendet werden kann um Kopfbewegungen
eines Benutzers zu verfolgen und die Ansicht entsprechend anzupassen.

\subsection{Hauptprogramm}

Die \emph{eqPly::Application} Klasse ist das Gerüst der Applikation. Zusammen 
mit der \emph{main} Funktion werden die notwendigen Klassen instanziert, sowie
die benötigten Schritte für das Aufsetzen einer Equalizer Applikation in die
Wege geleitet. Da diese ausführlich in \cite{equalizer:guide} beschrieben
sind, wird an dieser Stelle nicht weiter darauf eingegangen.

\begin{sidewaysfigure}[hp]
\centering
\includegraphics[width=1.00\textheight]{figures/UML.pdf}
\caption[Zusammenhänge der Klassen von eqPly und Equalizer]
{Zusammenhänge der Klassen von eqPly und Equalizer (Quelle: \cite{equalizer:guide})}
\label{fig:eq-classes}
\end{sidewaysfigure}

%
% EOF
%

### 2007-09-26 Developer/eqPly/chapters/conclusion.tex
Added LaTeX sources of semester thesis about the new eqPly in /doc/Developer/eqPly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

% Copyright (c) 2007, Tobias Wolf <twolf@access.unizh.ch>
% All rights reserved.

%
% Kapiteldatei
%

\chapter{Zusammenfassung}
\label{conclusion}

Diese Semesterarbeit hatte zwei Ziele. Zum ersten die Umsetzung einer neuen 
Datenstruktur für die effiziente Verwaltung von polygonalen Daten, und zum 
zweiten ein ebenfalls effizientes aber auch optisch ansprechendes Rendering 
dieser Daten.

Der erste Teil der Arbeit wurde vollständig umgesetzt, die Implementierung der
neuen kd-Tree Datenstruktur erfüllt sämtliche in Abschnitt \ref{requirements} 
gestellten Anforderungen. Der einzige spürbare Nachteil im Vergleich zur alten
Datenstruktur ist ein bedeutend langsamerer Aufbau des Baumes. Da nach der
erstmaligen Erstellung aber sowieso binäre Repräsentationen gespeichert werden
und für weitere Aufrufe verfügbar sind, ist dieser Punkt vernachlässigbar. Der
Speicherverbraucht wurde auf weniger als ein Viertel reduziert (siehe Tabelle
\ref{tbl:file-sizes}.

Momentan wird der kd-Tree von jedem Renderknoten vollständig geladen. Eine
denkbare Erweitung für die Verwendung mit der Equalizer DB Dekomposition wäre
beispielsweise, ein teilweises Laden nur der benötigen Bereiche zu ermöglichen.
Die Effizienz des View Frustum Cullings liesse sich unter Umständen verbessern
indem statt beim Aufbau die Sortierachse zu rotieren jeweils diejenige mit der
längsten Dimension gewählt würde.

Des weiteren steht die Ausgliederung der Mesh Klassen in eine eigenstände 
Biobliothek aus, um diese auch für andere Projekte verwenden zu können. Als 
Folge daraus wäre auch eine GLUT Applikation denkbar, welche die Benutzung 
der Mesh Klassen in einer stand-alone Umgebung demonstriert.

Der zweite Teil der Arbeit erreicht die definierten Ziele teilweise. Das neue
Rendering mit Vertex Buffers Objects ist implementiert und funktioniert. Auch
Rendering mit Display Listen wird weiterhin angeboten für Umgebungen ohne VBO 
Unterstützung.

Ein Vergleich der Renderingperformance zwischen der alten und der neuen 
Applikation findet sich in folgender Tabelle \ref{tbl:render-performance}.
Die verwendeten Modelle stammen aus \cite{stanford:repository} und 
\cite{cyberware}.

\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
\textbf{Modell} & \textbf{eqPly (alt)} & \textbf{eqPly (neu, DL)} & 
\textbf{eqPly (neu, VBO)}\\
xyz & 1.0 FPS & 2.0 FPS & 3.0 FPS
\end{tabular}
\caption{Vergleich Renderingperformance altes und neues eqPly}
\label{tbl:render-performance}
\end{table}

%Leider
%bleibt die Performance aber in vielen Fällen aus bisher ungeklärten Gründen 
%hinter den Erwartungen zurück. Das Rendering mit Display Listen steht aber nach
%wie vor als alternative Lösung zur Verfügung, und bleibt derzeit auch noch das 
%Standard Verfahren.

Das Rendering ist durch die standardmässige Nutzung von Gouraud anstelle von
Flat Shading bereits optisch ansprechender als in der alten Applikation. Dies
ist in Abbildungen \ref{fig:sample-flat} und \ref{fig:sample-smooth} deutlich
zu sehen.

\begin{figure}[hb]
\centering
\includegraphics[width=0.86\textwidth]{figures/FlatShading.png}
\caption{RockerArm Modell mit Flat Shading}
\label{fig:sample-flat}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=0.86\textwidth]{figures/SmoothShading.png}
\caption{RockerArm Modell mit Gouraud Shading}
\label{fig:sample-smooth}
\end{figure}

Der Einsatz von GLSL Shadern und Phong Shading verbessert dies je nach Licht- 
und Materialeigenschaften noch einmal. Weitere Algorithmen konnten aufgrund 
mangelnder Zeit leider nicht mehr realisiert werden, das vorhandene Gerüst 
sollte das Hinzufügen weiterer Shader Algorithmen zu einem späteren Zeitpunkt 
jedoch problemlos ermöglichen.

%
% EOF
%

### 2007-09-26 Developer/eqPly/chapters/datastructure.tex
Added LaTeX sources of semester thesis about the new eqPly in /doc/Developer/eqPly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

% Copyright (c) 2007, Tobias Wolf <twolf@access.unizh.ch>
% All rights reserved.

%
% Kapiteldatei
%

\chapter{Datenstruktur}
\label{data-structure}

In diesem Kapitel wird die erste Teilaufgabe der Arbeit behandelt, die 
Datenstruktur. In einem ersten Schritt wird die alte Datenstruktur analysiert
und erläutert sowie deren Probleme aufgezeigt. In einem zweiten Schritt werden
dann die Anforderungen an die neue Datenstruktur spezifiziert und das gewählte
Design sowie die schlussendliche Implementierung vorgestellt.

\section{Alte Datenstruktur}

Der Kern der alten Datenstruktur ist ein \emph{Bounding Box Octree 
\cite{octree:desc}}. Die \emph{Bounding Boxes} sind achsenparallele Quader 
welche die enthaltenen Daten minimal umschliessen. Der \emph{Octree} ist eine 
Baumstruktur zur binären Raumpartitionierung bei der ein Knoten entweder 
keine oder acht Kinder hat.

Ein Knoten des Baumes wird solange mit Daten befüllt, bis eine vordefinierte
Grenze erreicht ist. Bei Überschreiten dieser Grenze wird die Bounding Box des 
Knotens in acht gleich grosse Oktanten unterteilt, welche die neuen Kinder des
Knotens sind. Die im Knoten enthaltenen Daten werden gemäss ihrer neuen 
Zugehörigkeit auf die Kinder verteilt. Dieser Vorgang wird solange wiederholt, 
bis ein Blatt genug Platz für die Daten bietet, oder eine gewisse maximale
Baumtiefe erreicht wurde.

\begin{figure}[hb]
\centering
\includegraphics[width=0.32\textwidth]{figures/QuadTree.png}
\caption{2D Raumaufteilung mittels Quadtree}
\label{fig:quadtree}
\end{figure}

Abbildung \ref{fig:quadtree} soll dies anhand eines grafischen Beispiels
verdeutlichen. Um die Komplexität einer dreidimensionalen Darstellung zu 
vermeiden, wird der zweidimensionale Fall eines \emph{Quadtree} (Teilung in 
vier Quadranten) mit einzelnen Punkten als Daten gezeigt. Die Datemenge
umfasst hier 16 Punkte, und die Grenze pro Knoten beträgt 3 Punkte. Erzeugt 
wurde die Abbildung mit Hilfe von \cite{spatialdemos}.

\subsection{Analyse}

Das alte Modell wird primär repräsentiert durch eine Klasse \emph{PlyModel}.
Es handelt sich dabei um eine Template Klasse, die als Template Argument den
Typ der zu speichernden \emph{Faces} (in diesem Fall Triangles) benötigt. 
Dieser Typ sieht konkret wie folgt aus:

\glossary{name={Faces},description={Flächen aus denen sich ein Objekt 
zusammensetzt.}}

\begin{lstlisting}
template<class VertexType>
struct NormalFace
{
    float       normal[3];
    VertexType  vertices[3];
};
\end{lstlisting}

Es wird nur eine Flächennormale gespeichert, und der Typ der drei Vertices ist
wiederum ein Template Argument. Es stehen zwei verschiedene Vertextypen zur
Auswahl, eine erste Variante die nur Vertexpositionen beinhaltet, und eine 
zweite die zusätzlich noch eine Farbe speichert:

\begin{lstlisting}
struct Vertex
{
    float  pos[3];
};
\end{lstlisting}

\begin{lstlisting}
struct ColorVertex : public Vertex
{
    float color[3];
};
\end{lstlisting}

Der Kern der Klasse \emph{PlyModel} ist die Kapselung des Bounding Box Octrees,
entsprechend enthält sie einen eigenen Typ für die Bounding Boxes mit denen die
polygonalen Daten effektiv verwaltet werden:

\begin{lstlisting}
struct BBox 
{
    BBox*          parent;
    BBox*          next;
    BBox*          children;

    Vertex         pos[2];
    Vertex         cullBox[2];

    vmml::Vector4f cullSphere;

    size_t         nFaces;
    FaceType*      faces;

    float          range[2];

    void*          userData;
};
\end{lstlisting}

Die Datenstruktur enthält Zeiger zu ihrem Elternknoten, dem nächsten Nachbarn
auf gleicher Baumtiefe sowie den potentiellen Kindern. Darüber hinaus wird 
die dem Knoten entsprechende Bounding Box und eine davon abgeleitete Bounding 
Sphere (erlaubt effizienteres View Frustum Culling) gespeichert, ebenso ein 
dem Anteil der enthaltenen Daten entsprechendes eindimensionales Intervall 
(benötigt für Equalizer DB Dekomposition) sowie die Anzahl der enthaltenen 
Triangles und ein Zeiger auf eine Liste derselben.

Die Klasse \emph{PlyModel} bietet basierend auf dieser Struktur eine Reihe von 
Methoden an. Das Erstellen und Füllen des Bounding Box Octree wird dabei genau 
so unterstützt wie das Speichern und Laden einer binären Repräsentation. Die
Klasse verwaltet ebenfalls die Hauptkopie der Liste aller Triangles und bietet 
eine Reihe von Hilfsfunktionen wie zum Beispiel das Skalieren der Modelldaten
oder der Berechnung von Flächennormalen und Bounding Spheres.

Die Methoden für das Einlesen der Daten aus PLY Dateien sind in einer eigenen
Klasse \emph{PlyFileIO} untergebracht, welche die schon in Abschnitt 
\ref{helper-classes} erwähnte Biobliothek dafür nutzt. Das Rendering der
alten Datenstrukur ist vollständig von dieser externalisiert, und befindet
sich in \emph{eqPly::Channel}.

\subsection{Probleme}

Sowohl die verwendete Octree Baumstruktur als auch die Implementierung des
Polygonmodells offenbart eine Reihe von Schwächen. Im folgenden sollen zunächst
die Probleme des Octrees aufgezeigt werden.

Bei Überschreiten der Füllgrenze pro Knoten wird der Raum immer in acht gleich
grosse Oktanten unterteilt, dies geschieht ohne Rücksicht auf die tatsächliche
Verteilung der Daten. In der Folge ist die Blattgrösse nur sehr schlecht zu
kontrollieren, sie kann zwischen leer bis hinauf zur Füllgrenze variieren. Ein
weiterer Effekt dieser fehlenden Kontrolle in Bezug auf die Knotengrössen ist,
dass die Baumtiefe in verschiedenen Bereichen des Baumes stark schwanken kann,
je nach Dichte der Daten in diesem Bereich. Das Fazit ist, dass der Octree in
dieser Form äussert schlecht ausbalanciert ist.

Auch die konkrete Implementierung zeigt eine Reihe von Schwächen. Die PLY 
Biobliothek alloziert für mehrere Rückgabevariablen Speicher, der von den 
aufrufenden Methoden in \emph{PlyFileIO} nicht vollständig freigegeben wird. 
Die Folge sind mögliche Speicherlecks beim Einlesen der PLY Dateien.

Das Modell verwaltet direkt Triangles, und speichert darum gemeinsam genutzte
Vertices mehrfach. Informationen über den Aufbau der polygonalen Daten wie
welche Triangles welche Vertices teilen (und entsprechend benachbart sind) 
gehen verloren. Dies und die Tatsache dass nur Flächennormalen gespeichert 
werden, führt dazu dass beim Rendering nur Flat Shading verwendet werden kann, 
und andere Shading Verfahren aufgrund von fehlenden Informationen gar nicht 
realisiert werden können.

Die mehrfache Template Benutzung führt zur leicht kryptischen Deklaration 
eines konkreten Modells der Form
\lstinline!typedef PlyModel< NormalFace< ColorVertex > > Model;!. Darüber 
hinaus ist der Implementierung die C Herkunft noch deutlich anzusehen 
(Verwendung von C Arrays, malloc() usw.), und sie ist allgemein eher
unübersichtlich und ihre Funktionsweise nur schwer nachzuvollziehen. Ein
letztes Problem ist, dass die alte Datenstruktur in ihrer Implementierung
sehr viel Speicher verbraucht.

\section{Neue Datenstruktur}

Aufgrund der nicht vorhandenen Balancierung des Octrees und der schlechten
Kontrolle über die Blattgrössen, wurde ein kd-Tree \cite{kd-tree:desc} als
neuer Kern der Datenstruktur ausgewählt.

Der kd-Tree ist wie der Octree eine Sruktur zur binären Raumpartitionierung,
die Aufteilung erfolgt jedoch nach einem anderem Verfahren, was zu einer Reihe
von Vorteilen führt. Der grundlegende Algorithmus zur Erstellung eines kd-Trees
läuft wie folgt ab.

In jedem Schritt werden die Daten zunächst entlang einer bestimmten Achse
sortiert, die Sortierachse rotiert dabei mit der Rekursionstiefe. Danach wird 
der Median ermittelt, und die Datenmenge in zwei Häften geteilt. Die eine 
enthält Objekte kleiner als der Median, und die andere Objekte grösser oder 
gleich dem Median. Beide Hälften werden rekursiv weiter verarbeitet, bis die 
gewünschte Blattgrösse erreicht wird.

\begin{figure}[ht]
\centering
\includegraphics[width=0.32\textwidth]{figures/kdTree.png}
\caption{2D Raumaufteilung mittels kd-Tree}
\label{fig:kd-tree}
\end{figure}

Abbildung \ref{fig:kd-tree} soll dies anhand eines grafischen Beispiels
verdeutlichen. Er wird der zweidimensionale Fall gezeigt. Die verwendeten 
Daten sind die selben 16 Punkte wie im Quadtree Beispiel in Abbildung 
\ref{fig:quadtree}, und die Grenze für die Blattgrösse ist ebenfalls wieder 3.
Die erste Aufteilung spaltet den Raum auf der x-Achse, die zweite auf der
y-Achse, und die dritte und letzte wiederum auf der x-Achse.

Wie aus der Beschreibung des Algorithmus und dem Beispiel ersichtlich sein 
sollte, hat ein so erstellter kd-Tree mehrere Vorteile: Die Anzahl Knoten fällt
geringer aus als beim Quadtree/Octree, weil keine leeren Bereiche entstehen. Da
die Raumteilung immer anhand eines Datenwertes erfolgt, sind die geteilten
Bereiche (fast) gleich gross, und die Baumtiefe ist ebenfalls gleichmässig. Im
Endeffekt ist dieser Baum also beinahe perfekt ausbalanciert.

\subsection{Anforderungen}
\label{requirements}

Es werden an die neue Datenstruktur eine Reihe von Anforderungen gestellt, die
zum Teil schon durch die Wahl des kd-Tree erfüllt werden, zum anderen aber im
Design und in der Implementierung Berücksichtigung finden müssen. Die folgenden
Anforderungen sind zum Teil neu, und galten zum Teil auch schon für die alte
Datenstruktur und müssen entsprechend auch von der neuen erfüllt werden.

Als erstes soll die Baumstruktur besser ausbalanciert sein, und eine genauere
Kontrolle über die Blattgrössen aufgrund der Verwendung von VBOs beim Rendering
erreicht werden. Diese beiden Punkte werden direkt durch die kd-Tree Auswahl 
erfüllt.

Als zweites sollen die Informationen über die Struktur der polygonalen Daten 
erhalten bleiben, also die Nachbarschaftsbeziehungen zwischen den Triangles. 
Dies einerseits, damit bei der Speicherung und beim Rendering Vorteile aus
den gemeinsam genutzten Vertices gezogen werden können, und andererseits um neu
Vertexnormalen anstelle der Flächennormalen berechnen zu können. Diese erlauben 
erst die Verwendung besserer Shading Algorithmen.

Als drittes soll der neue Code heutigen objektrientierten Ansprüchen genügen,
die Möglichkeiten von C++ besser ausnutzen und natürlich einfach verständlich
und gut wartbar sein.

Da Equalizer selbst Unterstützung für mehrere Betriebssysteme (Mac OS X, 
Linux, Windows) sowohl in 32 als auch 64 Bit bietet und auf unterschiedlichen
Prozessorfamilien (mit Little- und Big-Endian Datenformat) läuft, müssen die
entsprechenden Eigenheiten auch in Hinblick auf die neue Datenstruktur mit 
berücksichtigt werden.

Schliesslich soll die neue Datenstruktur auch wieder effizientes View Frustum 
Culling und ein Mapping der Baumbereiche auf ein eindimensionales Intervall 
für die DB Dekomposition von Equalizer unterstützen.

\subsection{Design}

\subsubsection*{Basistypen}

Um die verwendeten Datentypen flexibel zu halten, aber auf eine übermässige
Verwendung von Templates zu verzichten, wurde entschieden die Basistypen der
Datenstruktur in Form von eigenen Typedefs zu bilden.

\begin{lstlisting}
typedef vmml::Vector3< GLfloat >    Vertex;
typedef vmml::Vector4< GLubyte >    Color;
typedef vmml::Vector3< GLfloat >    Normal;
typedef size_t                      Index;
typedef GLushort                    ShortIndex;
\end{lstlisting}

Besondere Erwähnung verdient hierbei die Definition von zwei verschiedenen 
Indextypen. Grund hierfür ist, dass für effizientes Rendering die verwendeten
Indices im Bereich eines Short Datentyps liegen sollen. Das Einsortieren der
polygonalen Daten in den kd-Tree arbeitet noch mit den globalen, grösseren 
Indices. Sobald jedoch ein Blatt erzeugt wird, werden die darin enthaltenen
Triangles neu indexiert. Dies impliziert natürlich, dass ein Blatt maximal so
viele unterschiedliche Vertices enthalten darf wie im Short Bereich Platz 
finden.

Auf Basis dieser Primärtypen sowie einer Wrapper Hilfsklasse für kleine Arrays
mit fester Grösse werden dann die folgenden zusammengesetzten Typen definiert.

\begin{lstlisting}
typedef vmml::Vector3< Index >      Triangle;
typedef ArrayWrapper< Vertex, 2 >   BoundingBox;
typedef vmml::Vector4< float >      BoundingSphere;
typedef ArrayWrapper< float, 2 >    Range;
\end{lstlisting}

Effizientes View Frustum Culling lässt sich einfacher mit Bounding Spheres als 
mit Bounding Boxes realisieren, da der Überlappungstest einfacher ausfällt. Die
Struktur wird selber nur die Bounding Spheres speichern, Boxes kommen aber noch
beim Aufbau temporär zum Einsatz, da sich diese besser zusammenfassen lassen.

\subsubsection*{Mesh Klassen}

Nachdem die Basistypen vorgestellt wurden, können nun die Klassen betrachtet
werden. Da die Datenstruktur Polygonmeshes verwaltet, wird für alle Belange 
der Datenstruktur der eigene Namespace \emph{mesh} verwendet.

\emph{mesh::VertexData} speichert die polygonalen Daten in ihrer Ursprungsform,
zur Verwendung kommen C++ STL Vektoren. Darüber hinaus besitzt die Klasse alle 
notwendigen Methoden um PLY Dateien einzulesen, die Bounding Box des gesamten
Modells zu ermitteln, das Modell zu skalieren, die Normalen zu berechnen sowie
die Triangles für die Erzeugung des kd-Trees zu sortieren.

\begin{lstlisting}
class VertexData
{
public:
    VertexData();

    bool readPlyFile( const char* file, const bool ignoreColors );
    void sort( const Index start, const Index length, const Axis axis );
    void scale( const float baseSize );
    void calculateNormals( const bool vertexNormals );
    void calculateBoundingBox();
    const BoundingBox& getBoundingBox() const;

    std::vector< Vertex >   vertices;
    std::vector< Color >    colors;
    std::vector< Normal >   normals;
    std::vector< Triangle > triangles;

private:
    ...

    BoundingBox _boundingBox;
};
\end{lstlisting}

\emph{mesh::VertexBufferData} verwaltet die sortierten und reindexierten Daten
für den kd-Tree. Während der Erstellung des Baumes werden die Elemente von 
\emph{mesh::VertexData} in die Vektoren dieser Klasse kopiert, entsprechend 
kann \emph{mesh::VertexData} nach Erstellen des Baums direkt entsorgt werden.
Da \emph{mesh::VertexBufferData} Bestandteil des kd-Tree ist, besitzt es wie
alle Klassen des Baumes selbst Methoden zum Speichern seiner Daten in einen 
Stream bzw. zum Laden aus einem memory-mapped File.

\begin{lstlisting}
class VertexBufferData
{
public:
    void clear();
    void toStream( std::ostream& os );
    void fromMemory( char** addr );

    std::vector< Vertex >       vertices;
    std::vector< Color >        colors;
    std::vector< Normal >       normals;
    std::vector< ShortIndex >   indices;

private:
    ...
};
\end{lstlisting}

\emph{mesh::VertexBufferState} ist eine Hilfsklasse, die den Renderstate für
den kd-Tree abstrahiert. Dies sind insbesondere der zu verwendende Rendermodus
(sieht Kapitel \ref{rendermodes}), die Verwendung von Farben sowie die
Verwaltung von OpenGL Objekten und Erweiterungsfunktionen. Die Klasse selbst 
ist abstrakt, konkrete Versionen sind für eine stand-alone Nutzung 
\emph{mesh::VertexBufferStateSimple} und \emph{mesh::EqVertexBufferState} für 
die Verwendung mit Equalizer.

\begin{lstlisting}
class VertexBufferState
{
public:
    virtual bool useColors() const;
    virtual void setColors( const bool colors );
    virtual RenderMode getRenderMode() const;
    virtual void setRenderMode( const RenderMode mode );

    virtual GLuint getDisplayList( const void* key );
    virtual GLuint newDisplayList( const void* key );
    virtual GLuint getBufferObject( const void* key );
    virtual GLuint newBufferObject( const void* key );

    virtual void deleteAll();

    virtual const GLFunctions* getGLFunctions() const;

protected:
    VertexBufferState( const GLFunctions* glFunctions ); 
    virtual ~VertexBufferState();

    const GLFunctions*  _glFunctions;
    bool                _useColors;
    RenderMode          _renderMode;

private:
};
\end{lstlisting}

\emph{mesh::VertexBufferBase} ist die abstrakte Basisklasse für alle anderen
Baumklassen. Sie definiert das Interface welches alle Typen von Knoten anbieten
müssen, und implementiert bereits Basisfunktionalität soweit dies möglich ist.
Im Gegensatz zur alten Datenstruktur ist das Rendering nun Bestandteil der
Struktur selbst. Entsprechend muss jeder Knoten eine Methode für das Rendering 
zur Verfügung stellen. Weitere öffentliche Methoden sind die Rückgabe der 
Kinder und der Bounding Spheres für das externe View Frustum Culling sowie des 
abgedeckten Intervalls für die Equalizer DB Dekomposition.

Zusätzlich besitzt die Klasse Methoden für das Speichern und Laden ihrer Daten,
für das Erstellen des kd-Trees sowie der Aktualisierung der Bounding Sphere und
des Intervalls. Da alle diese Methoden von Klienten nur auf dem Wurzelknoten 
aufgerufen werden sollen sind sie auf allgemeiner Knotenebene \emph{protected}.

\begin{lstlisting}
class VertexBufferBase
{
public:
    virtual void render( VertexBufferState& state ) const;

    const BoundingSphere& getBoundingSphere() const; 
    const float* getRange() const;

    virtual const VertexBufferBase* getLeft() const;
    virtual const VertexBufferBase* getRight() const;

protected:
    VertexBufferBase();
    virtual ~VertexBufferBase();

    virtual void toStream( std::ostream& os );
    virtual void fromMemory( char** addr, VertexBufferData& globalData );

    virtual void setupTree( VertexData& data, const Index start,
                            const Index length, const Axis axis,
                            VertexBufferData& globalData );
    virtual BoundingBox updateBoundingSphere();
    virtual void updateRange();

    void calculateBoundingSphere( const BoundingBox& bBox );

    BoundingSphere  _boundingSphere;
    Range           _range;

private:
};
\end{lstlisting}

\emph{mesh::VertexBufferNode} ist die Klasse für reguläre Knoten die kein Blatt
sind. Sie ist abgeleitet von \emph{mesh::VertexBufferBase} und selbst 
Basisklasse für \emph{mesh::VertexBufferRoot}. Die Klasse besitzt mit Ausnahme
einer Hilfsfunktion keine neuen Methoden, implementiert aber einerseits die
abstrakten Methoden von \emph{mesh::VertexBufferNode} und überschreibt 
andererseits sofern notwendig bestehende virtuelle Methoden.

\begin{lstlisting}
class VertexBufferNode : public VertexBufferBase
{
public:
    ...

protected:
    ...

private:
    size_t countUniqueVertices( VertexData& data, const Index start,
                                const Index length ) const;

    VertexBufferBase*   _left;
    VertexBufferBase*   _right;
};
\end{lstlisting}

\emph{mesh::VertexBufferRoot} ist die Klasse für den Wurzelknoten des Baumes, 
und abgeleitet von \emph{mesh::VertexBufferNode}. Sie verwaltet die Hauptkopie 
der polygonalen Daten, und stellt den öffentlichen Zugang für Klienten dar. 
Entsprechend zu den \emph{protected} Methoden der Basisklassen bietet sie 
zugehörige \emph{public} Methoden an.

Das Rendering der Datenstruktur ist entweder durch einen einzelnen Aufruf der
entsprechenden Methode auf dem Wurzelknoten möglich, oder durch mehrere direkte
Aufrufe der \emph{render()} Methoden der regulären bzw. Blattknoten, wie es zum
Beispiel bei Verwendung des externen View Frustum Culling geschieht. In beiden 
Fällen ist vor und nach dem Rendering auf der Wurzel \emph{begin-} und 
\emph{endRendering()} aufzurufen. Diese beiden Methoden kümmern sich um den 
Auf- und Abbau des für das Rendering notwendigen OpenGL States.

\begin{lstlisting}
class VertexBufferRoot : public VertexBufferNode
{
public:
    virtual void render( VertexBufferState& state ) const;

    void beginRendering( VertexBufferState& state ) const;
    void endRendering( VertexBufferState& state ) const;

    void setupTree( VertexData& data );
    bool writeToFile( const char* filename );
    bool readFromFile( const char* filename );
    bool hasColors() const;

protected:
    virtual void toStream( std::ostream& os );
    virtual void fromMemory( char* start );

private:
    bool constructFromPly( const char* filename );

    VertexBufferData _data;
};
\end{lstlisting}

Die letzte Mesh Klasse ist \emph{mesh::VertexBufferLeaf}, bei welcher es sich
um die Klasse für Blattknoten handelt. Sie hält Referenzen auf die globalen
polygonalen Daten in der Wurzel, sowie Start und Länge der Bereiche innerhalb 
der globalen Daten für welche sie zuständig ist. Während die \emph{render()} 
Methoden der übrigen Knoten nicht mehr machen als an die Kinder zu delegieren,
sind die Blätter für das tatsächliche Rendering zuständig.

\begin{lstlisting}
class VertexBufferLeaf : public VertexBufferBase
{
public:
    ...

protected:
    ...

private:
    void setupRendering( VertexBufferState& state ) const;
    void renderImmediate( VertexBufferState& state ) const;
    void renderDisplayList( VertexBufferState& state ) const;
    void renderBufferObject( VertexBufferState& state ) const;

    VertexBufferData&   _globalData;
    Index               _vertexStart;
    ShortIndex          _vertexLength;
    Index               _indexStart;
    Index               _indexLength;
    mutable bool        _isSetup;
};
\end{lstlisting}

\subsection{Implementierung}

Nachdem das Design der neuen Datenstruktur im Detail erklärt wurde, soll nun 
auf einige ausgewählte Details der Implementierung eingegangen werden. Für die
vollständigen Details der Implementierung sei auf den Source Code auf der
beiliegenden CD-ROM verwiesen.

Die neue Datenstruktur wurde zwar in erster Hinsicht für die Verwendung in der
Equalizer Beispielapplikation entwickelt, sollte aber dennoch unabhängig von
Equalizer selbst bleiben um unter Umständen auch in anderen Projekten einsetzbar
zu sein. Da für beide Einsatzzwecke an einigen Stellen unterschiedlicher Code
notwendig ist, wurde eine Lösung mit Makros und bedingt compiliertem Code 
verwendet. Die Datei \emph{typedefs.h} bietet einerseits die Einstellung ob 
Equalizer verwendet werden soll (\lstinline!#define EQUALIZER!), und 
andererseits die Definitionen welche Streams für Logausgaben 
(\lstinline!#define MESHINFO! etc.) und welche Form von Assertions 
(\lstinline!#define MESHASSERT!) genutzt werden sollen. In der Datei wird auch
ein NullStream definiert, der bei stand-alone Benutzung verwendet werden kann
um gewisse Logausgaben komplett zu verwerfen.

Generell ist zur Implementierung zu sagen, dass intern verwendete Methoden im 
Fehlerfall Ausnahmen vom Typ \emph{mesh::MeshException} werfen, während die 
vom Klienten aufrufbaren Methoden ihren Erfolg über die Rückgabe von 
\emph{boolean} Werten signalisieren. Besonders unter Windows verlangsamt der
Einsatz von Debugging Code verschiedene Bereiche des Baumaufbaus so stark, dass
selbst mittelgrosse PLY Modelle nicht mehr in nützlicher Frist verarbeitet
werden.

\subsubsection*{VertexData}

\emph{mesh::VertexData} verwendet für das Einlesen der PLY Dateien die selbe
Bibliothek wie die alte Beispielapplikation. Der Code selbst wurde aber ganz
neu geschrieben, und bemüht sich nun Speicherlecks zu vermeiden. In Bezug auf
die Sortierfunktion für Triangles ist anzumerken, dass diese nach willkürlich 
festgelegter Art vorgeht (siehe Code). Ein anderes Sortierverfahren würde den
Zweck ebenfalls erfüllen, es muss nur eine gleichbleibende Ordnung über die 
Triangles definiert werden.

\subsubsection*{VertexBufferNode}

Aus \emph{mesh::VertexBufferNode} soll kurz betrachtet werden, wie die
Umsetzung des kd-Tree Algorithmus in der Praxis nun tatsächlich aussieht.
\emph{countUniqueVertices()} ist eine Hilfsfunktion welche die Anzahl 
einzigartiger Vertices innerhalb des betrachteten Bereiches ermittelt. Das 
Reindexieren der Indices und der Aufbau der sortierten polygonalen Daten 
findet in der \emph{setupTree()} Methode von \emph{mesh::VertexBufferLeaf} 
statt. Die Konstrukte mit dem \emph{static\_cast} die hier und an einigen 
anderen Stellen verwendet werden, sind notwendig um die \emph{protected}
Eigenschaft der verwendeten Methoden zu umgehen.

\begin{lstlisting}
void VertexBufferNode::setupTree( VertexData& data, const Index start,
                                  const Index length, const Axis axis,
                                  VertexBufferData& globalData )
{
    ...

    data.sort( start, length, axis );
    Index median = start + ( length / 2 );

    // left child will include elements smaller than the median
    Index leftLength = length / 2;
    if( countUniqueVertices( data, start, leftLength ) > LEAF_SIZE )
        _left = new VertexBufferNode;
    else
        _left = new VertexBufferLeaf( globalData );

    // right child will include elements equal to or greater
    Index rightLength = ( length + 1 ) / 2;
    if( countUniqueVertices( data, median, rightLength ) > LEAF_SIZE )
        _right = new VertexBufferNode;
    else
        _right = new VertexBufferLeaf( globalData );

    // move to next axis and continue construction in the child nodes
    Axis newAxis = static_cast< Axis >( ( axis + 1 ) % 3 );
    static_cast< VertexBufferNode* >
        ( _left )->setupTree( data, start, leftLength, newAxis, 
                              globalData );
    static_cast< VertexBufferNode* >
        ( _right )->setupTree( data, median, rightLength, newAxis, 
                               globalData );
}
\end{lstlisting}

\subsubsection*{VertexBufferLeaf}

Die entsprechende \emph{setupTree()} Methode in \emph{mesh::VertexBufferLeaf}
sieht wie folgt aus. Sie sortiert die Daten ein letztes Mal nach der neuen
Achse, und reindexiert die Vertices dann damit alle Indices innerhalb des 
Short Bereichs liegen. Während dem Reindexieren werden die Daten ebenfalls 
sortiert in die globale \emph{mesh::VertexBufferData} Struktur eingefügt.
Werden vereinzelte Vertices auch über Blattgrenzen hinweg noch gemeinsam
genutzt, werden die entsprechenden Daten durch diesen Schritt dupliziert.

\begin{lstlisting}
void VertexBufferLeaf::setupTree( VertexData& data, const Index start,
                                  const Index length, const Axis axis,
                                  VertexBufferData& globalData )
{
    ...
    data.sort( start, length, axis );
    _vertexStart = globalData.vertices.size();
    _vertexLength = 0;
    _indexStart = globalData.indices.size();
    _indexLength = 0;

    const bool hasColors = ( data.colors.size() > 0 );

    // stores the new indices (relative to _start)
    map< Index, ShortIndex > newIndex;

    for( Index t = 0; t < length; ++t )
    {
        for( Index v = 0; v < 3; ++v )
        {
            Index i = data.triangles[start + t][v];
            if( newIndex.find( i ) == newIndex.end() )
            {
                newIndex[i] = _vertexLength++;
                // assert number does not exceed SmallIndex range
                MESHASSERT( _vertexLength );
                globalData.vertices.push_back( data.vertices[i] );
                if( hasColors )
                    globalData.colors.push_back( data.colors[i] );
                globalData.normals.push_back( data.normals[i] );
            }
            globalData.indices.push_back( newIndex[i] );
            ++_indexLength; 
        }
    }
    ...
}
\end{lstlisting}

Auf die für das Rendering relevanten Methoden von \emph{mesh::VertexBufferLeaf}
sei auf Kapitel \ref{rendering} verwiesen, sie werden dort genauer behandelt.

\subsubsection*{VertexBufferRoot}

Der letzte fehlende Bestandteil des Baumaufbaus ist die Methode die der Client
in \emph{mesh::VertexBufferRoot} aufruft um den Vorgang zu starten. Sie ist
nicht viel mehr als ein öffentlicher Wrapper für die \emph{protected} Methode.
Nach dem eigentlichen Baumaufbau werden ebenfalls rekursiv die Bounding Spheres
sowie die Intervalle für die Equalizer DB Dekomposition berechnet. Anzumerken
ist hierbei, dass die Methode \emph{updateBoundingSphere()} temporär mit
Bounding Boxes arbeitet und an den Aufrufer zurückliefert, weil das Verschmelzen
der Bounding Volumes damit einfacher ist als bei der direkten Verwendung von
Bounding Spheres.

\begin{lstlisting}
void VertexBufferRoot::setupTree( VertexData& data )
{
    // data is VertexData, _data is VertexBufferData
    _data.clear();
    VertexBufferNode::setupTree( data, 0, data.triangles.size(), 
                                 AXIS_X, _data );
    VertexBufferNode::updateBoundingSphere();
    VertexBufferNode::updateRange();
}
\end{lstlisting}

Es wurden bereits plattformspezifische Eigenheiten erwähnt, die es bei der
Implementierung ebenfalls zu berücksichtigen gilt. \emph{mesh::VertexBufferRoot}
verwendet in der Hinsicht beispielsweise zwei Hilfsfunktionen, um einerseits
zu testen ob es sich um ein 32 oder 64 Bit System handelt, und andererseits
ob das System die Daten im Little- oder Big-Endian Format verwaltet. Für jede
mögliche Kombination werden eigenständige binäre Repräsentationen des kd-Trees
abgespeichert.

\begin{lstlisting}
size_t getArchitectureBits()
{
    return ( sizeof( void* ) * 8 );
}
\end{lstlisting}

\begin{lstlisting}
bool isArchitectureLittleEndian()
{
    unsigned char test[2] = { 1, 0 };
    short x = *( reinterpret_cast< short* >( test ) );
    return ( x == 1 );
}
\end{lstlisting}

Mit diesen ausgewählten Betrachtungen einzelner Implementierungssdetails ist
das Ende der ersten Teilaufgabe der vorliegenden Arbeit erreicht.

Tabelle \ref{tbl:file-sizes} zeigt anhand der Dateigrössen der binären 32 Bit 
Repräsentationen (es findet keine Komprimierung statt) noch auf, wie sehr der 
Speicherbedarf durch den Wechsel von der alten auf die neue Datenstruktur 
reduziert wurde. Die für die Messungen verwendeten PLY Beispielmodelle 
stammen aus \cite{stanford:repository}.

\begin{table}[ht]
\centering
\begin{tabular}{lrr}
\textbf{Modell} & \textbf{Dateigrösse Octree} & \textbf{Dateigrösse kd-Tree} \\
Armadillo & 27.72 MB & 6.09 MB \\
Dragon & 69.83 MB & 15.34 MB \\
Happy Buddha & 87.16 MB & 19.05 MB \\
Lucy & n/a & 494.58 MB \\
Stanford Bunny & 5.56 MB & 1.20 MB \\
\end{tabular}
\caption{Unterschied Speicherbedarf der Datenstrukturen}
\label{tbl:file-sizes}
\end{table}

%
% EOF
%

### 2007-09-26 Developer/eqPly/chapters/introduction.tex
Added LaTeX sources of semester thesis about the new eqPly in /doc/Developer/eqPly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

% Copyright (c) 2007, Tobias Wolf <twolf@access.unizh.ch>
% All rights reserved.

%
% Kapiteldatei
%

\chapter{Einleitung}

Die Leistung von Grafikhardware hat sich in den letzten Jahren kontinuierlich 
und rasant weiter entwickelt. Mit höherer Leistungsfähigkeit gehen aber auch 
höhere Ansprüche einher. Es besteht der Wunsch, immer komplexer werdende Daten 
mit steigendem Datenvolumen in höheren Auflösungen zu visualisieren. Trotz des 
Fortschrittes gelangen tradionelle Computergrafik Systeme in gewissen 
Anwendungsgebieten an ihre Grenzen, denn obwohl sich die Hardwareleistung 
in regelmässigen Abständen verdoppelt, wachsen die produzierten und 
darzustellenden Datenmengen mindestens genau so schnell 
\cite{pajarola:datasize}.

Dem ähnlich gelagerten Problem einer zu geringen Rechenleistung einzelner 
Computersysteme wurde begegnet, indem mehrere verbundene Rechnersysteme die
gewünschten Berechnungen verteilt aber parallel ausführen. Analog dazu lässt
sich auch im Bereich der Computergrafik die notwendige Arbeit auf mehrere
Grafiksysteme verteilen, welche dann einzelne Teilbereiche parallel bearbeiten 
und die Teilergebnisse wo notwendig wieder zu einem Gesamtbild zusammensetzen.

\section{Aufgabenstellung}

\emph{Equalizer} \cite{equalizer:main}\cite{equalizer:tr} ist ein 
Lösungsansatz für die im letzten Abschnitt vorgestellte Problematik. Es handelt 
sich bei Equalizer um ein Open Source Framework für verteiltes, paralleles 
\emph{Rendering}\footnote{Englische Begriffe werden im Glossar am Ende des
Dokumentes übersetzt oder erklärt.}, welches sich zur Zeit in Entwicklung 
befindet. Es setzt auf \emph{OpenGL} \cite{opengl:main} auf, der bedeutendsten 
Programmierschnittstelle für plattformübergreifende, anspruchsvolle 
Grafikprogrammierung.

\glossary{name={Rendering},description={Erzeugung eines digitalen Bildes aus 
einer Bildbeschreibung.}}

Ziel der vorliegenden Semesterarbeit ist die Entwicklung einer Applikation für 
die effiziente und optisch ansprechende Darstellung von polygonalen Daten, auf
Basis des Equalizer Frameworks und der OpenGL Programmierschnittstelle. Ein 
bereits existierendes Beispielprogramm wird als Grundlage verwendet, und soll 
am Ende durch das Ergebnis dieser Arbeit vollständig ersetzt werden.

Die erste Teilaufgabe besteht darin, eine effiziente Datenstruktur für die 
Verwaltung von polygonalen Daten umzusetzen. Die Wahl fiel hierbei auf eine 
besondere Baumstruktur, den sogenannten \emph{kd-Tree} \cite{kd-tree:desc}.

\glossary{name={kd-Tree},description={K-dimensionaler Baum, eine Datenstruktur
aus der Informatik zur räumlichen Aufteilung.}}

Die polygonalen Daten in Form von \emph{Vertices} und \emph{Triangles} werden 
aus Dateien im PLY Dateiformat \cite{ply:desc} gelesen, und anschliessend in 
die Baumstruktur einsortiert. Die fertige Datenstruktur erlaubt sowohl ein 
einfaches \emph{View Frustum Culling} als auch ein effizientes Rendering. Da 
das Erstellen des kd-Trees ein sehr aufwändiger Prozess ist, werden die Daten 
für zukünftige Verwendung in einem eigenen Binärformat zwischengespeichert.

\glossary{name={Vertices},description={Eckpunkte grafischer Objekte.}}
\glossary{name={Triangles},description={Dreiecke. Elementare Objekte der
Computergrafik aus denen komplexere Objekte typischerweise zusammengesetzt
sind.}}
\glossary{name={View Frustum Culling},description={Eliminierung von 
Objekten die ausserhalb des sichtbaren Bereiches liegen vor dem Rendering.}}

Die zweite Teilaufgabe schliesslich ist die Implementierung eines Rendering 
Verfahrens für die optisch ansprechende Darstellung der polygonalen Daten.

Die Datenstruktur ist optimiert für die Verwendung von \emph{Vertex Buffer 
Objects} (VBOs) \cite{vbo:whitepaper}, welche entsprechend auch für das
Rendering verwendet werden sollen. Zur Verbesserung der visuellen Qualität
sollen verschiedene Algorithmen einsetzt werden, welche mit Hilfe der 
\emph{Open GL Shader Language} (GLSL) \cite{opengl:glsl} entwickelt werden 
und direkt auf der Grafikhardware ablaufen. Damit auch ältere OpenGL 
Umgebungen, welche weder VBOs (ab Version 1.5) noch GLSL (ab Version 2.0) 
unterstützen, die Applikation ausführen und darstellen können, wird eine 
OpenGL Version 1.1 kompatible Lösung ebenfalls implementiert.

\section{Gliederung}

Der Aufbau der Arbeit sieht wie folgt aus. Nach der Einführung in diesem
Kapitel beschäftigt sich Kapitel \ref{application} mit dem grundliegenden
Aufbau der Applikation. Weil dazu auch ein gewisses Verständnis der Equalizer
Architektur notwendig ist, werden wichtige Elemente daraus kurz vorgestellt.

Kapitel \ref{data-structure} widmet sich ausführlich dem ersten Teil der 
Aufgabe. Es wird die bisher verwendete Datenstruktur erklärt und deren
Schwächen aufgezeigt. Die neue Datenstruktur wird vorgestellt und auch auf
deren konkrekte Implementierung eingegangen.

Der zweite Teil der Aufgabe wird in Kapitel \ref{rendering} behandelt. Es
werden zunächst die verschiedenen möglichen OpenGL Rendermodi vorgestellt, 
und dann deren Verwendung in der bestehenden und neuen Applikation aufgezeigt.
Anschliessend werden die unterschiedlichen Algorithmen des \emph{Shading} 
diskutiert, sowie deren Anwendung sowohl in Standard OpenGL als auch mit GLSL.

\glossary{name={Shading},description={Simulation der Oberflächen grafischer
Objekte anhand der Beleuchtung und Materialeigenschaften.}}

Den Abschluss der Arbeit stellt Kapitel \ref{conclusion} dar, in welchem 
das Ergebnis der Arbeit kurz zusammengefasst sowie ein Ausblick auf weitere
zukünftige Erweiterungen der Applikation geworfen wird.

%
% EOF
%

### 2007-09-26 Developer/eqPly/chapters/rendering.tex
Added LaTeX sources of semester thesis about the new eqPly in /doc/Developer/eqPly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

% Copyright (c) 2007, Tobias Wolf <twolf@access.unizh.ch>
% All rights reserved.

%
% Kapiteldatei
%

\chapter{Rendering}
\label{rendering}

In diesem Kapitel wird die zweite Teilaufgabe der Arbeit behandelt, das 
Rendering. In der ersten Sektion werden die drei OpenGL Rendermodi beschrieben
welche die neue Applikation grundsätzlich unterstützt, sowie die konkrete
Implementierung der entsprechenden Rendermodi in Bezug auf die Applikation. In
der zweiten Sektion werden anschliessend die möglichen Shading Algorithmen
vorgestellt, und zu deren besseren Verständnis kurz das Beleuchtungsmodell
rekapituliert. Abgeschlossen wird das Kapitel mit einem Ausblick auf weitere
Algorithmen die das Rendering optisch attraktiver gestalten können.

\section{OpenGL Rendermodi}
\label{rendermodes}

Die alte Beispielapplikation hat für das Rendering ausschliesslich Display
Listen verwendet. Die neue Applikation wird nun um die Möglichkeit erweitert,
für das Rendering Vertex Buffer Objects \cite{vbo:whitepaper} zu verwenden.

\subsection{Immediate Mode}

Immediate Mode ist die einfachste Möglichkeit im Umgang mit OpenGL. Kommandos
werden einzeln abgesetzt und direkt ausgeführt. Dies bedeutet für den konkreten
Fall des Rendering unserer polygonalen Daten, dass für jedes zu zeichnende 
Triangle je drei Aufrufe für das Setzen von Vertexpositionen, Normalen und
Farbinformationen notwendig sind. Im folgenden der entsprechende Auszug aus
\emph{mesh::VertexBufferLeaf}.

\begin{lstlisting}
inline
void VertexBufferLeaf::renderImmediate( VertexBufferState& state )
{
    glBegin( GL_TRIANGLES );  
    for( Index offset = 0; offset < _indexLength; ++offset )
    {
        Index i = _vertexStart + 
                  _globalData.indices[_indexStart + offset];
        if( state.useColors() )
            glColor4ubv( &_globalData.colors[i][0] );
        glNormal3fv( &_globalData.normals[i][0] );
        glVertex3fv( &_globalData.vertices[i][0] );
    }
    glEnd();
}
\end{lstlisting}

Immediate Mode hat aber gravierende Nachteile in Hinblick auf die Performance.
Der Overhead für die häufigen Funktionsaufrufe und das einzelne Durchreichen 
der Daten ist signifikant. Immediate Mode taugt deshalb in anspruchsvollen
Anwendungen heute praktisch nur noch zu Testzwecken. Die entsprechende Methode
ist in \emph{mesh::VertexBufferLeaf} nur für Vergleichszwecke vorhanden, und
weil der selbe Code auch für die Nutzung von Display Listen Verwendung findet.

\subsection{Display Lists}

Display Listen bilden den ersten Ausweg zur signifikanten Leistungssteigerung.
Neue Display Listen werden mit \lstinline!glGenLists()! angefordert, und lassen
sich einfach erstellen indem vorhandener Immediate Mode Code mit den Befehlen
\lstinline!glNewList()! und \lstinline!glEndList()! umschlossen wird. Sämtliche
Kommandos die zwischen diesen Zeilen abgesetzt werden, kompiliert OpenGL zu 
einer Display Liste. Aufgerufen wird eine Display Liste durch einen einzelnen
Befehl, \lstinline!glCallList()!.

In unserer Applikation lässt sich auf Basis des oben bereits dargestellten
Immediate Mode Codes ohne Probleme eine neue Display Liste anlegen. Dies wird
einmalig in der Methode \emph{vertexBufferLeaf::setupRendering()} erledigt,
danach steht die Display Liste für weitere Rendervorgänge zur Verfügung.

\begin{lstlisting}
glNewList( displayList, GL_COMPILE );
renderImmediate( state );
glEndList();
\end{lstlisting}

Das Rendering in \emph{mesh::VertexBufferLeaf} reduziert sich damit auf:

\begin{lstlisting}
inline
void VertexBufferLeaf::renderDisplayList( VertexBufferState& state )
{
    GLuint displayList = state.getDisplayList( this );
    glCallList( displayList );
}
\end{lstlisting}

Aktuelle OpenGL Treiber optimieren die Kommandos beim Kompilieren einer Display
Liste stark, und speichern die Display Listen direkt auf der Grafikhardware. Das
Ergebnis ist in vielen Fällen eine kaum zu verbessernde Renderleistung, solange
die Display Listen nicht signifikant zu gross oder klein sind.

Display Listen haben aber auch Nachteile, auch wenn diese für diese Applikation
nicht ins Gewicht fallen. Display Listen können keine Befehle enthalten, welche
den OpenGL Client State verändern. Findet sich solch ein Befehl innerhalb
eines Display Listen Blocks, wird er direkt ausgeführt, aber nicht mit in die
Liste kompiliert. Weiterhin taugen Display Listen nur für statische Objekte, da 
eine kompilierte Display Liste nachträglich nicht mehr verändert werden kann.

\subsection{Vertex Buffer Objects}

Vertex Buffer Objects, kurz VBOs \cite{vbo:whitepaper}, sind ursprünglich eine 
Extension von OpenGL und ab Version 1.5 fester Bestandteil der Spezifikation. 
Kurz formuliert erlauben es VBOs, diverse Daten wie Vertexpositionen, Normalen 
und Farben nicht nur blockweise mit einem Aufruf zu spezifizieren, sondern 
diese werden im Idealfall auch direkt auf der Grafikhardware gespeichert. VBOs 
taugen dabei nicht nur für statische Daten, sondern können auch für Daten 
verwendet werden die sich dynamisch ändern. Damit der Treiber die Zugriffe 
optimieren kann, wird beim Erstellen eines VBO angegeben um welche Verwendung 
es sich handelt.

Obwohl praktisch alle aktuellen Grafiktreiber OpenGL in Versionen jenseits von 
1.5 unterstützen, ist die Methode die VBO Funktionen anzusprechen nicht immer
die gleiche. Unter Mac OS X sind die Funktionen beispielsweise direkt verfügbar.
Unter Windows hingegen stellt Microsoft nur veraltete OpenGL Biblioteheken zur
Verfügung und die Funktionen können nur über Funktionszeiger angesprochen 
werden, die vor der ersten Verwendung auch noch ermittelt werden müssen. Unter
Linux hat sich während der Entwicklung dieser Applikation ein gemischtes Bild
gezeigt. Während auch hier Funktionszeiger ermittelt werden müssen, sind die
Funktionen in einigen Fällen unter den standard OpenGL Namen (z.B. 
\lstinline!glBindBuffer()!) anzutreffen, während in anderen Fällen nur das 
Extension Namensschema (z.B. \lstinline!glBindBufferARB()!) verwendet werden 
kann.

Um diese Problematik vom Code der die VBOs nutzen möchte in der vorliegenden
Applikation zu entbinden, wurde die Initialisierung der Funktionszeiger und 
deren zentrale Speicherung in eine Klasse \emph{GLFunctions} ausgelagert, die 
von der Applikation über den \emph{VertexBufferState} angesprochen werden kann.

VBOs müssen vor der Verwendung ähnlich zu Display Listen erst initialisiert
werden. Eine neues VBO wird zunächst über \lstinline!glGenBuffers()!
angefordert. Danach wird ein VBO mit \lstinline!glBindBuffer()! aktiviert und 
via \lstinline!glBufferData()! mit Daten gefüllt. Es gibt dabei zwei Arten von 
Buffer Objects. Die erste Art speichert Vertexdaten wie Positionen, Normalen 
und Farben, die zweite speichert Indexdaten. Im letzteren Fall spricht man dann 
auch von Element Buffer Objects, kurz EBOs. Es können entweder einzelne VBOs
für die verschiedenen Typen von Vertexdaten angelegt werden, oder die Daten
verwoben in einem einzigen VBO gespeichert werden. Siehe dazu auch 
\cite{vbo:tutorial}.

Für das Rendering der Daten im kd-Tree wurde der erste Ansatz gewählt. Einmalig
werden die VBOs in \emph{mesh::VertexBufferLeaf::setupRendering()} 
initialisiert, und stehen fortan für das Rendering zur Verfügung.

\begin{lstlisting}
GLuint buffers[4];

buffers[VERTEX_OBJECT] = state.newBufferObject( ... );
gl->bindBuffer( GL_ARRAY_BUFFER, buffers[VERTEX_OBJECT] );
gl->bufferData( GL_ARRAY_BUFFER, _vertexLength * sizeof( Vertex ),
               &_globalData.vertices[_vertexStart], GL_STATIC_DRAW );

...

buffers[INDEX_OBJECT] = state.newBufferObject( ... );
gl->bindBuffer( GL_ELEMENT_ARRAY_BUFFER, buffers[INDEX_OBJECT] );
gl->bufferData( GL_ELEMENT_ARRAY_BUFFER, 
               _indexLength * sizeof( ShortIndex ),
               &_globalData.indices[_indexStart], GL_STATIC_DRAW );
\end{lstlisting}

Um die in den VBOs und dem EBO vorbereiteten Daten zu rendern, ist zunächst
jedes Buffer Object wieder mit \lstinline!glBindBuffer()! zu aktivieren. 
Für jedes VBO ist danach mit \lstinline!glXYZPointer()! das Format der Daten 
anzugeben. Nach dem Aktivieren des EBO können die Daten schliesslich mit 
einem Aufruf von \lstinline!glDrawElements()! gezeichnet werden.

\begin{lstlisting}
inline
void VertexBufferLeaf::renderBufferObject( VertexBufferState& state )
{
    const GLFunctions* gl = state.getGLFunctions();
    GLuint buffers[4];
    for( int i = 0; i < 4; ++i )
        buffers[i] = state.getBufferObject( ... );

    if( state.useColors() )
    {
        gl->bindBuffer( GL_ARRAY_BUFFER, buffers[COLOR_OBJECT] );
        glColorPointer( 4, GL_UNSIGNED_BYTE, 0, 0 );
    }
    gl->bindBuffer( GL_ARRAY_BUFFER, buffers[NORMAL_OBJECT] );
    glNormalPointer( GL_FLOAT, 0, 0 );
    gl->bindBuffer( GL_ARRAY_BUFFER, buffers[VERTEX_OBJECT] );
    glVertexPointer( 3, GL_FLOAT, 0, 0 );
    gl->bindBuffer( GL_ELEMENT_ARRAY_BUFFER, buffers[INDEX_OBJECT] );
    glDrawElements( GL_TRIANGLES, _indexLength, GL_UNSIGNED_SHORT, 0 );
}
\end{lstlisting}

Was ebenfalls noch beachtet werden muss ist, dass vor dem Rendering der VBOs
der Vertex Array State vom Client aktiviert werden muss. Dies erledigt in
der Applikation die Methode \emph{mesh::VertexBufferRoot::beginRendering()}.

Wie schon beim Vergleich des Rendering Code ersichtlich wird, benötigen VBOs
bedeutend mehr Funktionsaufrufe als Display Listen. Entsprechend gibt es für
die Grösse von VBOs eine Untergrenze bei deren Unterschreitung die Leistung
einbricht. Ebenfalls zu beachten ist dass die Leistung von VBO Rendering stark
zusammenfällt, wenn die falschen Datentypen verwendet werden. Der Hintergrund 
ist dabei, dass die Grafikhardware in diesem Fall bei jedem Rendering 
Konvertierungen ins korrekte, intern verwendete Format vornehmen muss. Für 
weitere Benchmarks und Hinweise zur Verwendung von VBOs, siehe auch 
\cite{vbo:benchmarks}.

%Obwohl bei der Entwicklung der vorliegenden Applikation alle Hinweise in Bezug 
%auf Datenformate und Puffergrössen beachtet wurden, bleibt die Leistung des VBO
%Rendering in vielen Fällen aus bisher unbekannten Gründen dennoch hinter den 
%Display Listen zurück.

\section{Shading Algorithmen}

Nachdem die grundlegenden Rendermöglichkeiten in der letzten Sektion gezeigt
wurden, sollen in dieser Sektion nun die verschiedenen Shading Algorithmen 
sowohl in der Theorie erklärt als auch in der praktischen Implementierung 
vorgestellt werden. Als Hintergrund wird dazu zunächst das Beleuchtungsmodell
diskutiert, welche in den meisten Echtzeit Computergrafik Systemen verwendet
wird. Eine ausführliche Behandlung dieser Themen findet sich beispielsweise in
\cite{angel:icg}.

\subsection{Beleuchtungsmodell}

Das heutzutage in der (Echtzeit) Computergrafik am häufigsten anzutreffende 
Beleuchtungsmodell stammt von Phong und wurde bereits vor mehr als 30 Jahren
publiziert. Es bildet die Realität nicht exakt ab, erreicht aber in vielen
Situationen akzeptable bis gute Ergebnisse.

Das Phong Modell ist ein lokales Beleuchtungsmodell, es berücksichtigt nur 
Reflektionen die ein Objekt selbst durch direkte Bestrahlung einer Lichtquelle
erzeugt. Lichtreflektionen die von anderen Objekten stammen werden nicht in
Betracht gezogen. Um diesen Verlust auszugleichen, enthält das Modell eine
Komponente für globales \emph{Ambient Light}. Das reflektierte Licht besteht
aus zwei Komponten, dem \emph{Diffuse Light} sowie dem \emph{Specular Light}).
\[I_{total} = I_{ambient} + I_{diffuse} + I_{specular}\]

\glossary{name={Light, Ambient},description={Umgebungslicht.}}
\glossary{name={Light, Diffuse},description={Von rauen Oberflächen 
gleichmässig in alle Richtungen reflektiertes Licht.}}
\glossary{name={Light, Specular},description={Von glatten Oberflächen
spiegelnd reflektiertes Licht.}}

Unterschiedliche Materialien reflektieren Licht auf unterschiedliche Weise. Um
dies zu modellieren, werden jedem Objekt drei Materialkoeffizienten zugewiesen,
welche den drei Lichtkomponenten entsprechen und bestimmen welches Licht vom 
Objekt wie stark reflektiert wird. Neben diesen drei Koeffizienten wird auch
noch ein vierter verwendet, der beeinflusst wie stark die Objektoberfläche
spiegelt und \emph{Shininess} genannt wird.

\glossary{name={Shininess},description={Glanz, ein Koeffizient der angibt wie 
stark eine Objektoberfläche einfallendes Licht spiegelt.}}

\subsubsection*{Ambient Light}

Das Ambient Light ist global, beleuchtet also jeden Punkt eines jeden Objektes
mit der gleichen Intensität. Wenn die Intensität des Ambient Light $L_{a}$ ist 
und der entsprechende Materialkoeffizient $k_{a}$, so beträgt die reflektierte 
Intensität 
\[I_{a} = k_{a}L_{a}\text{.}\]

\subsubsection*{Diffuse Light}

Diffuse Light wird von einer rauen Oberfläche in alle Richtungen gleichmässig
reflektiert, und erscheint somit für jeden Beobachter gleich. Allerdings ist 
die reflektierte Intensität abhängig von der Lage der Oberfläche und des
Lichtes zueinander. Im Phong Modell wird diese Komponente daher mit Hilfe des 
Winkels zwischen dem Lichtvektor $l$ und der Flächennormalen $n$ modelliert. 
Da der Cosinus des Winkels bei Einheitslänge der Vektoren dem Skalarprodukt 
entspricht, ergibt sich für die Intensität somit
\[I_{d} = k_{d}L_{d}(l \cdot n)\text{.}\]

\subsubsection*{Specular Light}

Spiegelnde Reflektionen stammen von glatten Oberflächen. Das Licht fällt aus
Richtung des Lichtvektors $l$ ein, und wird in Richtung des Reflektionsvektors
$r$ zurückgeworfen. Der davon sichtbare Anteil hängt vom Winkel zwischen dem
(perfekten) Reflektionsvektor und dem Beobachter ab. Nicht jede Oberfläche
reflektiert allerdings perfekt, und die daraus resultierende Streuung der 
tatsächlichen Reflektion und deren Einfluss auf die Intensität wird durch den
\emph{Shininess} Koeffizienten $\alpha$ modelliert. Mit $v$ als Einheitsvektor 
in Richtung des Beobachters, beträgt die Intensität also
\[I_{s} = k_{s}L_{s}(r \cdot v)^{\alpha}\text{.}\]

\subsubsection*{Vollständige Formel}

Wenn die drei verschiedenen Komponenten nun zusammengesetzt werden, und 
möglicherweise negative Komponenten ausgeschlossen werden, ergibt sich die 
folgende vollständige Formel für das Phong Beleuchtungsmodell (unter
Vernachlässigung dass Intensität mit zunehmender Entfernung zwischen der
Lichtquelle und dem Objekt abfällt):

\[I = k_{a}L_{a} + k_{d}L_{d}max(l \cdot n, 0) + 
      k_{s}L_{s}max((r \cdot v)^{\alpha}, 0)\]

Um die Berechnung von $r$ zu vermeiden, hat Blinn eine Modifikation des Phong
Modells vorgeschlagen. Er verwendet den sogenannten \emph{Halfway} Vektor
$h = \frac{l + v}{\left|l + v\right|}$ und approximiert dann $r \cdot v$ in 
der Formal mit $n \cdot h$. Um ein optisch ähnliches Ergebnis wie beim reinen
Phong Modell zu erhalten, muss nach dieser Approximation der \emph{Shininess} 
Koeffizient $\alpha$ angepasst werden.

\subsection{Flat Shading}

Theoretisch unterscheiden sich sowohl der Lichtvektor $l$ als auch die Richtung 
zum Beobachter $v$ für jeden Punkt einer Oberfläche. Um die Shading Berechnung 
zu vereinfachen, können sowohl die Lichtquelle als auch der Beobachter als 
unendlich weit entfernt betrachtet werden, womit aus beiden Vektoren konstante 
Richtungen werden. Für plane Oberflächen ist die Flächennormale $n$ ebenfalls
an jedem Punkt konstant.

Wenn alle drei Vektoren konstant sind, muss das Shading der Oberfläche nur an
einem Punkt berechnet werden. Das Ergebnis sind Flächen die über den ganzen
Bereich dieselbe Farbtönung tragen. Dieser Algorithmus wird als \emph{Flat 
Shading} bezeichnet.

Weil die alte Datenstruktur nur Flächennormalen gespeichert hat, war Flat 
Shading der einzige Algorithmus der mit der alten Beispielapplikation 
realisiert werden konnte. Da dies optisch nicht ansprechend und auch nicht 
mehr zeitgemäss ist, wird in der neuen Applikation auf Flat Shading ganz 
verzichtet.

\subsection{Gouraud Shading}

Eür ein korrektes Shading gekrümmter Oberflächen wäre es notwendig, die
betroffenen Vektoren an jedem einzelnen Punkt zu kennen. Eine brauchbare 
Approximation ist es, das Shading nur an Vertices zu berechnen und dann 
über die Oberfläche hinweg zu interpolieren. Dieses Verfahren wird 
\emph{Gouraud Shading} genannt.

Für Gourauf Shading wird je eine Normale pro Vertex benötigt. Diese lässt 
sich durch Mittelung der Flächennormalen sämtlicher angrenzender Flächen 
vernünftig schätzen. Die neue Applikation unterstützt Gouraud Shading direkt, 
da unmittelbar nach Einlesen des Modells die entsprechenden Vertexnormalen 
berechnet werden.

\subsection{Phong Shading}

Gouraud Shading hat immer noch einen entscheidenen Nachteil. Weil das Shading
nur an den Vertices berechnet und über die Öberfläche interpoliert wird, gehen 
spiegelnde Reflektionen inmitten einer Fläche verloren oder werden über die 
gesamte Fläche verschmiert.

Eine weitere Verbesserung ist, anstelle des Shading die Normalen selbst zu
interpolieren. Dieser Algorithmus ist das \emph{Phong Shading}. Das Problem
mit Phong Shading ist, dass Shading nicht mehr pro Vertex berechnet werden 
kann, sondern per Pixel berechnet werden muss. Dies war bis zur Einführung von
programmierbaren Fragment Shadern in der Grafikhardware nicht in Echtzeit
möglich.

Die neue Applikation enthält ein Gerüst für den Einsatz von Vertex und Fragment
Shadern, die mit der OpenGL Shading Language geschrieben wurden. Als einfache 
Beispiele werden Shader mitgeliefert, welche das Phong Shading implementieren.

\section{Weitere Algorithmen}

Aufgrund fehlender Zeit konnten leider keine weiteren Algorithmen recherchiert
und umgesetzt werden, welche die optische Attraktivität der Darstellung weiter 
erhöhen würden. Da aber ein Gerüst zur Unterstützung von GLSL Shadern vorhanden
ist, sollten sich weitere Algorithmen zu einem späteren Zeitpunkt ohne grosse 
Probleme zur Applikation hinzufügen lassen.

%
% EOF
%

### 2007-09-26 Developer/eqPly/chapters/datastructure.tex
Updated /doc/Developer/eqPly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 \centering
-\begin{tabular}{lrr}
-\textbf{Modell} & \textbf{Dateigrösse Octree} & \textbf{Dateigrösse kd-Tree} \\
-Armadillo & 27.72 MB & 6.09 MB \\
-Dragon & 69.83 MB & 15.34 MB \\
-Happy Buddha & 87.16 MB & 19.05 MB \\
-Lucy & n/a & 494.58 MB \\
-Stanford Bunny & 5.56 MB & 1.20 MB \\
+\begin{tabular}{lrrr}
+\textbf{Modell} & \textbf{Dateigrösse Octree} & \textbf{Dateigrösse kd-Tree} &
+\textbf{Verhältnis}\\
+Armadillo & 27.72 MB & 6.09 MB & 21.96 \%\\
+Dragon & 69.83 MB & 15.34 MB & 21.97 \%\\
+Happy Buddha & 87.16 MB & 19.05 MB & 21.86 \%\\
+Lucy & n/a & 494.58 MB & n/a\\
+Stanford Bunny & 5.56 MB & 1.20 MB & 21.56 \%\\
 \end{tabular}

### 2007-09-26 Developer/eqPly/chapters/conclusion.tex
Updated /doc/Developer/eqPly with benchmark results.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 Speicherverbraucht wurde auf weniger als ein Viertel reduziert (siehe Tabelle
-\ref{tbl:file-sizes}.
+\ref{tbl:file-sizes}).
 
@@
 Rendering mit Vertex Buffers Objects ist implementiert und funktioniert. Auch
-Rendering mit Display Listen wird weiterhin angeboten für Umgebungen ohne VBO 
-Unterstützung.
+Rendering mit Display Listen wird als OpenGL 1.1 kompatible Lösung weiterhin 
+angeboten, und bleibt zur Zeit noch die Standardeinstellung. VBO Verwendung
+kann über einen Kommandozeilenparameter aktiviert werden.
 
@@
 Applikation findet sich in folgender Tabelle \ref{tbl:render-performance}. 
-Die verwendeten Modelle stammen aus \cite{stanford:repository} und 
-\cite{cyberware}.
+In einigen Fällen bleibt die VBO Performance aus bisher ungeklärten Gründen 
+hinter den Erwartungen zurück, während sie in anderen Fällen einen markanten
+Leistungsvorteil zeigt. Die verwendeten Modelle stammen aus 
+\cite{stanford:repository} und \cite{cyberware}.
 
@@
 \centering
-\begin{tabular}{lrrr}
-\textbf{Modell} & \textbf{eqPly (alt)} & \textbf{eqPly (neu, DL)} & 
-\textbf{eqPly (neu, VBO)}\\
-xyz & 1.0 FPS & 2.0 FPS & 3.0 FPS
+\begin{tabular}{lrrrr}
+\textbf{Modell} & \textbf{Anzahl} & \textbf{eqPly (alt)} & 
+\textbf{eqPly (neu)} & \textbf{eqPly (neu)}\\
+\textbf{} & \textbf{Triangles} & \textbf{DL [FPS]} & 
+\textbf{DL [FPS]} & \textbf{VBO [FPS]} \\
+Stanford Bunny   &     69'451  &  406.82   &   436.67   &   617.00\\
+RockerArm        &     80'354  &  445.61   &   445.57   &   301.67\\
+Goddess1         &    274'822  &  142.20   &   162.45   &    70.94\\
+Armadillo        &    345'944  &  118.30   &   136.43   &    92.17\\
+Boat             &    776'374  &   29.25   &    20.06   &    28.47\\
+Dragon           &    871'414  &   17.06   &    45.02   &    38.19\\
+Goddess2         &  1'047'330  &   10.02   &     9.32   &    42.99\\
+Hip              &  1'060'346  &   11.10   &     9.24   &    29.37\\
+Happy Buddha     &  1'087'716  &    9.54   &    29.27   &    34.88\\
 \end{tabular}
@@
 
-%Leider
-%bleibt die Performance aber in vielen Fällen aus bisher ungeklärten Gründen 
-%hinter den Erwartungen zurück. Das Rendering mit Display Listen steht aber nach
-%wie vor als alternative Lösung zur Verfügung, und bleibt derzeit auch noch das 
-%Standard Verfahren.
-
-Das Rendering ist durch die standardmässige Nutzung von Gouraud anstelle von
-Flat Shading bereits optisch ansprechender als in der alten Applikation. Dies
-ist in Abbildungen \ref{fig:sample-flat} und \ref{fig:sample-smooth} deutlich
-zu sehen.
+Das Rendering ist durch die standardmässige Nutzung von Gouraud anstelle 
+von Flat Shading bereits optisch deutlich ansprechender als in der alten 
+Applikation. Dies ist in Abbildungen \ref{fig:sample-flat} und 
+\ref{fig:sample-smooth} am Beispiel des RockerArm Modells zu sehen.
 

### 2007-09-26 Developer/eqPly/chapters/datastructure.tex
Updated /doc/Developer/eqPly with benchmark results.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

--- 2007-09-26 Developer/eqPly/chapters/datastructure.tex
@@
 
-Als erstes soll die Baumstruktur besser ausbalanciert sein, und eine genauere
+Erstens soll die Baumstruktur besser ausbalanciert sein, und eine genauere
 Kontrolle über die Blattgrössen aufgrund der Verwendung von VBOs beim Rendering
@@
 
-Als zweites sollen die Informationen über die Struktur der polygonalen Daten 
-erhalten bleiben, also die Nachbarschaftsbeziehungen zwischen den Triangles. 
-Dies einerseits, damit bei der Speicherung und beim Rendering Vorteile aus
-den gemeinsam genutzten Vertices gezogen werden können, und andererseits um neu
-Vertexnormalen anstelle der Flächennormalen berechnen zu können. Diese erlauben 
-erst die Verwendung besserer Shading Algorithmen.
+Zweitens sollen die Informationen über die Struktur der polygonalen Daten 
+erhalten bleiben, also die Nachbarschaftsbeziehungen zwischen Triangles. Dies 
+einerseits, um bei Speicherung und Rendering Vorteile aus den gemeinsam
+genutzten Vertices ziehen zu können, und andererseits um Vertexnormalen 
+anstatt von Flächennormalen berechnen zu können. Diese erlauben erst die 
+Verwendung besserer Shading Algorithmen.
 
-Als drittes soll der neue Code heutigen objektrientierten Ansprüchen genügen,
+Drittens soll der neue Code heutigen objektrientierten Ansprüchen genügen,
 die Möglichkeiten von C++ besser ausnutzen und natürlich einfach verständlich
@@
     ...
+
     data.sort( start, length, axis );
@@
     }
+
     ...
@@
 \centering
-\begin{tabular}{lrr}
-\textbf{Modell} & \textbf{Dateigrösse Octree} & \textbf{Dateigrösse kd-Tree} \\
-Armadillo & 27.72 MB & 6.09 MB \\
-Dragon & 69.83 MB & 15.34 MB \\
-Happy Buddha & 87.16 MB & 19.05 MB \\
-Lucy & n/a & 494.58 MB \\
-Stanford Bunny & 5.56 MB & 1.20 MB \\
+\begin{tabular}{lrrr}
+\textbf{Modell} & \textbf{Binary Octree} & 
+\textbf{Binary kd-Tree} & \textbf{Verhältnis}\\
+\textbf{} & \textbf{[MB]} & 
+\textbf{[MB]} & \textbf{[\%]}\\
+Armadillo      &  27.72 &   6.09 & 21.96\\
+Dragon         &  69.83 &  15.34 & 21.97\\
+Happy Buddha   &  87.16 &  19.05 & 21.86\\
+Lucy           &    n/a & 494.58 &   n/a\\
+Stanford Bunny &   5.56 &   1.20 & 21.56\\
 \end{tabular}

### 2007-09-27 Developer/eqPly/chapters/application.tex
Updated /doc/Developer/eqPly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation
Fixed typos in /doc/Developer/eqPly and updated eq UML figure.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 Klasse \emph{eq::Window}, welche einzelne OpenGL \emph{Drawables} und 
-\emph{Contexts} verwalten. Jedes Fenster kann abschliessendend noch mehrere 
+\emph{Contexts} verwalten. Jedes Fenster kann abschliessend noch mehrere 
 \emph{Viewports} besitzen, die durch die Klasse \emph{eq::Channel} 
@@
 Aufgrund der relativ klar vorgegebenen Struktur für Equalizer Programme weicht 
-die Architektur der neuen Applikation kaum von der des alten Beispielsprogrammes
+die Architektur der neuen Applikation kaum von der des alten Beispielprogrammes
 ab. Abbildung \ref{fig:eq-classes} am Ende der Sektion zeigt die Zusammenhänge 
@@
 sind in \emph{eqPly} vor allem die Rotations- und Translationsmatrizen für
-die Kameraposition, die sich sowohl durch Benutzerinteraktion als auch durch
+die Kameraposition, welche sich sowohl durch Benutzerinteraktion als auch durch
 kontinuierliche Animation fortlaufend ändert.
@@
 Anwendung der verschiedenen Modell- und Projektionstransformationen sowie das
-View Frustom Culling und Rendering des Modells.
+View Frustum Culling und Rendering des Modells.
 
@@
 
-\begin{sidewaysfigure}[hp]
+\begin{figure}[ht]
 \centering
-\includegraphics[width=1.00\textheight]{figures/UML.pdf}
+\includegraphics[width=1.00\textwidth]{figures/UML.pdf}
 \caption[Zusammenhänge der Klassen von eqPly und Equalizer]
@@
 \label{fig:eq-classes}
-\end{sidewaysfigure}
+\end{figure}
 

### 2007-09-27 Developer/eqPly/chapters/conclusion.tex
Fixed typos in /doc/Developer/eqPly and updated eq UML figure.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 und für weitere Aufrufe verfügbar sind, ist dieser Punkt vernachlässigbar. Der
-Speicherverbraucht wurde auf weniger als ein Viertel reduziert (siehe Tabelle
+Speicherverbrauch wurde auf weniger als ein Viertel reduziert (siehe Tabelle
 \ref{tbl:file-sizes}).
@@
 beispielsweise, ein teilweises Laden nur der benötigen Bereiche zu ermöglichen.
-Die Effizienz des View Frustum Cullings liesse sich unter Umständen verbessern
+Die Effizienz des View Frustum Cullings liesse sich unter Umständen verbessern,
 indem statt beim Aufbau die Sortierachse zu rotieren jeweils diejenige mit der
@@
 Stanford Bunny   &      69'451  &  406.82   &   436.67   &   617.00\\
-RockerArm        &     80'354  &  445.61   &   445.57   &   301.67\\
-Goddess1         &    274'822  &  142.20   &   162.45   &    70.94\\
+Rocker Arm       &      80'354  &  445.61   &   445.57   &   301.67\\
+Goddess 1        &     274'822  &  142.20   &   162.45   &    70.94\\
 Armadillo        &     345'944  &  118.30   &   136.43   &    92.17\\
+Ganesh           &     413'236  &           &            &         \\
 Boat             &     776'374  &   29.25   &    20.06   &    28.47\\
 Dragon           &     871'414  &   17.06   &    45.02   &    38.19\\
-Goddess2         &  1'047'330  &   10.02   &     9.32   &    42.99\\
+Goddess 2        &   1'047'330  &   10.02   &     9.32   &    42.99\\
 Hip              &   1'060'346  &   11.10   &     9.24   &    29.37\\
 Happy Buddha     &   1'087'716  &    9.54   &    29.27   &    34.88\\
+Asian Dragon     &   7'219'045  &           &            &         \\
+Thai Statue      &  10'000'000  &           &            &         \\
 \end{tabular}

### 2007-09-27 Developer/eqPly/chapters/datastructure.tex
Fixed typos in /doc/Developer/eqPly and updated eq UML figure.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-Der Kern der alten Datenstruktur ist ein \emph{Bounding Box Octree 
-\cite{octree:desc}}. Die \emph{Bounding Boxes} sind achsenparallele Quader 
+Der Kern der alten Datenstruktur ist ein \emph{Bounding Box Octree} 
+\cite{octree:desc}. Die \emph{Bounding Boxes} sind achsenparallele Quader 
 welche die enthaltenen Daten minimal umschliessen. Der \emph{Octree} ist eine 
@@
 Die Klasse \emph{PlyModel} bietet basierend auf dieser Struktur eine Reihe von 
-Methoden an. Das Erstellen und Füllen des Bounding Box Octree wird dabei genau 
+Methoden an. Das Erstellen und Füllen des Bounding Box Octrees wird dabei genau
 so unterstützt wie das Speichern und Laden einer binären Repräsentation. Die
@@
 
-Drittens soll der neue Code heutigen objektrientierten Ansprüchen genügen,
+Drittens soll der neue Code heutigen objektorientierten Ansprüchen genügen,
 die Möglichkeiten von C++ besser ausnutzen und natürlich einfach verständlich
@@
 den kd-Tree abstrahiert. Dies sind insbesondere der zu verwendende Rendermodus
-(sieht Kapitel \ref{rendermodes}), die Verwendung von Farben sowie die
-Verwaltung von OpenGL Objekten und Erweiterungsfunktionen. Die Klasse selbst 
+(siehe Kapitel \ref{rendermodes}), die Verwendung von Farben sowie die
+Verwaltung von OpenGL Objekten und Extension Functions. Die Klasse selbst 
 ist abstrakt, konkrete Versionen sind für eine stand-alone Nutzung 
@@
 des Intervalls. Da alle diese Methoden von Klienten nur auf dem Wurzelknoten 
-aufgerufen werden sollen sind sie auf allgemeiner Knotenebene \emph{protected}.
+aufgerufen werden sollen, sind sie auf allgemeiner Knotenebene \emph{protected}.
 
@@
 
-Auf die für das Rendering relevanten Methoden von \emph{mesh::VertexBufferLeaf}
-sei auf Kapitel \ref{rendering} verwiesen, sie werden dort genauer behandelt.
+Die für das Rendering relevanten Methoden von \emph{mesh::VertexBufferLeaf}
+werden in Kapitel \ref{rendering} genauer betrachtet, und entsprechend hier
+nicht weiter behandelt.
 

### 2007-09-27 Developer/eqPly/chapters/rendering.tex
Updated /doc/Developer/eqPly with benchmark results.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation
Fixed typos in /doc/Developer/eqPly and updated eq UML figure.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 werden einzeln abgesetzt und direkt ausgeführt. Dies bedeutet für den konkreten
-Fall des Rendering unserer polygonalen Daten, dass für jedes zu zeichnende 
+Fall des Renderings unserer polygonalen Daten, dass für jedes zu zeichnende 
 Triangle je drei Aufrufe für das Setzen von Vertexpositionen, Normalen und
@@
 die gleiche. Unter Mac OS X sind die Funktionen beispielsweise direkt verfügbar.
-Unter Windows hingegen stellt Microsoft nur veraltete OpenGL Biblioteheken zur
+Unter Windows hingegen stellt Microsoft nur veraltete OpenGL Bibliotheken zur
 Verfügung und die Funktionen können nur über Funktionszeiger angesprochen 
@@
 
-%Obwohl bei der Entwicklung der vorliegenden Applikation alle Hinweise in Bezug 
-%auf Datenformate und Puffergrössen beachtet wurden, bleibt die Leistung des VBO
-%Rendering in vielen Fällen aus bisher unbekannten Gründen dennoch hinter den 
-%Display Listen zurück.
+Obwohl bei der Entwicklung der vorliegenden Applikation alle Hinweise in Bezug 
+auf Datenformate und Puffergrössen beachtet wurden, bleibt die Leistung des VBO
+Rendering in einigen Fällen aus bisher unbekannten Gründen dennoch hinter den 
+Display Listen zurück.
 
@@
 vorgestellt werden. Als Hintergrund wird dazu zunächst das Beleuchtungsmodell
-diskutiert, welche in den meisten Echtzeit Computergrafik Systemen verwendet
+diskutiert, welches in den meisten Echtzeit Computergrafik Systemen verwendet
 wird. Eine ausführliche Behandlung dieser Themen findet sich beispielsweise in

### 2007-09-28 Developer/ProgrammingGuide/paper.tex
PG: update to v0.8
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   \textbf{INCOMPLETE}\\[\medskipamount]
-  Version 0.7, \today
+  Version 0.8, \today
 }
@@
 \newcommand{\fig}[1]{Figure~\ref{#1}}
+\newcommand{\sref}[1]{Section~\ref{#1}}
 \newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}
@@
 \begin{figure}[ht]
-  \hfill
-  \includegraphics[width=4cm]{images/logo.pdf}\hfill
-  \includegraphics[width=4cm]{images/vmml.pdf}\hfill\vspace{-1em}\\
+  \center
+  \includegraphics[width=4cm]{images/logo.pdf}
+%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
+\end{figure}
+\vfill
+\begin{figure}[ht]
+  \center
+  \includegraphics[width=4cm]{images/vmml.pdf}
 %  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
@@
     \hline
+    0.8     & Sep 28, 2007 & added head tracking, finished channel,
+                             proof-reading pass\\
     0.7     & Sep 21, 2007 & started channel and image compositing,
@@
 Equalizer provides a framework for the development of parallel
-OpenGL\texttrademark\ 
-applications. Equalizer-based applications can run a single
-shared-memory system with multiple graphics cards (GPU's) or on a
-distributed graphics cluster. This Programming Guide introduces the
-programming interface using the \textsf{eqPly} example shipped with
-Equalizer as a guideline.
+OpenGL\texttrademark\ applications. Equalizer-based applications can run
+a single shared-memory system with multiple graphics cards (GPU's) or on
+graphics clusters. This Programming Guide introduces the programming
+interface using the \textsf{eqPly} example shipped with Equalizer as a
+guideline.
 
@@
 
-Equalizer is the next evolution of a generic parallel programming
-interface for visualization applications using OpenGL. Existing
+Equalizer is the next evolution of generic parallel programming
+interfaces for visualization applications using OpenGL. Existing
 solutions, such as SGI's OpenGL Multipipe\texttrademark\ SDK, VRCO's
@@
 so that the OpenGL rendering can be executed in parallel, potentially
-across multiple processes for cluster-based execution. Equalizer
-provides the domain-specific parallel rendering know-how and abstracts
-threading, synchronization, windowing and event handling for
-portability. It is a `GLUT on steroids', providing parallel and
+using multiple processes for cluster-based execution. Equalizer provides
+the domain-specific parallel rendering know-how and abstracts
+configuration, threading, synchronization, windowing and event handling
+for portability. It is a `GLUT on steroids', providing parallel and
 distributed execution, scalable rendering features and fully
@@
 All functionality of Equalizer is accessed through the Equalizer client
-library, which implements all functionality discussed in this document.
+library, which implements all functionality discussed in this
+document. The client library provides communication between the Equalizer
+processes.
 
@@
 An Equalizer server is responsible for managing one visualization
-system\footnote{a shared memory system or graphics cluster}. Currently
-it is only useful for running one application at a time, but it will be
+system, i.e., a shared memory system or graphics cluster. Currently it
+is only useful for running one application at a time, but it will be
 extended to support multiple applications concurrently and efficiently
@@
 
-The application connects to a server, which chooses a configuration for
-the application. It provides a render client, to be launched by the
-server. The application reacts on events and controls the rendering.
+The application connects to an Equalizer server, which chooses a
+configuration for the application. It also provides a render client, to
+be launched by the server. The application reacts on events and controls
+the rendering.
 
@@
 
-The render client implements the rendering part of an application. It is
-passive, and receives all its rendering tasks from the server. The tasks
-are executed by calling the appropriate task methods (see
-\ref{ssTaskMethods}).
+The render client implements --obviously-- the rendering part of an
+application. Its execution is passive, completely driven by rendering
+tasks received from the server. The tasks are executed by calling the
+appropriate task methods (see \sref{ssTaskMethods}) in the correct
+thread and context.
 
 The application might be a rendering client, in which case it can also
-contribute to the rendering. It can choose not to implement any render
-client-related code, in which case it is reduced to be the application's
-'master' process without any OpenGL windows.
+contribute to the rendering. If it does not implement any render
+client-related code it is reduced to be the application's 'master'
+process without any OpenGL windows.
 
 The rendering client can be the same executable as the application, as
-is the case with \textsf{eqPly}. Real-world applications often implement
-a separate, light-weight rendering client.
+is the case with \textsf{eqPly}. When it is started as a render client,
+the Equalizer initialization routine does not return and takes over the
+control by calling the render client task methods. Complex real-world
+applications often implement a separate, light-weight rendering client.
 
@@
 invasive, that is, Equalizer imposes only the minimal, natural execution
-framework upon the application. It does not impose a scene graph or does
-interfere in any way with the application's rendering code.
-
-Methods called by the application always have the form
-\textsf{verb[Noun]}, whereas methods called by Equalizer (`Task
-Methods') have the form \textsf{nounVerb}.
+framework upon the application. It does not provide a scene graph or
+does interfere in any other way with the application's rendering
+code. The restructuring work needed for Equalizer is needed anyway in
+order to parallelize the application for rendering.
+
+Methods called by the application have the form \textsf{verb[Noun]},
+whereas methods called by Equalizer (`Task Methods') have the form
+\textsf{nounVerb}. For example the application calls
+\textsf{Config::startFrame} to render a new frame, which causes --among
+other things-- \textsf{Node::frameStart} to be called in all active
+render clients.
 
@@
 The root of the resource tree is the \textsf{eq::Config}, which
-represents the current configuration of the application. It currently
-only holds the local node, not all nodes of the configuration.
+represents the current configuration of the application. The
+configuration is the session in which all render clients are
+registered. The config instance on a given node currently only holds the
+local node, not all nodes of the configuration.
 
@@
 An \textsf{eq::Node} is the representation of a single computer in the
-system. It is one operating system process of the render client. All
-node task methods are executed from the main application thread. Each
-configuration might also use an application node, in which case the
-application process is also used for rendering.
+system. One operating system process of the render client will be used
+for each node. Each configuration might also use an application node, in
+which case the application process is also used for rendering. All
+node-specific task methods are executed from the main application
+thread.
 
@@
 the current implementation it is also one operating system thread,
-unless the pipe's thread hint is set to \textsf{false}. All pipe and
-child window and channel task methods are executed from the pipe thread
-for threaded pipes or from the main application thread for non-threaded
+unless the pipe's thread hint is set to \textsf{false}. All pipe, window
+and channel task methods are executed from the pipe thread for threaded
+pipes or from the main application thread for non-threaded
 pipes\footnote{see also
@@
 
-An \textsf{eq::Window} is an drawable and OpenGL context. The drawable
-can be an on-screen window or an off-screen PBuffer or
+An \textsf{eq::Window} holds a drawable and OpenGL context. The
+drawable can be an on-screen window or an off-screen PBuffer or
 FBO\footnote{off-screen drawables are not yet implemented, but can be
-  created by the application and used with Equalizer}.
+  created by the application and used with Equalizer}. The window holds
+window-system-specific handles to the drawable and context, e.g., an X11
+window \textsf{XID} and \textsf{GLXContext} for the glX window system.
 
@@
 The \textsf{eq::Channel} is the abstraction of an OpenGL viewport within
-its parent window. It is the entity executing the actual rendering.
-
-\subsection{Resource Usage}
+its parent window. It is the entity executing the actual rendering. The
+channel's viewport might be overwritten when it is rendering for another
+channel.
+
+\subsection{Compounds}
 
 How the rendering resources are to be used is configured using a
-compound tree. Each compound has a channel, which it uses to execute the
-rendering tasks. The rendering tasks are computed by the server and send
-to the render clients. At no point the application or render clients
-have or need knowledge of compounds. The configuration of compounds is
-not in the scope of this document\footnote{see
+compound tree. Although the API does not currently expose compounds, the
+basic design behind compounds is explained here to increase the
+understanding of the architecture behind Equalizer.
+
+Each compound has a channel, which it uses to execute the rendering
+tasks. One channel might be used by multiple compounds. Unused channels
+are not instantiated. The rendering tasks for the channels are computed
+by the server and send to the appropriate render clients.
+
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/frustra.pdf}
+  {\caption{\small\label{fFrustra}Frustrum descriptions}}
+\end{wrapfigure}
+The channels of the leaf compounds (source channels) in the compound
+tree execute the rendering for the top-most compound's channel
+(destination channel). The top-most compound with a channel defines the
+2D pixel viewport rendered by all leaf compounds. The destination
+channel and pixel viewport can not be overridden.
+
+Compounds have a frustum description, which is inherited to the
+children. The frustum can be overridden by children, but this is rarely
+used since there is no use case for this behaviour. The frustrum can be
+specified as a wall, which is completely defined by the bottom-left,
+bottom-right and top-left coordinates relative to the origin, or as a
+projection, which is defined by the position and head-pitch-roll
+orientation of the projector, as well as the horizontal and vertical
+field-of-view and distance to the projection wall. \fig{fFrustra}
+illustrates the wall and projection frustum parameters.
+
+Compounds have attributes which configure the decomposition of the
+destination's channel viewport, frustum and database. A
+\textsf{viewport} decomposes the destination channel and frustum in
+screen space. A \textsf{range} tells the application to render a part of
+its database, and an eye pass can selectively render different stereo
+passes.
+
+Compounds have output and input frames which configure the recomposition
+of the resulting pixels from the source compound channels. An output
+frame connects to an input frame of the same name, and transports the
+selected frame buffer data from the output channel to the input
+channel. This composition process is extensively described in
+\sref{sCompositing}.
+
+More information on the configuration of compounds can be found on the
+Equalizer website\footnote{see
   \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.
@@
 The \textsf{eqPly} example is shipped with the Equalizer distribution
-and serves as a simple reference implementation of an Equalizer-based
-application. Its focus is not on rendering features or visual quality.
-It serves as a test bed for most of the Equalizer features.
+and serves as a reference implementation of an Equalizer-based
+application. Its focus is not on rendering features or visual quality,
+but it is used as a test bed for most of the Equalizer features.
 
@@
 All classes in the example are in the \textsf{eqPly} namespace to avoid
-type name ambiguities, in particular for the \textsf{Window} class.
+type name ambiguities, in particular for the \textsf{Window}
+class. \fig{fUml} shows the most important Equalizer classes and how
+they are subclassed in the \textsf{eqPly} example. Each of the classes
+in the \textsf{eqPly} namespace will be discussed in this section.
+
+\begin{figure}[ht!]\center
+  \includegraphics[width=\textwidth]{images/uml}
+  {\caption{\small\label{fUml}UML diagram of significant Equalizer and
+      eqPly classes and task methods.}}
+\end{figure}
 
@@
 
-Versioned objects have to override \textsf{isStatic} to return false, in
-order for Equalizer to know that they should be versioned. Right now,
-only the master instance of the object is writable, that is,
-\textsf{eqNet::Object::commit} can be called to generate a new
+Versioned objects have to override \textsf{isStatic} to return false, to
+indicate that they are versioned. The current implementation has the
+following characteristics:
+\begin{itemize}
+\item Only the master instance of the object is writable, that is,
+  \textsf{eqNet::Object::com\-mit} can be called to generate a new
   version.
+\item Slave instance versions can only be advanced, that is,
+  \textsf{eqNet::Object::sync( version )} with a version smaller than
+  the current version will fail.
+\end{itemize}
 
@@
 glX/X11, WGL and AGL/Carbon. Some operating systems, and therefore some
-Equalizer versions, support multiple window systems\footnote{see
+Equalizer versions, support multiple window systems
+concurrently\footnote{see
   \link{http://www.equalizergraphics.com/compatibility.html}}.
@@
 \end{lstlisting}}%>>
+
+Parts of the Carbon API used for window and event handling in the AGL
+window system are not thread safe. The application has to call
+\textsf{eq::Global::enterCarbon} before any thread-unsafe Carbon call,
+and \textsf{eq::Global::leaveCarbon} afterwards. These functions should
+be used only during window initialization and exit, not during
+rendering\footnote{for various reasons \textsf{enterCarbon} might block
+  up to 50 milliseconds}. Carbon call in the window event handling
+routine \textsf{Window::processEvent} are thread-safe. Please contact
+the Equalizer developer mailing list if you need to use Carbon calls on
+a per-frame basis.
 
@@
 routine contains the application's OpenGL rendering code, which is being
-rendered using the rendering context provided by Equalizer.
+rendered using the rendering context provided by Equalizer. As most of
+the other task methods, \textsf{frameDraw} is called in parallel by
+Equalizer on all pipe threads in the configuration. Therefore the
+rendering should not write to the database, which is the case for all
+major scene graph implementations.
 
@@
 \item[Frustum]: Frustum parameters as defined by
-  \textsf{glFrustum}. The frustum is influenced by the destination
+  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL
+  projection matrix. The frustum is influenced by the destination
   channel's view definition, compound viewports, head matrix and the
@@
   frustum. This is typically an identity matrix and is used for off-axis
-  frustra in immersive rendering.
+  frustra in immersive rendering. It is normally used to set up the
+  `view' part of the modelview matrix, before static light sources are
+  defined.
 \item[Range]: A one-dimensional range with the interval [0..1]. This
@@
   \includegraphics[width=.4\textwidth]{images/DB.pdf}
-  {\caption{\small\label{fDB}Final view of an DB compound}}
+  {\caption{\small\label{fDB}Destination view of an DB compound}}
 \end{wrapfigure}
 
-After the basic view setup, a directional light is set up, and the model
-is positioned using the camera parameters from the frame data. The
+After the basic view setup, a directional light is configured, and the
+model is positioned using the camera parameters from the frame data. The
 camera parameters are transported using the the frame data to ensure
-that all channels render the frame from the same position.
+that all channels render the frame in the same position.
 
@@
 Finally the model is rendered, if it could be loaded during
-initialization. Otherwise a quad is drawn in its place:
+node initialization. Otherwise a quad is drawn in its place:
 
@@
 
-\if 0
+In order to draw the model, a helper class for view frustum culling is
+set up using the view frustum from Equalizer and the camera
+position. The frustum helper computes the six frustum planes from the
+projection and modelView matrices. During rendering, the bounding
+spheres of the model are tested against these planes to determine the
+visibility with the frustum. Furthermore the render state from the
+window, and the database range from the channel is obtained. The render
+state manages display list or VBO allocation:
+
 {\footnotesize\begin{lstlisting}
@@
 
-{\footnotesize\begin{lstlisting}
-\end{lstlisting}}
-
-The model data is organized a in 3-dimensional kD-tree\footnote{See also
-  \link{http://en.wikipedia.org/wiki/Kd-tree}} for efficient view
-frustum culling. During preprocessing, each node of the tree gets a
-database range assigned.
-
-{\footnotesize\begin{lstlisting}
-\end{lstlisting}}
-\fi
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Rendering Loop}}
+\end{wrapfigure}
+The model data is spatially organized in an 3-dimensional
+kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
+for efficient view frustum culling. When the model is loaded by
+\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
+node of the tree gets a database range assigned, that is, the root node
+has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
+1], and so on for all nodes in the tree. The preprocessed model is saved
+in a binary format for accelerating subsequent use.
+
+The rendering loop maintains a list of candidates to render, which
+initially contains the root node. Each candidate of this list is tested
+for full visiblity against the frustum and range, and rendered if
+visible. It is dropped if it is fully invisible or fully out of
+range. If it is partially visible or partially in range, the children of
+the node are added to the candidate list:
+
+{\footnotesize\begin{lstlisting}
+    model->beginRendering( state );
+
+    // start with root node
+    vector< const VertexBufferBase* > candidates;
+    candidates.push_back( model );
+
+    while( !candidates.empty() )
+    {
+        const VertexBufferBase* treeNode = candidates.back();
+        candidates.pop_back();
+
+        // completely out of range check
+        if( treeNode->getRange()[0] >= range.end || 
+            treeNode->getRange()[1] < range.start )
+            continue;
+
+        // bounding sphere view frustum culling
+        switch( culler.testSphere( treeNode->getBoundingSphere() ) )
+        {
+            case vmml::VISIBILITY_FULL:
+                // if fully visible and fully in range, render it
+                if( treeNode->getRange()[0] >= range.start && 
+                    treeNode->getRange()[1] < range.end )
+                {
+                    treeNode->render( state );
+                    break;
+                }
+                // partial range, fall through to partial visibility
+            case vmml::VISIBILITY_PARTIAL:
+            {
+                const VertexBufferBase* left  = treeNode->getLeft();
+                const VertexBufferBase* right = treeNode->getRight();
+
+                if( !left && !right )
+                {
+                    if( treeNode->getRange()[0] >= range.start )
+                        treeNode->render( state );
+                    // else drop, to be drawn by 'previous' channel
+                }
+                else
+                {
+                    if( left )
+                        candidates.push_back( left );
+                    if( right )
+                        candidates.push_back( right );
+                }
+                break;
+            }
+            case vmml::VISIBILITY_NONE:
+                // do nothing
+                break;
+        }
+    }
+
+    model->endRendering( state );
+}
+\end{lstlisting}}
 
@@
 
-This section discusses some features of interest which are not covered
-by the \textsf{eqPly} section. Where possible, code examples from the
-Equalizer distribution are used.
+This section discusses some additional, important features not covered
+by the previous \textsf{eqPly} section. Where possible, code examples
+from the Equalizer distribution are used.
 
@@
 
-Event handling is one of the areas which requires a lot of
-flexibility. On one hand, the implementation differs slightly for each
-window system due to conceptual differences in the OS-specific
-implementation. On the other hand, each application and widget set has
-its own model on how events are to be handled. Therefore, event handling
-in Equalizer is customizable in any stage of the processing\footnote{\link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.
-
-The default implementation aims to provide a quick and easy
-implementation of event handling, while allowing all necessary
-customizations. It gathers all events in the application main thread, so
-that the developer only has to implement \textsf{Config::processEvent}
-to update its data based on the pre-processed, generic events.
+Event handling requires a lot of flexibility. On one hand, the
+implementation differs slightly for each operating and window system due
+to conceptual differences in the OS-specific implementation. On the
+other hand, each application and widget set has its own model on how
+events are to be handled. Therefore, event handling in Equalizer is
+customizable in any stage of the processing, to the extreme of making it
+possible to disable all event-related code in Equalizer. More
+information on event handling can be found on the Equalizer
+website\footnote{see
+  \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.
+
+The default implementation provides a convenient, easily accessible
+event framework, while allowing all necessary customizations. It gathers
+all events in the application main thread, so that the developer only
+has to implement \textsf{Config::processEvent} to update its data based
+on the pre-processed, generic keyboard and mouse events. It is very easy
+to use and similar to an GLUT-based implementation.
 
@@
 Where possible, events are received and processed by a separate per-node
-event thread to allow asynchronous event handling. Currently an event
-thread is only used by the X11/glX window system. WGL receives and
-processes the events from the pipe threads which created the
-windows. AGL receives the events from application or node main
-thread. Whenever the term \textbf{event thread} is used, it refers to
-the thread receiving the event, i.e., a per-node thread for glX, the pipe
-thread for WGL and the main thread for AGL.
+event thread to allow asynchronous\footnote{with respect to the
+  rendering} event handling. Currently an event thread is only used by
+the X11/glX window system. WGL receives and processes the events from
+the pipe threads which created the windows. AGL receives the events from
+application or node main thread. Whenever the term \textbf{event thread}
+is used, it refers to the thread receiving the event, i.e., a per-node
+thread for glX, the pipe thread for WGL and the main thread for AGL.
 
 \subsubsection{Initialization and Exit}
+
+During window and pipe initialization the event handling is set up. For
+both entities, \textsf{initEvent\-Handler} is called to register the
+pipe or window with an event handler. This method may be overwritten to
+use a custom event handler, or to not install an event handler to
+disable event handling. Likewise, \textsf{exitEventHandler} is called to
+de-initialize event handling.
 
@@
 \end{wrapfigure}
-
-During window and pipe initialization the event handling is set up. For
-both entities, \textsf{initEvent\-Handler} is called to register the pipe
-or window with an event handler. This method may be overwritten to use a
-custom event handler, or to not install an event handler to disable
-event handling in Equalizer. Likewise, \textsf{exitEventHandler} is
-called to de-initialize event handling.
-
 An event handler consists of two parts: the generic base class providing
-the interface and some generic functions, and the window-system-specific
-part providiing the actual implementation.
+the interface and generic functions, and the window-system-specific
+part providing the actual implementation.
 
@@
 
-WGL and AGL require the application to manually receive and dispatch
-(`pump') events. On WGL, this has to happen on each thread with windows,
-whereas on AGL it has to happen only on the main thread. By default,
-Equalizer pumps these events automatically for the application.
+For the WGL and AGL window system it is required to manually receive and
+dispatch (`pump') events. On WGL, this has to happen on each thread with
+windows, whereas on AGL it has to happen only on the main thread. By
+default, Equalizer pumps these events automatically for the application.
 
 The methods \textsf{Client::useMessagePump} and
-\textsf{Pipe::useMessagePump} can be overriden to return \textsf{false}
+\textsf{Pipe::useMessagePump} can be overridden to return \textsf{false}
 to disable this behaviour for their respective threads. On non-threaded
@@
 
+If the application disables message pumping in Equalizer, it has to make
+sure the events are pumped externally, as it often done within external
+widget sets, e.g., Trolltech's Qt.
+
 \subsubsection{Event Data Flow}
@@
 \begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/eventHandling.pdf}
+  \includegraphics[width=.4\textwidth]{images/eventFilter.pdf}
   {\caption{\small\label{fEventProcessing}Event Processing}}
 \end{wrapfigure}
-
 Events are received by an event handler. The event handler finds the
@@
 application thread. After a frame has been finished,
-\textsf{Config::finishFrame} calls \textsf{Config::processEvents}. The
+\textsf{Config::finishFrame} calls \textsf{Config::handleEvents}. The
 default implementation of this method provides non-blocking event
-processing, that is, it calls \textsf{Config::processEvent} for each
+processing, that is, it calls \textsf{Config::handleEvent} for each
 queued event. By overriding this function, event-driven execution can be
 implemented.
+
+Later Equalizer versions will introduce \textsf{Pipe::processEvent} and
+\textsf{PipeEvent} to communicate pipe-specific events, e.g, resolution
+changes.
 
@@
 
-\subsection{Image Compositing for Scalable Rendering}
+\subsection{\label{sCompositing}Image Compositing for Scalable Rendering}
 
@@
 
+\subsubsection{Parallel Direct Send Compositing}
+
+In order to provide a motivation for the design of the image compositing
+API, the parallel compositing algorithm for direct send is introduced
+here.
+
+The motivation for direct send is to parallelize the costly reassembly
+for database (sort-last) decomposition. With each additional source
+channel, the amount of pixel data to be composited grows
+linearly. Direct send distributes this workload evenly across all source
+channels, and thereby keeps the compositing work per channel constant.
+
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/directSend.pdf}
+  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+\end{wrapfigure}
+Each channel is responsible for the sort-last compositing of one 2D
+tile. He receives the framebuffer pixels for his tike from all the other
+channels. The size of this tile decreases linearly with the number of
+source channels.
+
+After performing the sort-last compositing, the color information is
+transferred to the destination channel, similarly to an 2D (sort-first)
+compound.
+
+\fig{fDirectSend} illustrates this algorithm for three channels. The
+Equalizer website contains a presentation for this
+algorithm\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}.
+
+The following operations have to be possible in order to perform this
+algorithm:
+\begin{itemize}
+\item Selection of color and/or depth readback
+\item Restricting the read back to a part of the rendered area
+\item Positioning the pixel data correctly on the receiving channels
+\end{itemize}
+
+Furthermore it should be possible for the application to implement a
+read back of only the relevant region of interest, that is, the 2D area
+of the framebuffer actually updated during rendering. This optimization
+will be fully supported by later versions of Equalizer.
+
 \subsubsection{Frame, FrameData and Images}
 
-A \textsf{eq::Frame} references a \textsf{eq::Fra\-me\-Data}. The frame data
-is the object connecting output with input frames. It is a holder for
-images and is used to signal the availability of the pixel data. An
-\textsf{eq::Image} holds a two-dimensional snapshot of the framebuffer
-and can contain color and/or depth information.
+An \textsf{eq::Frame} references an \textsf{eq::Fra\-me\-Data}. The
+frame data is the object connecting output with input frames. Output and
+input frames with the same name and from the same compound tree will
+reference the same frame data.
+
+The frame data is a holder for images and is used to signal the
+availability of the pixel data. An \textsf{eq::Image} holds a
+two-dimensional snapshot of the framebuffer and can contain color and/or
+depth information.
+
+The frame data transports the inherited range of its compound. The range
+is needed to compute the assembly order of multiple input frames, e.g.,
+for sorted-blend compositing in volume rendering applications.
 
 Readback and assemble operations on frames and images are designed to be
-asynchronous. They have a start and finish method for readback and
+asynchronous. They have a start and finish method for both readback and
 assemble to allow the initiation and synchronization of the operation.
@@
 
-\if 0
+The offset of input and output frames characterizes the position of the
+frame data with respect to the framebuffer, that is, the
+\textbf{window's} lower-left corner. For output frames this is the
+position of the channel with respect to the window.
 
@@
 \end{wrapfigure}
-
-The offset of input and output frames characterizes the position of the
-frame data with respect to the framebuffer, that is, the
-\textbf{window's} lower-left corner. For output frames this is the
-position of the channel in 
+For output frames, the frame data's pixel viewport (pvp) is the area of
+the frame buffer to read back. It will transport the offset from the
+source to the destination channel, that is, the frame data pvp for input
+frames position the pixel data on the destination. This has the effect
+that a partial framebuffer readback will end up in the same place in the
+destination channels.
+
+The image pixel viewport signifies the region of interest read
+back. Currently this pvp has no offset and is as big as the frame data
+pvp. Only one image will be read back.
+
+\fig{fAssembly} illustrates the relationship between frames, frame data
+and images.
+
+\subsubsection{Custom Assembly in eVolve}
+
+The \textsf{eVolve} example is a scalable volume renderer.
+
+\subsection{Head Tracking}
+
+The \textsf{eqPly} example contains rudimentary support for head
+tracking, in order to show how head tracking can be integrated with
+Equalizer. Supporting a wide range of tracking devices is not within the
+scope of Equalizer. Other open source and commercial implementations
+cover this functionality sufficiently and can be integrated with
+Equalizer.
+
+\begin{figure}[h!t]
+  \subfigure[]{\label{fMono}
+    \includegraphics[width=.3\textwidth]{images/mono}}\hfil
+  \subfigure[]{\label{fStereo}
+    \includegraphics[width=.3\textwidth]{images/stereo}}\hfil
+  \subfigure[]{\label{fTracked}
+    \includegraphics[width=.3\textwidth]{images/tracked}}%
+  {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked
+    frustra}}
+\end{figure}
+
+\fig{fMono} illustrates a monoscopic view frustum. The viewer is
+positioned at the origin, and the frustum is completely symmetric. This
+is the typical view frustum for non-stereoscopic applications.
+
+In stereo rendering, the scene is rendered twice, with the two frustra
+'moved' by the distance between the eyes, as shown in \fig{fStereo}.
+
+In immersive visualization, the observer is tracked in and the view
+frustra are adapted to the viewer's position and orientation, as shown
+in \fig{fTracked}. The transformation origin $\rightarrow$ viewer is set by
+the application using \textsf{Config::setHeadMatrix}, which is used by
+the server to compute the frustra. The resulting off-axis frustra are
+positioned using the channel's head transformation.
+
+\if 0
+{\footnotesize\begin{lstlisting}
+\end{lstlisting}}
 \fi
 
-\subsubsection{Custom Assembly in eVolve}
-
-\subsection{Head Tracking}
-\if 0
-{\footnotesize\begin{lstlisting}
-\end{lstlisting}}
-\fi
-
 \end{document}

### 2007-09-28 Developer/eqPly/chapters/application.tex
Updated /doc/Developer/eqPly to final version.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 \textbf{Equalizer Klasse} & \textbf{Bedeutung} \\
-eq::Config & Beschreibung der vorhandenen Resourcen und deren Verwendung \\
+eq::Config & Beschreibung vorhandener Resourcen und deren Verwendung \\
 eq::Node & Einzelner Computer innerhalb des Rendering Clusters \\
@@
 die Architektur der neuen Applikation kaum von der des alten Beispielprogrammes
-ab. Abbildung \ref{fig:eq-classes} am Ende der Sektion zeigt die Zusammenhänge 
-der Klassen sowohl untereinander als auch in Relation zu Equalizer. Da der 
-Aufbau von \emph{eqPly} in \cite{equalizer:guide} schon ausführlich behandelt 
-wird, beschränken sich die folgenden Ausführungen auf ein Minimum. Die Klassen 
-betreffend Datenstruktur und Rendering (\emph{mesh::Vertex*}) werden gesondert 
-in Kapitel \ref{data-structure} und \ref{rendering} behandelt.
+ab. Abbildung \ref{fig:eq-classes} zeigt die Zusammenhänge der Klassen sowohl 
+untereinander als auch in Relation zu Equalizer. Da der Aufbau von \emph{eqPly}
+in \cite{equalizer:guide} schon ausführlich behandelt wird, beschränken sich 
+die folgenden Ausführungen auf ein Minimum. Die Klassen betreffend 
+Datenstruktur und Rendering (\emph{mesh::Vertex*}) werden gesondert in Kapitel
+\ref{data-structure} und \ref{rendering} behandelt.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{figures/UML.pdf}
+\caption[Zusammenhänge der Klassen von eqPly und Equalizer]
+{Zusammenhänge der Klassen von eqPly und Equalizer (Quelle: \cite{equalizer:guide})}
+\label{fig:eq-classes}
+\end{figure}
 
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.00\textwidth]{figures/UML.pdf}
-\caption[Zusammenhänge der Klassen von eqPly und Equalizer]
-{Zusammenhänge der Klassen von eqPly und Equalizer (Quelle: \cite{equalizer:guide})}
-\label{fig:eq-classes}
-\end{figure}
-
 %

### 2007-09-28 Developer/eqPly/chapters/conclusion.tex
Updated /doc/Developer/eqPly to final version.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 und für weitere Aufrufe verfügbar sind, ist dieser Punkt vernachlässigbar. Der
-Speicherverbrauch wurde auf weniger als ein Viertel reduziert (siehe Tabelle
-\ref{tbl:file-sizes}).
+Speicherverbrauch wurde auf weniger als ein Viertel reduziert, wie die folgende
+Tabelle \ref{tbl:file-sizes} anhand der Grösse binärer 32 Bit Repräsentationen
+aufzeigt. Es findet keine Kompression statt. Die für die Messungen verwendeten 
+Modelle stammen aus \cite{stanford:repository}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrr}
+\textbf{Modell} & \textbf{Binary Octree} & 
+\textbf{Binary kd-Tree} & \textbf{Verhältnis}\\
+\textbf{} & \textbf{[MB]} & 
+\textbf{[MB]} & \textbf{[\%]}\\
+Armadillo      &  27.72 &   6.09 & 21.96\\
+Dragon         &  69.83 &  15.34 & 21.97\\
+Happy Buddha   &  87.16 &  19.05 & 21.86\\
+Lucy           &    n/a & 494.58 &   n/a\\
+Stanford Bunny &   5.56 &   1.20 & 21.56\\
+\end{tabular}
+\caption{Unterschied Speicherbedarf der Datenstrukturen}
+\label{tbl:file-sizes}
+\end{table}
 
@@
 hinter den Erwartungen zurück, während sie in anderen Fällen einen markanten
-Leistungsvorteil zeigt. Die verwendeten Modelle stammen aus 
-\cite{stanford:repository} und \cite{cyberware}.
+Leistungsvorteil zeigt. Neben den Modellen aus \cite{stanford:repository} 
+wurden hierfür auch Modelle von \cite{cyberware} verwendet.
 
@@
 Armadillo        &     345'944  &  118.30   &   136.43   &    92.17\\
-Ganesh           &     413'236  &           &            &         \\
 Boat             &     776'374  &   29.25   &    20.06   &    28.47\\
@@
 Happy Buddha     &   1'087'716  &    9.54   &    29.27   &    34.88\\
-Asian Dragon     &   7'219'045  &           &            &         \\
-Thai Statue      &  10'000'000  &           &            &         \\
 \end{tabular}
@@
 \label{tbl:render-performance}
+\end{table}
+
+Um den möglichen Performance Unterschieden des VBO Renderings auf die Spur zu
+kommen, wurde noch die erzeugte Anzahl Blattknoten sowie die durchschnittliche
+Anzahl Vertices und Triangles pro Blatt ermittelt. Die Ergebnisse in Tabelle
+\ref{tbl:tree-data} erlauben aber auch keine gesicherten Rückschlüsse.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrr}
+\textbf{Modell} & \textbf{Anzahl} & \textbf{Anzahl} & 
+\textbf{Triangles} & \textbf{Vertices}\\
+\textbf{} & \textbf{Triangles} & \textbf{Blätter} & 
+\textbf{pro Blatt} & \textbf{pro Blatt} \\
+Stanford Bunny   &      69'451  &   16   &    4'340   &    2'329\\
+Rocker Arm       &      80'354  &   16   &    5'022   &    2'797\\
+Goddess 1        &     274'822  &   16   &   17'176   &    9'252\\
+Armadillo        &     345'944  &   16   &   21'621   &   11'218\\
+Boat             &     776'374  &   32   &   24'261   &   12'769\\
+Dragon           &     871'414  &   32   &   27'231   &   14'138\\
+Goddess 2        &   1'047'330  &   32   &   32'729   &   17'430\\
+Hip              &   1'060'346  &   32   &   33'135   &   17'307\\
+Happy Buddha     &   1'087'716  &   32   &   33'991   &   17'509\\
+\end{tabular}
+\caption{Statistische Daten über die Ausnutzung der Blätter im kd-Tree}
+\label{tbl:tree-data}
 \end{table}
@@
 
-\begin{figure}[hb]
+\begin{figure}[ht]
 \centering
-\includegraphics[width=0.86\textwidth]{figures/FlatShading.png}
+\includegraphics[width=0.60\textwidth]{figures/FlatShading.png}
 \caption{RockerArm Modell mit Flat Shading}
 \label{fig:sample-flat}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.86\textwidth]{figures/SmoothShading.png}
-\caption{RockerArm Modell mit Gouraud Shading}
-\label{fig:sample-smooth}
 \end{figure}
@@
 
+\begin{figure}[hb]
+\centering
+\includegraphics[width=0.60\textwidth]{figures/SmoothShading.png}
+\caption{RockerArm Modell mit Gouraud Shading}
+\label{fig:sample-smooth}
+\end{figure}
+
 %

### 2007-09-28 Developer/eqPly/chapters/datastructure.tex
Updated /doc/Developer/eqPly to final version.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     bool readPlyFile( const char* file, const bool ignoreColors );
-    void sort( const Index start, const Index length, const Axis axis );
+    void sort( const Index start, const Index length, 
+               const Axis axis );
     void scale( const float baseSize );
@@
     ...
-
     BoundingBox _boundingBox;
@@
     virtual void toStream( std::ostream& os );
-    virtual void fromMemory( char** addr, VertexBufferData& globalData );
+    virtual void fromMemory( char** addr, 
+                             VertexBufferData& globalData );
 
@@
                             const Index length, const Axis axis,
+                            const size_t depth,
                             VertexBufferData& globalData );
@@
 {
-public:
-    ...
-
-protected:
     ...
@@
                                   const Index length, const Axis axis,
+                                  const size_t depth,
                                   VertexBufferData& globalData )
 {
-    ...
-
     data.sort( start, length, axis );
-    Index median = start + ( length / 2 );
+    const Index median = start + ( length / 2 );
 
     // left child will include elements smaller than the median
-    Index leftLength = length / 2;
-    if( countUniqueVertices( data, start, leftLength ) > LEAF_SIZE )
+    const Index leftLength    = length / 2;
+    const bool  subdivideLeft = 
+        countUniqueVertices( data, start, leftLength ) > LEAF_SIZE ||
+        depth < 3;
+
+    if( subdivideLeft )
         _left = new VertexBufferNode;
@@
 
-    // right child will include elements equal to or greater
-    Index rightLength = ( length + 1 ) / 2;
-    if( countUniqueVertices( data, median, rightLength ) > LEAF_SIZE )
+    // right child will include elements equal or greater median
+    const Index rightLength    = ( length + 1 ) / 2;
+    const bool  subdivideRight = 
+        countUniqueVertices( data, median, rightLength ) > LEAF_SIZE ||
+        depth < 3;
+
+    if( subdivideRight )
         _right = new VertexBufferNode;
@@
 
-    // move to next axis and continue construction in the child nodes
-    Axis newAxis = static_cast< Axis >( ( axis + 1 ) % 3 );
+    // move to next axis and continue contruction in the child nodes
+    const Axis newAxis = static_cast< Axis >( ( axis + 1 ) % 3 );
     static_cast< VertexBufferNode* >
         ( _left )->setupTree( data, start, leftLength, newAxis, 
-                              globalData );
+                              depth + 1, globalData );
     static_cast< VertexBufferNode* >
         ( _right )->setupTree( data, median, rightLength, newAxis, 
-                               globalData );
+                               depth + 1, globalData );
 }
@@
                                   const Index length, const Axis axis,
+                                  const size_t depth,
                                   VertexBufferData& globalData )
 {
-    ...
-
     data.sort( start, length, axis );
@@
     }
-
-    ...
 }
@@
     VertexBufferNode::setupTree( data, 0, data.triangles.size(), 
-                                 AXIS_X, _data );
+                                 AXIS_X, 0, _data );
     VertexBufferNode::updateBoundingSphere();
@@
 
-Tabelle \ref{tbl:file-sizes} zeigt anhand der Dateigrössen der binären 32 Bit 
-Repräsentationen (es findet keine Komprimierung statt) noch auf, wie sehr der 
-Speicherbedarf durch den Wechsel von der alten auf die neue Datenstruktur 
-reduziert wurde. Die für die Messungen verwendeten PLY Beispielmodelle 
-stammen aus \cite{stanford:repository}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrrr}
-\textbf{Modell} & \textbf{Binary Octree} & 
-\textbf{Binary kd-Tree} & \textbf{Verhältnis}\\
-\textbf{} & \textbf{[MB]} & 
-\textbf{[MB]} & \textbf{[\%]}\\
-Armadillo      &  27.72 &   6.09 & 21.96\\
-Dragon         &  69.83 &  15.34 & 21.97\\
-Happy Buddha   &  87.16 &  19.05 & 21.86\\
-Lucy           &    n/a & 494.58 &   n/a\\
-Stanford Bunny &   5.56 &   1.20 & 21.56\\
-\end{tabular}
-\caption{Unterschied Speicherbedarf der Datenstrukturen}
-\label{tbl:file-sizes}
-\end{table}
-
 %

### 2007-10-07 Developer/ProgrammingGuide/paper.tex
Programming Guide: update to v0.9
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 \title{Equalizer Programming Guide\\
-  {\footnotesize\mdseries
+  {\footnotesize\sffamily\mdseries
     \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
@@
 }
-\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
-  % Eyescale Software Sarl
+\author{Stefan Eilemann\\{\footnotesize Eyescale Software GmbH}
 }
 \date{
-  \textbf{INCOMPLETE}\\[\medskipamount]
-  Version 0.8, \today
+  \vfill
+  \vfill
+  \begin{figure}[ht]
+    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
+  \end{figure}
+  \vfill
+  \vfill
+  \textbf{DRAFT}\\[\medskipamount]
+  Version 0.9 for Equalizer v0.4\\[\medskipamount]
+  \today
 }
@@
 \thispagestyle{empty}
-\begin{figure}[ht]
-  \center
-  \includegraphics[width=4cm]{images/logo.pdf}
-%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
-\end{figure}
-\vfill
-\begin{figure}[ht]
-  \center
-  \includegraphics[width=4cm]{images/vmml.pdf}
-%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
-\end{figure}
-\vfill
 
@@
     \hline
-    0.8     & Sep 28, 2007 & added head tracking, finished channel,
+    0.9     & Oct 7, 2007  & add eqHello, draft assembly in eVolve\\
+    0.8     & Sep 28, 2007 & add head tracking, finish channel,
                              proof-reading pass\\
-    0.7     & Sep 21, 2007 & started channel and image compositing,
-                             added event handling, moved to new eqPly
+    0.7     & Sep 21, 2007 & start channel and image compositing,
+                             add event handling, move to new eqPly
                              code base\\
-    0.6     & Sep 14, 2007 & added pipe, window and object manager\\
-    0.5     & Sep 3,  2007 & added config and partly pipe\\
-    0.4     & Aug 31, 2007 & added distributed objects\\
-    0.3     & Aug 26, 2007 & added application and render client\\
-    0.2     & Aug 20, 2007 & added main function\\
-    0.1     & Aug 19, 2007 & outlined the basic concepts\\
+    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
+    0.5     & Sep 3,  2007 & add config and partly pipe\\
+    0.4     & Aug 31, 2007 & add distributed objects\\
+    0.3     & Aug 26, 2007 & add application and render client\\
+    0.2     & Aug 20, 2007 & add main function\\
+    0.1     & Aug 19, 2007 & outline the basic concepts\\
     \hline
@@
 
-A prerequisite for this Programming Guide is a working \textsf{eqPly}
-example. The Quickstart
+It is recommended to have a compiled version of Equalizer and the
+examples for this Programming Guide. The Quickstart
 Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
-explains how to run it. \textsf{eqPly} can also be executed without a
-server, which simplifies the development cycle. In this case it will be
-configured to use one window.
+should be executed once to get familar with the capabilities of the
+\textsf{eqPly} example.
+
+Compiling Equalizer is as simple as running \textsf{make} on Linux or
+building the Equalizer solution on Windows. On Mac OS X, some
+prerequisites have to be installed before running \textsf{make}, as
+explained in \textsf{README.Darwin}.
 
@@
 
+\section{Hello, World!}
+
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/eqHello.png}
+  {\caption{\small\label{fHello}Hello, World!}}
+\end{wrapfigure}
+
+The \textsf{eqHello} example is a minimal example to illustrate the
+basic principle of an Equalizer application: The application developer
+has to implement the rendering method \textsf{Channel::frameDraw},
+similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
+run as a stand-alone application from the command line.
+
+The \textsf{eqHello} redraw function renders six rotating, colored quads
+around the origin, where the viewer is usually positioned. The Equalizer
+\textsf{frameDraw} meth\-od is used as a convience function to setup the
+frustum and other OpenGL state. After setting up some additional
+lighting parameter, \textsf{eqHello} rotates the scene and render the
+quads using immediate mode:
+
+{\footnotesize\begin{lstlisting}
+void Channel::frameDraw( const uint32_t spin )
+{
+    // setup OpenGL State
+    eq::Channel::frameDraw( spin );
+
+    const float lightPos[] = { 0.0f, 0.0f, 1.0f, 0.0f };
+    glLightfv( GL_LIGHT0, GL_POSITION, lightPos );
+
+    const float lightAmbient[] = { 0.2f, 0.2f, 0.2f, 1.0f };
+    glLightfv( GL_LIGHT0, GL_AMBIENT, lightAmbient );
+
+    // rotate scene around the origin
+    glRotatef( static_cast< float >( spin ) * 0.5f, 1.0f, 0.5f, 0.25f );
+
+    // render six axis-aligned colored quads around the origin
+    [...]
+}
+\end{lstlisting}}
+
+The \textsf{eqHello} main function sets up the communication with the
+server, initializes and drives the rendering. The details of this setup
+are explained in \sref{sEqPly}.
+
 \section{The Programming Interface}
@@
 
-\subsection{The Resource Tree}
+\subsection{\label{sResource}The Resource Tree}
 
@@
 
-\section{The \textsf{eqPly} polygonal renderer}
+\section{\label{sEqPly}The eqPly polygonal renderer}
 
@@
 and serves as a reference implementation of an Equalizer-based
-application. Its focus is not on rendering features or visual quality,
-but it is used as a test bed for most of the Equalizer features.
+application of medium complexity. Its focus is not on rendering features
+or visual quality, but it is used as a show case for the integration of
+most of the Equalizer features.
 
@@
 All classes in the example are in the \textsf{eqPly} namespace to avoid
-type name ambiguities, in particular for the \textsf{Window}
-class. \fig{fUml} shows the most important Equalizer classes and how
-they are subclassed in the \textsf{eqPly} example. Each of the classes
-in the \textsf{eqPly} namespace will be discussed in this section.
+type name ambiguities, in particular for the \textsf{Window} class which
+is frequently used as a type by windowing systems. \fig{fUml} shows the
+most important Equalizer classes and how they are subclassed in the
+\textsf{eqPly} example.
+
+The derived classes fall into two categories: The subclassed rendering
+entities as discussed in \sref{sResource}, and two classes for
+distributing data. For each of the rendering entities, its function and
+typical usage will be discussed. The distributed data classes illustrate
+the typical usage of static and dynamic, frame-specific data.
 
@@
 and deserializing its data. The serialization is done in
-\textsf{getInstanceData}. The serialized data is cached using a
-(non-distributed) member variable, which is cleared by each setter of
-distributed data. Serializing the data is a simple matter of allocating
-a large enough memory buffer, and copying the data into the buffer:
-
-{\footnotesize\begin{lstlisting}
-const void* InitData::getInstanceData( uint64_t* size )
-{
-    *size = sizeof( StaticData ) + _filename.length() + 1;
-    if( _instanceData )
-        return _instanceData;
-
-    _instanceData = new char[ *size ];
-
-    reinterpret_cast< StaticData* >( _instanceData )[0] = _data;
-
-    const char* string = _filename.c_str();
-    memcpy( _instanceData + sizeof( StaticData ), string, _filename.length()+1);
-
-    return _instanceData;
-}
-\end{lstlisting}}
-
-After the data is no longer needed, \textsf{releaseInstanceData} is
-called by Equalizer to allow the object to free the data. Since the data
-is cached for further usage, the release method is not overwritten in
-\textsf{InitData}.
-
-The memory buffer returned by \textsf{getInstanceData} is transferred to
-the remote node during object mapping and passed to
-\textsf{applyInstanceData} for deserialization. The deserialization
-routine reads the data back into its own member variables:
-
-{\footnotesize\begin{lstlisting}
-void InitData::applyInstanceData( const void* data, const uint64_t size )
-{
-    _data     = reinterpret_cast< const StaticData* >( data )[0];
-    _filename = static_cast<const char*>( data ) + sizeof( StaticData );
-
-    EQASSERT( _data.frameDataID != EQ_ID_INVALID );
+\textsf{getInstanceData} and deserialization in
+\textsf{applyInstanceData} by streaming all member variable to or from
+the provided data streams. The data transport between nodes is
+implemented in the data streams and various other Equalizer classes:
+
+{\footnotesize\begin{lstlisting}
+void InitData::getInstanceData( eqNet::DataOStream& os )
+{
+    os << _frameDataID << _windowSystem << _useVBOs << _useShaders << _filename;
+}
+
+void InitData::applyInstanceData( eqNet::DataIStream& is )
+{
+    is >> _frameDataID >> _windowSystem >> _useVBOs >> _useShaders >> _filename;
+
+    EQASSERT( _frameDataID != EQ_ID_INVALID );
     EQINFO << "New InitData instance" << endl;
@@
 \end{lstlisting}}%>>
+
+The data output and input streams perform no type checking on the
+variables. It is the application's responsibility to match the order and
+types of variables during serialization and deserialization exactly.
 
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/configInit.pdf}
-  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
-\end{wrapfigure}
 The config initialization happens in parallel, that is, all config
@@
 
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/configInit.pdf}
+  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
+\end{wrapfigure}
 The \textsf{eqPly::Config} class holds the master versions of the
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
-  {\caption{\small\label{fWindowInit}Window Initialization}}
-\end{wrapfigure}
-
 The initialization sequence uses multiple, overrideable task
@@
 
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
+  {\caption{\small\label{fWindowInit}Window Initialization}}
+\end{wrapfigure}
 \fig{fWindowInit} shows a flow chart of the window initialization. The
@@
 class. Each of the \textsf{apply} methods uses the corresponding
-\textsf{get} method(s) and then calls the approriate OpenGL
+\textsf{get} method(s) and then calls the appropriate OpenGL
 function(s). It is also possible to just query the values from Equalizer
@@
 
-The following contextual information has to be used in order to render
-the view expected by Equalizer:
+The contextual information has to be used in order to render the view as
+expected by Equalizer. Failure to use certain information will result in
+incorrect rendering for some or all configurations. The channel render
+context consist of:
+
 \begin{description}
@@
   current eye pass.
-\item[Head Transformation:] A transformation matrix positioning the
+\item[Head Transformation]: A transformation matrix positioning the
   frustum. This is typically an identity matrix and is used for off-axis
@@
 
-Consequently, the first part of the \textsf{frameDraw} method in
-\textsf{eqPly} look as follows:
-
-{\footnotesize\begin{lstlisting}
-void Channel::frameDraw( const uint32_t frameID )
+The \textsf{frameDraw} method in \textsf{eqPly} calls the
+\textsf{frameDraw} method from the parent class, the Equalizer
+channel. The default \textsf{frameDraw} method uses the apply
+convenience functions to setup the OpenGL state for all render context
+information, with the exception of the range which will be used later
+during rendering:
+
+{\footnotesize\begin{lstlisting}
+void eq::Channel::frameDraw( const uint32_t frameID )
 {
@@
     glLoadIdentity();
-
     applyFrustum();
@@
     applyHeadTransform();
+}
+
+void Channel::frameDraw( const uint32_t frameID )
+{
+    // Setup OpenGL state
+    eq::Channel::frameDraw( frameID );
 \end{lstlisting}}
@@
 \end{wrapfigure}
-
 After the basic view setup, a directional light is configured, and the
@@
 camera parameters are transported using the the frame data to ensure
-that all channels render the frame in the same position.
-
-Furthermore, a color is set in case the model does not contain color
-information, or the color information is not to be used. In sort-last
-rendering, \textsf{eqPly} renders the individual parts with different
-colors to illustrate the algorithm, as shown in \fig{fDB}. The Equalizer
-channel provides a method to obtain a random, but unique color for all
-channels in the configuration:
-
-{\footnotesize\begin{lstlisting}
-    glLightfv( GL_LIGHT0, GL_POSITION, lightpos );
+that all channels render a given frame using the same position.
+
+Furthermore, a white color is set in case the model does not contain
+color information, or the color information is not to be used. In
+sort-last rendering, \textsf{eqPly} uses a different color for each
+channell to illustrate the database decomposition, as shown in
+\fig{fDB}. The Equalizer channel provides a method to obtain a random,
+but unique color for all channels in the configuration. This color is
+determined by the server to ensure uniqueness across all channels of the
+configuration:
+
+{\footnotesize\begin{lstlisting}
+    glLightfv( GL_LIGHT0, GL_POSITION, lightPosition );
+    glLightfv( GL_LIGHT0, GL_AMBIENT,  lightAmbient  );
+    glLightfv( GL_LIGHT0, GL_DIFFUSE,  lightDiffuse  );
+    glLightfv( GL_LIGHT0, GL_SPECULAR, lightSpecular );
+
+    glMaterialfv( GL_FRONT, GL_AMBIENT,   materialAmbient );
+    glMaterialfv( GL_FRONT, GL_DIFFUSE,   materialDiffuse );
+    glMaterialfv( GL_FRONT, GL_SPECULAR,  materialSpecular );
+    glMateriali(  GL_FRONT, GL_SHININESS, materialShininess );
 
@@
 
-Finally the model is rendered, if it could be loaded during
-node initialization. Otherwise a quad is drawn in its place:
+Finally the model loaded by the node is rendered. If the model was not
+loaded during node initialization, a quad is drawn in its place:
 
@@
 result image during scalable rendering, when multiple channels are
-rendering for a single view. The source channels producing an
-\textsf{outputFrame} use \textsf{Channel::frameReadback} to read the
+rendering for a single view. The source channels producing one or more
+\textsf{outputFrame}s use \textsf{Channel::frameReadback} to read the
 pixel data from the frame buffer. The channels receiving one or multiple
-\textsf{inputFrame} use \textsf{Channel::frameAssemb\-le} to assemble
-the frames into the framebuffer. Equalizer takes care of the network
+\textsf{inputFrame}s use \textsf{Channel::frameAssemb\-le} to assemble
+the pixel data into the framebuffer. Equalizer takes care of the network
 transport of frame buffer data between nodes, if needed.
 
+Normally the programmer does not need to interfere with the image
+compositing, or only at a high level, for example to order the input
+frames or to optimize the readback. The following sections provide a
+detailed description of the image compositing API in Equalizer.
+
 \subsubsection{Parallel Direct Send Compositing}
@@
 In order to provide a motivation for the design of the image compositing
-API, the parallel compositing algorithm for direct send is introduced
-here.
-
-The motivation for direct send is to parallelize the costly reassembly
-for database (sort-last) decomposition. With each additional source
-channel, the amount of pixel data to be composited grows
-linearly. Direct send distributes this workload evenly across all source
-channels, and thereby keeps the compositing work per channel constant.
+API, the direct send parallel compositing algorithm is introduced in this
+section.
+
+The motivation for direct send is to parallelize the costly
+recomposition for database (sort-last) decomposition. With each
+additional source channel, the amount of pixel data to be composited
+grows linearly. When using the naive approach of compositing all frames
+on the destination channel, this channel quickly becomes the bottleneck
+in the system. Direct send distributes this workload evenly across all
+source channels, and thereby keeps the compositing work per channel
+constant.
 
@@
 \end{wrapfigure}
-Each channel is responsible for the sort-last compositing of one 2D
-tile. He receives the framebuffer pixels for his tike from all the other
-channels. The size of this tile decreases linearly with the number of
-source channels.
+In direct send compositing, each rendering channel is also responsible
+for the sort-last composition of one screen-space tile. He receives the
+framebuffer pixels for his tile from all the other channels. The size of
+one tile decreases linearly with the number of source channels, which
+keeps the total amount of pixel data per channel constant.
 
@@
 transferred to the destination channel, similarly to an 2D (sort-first)
-compound.
+compound. The amount of pixel data for this part of the compositing
+pipeline also approaches a constant value, i.e., the full frame buffer.
 
@@
 \begin{itemize}
-\item Selection of color and/or depth readback
-\item Restricting the read back to a part of the rendered area
+\item Selection of frame buffer attachments: color and/or depth
+\item Restricting the read back area to a part of the rendered area
 \item Positioning the pixel data correctly on the receiving channels
@@
 
-\subsubsection{Frame, FrameData and Images}
+\subsubsection{Frame, Frame Data and Images}
 
@@
 
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/assembly.pdf}
+  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
+\end{wrapfigure}
 The offset of input and output frames characterizes the position of the
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/assembly.pdf}
-  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
-\end{wrapfigure}
 For output frames, the frame data's pixel viewport (pvp) is the area of
@@
 
-The \textsf{eVolve} example is a scalable volume renderer.
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/slices.pdf}
+  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
+      Volume Rendering}}
+\end{wrapfigure}
+The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
+texture-based volume rendering, where the 3D texture is intersected by
+view-aligned slices. The slices are rendered back-to-front and blended
+together to produce the final image, as shown in
+\fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382 of the German
+  Research Council (DFG)}.
+
+When using 2D (sort-first) or stereo decompositions, no special
+programming is needed to achieve good scalability, as \textsf{eVolve} is
+mostly fill-limited and therefore scales nicely in this mode.
+
+The full power of scalable volume rendering is however in DB (sort-last)
+compounds, where the full volume is divided into separate bricks. Each
+of the bricks is rendered the same as in the other modes. For
+recomposition, the \textsf{RGBA} frame buffer data resulting from these
+render passes then has to be assembled correctly.
+
+DB compounds have the advantage of scaling any part of the volume
+rendering pipeline: texture and main memory (smaller bricks for each
+channel), fill rate (less samples per channel) and IO bandwidth for
+time-dependend data (less data per time step and channel).
+
+\begin{figure}[h!t]
+  \subfigure[]{\label{fBlendOrtho}
+    \includegraphics[width=.45\textwidth]{images/b2f_ortho.pdf}
+  }\hfil
+  \subfigure[]{\label{fBlendPerspective}
+    \includegraphics[width=.55\textwidth]{images/b2f_perspective.pdf}
+  }%
+  {\caption{\small\label{fBlend}Back-to-Front Com\-po\-siting for
+      Orthogonal and Perspective View Frustra}}
+\end{figure}
+For recomposition, the 2D frame buffer contents are blended together to
+form a seamless picture. For currect blending, the frames are ordered in
+the same back-to-front order as the slices used for rendering using the
+same blending parameters. Simplified, the frame buffer images are
+`thick' slices which are `rendered' by writing their content to the
+destination frame buffer using the correct order.
+
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
+For orthographic rendering, determining the compositing order of the
+input frames is trivial. The screen-space orientation of the volume
+bricks determines the order in which they have to be composited. The
+bricks in \textsf{eVolve} are created by slicing the volume along one
+dimension. Therefore the range of the resulting frame buffer images,
+together with the sorting order, is used to arrange the frames during
+compositing. \fig{fBlendOrtho} shows this composition for one view.
+
+Finding the correct assembly order for a perspective frustra is more
+complex. The perspective distortion invalidates a simple orientation
+criteria like the one used for orthographic frustra. For the view and
+frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
+  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
+1-4-3-2. In order to compute the assembly order, \textsf{eVolve} uses
+the angle between the origin$\rightarrow$slice vector and the near
+plane, as shown in \fig{fBlendPerspective}. The result image naturally
+looks the same as the volume rendering would when rendered n a singly
+channel, as shown in \fig{fVolResult}.
 
@@
 scope of Equalizer. Other open source and commercial implementations
-cover this functionality sufficiently and can be integrated with
+cover this functionality sufficiently and can easily be integrated with
 Equalizer.

### 2007-10-08 Developer/ProgrammingGuide/paper.tex
Programming Guide:  update to v0.9.1
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 \date{
-  \vfill
-  \vfill
+  \vspace{4cm}
   \begin{figure}[ht]
@@
   \end{figure}
-  \vfill
-  \vfill
+  \vspace{1cm}
   \textbf{DRAFT}\\[\medskipamount]
-  Version 0.9 for Equalizer v0.4\\[\medskipamount]
+  Version 0.9.1 for Equalizer v0.4\\[\medskipamount]
   \today
@@
     \hline
+    0.9.1   & Oct 8, 2007  & review pass\\
     0.9     & Oct 7, 2007  & add eqHello, draft assembly in eVolve\\
@@
 the domain-specific parallel rendering know-how and abstracts
-configuration, threading, synchronization, windowing and event handling
-for portability. It is a `GLUT on steroids', providing parallel and
-distributed execution, scalable rendering features and fully
-customizable event handling.
+configuration, threading, synchronization, windowing and event
+handling. It is a `GLUT on steroids', providing parallel and distributed
+execution, scalable rendering features and fully customizable event
+handling.
 
@@
 It is recommended to have a compiled version of Equalizer and the
-examples for this Programming Guide. The Quickstart
+examples when working with this Programming Guide. The Quickstart
 Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
@@
 Compiling Equalizer is as simple as running \textsf{make} on Linux or
-building the Equalizer solution on Windows. On Mac OS X, some
-prerequisites have to be installed before running \textsf{make}, as
-explained in \textsf{README.Darwin}.
+building the Equalizer Visual Studio 2005 solution on Windows. On Mac OS
+X 10.4 (Tiger), some prerequisites have to be installed before running
+\textsf{make}, as explained in \textsf{README.Darwin}. Mac OS X 10.5
+Leopard does have all the prerequisites installed by default.
 
@@
 The application connects to an Equalizer server, which chooses a
-configuration for the application. It also provides a render client, to
-be launched by the server. The application reacts on events and controls
-the rendering.
+configuration for the application. Furthermore, the application also
+provides its render client, which will be controlled by the server. The
+application reacts on events, updates its database and controls the
+rendering.
 
@@
 The render client implements --obviously-- the rendering part of an
-application. Its execution is passive, completely driven by rendering
-tasks received from the server. The tasks are executed by calling the
-appropriate task methods (see \sref{ssTaskMethods}) in the correct
-thread and context.
+application. Its execution is passive, that is, it has no main loop and
+is completely driven by rendering tasks received from the server. The
+tasks are executed by calling the appropriate task methods (see
+\sref{ssTaskMethods}) in the correct thread and context.
 
@@
 contribute to the rendering. If it does not implement any render
-client-related code it is reduced to be the application's 'master'
+client-related code it is reduced to be the application's `master'
 process without any OpenGL windows.
@@
 The rendering client can be the same executable as the application, as
-is the case with \textsf{eqPly}. When it is started as a render client,
+it is the case with \textsf{eqPly}. When it is started as a render client,
 the Equalizer initialization routine does not return and takes over the
@@
 \textsf{eqPly} section will discuss the most important task methods. A
-full list can be found on the
-website\footnote{\link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.
+full list can be found on the website\footnote{see
+  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.
+
+\if 0
+\subsection{Thread Safety}
+
+Using threading correctly in OpenGL-based applications is easy with
+Equalizer. Equalizer creates one rendering thread for each graphics
+card. All task methods for a pipe, and therefore all OpenGL commands,
+are executed from this thread. This threading model is the OpenGL
+`threading model', which maintains one current context for each
+thread. It is difficult, if not impossible, to render from different
+threads to the same context.
+
+The rendering threads concurrently render the application's data
+base. The data base should be accessed in a read-only fashion during
+rendering, as it is the case with all modern scene graphs.
+
+The rendering threads run asynchronously to the application's main
+thread. Depending on the configuration's latency, the fall $n$ frames
+behind the last frame finished by the application thread.
+
+TBD support and explain per-node frame sync.
+\fi
 
@@
 unless the pipe's thread hint is set to \textsf{false}. All pipe, window
-and channel task methods are executed from the pipe thread for threaded
-pipes or from the main application thread for non-threaded
-pipes\footnote{see also
+and channel task methods are executed from the pipe thread pipes or from
+the main application thread for non-threaded pipes\footnote{see
   \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.
@@
 children. The frustum can be overridden by children, but this is rarely
-used since there is no use case for this behaviour. The frustrum can be
-specified as a wall, which is completely defined by the bottom-left,
-bottom-right and top-left coordinates relative to the origin, or as a
-projection, which is defined by the position and head-pitch-roll
-orientation of the projector, as well as the horizontal and vertical
-field-of-view and distance to the projection wall. \fig{fFrustra}
-illustrates the wall and projection frustum parameters.
+used since there is no use case for this behaviour.
+
+The frustrum can be specified as a wall, which is completely defined by
+the bottom-left, bottom-right and top-left coordinates relative to the
+origin, or as a projection, which is defined by the position and
+head-pitch-roll orientation of the projector, as well as the horizontal
+and vertical field-of-view and distance of the projection
+wall. \fig{fFrustra} illustrates the wall and projection frustum
+parameters.
 
@@
 of the resulting pixels from the source compound channels. An output
-frame connects to an input frame of the same name, and transports the
+frame connects to an input frame of the same name, which transports the
 selected frame buffer data from the output channel to the input
@@
 In this section the source code of \textsf{eqPly} is discussed in
-detail, and relevant design decision and remarks are raised.
+detail, and relevant design decision, caveat and remarks are raised.
 
@@
 type name ambiguities, in particular for the \textsf{Window} class which
-is frequently used as a type by windowing systems. \fig{fUml} shows the
-most important Equalizer classes and how they are subclassed in the
-\textsf{eqPly} example.
+is frequently used as a type in the global namespace by windowing
+systems. \fig{fUml} shows the most important Equalizer classes and how
+they are subclassed by the \textsf{eqPly} example.
 
@@
 The main function starts off with parsing the command line into the
-\textsf{LocalInitData} data structure, which in part will be distributed
-to all render client nodes. For actual command line parsing is done by
-the \textsf{LocalInitData} class and will be discussed there:
+\textsf{LocalInitData} data structure, part of which will be distributed
+to all render client nodes. The actual command line parsing is done by
+the \textsf{LocalInitData} class and will be discussed in
+\sref{sInitData}:
 
@@
 options\footnote{Equalizer-specific options always start with -\,-eq-},
-e.g., the default server location. Furthermore, a node factory is
+e.g., the default server address. Furthermore, a \textsf{NodeFactory} is
 provided. The \textsf{EQERROR} macro, and its counterparts
@@
 object, thus making it possible to selectively subclass individual
-entity types. For each rendering resource used in the configuration, one
-C++ object will be created:
+entity types, as it is done by \textsf{eqHello}. For each rendering
+resource used in the configuration, one C++ object will be created:
 
@@
 initialize it locally. The application is an \textsf{eq::Client}, which
-is an \textsf{eqNet::Node}. The underlying network distribution in
-Equalizer is a peer-to-peer network structure of
+in turn is an \textsf{eqNet::Node}. The underlying network distribution
+in Equalizer is a peer-to-peer network structure of
 \textsf{eqNet::Node}s. The application programmer rarely is aware of the
@@
 
-After it has finished, the application and Eqply is deinitialized and
-the \textsf{main} function returns:
+After it has finished, the application and \textsf{eqPly} is
+de-initialized and the \textsf{main} function returns:
 
@@
 
-In the \textbf{eqPly} case, the application is also the render client. It has
-three run-time behaviours:
+In the case of \textsf{eqPly} the application is also the render
+client. The \textsf{eqPly} executable has three run-time behaviours:
 
@@
     parameters, which cause \textsf{Client::initLocal} to never
-    return. During exit, the server terminates the process.
+    return. During exit, the server terminates the process. By default,
+    the server starts render client using \textsf{ssh}.
   \item \textbf{Resident render client}: Manually pre-started render
@@
 The second step is to ask the server for a configuration. The
-\textsf{ConfigParams} are a placeholder for later implementations to
-provide additional hints and information to the server for choosing a
-configuration. The configuration is created using
+\textsf{ConfigParams} are a placeholder for later Equalizer
+implementations to provide additional hints and information to the
+server for choosing a configuration. The configuration chosen by the
+server is created locally using
 \textsf{NodeFactory::createConfig}. Therefore it is of type
 \textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
-making q cast necessary:
+making cast necessary:
 
@@
 windows and channels are initialized. It returns \textsf{true} only if
-all init task methods were successful. The \textsf{EQLOG} macro allows
-topic-specific logging. The numeric topic values are specified in the
-respective \textsf{log.h} header files:
+all initialization task methods were successful.
+
+The \textsf{EQLOG} macro allows topic-specific logging. The numeric
+topic values are specified in the respective \textsf{log.h} header
+files, and logging for various topics is enables using the environment
+variable \textsf{EQ\_LOG\_TOPICS}:
 
@@
 
-When the configuration was successfully initialized, the actual main
-loop is executed. The main loop runs until the user exits the
-configuration or a maximum number of frames, specified as a command-line
-argument, has been rendered. The latter is useful for benchmarks. The
+When the configuration was successfully initialized, the main rendering
+loop is executed. The main loop runs until either the user exits the
+configuration or a maximum number of frames has been rendered, specified
+as a command-line argument. The latter is useful for benchmarks. The
 \textsf{Clock} is reused for measuring the overall performance. A new
@@
 \textsf{current - latency}. The latency is specified in the
-configuration file, and allows several outstanding frames for which the
-tasks are already queued in the node and pipe threads for
-execution. This enables overlapped execution and minimizes idle
-times. The first \textsf{latency Config::finishFrame} return
+configuration file, and allows several outstanding frames. This allows
+overlapped execution in the node and pipe threads and minimizes idle
+times. The first \textsf{nLatency Config::finishFrame} return
 immediately, since they have no frame to synchronize
-upon. \fig{fSyncAsync} shows the execution of (hypothetical)
-rendering tasks without latency~(\fig{fSync}) and with a latency of
+upon. \fig{fSyncAsync} shows the execution of (hypothetical) rendering
+tasks without latency~(\fig{fSync}) and with a latency of
 one~(\fig{fAsync}).
+
+With \textsf{eqPly} a speedup of 15\% has been observed on a five-node
+rendering
+cluster when using a latency of one instead of no latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}.
 
@@
 
-In the second and third case, when the executable is used as a render
-client, \textsf{Client::initLocal} never returns. Therefore the
-application's main loop is never executed. In order to keep the client
-resident, the \textsf{eqPly} example overrides the client loop to keep
-it running beyond one configuration run:
+In the second and third usage case of the \textsf{eqPly}, when the
+executable is used as a render client, \textsf{Client::initLocal} never
+returns. Therefore the application's main loop is never executed. In
+order to keep the client resident, the \textsf{eqPly} example overrides
+the client loop to keep it running beyond one configuration run:
 
@@
 
-Distributed object are created by subclassing from
+Distributed objects are created by subclassing from
 \textsf{eqNet::Object}. Distributed objects can be static (immutable) or
@@
 
-\subsubsection{InitData - a Static Distributed Object}
+\subsubsection{\label{sInitData}InitData - a Static Distributed Object}
 
@@
 
+On the application side, the class \textsf{LocalInitData} subclasses
+\textsf{InitData} to provide the command line parsing and to set the
+initial values. The render nodes only instantiate the distributed part
+in \textsf{InitData}
+
 A static distributed object either has to provide a pointer and size to
-its data using \textsf{setInstanceData} or it has to implement
+its data using \textsf{setInstanceData}, or it has to implement
 \textsf{getInstanceData} and \textsf{applyInstanceData}. The first
@@
 
-The data output and input streams perform no type checking on the
-variables. It is the application's responsibility to match the order and
+The data output and input streams perform no type checking on the data
+stream. It is the application's responsibility to match the order and
 types of variables during serialization and deserialization exactly.
@@
 
-The frame data is layed out in one anonymous structure in
+The \textsf{eqPly} frame data is layed out in one anonymous structure in
 memory. It also does not track changes since it is relatively small in
-size and changes frequently. Therefore, for the instance and delta
-date are the same and set in the constructor:
+size and changes frequently. Therefore, the instance and delta data are
+the same and set in the constructor:
 
@@
 
-The configuration is driving the applications rendering, that is, it is
+The configuration is driving the application's rendering, that is, it is
 responsible for updating the data based on received events, requesting
@@
 parent's initialization methods are always executed before any child
-initialization method. This is necessary to allow a speedy startup of
-the configuration on large-scale graphics clusters. On the other hand,
-it means that initialization functions are called even if the parent's
-initialization has failed.
+initialization method. This parallelization is necessary to allow a
+speedy startup of the configuration on large-scale graphics clusters. On
+the other hand, it means that initialization functions are called even
+if the parent's initialization has failed.
 
@@
 The frame data is registered first, since its identifier is transmitted
-using the initialization data, which is registered afterwards. The
+using the initialization data. The initialization data is registered
+afterwards so that it can transmit the frame data's identifier. The
 identifier of the initialization data is transmitted to the render
@@
 provide extensive support for tracking device, as this is an orthogonal
-problem to parallel rendering, and has been solved already by a number
-of implementations\footnote{VRCO Trackd, VRPN, etc.}. The example
-code in \textsf{eqPly} is one reference implementation for the
-integration of such a tracking library:
+problem to parallel rendering. Tracking device support has been solved
+already by a number of implementations\footnote{VRCO Trackd, VRPN,
+  etc.}, which can easily be integrated with Equalizer. The example code
+in \textsf{eqPly} is one reference implementation for the integration of
+such a tracking library:
 
@@
 The rendering frames are issued by the application. The \textsf{Config}
-only overrides \textsf{startFrame} in order to update the its data
-before forwarding the start frame request to the \textsf{eq::Config}.
+only overrides \textsf{startFrame} in order to update its data before
+forwarding the start frame request to the \textsf{eq::Config}.
 
@@
 The camera position is updated and the frame data is commited, which
-generates a new version. This version is passed to the rendering
-callbacks and will be used to synchronize the frame data to the state
-belonging to the current frame:
+generates a new version for this object. This version is passed to the
+rendering callbacks and will be used by the rendering threads to
+synchronize the frame data to the state belonging to the current frame:
 
@@
     {
-        case eq::ConfigEvent::WINDOW_CLOSE:
-            _running = false;
-            return true;
-
         [...]
-
         case eq::ConfigEvent::POINTER_MOTION:
@@
 
-During node initialization the initialization data is mapped to
-a local instance using the passed identifier from
-\textsf{Config::init}. The model is loaded based on the filename in the
-initialization data. No pipe, window or channel tasks methods are
-executed before \textsf{Node::configInit} has returned.
+During node initialization the init data is mapped to a local instance
+using the passed identifier from \textsf{Config::init}. The model is
+loaded based on the filename in the initialization data. No pipe, window
+or channel tasks methods are executed before \textsf{Node::configInit}
+has returned.
 
@@
 e.g., \textsf{Window::startFrame}, are currently empty since the
-synchronization is implicit due to the execution within the thread.
+synchronization is implicit due to the sequential execution within the
+thread.
 
@@
 \textsf{frameFinish} task methods. These task methods unlock their
-parents by calling \textsf{finishFrame}.
+parents by calling \textsf{releaseFrame}.
 
@@
 vital for the execution that \textsf{Node::startFrame} or
-\textsf{Node::finishFrame} are called, respectively. The default
-implementation of the node task methods does take care of that.
+\textsf{Node::releaseFrame} are called, respectively. The default
+implementation of the node task methods do take care of that.
 
@@
 command line option. Therefore the implementation of
-\textsf{selectWindowSystem} is overwritten, and return the specified
-window system, if it is valid:
+\textsf{selectWindowSystem} is overwritten, and returns the specified
+window system, if it is supported:
 
@@
 be used only during window initialization and exit, not during
-rendering\footnote{for various reasons \textsf{enterCarbon} might block
-  up to 50 milliseconds}. Carbon call in the window event handling
-routine \textsf{Window::processEvent} are thread-safe. Please contact
-the Equalizer developer mailing list if you need to use Carbon calls on
-a per-frame basis.
+rendering. For various reasons \textsf{enterCarbon} might block up to 50
+milliseconds. Carbon call in the window event handling routine
+\textsf{Window::processEvent} are thread-safe, since the global carbon
+lock is set in this method. Please contact the Equalizer developer
+mailing list if you need to use Carbon calls on a per-frame basis.
 
@@
 context. Subclassed windows should maintain all data specific to the
-OpenGL context. The Equalizer window creation routines share the OpenGL
-context with the first window of the pipe, this allowing the reuse of
-OpenGL objects, e.g., display lists and textures.
+OpenGL context. The Equalizer window creation routines shares the OpenGL
+context with the first window of the pipe, thus allowing the reuse of
+OpenGL objects between all windows of one pipe, e.g., display lists and
+textures.
 
@@
 task method to create the drawable and context. The child task method
-depends on the pipe's window system. The default implementations
+depends on the pipe's window system. The default implementations of
 \textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
@@
 drawable and context was created successfully, \textsf{configInit} calls
-\textsf{configInitGL}, which performs the generic OpenGL state
-initialization. The default implementation sets up some typical OpenGL
-state, e.g., it enables the depth test.
+\textsf{configInitGL}, which is emant to perform the generic OpenGL
+state initialization. The default implementation sets up some typical
+OpenGL state, e.g., it enables the depth test.
 
@@
 multipipe, multi-threaded execution environment. It uses an object
-manager, which is described in detail below. It is used in conjunction
-with a reference pointer here, since it is potentially `owned' by
-multiple windows at the same time.
+manager, which is described in detail in \sref{sObjectManager}. It is
+used in conjunction with a reference pointer here, since it is
+potentially `owned' by multiple windows at the same time.
 
@@
 ID used later by the channel for rendering. The details of this code are
-omitted here, since they are pretty straight-forward and not
+omitted here, since they are pretty straight-forward OpenGL code and not
 Equalizer-specific.
 
-The window exit deallocates all OpenGL objects if the state object is
+The window exit deallocates all OpenGL objects when the state object is
 about to be disposed. The object manager does not delete the object in
@@
 
-\subsubsection{Object Manager}
+\subsubsection{\label{sObjectManager}Object Manager}
 
@@
 can easily be replaced to use the rendering code in a stand-alone
-application, e.g., for an GLUT-based renderer.
+application, e.g., for a GLUT-based renderer.
 
@@
 The OM is a template class. The template type is the key type, by which
-objects can be identified. The same key is used by all contexts to get
-the OpenGL name of an object. In \textsf{eqPly}, a key of type
-\textsf{const void *} is used. The rendering code uses the address of
-the data item to be rendered to obtain the associated OpenGL object.
+objects are identified. The same key is used by all contexts to get the
+OpenGL name of an object. In \textsf{eqPly}, a key of type \textsf{const
+  void *} is used. The rendering code uses the address of the data item
+to be rendered as the key to obtain the associated OpenGL object.
 
@@
 
-Currently handling for display lists, textures and buffers is
+Currently handling for display lists, VBO's, textures and shaders is
 implemented. For each object, the following functions are available:
@@
 \item[supportsObjects()]: returns true if the usage for this particular
-  type of objects is supported. For basic objects, which are supported
-  on all OpenGL implementations, this function is not implemented.
+  type of objects is supported. For objects available in OpenGL 1.1 or
+  earlier, this function is not implemented.
 \item[getObject( key )]: returns the object associated with the given
@@
 actual rendering code. The channel is used to perform various rendering
-operations.
+operations for the compounds.
 
@@
 reasonable values to contain the whole model. During rendering, the near
-and far planes are adjusted to the current model position:
+and far planes are adjusted dynamically to the current model position:
 
@@
 routine contains the application's OpenGL rendering code, which is being
-rendered using the rendering context provided by Equalizer. As most of
-the other task methods, \textsf{frameDraw} is called in parallel by
+rendered using the contextual information provided by Equalizer. As most
+of the other task methods, \textsf{frameDraw} is called in parallel by
 Equalizer on all pipe threads in the configuration. Therefore the
-rendering should not write to the database, which is the case for all
+rendering should not write to shared data, which is the case for all
 major scene graph implementations.
@@
 sort-last rendering, \textsf{eqPly} uses a different color for each
-channell to illustrate the database decomposition, as shown in
+channel to illustrate the database decomposition, as shown in
 \fig{fDB}. The Equalizer channel provides a method to obtain a random,
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Rendering Loop}}
-\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 range. If it is partially visible or partially in range, the children of
-the node are added to the candidate list:
+the node are added to the candidate list. \fig{fRender} shows a flow
+chart of the rendering algorithm, which performs efficient view frustum
+and range culling.
+
+\begin{wrapfigure}{r}{.6\textwidth}
+  \includegraphics[width=.6\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Rendering Loop}}
+\end{wrapfigure}
+The actual rendering uses display lists or vertex buffer objects. The
+OpenGL objects are allocated using the object manager. The rendering is
+done by the leaf nodes, which are small enough to store the vertex
+indices in a \textsf{short} value for optimal performance with VBO's.
+
+The leaf nodes reuse the objects stored in the object manager, or create
+and set up new objects if it was not yet set up. Since one object
+manager is used per thread (pipe), this allows a thread-safe sharing of
+the compiled display lists or VBO's across all windows of a pipe.
+
+The main rendering loop in \textsf{eqPly} looks like this:
 
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 For orthographic rendering, determining the compositing order of the
@@
 
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 Finding the correct assembly order for a perspective frustra is more
@@
   courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
-1-4-3-2. In order to compute the assembly order, \textsf{eVolve} uses
-the angle between the origin$\rightarrow$slice vector and the near
-plane, as shown in \fig{fBlendPerspective}. The result image naturally
-looks the same as the volume rendering would when rendered n a singly
-channel, as shown in \fig{fVolResult}.
+1-4-3-2.
+
+In order to compute the assembly order, \textsf{eVolve} uses the angle
+between the origin$\rightarrow$slice vector and the near plane, as shown
+in \fig{fBlendPerspective}. The result image naturally looks the same as
+the volume rendering would when rendered n a single channel, as shown in
+\fig{fVolResult}.
+
+The assembly algorith described in this section also works with parallel
+compositing algorithms, such as direct-send.
 

### 2007-10-19 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-\usepackage{tabularx,graphicx,a4,listings,wrapfig,subfigure}
+\usepackage{tabularx,graphicx,a4,listings,wrapfig,subfigure,textcomp}
 
@@
   \textbf{DRAFT}\\[\medskipamount]
-  Version 0.9.1 for Equalizer v0.4\\[\medskipamount]
+  Version 0.9.2 for Equalizer v0.4\\[\medskipamount]
   \today
@@
     \hline
+    0.9.2   & Oct 19, 2007 & Jonas' feedback, changes related to
+                             per-node frame sync\\
     0.9.1   & Oct 8,  2007 & review pass\\
@@
 
-Any questions related to Equalizer programming and this Programming
-Guide should be directed to the \textsf{eq-dev} mailing
-list\footnote{see \link{http://www.equalizergraphics.com/lists.html}}.
-
-Equalizer is the next evolution of generic parallel programming
+Equalizer is the next step in the evolution of generic parallel programming
 interfaces for visualization applications using OpenGL. Existing
+% rephrase ( aber ich weiss auch nicht wie :) )
 solutions, such as SGI's OpenGL Multipipe\texttrademark\ SDK, VRCO's
@@
 
+If you have any question regarding Equalizer programming, this programming
+guide, or specific problems you encountered, please direct them to the 
+ \textsf{eq-dev} mailing list\footnote{see \link{http://www.equalizergraphics.com/lists.html}}.
+
 \section{Getting Started}
 
-\subsection{Compiling and running \textsf{eqPly}}
-
-It is recommended to have a compiled version of Equalizer and the
-examples when working with this Programming Guide. The Quickstart
+\subsection{Installing Equalizer and running \textsf{eqPly}}
+
+Equalizer can be installed by downloading the
+distribution\footnote{\link{http://www.equalizergraphics.com/downloads.html}}
+and by compiling the source code. After installing Equalizer, please
+take a look at the Quickstart
 Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
-should be executed once to get familar with the capabilities of the
-\textsf{eqPly} example.
+to get familiar with the capabilities of the \textsf{eqPly} example.
 
@@
 
+The Equalizer architecture is based on a client-server model. Its client
+library exposes all functionality discussed in this document to the
+programmer, and provides communication between the different Equalizer
+processes.
+
+\subsubsection{Server}
+Each Equalizer server is responsible for managing one visualization
+system, i.e., a shared memory system or graphics cluster.  It controls
+and launches the application's rendering clients.  Currently, Equalizer
+currently only supports one application per server, but it will provide
+concurrent and efficient multi-application support in future.
+
+\subsubsection{Application}
+
+The application connects to an Equalizer server and receives a
+configuration.  Furthermore, the application also provides its render
+client, which will be controlled by the server. The application reacts
+on events, updates its database and controls the rendering.
+
+\subsubsection{Render Clients}
+
 \begin{wrapfigure}{r}{.4\textwidth}
@@
 \end{wrapfigure}
-All functionality of Equalizer is accessed through the Equalizer client
-library, which implements all functionality discussed in this
-document. The client library provides communication between the Equalizer
-processes.
-
-\subsubsection{Server}
-
-An Equalizer server is responsible for managing one visualization
-system, i.e., a shared memory system or graphics cluster. Currently it
-is only useful for running one application at a time, but it will be
-extended to support multiple applications concurrently and efficiently
-on one system. The server controls and potentially launches the
-application's rendering clients.
-
-\subsubsection{Application}
-
-The application connects to an Equalizer server, which chooses a
-configuration for the application. Furthermore, the application also
-provides its render client, which will be controlled by the server. The
-application reacts on events, updates its database and controls the
-rendering.
-
-\subsubsection{Render Clients}
-
-The render client implements --obviously-- the rendering part of an
-application. Its execution is passive, that is, it has no main loop and
+The render client implements the rendering part of an
+application. Its execution is passive, it has no main loop and
 is completely driven by rendering tasks received from the server. The
@@
 
-The application might be a rendering client, in which case it can also
-contribute to the rendering. If it does not implement any render
-client-related code it is reduced to be the application's `master'
+The application might contain a rendering client, in which case it can 
+also contribute to the rendering. If it does not implement any render
+client-related code, it is reduced to be the application's `master'
 process without any OpenGL windows.
@@
 The \textsf{eqHello} redraw function renders six rotating, colored quads
-around the origin, where the viewer is usually positioned. The Equalizer
-\textsf{frameDraw} meth\-od is used as a convience function to setup the
+around the origin. The \textsf{frameDraw} meth\-od provided by the
+\textsf{eq::Channel} can be used as a convience function to setup the
 frustum and other OpenGL state. After setting up some additional
@@
 Equalizer uses a C++ programming interface. The API is minimally
-invasive, that is, Equalizer imposes only the minimal, natural execution
-framework upon the application. It does not provide a scene graph or
-does interfere in any other way with the application's rendering
-code. The restructuring work needed for Equalizer is needed anyway in
-order to parallelize the application for rendering.
+invasive, so Equalizer imposes only a minimal, natural execution
+framework upon the application. It does not provide a scene graph, or
+interfere in any other way with the application's rendering code. The
+restructuring work required for Equalizer is needed anyway in order to
+parallelize the application for rendering.
 
@@
 whereas methods called by Equalizer (`Task Methods') have the form
-\textsf{nounVerb}. For example the application calls
+\textsf{nounVerb}. For example, the application calls
 \textsf{Config::startFrame} to render a new frame, which causes --among
@@
 
-The application subclasses Equalizer objects and overrides virtual
-functions to implement certain functionality, e.g., the application's
+The application inherits from Equalizer classes and overrides virtual
+functions to implement certain functionality, such as the application's
 OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
-methods are in concept similar to C function callbacks. The
+methods are similar in concept to C function callbacks. The
 \textsf{eqPly} section will discuss the most important task methods. A
@@
 \fig{fConfig} shows one example configuration for a four-side
-CAVE\texttrademark, running on two machines (node) using three graphics
-cards (pipe) with one window each to render to the four output channels
+CAVE\texttrademark, running on two machines (nodes) using three graphics
+cards (pipes) with one window each to render to the four output channels
 connected to the projectors for each of the walls. The compound
@@
 For testing and development purposes it is possible to use multiple
-instances for one resource, e.g., to run multiple render client nodes on
-one computer. For deployment one node and pipe should be used for each
+instances for one resource, e.g. to run multiple render client nodes on
+one computer. For deployment, one node and pipe should be used for each
 computer and graphics card, respectively.
@@
 
+%FIXME confusing
 The root of the resource tree is the \textsf{eq::Config}, which
@@
 The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU).  In
-the current implementation it is also one operating system thread,
-unless the pipe's thread hint is set to \textsf{false}. All pipe, window
-and channel task methods are executed from the pipe thread pipes or from
-the main application thread for non-threaded pipes\footnote{see
+the current implementation it is also one operating system
+thread. Non-threaded pipes are supported for integrating with
+thread-unsafe libraries, but have various performance caveats. All pipe,
+window and channel task methods are executed from the pipe's thread or
+in the case of non-threaded pipes, from the main application
+thread\footnote{see
   \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.
@@
 
-An \textsf{eq::Window} holds a drawable and OpenGL context. The
-drawable can be an on-screen window or an off-screen PBuffer or
-FBO\footnote{off-screen drawables are not yet implemented, but can be
-  created by the application and used with Equalizer}. The window holds
-window-system-specific handles to the drawable and context, e.g., an X11
+An \textsf{eq::Window} holds a drawable and an OpenGL context. The
+drawable can be an on-screen window or an off-screen\footnote{Off-screen drawables are not implemented yet, but can be
+  created by the application and used with Equalizer} PBuffer or
+FBO(Framebuffer object). The window holds
+window-system-specific handles to the drawable and context, e.g. an X11
 window \textsf{XID} and \textsf{GLXContext} for the glX window system.
@@
 channel's viewport might be overwritten when it is rendering for another
-channel.
+channel during scalable rendering.
 
@@
 
-How the rendering resources are to be used is configured using a
-compound tree. Although the API does not currently expose compounds, the
-basic design behind compounds is explained here to increase the
-understanding of the architecture behind Equalizer.
+Usage of the rendering resources is configured using a compound 
+tree. Although the API does not currently expose compounds, the
+basic design behind compounds is explained here for a better 
+understanding of the Equalizer architecture.
 
@@
 \end{wrapfigure}
-The channels of the leaf compounds (source channels) in the compound
-tree execute the rendering for the top-most compound's channel
-(destination channel). The top-most compound with a channel defines the
-2D pixel viewport rendered by all leaf compounds. The destination
-channel and pixel viewport can not be overridden.
+The channels of the leaf compounds in the compound tree are designated
+as source channels, and the channel of the top-most compound is the
+destination channel. Source channels execute the rendering for the
+destination channel. The top-most compound with a channel defines the 2D
+pixel viewport rendered by all leaf compounds. The destination channel
+and pixel viewport can not be overridden.
 
 Compounds have a frustum description, which is inherited to the
-children. The frustum can be overridden by children, but this is rarely
-used since there is no use case for this behaviour.
-
-The frustrum can be specified as a wall, which is completely defined by
-the bottom-left, bottom-right and top-left coordinates relative to the
-origin, or as a projection, which is defined by the position and
-head-pitch-roll orientation of the projector, as well as the horizontal
-and vertical field-of-view and distance of the projection
+children. The frustrum can be specified as a wall, which is completely
+defined by the bottom-left, bottom-right and top-left coordinates
+relative to the origin, or as a projection, which is defined by the
+position and head-pitch-roll orientation of the projector, as well as
+the horizontal and vertical field-of-view and distance of the projection
 wall. \fig{fFrustra} illustrates the wall and projection frustum
@@
 and serves as a reference implementation of an Equalizer-based
-application of medium complexity. Its focus is not on rendering features
-or visual quality, but it is used as a show case for the integration of
-most of the Equalizer features.
+application of medium complexity. Its focus lies not on rendering features
+or visual quality, but on providing a working show case for the integration
+of most of the Equalizer features.
 
@@
 
-The node factory is used by Equalizer to create the object instances for
+The node factory is used by Equalizer to create the objects or instances for
 the rendering entities. Each of the classes inherits from the same type
@@
 in Equalizer is a peer-to-peer network structure of
-\textsf{eqNet::Node}s. The application programmer rarely is aware of the
-classes in the \textsf{eqNet} namespace, but both the
+\textsf{eqNet::Node}s. The application programmer does not have to be
+aware of the classes in the \textsf{eqNet} namespace, but both the
 \textsf{eq::Client} and the server are \textsf{eqNet::Node}s. The local
@@
 
-After it has finished, the application and \textsf{eqPly} is
+After finishing, the application and \textsf{eqPly} are
 de-initialized and the \textsf{main} function returns:
@@
 
-In the case of \textsf{eqPly} the application is also the render
+In the case of \textsf{eqPly}, the application is also the render
 client. The \textsf{eqPly} executable has three run-time behaviours:
@@
 Finally it is time to initialize the configuration. For statistics, the
-time for this operation is measures and printed. During initialization
+time for this operation is measured and printed. During initialization
 the server launches and connects all render client nodes, and calls the
@@
 topic values are specified in the respective \textsf{log.h} header
-files, and logging for various topics is enables using the environment
+files, and logging for various topics is enabled using the environment
 variable \textsf{EQ\_LOG\_TOPICS}:
@@
 sends them to the appropriate render client nodes. The render client
-nodes dispatch the tasks to the correct node or pipe thread, were they
-are executed in the order they arrive.
+nodes dispatch the tasks to the correct node or pipe thread, where they
+are executed in order of arrival.
 
@@
 configuration file, and allows several outstanding frames. This allows
-overlapped execution in the node and pipe threads and minimizes idle
+overlapping execution in the node and pipe threads and minimizes idle
 times. The first \textsf{nLatency Config::finishFrame} return
@@
 
-With \textsf{eqPly} a speedup of 15\% has been observed on a five-node
+With \textsf{eqPly}, a speedup of 15\% has been observed on a five-node
 rendering
@@
 up with the latency. It returns after all outstanding frames have been
-rendered, and is needed here to provide an accurate measurement of the
+rendered, and is needed to provide an accurate measurement of the
 framerate:
@@
 The remainder of the application code cleans up in the reverse order of
-the initialization. The config is exited, released and the connection to
+initialization. The config is exited, released and the connection to
 the server is closed:
@@
 
-In the second and third usage case of the \textsf{eqPly}, when the
+In the second and third use case of the \textsf{eqPly}, when the
 executable is used as a render client, \textsf{Client::initLocal} never
@@
 initial data to all rendering nodes, as well as a versioned object to
-provide frame-specific data, such as the camera position, to the
+provide frame-specific data such as camera position to the
 rendering methods.
@@
 \textsf{getInstanceData} and \textsf{applyInstanceData}. The first
-approach can be used if all distributed member variables can be outlayed
+approach can be used if all distributed member variables can be stored
 in one contiguous block of memory. The second approach is used
@@
 length. Therefore it uses the second approach of manually serializing
-and deserializing its data. The serialization is done in
+and deserializing its data. Serialization in
 \textsf{getInstanceData} and deserialization in
-\textsf{applyInstanceData} by streaming all member variable to or from
+\textsf{applyInstanceData} is performed by streaming all member variable to or from
 the provided data streams. The data transport between nodes is
@@
 
-The data output and input streams perform no type checking on the data
-stream. It is the application's responsibility to match the order and
-types of variables during serialization and deserialization exactly.
+The data input and output streams perform no type checking on the data.
+It is the application's responsibility to exactly match the order and
+types of variables during serialization and deserialization.
 
@@
 
-Versioned objects have to override \textsf{isStatic} to return false, to
+Versioned objects have to override \textsf{isStatic} to return false to
 indicate that they are versioned. The current implementation has the
@@
 
-The frame data is registered first, since its identifier is transmitted
-using the initialization data. The initialization data is registered
-afterwards so that it can transmit the frame data's identifier. The
-identifier of the initialization data is transmitted to the render
-client nodes using the \textsf{initID} parameter of
-\textsf{eq::Config::init}. Equalizer will pass this identifier to all
-\textsf{configInit} calls of the respective objects:
+The frame data is registered before the initialization data, since its
+identifier is transmitted using the \textsf{InitData}. The identifier of
+the initialization data is transmitted to the render client nodes using
+the \textsf{initID} parameter of \textsf{eq::Config::init}. Equalizer
+will pass this identifier to all \textsf{configInit} calls of the
+respective objects:
 
@@
   etc.}, which can easily be integrated with Equalizer. The example code
-in \textsf{eqPly} is one reference implementation for the integration of
+in \textsf{eqPly} provides a reference implementation for the integration of
 such a tracking library:
@@
 
-The exit of the configuration stops the render clients by calling
+The exit function of the configuration stops the render clients by calling
 \textsf{eq::Config::exit}, and then deregisters the initialization and
@@
 
-The rendering frames are issued by the application. The \textsf{Config}
-only overrides \textsf{startFrame} in order to update its data before
-forwarding the start frame request to the \textsf{eq::Config}.
+The rendering frames are issued by the application. The
+\textsf{eqPly::Config} only overrides \textsf{startFrame} in order to
+update its data before forwarding the start frame request to the
+\textsf{eq::Config}.
 
@@
 The camera position is updated and the frame data is commited, which
-generates a new version for this object. This version is passed to the
+generates a new version of this object. This version is passed to the
 rendering callbacks and will be used by the rendering threads to
@@
 
-Events are send by the render clients to the application using
+Events are sent by the render clients to the application using
 \textsf{eq::Config::sendEvent}. At the end of the frame,
@@
 
-Foreach active render client, one \textsf{eq::Node} instance is
+For each active render client, one \textsf{eq::Node} instance is
 created on the appropriate machine. Nodes are only instantiated on their
-render client processes, i.e., each process should have only one
+render client processes, i.e., each process should only have one
 instance of the \textsf{eq::Node} class. The application process might
@@
 During node initialization the init data is mapped to a local instance
-using the passed identifier from \textsf{Config::init}. The model is
+using the identifier passed from \textsf{Config::init}. The model is
 loaded based on the filename in the initialization data. No pipe, window
@@
 
-The node config exit deletes the loaded model and unmaps the
+The node config exit function deletes the loaded model and unmaps the
 initialization data:
@@
 during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
-\textsf{frameStart} task methods of the various entities. The entity
+\textsf{frameStart} task methods of the various entities. The entities
 unlock all its children by calling \textsf{startFrame}, e.g.,
@@
 The explicit synchronization of child or parent resources allows the
-application to optimize the processing, by doing certain, independent
+application to optimize the processing, by performing certain, independent
 operations when the child or parent resources are already unlocked.
+
+The default thread synchronization synchronizes all
+\textsf{Channel::frameDraw} operations on a \textbf{single} node with
+the node's main thread. This facilitates porting, since the scene
+database does not have to be multi-buffered as all threads of the node
+process render the same frame.
+
+The per-node frame synchronization does not break the asynchronous
+execution across rendering nodes.  Advanced applications can remove per-node
+frame synchronization.
+
+The per-node frame synchronization is achieved through the
+\textsf{startFrame\textrightarrow wait\-Fra\-me\-Started} and
+\textsf{releaseFrameLocal\textrightarrow waitFrameLocal} synchronization
+points. Note that this synchronization is only per-node, different nodes
+in the cluster still run asynchronously.
+
+The first synchronization ensures that the application (node) thread is
+done modifiying the data, typically during event processing. The pipe
+threads call \textsf{Node::waitFrameStarted} which blocks until the node
+calls \textsf{startFrame} in \textsf{Node::frame\-Start}.
+
+The second pair ensures that after \textsf{Config::finishFrame} all pipe
+threads are done rendering the current frame, and therefore do not
+access shared data. The node's \textsf{frameDrawFinish} waits for all
+local pipes to release the synchronization by calling
+\textsf{Pipe::releaseFrameLocal}, which happens by default in
+\textsf{Pipe::frameDrawFinish}. The \textsf{frameDrawFinish} methods are
+called after all \textsf{Channel::frameDraw} of the corresponding thread
+have been executed.
 
@@
 \fig{fFrameSync} outlines the synchronization for the application, node
-and pipe classes. The window and channel synchronization is similar and
-omitted for simplicity. The \textsf{eqPly} example does not override
-\textsf{Node::frameStart} or \textsf{frameFinish}, but it is absolutely
-vital for the execution that \textsf{Node::startFrame} or
-\textsf{Node::releaseFrame} are called, respectively. The default
-implementation of the node task methods do take care of that.
+and pipe classes. Please note that \textsf{Config::finishFrame} does
+block until the current frame has been released locally and until the
+frame \textsf{current - latency} has been released by all nodes. The
+window and channel synchronization is similar and omitted for
+simplicity.
+
+It is absolutely vital for the execution that \textsf{Node::start\-Fra\-me}
+and \textsf{Node::release\-Fra\-me} are called, respectively. The default
+implementation of the node task methods does take care of that.
+
+Since \textsf{eqPly} multi-buffers all dynamic data, it can completely
+remove frame synchronization by:
+
+\begin{itemize}
+\item releasing the local synchronization in \textsf{Node::frameStart}
+\item not calling \textsf{Node::waitFrameStarted} in
+  \textsf{Pipe::frameStart}
+\item not waiting for the pipe synchronization in \textsf{Node::frameDrawFinish}
+\end{itemize}
+
+Releasing the local synchronization early is done by calling
+\textsf{releaseFrameLocal}, and \textsf{Node::frameDrawFinish} is
+overwritten with an empty implementation:
+
+{\footnotesize\begin{lstlisting}
+void Node::frameDrawFinish( const uint32_t frameID,
+                            const uint32_t frameNumber )
+{ /* nop, see frameStart */ }
+
+void Node::frameStart( const uint32_t frameID, const uint32_t frameNumber )
+{
+    startFrame( frameNumber ); // unlock pipe threads
+
+    // Don't wait for pipes to release frame locally, sync not needed since all
+    // dynamic data is multi-buffered
+    releaseFrameLocal( frameNumber );
+}
+\end{lstlisting}}
 
@@
 All task methods for a pipe and its children are executed in a separate
-thread. This approach optimizes usage of the GPU resource, since all
-tasks are executed serially and do not compete for the usage, i.e.,
+thread. This approach optimizes usage of the GPU, since all
+tasks are executed serially and do not compete for resources and
 context switches are minimized. Later versions of Equalizer might
@@
 provided by the initialization data. The initialization in
-\textsf{eq::Pipe} does GPU-specific initialization, e.g., the display
-connection is opened when glX/X11 is used:
+\textsf{eq::Pipe} does GPU-specific initialization, which is
+window-system-dependent. On AGL the display ID is determined, and on glX
+the display connection, to be used by the child windows, is opened.
 
@@
 
-The config exit is again symmetric to the config initialization. The
+The config exit function is similar to the config initialization. The
 frame data is unmapped and GPU-specific data is de-initialized by
@@
 rendering. For various reasons \textsf{enterCarbon} might block up to 50
-milliseconds. Carbon call in the window event handling routine
+milliseconds. Carbon calls in the window event handling routine
 \textsf{Window::processEvent} are thread-safe, since the global carbon
@@
 {
+    // don't wait for node to start frame, local sync not needed
+    // node->waitFrameStarted( frameNumber );
     _frameData.sync( frameID );
@@
 context. Subclassed windows should maintain all data specific to the
-OpenGL context. The Equalizer window creation routines shares the OpenGL
+OpenGL context. The Equalizer window creation routines share the OpenGL
 context with the first window of the pipe, thus allowing the reuse of
-OpenGL objects between all windows of one pipe, e.g., display lists and
-textures.
+OpenGL objects such as display lists and textures between all 
+windows of one pipe.
 
@@
 task method to create the drawable and context. The child task method
-depends on the pipe's window system. The default implementations of
+depends on the window system of the pipe. The default implementations of
 \textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
@@
 drawable and context was created successfully, \textsf{configInit} calls
-\textsf{configInitGL}, which is emant to perform the generic OpenGL
+\textsf{configInitGL}, which will perform the generic OpenGL
 state initialization. The default implementation sets up some typical
@@
 attributes, such as stereo, doublebuffer and stencil can be set to
-\textsf{eq::AUTO}, in which case the default implementation tries to use
-them and gradually backs off upon failure.
+\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
+enable them if possible.
 
@@
 manager, which is described in detail in \sref{sObjectManager}. It is
-used in conjunction with a reference pointer here, since it is
+used in conjunction with a reference-counting smart-pointer here, since it is
 potentially `owned' by multiple windows at the same time.
@@
 The logo texture is loaded from the file system and bound to a texture
-ID used later by the channel for rendering. The details of this code are
-omitted here, since they are pretty straight-forward OpenGL code and not
+ID used later by the channel for rendering. A code listing is ommitted,
+since the code consists of standard OpenGL calls and is not
 Equalizer-specific.
 
-The window exit deallocates all OpenGL objects when the state object is
+The window exit function deallocates all OpenGL objects when the state object is
 about to be disposed. The object manager does not delete the object in
 its destructor, since it does not know if an OpenGL context is still
-current. In any case, \textsf{eq::Window::configExit} is called to
+current. Additionally, \textsf{eq::Window::configExit} is called to
 destroy the drawable and context:
@@
 
-The object manager is --strictly speaking-- not part of the window. It
+The object manager is not strictly a part of the window. It
 is discussed here since the \textsf{eqPly} window uses an object manager.
@@
 OpenGL objects across shared contexts. Typically one OM is used for each
-set of shared contexts, which normally spawns all contexts of a single
+set of shared contexts and spawns all contexts of a single
 GPU\footnote{\link{http://www.equalizergraphics.com/documents/design/objectManager.html}}.
 
-The OM is a template class. The template type is the key type, by which
-objects are identified. The same key is used by all contexts to get the
+The OM is a template class. The template type is the key which is used to
+identify objects. The same key is used by all contexts to get the
 OpenGL name of an object. In \textsf{eqPly}, a key of type \textsf{const
@@
 
-The usage of objects is reference counted. If an application releases
-the objects properly, they are automatically de-allocated. It is also
-possible to manually manage de-allocation of objects, which is often the
-more convenient use case.
-
-Currently handling for display lists, VBO's, textures and shaders is
+All objects managed by the OM are reference counted. If an application
+releases the objects properly, they are automatically de-allocated. It
+is also possible to manually manage de-allocation of objects, which
+might be more convenient in some cases.
+
+Currently, support for display lists, VBO's, textures and shaders is
 implemented. For each object, the following functions are available:
@@
 The channel is the heart of the application in that it contains the
-actual rendering code. The channel is used to perform various rendering
+actual rendering code. The channel is used to perform the various rendering
 operations for the compounds.
@@
 Equalizer on all pipe threads in the configuration. Therefore the
-rendering should not write to shared data, which is the case for all
-major scene graph implementations.
+rendering must not write to shared data, which is the case for all major
+scene graph implementations.
 
@@
 Furthermore, a white color is set in case the model does not contain
-color information, or the color information is not to be used. In
+color information, or the color information is not used. In
 sort-last rendering, \textsf{eqPly} uses a different color for each
@@
 spheres of the model are tested against these planes to determine the
-visibility with the frustum. Furthermore the render state from the
-window, and the database range from the channel is obtained. The render
+visibility with the frustum. Furthermore, the render state from the
+window and the database range from the channel are obtained. The render
 state manages display list or VBO allocation:
@@
 
+\begin{wrapfigure}{r}{.6\textwidth}
+  \vspace{-3ex}\includegraphics[width=.6\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Rendering Loop}}\vspace{-3ex}
+\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 \textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
-node of the tree gets a database range assigned, that is, the root node
+node of the tree gets a database range assigned. The root node
 has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Rendering Loop}}
-\end{wrapfigure}
-The actual rendering uses display lists or vertex buffer objects. The
+The actual rendering uses display lists or vertex buffer objects. These
 OpenGL objects are allocated using the object manager. The rendering is
@@
 indices in a \textsf{short} value for optimal performance with VBO's.
-
 The leaf nodes reuse the objects stored in the object manager, or create
@@
 
-This section discusses some additional, important features not covered
+This section discusses some additional important features not covered
 by the previous \textsf{eqPly} section. Where possible, code examples
@@
 event framework, while allowing all necessary customizations. It gathers
-all events in the application main thread, so that the developer only
+all events in the main thread of the application, so that the developer only
 has to implement \textsf{Config::processEvent} to update its data based
@@
 the X11/glX window system. WGL receives and processes the events from
-the pipe threads which created the windows. AGL receives the events from
+the pipe threads that created the windows. AGL receives the events from
 application or node main thread. Whenever the term \textbf{event thread}
-is used, it refers to the thread receiving the event, i.e., a per-node
+is used, it refers to the thread receiving the event, such as a per-node
 thread for glX, the pipe thread for WGL and the main thread for AGL.
@@
 \subsubsection{Initialization and Exit}
-
-During window and pipe initialization the event handling is set up. For
-both entities, \textsf{initEvent\-Handler} is called to register the
-pipe or window with an event handler. This method may be overwritten to
-use a custom event handler, or to not install an event handler to
-disable event handling. Likewise, \textsf{exitEventHandler} is called to
-de-initialize event handling.
 
@@
 \end{wrapfigure}
+During window and pipe initialization the event handling is set up. For
+both entities, \textsf{initEvent\-Handler} is called to register the
+pipe or window with an event handler. This method may be overwritten to
+use a custom event handler, or to not install an event handler at all and so
+disable event handling. \textsf{exitEventHandler} is called to
+de-initialize event handling.
+
 An event handler consists of two parts: the generic base class providing
@@
 
-For the WGL and AGL window system it is required to manually receive and
+For the WGL and AGL window systems, it is required to manually receive and
 dispatch (`pump') events. On WGL, this has to happen on each thread with
@@
 sure the events are pumped externally, as it often done within external
-widget sets, e.g., Trolltech's Qt.
+widget sets such as Trolltech Qt.
 
@@
 \textsf{eq::Window} associated to the event. It then creates a generic
-\textsf{WindowEvent}, which holds important event data in a format
-independent of the window system. The original event is attached to the
+\textsf{WindowEvent}, which holds important event data in a 
+window system-independent format. The original event is attached to the
 generic window event.
@@
 member has to be set to the full size of the event to be transmitted.
-Each event has a type, by which the config processing function can
-identify it.
+Each event has a type which is used to identify it by the config 
+processing function.
 
@@
 
-The motivation for direct send is to parallelize the costly
-recomposition for database (sort-last) decomposition. With each
-additional source channel, the amount of pixel data to be composited
-grows linearly. When using the naive approach of compositing all frames
-on the destination channel, this channel quickly becomes the bottleneck
-in the system. Direct send distributes this workload evenly across all
-source channels, and thereby keeps the compositing work per channel
-constant.
-
 \begin{wrapfigure}{r}{.6\textwidth}
@@
 \end{wrapfigure}
+The main idea for direct send is to parallelize the costly recomposition
+for database (sort-last) decomposition. With each additional source
+channel, the amount of pixel data to be composited grows linearly. When
+using the naive approach of compositing all frames on the destination
+channel, this channel quickly becomes the bottleneck in the
+system. Direct send distributes this workload evenly across all source
+channels, and thereby keeps the compositing work per channel constant.
+
 In direct send compositing, each rendering channel is also responsible
-for the sort-last composition of one screen-space tile. He receives the
-framebuffer pixels for his tile from all the other channels. The size of
-one tile decreases linearly with the number of source channels, which
-keeps the total amount of pixel data per channel constant.
+for the sort-last composition\footnote{depth-based for polygonal data or
+  sorted-blend for volume data} of one screen-space tile. He receives
+the framebuffer pixels for his tile from all the other channels. The
+size of one tile decreases linearly with the number of source channels,
+which keeps the total amount of pixel data per channel constant.
 
@@
 
-The image pixel viewport signifies the region of interest read
-back. Currently this pvp has no offset and is as big as the frame data
-pvp. Only one image will be read back.
+The image pixel viewport signifies the region of interest that will be read
+back. The default readback operation reads back one image using the
+frame data's pvp.
 
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/slices.pdf}
-  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
-      Volume Rendering}}
-\end{wrapfigure}
 The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
@@
 
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/slices.pdf}
+  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
+      Volume Rendering}}\vspace{-3ex}
+\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
@@
 \end{figure}
+\begin{wrapfigure}{r}{.4\textwidth}
+  \includegraphics[width=.4\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended together to
-form a seamless picture. For currect blending, the frames are ordered in
-the same back-to-front order as the slices used for rendering using the
+form a seamless picture. For correct blending, the frames are ordered in
+the same back-to-front order as the slices used for rendering, and use the
 same blending parameters. Simplified, the frame buffer images are
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 Finding the correct assembly order for a perspective frustra is more
@@
 in \fig{fBlendPerspective}. The result image naturally looks the same as
-the volume rendering would when rendered n a single channel, as shown in
+the volume rendering would when rendered on a single channel, as shown in
 \fig{fVolResult}.
 
-The assembly algorith described in this section also works with parallel
-compositing algorithms, such as direct-send.
+The assembly algorithm described in this section also works with parallel
+compositing algorithms such as direct-send.
 
@@
 tracking, in order to show how head tracking can be integrated with
-Equalizer. Supporting a wide range of tracking devices is not within the
+Equalizer. Support for a wide range of tracking devices is not within the
 scope of Equalizer. Other open source and commercial implementations

### 2007-10-21 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 \title{Equalizer Programming Guide\\
-  {\footnotesize\sffamily\mdseries
+  {\vspace{-1ex}\footnotesize\sffamily\mdseries
     \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
@@
 }
-\author{Stefan Eilemann\\{\footnotesize Eyescale Software GmbH}
-}
+\author{Eyescale Software GmbH}
+
 \date{
@@
   \textbf{DRAFT}\\[\medskipamount]
-  Version 0.9.2 for Equalizer v0.4\\[\medskipamount]
+  Version 0.9.3 for Equalizer 0.4\\[\medskipamount]
   \today
@@
 \clearpage
+
+\thispagestyle{empty}
+\section*{Equalizer 0.4 Programming Guide}
+\subsection*{Contributors}
+
+Written by Stefan Eilemann.\\
+Engineering contributions by Maxim Makhinya and Jonas B\"osch.
+
+\subsection*{Copyright}
+
+\textcopyright 2007 Eyescale Software GmbH. All rights reserved. No
+permission is granted to copy, distribute, or create derivative works
+from the contents of this electronic documentation in any manner, in
+whole or in part, without the prior written permission of Eyescale
+Software GmbH.
+
+\subsection*{Trademarks and Attributions}
+
+OpenGL is a registered trademark, OpenGL Multipipe is a trademark of
+Silicon Graphics, Inc. Linux is a registered trademark of Linus
+Torvalds. CAVELib is a registered trademark of the University of
+Illinois. The CAVE is a registered trademark of the Board of Trustees of
+the University of Illinois at Chicago. All other trademarks and
+copyrights herein are the property of their respective owners.
+
+\subsection*{Feedback}
+
+If you have comments about the content, accuracy or comprehensibility of
+this programming guide, please contact
+\htmladdnormallink{eile@equalizergraphics.com}
+{mailto:eile@equalizergraphics.com?subject=Equalizer\%20Programming\%20Guide}.
+
+\clearpage
+\thispagestyle{empty}
 \tableofcontents
-\thispagestyle{empty}
 \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
@@
     \hline
+    0.9.3   & Oct 21, 2007 & update title page, add front matter\\
     0.9.2   & Oct 19, 2007 & Jonas' feedback, changes related to
@@
 
-Equalizer provides a framework for the development of parallel
-OpenGL\texttrademark\ applications. Equalizer-based applications can run
-a single shared-memory system with multiple graphics cards (GPU's) or on
-graphics clusters. This Programming Guide introduces the programming
-interface using the \textsf{eqPly} example shipped with Equalizer as a
-guideline.
+Equalizer provides a framework for the development of parallel OpenGL
+applications. Equalizer-based applications can run a single
+shared-memory system with multiple graphics cards (GPU's) or on graphics
+clusters. This Programming Guide introduces the programming interface
+using the \textsf{eqPly} example shipped with Equalizer as a guideline.
 
@@
 % rephrase ( aber ich weiss auch nicht wie :) )
-solutions, such as SGI's OpenGL Multipipe\texttrademark\ SDK, VRCO's
-Cavelib\texttrademark\ and VRJuggler, implement a subset of concepts
-similar to Equalizer. In other areas, e.g., tracking device support,
-they provide more functionality.
+solutions, such as OpenGL Multipipe SDK, Cavelib and VRJuggler,
+implement a subset of concepts similar to Equalizer. In other areas,
+e.g., tracking device support, they provide more functionality.
 
@@
 \fig{fConfig} shows one example configuration for a four-side
-CAVE\texttrademark, running on two machines (nodes) using three graphics
+CAVE, running on two machines (nodes) using three graphics
 cards (pipes) with one window each to render to the four output channels
@@
 {
-    os << _frameDataID << _windowSystem << _useVBOs << _useShaders << _filename;
+    os << _frameDataID << _windowSystem << _precision << _brightness << _useGLSL
+       << _filename;
 }
@@
 {
-    is >> _frameDataID >> _windowSystem >> _useVBOs >> _useShaders >> _filename;
+    is >> _frameDataID >> _windowSystem >> _precision >> _brightness >> _useGLSL
+       >> _filename;
 

### 2007-10-26 Developer/ProgrammingGuide/paper.tex
Update Programming Guide, Release Notes and add release annoucement.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 \date{
-  \vspace{4cm}
+  \vspace{2cm}
   \begin{figure}[ht]
@@
   \vspace{1cm}
-  \textbf{DRAFT}\\[\medskipamount]
-  Version 0.9.3 for Equalizer 0.4\\[\medskipamount]
+  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
+  Version 0.9.4 for Equalizer 0.4\\[\medskipamount]
   \today
@@
 Silicon Graphics, Inc. Linux is a registered trademark of Linus
-Torvalds. CAVELib is a registered trademark of the University of
-Illinois. The CAVE is a registered trademark of the Board of Trustees of
-the University of Illinois at Chicago. All other trademarks and
-copyrights herein are the property of their respective owners.
+Torvalds.  Mac OS is a trademark of Apple Inc. CAVELib is a registered
+trademark of the University of Illinois. The CAVE is a registered
+trademark of the Board of Trustees of the University of Illinois at
+Chicago. Qt is a registered trademark of Trolltech. All other
+trademarks and copyrights herein are the property of their respective
+owners.
 
@@
 {mailto:eile@equalizergraphics.com?subject=Equalizer\%20Programming\%20Guide}.
+
+\vfill
+
+\subsection*{Previous Page}
+
+The images on the front page show: a terrain rendering application on a
+six-node display wall [top left], the \textsf{eqPly} polygonal renderer
+in a three-sided CAVE [middle], two volume rendering results from
+eVolve\footnote{Data sets courtesy of General Electric, USA and AVS,
+  USA} [bottom left and middle] and a six-node sort-last database
+decomposition with parallel direct-send recomposition\footnote{Data set
+  courtesy of Stanford University Computer Graphics Laboratory} [bottom
+right].
 
@@
     \hline
+    0.9.4   & Oct 26, 2007 & release candidate 1\\
     0.9.3   & Oct 21, 2007 & update title page, add front matter\\
@@
 Equalizer provides a framework for the development of parallel OpenGL
-applications. Equalizer-based applications can run a single
-shared-memory system with multiple graphics cards (GPU's) or on graphics
-clusters. This Programming Guide introduces the programming interface
-using the \textsf{eqPly} example shipped with Equalizer as a guideline.
+applications. Equalizer-based applications can run from a single
+shared-memory system with a one or multiple graphics cards up to
+large-scale graphics clusters. This Programming Guide introduces the
+programming interface, often using the Equalizer \textsf{eqPly} example
+as a guideline.
 
 Equalizer is the next step in the evolution of generic parallel programming
-interfaces for visualization applications using OpenGL. Existing
-% rephrase ( aber ich weiss auch nicht wie :) )
+interfaces for OpenGL-based visualization applications. Existing
 solutions, such as OpenGL Multipipe SDK, Cavelib and VRJuggler,
@@
 
-Equalizer implements the minimum necessary layer to build parallel and
-scalable OpenGL applications. The programmer structures the application
-so that the OpenGL rendering can be executed in parallel, potentially
-using multiple processes for cluster-based execution. Equalizer provides
-the domain-specific parallel rendering know-how and abstracts
-configuration, threading, synchronization, windowing and event
-handling. It is a `GLUT on steroids', providing parallel and distributed
-execution, scalable rendering features and fully customizable event
-handling.
-
-If you have any question regarding Equalizer programming, this programming
-guide, or specific problems you encountered, please direct them to the 
- \textsf{eq-dev} mailing list\footnote{see \link{http://www.equalizergraphics.com/lists.html}}.
+In order to adapt an application for Equalizer, the programmer
+structures the source code so that the OpenGL rendering can be executed
+in parallel, potentially using multiple processes for cluster-based
+execution. Equalizer provides the domain-specific parallel rendering
+know-how and abstracts configuration, threading, synchronization,
+windowing and event handling. It is a `GLUT on steroids', providing
+parallel and distributed execution, scalable rendering features and
+fully customizable event handling.
+
+If you have any question regarding Equalizer programming, this
+programming guide, or other specific problems you encountered, please
+direct them to the \textsf{eq-dev} mailing list\footnote{see
+  \link{http://www.equalizergraphics.com/lists.html}}.
 
@@
 distribution\footnote{\link{http://www.equalizergraphics.com/downloads.html}}
-and by compiling the source code. After installing Equalizer, please
-take a look at the Quickstart
+and compiling the source code. After installing Equalizer, please take a
+look at the Quickstart
 Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
@@
 
-The Equalizer architecture is based on a client-server model. Its client
+The Equalizer architecture is based on a client-server model. The client
 library exposes all functionality discussed in this document to the
@@
 and launches the application's rendering clients.  Currently, Equalizer
-currently only supports one application per server, but it will provide
+only supports one application per server, but it will provide
 concurrent and efficient multi-application support in future.
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/processes.pdf}
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/processes.pdf}
   {\caption{\small\label{fProcesses}Equalizer Processes}}
 \end{wrapfigure}
-The render client implements the rendering part of an
-application. Its execution is passive, it has no main loop and
-is completely driven by rendering tasks received from the server. The
+The render client implements the rendering part of an application. Its
+execution is passive, it has no main loop and is completely driven by
+Equalizer, based on the rendering tasks received from the server. The
 tasks are executed by calling the appropriate task methods (see
-\sref{ssTaskMethods}) in the correct thread and context.
-
-The application might contain a rendering client, in which case it can 
+\sref{ssTaskMethods}) in the correct thread and context. The application
+either implements the task methods with application-specific code or
+uses the default methods provided by Equalizer.
+
+The application can also be a rendering client, in which case it can
 also contribute to the rendering. If it does not implement any render
 client-related code, it is reduced to be the application's `master'
-process without any OpenGL windows.
+process without any OpenGL windows and rendering code.
 
 The rendering client can be the same executable as the application, as
-it is the case with \textsf{eqPly}. When it is started as a render client,
-the Equalizer initialization routine does not return and takes over the
-control by calling the render client task methods. Complex real-world
-applications often implement a separate, light-weight rendering client.
+it is the case with all provided examples. When it is started as a
+render client, the Equalizer initialization routine does not return and
+takes over the control by calling the render client task
+methods. Complex applications usually implement a separate, light-weight
+rendering client.
 
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/eqHello.png}
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eqHello.png}
   {\caption{\small\label{fHello}Hello, World!}}
@@
 
-The \textsf{eqHello} example is a minimal example to illustrate the
+The \textsf{eqHello} example is a minimal application to illustrate the
 basic principle of an Equalizer application: The application developer
@@
 interfere in any other way with the application's rendering code. The
-restructuring work required for Equalizer is needed anyway in order to
-parallelize the application for rendering.
+restructuring work required for Equalizer is the minimal refactoring
+needed to parallelize the application for rendering.
 
@@
 The application inherits from Equalizer classes and overrides virtual
-functions to implement certain functionality, such as the application's
+functions to implement certain functionality, e.g., the application's
 OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
@@
 
-\if 0
-\subsection{Thread Safety}
+\subsection{Execution Modeal and Thread Safety}
 
@@
 are executed from this thread. This threading model is the OpenGL
-`threading model', which maintains one current context for each
-thread. It is difficult, if not impossible, to render from different
-threads to the same context.
-
-The rendering threads concurrently render the application's data
-base. The data base should be accessed in a read-only fashion during
-rendering, as it is the case with all modern scene graphs.
-
-The rendering threads run asynchronously to the application's main
-thread. Depending on the configuration's latency, the fall $n$ frames
-behind the last frame finished by the application thread.
-
-TBD support and explain per-node frame sync.
-\fi
-
-\subsection{\label{sResource}The Resource Tree}
+`threading model', which maintains a current context for each
+thread. If structured correctly, the application rarely has to take care
+of thread synchonization or protection of shared data.
+
+The main thread is responsible for maintaining the application logic. It
+reacts on user events, updates the data model and requests new frames to
+be rendered. It drives the whole application, as shown in \fig{model}
+
+The rendering threads concurrently render the application's
+database. The data\-base should be accessed in a read-only fashion
+during rendering to avoid threading problems. This is normally the case,
+for example all modern scene graphs use read-only render traversals.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/model.pdf}
+  {\caption{\small\label{fModel}Simplified execution model}}
+\end{wrapfigure}
+All rendering threads in the configuration run asynchronously to the
+application's main thread. Depending on the configuration's latency,
+they can fall $n$ frames behind the last frame finished by the
+application thread. A latency of one frame is usually not perceived by
+the user, but can increase rendering performance substantially.
+
+Rendering threads on a single node are by default synchronized. When a
+frame is finished, all local rendering threads are done
+drawing. Therefore the application can safely modify the data between
+the end of a frame and the beginning of a new frame. Furthermore, only
+one instance of the application data has to be maintained within a
+process since all rendering threads are quaranteed to draw the same frame.
+
+This per-node frame synchronization does not inhibit latency across
+rendering nodes. Furthermore, advanced rendering software which
+multi-buffers the dynamic parts of the database can disable the per-node
+frame synchronization, as explained in \sref{sNodeFrame}. Some scene
+graphs do implement multi-buffered data.
+
+\subsection{Config}
+
+The \textsf{eq::Config} represents the current configuration of the
+application. The configuration is the session in which all render
+clients are registered. A configuration consists of the description of
+the rendering resources and the usage description for these resources.
 
@@
 
+The resource usage is configured using a compound tree, which is a
+hierarchical representation of the rendering decomposition and
+recomposition across the resources. It is explained in \sref{compounds}.
+
 \begin{figure}[ht!]\center
-  \includegraphics[width=.9\textwidth]{images/cave.pdf}
+  \includegraphics[width=\textwidth]{images/cave.pdf}
   {\caption{\small\label{fConfig}An example configuration}}
@@
 
-\subsubsection{Configuration}
-
-%FIXME confusing
-The root of the resource tree is the \textsf{eq::Config}, which
-represents the current configuration of the application. The
-configuration is the session in which all render clients are
-registered. The config instance on a given node currently only holds the
-local node, not all nodes of the configuration.
-
 \subsubsection{Node}
 
-An \textsf{eq::Node} is the representation of a single computer in the
-system. One operating system process of the render client will be used
+An \textsf{eq::Node} is the representation of a single computer in a
+cluster. One operating system process of the render client will be used
 for each node. Each configuration might also use an application node, in
@@
 thread-unsafe libraries, but have various performance caveats. All pipe,
-window and channel task methods are executed from the pipe's thread or
-in the case of non-threaded pipes, from the main application
+window and channel task methods are executed from the pipe's thread, or
+in the case of non-threaded pipes from the main application
 thread\footnote{see
@@
   created by the application and used with Equalizer} PBuffer or
-FBO(Framebuffer object). The window holds
+framebuffer object (FBO). The window holds
 window-system-specific handles to the drawable and context, e.g. an X11
@@
 
-\subsection{Compounds}
+\subsection{\label{compounds}Compounds}
 
 Usage of the rendering resources is configured using a compound
-tree. Although the API does not currently expose compounds, the
-basic design behind compounds is explained here for a better 
-understanding of the Equalizer architecture.
-
+tree. Although the API does not currently expose compounds, the basic
+design behind compounds is explained here for a better understanding of
+the Equalizer architecture. Further information on the configuration of
+compounds can be found on the Equalizer website\footnote{see
+  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.
+
+\subsubsection{Compound Channels}
 Each compound has a channel, which it uses to execute the rendering
 tasks. One channel might be used by multiple compounds. Unused channels
-are not instantiated. The rendering tasks for the channels are computed
-by the server and send to the appropriate render clients.
-
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/frustra.pdf}
-  {\caption{\small\label{fFrustra}Frustrum descriptions}}
+are not instantiated during initialization. The rendering tasks for the
+channels are computed by the server and send to the appropriate render
+clients.
+
+\subsubsection{Frustum}
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frustra.pdf}
+  {\caption{\small\label{fFrustra}Wall and projection parameters}}
 \end{wrapfigure}
+Compounds have a frustum description to define the physical layout of
+the display environment.
+
+The frustum description is inherited by the children, therefore the
+frustum is typically defined on the top-most compound.
+
+The frustrum can be specified as a wall or projection description.
+
+A wall is completely defined by the bottom-left, bottom-right and
+top-left coordinates relative to the origin.
+
+The projection is defined by the position and head-pitch-roll
+orientation of the projector, as well as the horizontal and vertical
+field-of-view and distance of the projection wall.
+
+\fig{fFrustra} illustrates the wall and projection frustum parameters.
+
+\subsubsection{Compound Classification}
 The channels of the leaf compounds in the compound tree are designated
-as source channels, and the channel of the top-most compound is the
-destination channel. Source channels execute the rendering for the
-destination channel. The top-most compound with a channel defines the 2D
-pixel viewport rendered by all leaf compounds. The destination channel
-and pixel viewport can not be overridden.
-
-Compounds have a frustum description, which is inherited to the
-children. The frustrum can be specified as a wall, which is completely
-defined by the bottom-left, bottom-right and top-left coordinates
-relative to the origin, or as a projection, which is defined by the
-position and head-pitch-roll orientation of the projector, as well as
-the horizontal and vertical field-of-view and distance of the projection
-wall. \fig{fFrustra} illustrates the wall and projection frustum
-parameters.
-
+as source channels. The top-most channel in the tree is the destination
+channel. Only source channels execute rendering task. All channels in a
+compound tree work for the destination channel. The destination channel
+defines the 2D pixel viewport rendered by all leaf compounds. The
+destination channel and pixel viewport can not be overridden by child
+compounds.
+
+\subsubsection{Decomposition - Attributes}
 Compounds have attributes which configure the decomposition of the
@@
 screen space. A \textsf{range} tells the application to render a part of
-its database, and an eye pass can selectively render different stereo
-passes.
-
-Compounds have output and input frames which configure the recomposition
-of the resulting pixels from the source compound channels. An output
-frame connects to an input frame of the same name, which transports the
-selected frame buffer data from the output channel to the input
-channel. This composition process is extensively described in
+its database, and an \textsf{eye} pass can selectively render different
+stereo passes. Setting one or multiple attributes causes the parent's
+view to be decomposed accordingly. Attributes are cumulative, that is,
+intermediate compound attributes affect and therefore decompose the
+rendering of all their children.
+
+\subsubsection{Recomposition - Frames}
+Compounds use output and input frames to configure the recomposition of
+the resulting pixel data from the source channels. An output frame
+connects to an input frame of the same name. The selected frame buffer
+data is transported from the output channel to the input channel. The
+assembly routine of the input channel will block on the availability of
+the output frame. This composition process is extensively described in
 \sref{sCompositing}.
 
-More information on the configuration of compounds can be found on the
-Equalizer website\footnote{see
-  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.
-
 \section{\label{sEqPly}The eqPly polygonal renderer}
+
+In this section the source code of \textsf{eqPly} is discussed in
+detail, and relevant design decision, caveats and other remarks are
+raised.
 
@@
 and serves as a reference implementation of an Equalizer-based
-application of medium complexity. Its focus lies not on rendering features
-or visual quality, but on providing a working show case for the integration
-of most of the Equalizer features.
-
-In this section the source code of \textsf{eqPly} is discussed in
-detail, and relevant design decision, caveat and remarks are raised.
-
-All classes in the example are in the \textsf{eqPly} namespace to avoid
-type name ambiguities, in particular for the \textsf{Window} class which
-is frequently used as a type in the global namespace by windowing
-systems. \fig{fUml} shows the most important Equalizer classes and how
-they are subclassed by the \textsf{eqPly} example.
-
-The derived classes fall into two categories: The subclassed rendering
-entities as discussed in \sref{sResource}, and two classes for
-distributing data. For each of the rendering entities, its function and
-typical usage will be discussed. The distributed data classes illustrate
-the typical usage of static and dynamic, frame-specific data.
+application of medium complexity. It focuses on the example usage of
+core Equalizer features, not on rendering features or visual quality.
 
@@
 
+All classes in the example are in the \textsf{eqPly} namespace to avoid
+type name ambiguities, in particular for the \textsf{Window} class which
+is frequently used as a type in the global namespace by windowing
+systems. \fig{fUml} shows how the most important Equalizer classes are
+used through inheritance by the \textsf{eqPly} example.
+
+The \textsf{eqPly} classes fall into two categories: Subclasses of the
+rendering entities introduced in \sref{sResource}, and two other classes
+for distributing data. The function and typical usage for each of the
+rendering entities is discussed in this section.
+
+The distributed data classes are helper classes based on
+\textsf{eqNet::Object}. They illustrate the typical usage of distributed
+objects for static as well as dynamic, frame-specific data.
+
 \subsection{The main Function}
@@
 The main function starts off with parsing the command line into the
-\textsf{LocalInitData} data structure, part of which will be distributed
-to all render client nodes. The actual command line parsing is done by
-the \textsf{LocalInitData} class and will be discussed in
-\sref{sInitData}:
+\textsf{LocalInitData} data structure. A part of it, the base class
+\textsf{InitData}, will be distributed to all render client nodes. The
+command line parsing is done by the \textsf{LocalInitData} class, which
+is discussed in \sref{sInitData}:
 
@@
 
-The node factory is used by Equalizer to create the objects or instances for
-the rendering entities. Each of the classes inherits from the same type
-provided by Equalizer in the \textsf{eq} namespace. The provided
-\textsf{eq::NodeFactory} base class instantiates a 'plain' Equalizer
-object, thus making it possible to selectively subclass individual
-entity types, as it is done by \textsf{eqHello}. For each rendering
-resource used in the configuration, one C++ object will be created:
+The node factory is used by Equalizer to create the object instances of
+the configured rendering entities. Each of the classes inherits from the
+same type provided by Equalizer in the \textsf{eq} namespace. The
+provided \textsf{eq::NodeFactory} base class instantiates 'plain'
+Equalizer objects, thus making it possible to selectively subclass
+individual entity types, as it is done by \textsf{eqHello}. For each
+rendering resources used in the configuration, one C++ object will be
+created during initialization:
 
@@
 initialize it locally. The application is an \textsf{eq::Client}, which
-in turn is an \textsf{eqNet::Node}. The underlying network distribution
-in Equalizer is a peer-to-peer network structure of
-\textsf{eqNet::Node}s. The application programmer does not have to be
-aware of the classes in the \textsf{eqNet} namespace, but both the
-\textsf{eq::Client} and the server are \textsf{eqNet::Node}s. The local
-initialization of nodes creates a local listening socket, so that the
-node, and therefore the \textsf{eq::Client} can communicate over the
-network with other nodes, such as the server and the rendering clients.
+in turn is an \textsf{eqNet::Node}. The underlying network layer in
+Equalizer is a peer-to-peer network of \textsf{eqNet::Node}s. The
+application programmer does not usually have to be aware of the classes
+in the \textsf{eqNet} namespace, but both the \textsf{eq::Client} and
+the server are \textsf{eqNet::Node}s.
+
+The local initialization of a node creates a local listening socket, so
+that the node. Therefore the \textsf{eq::Client} can communicate over
+the network with other nodes, such as the server and the rendering
+clients.
 
@@
 
-Finally everything is set up to run the \textsf{eqPly} application:
+Finally everything is set up, and the \textsf{eqPly} application is executed:
 
@@
 
-After finishing, the application and \textsf{eqPly} are
-de-initialized and the \textsf{main} function returns:
+After the application has finished, it is de-initialized and the
+\textsf{main} function returns:
 
@@
 \begin{enumerate}
-  \item \textbf{Application}: The executable started by the user, which
-    is the controlling entity in the rendering session.
+\item \textbf{Application}: The executable started by the user, the
+  controlling entity in the rendering session.
 \item \textbf{Auto-launched render client}: The typical render client,
@@
   is selected using the command-line option \textsf{--eq-client} and
-    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
+  potentially \textsf{--eq-listen \textless address\textgreater} and
+  \textsf{-r}\footnote{see
+    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
 \end{enumerate}
@@
 explicitly specifies a server address. If no server was specified,
-\textsf{Client::connectServer} try first to connect to a server on the
-local machine using the default port 4242. If that fails, it will create
-a server running within the applications process with a default
-1-channel configuration\footnote{see
+\textsf{Client::connectServer} tries first to connect to a server on the
+local machine using the default port. If that fails, it will create
+a server running within the application process with a default
+one-channel configuration\footnote{see
   \link{http://www.equalizergraphics.com/documents/design/standalone.html}}.
@@
 implementations to provide additional hints and information to the
-server for choosing a configuration. The configuration chosen by the
+server for choosing the configuration. The configuration chosen by the
 server is created locally using
@@
 \textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
-making cast necessary:
+making the cast necessary:
 
@@
 When the configuration was successfully initialized, the main rendering
-loop is executed. The main loop runs until either the user exits the
-configuration or a maximum number of frames has been rendered, specified
-as a command-line argument. The latter is useful for benchmarks. The
-\textsf{Clock} is reused for measuring the overall performance. A new
-frame is started using \textsf{Config::startFrame} and a frame is
-finished using \textsf{Config::finishFrame}.
+loop is executed. The main loop runs until the user exits the
+configuration, or when a maximum number of frames has been rendered,
+when specified by a command-line argument. The latter is useful for
+benchmarks. The \textsf{Clock} is reused for measuring the overall
+performance. A new frame is started using \textsf{Config::startFrame}
+and a frame is finished using \textsf{Config::finishFrame}.
 
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
+\textsf{Config::finishFrame} synchronizes on the completion of the frame
+\textsf{current - latency}. The latency is specified in the
+configuration file, and allows several outstanding frames. This allows
+overlapping execution in the node and pipe threads and minimizes idle
+times.
+
+By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
+completion of all local rendering tasks for the current frame. This
+facilitates porting of exising rendering codes, since the database does
+not have to be multi-buffered. Applications such as \textsf{eqPly}, which
+do not need this per-node frame synchronization can disable it, as
+explained in \sref{sNodeFrame}.
+
+\begin{wrapfigure}{r}{.618\textwidth}
   \subfigure[]{\label{fSync}
@@
 \end{wrapfigure}
-\textsf{Config::finishFrame} synchronizes on the completion of the frame
-\textsf{current - latency}. The latency is specified in the
-configuration file, and allows several outstanding frames. This allows
-overlapping execution in the node and pipe threads and minimizes idle
-times. The first \textsf{nLatency Config::finishFrame} return
-immediately, since they have no frame to synchronize
-upon. \fig{fSyncAsync} shows the execution of (hypothetical) rendering
-tasks without latency~(\fig{fSync}) and with a latency of
-one~(\fig{fAsync}).
-
-With \textsf{eqPly}, a speedup of 15\% has been observed on a five-node
-rendering
-cluster when using a latency of one instead of no latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}.
+\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
+without latency (\fig{fSync}) and with a latency of
+one (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
+observed on a five-node rendering cluster when using a latency of one
+instead of no
+latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
+latency of one or two frames is normally not perceived by the user.
 
@@
 Equalizer provides distributed objects which help implementing data
-distribution in a graphics cluster. The master version of a distributed
+distribution in a cluster environment. The master version of a distributed
 object is registered with a \textsf{eqNet::Session}, which assigns a
@@
 instance of the object to this identifier, thus synchronizing the
-object's data with the remotely registered object.
+object's data with the remotely registered master version.
 
@@
 initial data to all rendering nodes, as well as a versioned object to
-provide frame-specific data such as camera position to the
+provide frame-specific data such as the camera position to the
 rendering methods.
@@
 \textsf{InitData} to provide the command line parsing and to set the
-initial values. The render nodes only instantiate the distributed part
-in \textsf{InitData}
+default values. The render nodes only instantiate the distributed part
+in \textsf{InitData}.
 
@@
 \textsf{getInstanceData} and \textsf{applyInstanceData}. The first
-approach can be used if all distributed member variables can be stored
-in one contiguous block of memory. The second approach is used
-otherwise.
+approach can be used if all distributed member variables are stored in
+one contiguous block of memory. The second approach is used otherwise.
 
@@
 length. Therefore it uses the second approach of manually serializing
-and deserializing its data. Serialization in
-\textsf{getInstanceData} and deserialization in
-\textsf{applyInstanceData} is performed by streaming all member variable to or from
-the provided data streams. The data transport between nodes is
-implemented in the data streams and various other Equalizer classes:
+and deserializing its data. Serialization in \textsf{getInstanceData}
+and deserialization in \textsf{applyInstanceData} is performed by
+streaming all member variable to or from the provided data
+streams. Efficient buffering and data transport between nodes is
+implemented in the data streams:
 
@@
 {
-    os << _frameDataID << _windowSystem << _precision << _brightness << _useGLSL
-       << _filename;
+    os << _frameDataID << _windowSystem << _useVBOs << _useGLSL << _filename;
 }
@@
 {
-    is >> _frameDataID >> _windowSystem >> _precision >> _brightness >> _useGLSL
-       >> _filename;
+    is >> _frameDataID >> _windowSystem >> _useVBOs >> _useGLSL >> _filename;
 
@@
 
-Versioned objects have to override \textsf{isStatic} to return false to
-indicate that they are versioned. The current implementation has the
-following characteristics:
+Versioned objects have to override \textsf{isStatic} to return
+\textsf{false}, which indicates that they are versioned. The current
+implementation has the following characteristics:
 \begin{itemize}
 \item Only the master instance of the object is writable, that is,
-  \textsf{eqNet::Object::com\-mit} can be called to generate a new
-  version.
+  \textsf{eqNet::Object::com\-mit} can be called only on the master
+  instance to generate a new version.
 \item Slave instance versions can only be advanced, that is,
@@
   the current version will fail.
+\item Newly mapped slave instance are mapped to the oldest available
+  version.
 \end{itemize}
@@
 
-In addition to the instance data (de)serialization methods needed to map
+In addition to the instance data (de)serialization methods used to map
 an object, versioned objects may implement \textsf{pack} and
 \textsf{unpack} to serialize or deserialize the changes since the last
-version.
+version. These methods can
 
@@
 size and changes frequently. Therefore, the instance and delta data are
-the same and set in the constructor:
+the same and set in the constructor. The default implementation will
+take care of the distribution of the data:
 
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/configInit.pdf}
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
   {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
@@
 to set up a tracking device for head tracking. Equalizer does not
-provide extensive support for tracking device, as this is an orthogonal
-problem to parallel rendering. Tracking device support has been solved
-already by a number of implementations\footnote{VRCO Trackd, VRPN,
-  etc.}, which can easily be integrated with Equalizer. The example code
-in \textsf{eqPly} provides a reference implementation for the integration of
-such a tracking library:
+provide extensive support for tracking devices, as this is an orthogonal
+problem to parallel rendering. Tracking device support has already been
+solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
+which can easily be integrated with Equalizer. The example code in
+\textsf{eqPly} provides a reference implementation for the integration
+of such a tracking library. \sref{sTracking} provides more background on
+head tracking.
 
@@
 The exit function of the configuration stops the render clients by calling
-\textsf{eq::Config::exit}, and then deregisters the initialization and
+\textsf{eq::Con\-fig::exit}, and then deregisters the initialization and
 frame data objects with the session:
@@
 
-\subsubsection{Frame Control}
+\subsubsection{\label{sNodeFrame}Frame Control}
 
@@
 
-The first synchronization ensures that the application (node) thread is
-done modifiying the data, typically during event processing. The pipe
-threads call \textsf{Node::waitFrameStarted} which blocks until the node
-calls \textsf{startFrame} in \textsf{Node::frame\-Start}.
+The first synchronization point ensures that the application (node)
+thread is done modifiying the data, typically during event
+processing. The pipe threads call \textsf{Node::waitFrameStarted} which
+blocks until the node calls \textsf{startFrame} in
+\textsf{Node::frame\-Start}.
 
@@
 threads are done rendering the current frame, and therefore do not
-access shared data. The node's \textsf{frameDrawFinish} waits for all
+access shared data. At this point the application thread can safely
+modify scene data. The node's \textsf{frameDrawFinish} waits for all
 local pipes to release the synchronization by calling
-\textsf{Pipe::releaseFrameLocal}, which happens by default in
+\textsf{Pipe::release\-Fra\-me\-Lo\-cal}, which happens by default in
 \textsf{Pipe::frameDrawFinish}. The \textsf{frameDrawFinish} methods are
@@
 \begin{figure}[ht!]\center
-  \includegraphics[width=.9\textwidth]{images/mainloop.pdf}
+  \includegraphics[width=\textwidth]{images/mainloop.pdf}
   {\caption{\small\label{fFrameSync}Synchronization of frame tasks}}
@@
 \begin{itemize}
-\item releasing the local synchronization in \textsf{Node::frameStart}
-\item not calling \textsf{Node::waitFrameStarted} in
-  \textsf{Pipe::frameStart}
+\item releasing the local synchronization early in \textsf{Node::frameStart}
+\item not waiting for the node to start the frame by not calling
+  \textsf{Node::waitFrameStarted} in \textsf{Pipe::frameStart}
 \item not waiting for the pipe synchronization in \textsf{Node::frameDrawFinish}
@@
 thread. This approach optimizes usage of the GPU, since all
-tasks are executed serially and do not compete for resources and
-context switches are minimized. Later versions of Equalizer might
+tasks are executed serially and therefore do not compete for resources
+or cause OpenGL context switches. Later versions of Equalizer might
 introduce threaded windows to allow the parallel and independent
@@
 \end{lstlisting}}%>>
+
+\subsubsection{Carbon/AGL Thread Safety}
 
@@
 The Equalizer window holds an OpenGL drawable and rendering
-context. Subclassed windows should maintain all data specific to the
-OpenGL context. The Equalizer window creation routines share the OpenGL
-context with the first window of the pipe, thus allowing the reuse of
-OpenGL objects such as display lists and textures between all 
-windows of one pipe.
+context. When using the default window initialization functions, all
+windows of a pipe share the OpenGL context. This allows reuse of OpenGL
+objects such as display lists and textures between all windows of one
+pipe.
+
+The window class is the natural place for the application to maintain
+all data specific to the OpenGL context.
 
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
   {\caption{\small\label{fWindowInit}Window Initialization}}
@@
 The state object in \textsf{eqPly} gathers all rendering state, which
-includes an object manager for object state handling. This state object
-can easily be replaced to use the rendering code in a stand-alone
-application, e.g., for a GLUT-based renderer.
+includes an object manager for OpenGL object allocation.
 
@@
 \begin{description}
-\item[supportsObjects()]: returns true if the usage for this particular
+\item[supportsObjects()] returns true if the usage for this particular
   type of objects is supported. For objects available in OpenGL 1.1 or
   earlier, this function is not implemented.
-\item[getObject( key )]: returns the object associated with the given
+\item[getObject( key )] returns the object associated with the given
   key, or FAILED. Increases the reference count of existing objects.
-\item[newObject( key )]: allocates a new object for the given
+\item[newObject( key )] allocates a new object for the given
   key. Returns FAILED if the object already exists or if the allocation
   failed. Sets the reference count of a newly created object to one.
-\item[obtainObject( key )]: convenience function which gets or obtains
+\item[obtainObject( key )] convenience function which gets or obtains
   the object associated with the given key. Returns FAILED only if the
   object allocation failed.
-\item[releaseObject( key \textbar\ name )]: decreases the reference count and
+\item[releaseObject( key \textbar\ name )] decreases the reference count and
   deletes the object if the reference count reaches zero.
-\item[deleteObject( key \textbar\ name )]: manually deletes the object. To be
+\item[deleteObject( key \textbar\ name )] manually deletes the object. To be
   used if reference counting is not used.
@@
 \begin{description}
-\item[Buffer]: The OpenGL read and draw buffer as well as color mask.
+\item[Buffer] The OpenGL read and draw buffer as well as color mask.
   These parameters are influenced by the current eye pass, eye
-  seperation and anaglyphic stereo settings.
-\item[Viewport]: The two-dimensional pixel viewport restricting the
+  separation and anaglyphic stereo settings.
+\item[Viewport] The two-dimensional pixel viewport restricting the
   rendering area within the channel. For correct operations, both
@@
   definition and compound viewports set for sort-first/2D decompositions.
-\item[Frustum]: Frustum parameters as defined by
+\item[Frustum] The same frustum parameters as defined by
   \textsf{glFrustum}. Typically the frustum used to set up the OpenGL
@@
   current eye pass.
-\item[Head Transformation]: A transformation matrix positioning the
+\item[Head Transformation] A transformation matrix positioning the
   frustum. This is typically an identity matrix and is used for off-axis
@@
   defined.
-\item[Range]: A one-dimensional range with the interval [0..1]. This
+\item[Range] A one-dimensional range with the interval [0..1]. This
   parameter is optional and should be used by the application to render
-  only the appropriate subset of its data.
+  only the appropriate subset of its data. It is influenced by the
+  compound range attribute.
 \end{description}
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/DB.pdf}
+\begin{wrapfigure}{r}{.382\textwidth}
+  \vspace{-8ex}\includegraphics[width=.382\textwidth]{images/DB.png}
   {\caption{\small\label{fDB}Destination view of an DB compound}}
@@
 
-Finally the model loaded by the node is rendered. If the model was not
-loaded during node initialization, a quad is drawn in its place:
+Finally the model, which has been loaded by the node, is rendered. If
+the model was not loaded during node initialization, a quad is drawn in
+its place:
 
@@
 In order to draw the model, a helper class for view frustum culling is
-set up using the view frustum from Equalizer and the camera
-position. The frustum helper computes the six frustum planes from the
-projection and modelView matrices. During rendering, the bounding
-spheres of the model are tested against these planes to determine the
-visibility with the frustum. Furthermore, the render state from the
-window and the database range from the channel are obtained. The render
-state manages display list or VBO allocation:
+set up using the view frustum from Equalizer and the camera position
+from the frame data. The frustum helper computes the six frustum planes
+from the projection and modelView matrices. During rendering, the
+bounding spheres of the model are tested against these planes to
+determine the visibility with the frustum.
+
+Furthermore, the render state from the window and the database range
+from the channel is obtained. The render state manages display list or
+VBO allocation:
 
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \vspace{-3ex}\includegraphics[width=.6\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Rendering Loop}}\vspace{-3ex}
-\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Rendering Loop}}
+\end{wrapfigure}
 The rendering loop maintains a list of candidates to render, which
@@
 customizable in any stage of the processing, to the extreme of making it
-possible to disable all event-related code in Equalizer. More
+possible to disable all event-related code in Equalizer. In this aspect,
+Equalizer substantially differs from GLUT, which imposes an event model
+and hides most of the event handling in \textsf{glutMainLoop}. More
 information on event handling can be found on the Equalizer
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/eventInit.pdf}
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventInit.pdf}
   {\caption{\small\label{fEventInit}Event Handling Initialization}}
@@
 pipe or window with an event handler. This method may be overwritten to
-use a custom event handler, or to not install an event handler at all and so
-disable event handling. \textsf{exitEventHandler} is called to
-de-initialize event handling.
+use a custom event handler, or to not install an event handler at all
+and so disable event handling. Likewise, \textsf{exitEventHandler} is
+called to de-initialize event handling.
 
@@
 
-Pipe event handling is only used for glX, where one \textsf{Display}
-connection is created to subscribe to window events. Event handling is
-initialized whenever a new, window-system-specific pipe or window handle
-is set. First, \textsf{exitEventHandler} is called to de-initialize
-event handling for the old handle (if set), and then
-\textsf{initEvent\-Handler} is called for the new handle. AGL and glX use
-an event handler singleton, whereas WGL uses one event handler per
-window.
+Event handling is initialized whenever a new, window-system-specific
+pipe or window handle is set. First, \textsf{exitEventHandler} is called
+to de-initialize event handling for the old handle (if set), and then
+\textsf{initEvent\-Handler} is called for the new handle.
+
+AGL and glX use an event handler singleton, whereas WGL uses one event
+handler per window. Pipe event handling is only used for glX, where one
+\textsf{Display} connection is created and used to subscribe to window
+events.
 
@@
 sure the events are pumped externally, as it often done within external
-widget sets such as Trolltech Qt.
+widget sets such as Qt.
 
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/eventFilter.pdf}
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
   {\caption{\small\label{fEventProcessing}Event Processing}}
@@
 \textsf{ConfigEvent}. The config events are send to the application
-thread using \textsf{Config::sendEvent}. If the event was processed, the
-function has to return \textsf{true}. If \textsf{false} is returned, the
-event will be passed to a previously installed, window-system-specific
-event handling function. The default implementation of
-\textsf{Window::processEvent} passes most events on to the application.
+thread using \textsf{Config::sendEvent}.
+
+If the event was processed by \textsf{processEvent}, the function has to
+return \textsf{true}. If \textsf{false} is returned, the event will be
+passed to a previously installed, window-system-specific event handling
+function. The default implementation of \textsf{Window::processEvent}
+passes most events on to the application.
 
@@
 processing, that is, it calls \textsf{Config::handleEvent} for each
-queued event. By overriding this function, event-driven execution can be
-implemented.
+queued event. By overriding \textsf{handleEvents}, event-driven
+execution can be implemented.
 
 Later Equalizer versions will introduce \textsf{Pipe::processEvent} and
-\textsf{PipeEvent} to communicate pipe-specific events, e.g, resolution
-changes.
+\textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
+resolution changes.
 
@@
 Two task methods are responsible for collecting and compositing the
-result image during scalable rendering, when multiple channels are
-rendering for a single view. The source channels producing one or more
-\textsf{outputFrame}s use \textsf{Channel::frameReadback} to read the
-pixel data from the frame buffer. The channels receiving one or multiple
-\textsf{inputFrame}s use \textsf{Channel::frameAssemb\-le} to assemble
-the pixel data into the framebuffer. Equalizer takes care of the network
-transport of frame buffer data between nodes, if needed.
+result image during scalable rendering. Scalable rendering is a use case
+of parallel rendering, when multiple channels are contributing to a single
+view.
+
+The source channels producing one or more \textsf{outputFrame}s use
+\textsf{Channel::frame\-Read\-back} to read the pixel data from the frame
+buffer. The channels receiving one or multiple \textsf{inputFrame}s use
+\textsf{Channel::frameAssemb\-le} to assemble the pixel data into the
+framebuffer. Equalizer takes care of the network transport of frame
+buffer data between nodes, if needed.
 
@@
 
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/directSend.pdf}
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
   {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
@@
 frame data is the object connecting output with input frames. Output and
-input frames with the same name and from the same compound tree will
+input frames with the same name within the same compound tree will
 reference the same frame data.
 
-The frame data is a holder for images and is used to signal the
-availability of the pixel data. An \textsf{eq::Image} holds a
+The frame data is a holder for images and additional information, such
+as output frame attributes and pixel data availability.
+
+An \textsf{eq::Image} holds a
 two-dimensional snapshot of the framebuffer and can contain color and/or
@@
 
-The frame data transports the inherited range of its compound. The range
-is needed to compute the assembly order of multiple input frames, e.g.,
-for sorted-blend compositing in volume rendering applications.
+The frame synchronization through the frame data allows the input frame
+to wait for the pixel data to become ready, which is signalled by the
+output frame after readback.
+
+Furthermore, the frame data transports the inherited range of the output
+frame's compound. The range can be used to compute the assembly order of
+multiple input frames, e.g., for sorted-blend compositing in volume
+rendering applications.
 
@@
 \textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
-respective start method of the image.
-
-\begin{wrapfigure}{r}{.6\textwidth}
-  \includegraphics[width=.6\textwidth]{images/assembly.pdf}
+respective start method of the image. Later versions of Equalizer will
+implement asynchronous pixel transfers.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
   {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
@@
 The image pixel viewport signifies the region of interest that will be read
-back. The default readback operation reads back one image using the
-frame data's pvp.
+back. The default readback operation reads back one image using the full
+pixel viewport of the frame data.
 
@@
 
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/slices.pdf}
-  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
-      Volume Rendering}}\vspace{-3ex}
-\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
+      Volume Rendering}}
+\end{wrapfigure}
 The full power of scalable volume rendering is however in DB (sort-last)
 compounds, where the full volume is divided into separate bricks. Each
-of the bricks is rendered the same as in the other modes. For
-recomposition, the \textsf{RGBA} frame buffer data resulting from these
-render passes then has to be assembled correctly.
+of the bricks is rendered like a separate volume. For recomposition, the
+\textsf{RGBA} frame buffer data resulting from these render passes then
+has to be assembled correctly.
+
+Conceptually, the individual volume bricks of each of the source
+channels produces pixel data which can be handled like one big 'slice'
+through the full texture. Therefore they have to be blended
+back-to-front in the same way as the slice planes during rendering.
 
@@
 channel), fill rate (less samples per channel) and IO bandwidth for
-time-dependend data (less data per time step and channel).
+time-dependend data (less data per time step and channel). Since the
+amount of texture memory needed for each node decreases linearly, they
+make it possible to render data sets which are not feasible to
+visualize with any other approach.
 
@@
 \end{figure}
-\begin{wrapfigure}{r}{.4\textwidth}
-  \includegraphics[width=.4\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended together to
@@
 
-Finding the correct assembly order for a perspective frustra is more
+Finding the correct assembly order for perspective frustra is more
 complex. The perspective distortion invalidates a simple orientation
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \vspace{-4ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 In order to compute the assembly order, \textsf{eVolve} uses the angle
 between the origin$\rightarrow$slice vector and the near plane, as shown
-in \fig{fBlendPerspective}. The result image naturally looks the same as
-the volume rendering would when rendered on a single channel, as shown in
-\fig{fVolResult}.
+in \fig{fBlendPerspective}. When the angle becomes bigger than
+90\textdegree, the compositing order of the remaining frames has to be
+changed. The result image of this composition naturally looks the same
+as the volume rendering would when rendered on a single
+channel. \fig{fVolResult} shows the result of the composition from
+\fig{fBlendPerspective}.
 
@@
 
-\subsection{Head Tracking}
+\subsection{\label{sTracking}Head Tracking}
 
@@
   \subfigure[]{\label{fMono}
-    \includegraphics[width=.3\textwidth]{images/mono}}\hfil
+    \includegraphics[height=3cm]{images/mono}}\hfil
   \subfigure[]{\label{fStereo}
-    \includegraphics[width=.3\textwidth]{images/stereo}}\hfil
+    \includegraphics[height=3cm]{images/stereo}}\hfil
   \subfigure[]{\label{fTracked}
-    \includegraphics[width=.3\textwidth]{images/tracked}}%
+    \includegraphics[height=3cm]{images/tracked}}%
   {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked

### 2007-10-28 Developer/ProgrammingGuide/paper.tex
Update Programming Guide and website for 0.4 release
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   \vspace{1cm}
-  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 0.9.4 for Equalizer 0.4\\[\medskipamount]
+%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
+  Version 1.0 for Equalizer 0.4\\[\medskipamount]
   \today
@@
     \hline
-    0.9.4   & Oct 26, 2007 & release candidate 1\\
-    0.9.3   & Oct 21, 2007 & update title page, add front matter\\
-    0.9.2   & Oct 19, 2007 & Jonas' feedback, changes related to
-                             per-node frame sync\\
-    0.9.1   & Oct 8,  2007 & review pass\\
-    0.9     & Oct 7,  2007 & add eqHello, draft assembly in eVolve\\
-    0.8     & Sep 28, 2007 & add head tracking, finish channel,
-                             proof-reading pass\\
-    0.7     & Sep 21, 2007 & start channel and image compositing,
-                             add event handling, move to new eqPly
-                             code base\\
-    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
-    0.5     & Sep 3,  2007 & add config and partly pipe\\
-    0.4     & Aug 31, 2007 & add distributed objects\\
-    0.3     & Aug 26, 2007 & add application and render client\\
-    0.2     & Aug 20, 2007 & add main function\\
-    0.1     & Aug 19, 2007 & outline the basic concepts\\
+    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
     \hline
@@
 applications. Equalizer-based applications can run from a single
-shared-memory system with a one or multiple graphics cards up to
+shared-memory system with one or multiple graphics cards up to
 large-scale graphics clusters. This Programming Guide introduces the
@@
 reacts on user events, updates the data model and requests new frames to
-be rendered. It drives the whole application, as shown in \fig{model}
+be rendered. It drives the whole application, as shown in \fig{fModel}.
 
@@
 one instance of the application data has to be maintained within a
-process since all rendering threads are quaranteed to draw the same frame.
+process since all rendering threads are guaranteed to draw the same frame.
 
@@
 
-\subsection{Config}
+\subsection{\label{sConfig}Config}
 
@@
 
-\fig{fConfig} shows one example configuration for a four-side
+\fig{fConfig} shows an example configuration for a four-side
 CAVE, running on two machines (nodes) using three graphics
@@
 
-An \textsf{eq::Node} is the representation of a single computer in a
-cluster. One operating system process of the render client will be used
-for each node. Each configuration might also use an application node, in
-which case the application process is also used for rendering. All
-node-specific task methods are executed from the main application
-thread.
+The \textsf{eq::Node} class is the representation of a single computer
+in a cluster. One operating system process of the render client will be
+used for each node. Each configuration might also use an application
+node, in which case the application process is also used for
+rendering. All node-specific task methods are executed from the main
+application thread.
 
@@
 
-The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU).  In
-the current implementation it is also one operating system
+The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU).
+In the current implementation it is also one operating system
 thread. Non-threaded pipes are supported for integrating with
@@
 
-An \textsf{eq::Window} holds a drawable and an OpenGL context. The
-drawable can be an on-screen window or an off-screen\footnote{Off-screen drawables are not implemented yet, but can be
-  created by the application and used with Equalizer} PBuffer or
-framebuffer object (FBO). The window holds
-window-system-specific handles to the drawable and context, e.g. an X11
-window \textsf{XID} and \textsf{GLXContext} for the glX window system.
+The \textsf{eq::Window} class holds a drawable and an OpenGL
+context. The drawable can be an on-screen window or an
+off-screen\footnote{Off-screen drawables are not implemented yet, but
+  can be created by the application and used with Equalizer} PBuffer or
+framebuffer object (FBO). The window holds window-system-specific
+handles to the drawable and context, e.g., an X11 window \textsf{XID} and
+\textsf{GLXContext} for the glX window system.
 
@@
 
-The \textsf{eq::Channel} is the abstraction of an OpenGL viewport within
-its parent window. It is the entity executing the actual rendering. The
-channel's viewport might be overwritten when it is rendering for another
-channel during scalable rendering.
+The \textsf{eq::Channel} class is the abstraction of an OpenGL viewport
+within its parent window. It is the entity executing the actual
+rendering. The channel's viewport might be overwritten when it is
+rendering for another channel during scalable rendering.
 
@@
 \subsubsection{Compound Channels}
-Each compound has a channel, which it uses to execute the rendering
-tasks. One channel might be used by multiple compounds. Unused channels
-are not instantiated during initialization. The rendering tasks for the
-channels are computed by the server and send to the appropriate render
-clients.
+Each compound has a channel, which is used by the compound to execute
+the rendering tasks. One channel might be used by multiple
+compounds. Unused channels are not instantiated during
+initialization. The rendering tasks for the channels are computed by the
+server and send to the appropriate render clients.
 
@@
 screen space. A \textsf{range} tells the application to render a part of
-its database, and an \textsf{eye} pass can selectively render different
-stereo passes. Setting one or multiple attributes causes the parent's
-view to be decomposed accordingly. Attributes are cumulative, that is,
-intermediate compound attributes affect and therefore decompose the
-rendering of all their children.
+its database, and an \textsf{eye} rendering pass can selectively render
+different stereo passes. Setting one or multiple attributes causes the
+parent's view to be decomposed accordingly. Attributes are cumulative,
+that is, intermediate compound attributes affect and therefore decompose
+the rendering of all their children.
 
@@
 
-In this section the source code of \textsf{eqPly} is discussed in
-detail, and relevant design decision, caveats and other remarks are
+In this section the source code of \textsf{eqPly} is explained in
+detail, and relevant design decisions, caveats and other remarks are
 raised.
@@
 The \textsf{eqPly} classes fall into two categories: Subclasses of the
-rendering entities introduced in \sref{sResource}, and two other classes
+rendering entities introduced in \sref{sConfig}, and two other classes
 for distributing data. The function and typical usage for each of the
@@
 
-The local initialization of a node creates a local listening socket, so
-that the node. Therefore the \textsf{eq::Client} can communicate over
-the network with other nodes, such as the server and the rendering
-clients.
+The local initialization of a node creates a local listening socket,
+which allows the \textsf{eq::Client} to communicate over the network
+with other nodes, such as the server and the rendering clients.
 
@@
   return. During exit, the server terminates the process. By default,
-  the server starts render client using \textsf{ssh}.
+  the server starts the render client using \textsf{ssh}.
 \item \textbf{Resident render client}: Manually pre-started render
@@
 without latency (\fig{fSync}) and with a latency of
-one (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
-observed on a five-node rendering cluster when using a latency of one
+one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
+observed on a five-node rendering cluster when using a latency of one frame
 instead of no
@@
 
-When the main loop is finished, \textsf{Config::finishAllFrames} catches
-up with the latency. It returns after all outstanding frames have been
-rendered, and is needed to provide an accurate measurement of the
-framerate:
+When the main rendering loop has finished,
+\textsf{Config::finishAllFrames} is called to catch up with the
+latency. It returns after all outstanding frames have been rendered, and
+is needed to provide an accurate measurement of the framerate:
 
@@
 length. Therefore it uses the second approach of manually serializing
-and deserializing its data. Serialization in \textsf{getInstanceData}
-and deserialization in \textsf{applyInstanceData} is performed by
+and de-serializing its data. Serialization in \textsf{getInstanceData}
+and de-serialization in \textsf{applyInstanceData} is performed by
 streaming all member variable to or from the provided data
@@
 It is the application's responsibility to exactly match the order and
-types of variables during serialization and deserialization.
+types of variables during serialization and de-serialization.
 
@@
 an object, versioned objects may implement \textsf{pack} and
-\textsf{unpack} to serialize or deserialize the changes since the last
-version. These methods can
+\textsf{unpack} to serialize or de-serialize the changes since the last
+version.
 
@@
 initialization tasks are transmitted by the server at once and their
-execution is synchronized later.
+completion is synchronized afterwards.
 
@@
 the initialization data is transmitted to the render client nodes using
-the \textsf{initID} parameter of \textsf{eq::Config::init}. Equalizer
-will pass this identifier to all \textsf{configInit} calls of the
-respective objects:
+the \textsf{initID} parameter of \textsf{eq::Config::init}.
+
+Equalizer will pass this identifier to all \textsf{configInit} calls of
+the respective objects:
 
@@
 The exit function of the configuration stops the render clients by calling
-\textsf{eq::Con\-fig::exit}, and then deregisters the initialization and
+\textsf{eq::Con\-fig::exit}, and then de-registers the initialization and
 frame data objects with the session:
@@
 If a tracker is used, the current head position and orientation is
-retrieved and given to Equalizer, which uses the head matrix together
+retrieved and passed to Equalizer, which uses the head matrix together
 with the wall or projection description to compute the view
@@
 frame \textsf{current - latency} has been released by all nodes. The
-window and channel synchronization is similar and omitted for
+window and channel synchronization are similar and omitted for
 simplicity.
@@
 Pipe threads are not explicitely synchronized with each other, that is,
-pipes might be rendering different frames at one given time. Therefore
+pipes might be rendering different frames at any given time. Therefore
 frame-specific data has to be allocated for each pipe thread, which in
@@
 provided by the initialization data. The initialization in
-\textsf{eq::Pipe} does GPU-specific initialization, which is
+\textsf{eq::Pipe} does the GPU-specific initialization, which is
 window-system-dependent. On AGL the display ID is determined, and on glX
-the display connection, to be used by the child windows, is opened.
+the display connection is opened.
 
@@
 \textsf{selectWindowSystem} loops over all window systems and returns
-the first supported window system, determined using
+the first supported window system, determined by using
 \textsf{supportsWindowSystem}.
@@
 command line option. Therefore the implementation of
-\textsf{selectWindowSystem} is overwritten, and returns the specified
-window system, if it is supported:
+\textsf{selectWindowSystem} is overwritten and returns the specified
+window system, if supported:
 
@@
 
-The Equalizer window holds an OpenGL drawable and rendering
+The Equalizer window holds an OpenGL drawable and a rendering
 context. When using the default window initialization functions, all
@@
 create an on-screen window using OS-specific methods. If the OpenGL
-drawable and context was created successfully, \textsf{configInit} calls
-\textsf{configInitGL}, which will perform the generic OpenGL
+drawable and context were created successfully, \textsf{configInit} calls
+\textsf{configInitGL}, which performs the generic OpenGL
 state initialization. The default implementation sets up some typical
@@
 
-The window exit function deallocates all OpenGL objects when the state object is
-about to be disposed. The object manager does not delete the object in
-its destructor, since it does not know if an OpenGL context is still
-current. Additionally, \textsf{eq::Window::configExit} is called to
-destroy the drawable and context:
+The window exit function de-allocates all OpenGL objects when the state
+object is about to be disposed. The object manager does not delete the
+object in its destructor, since it does not know if an OpenGL context is
+still current. Additionally, \textsf{eq::Window::configExit} is called
+to destroy the drawable and context:
 
@@
 The object manager is not strictly a part of the window. It
-is discussed here since the \textsf{eqPly} window uses an object manager.
+is mentioned here since the \textsf{eqPly} window uses an object manager.
 
@@
 
-The OM is a template class. The template type is the key which is used to
+The OM is a template class. The template type is the key used to
 identify objects. The same key is used by all contexts to get the
@@
 This section discusses some additional important features not covered
-by the previous \textsf{eqPly} section. Where possible, code examples
+by the previous \textsf{eqPly} section. Where possible code examples
 from the Equalizer distribution are used.
@@
 events are to be handled. Therefore, event handling in Equalizer is
-customizable in any stage of the processing, to the extreme of making it
+customizable at any stage of the processing, to the extreme of making it
 possible to disable all event-related code in Equalizer. In this aspect,
@@
 
-Where possible, events are received and processed by a separate per-node
+Where possible events are received and processed by a separate per-node
 event thread to allow asynchronous\footnote{with respect to the
@@
 
-Event handling is initialized whenever a new, window-system-specific
+Event handling is initialized whenever a new window-system-specific
 pipe or window handle is set. First, \textsf{exitEventHandler} is called
@@
 
-The \textsf{Config::sendEvent} method transmits a
+The \textsf{Config::sendEvent} method transmits an
 \textsf{eq::ConfigEvent} or derived class to the application. The
@@
 
-On the application end, \textsf{Config::handleEvent} uses the ostream
-operator for the derived config event to output these events in a nicely
-formatted way:
+On the application end, \textsf{Config::handleEvent} uses the
+\textsf{ostream} operator for the derived config event to output these
+events in a nicely formatted way:
 
@@
 Normally the programmer does not need to interfere with the image
-compositing, or only at a high level, for example to order the input
-frames or to optimize the readback. The following sections provide a
-detailed description of the image compositing API in Equalizer.
+compositing. Changes are sometimes required at a high level, for example
+to order the input frames or to optimize the readback. The following
+sections provide a detailed description of the image compositing API in
+Equalizer.
 
@@
 channel, the amount of pixel data to be composited grows linearly. When
-using the naive approach of compositing all frames on the destination
+using the simple approach of compositing all frames on the destination
 channel, this channel quickly becomes the bottleneck in the
@@
 \fig{fDirectSend} illustrates this algorithm for three channels. The
-Equalizer website contains a presentation for this
+Equalizer website contains a presentation explaining this
 algorithm\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}.
@@
 
-For output frames, the frame data's pixel viewport (pvp) is the area of
-the frame buffer to read back. It will transport the offset from the
-source to the destination channel, that is, the frame data pvp for input
-frames position the pixel data on the destination. This has the effect
-that a partial framebuffer readback will end up in the same place in the
-destination channels.
+For output frames, the frame data's pixel viewport is the area of the
+frame buffer to read back. It will transport the offset from the source
+to the destination channel, that is, the frame data pixel viewport for
+input frames position the pixel data on the destination. This has the
+effect that a partial framebuffer readback will end up in the same place
+in the destination channels.
 
@@
 channels produces pixel data which can be handled like one big 'slice'
-through the full texture. Therefore they have to be blended
+through the full texture. Therefore they have to be blen\-ded
 back-to-front in the same way as the slice planes during rendering.
@@
 channel), fill rate (less samples per channel) and IO bandwidth for
-time-dependend data (less data per time step and channel). Since the
+time-dependent data (less data per time step and channel). Since the
 amount of texture memory needed for each node decreases linearly, they
@@
 between the origin$\rightarrow$slice vector and the near plane, as shown
-in \fig{fBlendPerspective}. When the angle becomes bigger than
+in \fig{fBlendPerspective}. When the angle becomes greater than
 90\textdegree, the compositing order of the remaining frames has to be
@@
 \fig{fMono} illustrates a monoscopic view frustum. The viewer is
-positioned at the origin, and the frustum is completely symmetric. This
-is the typical view frustum for non-stereoscopic applications.
+positioned at the origin of the global coordinate system, and the
+frustum is completely symmetric. This is the typical view frustum for
+non-stereoscopic applications.
 

### 2007-11-13 Developer/ProgrammingGuide/paper.tex
Update Quickstart Guide, fix some typos
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation
- Implement OpenMP-based parallel image compression - Make object change manager configuration explicit - Programming Guide and documentation for the above
Existing eqNet::Object subclasses have now to use getChangeType()
instead of isStatic(), usePack() and/or isUnbuffered().


[ ] May break build
[x] Breaks existing applications
[ ] Bugfix
[x] New Feature
[x] Cleanup
[ ] Optimization
[ ] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.0 for Equalizer 0.4\\[\medskipamount]
+  Version 1.1.0 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
     1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
+    1.0.1   & Nov 06, 2007 & Fixed Typo\\
+    1.1     & Nov 12, 2007 & Adapted distributed objects\\
     \hline
@@
 
-\subsection{Execution Modeal and Thread Safety}
+\subsection{Execution Model and Thread Safety}
 
@@
 
+\subsubsection{Change Handling}
+
+The way changes are to be handled is determined by calling
+\textsf{Object::getChangeType} during the registration of the master
+version of a distributed object . The change type determines the memory
+footprint and the contract for calling the serialization methods. The
+following change types are possible:
+
+\begin{description}
+  \item[STATIC] The object is not versioned. The instance data is
+    serialized whenever a new slave instance is mapped. No additional
+    data is stored.
+  \item[INSTANCE] The object is versioned, and the instance and delta
+    data is identical, that is, only instance data is
+    serialized. Previous instance data is saved to be able to map old
+    versions.
+  \item[DELTA] The object is versioned, and the delta data is typically
+    smaller than the instance data. Both the delta and instance data are
+    serialized and saved to map old versions.
+  \item[DELTA\_UNBUFFERED] The object is versioned, and delta data is
+    used to update slave instances. No data is stored, and no previous
+    versions can be mapped. The instance data is serialized whenever a
+    new slave instance is mapped.
+\end{description}
+
 \subsubsection{\label{sInitData}InitData - a Static Distributed Object}
@@
 
-Versioned objects have to override \textsf{isStatic} to return
-\textsf{false}, which indicates that they are versioned. The current
+Versioned objects have to override \textsf{getChangeType} to indicate
+how they want to have changes to be handled. The current
 implementation has the following characteristics:
@@
 size and changes frequently. Therefore, the instance and delta data are
-the same and set in the constructor. The default implementation will
-take care of the distribution of the data:
+the same and set in the constructor. Furthermore, the change type is
+\textsf{INSTANCE}. The default \textsf{Object} implementation will take
+care of the distribution of the data:
 
@@
 \begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-8ex}\includegraphics[width=.382\textwidth]{images/DB.png}
+  \includegraphics[width=.382\textwidth]{images/DB.png}
   {\caption{\small\label{fDB}Destination view of an DB compound}}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Rendering Loop}}
-\end{wrapfigure}
 The rendering loop maintains a list of candidates to render, which
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Rendering Loop}}
+\end{wrapfigure}
 The actual rendering uses display lists or vertex buffer objects. These
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
+  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
+\end{wrapfigure}
 Furthermore, the frame data transports the inherited range of the output
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
-  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
-\end{wrapfigure}
 The offset of input and output frames characterizes the position of the

### 2007-11-16 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-\title{Equalizer Programming Guide\\
-  {\vspace{-1ex}\footnotesize\sffamily\mdseries
-    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
-    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
-}
+\title{Equalizer Programming Guide}
 \author{Eyescale Software GmbH}
@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.0 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.1 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
     1.1     & Nov 12, 2007 & Adapted distributed objects\\
+    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
     \hline

### 2007-11-23 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.1 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.2 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
 Written by Stefan Eilemann.\\
-Engineering contributions by Maxim Makhinya and Jonas B\"osch.
+Engineering contributions by Maxim Makhinya, Jonas B\"osch and Christian Marten.
 
@@
     1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
+    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
     \hline
@@
 
-The \textsf{eqHello} redraw function renders six rotating, colored quads
+The \textsf{eqHello} redraw function renders six colored quads, rotating
 around the origin. The \textsf{frameDraw} meth\-od provided by the
@@
 frustum and other OpenGL state. After setting up some additional
-lighting parameter, \textsf{eqHello} rotates the scene and render the
+lighting parameter, \textsf{eqHello} rotates the scene and renders the
 quads using immediate mode:
@@
 appropriate initialization task methods, as explained in later
-sections. \textsf{Config::init} does return after all nodes, pipes,
+sections. \textsf{Config::init} returns after all nodes, pipes,
 windows and channels are initialized. It returns \textsf{true} only if
@@
 When the configuration was successfully initialized, the main rendering
-loop is executed. The main loop runs until the user exits the
+loop is executed. It runs until the user exits the
 configuration, or when a maximum number of frames has been rendered,
-when specified by a command-line argument. The latter is useful for
+specified by a command-line argument. The latter is useful for
 benchmarks. The \textsf{Clock} is reused for measuring the overall

### 2007-11-23 Developer/ProgrammingGuide/paper.tex
Add GLEW-related documentation, minor website tweaks
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.2 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.3 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
     1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
+    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
     \hline
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventInit.pdf}
-  {\caption{\small\label{fEventInit}Event Handling Initialization}}
-\end{wrapfigure}
 During window and pipe initialization the event handling is set up. For
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventInit.pdf}
+  {\caption{\small\label{fEventInit}Event Handling Initialization}}
+\end{wrapfigure}
 An event handler consists of two parts: the generic base class providing
@@
 
+\subsection{OpenGL Extension Handling}
+
+Equalizer uses GLEW\footnote{\link{http://glew.sourceforge.net}} for
+OpenGL extension handling, particularly the GLEW MX implementation for
+multi-context support.
+
+Each \textsf{eq::Window} has a \textsf{GLEWContext}. This context can be
+obtained by using \textsf{glewGetContext} on the window or channel. GLEW
+MX uses this function to dispatch the functions to the correct
+context. Equalizer (re-)initializes the GLEW context whenever a new
+OpenGL context is set on the window.
+
+Extended OpenGL functions called from a window or channel instance can
+be called directly. GLEW will call the object's \textsf{glewGetContext}
+to obtain the correct context:
+
+{\footnotesize\begin{lstlisting}
+void eqPly::Channel::_drawModel( const Model* model )
+{
+    ...
+    glUseProgram( program );
+    ...
+}
+\end{lstlisting}}
+
+Functions called from another place need to have the macro or function
+\textsf{glewGetContext} that returns the pointer to the GLEWContext of
+the appropriate window.
+
+{\footnotesize\begin{lstlisting}
+// state has GLEWContext* from window
+#define glewGetContext state.glewGetContext
+
+/*  Set up rendering of the leaf nodes.  */
+void VertexBufferLeaf::setupRendering( VertexBufferState& state,
+                                       GLuint* data ) const
+{
+    ...
+    glBindBuffer( GL_ARRAY_BUFFER, data[VERTEX_OBJECT] );
+    glBufferData( GL_ARRAY_BUFFER, _vertexLength * sizeof( Normal ),
+                    &_globalData.normals[_vertexStart], GL_STATIC_DRAW );
+    ...
+}
+\end{lstlisting}}
+
+\subsection{\label{sTracking}Head Tracking}
+
+The \textsf{eqPly} example contains rudimentary support for head
+tracking, in order to show how head tracking can be integrated with
+Equalizer. Support for a wide range of tracking devices is not within the
+scope of Equalizer. Other open source and commercial implementations
+cover this functionality sufficiently and can easily be integrated with
+Equalizer.
+
+\begin{figure}[h!t]
+  \subfigure[]{\label{fMono}
+    \includegraphics[height=3cm]{images/mono}}\hfil
+  \subfigure[]{\label{fStereo}
+    \includegraphics[height=3cm]{images/stereo}}\hfil
+  \subfigure[]{\label{fTracked}
+    \includegraphics[height=3cm]{images/tracked}}%
+  {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked
+    frustra}}
+\end{figure}
+
+\fig{fMono} illustrates a monoscopic view frustum. The viewer is
+positioned at the origin of the global coordinate system, and the
+frustum is completely symmetric. This is the typical view frustum for
+non-stereoscopic applications.
+
+In stereo rendering, the scene is rendered twice, with the two frustra
+'moved' by the distance between the eyes, as shown in \fig{fStereo}.
+
+In immersive visualization, the observer is tracked in and the view
+frustra are adapted to the viewer's position and orientation, as shown
+in \fig{fTracked}. The transformation origin $\rightarrow$ viewer is set by
+the application using \textsf{Config::setHeadMatrix}, which is used by
+the server to compute the frustra. The resulting off-axis frustra are
+positioned using the channel's head transformation.
+
 \subsection{\label{sCompositing}Image Compositing for Scalable Rendering}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
-\end{wrapfigure}
 The main idea for direct send is to parallelize the costly recomposition
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+\end{wrapfigure}
 The following operations have to be possible in order to perform this
@@
 \begin{itemize}
-\item Selection of frame buffer attachments: color and/or depth
+\item Selection of color and/or depth frame buffer attachments
 \item Restricting the read back area to a part of the rendered area
@@
 
-The frame synchronization through the frame data allows the input frame
-to wait for the pixel data to become ready, which is signalled by the
-output frame after readback.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+The frame synchronization through the frame data allows the input frame
+to wait for the pixel data to become ready, which is signalled by the
+output frame after readback.
+
 Furthermore, the frame data transports the inherited range of the output
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 Finding the correct assembly order for perspective frustra is more
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-4ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 In order to compute the assembly order, \textsf{eVolve} uses the angle
@@
 
-\subsection{\label{sTracking}Head Tracking}
-
-The \textsf{eqPly} example contains rudimentary support for head
-tracking, in order to show how head tracking can be integrated with
-Equalizer. Support for a wide range of tracking devices is not within the
-scope of Equalizer. Other open source and commercial implementations
-cover this functionality sufficiently and can easily be integrated with
-Equalizer.
-
-\begin{figure}[h!t]
-  \subfigure[]{\label{fMono}
-    \includegraphics[height=3cm]{images/mono}}\hfil
-  \subfigure[]{\label{fStereo}
-    \includegraphics[height=3cm]{images/stereo}}\hfil
-  \subfigure[]{\label{fTracked}
-    \includegraphics[height=3cm]{images/tracked}}%
-  {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked
-    frustra}}
-\end{figure}
-
-\fig{fMono} illustrates a monoscopic view frustum. The viewer is
-positioned at the origin of the global coordinate system, and the
-frustum is completely symmetric. This is the typical view frustum for
-non-stereoscopic applications.
-
-In stereo rendering, the scene is rendered twice, with the two frustra
-'moved' by the distance between the eyes, as shown in \fig{fStereo}.
-
-In immersive visualization, the observer is tracked in and the view
-frustra are adapted to the viewer's position and orientation, as shown
-in \fig{fTracked}. The transformation origin $\rightarrow$ viewer is set by
-the application using \textsf{Config::setHeadMatrix}, which is used by
-the server to compute the frustra. The resulting off-axis frustra are
-positioned using the channel's head transformation.
-
 \if 0

### 2007-11-27 Developer/ProgrammingGuide/paper.tex
Precursory work to support RenderContext association for window events. Remove glX event thread and implement message pumping for glx. Update doc accordingly.
Needs make clean.

[x] May break build
[ ] Breaks existing applications
[x] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.3 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.4 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
     1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
+    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
     \hline
@@
 
-Where possible events are received and processed by a separate per-node
-event thread to allow asynchronous\footnote{with respect to the
-  rendering} event handling. Currently an event thread is only used by
-the X11/glX window system. WGL receives and processes the events from
-the pipe threads that created the windows. AGL receives the events from
-application or node main thread. Whenever the term \textbf{event thread}
-is used, it refers to the thread receiving the event, such as a per-node
-thread for glX, the pipe thread for WGL and the main thread for AGL.
+In general, events are received and processed by the pipe thread a
+window belongs to. An exception to this rule is AGL, where all events
+are dispatched from the main node thread. WGL and GLX receive and
+process the events from the pipe threads that created the
+windows. Whenever the term \textbf{event thread} is used, it refers to
+the thread receiving the event, i.e., the pipe thread for WGL and GLX,
+and the main thread for AGL.
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventInit.pdf}
-  {\caption{\small\label{fEventInit}Event Handling Initialization}}
-\end{wrapfigure}
 An event handler consists of two parts: the generic base class providing
@@
 
-AGL and glX use an event handler singleton, whereas WGL uses one event
-handler per window. Pipe event handling is only used for glX, where one
-\textsf{Display} connection is created and used to subscribe to window
-events.
+AGL uses an event handler singleton, GLX uses one event handler per pipe
+and WGL uses one event handler per window.
 
@@
 
-For the WGL and AGL window systems, it is required to manually receive and
-dispatch (`pump') events. On WGL, this has to happen on each thread with
-windows, whereas on AGL it has to happen only on the main thread. By
-default, Equalizer pumps these events automatically for the application.
+In order to dispatch the events, Equalizer 'pumps' the native events. On
+WGL and GLX, this has happens on each thread with windows, whereas on
+AGL it has to happen only on the main thread. By default, Equalizer
+pumps these events automatically for the application in-between
+executing task methods.
 
@@
 
-Events sent using \textsf{Config::sendEvent} are queued in the
+Events send using \textsf{Config::sendEvent} are queued in the
 application thread. After a frame has been finished,
@@
 
-The main idea for direct send is to parallelize the costly recomposition
-for database (sort-last) decomposition. With each additional source
-channel, the amount of pixel data to be composited grows linearly. When
-using the simple approach of compositing all frames on the destination
-channel, this channel quickly becomes the bottleneck in the
-system. Direct send distributes this workload evenly across all source
-channels, and thereby keeps the compositing work per channel constant.
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+\end{wrapfigure}
+The main idea behind direct send is to parallelize the costly
+recomposition for database (sort-last) decomposition. With each
+additional source channel, the amount of pixel data to be composited
+grows linearly. When using the simple approach of compositing all frames
+on the destination channel, this channel quickly becomes the bottleneck
+in the system. Direct send distributes this workload evenly across all
+source channels, and thereby keeps the compositing work per channel
+constant.
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
-\end{wrapfigure}
 The following operations have to be possible in order to perform this

### 2007-12-10 Developer/ProgrammingGuide/paper.tex
Update Programming Guide after last change.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.4 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.5 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
     1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
+    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
     \hline
@@
 public:
-    virtual eq::Config*  createConfig()  { return new eqPly::Config; }
-    virtual eq::Node*    createNode()    { return new eqPly::Node; }
-    virtual eq::Pipe*    createPipe()    { return new eqPly::Pipe; }
-    virtual eq::Window*  createWindow()  { return new eqPly::Window; }
-    virtual eq::Channel* createChannel() { return new eqPly::Channel; }
+    virtual eq::Config*  createConfig()  
+        { return new eqPly::Config; }
+    virtual eq::Node*    createNode( eq::Config* parent )  
+        { return new eqPly::Node( parent ); }
+    virtual eq::Pipe*    createPipe( eq::Node* parent )
+        { return new eqPly::Pipe( parent ); }
+    virtual eq::Window*  createWindow( eq::Pipe* parent )
+        { return new eqPly::Window( parent ); }
+    virtual eq::Channel* createChannel( eq::Window* parent )
+        { return new eqPly::Channel( parent ); }
 };

### 2007-12-17 Developer/ProgrammingGuide/paper.tex
Update Programming Guide and scalability page
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.5 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.6 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
     1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
+    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
     \hline
@@
 \end{wrapfigure}
-
 The \textsf{eqHello} example is a minimal application to illustrate the
@@
 are explained in \sref{sEqPly}.
+
+\section{Scalable Rendering}
+
+Real-time visualization is an inherently parallel
+problem. Unfortunately, different applications have different rendering
+algorithms, which require different scalable rendering modes to address
+the right bottlenecks. Equalizer supports the most important algorithms,
+and will continue to add new ones over time to meet application
+requirements.
+
+This section gives an introduction to scalable rendering, in order to
+provide some background for application developers. The scalability
+modes offered by Equalizer are discussed, along with their advantages
+and disadvantages.
+
+Choosing the right mode for the application profile is critical for
+performance. Equalizer uses the concept of compounds to describe the
+task decomposition and result recomposition. It allows the combination
+of the different compound modes in any possible way, which allows the
+address different bottlenecks in a flexible way.
+
+\subsection{2D or Sort-First Compounds}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/2D.pdf}
+  {\caption{\small A 2D compound}}
+\end{wrapfigure}
+2D or sort-first decomposes the rendering in screen-space, that is, each
+contributing rendering unit processes a tile of the final view. The
+recomposition simply assembles the tiles side-by-side on the destination
+view.
+
+The advantage of this mode is a low, constant IO overhead for the pixel
+transfers, since only color information has to be transmitted. The upper
+limit is the number of pixels of the destination view.
+
+It's disadvantage is that it relies on view frustum culling to reduce
+the amount of data rendered. Depending on the application data
+structure, the overlap of some primitives between individual tiles
+limits the scalability of this mode, typically to around eight graphics
+cards. Each node has to potentially hold the full database for rendering.
+
+2D decompositions can be used by all types of applications, but should
+be combined with DB compounds to reduce the data per node, if possible.
+
+\subsection{DB or Sort-Last Compounds}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/DB.pdf}
+  {\caption{\small A database compound}}
+\end{wrapfigure}
+DB or sort-last\footnote{3D model courtesy of AVS, USA.} decomposes the
+rendered database so that all rendering units process a part of the
+scene in parallel.
+
+Volume rendering applications use an ordered alpha-based blending to
+composite the result image. The depth buffer information is used to
+composite the individual images correctly for polygonal data.
+
+This mode provides very good scalability, since each rendering unit
+processes a part of the database. This allows to lower the requirements
+on all parts of the rendering pipeline: main memory usage, IO bandwidth,
+GPU memory usage, vertex processing and fill rate.
+
+Unfortunately, the database recomposition has linear increasing IO
+requirements for the pixel transfer. Parallel recomposition algorithms,
+such as direct-send address this problem by keeping the per-node IO
+constant (see \fig{fDirectSend}).
+
+The application has to partion the database so that the rendering units
+render only part of the database. Some OpenGL features, do not work
+correctly (anti-aliasing) or need special attention (transparency).
+
+The best use of database compounds is to divide the data to a managable
+size, and then to use other decomposition modes to achieve more
+scalability. Volume rendering is one of the applications which can
+profit from database compounds.
+
+\subsection{Stereo Compounds}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
+  {\caption{\small A stereo compound}}
+\end{wrapfigure}
+Stereo compounds\footnote{3D model courtesy of Stereolithography Archive
+  at Clemson University.} assign each eye pass to individual rendering
+units. The resulting images are copied to the appropriate stereo
+buffer. This modes supports a variety of stereo modes, including active
+(quad-buffered) stereo, anaglyphic stereo and auto-stereo displays with
+multiple eye passes.
+
+Due to the frame consistency between the eye views, this modes scales
+very well. The IO requirements for pixel transfer are small and
+constant.
+
+The number of rendering resources used by stereo compounds is limited by
+the number of eye passes, typically two.
+
+Stereo compounds are used by all applications when rendering in stereo,
+and is often combined with other modes.
+
+\subsection{Pixel Compounds}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
+  {\caption{\small A pixel compound}}
+\end{wrapfigure}
+Pixel compounds are similar to 2D compounds. The frustra of the
+source rendering units are modified so that each unit renders an evenly
+distributed subset of pixels.
+
+As 2D compounds, pixels compounds have low, constant IO requirements for
+the pixel transfers during recomposition.
+
+Pixel compounds only work very well for purely fill-limited
+applications, techniques like frustum culling do not reduce the rendered
+data for the source rendering resources.
+
+OpenGL functionality influenced by the raster position will not work
+correctly with pixel compounds, or needs at least special
+attention. Among them are: lines, points, sprites, glDrawPixels,
+glBitmap, glPolygonStipple.
+
+Pixel compounds are ideal for raytracing, which is highly fill-limited
+and needs the full database for rendering anyway. Volume rendering
+applications should choose this mode over 2D compounds.
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
-  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
-\end{wrapfigure}
 The \textsf{eqPly::Config} class holds the master versions of the
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
+  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
+\end{wrapfigure}
 If the configuration was initialized correctly, the configuration tries
@@
 
+\fig{fWindowInit} shows a flow chart of the window initialization. The
+colored functions are task methods and can be replaced by
+application-specific implementations.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
   {\caption{\small\label{fWindowInit}Window Initialization}}
-\end{wrapfigure}
-\fig{fWindowInit} shows a flow chart of the window initialization. The
-colored functions are task methods and can be replaced by
-application-specific implementations.
-
+\vspace{-4ex}\end{wrapfigure}
 The window-system specific initialization takes into account various
@@
 was successful, it initializes a state object and an overlay logo:
+\vspace{4ex}
 
@@
     eq::Pipe*  pipe        = getPipe();
-    Window*    firstWindow = static_cast< Window* >( pipe->getWindow( 0 ));
+    Window*    firstWindow = static_cast< Window* >( pipe->getWindows()[0] );
 
@@
     {
-        _state = new VertexBufferState( getGLFunctions( ));
-        _loadLogo();
+        _state = new VertexBufferState( glewGetContext( ));
 
@@
         {
-            const eq::GLFunctions* glFunctions = getGLFunctions();
-            // Check if all VBO funcs available, else leave DISPLAY_LIST_MODE on
-            if( glFunctions->hasGenBuffers() && glFunctions->hasBindBuffer() &&
-                glFunctions->hasBufferData() && glFunctions->hasDeleteBuffers())
+            // Check if VBO funcs available, else leave DISPLAY_LIST_MODE on
+            if( GLEW_VERSION_1_5 )
             {
@@
         }
+
+        if( initData.useGLSL() )
+        {
+            // Check if all functions are available
+            if( GLEW_VERSION_2_0 )
+            {
+                EQINFO << "Shaders supported, attempting to load" << endl;
+                _loadShaders();
+            }
+            else
+                EQWARN << "Shader function pointers missing, using fixed "
+                       << "pipeline" << endl;
+        }
+
+        _loadLogo();
     }
@@
     if( !_state ) // happens if first window failed to initialize
+    {
+        setErrorMessage( "No state handler object" );
         return false;
+    }
+
+    // turn off OpenGL lighting if we are using our own shaders
+    if( _state->getProgram( pipe ) != VertexBufferState::FAILED )
+        glDisable( GL_LIGHTING );
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Rendering Loop}}
+\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Rendering Loop}}
-\end{wrapfigure}
 The actual rendering uses display lists or vertex buffer objects. These

### 2007-12-20 Developer/ProgrammingGuide/paper.tex
Documentation updates for last change
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.6 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.7 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
     1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
+    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
     \hline
@@
 identify objects. The same key is used by all contexts to get the
-OpenGL name of an object. In \textsf{eqPly}, a key of type \textsf{const
-  void *} is used. The rendering code uses the address of the data item
-to be rendered as the key to obtain the associated OpenGL object.
-
-All objects managed by the OM are reference counted. If an application
-releases the objects properly, they are automatically de-allocated. It
-is also possible to manually manage de-allocation of objects, which
-might be more convenient in some cases.
+OpenGL name of an object.
+
+Each \textsf{eq::Window} has an object manager with the key type
+\textsf{const void*} for as long as it has an OpenGL context. The OM is
+shared between all windows of a pipe, if the windows are created using
+the default \textsf{configInit} functions. If the window is created by
+the application, the OM is not shared since no assumption can be made
+about OpenGL context sharing. Later version of Equalizer will introduce
+an API to configure OpenGL context sharing, and therefore OM sharing, for
+application-created windows.
+
+\textsf{eqPly} uses the window's object manager in the rendering code to
+obtain the OpenGL objects for a given data item. The address of the data
+item to be rendered is used as the key. All objects managed by the OM
+are reference counted. If an application releases the objects properly,
+they are automatically de-allocated. It is also possible to manually
+manage de-allocation of objects, which might be more convenient in some
+cases.
 

### 2008-01-15 Developer/ProgrammingGuide/paper.tex
Update Programmers Guide (window init refactorings)
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.7 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.8 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
     1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
+    1.1.8   & Jan 15, 2007 & Detailed window initialization based on
+                             latest refactoring\\
     \hline
@@
 
-\subsubsection{Carbon/AGL Thread Safety}
+\subsubsection{\label{sCarbonThread}Carbon/AGL Thread Safety}
 
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Rendering Loop}}
+  \vspace{-2.5ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Rendering Loop}\vspace{-2ex}}
 \end{wrapfigure}
@@
 
+\subsection{Advanced Window Initialization}
+
+This section explains window initialization in detail. It discussed in
+detail the handling on the different window systems. The entry point for
+the window initialization is the task method
+\textsf{Window::configInit}. This task method first calls a
+window-system-specific task method, and then
+\textsf{Window::configInitGL} to do generic OpenGL state setup, as shown
+in \fig{fWindowInit}.
+
+Since window initialization is notoriously error-prone and hard to
+debug, the default Equalizer functions propagate the reason for errors
+from the render clients back to the application. The \textsf{Pipe} and
+\textsf{Window} classes have a \textsf{setErrorMessage} method, which is
+used to set an error string. This string is passed to the
+\textsf{Config} instance on the application node, where it can be
+queried using \textsf{getErrorMessage}.
+
+The window-system-specific initialization methods use overrideable
+methods for all sub-tasks. This allows partial customization, without
+the need of rewriting tedious window initialization code, e.g., the
+OpenGL pixel format selection.
+
+\subsubsection{AGL Window Initialization}
+
+AGL initialization happens in three steps: choosing a pixel format,
+creating the context and creating a drawable.
+
+Most AGL and Carbon calls are not thread-safe. The Equalizer methods
+calling these functions use \textsf{Global::enterCarbon} and
+\textsf{Global::leaveCarbon} to protect the API calls. Please refer to
+\sref{sCarbonThread} for more details.
+
+The pixel format is chosen based on the window's attributes. Some
+attributes set to auto, e.g., stereo, cause the method first to request
+the feature and then to back off and retry if it is not available. The
+returned pixel format has to be destroyed using
+\textsf{Window::destroyAGLPixelFormat}. When no matching pixel format is
+found, \textsf{chooseAGLPixelFormat} returns \textsf{0} and the AGL
+window initialization returns with a failure.
+
+The context creation also uses the global Carbon lock. Furthermore, it
+sets up the swap buffer synchronization with the vertical retrace, if
+enabled by the corresponding window attribute hint. Again the window
+initialization fails if the context could not be created.
+
+The drawable creation method \textsf{configInitAGLDrawable} calls either
+\textsf{configInitAGLFullscreen} or \textsf{configInitAGLWindow}.
+
+The top-level AGL window initialization code therefore looks as follows:
+
+{\footnotesize\begin{lstlisting}
+bool Window::configInitAGL()
+{
+    AGLPixelFormat pixelFormat = chooseAGLPixelFormat();
+    if( !pixelFormat )
+        return false;
+
+    AGLContext context = createAGLContext( pixelFormat );
+    destroyAGLPixelFormat ( pixelFormat );
+    setAGLContext( context );
+
+    if( !context )
+        return false;
+
+    const bool success = configInitAGLDrawable();
+    if( success && !_xDrawable )
+    {
+        setErrorMessage( "configInitAGLDrawable did set no AGL drawable" );
+        return false;
+    }
+
+    return success;
+}
+\end{lstlisting}}
+
+\subsubsection{GLX Window Initialization}
+
+GLX initialization is very similar to AGL initialization. Again the
+steps are: choose visual (pixel format), create OpenGL context and then
+create drawable. The only difference is that the data returned by
+\textsf{chooseXVisualInfo} has to be freed using \textsf{XFree}:
+
+{\footnotesize\begin{lstlisting}
+bool Window::configInitGLX()
+{
+    XVisualInfo* visualInfo = chooseXVisualInfo();
+    if( !visualInfo )
+        return false;
+
+    GLXContext context = createGLXContext( visualInfo );
+    setGLXContext( context );
+
+    if( !context )
+        return false;
+
+    const bool success = configInitGLXDrawable( visualInfo );
+    XFree( visualInfo );
+
+    if( success && !_xDrawable )
+    {
+        setErrorMessage( "configInitGLXDrawable did set no X11 drawable" );
+        return false;
+    }
+
+    return success;    
+}
+\end{lstlisting}}
+
+\subsubsection{WGL Window Initialization}
+
+The WGL initialization requires another order of operations compared to
+AGL or GLX. The following functions are used to initialize a WGL window:
+
+\begin{enumerate}
+\item\textsf{getWGLDeviceContext} is used to get a device context which
+  might be needed for window creation. Ordinary windows do not need a
+  device context during creation, but PBuffers do need one. If a context
+  is returned, a function pointer to delete the returned device context
+  is set as well.
+\item\textsf{configInitWGLDrawable} creates the drawable. The device
+  context passed to \textsf{configInitWGLDrawable} is not used when
+  creating a window. On success, it sets the window handle.
+\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
+  window attributes. If no device context is given, it uses the device
+  context from this window's window handle. The chosen pixel format is
+  set on the window's device context as well as on the passed device
+  context. The window needs a window handle set before this method is
+  called.
+\item\textsf{createWGLContext} creates an OpenGL rendering context using
+  the given device context. The device context can not be 0. It does not
+  set the window's OpenGL context.
+\end{enumerate}
+
+The full \textsf{configInitWGL} task method, including error handling
+and cleanup, looks as follows:
+
+{\footnotesize\begin{lstlisting}
+bool Window::configInitWGL()
+{
+    PFNEQDELETEDCPROC deleteDCProc = 0;
+    HDC dc = getWGLDeviceContext( deleteDCProc );
+    EQASSERT( !dc || deleteDCProc );
+
+    if( !configInitWGLDrawable( dc ))
+    {
+        if( dc )
+            deleteDCProc( dc );
+        return false;
+    }
+
+    if( !_wglWindowHandle )
+    {
+        if( dc )
+            deleteDCProc( dc );
+        setErrorMessage( "configInitWGLDrawable did set no WGL drawable" );
+        return false;
+    }
+
+    int pixelFormat = chooseWGLPixelFormat( dc );
+    if( pixelFormat == 0 )
+    {
+        if( dc )
+            deleteDCProc( dc );
+        return false;
+    }
+
+    HDC windowDC = GetDC( _wglWindowHandle );
+    HGLRC context = createWGLContext( dc ? dc : windowDC );
+    ReleaseDC( _wglWindowHandle, windowDC );
+
+    setWGLContext( context );
+
+    if( !context )
+    {
+        if( dc )
+            deleteDCProc( dc );
+        return false;
+    }
+
+    if( dc )
+        deleteDCProc( dc );
+    return true;
+}
+\end{lstlisting}}
+
 \subsection{\label{sTracking}Head Tracking}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
-\end{wrapfigure}
 The main idea behind direct send is to parallelize the costly
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+\end{wrapfigure}
 In direct send compositing, each rendering channel is also responsible
@@
 
+The frame synchronization through the frame data allows the input frame
+to wait for the pixel data to become ready, which is signalled by the
+output frame after readback.
+
+Furthermore, the frame data transports the inherited range of the output
+frame's compound. The range can be used to compute the assembly order of
+multiple input frames, e.g., for sorted-blend compositing in volume
+rendering applications.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The frame synchronization through the frame data allows the input frame
-to wait for the pixel data to become ready, which is signalled by the
-output frame after readback.
-
-Furthermore, the frame data transports the inherited range of the output
-frame's compound. The range can be used to compute the assembly order of
-multiple input frames, e.g., for sorted-blend compositing in volume
-rendering applications.
-
 Readback and assemble operations on frames and images are designed to be
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
+      Volume Rendering}}
+\end{wrapfigure}
 The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
-      Volume Rendering}}
-\end{wrapfigure}
 The full power of scalable volume rendering is however in DB (sort-last)
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 For orthographic rendering, determining the compositing order of the
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 Finding the correct assembly order for perspective frustra is more

### 2008-01-16 Developer/ProgrammingGuide/paper.tex
Update Programmers Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.8 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.9 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
                              latest refactoring\\
+    1.1.9   & Jan 16, 2007 & Detail distributed object usage\\
     \hline
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/model.pdf}
+  {\caption{\small\label{fModel}Simplified execution model}}
+\end{wrapfigure}
 The main thread is responsible for maintaining the application logic. It
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/model.pdf}
-  {\caption{\small\label{fModel}Simplified execution model}}
-\end{wrapfigure}
 All rendering threads in the configuration run asynchronously to the
@@
 \subsubsection{Frustum}
+Compounds have a frustum description to define the physical layout of
+the display environment.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-Compounds have a frustum description to define the physical layout of
-the display environment.
-
 The frustum description is inherited by the children, therefore the
@@
 
-By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
-completion of all local rendering tasks for the current frame. This
-facilitates porting of exising rendering codes, since the database does
-not have to be multi-buffered. Applications such as \textsf{eqPly}, which
-do not need this per-node frame synchronization can disable it, as
-explained in \sref{sNodeFrame}.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
+completion of all local rendering tasks for the current frame. This
+facilitates porting of exising rendering codes, since the database does
+not have to be multi-buffered. Applications such as \textsf{eqPly}, which
+do not need this per-node frame synchronization can disable it, as
+explained in \sref{sNodeFrame}.
+
 \fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
@@
 rendering methods.
+
+\subsubsection{Best Practices}
+
+Distributed objects are addressed using session-unique identifiers,
+because pointers to other objects can not be distributed directly, since
+they have no meaning on remote nodes. The session with which distributed
+objects are registered is normally the \textsf{eq::Config}.
+
+The shared data entry point for the render clients is the identifier
+passed by the application to \textsf{Config::init}. This identifier
+typically contains the identifier of a static distributed object, and is
+passed by Equalizer to all \textsf{configInit} task methods. Normally
+this initial object is mapped by the \textsf{eq::Node}, and contains
+identifiers to other shared data objects.
+
+The distributed data objects referenced by the initial data object are
+often versioned objects, to keep them in sync with the rendered
+frames. Similar to the initial identifier passed to
+\textsf{Config::init}, an object identifier or version can be passed to
+\textsf{Config::startFrame}. Equalizer will pass this identifier to all
+\textsf{frameStart} task methods. In \textsf{eqPly}, the frame-specific
+data, e.g., the global camera data, is versioned. The frame data
+identifier is passed in the initial data, and the frame data version is
+passed with each start frame request.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/objects.pdf}
+  {\caption{\small\label{fObjects}Distributed Data Handling}}
+\end{wrapfigure}
+Applications distributing their scene graph, rather then relying on a
+shared filesystem like \textsf{eqPly}, use the frame data also as an
+entry point to their scene graph data structure. \fig{fObjects} shows
+one possible implementation, where the identifier and version of the
+scene graph root are transported using the frame data. The scene graph
+root then serializes and de-serializes his immediate children by
+transferring their identifier and current version.
+
+The objects are still created by the application, and then registered or
+mapped with the session in order to distribute them. When mapping
+objects in a hierarchical data structure, their type often has to be
+known in order to create them. Equalizer does not currently provide
+object typing, this has to be done by the application, either implicitly
+in the current implementation context, or by transferring a type
+identifier. In \textsf{eqPly}, object typing is implicit since it is
+well-defined which object is mapped in which task method.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
+  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
+\end{wrapfigure}
 The config initialization happens in parallel, that is, all config
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
-  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
-\end{wrapfigure}
 If the configuration was initialized correctly, the configuration tries
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
+  {\caption{\small\label{fWindowInit}Window Initialization}}
+\end{wrapfigure}
 The initialization sequence uses multiple, overrideable task
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
-  {\caption{\small\label{fWindowInit}Window Initialization}}
-\vspace{-4ex}\end{wrapfigure}
 The window-system specific initialization takes into account various
@@
 was successful, it initializes a state object and an overlay logo:
-\vspace{4ex}
 
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2.5ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Rendering Loop}\vspace{-2ex}}
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Rendering Loop}}
 \end{wrapfigure}
@@
 API, the direct send parallel compositing algorithm is introduced in this
-section.
+section. Other parallel compositing algorithms, e.g. binary-swap, can
+also be expressed through an Equalizer configuration file.
 

### 2008-01-29 Developer/ProgrammingGuide/paper.tex
Update programming guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.9 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.10 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
     1.1.9   & Jan 16, 2007 & Detail distributed object usage\\
+    1.1.10  & Jan 28, 2007 & Add creation thread for NodeFactory entities\\
     \hline
@@
 rendering resources used in the configuration, one C++ object will be
-created during initialization:
+created during initialization. Config, node and pipe objects are created and
+destroyed in the node thread, whereas window and channel objects are
+created and destroyed in the pipe thread:
 

### 2008-02-20 Developer/ProgrammingGuide/paper.tex
Programming Guide: add advance chapter on thread synchronization
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.10 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.11 for Equalizer 0.4+svn\\[\medskipamount]
   \today
@@
 
-\textcopyright 2007 Eyescale Software GmbH. All rights reserved. No
+\textcopyright 2007-2008 Eyescale Software GmbH. All rights reserved. No
 permission is granted to copy, distribute, or create derivative works
@@
     1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
-    1.1.8   & Jan 15, 2007 & Detailed window initialization based on
+    1.1.8   & Jan 15, 2008 & Detailed window initialization based on
                              latest refactoring\\
-    1.1.9   & Jan 16, 2007 & Detail distributed object usage\\
-    1.1.10  & Jan 28, 2007 & Add creation thread for NodeFactory entities\\
+    1.1.9   & Jan 16, 2008 & Detail distributed object usage\\
+    1.1.10  & Jan 28, 2008 & Add creation thread for NodeFactory entities\\
+    1.1.11  & Feb 20, 2008 & Add advanced chapter for thread synchronization\\
     \hline
@@
 
+\subsubsection{Parallel Rendering}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-1em}
+  \includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
+  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
+  \vspace{-1em}
+\end{wrapfigure}
+
+\fig{fExecutionFlow} illustrates the basic principle of any parallel
+rendering application. The typical OpenGL application, for example
+GLUT-based applications, has an event loop which redraws the scene,
+updates data based on received events, and eventually redraws a new
+frame.
+
+Any parallel rendering application keeps this basic execution model and
+extends it by separating the rendering code from the main event
+loop. This rendering coded is then executed in parallel on different
+resources. This model is followed by Equalizer, thus making application
+development as easy as possible.
+
 \section{Hello, World!}
@@
 
-\subsection{Execution Model and Thread Safety}
-
+\subsection{\label{sExecution}Execution Model and Thread Safety}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/model.pdf}
+  {\caption{\small\label{fModel}Simplified execution model}}
+\end{wrapfigure}
 Using threading correctly in OpenGL-based applications is easy with
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/model.pdf}
-  {\caption{\small\label{fModel}Simplified execution model}}
-\end{wrapfigure}
 The main thread is responsible for maintaining the application logic. It
@@
 frame synchronization, as explained in \sref{sNodeFrame}. Some scene
-graphs do implement multi-buffered data.
+graphs for example do implement multi-buffered data.
 
@@
 
+By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
+completion of all local rendering tasks for the current frame. This
+facilitates porting of exising rendering codes, since the database does
+not have to be multi-buffered. Applications such as \textsf{eqPly}, which
+do not need this per-node frame synchronization can disable it, as
+explained in \sref{sNodeFrame}.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
-completion of all local rendering tasks for the current frame. This
-facilitates porting of exising rendering codes, since the database does
-not have to be multi-buffered. Applications such as \textsf{eqPly}, which
-do not need this per-node frame synchronization can disable it, as
-explained in \sref{sNodeFrame}.
-
 \fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
@@
 
-\subsubsection{Best Practices}
-
+\subsubsection{Object Usage for Rendering}
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/objects.pdf}
+  {\caption{\small\label{fObjects}Distributed Data Handling}}
+\end{wrapfigure}
 Distributed objects are addressed using session-unique identifiers,
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/objects.pdf}
-  {\caption{\small\label{fObjects}Distributed Data Handling}}
-\end{wrapfigure}
 Applications distributing their scene graph, rather then relying on a
@@
 
-\subsubsection{\label{sNodeFrame}Frame Control}
-
-The application has extended control over the task synchronization
-during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
-\textsf{frameStart} task methods of the various entities. The entities
-unlock all its children by calling \textsf{startFrame}, e.g.,
-\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
-to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
-e.g., \textsf{Window::startFrame}, are currently empty since the
-synchronization is implicit due to the sequential execution within the
-thread.
-
-Likewise, \textsf{Config::finishFrame} causes the invokation of the
-\textsf{frameFinish} task methods. These task methods unlock their
-parents by calling \textsf{releaseFrame}.
-
-The explicit synchronization of child or parent resources allows the
-application to optimize the processing, by performing certain, independent
-operations when the child or parent resources are already unlocked.
-
-The default thread synchronization synchronizes all
-\textsf{Channel::frameDraw} operations on a \textbf{single} node with
-the node's main thread. This facilitates porting, since the scene
-database does not have to be multi-buffered as all threads of the node
-process render the same frame.
-
-The per-node frame synchronization does not break the asynchronous
-execution across rendering nodes.  Advanced applications can remove per-node
-frame synchronization.
-
-The per-node frame synchronization is achieved through the
-\textsf{startFrame\textrightarrow wait\-Fra\-me\-Started} and
-\textsf{releaseFrameLocal\textrightarrow waitFrameLocal} synchronization
-points. Note that this synchronization is only per-node, different nodes
-in the cluster still run asynchronously.
-
-The first synchronization point ensures that the application (node)
-thread is done modifiying the data, typically during event
-processing. The pipe threads call \textsf{Node::waitFrameStarted} which
-blocks until the node calls \textsf{startFrame} in
-\textsf{Node::frame\-Start}.
-
-The second pair ensures that after \textsf{Config::finishFrame} all pipe
-threads are done rendering the current frame, and therefore do not
-access shared data. At this point the application thread can safely
-modify scene data. The node's \textsf{frameDrawFinish} waits for all
-local pipes to release the synchronization by calling
-\textsf{Pipe::release\-Fra\-me\-Lo\-cal}, which happens by default in
-\textsf{Pipe::frameDrawFinish}. The \textsf{frameDrawFinish} methods are
-called after all \textsf{Channel::frameDraw} of the corresponding thread
-have been executed.
-
-\begin{figure}[ht!]\center
-  \includegraphics[width=\textwidth]{images/mainloop.pdf}
-  {\caption{\small\label{fFrameSync}Synchronization of frame tasks}}
-\end{figure}
-
-\fig{fFrameSync} outlines the synchronization for the application, node
-and pipe classes. Please note that \textsf{Config::finishFrame} does
-block until the current frame has been released locally and until the
-frame \textsf{current - latency} has been released by all nodes. The
-window and channel synchronization are similar and omitted for
-simplicity.
-
-It is absolutely vital for the execution that \textsf{Node::start\-Fra\-me}
-and \textsf{Node::release\-Fra\-me} are called, respectively. The default
-implementation of the node task methods does take care of that.
-
-Since \textsf{eqPly} multi-buffers all dynamic data, it can completely
-remove frame synchronization by:
-
-\begin{itemize}
-\item releasing the local synchronization early in \textsf{Node::frameStart}
-\item not waiting for the node to start the frame by not calling
-  \textsf{Node::waitFrameStarted} in \textsf{Pipe::frameStart}
-\item not waiting for the pipe synchronization in \textsf{Node::frameDrawFinish}
-\end{itemize}
-
-Releasing the local synchronization early is done by calling
-\textsf{releaseFrameLocal}, and \textsf{Node::frameDrawFinish} is
-overwritten with an empty implementation:
+The two remaining functions in the node relax the thread synchronization
+between the node and pipe threads, since all dynamic data is
+multi-buffered in \textsf{eqPly}. \sref{sThreads} provides a detailed
+explanation of thread synchronization in Equalizer:
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
-  {\caption{\small\label{fWindowInit}Window Initialization}}
-\end{wrapfigure}
 The initialization sequence uses multiple, overrideable task
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
+  {\caption{\small\label{fWindowInit}Window Initialization}}
+\end{wrapfigure}
 The window-system specific initialization takes into account various
 attributes set in the configuration file. Attributes include the size of
-various frame buffer attachments (color, alpha, depth, stencil) as well
-as other framebuffer properties, such as quad-buffered stereo,
+the various frame buffer planes  (color, alpha, depth, stencil) as well
+as other framebuffer attributes, such as quad-buffered stereo,
 doublebuffering, fullscreen mode and window decorations. Some of the
@@
 
-The \textsf{eqPly} window initialization function first calls
-\textsf{eq::Window::configInit} to use the generic window setup. If that
-was successful, it initializes a state object and an overlay logo:
-
-{\footnotesize\begin{lstlisting}
-bool Window::configInit( const uint32_t initID )
-{
-    if( !eq::Window::configInit( initID ))
+For the window-system specific initialization, \textsf{eqPly} uses the
+default Equalizer implemenation. The \textsf{eqPly} window
+initialization only overrides the OpenGL-specific initialization
+function in order to initialize a state object and an overlay logo. This
+function is only called if an OpenGL context was created and made current:
+
+{\footnotesize\begin{lstlisting}
+bool Window::configInitGL( const uint32_t initID )
+{
+    if( !eq::Window::configInitGL( initID ))
         return false;
@@
 \begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2em}
   \includegraphics[width=.618\textwidth]{images/render.pdf}
   {\caption{\small\label{fRender}Main Rendering Loop}}
+  \vspace{-2em}
 \end{wrapfigure}
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\small\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 Events are received by an event handler. The event handler finds the
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\small\label{fEventProcessing}Event Processing}}
+\end{wrapfigure}
 The event handler then passes the window event to
@@
 passed to a previously installed, window-system-specific event handling
-function. The default implementation of \textsf{Window::processEvent}
+function. The default implementation of \textsf{Window::pro\-cess\-Event}
 passes most events on to the application.
@@
 
-Later Equalizer versions will introduce \textsf{Pipe::processEvent} and
+Later Equalizer versions will introduce \textsf{Pipe::pro\-cess\-Event} and
 \textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
@@
 \end{lstlisting}}%>>
+
+\subsection{\label{sThreads}Multi-Threading and Synchronization}
+
+Equalizer applications use multiple, asynchronous execution threads. The
+default execution model is focused on making the porting of existing
+applications as easy as possible, as described in \sref{sExecution}. The
+default, per-node thread synchronization provided by Equalizer can
+gradually be relaxed by advanced applications to gain better performance
+through higher asynchronicity.
+
+This section explains the multi-threading in Equalizer in detail and
+gives advice on how to optimize applications for performance.
+
+\subsubsection{Threads}
+
+The application or node main thread is the primary thread of each
+process and executes the \textsf{main} function. The application and
+render clients initialize the local node for communications with other
+nodes, including the server, using \textsf{Client::initLocal}.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/threads.pdf}
+  {\caption{\small\label{fThreads}Threads within one node process}}
+\end{wrapfigure}
+During this initialization, Equalizer creates and manages two threads
+for communication, the receiver thread and the command thread. Normally
+no application code is executed from these two threads.
+
+The receiver thread manages the network connections to other nodes and
+receives data. It dispatches the received data either to the application
+threads, or to the command thread.
+
+The command thread processes Equalizer-related request from other nodes,
+for example during \textsf{eqNet::Object} mapping. In some special cases
+the command thread executes application code, for example when a remote
+node maps a static or unbuffered object,
+\textsf{Object::getInstanceData} is called from the command thread.
+
+The receiver and command thread are terminated when the application
+stops network communications using \textsf{Client::exitLocal}.
+
+During config initialization, one pipe thread is created for each
+pipe. The pipe threads execute all render task methods for this pipe,
+and therefore executes the application's rendering code. The pipe
+threads are terminated during \textsf{Config::exit}.
+
+The rest of this section discusses the thread synchronization between
+the main thread and the pipe threads.
+
+\subsubsection{\label{sNodeFrame}Thread Synchronization Model}
+
+The node main thread and the pipe threads are synchronized with each
+other, so that all draw task methods of one rendered frame are executed
+at the same time. This model allows to use the same database for
+rendering, and safe modifications of this database are possible from the
+node thread, since the pipe threads do not execute any rendering tasks
+between frames.
+
+Please note that the default thread synchronization synchronizes all
+\textsf{Channel::frameDraw} operations on a \textbf{single} node with
+the node's main thread. The per-node frame synchronization does not
+break the asynchronous execution across nodes. Advanced applications can
+remove the per-node frame synchronization.
+
+The application has extended control over the task synchronization
+during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
+\textsf{frameStart} task methods of the various entities. The entities
+unlock all their children by calling \textsf{startFrame}, e.g.,
+\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
+to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
+e.g., \textsf{Window::startFrame}, are currently empty since the
+synchronization is implicit due to the sequential execution within the
+thread.
+
+Each entity uses \textsf{waitFrameStarted} to block on the parent's
+\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
+\textsf{Node::waitFrameStarted} to wait for the corresponding
+\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
+update non-critical data before synchronizing with
+\textsf{waitFrameStarted}, or after unlocking using
+\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
+synchronization model.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
+At the end of the frame, two similar sets of synchronization methods are
+used. The first set synchronizes the local execution, while the second
+set synchronizes the global execution.
+
+The local synchronization consists of \textsf{releaseFrameLocal} to
+unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
+unlock. For the default synchronization model, Equalizer uses the task
+method \textsf{frameDrawFinish} which is called on each resource entity
+after the last \textsf{Channel::frameDraw} invocation for this
+frame. Consequently, \textsf{Pipe::frameDrawFinish} calls
+\textsf{Pipe::releaseFrame\-Lo\-cal} to signal that it is done drawing the
+current frame, and \textsf{Node::frameDrawFinish} calls
+\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
+thread until the current frame has been drawn.
+
+\fig{fFrameSync} illustrates the local frame synchronization. By
+removing the calls to \textsf{waitFrameLocal}, the node thread can
+disable the synchronization with the pipe threads, which causes the node
+operations to overlap with the rendering.
+
+The second, global synchronization is used for the frame completion
+during \textsf{Config::finishFrame}, which causes \textsf{frameFinish}
+to be called on all entities, passing the oldest frame number, i.e., frame
+\textsf{current-latency}. The \textsf{frameFinish} task methods have to
+call \textsf{releaseFrame} to signal that the entitity is done with the
+frame. The release causes the parent's \textsf{frameFinish} to be
+invoked, which is so far synchronized internally. Once all
+\textsf{Node::releaseFrame} have been called,
+\textsf{Config::finishFrame} returns.
+
+\begin{figure}[ht!]\center
+  \includegraphics[width=\textwidth]{images/mainloop.pdf}
+  {\caption{\small\label{fMainLoop}Synchronization of frame tasks}}
+\end{figure}
+
+\fig{fMainLoop} outlines the synchronization for the application, node
+and pipe classes for an application node and one render client. Please
+note that \textsf{Config::finishFrame} does block until the current
+frame has been released locally and until the frame \textsf{current -
+  latency} has been released by all nodes. The window and channel
+synchronization are similar and omitted for simplicity.
+
+It is absolutely vital for the execution that \textsf{Node::start\-Fra\-me}
+and \textsf{Node::release\-Fra\-me} are called, respectively. The default
+implementation of the node task methods does take care of that.
+
+Since \textsf{eqPly} multi-buffers all dynamic data, it completely
+removes frame synchronization by:
+
+\begin{itemize}
+\item releasing the local synchronization early in \textsf{Node::frameStart}
+\item not waiting for the node to start the frame by not calling
+  \textsf{Node::waitFrame\-Star\-ted} in \textsf{Pipe::frameStart}
+\item not waiting for the pipe synchronization in
+  \textsf{Node::frameDrawFinish} by not calling \textsf{Pipe::waitFrameLocal}
+\end{itemize}
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+\end{wrapfigure}
 The main idea behind direct send is to parallelize the costly
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
-\end{wrapfigure}
 In direct send compositing, each rendering channel is also responsible
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
+  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
+\end{wrapfigure}
 The frame synchronization through the frame data allows the input frame
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
-  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
-\end{wrapfigure}
 Readback and assemble operations on frames and images are designed to be

### 2008-03-10 Developer/ProgrammingGuide/paper.tex
Update Programming Guide and Quickstart presentation, minor code cleanup
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.11 for Equalizer 0.4+svn\\[\medskipamount]
+  Version 1.1.12 for Equalizer 0.5-RC\\[\medskipamount]
   \today
@@
 \thispagestyle{empty}
-\section*{Equalizer 0.4 Programming Guide}
+\section*{Equalizer 0.5 Programming Guide}
 \subsection*{Contributors}
@@
     1.1.11  & Feb 20, 2008 & Add advanced chapter for thread synchronization\\
+    1.1.12  & Mar 10, 2008 & General review for 0.5 release\\
     \hline
@@
 large-scale graphics clusters. This Programming Guide introduces the
-programming interface, often using the Equalizer \textsf{eqPly} example
+programming interface, often using the Equalizer example \textsf{eqPly}
 as a guideline.
@@
 
-Compiling Equalizer is as simple as running \textsf{make} on Linux or
-building the Equalizer Visual Studio 2005 solution on Windows. On Mac OS
-X 10.4 (Tiger), some prerequisites have to be installed before running
-\textsf{make}, as explained in \textsf{README.Darwin}. Mac OS X 10.5
-Leopard does have all the prerequisites installed by default.
+Compiling Equalizer is as simple as running \textsf{make} on Linux and
+Mac OS X or building the Equalizer Visual Studio 2005 solution on
+Windows. Note that on Mac OS X 10.4 (Tiger), some prerequisites have to
+be installed before running \textsf{make}, as explained in
+\textsf{README.Darwin}.
 
@@
 
-Any parallel rendering application keeps this basic execution model and
+A parallel rendering application uses the same basic execution model and
 extends it by separating the rendering code from the main event
-loop. This rendering coded is then executed in parallel on different
-resources. This model is followed by Equalizer, thus making application
-development as easy as possible.
+loop. The rendering code is then executed in parallel on different
+resources. This model is naturally followed by Equalizer, thus making
+application development as easy as possible.
 
@@
 \textsf{eq::Channel} can be used as a convience function to setup the
-frustum and other OpenGL state. After setting up some additional
-lighting parameter, \textsf{eqHello} rotates the scene and renders the
-quads using immediate mode:
+frustum and other OpenGL state. After setting up some lighting
+parameters, \textsf{eqHello} rotates the scene and renders the quads
+using immediate mode:
 
@@
 task decomposition and result recomposition. It allows the combination
-of the different compound modes in any possible way, which allows the
+of the different compound modes in any possible way, which allows to
 address different bottlenecks in a flexible way.
@@
 transfers, since only color information has to be transmitted. The upper
-limit is the number of pixels of the destination view.
-
-It's disadvantage is that it relies on view frustum culling to reduce
-the amount of data rendered. Depending on the application data
-structure, the overlap of some primitives between individual tiles
+limit is the amount of pixel data for the destination view.
+
+Its disadvantage is that it relies on view frustum culling to reduce the
+amount of data submitted for rendering. Depending on the application
+data structure, the overlap of some primitives between individual tiles
 limits the scalability of this mode, typically to around eight graphics
-cards. Each node has to potentially hold the full database for rendering.
+cards. Each node has to potentially hold the full database for
+rendering.
 
@@
 This mode provides very good scalability, since each rendering unit
-processes a part of the database. This allows to lower the requirements
-on all parts of the rendering pipeline: main memory usage, IO bandwidth,
-GPU memory usage, vertex processing and fill rate.
+processes only a part of the database. This allows to lower the
+requirements on all parts of the rendering pipeline: main memory usage,
+IO bandwidth, GPU memory usage, vertex processing and fill rate.
 
@@
 The application has to partion the database so that the rendering units
-render only part of the database. Some OpenGL features, do not work
+render only part of the database. Some OpenGL features do not work
 correctly (anti-aliasing) or need special attention (transparency).
@@
 The best use of database compounds is to divide the data to a managable
-size, and then to use other decomposition modes to achieve more
+size, and then to use other decomposition modes to achieve further
 scalability. Volume rendering is one of the applications which can
@@
 
-Pixel compounds only work very well for purely fill-limited
-applications, techniques like frustum culling do not reduce the rendered
-data for the source rendering resources.
+Pixel compounds work well for purely fill-limited applications,
+techniques like frustum culling do not reduce the rendered data for the
+source rendering resources.
 
@@
 frame synchronization, as explained in \sref{sNodeFrame}. Some scene
-graphs for example do implement multi-buffered data.
+graphs for example do implement multi-buffered data, and can profit from
+relaxing the frame synchronization.
 
@@
 instances for one resource, e.g. to run multiple render client nodes on
-one computer. For deployment, one node and pipe should be used for each
-computer and graphics card, respectively.
+one computer. For optimal performance during deployment, one node and
+pipe should be used for each computer and graphics card, respectively.
 
@@
 The \textsf{eq::Window} class holds a drawable and an OpenGL
-context. The drawable can be an on-screen window or an
-off-screen\footnote{Off-screen drawables are not implemented yet, but
-  can be created by the application and used with Equalizer} PBuffer or
-framebuffer object (FBO). The window holds window-system-specific
-handles to the drawable and context, e.g., an X11 window \textsf{XID} and
+context. The drawable can be an on-screen window or an off-screen
+PBuffer. Framebuffer object (FBO) support will be implement by later
+version of Equalizer. The window holds window-system-specific handles to
+the drawable and context, e.g., an X11 window \textsf{XID} and
 \textsf{GLXContext} for the glX window system.
@@
 within its parent window. It is the entity executing the actual
-rendering. The channel's viewport might be overwritten when it is
-rendering for another channel during scalable rendering.
+rendering. The channel's viewport is overwritten when it is rendering
+for another channel during scalable rendering.
 
@@
 as source channels. The top-most channel in the tree is the destination
-channel. Only source channels execute rendering task. All channels in a
+channel. Only source channels execute rendering tasks. All channels in a
 compound tree work for the destination channel. The destination channel
@@
 its database, and an \textsf{eye} rendering pass can selectively render
-different stereo passes. Setting one or multiple attributes causes the
-parent's view to be decomposed accordingly. Attributes are cumulative,
-that is, intermediate compound attributes affect and therefore decompose
-the rendering of all their children.
+different stereo passes. A \textsf{pixel} parameter adapts the frustum
+so that the source channel renders every n$th$ line out of m
+lines. Setting one or multiple attributes causes the parent's view to be
+decomposed accordingly. Attributes are cumulative, that is, intermediate
+compound attributes affect and therefore decompose the rendering of all
+their children.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \subfigure[]{\label{fSync}
+    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
+  \subfigure[]{\label{fAsync}
+    \includegraphics[width=.25\textwidth]{images/async}}%
+  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
+\end{wrapfigure}
 When the configuration was successfully initialized, the main rendering
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \subfigure[]{\label{fSync}
-    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
-  \subfigure[]{\label{fAsync}
-    \includegraphics[width=.25\textwidth]{images/async}}%
-  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
-\end{wrapfigure}
 \fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
@@
 
+The config initialization happens in parallel, that is, all config
+initialization tasks are transmitted by the server at once and their
+completion is synchronized afterwards.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The config initialization happens in parallel, that is, all config
-initialization tasks are transmitted by the server at once and their
-completion is synchronized afterwards.
-
 The tasks are executed by the node and pipe threads in parallel. The
@@
 This section discusses some additional important features not covered
-by the previous \textsf{eqPly} section. Where possible code examples
-from the Equalizer distribution are used.
+by the previous \textsf{eqPly} section. Where possible, code examples
+from the Equalizer distribution are used to illustrate one use case of
+the feature.
 
@@
 implementation differs slightly for each operating and window system due
-to conceptual differences in the OS-specific implementation. On the
+to conceptual differences in the specific implementation. On the
 other hand, each application and widget set has its own model on how
@@
 pipe or window with an event handler. This method may be overwritten to
-use a custom event handler, or to not install an event handler at all
-and so disable event handling. Likewise, \textsf{exitEventHandler} is
+use a custom event handler, or to not install an event handler at all in
+order to disable event handling. Likewise, \textsf{exitEventHandler} is
 called to de-initialize event handling.
@@
 \textsf{eq::Window} associated to the event. It then creates a generic
-\textsf{WindowEvent}, which holds important event data in a 
-window system-independent format. The original event is attached to the
-generic window event.
+\textsf{WindowEvent}, which holds important event data in an independent
+format. The original event is attached to the generic window event.
 
@@
 User-defined types start at \textsf{eq::ConfigEvent::USER}, and the
-member variable \textsf{ConfigEvent::user} can be used to store
+member variable \textsf{ConfigEvent::user} can be used to store up to
 \textsf{EQ\_USER\_EVENT\_SIZE}\footnote{currently 32 bytes} bytes. In
@@
 Please note that the default thread synchronization synchronizes all
-\textsf{Channel::frameDraw} operations on a \textbf{single} node with
+\textsf{Channel::frame\-Draw} operations on a \textbf{single} node with
 the node's main thread. The per-node frame synchronization does not
@@
 unlock. For the default synchronization model, Equalizer uses the task
-method \textsf{frameDrawFinish} which is called on each resource entity
+method \textsf{frameDrawFinish} which is called on each resource
 after the last \textsf{Channel::frameDraw} invocation for this
@@
 frame. The release causes the parent's \textsf{frameFinish} to be
-invoked, which is so far synchronized internally. Once all
+invoked, which is synchronized internally. Once all
 \textsf{Node::releaseFrame} have been called,
@@
 Equalizer uses GLEW\footnote{\link{http://glew.sourceforge.net}} for
-OpenGL extension handling, particularly the GLEW MX implementation for
-multi-context support.
+OpenGL extension handling, particularly the GLEW MX implementation
+providing multi-context support.
 
@@
 
-Functions called from another place need to have the macro or function
+Functions called from another place need to define a macro or function
 \textsf{glewGetContext} that returns the pointer to the GLEWContext of
-the appropriate window.
+the appropriate window:
 
@@
 
-This section explains window initialization in detail. It discussed in
+This section explains window initialization in detail. It discusses in
 detail the handling on the different window systems. The entry point for
@@
 The drawable creation method \textsf{configInitAGLDrawable} calls either
-\textsf{configInitAGLFullscreen} or \textsf{configInitAGLWindow}.
+\textsf{configInitAGLFullscreen}, \textsf{configInitAGLWindow} or \textsf{configInitAGLPBuffer}
 
@@
 
-    const bool success = configInitAGLDrawable();
-    if( success && !_xDrawable )
-    {
-        setErrorMessage( "configInitAGLDrawable did set no AGL drawable" );
-        return false;
-    }
-
-    return success;
+    return configInitAGLDrawable();
 }
@@
 \begin{enumerate}
-\item\textsf{getWGLDeviceContext} is used to get a device context which
-  might be needed for window creation. Ordinary windows do not need a
-  device context during creation, but PBuffers do need one. If a context
-  is returned, a function pointer to delete the returned device context
-  is set as well.
+\item\textsf{getWGLPipeDC} is used to get an affinity device context,
+  which might be needed for window creation. If a context is returned, a
+  function pointer to delete the returned device context is set as
+  well. This function might return 0, which is not an error.
+\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
+  window attributes. If no device context is given, it uses the system
+  device context. The chosen pixel format is set on the passed device
+  context.
 \item\textsf{configInitWGLDrawable} creates the drawable. The device
-  context passed to \textsf{configInitWGLDrawable} is not used when
-  creating a window. On success, it sets the window handle.
-\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
-  window attributes. If no device context is given, it uses the device
-  context from this window's window handle. The chosen pixel format is
-  set on the window's device context as well as on the passed device
-  context. The window needs a window handle set before this method is
-  called.
+  context passed to \textsf{configInitWGLDrawable} is used to query the
+  pixel format and is used as the device context for creating a
+  PBuffer. If no device context is given, the display device context is
+  used. On success, it sets the window handle. Setting a window handle
+  also sets the window's device context.
 \item\textsf{createWGLContext} creates an OpenGL rendering context using
-  the given device context. The device context can not be 0. It does not
-  set the window's OpenGL context.
+  the given device context. If no device context is given, the window's
+  device context is used. This function does not set the window's OpenGL
+  context.
 \end{enumerate}
@@
     PFNEQDELETEDCPROC deleteDCProc = 0;
-    HDC dc = getWGLDeviceContext( deleteDCProc );
+    HDC dc = getWGLPipeDC( deleteDCProc );
     EQASSERT( !dc || deleteDCProc );
 
-    if( !configInitWGLDrawable( dc ))
+    int pixelFormat = chooseWGLPixelFormat( dc );
+    if( pixelFormat == 0 )
     {
@@
 
-    if( !_wglWindowHandle )
-    {
-        if( dc )
-            deleteDCProc( dc );
-        setErrorMessage( "configInitWGLDrawable did set no WGL drawable" );
-        return false;
-    }
-
-    int pixelFormat = chooseWGLPixelFormat( dc );
-    if( pixelFormat == 0 )
+    if( !configInitWGLDrawable( dc, pixelFormat ))
     {
@@
 
-    HDC windowDC = GetDC( _wglWindowHandle );
-    HGLRC context = createWGLContext( dc ? dc : windowDC );
-    ReleaseDC( _wglWindowHandle, windowDC );
-
+    if( !_wglDC )
+    {
+        if( dc )
+            deleteDCProc( dc );
+        setErrorMessage( "configInitWGLDrawable did not set a WGL drawable" );
+        return false;
+    }
+
+    HGLRC context = createWGLContext( dc );
     setWGLContext( context );
@@
 frustra are adapted to the viewer's position and orientation, as shown
-in \fig{fTracked}. The transformation origin $\rightarrow$ viewer is set by
+in \fig{fTracked}. The transformation $origin \rightarrow viewer$ is set by
 the application using \textsf{Config::setHeadMatrix}, which is used by
 the server to compute the frustra. The resulting off-axis frustra are
-positioned using the channel's head transformation.
+positioned using the channel's head transformation, which can be
+retrieved using \textsf{Channel::getHeadTransform}.
 
@@
 framebuffer. Equalizer takes care of the network transport of frame
-buffer data between nodes, if needed.
+buffer data between nodes.
 
@@
 In direct send compositing, each rendering channel is also responsible
-for the sort-last composition\footnote{depth-based for polygonal data or
-  sorted-blend for volume data} of one screen-space tile. He receives
+for the sort-last composition of one screen-space tile. He receives
 the framebuffer pixels for his tile from all the other channels. The
@@
 After performing the sort-last compositing, the color information is
-transferred to the destination channel, similarly to an 2D (sort-first)
+transferred to the destination channel, similarly to a 2D (sort-first)
 compound. The amount of pixel data for this part of the compositing
@@
 \fig{fDirectSend} illustrates this algorithm for three channels. The
-Equalizer website contains a presentation explaining this
-algorithm\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}.
+Equalizer website contains a
+presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
+explaining and comparing this algorithm with the binary-swap algorithm.
 
@@
 \item Selection of color and/or depth frame buffer attachments
-\item Restricting the read back area to a part of the rendered area
+\item Restricting the read-back area to a part of the rendered area
 \item Positioning the pixel data correctly on the receiving channels
@@
 programming is needed to achieve good scalability, as \textsf{eVolve} is
-mostly fill-limited and therefore scales nicely in this mode.
+mostly fill-limited and therefore scales nicely in these modes.
 
@@
 through the full texture. Therefore they have to be blen\-ded
-back-to-front in the same way as the slice planes during rendering.
-
+back-to-front in the same way as the slice planes are blended during
+rendering.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 DB compounds have the advantage of scaling any part of the volume
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 For orthographic rendering, determining the compositing order of the
@@
 In order to compute the assembly order, \textsf{eVolve} uses the angle
-between the origin$\rightarrow$slice vector and the near plane, as shown
+between the $origin \rightarrow slice$ vector and the near plane, as shown
 in \fig{fBlendPerspective}. When the angle becomes greater than

### 2008-04-14 Developer/ProgrammingGuide/paper.tex
Programming Guide: Add custom cover page for print version
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation
Programming Guide: prepare 0.5 version
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.1.12 for Equalizer 0.5-RC\\[\medskipamount]
+  Version 1.2 for Equalizer 0.5\\[\medskipamount]
   \today
@@
 \thispagestyle{empty}
-
-%\abstract{
-%  Equalizer is a project to develop software to simplify the creation of
-%  scalable graphics applications and to improve the usability of
-%  multipipe visualization systems.
-%}
+\maketitle
 
@@
     1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
-    1.0.1   & Nov 06, 2007 & Fixed Typo\\
-    1.1     & Nov 12, 2007 & Adapted distributed objects\\
-    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
-    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
-    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
-    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
-    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
-    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
-    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
-    1.1.8   & Jan 15, 2008 & Detailed window initialization based on
-                             latest refactoring\\
-    1.1.9   & Jan 16, 2008 & Detail distributed object usage\\
-    1.1.10  & Jan 28, 2008 & Add creation thread for NodeFactory entities\\
-    1.1.11  & Feb 20, 2008 & Add advanced chapter for thread synchronization\\
-    1.1.12  & Mar 10, 2008 & General review for 0.5 release\\
+    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
     \hline
@@
 programming guide, or other specific problems you encountered, please
-direct them to the \textsf{eq-dev} mailing list\footnote{see
-  \link{http://www.equalizergraphics.com/lists.html}}.
+direct them to the \textsf{eq-dev} mailing
+list\footnote{\link{http://www.equalizergraphics.com/lists.html}}.
 

### 2008-05-02 Developer/ProgrammingGuide/paper.tex
Website: improve statistics docs, add statistics chapter to Programming Guide, improve PG images based on print version
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.2 for Equalizer 0.5\\[\medskipamount]
+  Version 1.3.1 for Equalizer 0.5+svn\\[\medskipamount]
   \today
@@
     1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
+    1.3.1   & May 02, 2008 & Add statistics, improved images based on 1.2 print\\
     \hline
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2em}
+  \vspace{-1.5em}
   \includegraphics[width=.618\textwidth]{images/render.pdf}
   {\caption{\small\label{fRender}Main Rendering Loop}}
-  \vspace{-2em}
+  \vspace{-1.5em}
 \end{wrapfigure}
@@
 
-\subsection{Event Handling}
+\subsection{\label{sEventHandling}Event Handling}
 
@@
   \subfigure[]{\label{fMono}
-    \includegraphics[height=3cm]{images/mono}}\hfil
+    \includegraphics[height=3.5cm]{images/mono}}\hfil
   \subfigure[]{\label{fStereo}
-    \includegraphics[height=3cm]{images/stereo}}\hfil
+    \includegraphics[height=3.5cm]{images/stereo}}\hfil
   \subfigure[]{\label{fTracked}
-    \includegraphics[height=3cm]{images/tracked}}%
+    \includegraphics[height=3.5cm]{images/tracked}}%
   {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked
@@
 
+\subsection{Statistics}
+
+\subsubsection{Statistics Gathering}
+
+Statistics are measured in milliseconds since the configuration was
+initialized. On each node, the global configuration clock is reset at
+the same time by the server during \textsf{Config::init}. Each statistic
+event records the originator's (channel, window or config) unique
+identifier.
+
+Statistics are enabled per entity using an attribute hint. The hint
+determines how precise the gathered statistics are. When set to
+\textsf{fastest}, the per-frame clock is sampled directly when the event
+occurs. When set to \textsf{nicest}, all OpenGL commands will be
+finished before sampling the event. This incurs a performance penalty,
+but gives more correct results. The default setting is fastest in
+release builds, and nicest in debug builds.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/statistics.png}
+  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
+\end{wrapfigure}
+The events are processed by the channel's and window's
+\textsf{processEvent} method. The default implementation sends these
+events to the config, using \textsf{Config::sendEvent}, as explained in
+\sref{sEventHandling}. When the default implementation of
+\textsf{Config::handleEvent} receives the statistics event, it sorts the
+event per frame and per originator. When a frame has been finished, the
+events are pushed to the local (app-)node for visualization.
+
+\fig{fStatistics} shows the visualization of statistics events in an
+overlay\footnote{3D model courtesy Stanford University Computer Graphics
+  Laboratory.}.
+
+\subsubsection{Statistics Overlay}
+
+In debug builds, a statistics overlay using the gathered statistics
+events is rendered directly before \textsf{Window::swapBuffers}. The
+overlay rendering method, \textsf{Channel::drawStatistics}, can be
+called directly to render the overlay. It can also be overwritten by a
+custom implementation.
+
+\begin{figure}[h!t]
+  \includegraphics[width=\textwidth]{images/statisticsDetail.pdf}
+    {\caption{\small\label{fStatisticsDetail}Annotated Detail of the
+        Statistics from \fig{fStatistics}.}}
+\end{figure}
+
+\fig{fStatisticsDetail} shows a detail view of \fig{fStatistics},
+together with some annotations for the overlay. The statistics shown are
+for a two-node 2D compound. The destination channel is on the
+\textsf{appNode} and contributes to the rendering.
+
+This compound executes two \textsf{Channel::frameDraw} tasks, one
+\textsf{Channel::frame\-Read\-back} task on the remote node and one
+\textsf{Channel::frameAssemble} task on the local node.
+
+The X axis is the time. One pixel on the screen corresponds to one
+millisecond. The right-most pixel is the most current time. On the Y
+axis are the rendering entities: channels, windows and the config. The
+upper channel is the local channel since it executes
+\textsf{frameAssemble}, and the lower channel is the remote channel,
+executing {frameReadback}.
+
+The remote machine is a Mac Mini, the local a MacBook Pro. The draw
+times are unbalanced due to the speed difference of the machines, and
+the transmit time is relatively long since 100 MBit Ethernet was used.
+
+In order to facilitate the understanding, subsequent frames alternate in
+brightness. The right-most, current frame is brighter than the
+frame-before-last.
+
+The configuration used has a latency of one frame. Consequently, the
+execution of two frames overlaps. This can be observed in the early
+execution of the remote channel's \textsf{frameDraw}, which starts while
+the local channel is still assembling the previous frame.
+
+The beginning of a frame is marked by a vertical green line, and the end
+of a frame by a vertical gray line. These lines also alternate in
+brightness. The brightness and color matches the event for
+\textsf{Config::startFrame} and \textsf{Config::finishFrame},
+respectively. The event for \textsf{startFrame} is typically not
+visible, since it takes less than one millisecond to execute. If no idle
+processing is done by the application, the event for
+\textsf{finishFrame} occupies a full frame, since the config is blocked
+here waiting for the frame to complete.
+
+In the above example, the local channel finishes drawing the frame
+early, and therefore spends a considerable amount of time waiting for
+the input frame from the remote channel. These wait events, rendered
+red, are a sub-event of the yellow \textsf{frameAssemble} task.
+
+The white \textsf{Window::swapBuffers} task takes also a long time in
+debug builds, since the execution of the swap buffer is locked to the
+vertical retrace of the display, and enforced by \textsf{glFinish}
+before sampling the end time of the event. In release builds, this does
+not happen and all OpenGL commands are pipelined with the execution on
+the CPU.
+
 \if 0

### 2008-05-05 Developer/ProgrammingGuide/paper.tex
Website: Add ohloh project metrics to download site
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.1 for Equalizer 0.5+svn\\[\medskipamount]
+  Version 1.3.2 for Equalizer 0.5+svn\\[\medskipamount]
   \today
@@
 Written by Stefan Eilemann.\\
-Engineering contributions by Maxim Makhinya, Jonas B\"osch and Christian Marten.
+Engineering contributions by Maxim Makhinya, Jonas B\"osch, Christian
+Marten and Patrick Bouchaud.
 
@@
     1.3.1   & May 02, 2008 & Add statistics, improved images based on 1.2 print\\
+    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
     \hline
@@
 
-The \textsf{eqPly} example has a static distributed object to provide
-initial data to all rendering nodes, as well as a versioned object to
-provide frame-specific data such as the camera position to the
-rendering methods.
-
-\subsubsection{Object Usage for Rendering}
-
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/objects.pdf}
-  {\caption{\small\label{fObjects}Distributed Data Handling}}
-\end{wrapfigure}
+The \textsf{eqPly} example uses static distributed objects to provide
+initial data and the model to all rendering nodes, as well as a
+versioned object to provide frame-specific data such as the camera
+position to the rendering methods.
+
+\subsubsection{Common Usage}
+
 Distributed objects are addressed using session-unique identifiers,
@@
 passed by Equalizer to all \textsf{configInit} task methods. Normally
-this initial object is mapped by the \textsf{eq::Node}, and contains
+this initial object is mapped by the render clients, and it contains
 identifiers to other shared data objects.
@@
 
-Applications distributing their scene graph, rather then relying on a
-shared filesystem like \textsf{eqPly}, use the frame data also as an
-entry point to their scene graph data structure. \fig{fObjects} shows
-one possible implementation, where the identifier and version of the
-scene graph root are transported using the frame data. The scene graph
-root then serializes and de-serializes his immediate children by
-transferring their identifier and current version.
-
-The objects are still created by the application, and then registered or
-mapped with the session in order to distribute them. When mapping
-objects in a hierarchical data structure, their type often has to be
-known in order to create them. Equalizer does not currently provide
-object typing, this has to be done by the application, either implicitly
-in the current implementation context, or by transferring a type
-identifier. In \textsf{eqPly}, object typing is implicit since it is
-well-defined which object is mapped in which task method.
-
 \subsubsection{Change Handling}
@@
 \textsf{Object::getChangeType} during the registration of the master
-version of a distributed object . The change type determines the memory
+version of a distributed object. The change type determines the memory
 footprint and the contract for calling the serialization methods. The
@@
 {
-    os << _frameDataID << _windowSystem << _useVBOs << _useGLSL << _filename;
+    os << _frameDataID << _modelID << _windowSystem << _useVBOs << _useGLSL
+       << _invFaces ;
 }
@@
 {
-    is >> _frameDataID >> _windowSystem >> _useVBOs >> _useGLSL >> _filename;
+    is >> _frameDataID >> _modelID >> _windowSystem >> _useVBOs >> _useGLSL
+       >> _invFaces;
 
@@
 
+\subsubsection{\label{sSceneData}Scene Data}
+
+Some applications might rely on a shared filesystem to access the data,
+for example when out-of-core algorithms are used. Other applications
+prefer to load the data only on the application process, and use
+distributed objects to synchronize the scene data with the render
+clients.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
+  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
+\end{wrapfigure}
+\textsf{eqPly} uses static distributed objects to distribute the model
+loaded by the application. This model can be easily extended to
+versioned objects, in order to support dynamic data changes.
+
+The kD-tree data structure and rendering code for the model is strongly
+separated from Equalizer, and kept in the separate namespace
+\textsf{mesh}. It could also be used in a GLUT application. To keep this
+separation, an external 'mirror' hierarchy is constructed around the
+tree. This hierarchy of \textsf{VertexBufferDist} nodes is responsible
+for cloning the model data on the remote render clients.
+
+The identifer of the root object of this distributed hierarchy is passed
+as part of the \textsf{InitData}, and used during remote node
+initialization to map the model. \fig{fModelDist} shows the UML
+hierarchy of the model and distribution classes.
+
+Each \textsf{VertexBufferDist} object has one node of the model's data
+tree. It is serializing the data for this node. Furthermore, it mirrors
+the kD-tree by having a \textsf{VertexBufferDist} child for each child
+of its corresponding tree node. During serialization, the identifier of
+these children is sent to the remote nodes, which reconstruct the mirror
+distribution hierarchy and model data tree based on this data.
+
+The serialization function \textsf{getInstanceData} sends all the data
+needed to reconstruct the model tree: the object identifiers of its
+children, vertex data for the tree root and vertex indices for the leaf
+nodes, as well as the bounding sphere and database range of each
+node. The deserialization function \textsf{applyInstanceData} retrieves
+the data in multiple steps, and constructs the model tree on the fly
+based on this information. It is omitted here for brevity:
+
+{\footnotesize\begin{lstlisting}
+void VertexBufferDist::getInstanceData( eqNet::DataOStream& os )
+{
+    EQASSERT( _node );
+    os << _isRoot;
+
+    if( _left && _right )
+    {
+        os << _left->getID() << _right->getID();
+
+        if( _isRoot )
+        {
+            EQASSERT( _root );
+            const mesh::VertexBufferData& data = _root->_data;
+
+            os << data.vertices << data.colors << data.normals << data.indices;
+        }
+    }
+    else
+    {
+        os << EQ_ID_INVALID << EQ_ID_INVALID;
+
+        EQASSERT( dynamic_cast< const mesh::VertexBufferLeaf* >( _node ));
+        const mesh::VertexBufferLeaf* leaf = 
+            static_cast< const mesh::VertexBufferLeaf* >( _node );
+
+        os << leaf->_vertexStart << leaf->_vertexLength << leaf->_indexStart
+           << leaf->_indexLength;
+    }
+
+    os << _node->_boundingSphere << _node->_range;
+}
+\end{lstlisting}}%>>
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/objects.pdf}
+  {\caption{\small\label{fObjects}Distributed Data Handling}}
+\end{wrapfigure}
+Applications distributing a dynamic scene graph use the frame data
+instead of the init data as the entry point to their scene graph data
+structure. \fig{fObjects} shows one possible implementation, where the
+identifier and version of the scene graph root are transported using the
+frame data. The scene graph root then serializes and de-serializes his
+immediate children by transferring their identifier and current version,
+similar to the static distribution done by \textsf{eqPly}.
+
+The objects are still created by the application, and then registered or
+mapped with the session in order to distribute them. When mapping
+objects in a hierarchical data structure, their type often has to be
+known in order to create them. Equalizer does not currently provide
+object typing, this has to be done by the application, either implicitly
+in the current implementation context, or by transferring a type
+identifier. In \textsf{eqPly}, object typing is implicit since it is
+well-defined which object is mapped in which context.
+
 \subsection{Config}
@@
 problem to parallel rendering. Tracking device support has already been
-solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
+solved by a number of implementations\footnote{VRCO Trackd VRPN, etc.},
 which can easily be integrated with Equalizer. The example code in
@@
 also have a node class, which is handled in exactly the same way as the
-render client nodes.
-
-During node initialization the init data is mapped to a local instance
-using the identifier passed from \textsf{Config::init}. The model is
-loaded based on the filename in the initialization data. No pipe, window
-or channel tasks methods are executed before \textsf{Node::configInit}
-has returned.
+render client nodes. The application and render clients might use a
+different node factory, instantiating a different type of \textsf{eq::Node}.
+
+During node initialization the static, per-config data is mapped to a
+local instance using the identifier passed from
+\textsf{Config::init}. No pipe, window or channel tasks methods are
+executed before \textsf{Node::configInit} has returned.
 
@@
 {
-    eq::Config* config = getConfig();
-    const bool  mapped = config->mapObject( &_initData, initID );
+    if( !eq::Node::configInit( initID ))
+        return false;
+
+    Config*    config = static_cast< Config* >( getConfig( ));
+    return config->mapData( initID );
+}
+\end{lstlisting}}
+
+The actual mapping of the static data is done by the config. The config
+first retrieves the distributed \textsf{InitData}. The object is
+directly unmapped since it is static, and the data has be retrieved.
+
+The init data contains, among other things, the identifier of the
+root of the model's kD-tree. Using this identifier, the model is mapped
+to the node, as explained in \sref{sSceneData}:
+
+{\footnotesize\begin{lstlisting}
+bool Config::mapData( const uint32_t initDataID )
+{
+    if( _initData.getID() == EQ_ID_INVALID )
+    {
+        const bool mapped = mapObject( &_initData, initDataID );
         EQASSERT( mapped );
-
-    const string& filename = _initData.getFilename();
-    EQINFO << "Loading model " << filename << endl;
-
-    _model = new Model();
-    if ( !_model->readFromFile( filename.c_str() ) )
-    {
-        EQWARN << "Can't load model: " << filename << endl;
-        delete _model;
-        _model = 0;
-    }
-
-    return eq::Node::configInit( initID );
+        unmapObject( &_initData ); // data was retrieved, unmap immediately
+    }
+    else  // appNode, _initData is registered already
+        EQASSERT( _initData.getID() == initDataID );
+
+    const uint32_t modelID = _initData.getModelID();
+    if( modelID == EQ_ID_INVALID ) // no model loaded by application
+        return true;
+
+    if( !_model )
+        _model = ModelDist::mapModel( this, modelID );
+
+    return (_model != 0);
+}
+\end{lstlisting}}
+
+Since the model data is static, the mirror tree for data distribution
+is only created temporarily to retrieve the data. This is done in a
+static function of the \textsf{VertexBufferDist} class. One instance of
+the distribution class is mapped to the model's identifier. During
+application of the instance data, it will map its children recursively:
+
+{\footnotesize\begin{lstlisting}
+mesh::VertexBufferRoot* VertexBufferDist::mapModel(eqNet::Session* session,
+                                                   const uint32_t modelID )
+{
+    VertexBufferDist modelDist( 0, 0 );
+
+    if( !session->mapObject( &modelDist, modelID ))
+    {
+        EQWARN << "Mapping of model failed" << endl;
+        return 0;
+    }
+
+    modelDist._unmapTree(); // No longer needed
+
+    return const_cast< mesh::VertexBufferRoot* >( modelDist._root );
 }
 \end{lstlisting}}%>>
-
-The node config exit function deletes the loaded model and unmaps the
-initialization data:
-
-{\footnotesize\begin{lstlisting}
-bool Node::configExit()
-{
-    delete _model;
-    _model = NULL;
-
-    eq::Config* config = getConfig();
-    config->unmapObject( &_initData );
-
-    return eq::Node::configExit();
-}
-\end{lstlisting}}
 
@@
 {
-    const Node*     node        = static_cast<Node*>( getNode( ));
-    const InitData& initData    = node->getInitData();
+    const Config*     config    = static_cast<Config*>( getConfig( ));
+    const InitData& initData    = config->getInitData();
     const uint32_t  frameDataID = initData.getFrameDataID();
@@
 {
-    const Node*            node     = static_cast<Node*>( getNode( ));
-    const InitData&        initData = node->getInitData();
+    const Config*            config = static_cast<Config*>( getConfig( ));
+    const InitData&        initData = config->getInitData();
     const eq::WindowSystem ws       = initData.getWindowSystem();
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/windowInit.pdf}
+  {\caption{\small\label{fWindowInit}Window Initialization}}
+\end{wrapfigure}
 \fig{fWindowInit} shows a flow chart of the window initialization. The
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
-  {\caption{\small\label{fWindowInit}Window Initialization}}
-\end{wrapfigure}
 The window-system specific initialization takes into account various
@@
 
-        const Node*     node     = static_cast< const Node* >( getNode( ));
-        const InitData& initData = node->getInitData();
+        const Config*   config   = static_cast< const Config* >( getConfig( ));
+        const InitData& initData = config->getInitData();
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/DB.png}
-  {\caption{\small\label{fDB}Destination view of an DB compound}}
-\end{wrapfigure}
 After the basic view setup, a directional light is configured, and the
@@
 
-    Node*            node  = (Node*)getNode();
-    const Model*     model = node->getModel();
+    Config*          config = static_cast< Config* >( getConfig( ));
+    const Model*     model  = config->getModel();
     const eq::Range& range  = getRange();
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/DB.png}
+  {\caption{\small\label{fDB}Destination view of a DB compound}}
+\end{wrapfigure}
 In order to draw the model, a helper class for view frustum culling is
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-1.5em}
   \includegraphics[width=.618\textwidth]{images/render.pdf}
   {\caption{\small\label{fRender}Main Rendering Loop}}
-  \vspace{-1.5em}
 \end{wrapfigure}
@@
 
+During this initialization, Equalizer creates and manages two threads
+for communication, the receiver thread and the command thread. Normally
+no application code is executed from these two threads.
+
+The receiver thread manages the network connections to other nodes and
+receives data. It dispatches the received data either to the application
+threads, or to the command thread.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-During this initialization, Equalizer creates and manages two threads
-for communication, the receiver thread and the command thread. Normally
-no application code is executed from these two threads.
-
-The receiver thread manages the network connections to other nodes and
-receives data. It dispatches the received data either to the application
-threads, or to the command thread.
-
 The command thread processes Equalizer-related request from other nodes,
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
 Please note that the default thread synchronization synchronizes all
@@
 \textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
-\textsf{Node::waitFrameStarted} to wait for the corresponding
+\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
 \textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
 At the end of the frame, two similar sets of synchronization methods are
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
-      Volume Rendering}}
-\end{wrapfigure}
 The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
+      Volume Rendering}}
+\end{wrapfigure}
 The full power of scalable volume rendering is however in DB (sort-last)
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 DB compounds have the advantage of scaling any part of the volume
@@
 visualize with any other approach.
+
+For recomposition, the 2D frame buffer contents are blended together to
+form a seamless picture. For correct blending, the frames are ordered in
+the same back-to-front order as the slices used for rendering, and use the
+same blending parameters. Simplified, the frame buffer images are
+`thick' slices which are `rendered' by writing their content to the
+destination frame buffer using the correct order.
 
@@
 \end{figure}
-For recomposition, the 2D frame buffer contents are blended together to
-form a seamless picture. For correct blending, the frames are ordered in
-the same back-to-front order as the slices used for rendering, and use the
-same blending parameters. Simplified, the frame buffer images are
-`thick' slices which are `rendered' by writing their content to the
-destination frame buffer using the correct order.
-
 For orthographic rendering, determining the compositing order of the
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/statistics.png}
+  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
+\end{wrapfigure}
 Statistics are enabled per entity using an attribute hint. The hint
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/statistics.png}
-  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
-\end{wrapfigure}
 The events are processed by the channel's and window's

### 2008-05-10 Developer/ProgrammingGuide/paper.tex
Website: Start file format section in Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-\usepackage{tabularx,graphicx,a4,listings,wrapfig,subfigure,textcomp}
+\usepackage{tabularx,longtable,graphicx,a4,listings,wrapfig,subfigure,textcomp}
 
@@
 
-\subsection*{Previous Page}
-
-The images on the front page show: a terrain rendering application on a
-six-node display wall [top left], the \textsf{eqPly} polygonal renderer
-in a three-sided CAVE [middle], two volume rendering results from
-eVolve\footnote{Data sets courtesy of General Electric, USA and AVS,
-  USA} [bottom left and middle] and a six-node sort-last database
-decomposition with parallel direct-send recomposition\footnote{Data set
-  courtesy of Stanford University Computer Graphics Laboratory} [bottom
-right].
+\subsection*{Front Page}
+
+The images on the front page show the following Equalizer applications:
+a terrain rendering application on a six-node display wall [top left], a
+flow visualization application for climate research\footnote{Image
+  courtesy of Computer Graphics and Multimedia Systems, University of
+  Siegen}, the \textsf{eVolve} volume renderer\footnote{Data sets
+  courtesy of General Electric, USA and AVS, USA} [bottom left] the
+\textsf{eqPly} polygonal renderer in a three-sided CAVE [bottom center],
+and a six-node sort-last database decomposition with parallel
+direct-send recomposition\footnote{Data set courtesy of Stanford
+  University Computer Graphics Laboratory} [bottom right].
 
@@
     1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
+    1.3.3   & May 10, 2008 & File Format Appendix [incomplete]\\
     \hline
@@
 hierarchical representation of the rendering decomposition and
-recomposition across the resources. It is explained in \sref{compounds}.
+recomposition across the resources. It is explained in \sref{sCompounds}.
 
@@
 
-\subsection{\label{compounds}Compounds}
+\subsection{\label{sCompounds}Compounds}
 
@@
 
+\newpage
+\appendix
+\section{File Format}
+
+The currently used file format is the one-to-one representation of the
+server's internal data structures. Its purpose is intermediate, that
+is, it will gradually be replaced by automatic resource dedection and
+configuration. However the core scalability engine will always use a
+similar structure as currently exposed by the file format.
+
+The file format represents an ascii deserialization of the
+server. Streaming an \textsf{eqs::Server} to an \textsf{eqBase::Log}
+ostream produces a valid configuration file. Likewise, loading a
+configuration file produces an \textsf{eqs::Server}.
+
+The file format uses the same syntactical structure as VRML. If your
+editor supports syntax highlighting and formatting for VRML, you can use
+this mode for \textsf{.eqc} files.
+
+The configuration file consist of an optional global section and a
+server configuration. The global section defines default values for
+various attributes. The server section represents an
+\textsf{eqs::Server}.
+
+\subsection{Global Section}
+
+The global section defines default values for attributes used by the
+individual entities of the server section. The naming convention for
+attributes is:
+
+{\footnotesize\begin{lstlisting}
+EQ_<ENTITY>_<DATATYPE>ATTR_<ATTR_NAME>
+\end{lstlisting}}
+
+The entity is the capitalized name of the corresponding section later in
+the configuration file: connection, config, pipe, window, channel or
+compound. The connection is used by the server and nodes.
+
+The datatype is one capital letter for the type of the attribute's
+value: \textsf{S} for strings, \textsf{C} for a character, \textsf{I}
+for an integer and \textsf{F} for floating-point values. Enumation
+values are handled as integers. Strings have always to be surrounded by
+double quotes '\textsf{''}'. A character has to surrounded by single
+quotes '\textsf{'}'.
+
+The attribute name is the capitalized name of the entities attribute, as
+discussed in the following sections.
+
+Global attribute values have useful default parameters, which can be
+overriden with an environment variable of the same name. For enumeration
+values the corresponding integer value has to be used. The global values
+in the config file override these defaults, and are in turn overriden by
+the corresponding attributes sections of the associated entities.
+
+The globals section starts with the token \textsf{global} and an open
+curly brace '\textsf{\{}', and is terminated with a closing curly brace
+'\textsf{\}}'. Within the braces, globals are set using the attribute's
+name followed by its value. The following attributes are available:
+
+\begin{center}
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONNECTION\_SATTR\_HOSTNAME\\
+  \textbf{Value} & string\\
+  \textbf{Default} & ''localhost''\\
+  \textbf{Description} & The hostname or IP address used to connect the
+  server or node. When used for the server, the listening port of the
+  server is bound to this address. When used for a node, the server
+  first tries to connect to the render client node using this hostname,
+  and then tries to launch the render client executable on this host.\\
+  \hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONNECTION\_SATTR\_LAUNCH\_COMMAND\\
+  \textbf{Value} & string\\
+  \textbf{Default} & ssh -n \%h \%c \textgreater\& \%h.\%n.log [POSIX]\\
+                   & ssh -n \%h \%c [WIN32]\\
+  \textbf{Description} & The command used by the server to launch nodes
+  which could not be connected. The launch command is executed from a
+  process forked of the server process. The \% tokens are replaced by the
+  server at runtime with concrete data: \%h is replaced by the hostname,
+  \%c by the command to launch, including command line arguments and \%n
+  by a node-unique identifier. Each command line argument are surrounded
+  by launch command quotes.\\
+  \hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONNECTION\_CATTR\_LAUNCH\_COMMAND\_QUOTE\\
+  \textbf{Value} & character\\
+  \textbf{Default} & ' [POSIX]\\
+                   & '' [WIN32]\\
+  \textbf{Description} & The server uses command line arguments to
+  launch render client nodes correctly. This arguments might contain
+  white spaces, and therefore have to be surrounded by quotes to
+  identify their borders. Certain launch commands or shells use
+  different conventions to separate command line arguments.\\
+  \hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONNECTION\_IATTR\_TYPE\\
+  \textbf{Value} & TCPIP \textbar \ SDP\\
+  \textbf{Default} & TCPIP\\
+  \textbf{Description} & The protocol for connections. SDP
+  programmatically selects the socket direct protocol (AF\_INET\_SDP)
+  provided by most InfiniBand protocol stacks.\\
+  \hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONNECTION\_IATTR\_TCPIP\_PORT\\
+  \textbf{Value} & unsigned\\
+  \textbf{Default} & 0\\
+  \textbf{Description} & The listening port used by the server or
+  node. For nodes, the port can be used to contact pre-started, resident
+  render client nodes or to use a specific port for the node. If 0 is
+  specified, a random port is chosen. Note that a server with no
+  connections automatically creates a default connection using the
+  server's default port.\\
+  \hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_CONNECTION\_IATTR\_LAUNCH\_TIMEOUT\\
+\textbf{Value} & unsigned\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_CONFIG\_FATTR\_EYE\_BASE\\
+\textbf{Value} & float\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_PIPE\_IATTR\_HINT\_THREAD\\
+\textbf{Value} & OFF \textbar \ ON\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_STEREO\\
+\textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_DOUBLEBUFFER\\
+\textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_DECORATION\\
+\textbf{Value} & OFF \textbar \ ON\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_FULLSCREEN\\
+\textbf{Value} & OFF \textbar \ ON\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_SWAPSYNC\\
+\textbf{Value} & OFF \textbar \ ON // AGL, WGL only\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_DRAWABLE\\
+\textbf{Value} & window \textbar \ pbuffer\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_STATISTICS\\
+\textbf{Value} & off \textbar \ fastest [on] \textbar \ nicest\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_COLOR\\
+\textbf{Value} & unsigned\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_ALPHA\\
+\textbf{Value} & unsigned\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_DEPTH\\
+\textbf{Value} & unsigned\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_STENCIL\\
+\textbf{Value} & unsigned\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_CHANNEL\_IATTR\_HINT\_STATISTICS\\
+\textbf{Value} & OFF \textbar \ FASTEST [ ON ] \textbar \ NICEST\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
+\textbf{Value} & QUAD \textbar \ ANAGLYPH\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_COMPOUND\_IATTR\_STEREO\_ANAGLYPH\_LEFT\_MASK\\
+\textbf{Value} & [ RED GREEN BLUE ]\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_COMPOUND\_IATTR\_STEREO\_ANAGLYPH\_RIGHT\_MASK\\
+\textbf{Value} & [ RED GREEN BLUE ]\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+
+\begin{tabularx}{\textwidth}{|l|X|}
+\hline
+\textbf{Name} & EQ\_COMPOUND\_IATTR\_UPDATE\_FOV\\
+\textbf{Value} & HORIZONTAL [ON] \textbar \ VERTICAL \textbar \ OFF\\
+\textbf{Default} & \\
+\textbf{Description} & \\
+\hline
+\end{tabularx}\\[\smallskipamount]
+\end{center}
+
+\subsection{Server Section}
+
+{\footnotesize\begin{lstlisting}
+server
+{
+    connection // 0-n times, listening connections of the server
+    {
+        type       TCPIP | SDP
+        TCPIP_port unsigned
+        hostname   string
+    }
+
+    config {...} // 1-n times, currently only the first one is used by the server
+}
+\end{lstlisting}}
+
+For an introduction on the concepts of nodes, pipes, windows, channels
+and compounds please refer to \sref{sConfig} and \sref{sCompounds}.
+
+\subsubsection{Config Section}
+
+{\footnotesize\begin{lstlisting}
+config
+{
+    latency int    // # of frames nodes may fall behind application
+    attributes
+    {
+         eye_base float // distance between left and right eye
+    }
+
+    (node|appNode) {...}// 1-n times, a system in the cluster
+                   // 0|1 appNode: launches render thread within app process
+    compound {...} // 1-n times
+}
+\end{lstlisting}}
+
+\subsubsection{Node Section}
+
+{\footnotesize\begin{lstlisting}
+{
+    name     string
+    connection // 0-n times, possible connections to this node
+    {
+        type          TCPIP | SDP
+        TCPIP_port    unsigned
+        hostname      string
+        command       string      // render client launch command
+        command_quote 'character' // launch command argument quote char
+        timeout       unsigned    // timeout in milliseconds for launch
+    }
+    pipe {...} // 1-n times
+}
+\end{lstlisting}}
+
+\subsubsection{Pipe Section}
+
+{\footnotesize\begin{lstlisting}
+pipe
+{
+    name     string
+    port     unsigned     // X server number or ignored
+    device   unsigned     // graphics adapter number
+    viewport [ viewport ] // default: autodetect
+    attributes
+    {
+        hint_thread   OFF | ON   // default ON
+    }
+
+    window {...} // 1-n times
+}
+\end{lstlisting}}
+
+\subsubsection{Window Section}
+
+{\footnotesize\begin{lstlisting}
+window
+{
+    name     string
+    viewport [ viewport ] // wrt pipe, default full screen
+
+    attributes
+    {
+        hint_stereo         OFF | ON | AUTO
+        hint_doublebuffer   OFF | ON | AUTO
+        hint_decoration     OFF | ON
+        hint_fullscreen     OFF | ON
+        hint_swapsync       OFF | ON         // AGL, WGL only
+        hint_drawable       window | pbuffer
+        hint_statistics     off | fastest [on] | nicest
+        planes_color        unsigned
+        planes_alpha        unsigned
+        planes_depth        unsigned
+        planes_stencil      unsigned
+    }
+
+    channel {...} // 1-n times
+}
+\end{lstlisting}}
+
+\subsubsection{Channel Section}
+
+{\footnotesize\begin{lstlisting}
+channel
+{
+    name     string
+    viewport [ viewport ] //wrt window, default full window
+
+    attributes
+    {
+        hint_statistics     OFF | FASTEST [ON] | NICEST
+    }
+}
+\end{lstlisting}}
+
+\subsubsection{Compound Section}
+
+{\footnotesize\begin{lstlisting}
+compound
+{
+    name     string
+    channel  string   // where the compound's tasks are executed
+    task     [ CLEAR DRAW READBACK ASSEMBLE ] // CULL later
+    buffer   [ COLOR DEPTH ]         // default COLOR
+    viewport [ viewport ]            // wrt parent compound, sort-first
+    range    [ float float ]         // DB-range for sort-last
+    pixel    [ int int ]             // pixel decomposition (step size)
+    eye      [ CYCLOP LEFT RIGHT ]   // monoscopic or stereo view
+    attributes
+    {
+        stereo_mode                  QUAD | ANAGLYPH    // default QUAD
+        // enable visible color components
+        stereo_anaglyph_left_mask    [ RED GREEN BLUE ] // default red
+        stereo_anaglyph_right_mask   [ RED GREEN BLUE ] // df green blue
+        update_FOV                   HORIZONTAL [ON] | VERTICAL | OFF
+        assemble                     GPU | CPU          // default GPU
+    }
+
+    wall                             // frustum description
+    {
+        bottom_left  [ float float float ]
+        bottom_right [ float float float ]
+        top_left     [ float float float ]
+    }
+
+    child-compounds
+
+    swapbarrier  // compounds with the same barriername sync swap
+    {
+        name string
+    }
+    outputframe
+    {
+        name string
+    }
+    inputframe
+    {
+        name string // corresponding output frame
+    }
+}
+\end{lstlisting}}
+
 \if 0

### 2008-05-15 Developer/ProgrammingGuide/paper.tex
Website: Update Equalizer presentation
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.2 for Equalizer 0.5+svn\\[\medskipamount]
+  Version 1.3.3 for Equalizer 0.5+svn\\[\medskipamount]
   \today
@@
     1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
-    1.3.3   & May 10, 2008 & File Format Appendix [incomplete]\\
+    1.3.3   & May 10-13, 2008 & File Format Appendix [incomplete]\\
     \hline
@@
   \textbf{Default} & ''localhost''\\
-  \textbf{Description} & The hostname or IP address used to connect the
+  \textbf{Details} & The hostname or IP address used to connect the
   server or node. When used for the server, the listening port of the
@@
   \hline
-\end{tabularx}\\[\smallskipamount]
+\end{tabularx}\\\vfill
 
@@
                    & ssh -n \%h \%c [WIN32]\\
-  \textbf{Description} & The command used by the server to launch nodes
+  \textbf{Details} & The command used by the server to launch nodes
   which could not be connected. The launch command is executed from a
@@
                    & '' [WIN32]\\
-  \textbf{Description} & The server uses command line arguments to
+  \textbf{Details} & The server uses command line arguments to
   launch render client nodes correctly. This arguments might contain
@@
   \textbf{Default} & TCPIP\\
-  \textbf{Description} & The protocol for connections. SDP
+  \textbf{Details} & The protocol for connections. SDP
   programmatically selects the socket direct protocol (AF\_INET\_SDP)
@@
   \textbf{Default} & 0\\
-  \textbf{Description} & The listening port used by the server or
+  \textbf{Details} & The listening port used by the server or
   node. For nodes, the port can be used to contact pre-started, resident
@@
   \textbf{Value} & unsigned\\
-\textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+  \textbf{Default} & 60'000\\
+  \textbf{Details} & Defines the timeout in milliseconds to wait for
+  an auto-launched node. If the render client process did not contact
+  the server within that time, the node is considered to be unreachable
+  and the initialization of the configuration fails.\\
+  \hline
+\end{tabularx}\\\vfill
 
@@
   \textbf{Value} & float\\
-\textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+  \textbf{Default} & 0.05\\
+  \textbf{Details} & The distance in meters between the left and the
+  right eye, i.e., the eye separation. The eye base influences the
+  frustum during stereo rendering. See \sref{sTracking} for details.\\
+  \hline
+\end{tabularx}\\\vfill
 
@@
   \textbf{Value} & OFF \textbar \ ON\\
-\textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+  \textbf{Default} & ON\\
+  \textbf{Details} & Determines if all task methods for a pipe and its
+  children are executed from a separate operating system thread
+  (default) or from the node main thread. Non-threaded pipes have
+  certain performance
+  limitations and should only be used if really necessary.\\
+  \hline
+\end{tabularx}\\\vfill
 
@@
   \textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\\
-\textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+  \textbf{Default} & AUTO\\
+  \textbf{Details} & Determines if the window selects a quad-buffered
+  stereo visual. When set to AUTO, the window initialization methods try
+  to allocate a stereo visual, but fall back to a mono visual if
+  allocation fails.\\
+  \hline
+\end{tabularx}\\\vfill
 
@@
   \textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\\
-\textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+  \textbf{Default} & AUTO\\
+  \textbf{Details} & Determines if the window selects a double-buffered
+  stereo visual. When set to AUTO, the window initialization methods try
+  to allocate a double-buffered visual, but fall back to a
+  single-buffered visual if allocation fails.\\
+  \hline
+\end{tabularx}\\\vfill
 
@@
   \textbf{Value} & OFF \textbar \ ON\\
-\textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+  \textbf{Default} & ON\\
+  \textbf{Details} & When set to OFF, window borders and other
+  decorations are disabled, and typically the window can not be moved or
+  resized. This option is useful for source windows during
+  decomposition. The implementation is window-system specific.\\
+  \hline
+\end{tabularx}\\\vfill
 
@@
   \textbf{Value} & OFF \textbar \ ON\\
-\textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+  \textbf{Default} & OFF\\
+  \textbf{Details} & When set to ON, the window displays in
+  fullscreen. This option forces window decorations to be OFF. The
+  implementation is window-system specific.\\
+  \hline
+\end{tabularx}\\\vfill
 
@@
   \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_SWAPSYNC\\
-\textbf{Value} & OFF \textbar \ ON // AGL, WGL only\\
-\textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+  \textbf{Value} & OFF \textbar \ ON\\
+  \textbf{Default} & ON\\
+  \textbf{Details} & Determines if the buffer swap is synchronized with
+  the vertical retrace of the display. This option is currently not
+  implemented for GLX. For WGL, the WGL\_EXT\_swap\_control extension is
+  required. For optimal performance, set swap synchronization to OFF for
+  source-only windows.\\
+  \hline
+\end{tabularx}\\\vfill
 
@@
 \textbf{Value} & window \textbar \ pbuffer\\
-\textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+\textbf{Default} & window\\
+\textbf{Details} & Selects the window's drawable type.\\
+\hline
+\end{tabularx}\\\vfill
 
@@
 \textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+\textbf{Details} & \\
+\hline
+\end{tabularx}\\\vfill
 
@@
 \textbf{Default} & \\
-\textbf{Description} & \\
-\hline
-\end{tabularx}\\[\smallskipamount]
+\textbf{Details} & \\
+\hline
+\end{tabularx}\\
 \end{center}

### 2008-05-16 Developer/ProgrammingGuide/paper.tex
Fill in more file format content in the Programming Guide.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
-    1.3.3   & May 10-13, 2008 & File Format Appendix [incomplete]\\
+    1.3.3   & May 10-16, 2008 & File Format Appendix [incomplete]\\
     \hline
@@
 the rendering tasks. One channel might be used by multiple
-compounds. Unused channels are not instantiated during
-initialization. The rendering tasks for the channels are computed by the
-server and send to the appropriate render clients.
+compounds. Unused channels, windows, pipes and nodes are not
+instantiated during config initialization. The rendering tasks for the
+channels are computed by the server and send to the appropriate render
+client nodes at the beginning of each frame.
 
@@
 as source channels. The top-most channel in the tree is the destination
-channel. Only source channels execute rendering tasks. All channels in a
+channel. Only source channels execute the draw task. All channels in a
 compound tree work for the destination channel. The destination channel
@@
 the output frame. This composition process is extensively described in
-\sref{sCompositing}.
+\sref{sCompositing}. Frame names are only valid within the compound
+tree, that is, an output frame from one compound tree can not be used as
+an input frame of another compound tree.
+
+\subsubsection{Synchronization - Swapbarrier}
+Compounds may have a swapbarrier to synchronize the buffer swap of
+multiple channels. Windows with a swapbarrier of the same name enter a
+barrier before executing the swap buffer task. Before entering the
+barrier, \textsf{Window::finish} is called to ensure that all OpenGL
+commands have been executed.
+
+Note that this software swap barrier is not accurate enough for
+edge-blended projection system. Such installations typically use a
+hardware synchronization, e.g., nVidia G-Sync cards. It is however
+sufficient for display walls made out of LCD's.
+
+Swapbarrier names are only valid within the compound tree, that is, a
+compound from one compound tree can not be synchronized with a compound
+from another compound tree.
 
@@
 
+When the configuration was successfully initialized, the main rendering
+loop is executed. It runs until the user exits the
+configuration, or when a maximum number of frames has been rendered,
+specified by a command-line argument. The latter is useful for
+benchmarks. The \textsf{Clock} is reused for measuring the overall
+performance. A new frame is started using \textsf{Config::startFrame}
+and a frame is finished using \textsf{Config::finishFrame}.
+
+When the frame is started, the server computes all rendering tasks and
+sends them to the appropriate render client nodes. The render client
+nodes dispatch the tasks to the correct node or pipe thread, where they
+are executed in order of arrival.
+
+\textsf{Config::finishFrame} synchronizes on the completion of the frame
+\textsf{current - latency}. The latency is specified in the
+configuration file, and allows several outstanding frames. This allows
+overlapping execution in the node and pipe threads and minimizes idle
+times.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-When the configuration was successfully initialized, the main rendering
-loop is executed. It runs until the user exits the
-configuration, or when a maximum number of frames has been rendered,
-specified by a command-line argument. The latter is useful for
-benchmarks. The \textsf{Clock} is reused for measuring the overall
-performance. A new frame is started using \textsf{Config::startFrame}
-and a frame is finished using \textsf{Config::finishFrame}.
-
-When the frame is started, the server computes all rendering tasks and
-sends them to the appropriate render client nodes. The render client
-nodes dispatch the tasks to the correct node or pipe thread, where they
-are executed in order of arrival.
-
-\textsf{Config::finishFrame} synchronizes on the completion of the frame
-\textsf{current - latency}. The latency is specified in the
-configuration file, and allows several outstanding frames. This allows
-overlapping execution in the node and pipe threads and minimizes idle
-times.
-
 By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
-  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
-\end{wrapfigure}
 \textsf{eqPly} uses static distributed objects to distribute the model
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
+  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
+\end{wrapfigure}
 The identifer of the root object of this distributed hierarchy is passed
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/objects.pdf}
-  {\caption{\small\label{fObjects}Distributed Data Handling}}
-\end{wrapfigure}
 Applications distributing a dynamic scene graph use the frame data
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/objects.pdf}
+  {\caption{\small\label{fObjects}Scene Graph Distribution}}
+  \vspace{-4em}
+\end{wrapfigure}
 The objects are still created by the application, and then registered or
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/windowInit.pdf}
-  {\caption{\small\label{fWindowInit}Window Initialization}}
-\end{wrapfigure}
 \fig{fWindowInit} shows a flow chart of the window initialization. The
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/windowInit.pdf}
+  {\caption{\small\label{fWindowInit}Window Initialization}}
+%  \vspace{-1em}
+\end{wrapfigure}
 The state object is used to handle the creation of OpenGL objects in a
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/DB.png}
+  {\caption{\small\label{fDB}Destination view of a DB compound}}
+\end{wrapfigure}
 After the basic view setup, a directional light is configured, and the
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/DB.png}
-  {\caption{\small\label{fDB}Destination view of a DB compound}}
-\end{wrapfigure}
 In order to draw the model, a helper class for view frustum culling is
@@
 \begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-1em}
   \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Rendering Loop}}
+  {\caption{\small\label{fRender}Main Render Loop}}
+  \vspace{-2em}
 \end{wrapfigure}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
 Please note that the default thread synchronization synchronizes all
@@
 
-Each entity uses \textsf{waitFrameStarted} to block on the parent's
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
+Each entity uses \textsf{waitFrame\-Started} to block on the parent's
 \textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 DB compounds have the advantage of scaling any part of the volume
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/statistics.png}
+  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
+\end{wrapfigure}
 Statistics are measured in milliseconds since the configuration was
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/statistics.png}
-  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
-\end{wrapfigure}
 Statistics are enabled per entity using an attribute hint. The hint
@@
 
-\subsection{Global Section}
+\subsection{\label{sGlobal}Global Section}
 
@@
   \textbf{Default} & window\\
-\textbf{Details} & Selects the window's drawable type.\\
+  \textbf{Details} & Selects the window's drawable type. A window is an
+  on-screen, window system-dependent window with a full-window OpenGL
+  drawable. PBuffers are off-screen drawables created using window
+  system-dependent PBuffer API's. In order to calculate the PBuffer size
+  on unconnected devices, a pipe viewport size of 4096x4096 is assumed,
+  unless specified otherwise.\\
   \hline
@@
   \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_STATISTICS\\
-\textbf{Value} & off \textbar \ fastest [on] \textbar \ nicest\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Value} & OFF \textbar \ FASTEST [ON] \textbar \ NICEST\\
+  \textbf{Default} & FASTEST [Release Build]\\
+                   & NICEST [Debug Build]\\
+  \textbf{Details} & Determines how statistics are gathered. OpenGL
+  buffers commands, which causes the rendering to be executed at an
+  arbitrary point in time. Nicest statistics gathering executes a
+  \textsf{Window::finish}, which calls by default \textsf{glFinish}, in
+  order to accurately account the rendering operations to the sampled
+  task method. However, calling \textsf{glFinish} has a performance
+  impact. Therefore, the fastest statistics gathering samples the task
+  statistics directly, without finishing the OpenGL commands.\\
   \hline
@@
   \textbf{Value} & unsigned\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Default} & AUTO\\
+  \textbf{Details} & Determines the number of color planes for the
+  window. The interpretation of this value is window system-specific, as
+  some window systems select a visual with the closest match to this
+  value, and some select a visual with at least the number of color
+  planes specified. AUTO select a visual with a reasonable quality,
+  typically eight bits per color.\\
   \hline
@@
   \textbf{Value} & unsigned\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Default} & UNDEFINED\\
+  \textbf{Details} & Determines the number of alpha planes for the
+  window. The interpretation of this value is window system-specific, as
+  some window systems select a visual with the closest match to this
+  value, and some select a visual with at least the number of alpha
+  planes specified. By default no alpha planes are requested.\\
   \hline
@@
   \textbf{Value} & unsigned\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Default} & AUTO\\
+  \textbf{Details} & Determines the precision of the depth buffer. The
+  interpretation of this value is window system-specific, as some window
+  systems select a visual with the closest match to this value, and some
+  select a visual with at least the number of depth bits specified. AUTO
+  select a visual with a reasonable depth precision, typically 24 bits.\\
   \hline
@@
   \textbf{Value} & unsigned\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Default} & UNDEFINED\\
+  \textbf{Details} & Determines the number of stencil planes for the
+  window. The interpretation of this value is window system-specific, as
+  some window systems select a visual with the closest match to this
+  value, and some select a visual with at least the number of stencil
+  planes specified. By default no stencil planes are requested. Note
+  that for depth-compositing and pixel-compositing at least one stencil
+  plane is needed.\\
   \hline
@@
   \textbf{Value} & OFF \textbar \ FASTEST [ ON ] \textbar \ NICEST\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Default} & FASTEST [Release Build]\\
+                   & NICEST [Debug Build]\\
+  \textbf{Details} & See EQ\_WINDOW\_IATTR\_HINT\_STATISTICS.\\
   \hline
@@
   \textbf{Value} & QUAD \textbar \ ANAGLYPH\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Default} & QUAD\\
+  \textbf{Details} & Selects the algorithm used for stereo
+  rendering. QUAD-buffered stereo use the left and right buffers of a
+  stereo window (active stereo). Anaglyphics stereo uses
+  \textsf{glColorMask} to mask colors for individual eye passes, used in
+  conjunction with colored glasses.\\
   \hline
@@
   \textbf{Value} & [ RED GREEN BLUE ]\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Default} & [ RED ] \\
+  \textbf{Details} & Select the color mask for the left eye pass during
+  anaglyphic stereo rendering.\\
   \hline
@@
   \textbf{Value} & [ RED GREEN BLUE ]\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Default} & [ GREEN BLUE ]\\
+  \textbf{Details} & Select the color mask for the right eye pass during
+  anaglyphic stereo rendering.\\
   \hline
@@
   \textbf{Value} & HORIZONTAL [ON] \textbar \ VERTICAL \textbar \ OFF\\
-\textbf{Default} & \\
-\textbf{Details} & \\
+  \textbf{Default} & HORIZONTAL\\
+  \textbf{Details} & Define how the wall or projection description is
+  updated when a window is resized. Horizontal increases or decreases
+  the horizontal field of view, while keeping the vertical size
+  constant. Vertical updates the vertical field of view, while keeping
+  the horizontal size constant. OFF does not update the wall or
+  projection description at all.\\
   \hline
@@
 
+The server section consists of connection description parameters for the
+server listening sockets, and a number of configurations for this
+server. Currently only the first configuration is used.
+
+\subsubsection{\label{sConnectionDescription}Connection Description}
+
+A connection description defines the network parameters of an Equalizer
+process. Currently two connection types, TCP/IP and SDP, are
+supported. TCP/IP creates a TCP socket. SDP is very similar, except that
+the address family \textsf{AF\_INET\_SDP} instead of \textsf{AF\_INET}
+is used to enforce a SDP connection. Note that you can also use the
+transparent mode provided by most InfiniBand implementations to used SDP
+with TCP connections.
+
+Furthermore, a port for the socket can be specified. When no port is
+specified for the server, the default port 4242 (+UID on Posix systems)
+is used. When no port is specified for a node, a random port will be
+chosen by the operating system. For pre-launched render clients, a port
+has to be specified for the server to find the client node.
+
+The hostname is the IP address or (resolvable) host name. When used for
+the server, the server binds its listening socket only to the specified
+name.
+
+A server or node may have multiple connection descriptions.
+\clearpage
 {\footnotesize\begin{lstlisting}
@@
 
-For an introduction on the concepts of nodes, pipes, windows, channels
-and compounds please refer to \sref{sConfig} and \sref{sCompounds}.
-
 \subsubsection{Config Section}
+
+A configuration has a number of parameters, nodes and compounds.
+
+The nodes and their children describe the rendering resources in a
+natural, hierarchical way. Compounds use rendering resources (channels)
+to execute rendering tasks. For an introduction on the concepts of
+nodes, pipes, windows, channels and compounds please refer to
+\sref{sConfig} and \sref{sCompounds}.
+
+The latency of a config defines the maximum number of frames the
+slowest operation may fall behind the application thread. A latency of 0
+synchronizes all rendering tasks started by \textsf{Config::startFrame}
+in \textsf{Config::finishFrame}. A latency of one synchronizes all
+rendering tasks started one frame ago in \textsf{finishFrame}.
+
+For a description of config attributes please refer to \sref{sGlobal}.
 
@@
 
-{\footnotesize\begin{lstlisting}
+For a description of connection attributes please refer to \sref{sGlobal}.
+
+{\footnotesize\begin{lstlisting}
+node
 {

### 2008-05-19 Developer/ProgrammingGuide/paper.tex
Finish Programming Guide File Format section, and homogenize file format spec, PG and implementation.
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.3 for Equalizer 0.5+svn\\[\medskipamount]
+  Version 1.3.3 for Equalizer 0.5.1+svn\\[\medskipamount]
   \today
@@
     1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
-    1.3.3   & May 10-16, 2008 & File Format Appendix [incomplete]\\
+    1.3.3   & May 10-19, 2008 & File Format Appendix\\
     \hline
@@
 
-The currently used file format is the one-to-one representation of the
-server's internal data structures. Its purpose is intermediate, that
-is, it will gradually be replaced by automatic resource dedection and
-configuration. However the core scalability engine will always use a
-similar structure as currently exposed by the file format.
+The file format currently used is the one-to-one representation of the
+server's internal data structures. Its purpose is intermediate, that is,
+it will gradually be replaced by automatic resource dedection and
+automatic configuration. However the core scalability engine will
+always use a similar structure as currently exposed by the file format.
 
@@
 editor supports syntax highlighting and formatting for VRML, you can use
-this mode for \textsf{.eqc} files.
+this mode for editing \textsf{.eqc} files.
 
@@
 The global section defines default values for attributes used by the
-individual entities of the server section. The naming convention for
+individual entities in the server section. The naming convention for
 attributes is:
@@
 values are handled as integers. Strings have always to be surrounded by
-double quotes '\textsf{''}'. A character has to surrounded by single
+double quotes '\textsf{''}'. A character has to be surrounded by single
 quotes '\textsf{'}'.
@@
 values the corresponding integer value has to be used. The global values
-in the config file override these defaults, and are in turn overriden by
-the corresponding attributes sections of the associated entities.
+in the config file override environment variables, and are in turn
+overridden by the corresponding attributes sections of the specific
+entities.
 
@@
   which could not be connected. The launch command is executed from a
-  process forked of the server process. The \% tokens are replaced by the
-  server at runtime with concrete data: \%h is replaced by the hostname,
-  \%c by the command to launch, including command line arguments and \%n
-  by a node-unique identifier. Each command line argument are surrounded
-  by launch command quotes.\\
+  process forked from the server process. The \% tokens are replaced by
+  the server at runtime with concrete data: \%h is replaced by the
+  hostname, \%c by the command to launch, including command line
+  arguments and \%n by a node-unique identifier. Each command line
+  argument is surrounded by launch command quotes.\\
   \hline
@@
                    & '' [WIN32]\\
-  \textbf{Details} & The server uses command line arguments to
-  launch render client nodes correctly. This arguments might contain
-  white spaces, and therefore have to be surrounded by quotes to
-  identify their borders. Certain launch commands or shells use
-  different conventions to separate command line arguments.\\
+  \textbf{Details} & The server uses command line arguments to launch
+  render client nodes correctly. Certain launch commands or shells use
+  different conventions to separate command line arguments. These
+  arguments might contain white spaces, and therefore have to be
+  surrounded by quotes to
+  identify their borders.\\
   \hline
@@
   programmatically selects the socket direct protocol (AF\_INET\_SDP)
-  provided by most InfiniBand protocol stacks.\\
+  provided by most InfiniBand protocol stacks, TCPIP uses normal TCP
+  sockets (AF\_INET).\\
   \hline
@@
   (default) or from the node main thread. Non-threaded pipes have
-  certain performance
-  limitations and should only be used if really necessary.\\
+  certain performance limitations and should only be used where necessary.\\
   \hline
@@
   stereo visual. When set to AUTO, the window initialization methods try
-  to allocate a stereo visual, but fall back to a mono visual if
-  allocation fails.\\
+  to allocate a stereo visual for windows, but fall back to a mono
+  visual if allocation fails. For PBuffers, AUTO selects a mono visual.\\
   \hline
@@
   stereo visual. When set to AUTO, the window initialization methods try
-  to allocate a double-buffered visual, but fall back to a
-  single-buffered visual if allocation fails.\\
+  to allocate a double-buffered visual for windows, but fall back to a
+  single-buffered visual if allocation fails. For PBuffers, AUTO selects
+  a single-buffered visual.\\
   \hline
@@
   impact. Therefore, the fastest statistics gathering samples the task
-  statistics directly, without finishing the OpenGL commands.\\
+  statistics directly, without finishing the OpenGL commands first. Some
+  operations, e.g., frame buffer readback, inherently finish all
+  previous OpenGL commands.\\
   \hline
@@
   value, and some select a visual with at least the number of color
-  planes specified. AUTO select a visual with a reasonable quality,
+  planes specified. AUTO selects a visual with a reasonable quality,
   typically eight bits per color.\\
@@
   \textbf{Value} & unsigned\\
-  \textbf{Default} & UNDEFINED\\
+  \textbf{Default} & AUTO\\
   \textbf{Details} & Determines the number of stencil planes for the
@@
   value, and some select a visual with at least the number of stencil
-  planes specified. By default no stencil planes are requested. Note
-  that for depth-compositing and pixel-compositing at least one stencil
-  plane is needed.\\
+  planes specified. AUTO tries to select a visual with at least one
+  stencil plane, but falls back to no stencil planes if allocation
+  fails. Note that for depth-compositing and pixel-compositing at least
+  one stencil plane is needed.\\
   \hline
@@
   \textbf{Details} & Selects the algorithm used for stereo
-  rendering. QUAD-buffered stereo use the left and right buffers of a
+  rendering. QUAD-buffered stereo uses the left and right buffers of a
   stereo window (active stereo). Anaglyphics stereo uses
@@
 name.
-
-A server or node may have multiple connection descriptions.
 \clearpage
+
+A server or node may have multiple connection descriptions. A server
+binds to all provided connection descriptions. For a node, all
+connection descriptions are used while trying to establish a connection
+to the node. When auto-launched by the server, all connection
+descriptions are passed to the launched node process, which binds to all
+provided descriptions.
+
 {\footnotesize\begin{lstlisting}
@@
 
-For a description of connection attributes please refer to \sref{sGlobal}.
+A node represents a machine in the cluster, and is one process. It has a
+name, a number of connection descriptions and at least one pipe. The
+name of the node can be used for debugging, it has no influence on the
+execution of Equalizer. For a description of connection attributes
+please refer to \sref{sGlobal}.
 
@@
 
+A pipe represents a graphics card (GPU), and is one execution thread. It
+has a name, GPU settings and attributes. The name of a pipe can be used
+for debugging, it has no influence on the execution of Equalizer.
+
+The GPU is identified by two parameters, a port and a device. The port
+is only used for the GLX window system, and identifies the port number
+of the X server, i.e., the number after the colon in the DISPLAY
+description (':\textbf{0}.1').
+
+The device identifies the graphics adapter. For the GLX window system
+this is the screen number, i.e., the number after the dot in the DISPLAY
+description (:0.\textbf{1}). The OpenGL output is always restricted by
+glX to the GPU attached to selected screen.
+
+For the AGL window system, the device selects the \textit{n}th display
+in the list of online displays. The OpenGL output is optimized for the
+selected display, but not restricted to the attached GPU.
+
+For the WGL window system, the device selects the \textit{n}th GPU in
+the system. The GPU can be offline, in this case only PBuffer windows
+can be used. In order to restrict the OpenGL output to the GPU, the
+\textsf{WGL\_NV\_gpu\_affinity} extension is used. If the extension is
+not present, the window is opened on the \textit{n}th monitor, but
+OpenGL commands are sent to all GPU's.
+
+The viewport of the pipe can be used to override the pipe
+resolution. The viewport is defined in pixels. The x and y parameter of
+the viewport are currently ignored. The default viewport is
+automatically detected. For offline GPU's, a default of 4096x4096 is
+used.
+
+For a description of pipe attributes please refer to \sref{sGlobal}.
+
 {\footnotesize\begin{lstlisting}
@@
 \subsubsection{Window Section}
+
+A window represents an OpenGL drawable and holds an OpenGL context. It
+has a name, a viewport and attributes. The name of a window can be used
+for debugging, it has no influence on the execution of Equalizer.
+
+The viewport of the window is relative to the pipe. It can be specified
+in relative or absolute coordinates. Relative coordinates are normalized
+coordinates with respect to the pipe, e.g., a viewport of \textsf{[ 0.25
+  0.25 0.5 0.5 ]} creates a window in the middle of the screen, using
+50\% of the pipe's size. Absolute coordinates are integer pixel values,
+e.g., a viewport of \textsf{[ 50 50 800 600 ]} creates a window 50 pixels
+from the upper-left corner, sized 800x600 pixels, regardless of the
+pipe's size. The default viewport is \textsf{[ 0 0 1 1 ]}, i.e., a
+full-screen window.
+
+For a description of window attributes please refer to \sref{sGlobal}.
 
@@
 
+A channel is a two-dimensional area within a window. It has a name,
+viewport and attributes. The name of the channel is used to identify the
+channel in the respective compounds. It should be unique within the config.
+
+The viewport of the channel is relative to the window. As for windows,
+it can be specified in relative or absolute coordinates. The default
+viewport is \textsf{[ 0 0 1 1 ]}, i.e., fully covering its window.
+
+For a description of channel attributes please refer to \sref{sGlobal}.
+
 {\footnotesize\begin{lstlisting}
@@
 \subsubsection{Compound Section}
+
+Compounds are the basic datastructure describing the rendering
+setup. They use channels for rendering. Please refer to
+\sref{sCompounds} for a description of compound operation logics.
+
+The name of the compound is used for the default names of swap barriers
+and output frames. The channel references the first channel of the same
+name in the configuration.
+
+Compound tasks describe the operations the compound executes. The
+default is all tasks for compounds with no children (leaf compounds) and
+\textsf{READBACK ASSEMBLE} for all others. The readback and assemble
+tasks are only executed if the compound has output frames or input
+frames, respectively. Tasks are not inherited by the children of a
+compound.
+
+The buffer defines the default frame buffer attachments read back by
+output frames. The viewport restricts the rendering to the area relative
+to the parent compound. The range restricts the database range, relative
+to the parent. The pixel setting selects the pixel decomposition
+parameters, relative to the parent. The eye pass select the mono, left
+or right eye passes rendered by the compound. All these attributes are
+inherited by the children of a compound. Viewport, range and pixel
+parameters are cumulative.
+
+For a description of compound attributes please refer to \sref{sGlobal}.
+
+A wall or projection description is used to define the view frustum of
+the compound. The frustum is inherited and typically only defined on the
+top-most compound. The last specified frustum description is used. Sizes
+are specified in meters. \fig{fFrustra} illustrates the frustum
+parameters.
+
+A swap barrier is used to synchronize the output of multiple
+compounds. All swap barriers of the same name are synchronized The
+default name is \textsf{barrier[.root\-Com\-pound\-Name]}.
+
+Output frames transport frame buffer contents to input frames of the
+same name. If the compound has a name, the default frame name is
+\textsf{frame.compoundName}, otherwise the default name is
+\textsf{frame.channelName}. The frame buffer attachments to read back
+are inherited from the compound, but can be overridden by output frames.
 
@@
         stereo_mode                  QUAD | ANAGLYPH    // default QUAD
-        // enable visible color components
         stereo_anaglyph_left_mask    [ RED GREEN BLUE ] // default red
@@
         update_FOV                   HORIZONTAL [ON] | VERTICAL | OFF
-        assemble                     GPU | CPU          // default GPU
     }
@@
     }
+    projection // alternate frustum description, last one wins
+    {
+        origin       [ float float float ]
+        distance     float
+        fov          [ float float ]
+        hpr          [ float float float ]
+    }
 
@@
         name   string
+        buffer [ COLOR DEPTH ]
     }

### 2008-07-03 Developer/ProgrammingGuide/paper.tex
Update statistics overlay documentation in Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.3 for Equalizer 0.5.1+svn\\[\medskipamount]
+  Version 1.3.4 for Equalizer 0.5.2+svn\\[\medskipamount]
   \today
@@
     1.3.3   & May 10-19, 2008 & File Format Appendix\\
+    1.3.4   & July 3, 2008 & Update statistics section\\
     \hline
@@
 
-In debug builds, a statistics overlay using the gathered statistics
-events is rendered directly before \textsf{Window::swapBuffers}. The
-overlay rendering method, \textsf{Channel::drawStatistics}, can be
-called directly to render the overlay. It can also be overwritten by a
+The Equalizer examples render a statistics overlay using the gathered
+statistics events directly before
+\textsf{Window::swapBuffers}. Statistics are toggled on and off using
+the 's' key. The overlay rendering method,
+\textsf{Channel::drawStatistics}, can be called directly by any
+application to render the overlay. It can also be overwritten by a
 custom implementation.
@@
 
-In order to facilitate the understanding, subsequent frames alternate in
-brightness. The right-most, current frame is brighter than the
+In order to facilitate the understanding, older frames are gradually
+greyed out. The right-most, current frame is brighter than the
 frame-before-last.
@@
 The beginning of a frame is marked by a vertical green line, and the end
-of a frame by a vertical gray line. These lines also alternate in
-brightness. The brightness and color matches the event for
-\textsf{Config::startFrame} and \textsf{Config::finishFrame},
-respectively. The event for \textsf{startFrame} is typically not
-visible, since it takes less than one millisecond to execute. If no idle
-processing is done by the application, the event for
-\textsf{finishFrame} occupies a full frame, since the config is blocked
-here waiting for the frame to complete.
+of a frame by a vertical gray line. These lines are also attenuated. The
+brightness and color matches the event for \textsf{Config::startFrame}
+and \textsf{Config::finishFrame}, respectively. The event for
+\textsf{startFrame} is typically not visible, since it takes less than
+one millisecond to execute. If no idle processing is done by the
+application, the event for \textsf{finishFrame} occupies a full frame,
+since the config is blocked here waiting for the frame to complete.
 
@@
 the CPU.
+
+The configuration used in this examples does not use a swap
+barrier. Swap barrier tasks are solid red events directly in front of
+the white swap buffer task.
 

### 2008-07-08 Developer/ProgrammingGuide/paper.tex
Update docs (namespace renaming)
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.4 for Equalizer 0.5.2+svn\\[\medskipamount]
+  Version 1.3.5 for Equalizer 0.5.3+svn\\[\medskipamount]
   \today
@@
     1.3.4   & July 3, 2008 & Update statistics section\\
+    1.3.4   & July 7, 2008 & Update after namespace naming cleanup\\
     \hline
@@
 The distributed data classes are helper classes based on
-\textsf{eqNet::Object}. They illustrate the typical usage of distributed
+\textsf{eq::net::Object}. They illustrate the typical usage of distributed
 objects for static as well as dynamic, frame-specific data.
@@
 initialize it locally. The application is an \textsf{eq::Client}, which
-in turn is an \textsf{eqNet::Node}. The underlying network layer in
-Equalizer is a peer-to-peer network of \textsf{eqNet::Node}s. The
+in turn is an \textsf{eq::net::Node}. The underlying network layer in
+Equalizer is a peer-to-peer network of \textsf{eq::net::Node}s. The
 application programmer does not usually have to be aware of the classes
-in the \textsf{eqNet} namespace, but both the \textsf{eq::Client} and
-the server are \textsf{eqNet::Node}s.
+in the \textsf{eq::net} namespace, but both the \textsf{eq::Client} and
+the server are \textsf{eq::net::Node}s.
 
@@
     // 3. init config
-    eqBase::Clock clock;
+    eq::base::Clock clock;
 
@@
 distribution in a cluster environment. The master version of a distributed
-object is registered with a \textsf{eqNet::Session}, which assigns a
+object is registered with a \textsf{eq::net::Session}, which assigns a
 session-unique identifier to the object. Other nodes can map their
@@
 Distributed objects are created by subclassing from
-\textsf{eqNet::Object}. Distributed objects can be static (immutable) or
+\textsf{eq::net::Object}. Distributed objects can be static (immutable) or
 dynamic. Dynamic objects are versioned.
@@
 {\footnotesize\begin{lstlisting}
-void InitData::getInstanceData( eqNet::DataOStream& os )
+void InitData::getInstanceData( eq::net::DataOStream& os )
 {
@@
 
-void InitData::applyInstanceData( eqNet::DataIStream& is )
+void InitData::applyInstanceData( eq::net::DataIStream& is )
 {
@@
 \item Only the master instance of the object is writable, that is,
-  \textsf{eqNet::Object::com\-mit} can be called only on the master
+  \textsf{eq::net::Object::com\-mit} can be called only on the master
   instance to generate a new version.
 \item Slave instance versions can only be advanced, that is,
-  \textsf{eqNet::Object::sync( version )} with a version smaller than
+  \textsf{eq::net::Object::sync( version )} with a version smaller than
   the current version will fail.
@@
 {\footnotesize\begin{lstlisting}
-void VertexBufferDist::getInstanceData( eqNet::DataOStream& os )
+void VertexBufferDist::getInstanceData( eq::net::DataOStream& os )
 {
@@
 initialization and frame data. Both objects are registered with the
-\textsf{eq::Config}, which is the \textsf{eqNet::Session} used for
+\textsf{eq::Config}, which is the \textsf{eq::net::Session} used for
 rendering. Equalizer takes care of the session setup and exit in
@@
 {\footnotesize\begin{lstlisting}
-mesh::VertexBufferRoot* VertexBufferDist::mapModel(eqNet::Session* session,
+mesh::VertexBufferRoot* VertexBufferDist::mapModel(eq::net::Session* session,
                                                    const uint32_t modelID )
@@
 The command thread processes Equalizer-related request from other nodes,
-for example during \textsf{eqNet::Object} mapping. In some special cases
+for example during \textsf{eq::net::Object} mapping. In some special cases
 the command thread executes application code, for example when a remote
@@
 The file format represents an ascii deserialization of the
-server. Streaming an \textsf{eqs::Server} to an \textsf{eqBase::Log}
+server. Streaming an \textsf{eq::server::Server} to an \textsf{eq::base::Log}
 ostream produces a valid configuration file. Likewise, loading a
-configuration file produces an \textsf{eqs::Server}.
+configuration file produces an \textsf{eq::server::Server}.
 
@@
 various attributes. The server section represents an
-\textsf{eqs::Server}.
+\textsf{eq::server::Server}.
 

### 2008-07-21 Developer/ProgrammingGuide/paper.tex
Implement automatic image compression on 'slow' links, add accum and samples hint to window attributes. Amend the documentation accordingly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[x] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.5 for Equalizer 0.5.3+svn\\[\medskipamount]
+  Version 1.3.6 for Equalizer 0.5.3+svn\\[\medskipamount]
   \today
@@
 \newcommand{\sref}[1]{Section~\ref{#1}}
+\newcommand{\aref}[1]{Appendix~\ref{#1}}
 \newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}
@@
   courtesy of Computer Graphics and Multimedia Systems, University of
-  Siegen}, the \textsf{eVolve} volume renderer\footnote{Data sets
-  courtesy of General Electric, USA and AVS, USA} [bottom left] the
-\textsf{eqPly} polygonal renderer in a three-sided CAVE [bottom center],
-and a six-node sort-last database decomposition with parallel
-direct-send recomposition\footnote{Data set courtesy of Stanford
-  University Computer Graphics Laboratory} [bottom right].
+  Siegen}, the \textsf{eVolve} volume renderer\footnote{Data set
+  courtesy of General Electric, USA} [bottom left] the \textsf{eqPly}
+polygonal renderer in a three-sided CAVE [bottom center], and a six-node
+sort-last database decomposition with parallel direct-send
+recomposition\footnote{Data set courtesy of Stanford University Computer
+  Graphics Laboratory} [bottom right].
 
@@
     1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
-    1.3.1   & May 02, 2008 & Add statistics, improved images based on 1.2 print\\
+    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
     1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
-    1.3.3   & May 10-19, 2008 & File Format Appendix\\
-    1.3.4   & July 3, 2008 & Update statistics section\\
-    1.3.4   & July 7, 2008 & Update after namespace naming cleanup\\
+    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
+    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
+    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
+    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
     \hline
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-1em}
   \includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
   {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
-  \vspace{-1em}
 \end{wrapfigure}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \subfigure[]{\label{fSync}
-    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
-  \subfigure[]{\label{fAsync}
-    \includegraphics[width=.25\textwidth]{images/async}}%
-  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
-\end{wrapfigure}
 By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
@@
 
-\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
-without latency (\fig{fSync}) and with a latency of
-one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
-observed on a five-node rendering cluster when using a latency of one frame
-instead of no
+\begin{figure}[ht!]\center
+  \includegraphics[width=\textwidth]{images/syncAsync}
+  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
+\end{figure}
+\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
+2D compound without latency and with a latency of one frame. The
+asynchronous execution pipelines certain rendering operations and hides
+imbalances in the load distribution, resulting in an improved
+framerate. With \textsf{eqPly}, a speedup of 15\% has been observed on a
+five-node rendering cluster when using a latency of one frame instead of
+no
 latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/objects.pdf}
+  {\caption{\small\label{fObjects}Scene Graph Distribution}}
+\end{wrapfigure}
 Applications distributing a dynamic scene graph use the frame data
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/objects.pdf}
-  {\caption{\small\label{fObjects}Scene Graph Distribution}}
-  \vspace{-4em}
-\end{wrapfigure}
 The objects are still created by the application, and then registered or
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
-  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
-\end{wrapfigure}
 The tasks are executed by the node and pipe threads in parallel. The
@@
   {\caption{\small\label{fWindowInit}Window Initialization}}
-%  \vspace{-1em}
 \end{wrapfigure}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-1em}
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Render Loop}}
-  \vspace{-2em}
-\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Render Loop}}
+\end{wrapfigure}
 The rendering loop maintains a list of candidates to render, which
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\small\label{fEventProcessing}Event Processing}}
+  \vspace{-3ex}
+\end{wrapfigure}
 Events are received by an event handler. The event handler finds the
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\small\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 The event handler then passes the window event to
@@
 OpenGL pixel format selection.
+
+\subsubsection{\label{sDrawableConfig}Drawable Configuration}
+
+OpenGL drawables have a multitude of buffer modes. A drawable might be
+single-buffered, double-buffered or quad-buffered, have auxilary image
+planes such as stencil, accumulation and depth buffer or multisampling.
+
+The OpenGL drawable is configured using window attributes. These
+attributes are used by the method choosing the pixel format (or visual
+in X11 speak) to select the correct drawable configuration.
+
+Window attributes can either be configured through the configuration
+file (see \aref{aFileFormat}), or programmatically. In the configuration
+file, modes are selected which are not application-specific, for example
+stereo formats for active stereo displays.
+
+Applications which require certain drawable attributes can set the
+corresponding window attribute hint during window initialization. The
+Equalizer volume rendering example, \textsf{eVolve}, is such an
+example. It does need alpha planes for rendering and compositing. The
+window initialization of \textsf{eVolve} sets the attribute before
+calling the default initialization method of Equalizer:
+
+{\footnotesize\begin{lstlisting}
+bool eVolve::Window::configInit( const uint32_t initID )
+{
+    // Enforce alpha channel, since we need one for rendering
+    setIAttribute( IATTR_PLANES_ALPHA, 8 );
+
+    return eq::Window::configInit( initID );
+}
+\end{lstlisting}}
 
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
+      Volume Rendering}}
+\end{wrapfigure}
  When using 2D (sort-first) or stereo decompositions, no special
@@
 
-\subsection{Statistics}
+\subsection{\label{sStatistics}Statistics}
 
@@
 \appendix
-\section{File Format}
+\section{\label{aFileFormat}File Format}
 
@@
   \hline
+  \textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_ACCUM\\
+  \textbf{Value} & unsigned\\
+  \textbf{Default} & UNDEFINED\\
+  \textbf{Details} & Determines the number of color accumulation buffer
+  planes for the window. The interpretation of this value is window
+  system-specific, as some window systems select a visual with the
+  closest match to this value, and some select a visual with at least
+  the number of accumulation buffer planes specified.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_ACCUM\_ALPHA\\
+  \textbf{Value} & unsigned\\
+  \textbf{Default} & UNDEFINED\\
+  \textbf{Details} & Determines the number of alpha accumulation buffer
+  planes for the window. The interpretation of this value is window
+  system-specific, as some window systems select a visual with the
+  closest match to this value, and some select a visual with at least
+  the number of accumulation buffer planes specified. If this attribute
+  is undefined, the value of EQ\_WINDOW\_IATTR\_PLANES\_ACCUM is used to
+  determine the number of alpha accumulation buffer planes.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_SAMPLES\\
+  \textbf{Value} & unsigned\\
+  \textbf{Default} & UNDEFINED\\
+  \textbf{Details} & Determines the number of samples used for multisampling.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
   \textbf{Name} & EQ\_CHANNEL\_IATTR\_HINT\_STATISTICS\\
@@
 name.
-\clearpage
 
@@
         planes_stencil      unsigned
+        planes_accum        unsigned
+        planes_accum_alpha  unsigned
+        planes_samples      unsigned
     }

### 2008-08-22 Developer/ProgrammingGuide/paper.tex
Programming Guide: Partial update for OSWindow changes.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.6 for Equalizer 0.5.3+svn\\[\medskipamount]
+  Version 1.3.7 for Equalizer 0.5.4+svn\\[\medskipamount]
   \today
@@
     1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
+    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes (event handling updates still missing)\\
     \hline
@@
 
-The \textsf{eq::Window} class holds a drawable and an OpenGL
+The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
 context. The drawable can be an on-screen window or an off-screen
-PBuffer. Framebuffer object (FBO) support will be implement by later
-version of Equalizer. The window holds window-system-specific handles to
-the drawable and context, e.g., an X11 window \textsf{XID} and
-\textsf{GLXContext} for the glX window system.
+PBuffer. Framebuffer object (FBO) support will be implement by a later
+version of Equalizer. The window uses an \textsf{eq::OSWindow}, which
+abstracts and manages window-system-specific handles to the drawable and
+context, e.g., an X11 window \textsf{XID} and \textsf{GLXContext} for
+the glX window system.
 
@@
 
-\subsection{Window}
-
-The Equalizer window holds an OpenGL drawable and a rendering
+\subsection{\label{sEqplyWIndow}Window}
+
+The Equalizer window abstracts an OpenGL drawable and a rendering
 context. When using the default window initialization functions, all
@@
 
+\subsubsection{Window System Interface}
+
+The particulars of creating a window and OpenGL context depend on the
+window system used. One can either use the implementation provided by
+the operating system, e.g., AGL, WGL or glX, or some higher-level
+toolkit, e.g., Qt.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
+  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
+\end{wrapfigure}
+All window-system specific functionality is implemented by a
+specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
+defines the minimal interface to be implemented for a new window
+system. Each \textsf{Window} uses one \textsf{OSWindow} during
+execution. This separation allows an easy implementation and adaption to
+another window system or application.
+
+Equalizer provides a generic interface and implementation for the three
+most common window systems: AGL, WGL and glX. The interfaces define the
+minimal functionality needed to reuse other window system specific
+classes, for example the AGL, WGL and glX event handlers. The
+implementation derived from these interfaces provides a sample
+implementation which honors all configurable window attributes.
+
 \subsubsection{Initialization and Exit}
@@
 The initialization sequence uses multiple, overrideable task
-methods. The main task method \textsf{configInit} executes a `child'
-task method to create the drawable and context. The child task method
-depends on the window system of the pipe. The default implementations of
-\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
-create an on-screen window using OS-specific methods. If the OpenGL
-drawable and context were created successfully, \textsf{configInit} calls
-\textsf{configInitGL}, which performs the generic OpenGL
-state initialization. The default implementation sets up some typical
-OpenGL state, e.g., it enables the depth test.
-
-\fig{fWindowInit} shows a flow chart of the window initialization. The
-colored functions are task methods and can be replaced by
-application-specific implementations.
-
-The window-system specific initialization takes into account various
+methods. The main task method \textsf{configInit} calls first
+\textsf{configInitOSWindow}, which creates and initializes the
+\textsf{OSWindow} for this window. The \textsf{OSWindow} initialization
+code is implementation specific. If the \textsf{OSWindow} was
+initialized successfully, \textsf{configInit} calls
+\textsf{configInitGL}, which performs the generic OpenGL state
+initialization. The default implementation sets up some typical OpenGL
+state, e.g., it enables the depth test.
+
+The \textsf{OSWindow} initialization takes into account various
 attributes set in the configuration file. Attributes include the size of
@@
 attributes, such as stereo, doublebuffer and stencil can be set to
-\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
-enable them if possible.
+\textsf{eq::AUTO}, in which case the Equalizer default implementation
+will test for their availability and enable them if possible.
 
 For the window-system specific initialization, \textsf{eqPly} uses the
-default Equalizer implemenation. The \textsf{eqPly} window
+default Equalizer implementation. The \textsf{eqPly} window
 initialization only overrides the OpenGL-specific initialization
-function in order to initialize a state object and an overlay logo. This
-function is only called if an OpenGL context was created and made current:
+function \textsf{configInitGL} in order to initialize a state object and
+an overlay logo. This function is only called if an OpenGL context was
+created and made current:
 
@@
 
-    eq::Pipe*  pipe        = getPipe();
-    Window*    firstWindow = static_cast< Window* >( pipe->getWindows()[0] );
+    glLightModeli( GL_LIGHT_MODEL_LOCAL_VIEWER, 1 );
+    glEnable( GL_CULL_FACE ); // OPT - produces sparser images in DB mode
+    glCullFace( GL_BACK );
 
     EQASSERT( !_state );
-
-    if( firstWindow == this )
-    {
-        _state = new VertexBufferState( glewGetContext( ));
+    _state = new VertexBufferState( getObjectManager( ));
+    _loadLogo();
 
@@
     const InitData& initData = config->getInitData();
-
-        if( initData.useVBOs() )
-        {
-            // Check if VBO funcs available, else leave DISPLAY_LIST_MODE on
-            if( GLEW_VERSION_1_5 )
-            {
-                _state->setRenderMode( mesh::BUFFER_OBJECT_MODE );
-                EQINFO << "VBO rendering enabled" << endl;
-            }
-            else
-                EQWARN << "VBO function pointers missing, using display lists" 
-                       << endl;
-        }
-
     if( initData.useGLSL() )
-        {
-            // Check if all functions are available
-            if( GLEW_VERSION_2_0 )
-            {
-                EQINFO << "Shaders supported, attempting to load" << endl;
         _loadShaders();
-            }
-            else
-                EQWARN << "Shader function pointers missing, using fixed "
-                       << "pipeline" << endl;
-        }
-
-        _loadLogo();
-    }
-    else
-    {
-        _state       = firstWindow->_state;
-        _logoTexture = firstWindow->_logoTexture;
-        _logoSize    = firstWindow->_logoSize;
-    }
-
-    if( !_state ) // happens if first window failed to initialize
-    {
-        setErrorMessage( "No state handler object" );
-        return false;
-    }
-
-    // turn off OpenGL lighting if we are using our own shaders
-    if( _state->getProgram( pipe ) != VertexBufferState::FAILED )
-        glDisable( GL_LIGHTING );
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/windowInit.pdf}
-  {\caption{\small\label{fWindowInit}Window Initialization}}
-\end{wrapfigure}
 The state object is used to handle the creation of OpenGL objects in a
-multipipe, multi-threaded execution environment. It uses an object
-manager, which is described in detail in \sref{sObjectManager}. It is
-used in conjunction with a reference-counting smart-pointer here, since it is
-potentially `owned' by multiple windows at the same time.
+multipipe, multi-threaded execution environment. It uses the object
+manager of the \textsf{eq::Window}, which is described in detail in
+\sref{sObjectManager}.
 
@@
 
-The window exit function de-allocates all OpenGL objects when the state
-object is about to be disposed. The object manager does not delete the
-object in its destructor, since it does not know if an OpenGL context is
-still current. Additionally, \textsf{eq::Window::configExit} is called
-to destroy the drawable and context:
-
-{\footnotesize\begin{lstlisting}
-bool Window::configExit()
-{
-    if( _state.isValid() && _state->getRefCount() == 1 )
+The window exit happens in the reverse order of the
+initialization. First, \textsf{configExitGL} is called to de-initialize
+OpenGL, followed by \textsf{configExitOSWindow} which de-initializes the
+drawable and context and deletes the \textsf{OSWindow} allocated in
+\textsf{configInitOSWindow}.
+
+The window OpenGL exit function of \textsf{eqPly} de-allocates all
+OpenGL objects. The object manager does not delete the object in its
+destructor, since it does not know if an OpenGL context is still
+current.
+
+{\footnotesize\begin{lstlisting}
+bool Window::configExitGL()
+{
+    if( _state )
         _state->deleteAll();
 
+    delete _state;
     _state = 0;
-    return eq::Window::configExit();
+
+    return eq::Window::configExitGL();
 }
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Render Loop}}
+\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Render Loop}}
-\end{wrapfigure}
 The rendering loop maintains a list of candidates to render, which
@@
 
-During this initialization, Equalizer creates and manages two threads
-for communication, the receiver thread and the command thread. Normally
-no application code is executed from these two threads.
-
-The receiver thread manages the network connections to other nodes and
-receives data. It dispatches the received data either to the application
-threads, or to the command thread.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+During this initialization, Equalizer creates and manages two threads
+for communication, the receiver thread and the command thread. Normally
+no application code is executed from these two threads.
+
+The receiver thread manages the network connections to other nodes and
+receives data. It dispatches the received data either to the application
+threads, or to the command thread.
+
 The command thread processes Equalizer-related request from other nodes,
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
 Each entity uses \textsf{waitFrame\-Started} to block on the parent's
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
 \fig{fFrameSync} illustrates the local frame synchronization. By
@@
 window-system-specific task method, and then
-\textsf{Window::configInitGL} to do generic OpenGL state setup, as shown
-in \fig{fWindowInit}.
+\textsf{Window::configInitGL} to do generic OpenGL state setup.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+\end{wrapfigure}
 \fig{fDirectSend} illustrates this algorithm for three channels. The
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
+  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
+\end{wrapfigure}
 The offset of input and output frames characterizes the position of the

### 2008-08-25 Developer/ProgrammingGuide/paper.tex
Programming Guide: Update for OSWindow changes.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.7 for Equalizer 0.5.4+svn\\[\medskipamount]
+  Version 1.3.8 for Equalizer 0.5.4+svn\\[\medskipamount]
   \today
@@
     1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
-    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes (event handling updates still missing)\\
+    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
+    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
     \hline
@@
 event framework, while allowing all necessary customizations. It gathers
-all events in the main thread of the application, so that the developer only
-has to implement \textsf{Config::processEvent} to update its data based
-on the pre-processed, generic keyboard and mouse events. It is very easy
-to use and similar to an GLUT-based implementation.
+all events from all node processes in the main thread of the
+application, so that the developer only has to implement
+\textsf{Config::processEvent} to update its data based on the
+pre-processed, generic keyboard and mouse events. It is very easy to use
+and similar to an GLUT-based implementation.
 
@@
 
-\subsubsection{Initialization and Exit}
-
-During window and pipe initialization the event handling is set up. For
-both entities, \textsf{initEvent\-Handler} is called to register the
-pipe or window with an event handler. This method may be overwritten to
-use a custom event handler, or to not install an event handler at all in
-order to disable event handling. Likewise, \textsf{exitEventHandler} is
-called to de-initialize event handling.
-
-An event handler consists of two parts: the generic base class providing
-the interface and generic functions, and the window-system-specific
-part providing the actual implementation.
-
-Event handling is initialized whenever a new window-system-specific
-pipe or window handle is set. First, \textsf{exitEventHandler} is called
-to de-initialize event handling for the old handle (if set), and then
-\textsf{initEvent\-Handler} is called for the new handle.
-
-AGL uses an event handler singleton, GLX uses one event handler per pipe
-and WGL uses one event handler per window.
-
 \subsubsection{Message Pump}
 
-In order to dispatch the events, Equalizer 'pumps' the native events. On
-WGL and GLX, this has happens on each thread with windows, whereas on
-AGL it has to happen only on the main thread. By default, Equalizer
+In order to dispatch events, Equalizer 'pumps' the native events. On
+WGL and GLX, this happens on each thread with windows, whereas on
+AGL it happens only on the main thread. By default, Equalizer
 pumps these events automatically for the application in-between
@@
 
+Events are received by an event handler. The event handler finds the
+\textsf{eq::OSWindow} associated to the event. It then creates a generic
+\textsf{WindowEvent}, which holds the event data in an independent
+format. The original native event and this generic \textsf{WindowEvent}
+form the \textsf{OSWindowEvent}, which is passed to the concrete
+\textsf{OSWindow} for processing.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
   {\caption{\small\label{fEventProcessing}Event Processing}}
-  \vspace{-3ex}
 \end{wrapfigure}
-Events are received by an event handler. The event handler finds the
-\textsf{eq::Window} associated to the event. It then creates a generic
-\textsf{WindowEvent}, which holds important event data in an independent
-format. The original event is attached to the generic window event.
-
-The event handler then passes the window event to
-\textsf{Window::processEvent}, which is responsible for either handling
+The purpose of the \textsf{OSWindow} processing method,
+\textsf{processEvent}, is to perform window-system specific event
+processing. For example, \textsf{AGLWindow::processEvent} calls
+\textsf{aglUpdateContext} whenever a resize event is received. For
+further processing, the \textsf{WindowEvent} is passed on to
+\textsf{Window::processEvent}. This \textsf{WindowEvent} no longer contains
+the native event.
+
+\textsf{Window::processEvent} is responsible for either handling
 the event locally, or for translating it into a generic
@@
 thread using \textsf{Config::sendEvent}.
-
-If the event was processed by \textsf{processEvent}, the function has to
-return \textsf{true}. If \textsf{false} is returned, the event will be
-passed to a previously installed, window-system-specific event handling
-function. The default implementation of \textsf{Window::pro\-cess\-Event}
-passes most events on to the application.
 
@@
 queued event. By overriding \textsf{handleEvents}, event-driven
-execution can be implemented.
+execution can easily be implemented.
+
+If the event was processed, \textsf{processEvent} has to return
+\textsf{true}. If \textsf{false} is returned to the event handler, the
+event will be passed to the previously installed, window-system-specific
+event handling function.
+
+\fig{fEventProcessing} shows the overall data flow of an event.
+
+\subsubsection{Default Implementation}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
+  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
+\end{wrapfigure}
+Currently, Equalizer provides an \textsf{AGLEvent\-Handler},
+\textsf{GLXEventHandler} and \textsf{WGL\-Event\-Handler}, which handle
+events for an \textsf{AGLWindowIF}, \textsf{GLXWindowIF} and
+\textsf{WGLWindowIF}, respectively. \fig{fEventUML} illustrates the
+class hierarchy for event processing.
+
+The \textsf{OSWindow} implementation is responsible for setting up and
+de-initializing event handling.
+
+Carbon events issued for AGL windows are processed from the process main
+thread. One \textsf{AGLEventHandler} is used. When a Carbon window is
+set on an \textsf{AGLWindow}, the window is registered with the
+\textsf{AGLEventHandler}. The event handler installs a Carbon event
+handler for all important events. The event handler uses an
+\textsf{AGLWindowEvent} to pass the Carbon \textsf{EventRef} to
+\textsf{AGLWindowIF::processEvent}. During window exit, the installed
+Carbon handler is removed when the window is deregistered with the
+\textsf{AGLEventHandler}. No event handling is set up for PBuffers.
+
+GLX uses one X11 Display connection for all windows on a GPU, which is
+managed by the \textsf{eq::Pipe}. For each thread, one
+\textsf{GLXEventHandler} is allocated. The pipe registers the display
+connection with this per-thread event handler. Events for windows
+created using this display connection are automatically received by the
+event handler. No per-window event handling setup needs to be
+done. PBuffers are handled in the same way as window drawables.
+
+The glX event handler iterates over all windows of the registered pipe
+to find the originating \textsf{GLXWindowIF}, which is then used to
+process the event in a fashion similar to AGL. The
+\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
+\textsf{GLXWindowIF::processEvent}.
+
+Each \textsf{WGLWindow} allocates one \textsf{WGLEventHandler} when the
+window handle is set. The WGLEventHandler passes the native event
+parameters \textsf{uMsg}, \textsf{wParam} and \textsf{lParam} to
+\textsf{WGLWindowIF::processEvent} as part of the
+\textsf{WGLWindowEvent}. No event handling is set up for PBuffers.
 
@@
 \textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
-resolution changes.
+resolution changes. It is also likely that an \textsf{OSPipe}
+abstraction, similar to the \textsf{OSWindow}, will be created to
+separate the window system specific code from the \textsf{eq::Pipe}.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
 Each entity uses \textsf{waitFrame\-Started} to block on the parent's
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
 \fig{fFrameSync} illustrates the local frame synchronization. By
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
-\end{wrapfigure}
 \fig{fDirectSend} illustrates this algorithm for three channels. The
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
+  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
+\end{wrapfigure}
 Readback and assemble operations on frames and images are designed to be
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+  \vspace{-1em}
+\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended together to

### 2008-08-29 Developer/ProgrammingGuide/paper.tex
Clean up event interface: - Remove ChannelEvent and WindowEvent: no need to pass pointer of window   and channel, since the receiver is the window or channel - Window::processEvent and Channel::processEvent now receive an eq::Event - rename event.pointerEvent and event.keyEvent to event.pointer and   event.key for consistency with other event names. - Update Programming Guide accordingly
Will test WGL version in a minute.

[ ] May break build
[x] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.8 for Equalizer 0.5.4+svn\\[\medskipamount]
+  Version 1.3.9 for Equalizer 0.5.4+svn\\[\medskipamount]
   \today
@@
     1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
+    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
     \hline
@@
 \textsf{eq::OSWindow} associated to the event. It then creates a generic
-\textsf{WindowEvent}, which holds the event data in an independent
-format. The original native event and this generic \textsf{WindowEvent}
+\textsf{Event}, which holds the event data in an independent
+format. The original native event and this generic \textsf{Event}
 form the \textsf{OSWindowEvent}, which is passed to the concrete
@@
 \textsf{aglUpdateContext} whenever a resize event is received. For
-further processing, the \textsf{WindowEvent} is passed on to
-\textsf{Window::processEvent}. This \textsf{WindowEvent} no longer contains
+further processing, the \textsf{Event} is passed on to
+\textsf{Window::processEvent}. This \textsf{Event} no longer contains
 the native event.

### 2008-10-07 Developer/ProgrammingGuide/paper.tex
Programming Guide: update for 2D pixel compounds
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.9 for Equalizer 0.5.4+svn\\[\medskipamount]
+  Version 1.3.10 for Equalizer 0.5.5\\[\medskipamount]
   \today
@@
     1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
+    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
     \hline
@@
 2D decompositions can be used by all types of applications, but should
-be combined with DB compounds to reduce the data per node, if possible.
+be combined with DB compounds to reduce the data per node, if
+possible. In most cases, a \textsf{loadBalancer} should be used to
+automatically adapt the tiling each frame, based on the current
+rendering load.
+
+2D compounds in Equalizer are configured using the \textsf{viewport}
+parameter, using the values \textsf{[ x y width height]} in normalized
+coordinates. The viewport defines the area of the parent (destination)
+channel to be rendered for each child. Each child compound uses an
+output frame, which is connected to an input frame on the destination
+channel. The destination channel can also be used as a source channel,
+in which case it renders in place and no output frame is needed.
 
@@
 profit from database compounds.
+
+DB compounds in Equalizer are configured using the \textsf{range}
+parameter, using the values \textsf{[ begin end ]} in normalized
+coordinates. The range defines the start and end point of the
+application's database to be rendered. The value has to be interpreted
+by the application's rendering code accordingly.  Each child compound
+uses an output frame, which is connected to an input frame on the
+destination channel. For more than two contributing channels, it is
+recommended to configure parallel direct send compositing, as described
+in \sref{sDirectSend}.
 
@@
 
+Eye compounds in Equalizer are configured using the \textsf{eye}
+parameter, limiting the child to render the \textsf{[ LEFT ]} or
+\textsf{[ RIGHT ]} eye. Each child compound uses an output frame, which
+is connected to an input frame on the destination channel. The
+destination channel can also be used to render an eye pass, in which case
+it renders in the correct stereo buffer and no output frame is needed.
+
 \subsection{Pixel Compounds}
+
+Pixel compounds are similar to 2D compounds. The frustra of the
+source rendering units are modified so that each unit renders an evenly
+distributed subset of pixels.
+
+As 2D compounds, pixels compounds have low, constant IO requirements for
+the pixel transfers during recomposition.
 
@@
 \end{wrapfigure}
-Pixel compounds are similar to 2D compounds. The frustra of the
-source rendering units are modified so that each unit renders an evenly
-distributed subset of pixels.
-
-As 2D compounds, pixels compounds have low, constant IO requirements for
-the pixel transfers during recomposition.
-
-Pixel compounds work well for purely fill-limited applications,
-techniques like frustum culling do not reduce the rendered data for the
+Pixel compounds work well for purely fill-limited applications.
+Techniques like frustum culling do not reduce the rendered data for the
 source rendering resources.
@@
 attention. Among them are: lines, points, sprites, glDrawPixels,
-glBitmap, glPolygonStipple.
+glBitmap, glPolygonStipple. The application can query the current pixel
+parameters at run-time to adjust the rendering accordingly.
 
@@
 and needs the full database for rendering anyway. Volume rendering
-applications should choose this mode over 2D compounds.
+applications are also well suited for this mode, and should choose it
+over 2D compounds.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
+  {\caption{\label{fPixelKernel}\small Pixel compound kernel}}
+\end{wrapfigure}
+Pixel compounds in Equalizer are configured using the \textsf{pixel}
+parameter, using the values \textsf{[ x y width height]}  to configure
+the size and offset of the sampling kernel. The width and height of the
+sampling kernel define how many pixels are skipped in the x and y
+direction, respectively. The x and y offset define the index of the
+source channel within the kernel. They have to be smaller than the size
+of the kernel. \fig{fPixelKernel} illustrates these parameters, and
+\fig{fPixelKernels} shows some example kernels for a four-to-one pixel decomposition.
+
+The destination channel can also be used as a source channel. Contrary
+to the other compound modes, it also has to use an output and
+corresponding input frame. During rendering, the frustum is 'squeezed'
+to configure the pixel decomposition. The destination channel can
+therefore not be rendered in place, like with the other compound modes.
+\begin{figure}[ht!]\center
+  \includegraphics[width=.618\textwidth]{images/pixelKernels.pdf}
+  {\caption{\small\label{fPixelKernels}Example pixel kernels for a
+      four-to-one pixel compound}}
+\end{figure}
 
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \vspace{-2em}\includegraphics[width=.382\textwidth]{images/modelDist.pdf}
+  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
+  \vspace{-2em}
+\end{wrapfigure}
 Some applications might rely on a shared filesystem to access the data,
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
-  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
-\end{wrapfigure}
 The identifer of the root object of this distributed hierarchy is passed
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/objects.pdf}
+  {\caption{\small\label{fObjects}Scene Graph Distribution}}
+\end{wrapfigure}
 Each \textsf{VertexBufferDist} object has one node of the model's data
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/objects.pdf}
-  {\caption{\small\label{fObjects}Scene Graph Distribution}}
-\end{wrapfigure}
 Applications distributing a dynamic scene graph use the frame data
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
-  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
-\end{wrapfigure}
 All window-system specific functionality is implemented by a
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
+  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
+\end{wrapfigure}
 The initialization sequence uses multiple, overrideable task
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/DB.png}
-  {\caption{\small\label{fDB}Destination view of a DB compound}}
-\end{wrapfigure}
 After the basic view setup, a directional light is configured, and the
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Render Loop}}
-\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\small\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 The purpose of the \textsf{OSWindow} processing method,
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\small\label{fEventProcessing}Event Processing}}
+\end{wrapfigure}
 \textsf{Window::processEvent} is responsible for either handling
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
-\end{wrapfigure}
 Currently, Equalizer provides an \textsf{AGLEvent\-Handler},
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
+  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
+\end{wrapfigure}
 The \textsf{OSWindow} implementation is responsible for setting up and
@@
 
-\subsubsection{Parallel Direct Send Compositing}
+\subsubsection{\label{sDirectSend}Parallel Direct Send Compositing}
 

### 2008-10-14 Developer/ProgrammingGuide/paper.tex
Programming Guide: Update for net::Object cleanups
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.10 for Equalizer 0.5.5\\[\medskipamount]
+  Version 1.3.11 for Equalizer 0.5.5+svn\\[\medskipamount]
   \today
@@
     1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
+    1.3.11  & Oct 13, 2008 & Updated to reflect eq::net::Object cleanup\\
     \hline
@@
 \begin{figure}[ht!]\center
-  \includegraphics[width=.618\textwidth]{images/pixelKernels.pdf}
+  \includegraphics[width=.9\textwidth]{images/pixelKernels.pdf}
   {\caption{\small\label{fPixelKernels}Example pixel kernels for a
@@
 different stereo passes. A \textsf{pixel} parameter adapts the frustum
-so that the source channel renders every n$th$ line out of m
-lines. Setting one or multiple attributes causes the parent's view to be
-decomposed accordingly. Attributes are cumulative, that is, intermediate
-compound attributes affect and therefore decompose the rendering of all
-their children.
+so that the source channel renders an evenly subset of the parent's
+pixels. Setting one or multiple attributes causes the parent's view to
+be decomposed accordingly. Attributes are cumulative, that is,
+intermediate compound attributes affect and therefore decompose the
+rendering of all their children.
 
@@
 Equalizer provides distributed objects which help implementing data
-distribution in a cluster environment. The master version of a distributed
-object is registered with a \textsf{eq::net::Session}, which assigns a
-session-unique identifier to the object. Other nodes can map their
-instance of the object to this identifier, thus synchronizing the
-object's data with the remotely registered master version.
+distribution in a cluster environment. The master version of a
+distributed object is registered with a \textsf{eq::net::Session}, which
+assigns a session-unique identifier to the object. This identifier can
+be used by other nodes to map their instance of the object, thus
+synchronizing the object's data with the remotely registered master
+version.
 
@@
 
-A static distributed object either has to provide a pointer and size to
-its data using \textsf{setInstanceData}, or it has to implement
-\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
-approach can be used if all distributed member variables are stored in
-one contiguous block of memory. The second approach is used otherwise.
-
-The \textsf{InitData} class contains a string of variable
-length. Therefore it uses the second approach of manually serializing
-and de-serializing its data. Serialization in \textsf{getInstanceData}
+A static distributed object has to implement \textsf{getInstanceData}
+and \textsf{applyInstanceData} to serialize and deserialize the object's
+distributed data. These method provide an output or input stream as a
+parameter, which abstract the data transmission and can be used like a
+\textsf{std::stream}.
+
+For \textsf{InitData}, serialization in \textsf{getInstanceData}
 and de-serialization in \textsf{applyInstanceData} is performed by
-streaming all member variable to or from the provided data
+streaming all member variables to or from the provided data
 streams. Efficient buffering and data transport between nodes is
@@
 Versioned objects have to override \textsf{getChangeType} to indicate
-how they want to have changes to be handled. The current
-implementation has the following characteristics:
+how they want to have changes to be handled. All versioned objects
+currently implemented have the following characteristics:
 \begin{itemize}
@@
 \item Newly mapped slave instance are mapped to the oldest available
-  version.
+  version by default, or to the specified version in
+  \textsf{Session::mapObject}.
 \end{itemize}
@@
 
-In addition to the instance data (de)serialization methods used to map
+In addition to the instance data (de-)serialization methods used to map
 an object, versioned objects may implement \textsf{pack} and
 \textsf{unpack} to serialize or de-serialize the changes since the last
-version.
-
-If the delta data happens to be layed out contiguously in memory,
-\textsf{setDeltaData} might be used. The default implementation of
-\textsf{pack} and \textsf{unpack} (de)serialize the delta data or the
-instance data if no delta data has been specified.
-
-The \textsf{eqPly} frame data is layed out in one anonymous structure in
-memory. It also does not track changes since it is relatively small in
-size and changes frequently. Therefore, the instance and delta data are
-the same and set in the constructor. Furthermore, the change type is
-\textsf{INSTANCE}. The default \textsf{Object} implementation will take
-care of the distribution of the data:
-
-{\footnotesize\begin{lstlisting}
-        FrameData()
-            {
-                reset();
-                setInstanceData( &data, sizeof( Data ));
-                EQINFO << "New FrameData " << std::endl;
+version. If these methods are not implemented, their default
+implementation forwards the (de-)serialization request to
+\textsf{getInstanceData} and \textsf{applyInstanceData}, respectively.
+
+The \textsf{eqPly} does not track changes since it is relatively small
+in size and changes frequently. Therefore, the instance and delta data
+are the same, and \textsf{pack} and \textsf{unpack} are not
+implemented. The change type is \textsf{INSTANCE}. The serialization
+functions exploit the fact that the frame data is layed
+out in one anonymous structure in memory, and transmit all the data in
+one call:
+
+{\footnotesize\begin{lstlisting}
+void FrameData::getInstanceData( eq::net::DataOStream& os )
+{ 
+    os.writeOnce( &data, sizeof( data )); 
+}
+
+void FrameData::applyInstanceData( eq::net::DataIStream& is )
+{
+    memcpy( &data, is.getRemainingBuffer(), sizeof( data ));
+    is.advanceBuffer( sizeof( data ));
 }
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-2em}\includegraphics[width=.382\textwidth]{images/modelDist.pdf}
-  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
-  \vspace{-2em}
-\end{wrapfigure}
 Some applications might rely on a shared filesystem to access the data,
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/objects.pdf}
-  {\caption{\small\label{fObjects}Scene Graph Distribution}}
-\end{wrapfigure}
 Each \textsf{VertexBufferDist} object has one node of the model's data
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
-  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
-\end{wrapfigure}
 The initialization sequence uses multiple, overrideable task
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
+  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
+\end{wrapfigure}
 The \textsf{OSWindow} initialization takes into account various
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Render Loop}}
+  \vspace{-5ex}
+\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 for efficient view frustum culling. When the model is loaded by
-\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
-node of the tree gets a database range assigned. The root node
-has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
-1], and so on for all nodes in the tree. The preprocessed model is saved
-in a binary format for accelerating subsequent use.
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Render Loop}}
-\end{wrapfigure}
+\textsf{Node::config\-Init}, it is preprocessed into the kD-tree During
+this preprocessing step, each node of the tree gets a database range
+assigned. The root node has the range [0, 1], its left child [0, 0.5]
+and its right child [0.5, 1], and so on for all nodes in the tree. The
+preprocessed model is saved in a binary format for accelerating
+subsequent loading.
+
 The rendering loop maintains a list of candidates to render, which
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
-\end{wrapfigure}
-The \textsf{OSWindow} implementation is responsible for setting up and
-de-initializing event handling.
-
-Carbon events issued for AGL windows are processed from the process main
-thread. One \textsf{AGLEventHandler} is used. When a Carbon window is
-set on an \textsf{AGLWindow}, the window is registered with the
-\textsf{AGLEventHandler}. The event handler installs a Carbon event
-handler for all important events. The event handler uses an
+The concrete implementation of these window interfaces is responsible
+for setting up and de-initializing event handling.
+
+Carbon events issued for AGL windows are processed from the node's main
+thread. One \textsf{AGLEventHandler} per process is used. When a Carbon
+window is set on an \textsf{AGLWindow}, the window is registered with
+the global \textsf{AGLEventHandler}. The event handler installs a Carbon
+event handler for all important events. The event handler uses an
 \textsf{AGLWindowEvent} to pass the Carbon \textsf{EventRef} to
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
+  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
+\end{wrapfigure}
 The glX event handler iterates over all windows of the registered pipe
@@
 \textsf{GLXWindowEvent} passes the \textsf{XEvent} to
-\textsf{GLXWindowIF::processEvent}.
+\textsf{GLXWindowIF::\-process\-Event}.
 
@@
 parameters \textsf{uMsg}, \textsf{wParam} and \textsf{lParam} to
-\textsf{WGLWindowIF::processEvent} as part of the
+\textsf{WGL\-Win\-dowIF::processEvent} as part of the
 \textsf{WGLWindowEvent}. No event handling is set up for PBuffers.
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/threads.pdf}
-  {\caption{\small\label{fThreads}Threads within one node process}}
-\end{wrapfigure}
 During this initialization, Equalizer creates and manages two threads
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/threads.pdf}
+  {\caption{\small\label{fThreads}Threads within one node process}}
+\end{wrapfigure}
 The receiver and command thread are terminated when the application
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
 The application has extended control over the task synchronization
@@
 This section explains window initialization in detail. It discusses in
-detail the handling on the different window systems. The entry point for
-the window initialization is the task method
-\textsf{Window::configInit}. This task method first calls a
-window-system-specific task method, and then
-\textsf{Window::configInitGL} to do generic OpenGL state setup.
+detail the handling of the different window systems. Window systems are
+integrated through the \textsf{OSWindow} interface, which externalizes
+the details of the windowing API from the \textsf{eq::Window}
+implementation and facilitates the integration with new windowing API's
+and custom applications.
+
+Equalizer provides sample implementations for the supported window
+systems. These sample implementations are intended to be sub-classed and
+various steps in the window initialization can be overwritten and
+customized.
+
+An application typically choses to subclass the sample implementation if
+only minor tweaks are needed for integration. For major changes or new
+window systems, it is often easier to subclass directly from
+\textsf{OSWindow} and implement the abstract methods of this interface
+class.
+
+The method \textsf{Window::configInitOSWindow} is used to instantiate
+and initialize the \textsf{OSWindow} implementation during config
+initialization. After an successful \textsf{OSWindow} initialization,
+\textsf{Window::configInitGL} is called for the generic OpenGL state
+setup.
 
 Since window initialization is notoriously error-prone and hard to
-debug, the default Equalizer functions propagate the reason for errors
+debug, the sample implementation propagates the reason for errors
 from the render clients back to the application. The \textsf{Pipe} and
@@
 
-The window-system-specific initialization methods use overrideable
-methods for all sub-tasks. This allows partial customization, without
-the need of rewriting tedious window initialization code, e.g., the
-OpenGL pixel format selection.
+The sample implementations \textsf{AGLWindow}, \textsf{GLXWindow} and
+\textsf{WGLWindow} all have similar, overrideable methods for all
+sub-tasks. This allows partial customization, without the need of
+rewriting tedious window initialization code, e.g., the OpenGL pixel
+format selection.
 
@@
 the feature and then to back off and retry if it is not available. The
-returned pixel format has to be destroyed using
-\textsf{Window::destroyAGLPixelFormat}. When no matching pixel format is
-found, \textsf{chooseAGLPixelFormat} returns \textsf{0} and the AGL
-window initialization returns with a failure.
+pixel format returned by \textsf{chooseAGLPixelFormat} has to be
+destroyed using \textsf{destroyAGLPixelFormat}. When no matching pixel
+format is found, \textsf{chooseAGLPixelFormat} returns \textsf{0} and
+the AGL window initialization returns with a failure.
 
@@
 {\footnotesize\begin{lstlisting}
-bool Window::configInitAGL()
+bool AGLWindow::configInit( )
 {
@@
 {\footnotesize\begin{lstlisting}
-bool Window::configInitGLX()
+bool GLXWindow::configInit( )
 {
@@
     {
-        setErrorMessage( "configInitGLXDrawable did set no X11 drawable" );
+        _window->setErrorMessage( 
+            "configInitGLXDrawable did not set a X11 drawable");
         return false;
@@
   function pointer to delete the returned device context is set as
-  well. This function might return 0, which is not an error.
+  well. This function might return 0, which is not an error, but
+  signifies that no affinity context is used.
 \item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
@@
 {\footnotesize\begin{lstlisting}
-bool Window::configInitWGL()
+bool WGLWindow::configInit()
 {
@@
             deleteDCProc( dc );
-        setErrorMessage( "configInitWGLDrawable did not set a WGL drawable" );
+        _window->setErrorMessage(
+            "configInitWGLDrawable did not set a WGL drawable" );
         return false;
@@
     setWGLContext( context );
+
+    if( getIAttribute( Window::IATTR_HINT_SWAPSYNC ) != AUTO )
+    {
+        if( WGLEW_EXT_swap_control )
+        {
+            // set vsync on/off
+            const GLint vsync = 
+                ( getIAttribute( Window::IATTR_HINT_SWAPSYNC )==OFF ) ? 0 : 1;
+            wglSwapIntervalEXT( vsync );
+        }
+        else
+            EQWARN << "WGLEW_EXT_swap_control not supported, ignoring window "
+                   << "swapsync hint" << std::endl;
+    }
 
@@
 
+When using 2D (sort-first) or stereo decompositions, no special
+programming is needed to achieve good scalability, as \textsf{eVolve} is
+mostly fill-limited and therefore scales nicely in these modes.
+
+The full power of scalable volume rendering is however in DB (sort-last)
+compounds, where the full volume is divided into separate bricks. Each
+of the bricks is rendered like a separate volume. For recomposition, the
+\textsf{RGBA} frame buffer data resulting from these render passes then
+has to be assembled correctly.
+
+Conceptually, the individual volume bricks of each of the source
+channels produces pixel data which can be handled like one big 'slice'
+through the full texture. Therefore they have to be blen\-ded
+back-to-front in the same way as the slice planes are blended during
+rendering.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
- When using 2D (sort-first) or stereo decompositions, no special
-programming is needed to achieve good scalability, as \textsf{eVolve} is
-mostly fill-limited and therefore scales nicely in these modes.
-
-The full power of scalable volume rendering is however in DB (sort-last)
-compounds, where the full volume is divided into separate bricks. Each
-of the bricks is rendered like a separate volume. For recomposition, the
-\textsf{RGBA} frame buffer data resulting from these render passes then
-has to be assembled correctly.
-
-Conceptually, the individual volume bricks of each of the source
-channels produces pixel data which can be handled like one big 'slice'
-through the full texture. Therefore they have to be blen\-ded
-back-to-front in the same way as the slice planes are blended during
-rendering.
-
 DB compounds have the advantage of scaling any part of the volume
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-  \vspace{-1em}
-\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended together to
@@
 \end{figure}
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+  \vspace{-2ex}
+\end{wrapfigure}
 For orthographic rendering, determining the compositing order of the
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/statistics.png}
-  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
-\end{wrapfigure}
 Statistics are measured in milliseconds since the configuration was
@@
 to the parent compound. The range restricts the database range, relative
-to the parent. The pixel setting selects the pixel decomposition
-parameters, relative to the parent. The eye pass select the mono, left
-or right eye passes rendered by the compound. All these attributes are
-inherited by the children of a compound. Viewport, range and pixel
-parameters are cumulative.
+to the parent. The pixel setting selects the pixel decomposition kernel,
+relative to the parent. The eye pass select the mono, left or right eye
+passes rendered by the compound. All these attributes are inherited by
+the children of a compound. Viewport, range and pixel parameters are
+cumulative.
 

### 2008-10-16 Developer/ProgrammingGuide/paper.tex
Programming Guide: update for 2D pixel compounds
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.11 for Equalizer 0.5.5+svn\\[\medskipamount]
+  Version 1.3.12 for Equalizer 0.5.5+svn\\[\medskipamount]
   \today
@@
 
-\textcopyright 2007-2008 Eyescale Software GmbH. All rights reserved. No
-permission is granted to copy, distribute, or create derivative works
-from the contents of this electronic documentation in any manner, in
-whole or in part, without the prior written permission of Eyescale
-Software GmbH.
+\textcopyright 2007-2008 \htmladdnormallink{Eyescale Software GmbH}
+{http://www.eyescale.ch}. All rights reserved. No permission is
+granted to copy, distribute, or create derivative works from the
+contents of this electronic documentation in any manner, in whole or in
+part, without the prior written permission of Eyescale Software GmbH.
 
@@
 this programming guide, please contact
-\htmladdnormallink{eile@equalizergraphics.com}
-{mailto:eile@equalizergraphics.com?subject=Equalizer\%20Programming\%20Guide}.
+\htmladdnormallink{eile@eyescale.ch}
+{mailto:eile@eyescale.ch?subject=Equalizer\%20Programming\%20Guide}.
 
@@
     1.3.11  & Oct 13, 2008 & Updated to reflect eq::net::Object cleanup\\
+    1.3.12  & Oct 16, 2008 & Added \sref{sView}\\
     \hline
@@
 
-\subsubsection{Frustum}
+\subsubsection{\label{sFrustum}Frustum}
 Compounds have a frustum description to define the physical layout of
@@
 
+\subsection{\label{sView}View Management}
+
+A view describes the projection of the scene onto a coherent 2D area,
+typically a display or window rendered by a destination channel. The
+application has read and write access to all the views of a
+configuration.
+
+Changing view parameters is typically done for non-fullscreen
+application windows, where the rendering is not meant to be viewed in
+real-world size in an immersive environment. A typical use case is
+changing the field-of-view of the rendering.
+
+A configuration has one or more views. A view is determined by the
+frustum (wall or projection description) and eye separation. A view is
+uniquely assigned to one compound and possibly one destination
+channel. For a description of wall and projection parameters please
+refer to \sref{sFrustum}.
+
+The application node can query all views of the \textsf{eq::Config}. It
+can change all parameters of each view (wall, projection and eye
+separation). The new view parameters are applied for the next frame.
+
+Destination channels on the render clients have access to a read-only
+version of their view. The view data is automatically kept up-to-date
+with the application's instance by Equalizer.
+
+The view API is not finalized and may change in the future.
+
+\subsubsection{View Update}
+
+The default Equalizer event handling is using the view API to maintain
+the aspect ratio of destination channels after a window resize. Without
+updating the wall or projection description, the rendering would become distorted.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
+  {\caption{\small\label{fViewUpdate}Event Flow during a View Update}}
+\end{wrapfigure}
+When a window is resized, a \textsf{CHANNEL\_RESIZE} event is eventually
+generated. If the corresponding channel has a view,
+\textsf{Channel::processEvent} sends a \textsf{VIEW\_RESIZE} event to
+the application. This event contains the identifier of the view.
+
+The config event is dispatched to \textsf{Config::handleView\-Resize},
+which uses the view identifier to find the corresponding view. Based on
+the original size and wall or projection description of the view, a new
+wall or projection is computed, keeping the aspect ratio and  the height
+of the frustum constant. This new frustum is automatically applied by
+Equalizer at the next config frame.
+
+\fig{fViewUpdate} shows a sequence diagram of such a view update.
+
 \subsection{\label{sCompositing}Image Compositing for Scalable Rendering}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+  \vspace{-2ex}
+\end{wrapfigure}
 After performing the sort-last compositing, the color information is
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
-  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
-\end{wrapfigure}
 The frame synchronization through the frame data allows the input frame
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
+      Volume Rendering}}
+\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
-      Volume Rendering}}
-\end{wrapfigure}
 DB compounds have the advantage of scaling any part of the volume
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-  \vspace{-2ex}
-\end{wrapfigure}
 For orthographic rendering, determining the compositing order of the

### 2008-10-17 Developer/ProgrammingGuide/paper.tex
Implement event-driven execution for eqPly. Fix AGL event handling. Remove unneeded Client::useMessagePump and use existing Pipe::useMessagePump instead.
[ ] May break build
[ ] Breaks existing applications
[x] Bugfix
[x] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.12 for Equalizer 0.5.5+svn\\[\medskipamount]
+  Version 1.3.13 for Equalizer 0.5.5+svn\\[\medskipamount]
   \today
@@
     1.3.12  & Oct 16, 2008 & Added \sref{sView}\\
+    1.3.13  & Oct 17, 2008 & Minor update in \sref{sMessagePump}\\
     \hline
@@
 
-\subsubsection{Message Pump}
+\subsubsection{\label{sMessagePump}Message Pump}
 
@@
 
-The methods \textsf{Client::useMessagePump} and
-\textsf{Pipe::useMessagePump} can be overridden to return \textsf{false}
-to disable this behaviour for their respective threads. On non-threaded
-pipes, \textsf{Pipe::useMessagePump} is not called.
+The method \textsf{Pipe::useMessagePump} can be overridden to return
+\textsf{false} to not enable message dispatch for the respective thread.
+
+The method \textsf{useMessagePump} is called by Equalizer during
+application and pipe thread initialization (before configInit!) to
+determine if the pipe's thread shall automatically dispatch OS
+events. For non-threaded and AGL pipes, the method affects the node
+thread. On WGL and GLX, messages are dispatched in the application and
+each pipe thread. On AGL, events are dispatched only on the application
+thread.
 

### 2008-10-21 Developer/ProgrammingGuide/paper.tex
Programming Guide: Minor tweak.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.13 for Equalizer 0.5.5+svn\\[\medskipamount]
+  Version 1.3.14 for Equalizer 0.5.5+svn\\[\medskipamount]
   \today
@@
     1.3.13  & Oct 17, 2008 & Minor update in \sref{sMessagePump}\\
+    1.3.14  & Oct 21, 2008 & Improved \sref{sConnectionDescription}\\
     \hline
@@
 
-The hostname is the IP address or (resolvable) host name. When used for
-the server, the server binds its listening socket only to the specified
-name.
-
-A server or node may have multiple connection descriptions. A server
-binds to all provided connection descriptions. For a node, all
-connection descriptions are used while trying to establish a connection
-to the node. When auto-launched by the server, all connection
-descriptions are passed to the launched node process, which binds to all
-provided descriptions.
+The hostname is the IP address or (resolvable) host name. A server or
+node may have multiple connection descriptions.
+
+A server listens on all provided connection descriptions. If no hostname
+is specified for a server connection description, it listens to
+INADDR\_ANY, and is therefore reachable on all interfaces. If the
+server's hostname is specified, the listening socket is bound to this
+address. If any hostname given is not resolvable, or any port can not be
+used, server initialization will fail.
+
+For a node, all connection descriptions are used while trying to
+establish a connection to the node. When auto-launched by the server,
+all connection descriptions of the node are passed to the launched node
+process, which binds to all provided descriptions.
 

### 2008-10-27 Developer/ProgrammingGuide/paper.tex
Programming Guide: Add DPlex documentation
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.14 for Equalizer 0.5.5+svn\\[\medskipamount]
+  Version 1.3.15 for Equalizer 0.5.5+svn\\[\medskipamount]
   \today
@@
     1.3.14  & Oct 21, 2008 & Improved \sref{sConnectionDescription}\\
+    1.3.15  & Oct 27, 2008 & Added \sref{sDPlex}\\
     \hline
@@
 
+2D decomposes the rendering in screen-space, that is, each contributing
+rendering unit processes a tile of the final view. The recomposition
+simply assembles the tiles side-by-side on the destination view. This
+mode is also known as sort-first or SFR.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-2D or sort-first decomposes the rendering in screen-space, that is, each
-contributing rendering unit processes a tile of the final view. The
-recomposition simply assembles the tiles side-by-side on the destination
-view.
-
 The advantage of this mode is a low, constant IO overhead for the pixel
@@
 \end{wrapfigure}
-DB or sort-last\footnote{3D model courtesy of AVS, USA.} decomposes the
+DB\footnote{3D model courtesy of AVS, USA.} decomposes the
 rendered database so that all rendering units process a part of the
-scene in parallel.
+scene in parallel. This mode is also known as sort-last.
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
-  {\caption{\small A stereo compound}}
-\end{wrapfigure}
 Stereo compounds\footnote{3D model courtesy of Stereolithography Archive
@@
 units. The resulting images are copied to the appropriate stereo
-buffer. This modes supports a variety of stereo modes, including active
+buffer. This mode supports a variety of stereo modes, including active
 (quad-buffered) stereo, anaglyphic stereo and auto-stereo displays with
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
+  {\caption{\small A Stereo compound}}
+\end{wrapfigure}
 Due to the frame consistency between the eye views, this modes scales
@@
 it renders in the correct stereo buffer and no output frame is needed.
+
+\subsection{\label{sDPlex}DPlex Compounds}
+
+DPlex compounds\footnote{3D model courtesy of Cyberware, Inc.} assign
+full, alternating frames to individual rendering units. The resulting
+images are copied to the destination channel, and Equalizer
+load-balancing is used to ensure a steady framerate on the destination
+window. This mode is also known as time-multiplex or AFR.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
+  {\caption{\small A DPlex compound}}
+\end{wrapfigure}
+
+Due to the frame consistency between consecutive frames, this mode scales
+very well. The IO requirements for pixel transfer are small and
+constant.
+
+DPlex requires a latency of at least n frames. This increased latency
+might be disturbing to the user, but it is often compensated by the
+higher frame rate. The frame rate typically increases linearly with the
+number of source channels, and therefore linearly with the latency.
+
+DPlex compounds in Equalizer are configured using the \textsf{period}
+and \textsf{phase} parameter, limiting each child to render a subset of
+the frames. Each child compound uses an output frame, which is connected
+to an input frame on the destination channel. The destination channel
+uses a DPlex load-balancer to smoothen the framerate.
 
@@
 
+The frustum description is inherited by the children, therefore the
+frustum is typically defined on the top-most compound.
+
+The frustrum can be specified as a wall or projection description.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The frustum description is inherited by the children, therefore the
-frustum is typically defined on the top-most compound.
-
-The frustrum can be specified as a wall or projection description.
-
 A wall is completely defined by the bottom-left, bottom-right and
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
+  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
+\end{wrapfigure}
 The tasks are executed by the node and pipe threads in parallel. The
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
+  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
+\end{wrapfigure}
 All window-system specific functionality is implemented by a
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
-  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
-\end{wrapfigure}
 The \textsf{OSWindow} initialization takes into account various
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Render Loop}}
-  \vspace{-5ex}
-\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\small\label{fEventProcessing}Event Processing}}
+\end{wrapfigure}
 The purpose of the \textsf{OSWindow} processing method,
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\small\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 \textsf{Window::processEvent} is responsible for either handling
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
+  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
+\end{wrapfigure}
 The concrete implementation of these window interfaces is responsible
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
-\end{wrapfigure}
 The glX event handler iterates over all windows of the registered pipe
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/threads.pdf}
+  {\caption{\small\label{fThreads}Threads within one node process}}
+\end{wrapfigure}
 During this initialization, Equalizer creates and manages two threads
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/threads.pdf}
-  {\caption{\small\label{fThreads}Threads within one node process}}
-\end{wrapfigure}
 The receiver and command thread are terminated when the application
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
 The application has extended control over the task synchronization
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+\end{wrapfigure}
 The main idea behind direct send is to parallelize the costly
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
-  \vspace{-2ex}
-\end{wrapfigure}
 After performing the sort-last compositing, the color information is
@@
 relative to the parent. The eye pass select the mono, left or right eye
-passes rendered by the compound. All these attributes are inherited by
-the children of a compound. Viewport, range and pixel parameters are
+passes rendered by the compound. The DPlex period defines that
+$\frac{1}{period}$ frames are rendered, and the phase defines when in the period
+the rendering starts. All these attributes are inherited by the children
+of a compound. Viewport, range, pixel and period parameters are
 cumulative.
+
+A load-balancer is used to automatically optimize the decomposition. A
+2D, horizontal or vertical load-balancer adapts the viewport of all
+direct children of the compound each frame. A DB load-balancer adapts
+the range of all direct children. A DPlex load-balancer smoothens the
+framerate of the compound's window to produce a steady output framerate.
 
@@
     eye      [ CYCLOP LEFT RIGHT ]   // monoscopic or stereo view
+    period   int                     // DPlex period, every nth frame rendered
+    phase    int                     // DPlex phase, when to start rendering
+
+    loadBalancer
+    {
+        mode   2D | HORIZONTAL | VERTICAL | DB | DPLEX
+    }
+
     attributes

### 2008-11-10 Developer/ProgrammingGuide/paper.tex
Programming Guide: Threading model updates
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 \usepackage{tabularx,longtable,graphicx,a4,listings,wrapfig,subfigure,textcomp}
+\usepackage[absolute]{textpos}
 
@@
 \date{
-  \vspace{2cm}
-  \begin{figure}[ht]
-    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
-  \end{figure}
-  \vspace{1cm}
+  \vspace{15cm}
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.15 for Equalizer 0.5.5+svn\\[\medskipamount]
+  Version 1.3.16 for Equalizer 0.6-RC1\\[\medskipamount]
   \today
@@
 \maketitle
+\begin{textblock}{}(0,5)
+  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
+\end{textblock}
 
@@
 
-\subsection*{Front Page}
+\subsection*{Front and Back Page}
 
@@
   Graphics Laboratory} [bottom right].
+
+The images on the back page show the following scalable rendering modes:
+2D (sort-first, SFR), DB (sort-last), stereo, pixel and DPlex
+(time-multiplex, AFR).
 
@@
     1.3.15  & Oct 27, 2008 & Added \sref{sDPlex}\\
+    1.3.16  & Nov 10, 2008 & Updated \sref{sThreadModel}, added back page\\
     \hline
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/model.pdf}
-  {\caption{\small\label{fModel}Simplified execution model}}
-\end{wrapfigure}
 Using threading correctly in OpenGL-based applications is easy with
@@
 the end of a frame and the beginning of a new frame. Furthermore, only
-one instance of the application data has to be maintained within a
-process since all rendering threads are guaranteed to draw the same frame.
+one instance of the scene data has to be maintained within a process
+since all rendering threads are guaranteed to draw the same frame.
 
@@
 multi-buffers the dynamic parts of the database can disable the per-node
-frame synchronization, as explained in \sref{sNodeFrame}. Some scene
-graphs for example do implement multi-buffered data, and can profit from
-relaxing the frame synchronization.
+frame synchronization, as explained in \sref{sThreadModel}. Some scene
+graphs implement multi-buffered data, and can profit from relaxing the
+frame synchronization.
 
@@
 do not need this per-node frame synchronization can disable it, as
-explained in \sref{sNodeFrame}.
+explained in \sref{sThreadModel}.
 
@@
 applications as easy as possible, as described in \sref{sExecution}. The
-default, per-node thread synchronization provided by Equalizer can
-gradually be relaxed by advanced applications to gain better performance
-through higher asynchronicity.
-
-This section explains the multi-threading in Equalizer in detail and
-gives advice on how to optimize applications for performance.
+default, per-node thread synchronization provided by Equalizer can be
+relaxed by advanced applications to gain better performance through
+higher asynchronicity.
 
@@
 
-\subsubsection{\label{sNodeFrame}Thread Synchronization Model}
-
-The node main thread and the pipe threads are synchronized with each
-other, so that all draw task methods of one rendered frame are executed
-at the same time. This model allows to use the same database for
-rendering, and safe modifications of this database are possible from the
-node thread, since the pipe threads do not execute any rendering tasks
-between frames.
-
-Please note that the default thread synchronization synchronizes all
-\textsf{Channel::frame\-Draw} operations on a \textbf{single} node with
-the node's main thread. The per-node frame synchronization does not
-break the asynchronous execution across nodes. Advanced applications can
-remove the per-node frame synchronization.
+\subsubsection{\label{sThreadModel}Thread Synchronization Models}
+Equalizer supports three threading models, which can be set
+programmatically by the application or through the configuration file
+format. Applications typically hard-code their threading model. The file
+format is commonly used to change the threading model for benchmarking
+and experimentation.
+
+The following thread synchronization models are implemented:
+\begin{description}
+\item[ASYNC:] No synchronization happens between render threads, except
+  for synchonizing the finish of frame \textsf{current-latency}. The
+  \textsf{eqPly} and \textsf{eVolve} Equalizer examples use this
+  threading model, by setting it in \textsf{Node::configInit}. This
+  synchonization model provides the best performance and should be used
+  by all applications which multi-buffer all non-static, frame-specific
+  data.
+\item[DRAW\_SYNC:] In addition to async thread model, all local render
+  threads are synchronized, so that the draw operations happen
+  synchronously with the node main loop. This model allows to use the
+  same database for rendering, and safe modifications of this database
+  are possible from the node thread, since the pipe threads do not
+  execute any rendering tasks between frames. This is the default
+  threading model. This threading model should be used by applications
+  which keep one copy of the scene graph per node.
+\item[LOCAL\_SYNC:] In addition to the async threas model, all local
+  frame operations, including readback, assemble and swap buffer are
+  synchronized with the node main loop. This threading model should be
+  used by applications which need to access non-buffered, frame-specific
+  data after rendering, e.g., during \textsf{Channel::frameAssemble}.
+\end{description}
+
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
+  {\caption{\small\label{fThreadModels}Async, draw\_sync and local\_sync
+    thread sunchronization models}}
+\end{figure}
+
+\fig{fThreadModels} illustrates the synchronization and task execution
+for the thread synchronization models. Please note that the thread
+synchronization synchronizes all pipe render threads on a
+\textbf{single} node with the node's main thread. The per-node frame
+synchronization does not break the asynchronous execution across
+nodes.
+
+\subsubsection{Thread Synchronization in Detail}
 
@@
 unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
-unlock. For the default synchronization model, Equalizer uses the task
-method \textsf{frameDrawFinish} which is called on each resource
-after the last \textsf{Channel::frameDraw} invocation for this
-frame. Consequently, \textsf{Pipe::frameDrawFinish} calls
-\textsf{Pipe::releaseFrame\-Lo\-cal} to signal that it is done drawing the
-current frame, and \textsf{Node::frameDrawFinish} calls
+unlock. For the default synchronization model \textsf{sync\_draw},
+Equalizer uses the task method \textsf{frameDrawFinish} which is called
+on each resource after the last \textsf{Channel::frameDraw} invocation
+for this frame. Consequently, \textsf{Pipe::frameDraw\-Fi\-nish} calls
+\textsf{Pipe::\-release\-Frame\-Lo\-cal} to signal that it is done
+drawing the current frame, and \textsf{Node::frameDrawFinish} calls
 \textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
@@
 
-\fig{fFrameSync} illustrates the local frame synchronization. By
-removing the calls to \textsf{waitFrameLocal}, the node thread can
-disable the synchronization with the pipe threads, which causes the node
-operations to overlap with the rendering.
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
+\fig{fFrameSync} illustrates the local frame synchronization.
 
@@
 \fig{fMainLoop} outlines the synchronization for the application, node
-and pipe classes for an application node and one render client. Please
-note that \textsf{Config::finishFrame} does block until the current
-frame has been released locally and until the frame \textsf{current -
-  latency} has been released by all nodes. The window and channel
-synchronization are similar and omitted for simplicity.
+and pipe classes for an application node and one render client when
+using the default \textsf{draw\_sync} thread model. Please note that
+\textsf{Config::finishFrame} does block until the current frame has been
+released locally and until the frame \textsf{current - latency} has been
+released by all nodes. The window and channel synchronization are
+similar and omitted for simplicity.
 
@@
 and \textsf{Node::release\-Fra\-me} are called, respectively. The default
-implementation of the node task methods does take care of that.
-
-Since \textsf{eqPly} multi-buffers all dynamic data, it completely
-removes frame synchronization by:
-
-\begin{itemize}
-\item releasing the local synchronization early in \textsf{Node::frameStart}
-\item not waiting for the node to start the frame by not calling
-  \textsf{Node::waitFrame\-Star\-ted} in \textsf{Pipe::frameStart}
-\item not waiting for the pipe synchronization in
-  \textsf{Node::frameDrawFinish} by not calling \textsf{Pipe::waitFrameLocal}
-\end{itemize}
+implementations of the node task methods do take care of that.
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
-\end{wrapfigure}
 The main idea behind direct send is to parallelize the costly
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+  \vspace{-4ex}
+\end{wrapfigure}
 \fig{fDirectSend} illustrates this algorithm for three channels. The
@@
   frustum during stereo rendering. See \sref{sTracking} for details.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_NODE\_IATTR\_THREAD\_MODEL\\
+  \textbf{Value} & ASYNC \textbar \ DRAW\_SYNC \textbar \ LOCAL\_SYNC\\
+  \textbf{Default} & DRAW\_SYNC\\
+  \textbf{Details} & The threading model for node synchronization. See 
+  \sref{sThreadModel} for details.\\
   \hline
@@
 name of the node can be used for debugging, it has no influence on the
-execution of Equalizer. For a description of connection attributes
-please refer to \sref{sGlobal}.
+execution of Equalizer. For a description of node and connection
+attributes please refer to \sref{sGlobal}.
 
@@
         timeout       unsigned    // timeout in milliseconds for launch
+    }
+    attributes
+    {
+        thread_model ASYNC | DRAW_SYNC | LOCAL_SYNC
     }
@@
 
+\cleardoublepage
+\thispagestyle{empty}
+\begin{textblock}{}(0,5)
+  \hspace{-1cm}\includegraphics[width=22cm]{images/teaserBack.pdf}
+\end{textblock}
+
 \end{document}

### 2008-11-13 Developer/ProgrammingGuide/paper.tex
Programming Guide: Prepare for 0.6-rc1 release
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 %  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.16 for Equalizer 0.6-RC1\\[\medskipamount]
+  Version 1.3.17 for Equalizer 0.6-RC1\\[\medskipamount]
   \today
@@
 
-\thispagestyle{empty}
+\pagestyle{empty}
 \maketitle
@@
 
-\thispagestyle{empty}
-\section*{Equalizer 0.5 Programming Guide}
+\section*{Equalizer 0.6 Programming Guide}
 \subsection*{Contributors}
@@
 \clearpage
-\thispagestyle{empty}
 \tableofcontents
+\listoffigures
 \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
@@
     1.3.16  & Nov 10, 2008 & Updated \sref{sThreadModel}, added back page\\
+    1.3.17  & Nov 13, 2008 & Added \sref{sLoadBalancing} and \sref{sAdvDPlex}, final touches\\
     \hline
   \end{tabularx}}
-\thispagestyle{empty}
 \clearpage
@@
 \pagenumbering{arabic}
+\pagestyle{headings}
 
@@
 
+\subsection{\label{sLoadBalancing}Load-Balancing}
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/lb.pdf}
+  {\caption{\label{fLoadBalancing}\small Static 2D compound vs 2D
+      load-balancing}}
+\end{wrapfigure}
+While pixel and stereo compounds are naturally load-balanced, 2D, DB and
+DPlex compounds might need load-balancing to provide optimal
+performance.
+
+Equalizer provides automatic, application-transparent load-balancing for
+2D and DB compounds. Using an appropriate load-balancer for these
+compounds will adapt the 2D split or database range each frame. The
+current load-balancing algorithm is a technology preview and will
+improve in future versions.
+
+2D load-balancing gives better framerates than static decomposition in
+virtually all cases. DB load-balancing is benefitial for applications
+which can not precisely predict the load for their scene data. Volume
+rendering is a counter-example, where the data is uniform and a static
+DB decomposition typically results in a better performance.
+
+DPlex compounds require a smoothing of the framerate at the destination
+channel, otherwise the framerate will become periodically faster and
+slower. Using a DPlex load-balancer will smoothen the swap buffer rate
+on the destination channel for optimal user experience.
+
 \section{The Programming Interface}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/model.pdf}
+  {\caption{\small\label{fModel}Simplified execution model}}
+\end{wrapfigure}
 Using threading correctly in OpenGL-based applications is easy with
@@
 
-The frustum description is inherited by the children, therefore the
-frustum is typically defined on the top-most compound.
-
-The frustrum can be specified as a wall or projection description.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+The frustum description is inherited by the children, therefore the
+frustum is typically defined on the top-most compound.
+
+The frustrum can be specified as a wall or projection description.
+
 A wall is completely defined by the bottom-left, bottom-right and
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
+  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
+\end{wrapfigure}
 The \textsf{eqPly::Config} class holds the master versions of the
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Render Loop}}\vspace{-5ex}
+\end{wrapfigure}
 The model data is spatially organized in an 3-dimensional
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
-\end{wrapfigure}
 The concrete implementation of these window interfaces is responsible
@@
 
+\subsubsection{\label{sAdvDPlex}DPlex Compounds}
+
+DPlex decomposition requires multiple frames to be rendered
+concurrently. Applications using the thread synchronization models
+\textsf{draw\_sync} (the default) or \textsf{local\_sync} have to use
+one render client process per GPU to benefit from DPlex task
+decomposition.
+
+Non-threaded pipes should not be used for DPlex source and destination
+channels. Applications using the \textsf{local\_sync} thread model can
+not benefit from DPlex if the application node uses a DPlex source or
+destination channel.
+
+Applications using the \textsf{async} thread synchronization model can
+fully profit from DPlex using multiple render threads on a multi-GPU
+system. For these applications, all render threads run asynchronously
+and can render different frames at the same time.
+
+Synchronizing the draw operations between multiple pipe render threads,
+and potentially the application thread, breaks DPlex decomposition. At
+any given time, only one frame can be rendered from the same
+process. The speed-up of DPlex relies however on the capability to
+render different frames concurrently.
+
+If one process per GPU is configured, draw-synchronous applications can
+scale the performance using DPlex compounds. The processes are not
+synchronized with each other, since each process keeps its own version
+of the scene data.
+
 \subsubsection{Thread Synchronization in Detail}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
 The local synchronization consists of \textsf{releaseFrameLocal} to
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
 \fig{fFrameSync} illustrates the local frame synchronization.
@@
   \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
-      Volume Rendering}}
+  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
 \end{wrapfigure}
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended together to
@@
 destination frame buffer using the correct order.
+
+For orthographic rendering, determining the compositing order of the
+input frames is trivial. The screen-space orientation of the volume
+bricks determines the order in which they have to be composited. The
+bricks in \textsf{eVolve} are created by slicing the volume along one
+dimension. Therefore the range of the resulting frame buffer images,
+together with the sorting order, is used to arrange the frames during
+compositing. \fig{fBlendOrtho} shows this composition for one view.
+
+Finding the correct assembly order for perspective frustra is more
+complex. The perspective distortion invalidates a simple orientation
+criteria like the one used for orthographic frustra. For the view and
+frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
+  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
+1-4-3-2.
 
@@
 \end{figure}
-
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
-For orthographic rendering, determining the compositing order of the
-input frames is trivial. The screen-space orientation of the volume
-bricks determines the order in which they have to be composited. The
-bricks in \textsf{eVolve} are created by slicing the volume along one
-dimension. Therefore the range of the resulting frame buffer images,
-together with the sorting order, is used to arrange the frames during
-compositing. \fig{fBlendOrtho} shows this composition for one view.
-
-Finding the correct assembly order for perspective frustra is more
-complex. The perspective distortion invalidates a simple orientation
-criteria like the one used for orthographic frustra. For the view and
-frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
-  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
-1-4-3-2.
 
@@
 \cleardoublepage
-\thispagestyle{empty}
+\pagestyle{empty}
 \begin{textblock}{}(0,5)

### 2008-12-01 Developer/ProgrammingGuide/paper.tex
0.6 release updates
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   \vspace{15cm}
-%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
-  Version 1.3.17 for Equalizer 0.6-RC1\\[\medskipamount]
-  \today
+  {\Large\sffamily The official reference for creating parallel,
+    scalable OpenGL\texttrademark\ applications with the Equalizer
+    parallel rendering framework}\\\vspace{1cm}
+  Version 1.4 for Equalizer 0.6\\[\medskipamount]
+%  \today
 }
@@
 
-\maketitle
-\vfill
-\lstset{language=C++}
-
 \pagestyle{empty}
 \maketitle
-\begin{textblock}{}(0,5)
+\thispagestyle{empty}
+\begin{textblock}{}(0,4.5)
   \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
@@
 
+\lstset{language=C++}
 \section*{Equalizer 0.6 Programming Guide}
+\today
 \subsection*{Contributors}
@@
 
-The images on the front page show the following Equalizer applications:
-a terrain rendering application on a six-node display wall [top left], a
-flow visualization application for climate research\footnote{Image
-  courtesy of Computer Graphics and Multimedia Systems, University of
-  Siegen}, the \textsf{eVolve} volume renderer\footnote{Data set
-  courtesy of General Electric, USA} [bottom left] the \textsf{eqPly}
-polygonal renderer in a three-sided CAVE [bottom center], and a six-node
-sort-last database decomposition with parallel direct-send
-recomposition\footnote{Data set courtesy of Stanford University Computer
-  Graphics Laboratory} [bottom right].
+The images on the front page show the following applications build using
+Equalizer: RTT DeltaGen\footnote{Image copyright Realtime Technology AG,
+  2008} [top left], a flow visualization application for climate
+research\footnote{Image courtesy of Computer Graphics and Multimedia
+  Systems, University of Siegen}, the \textsf{eVolve} volume
+renderer\footnote{Data set courtesy of General Electric, USA} [bottom
+left] the \textsf{eqPly} polygonal renderer in a three-sided CAVE
+[bottom center], and a six-node sort-last database decomposition with
+parallel direct-send recomposition\footnote{Data set courtesy of
+  Stanford University Computer Graphics Laboratory} [bottom right].
 
@@
 \clearpage
+\pagenumbering{roman}
+\pagestyle{headings}
 \tableofcontents
@@
     1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
-    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
-    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
-    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
-    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
-    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
-    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
-    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
-    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
-    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
-    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
-    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
-    1.3.11  & Oct 13, 2008 & Updated to reflect eq::net::Object cleanup\\
-    1.3.12  & Oct 16, 2008 & Added \sref{sView}\\
-    1.3.13  & Oct 17, 2008 & Minor update in \sref{sMessagePump}\\
-    1.3.14  & Oct 21, 2008 & Improved \sref{sConnectionDescription}\\
-    1.3.15  & Oct 27, 2008 & Added \sref{sDPlex}\\
-    1.3.16  & Nov 10, 2008 & Updated \sref{sThreadModel}, added back page\\
-    1.3.17  & Nov 13, 2008 & Added \sref{sLoadBalancing} and \sref{sAdvDPlex}, final touches\\
+    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
+    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
     \hline
@@
 
+Equalizer is used for scientific simulation visualization at various
+research institutes, for creating electronic art exhibitions as well as
+in commercial 3D design applications.
+
 Equalizer is the next step in the evolution of generic parallel programming
@@
 
-In order to adapt an application for Equalizer, the programmer
-structures the source code so that the OpenGL rendering can be executed
-in parallel, potentially using multiple processes for cluster-based
-execution. Equalizer provides the domain-specific parallel rendering
-know-how and abstracts configuration, threading, synchronization,
-windowing and event handling. It is a `GLUT on steroids', providing
-parallel and distributed execution, scalable rendering features and
+To adapt an application for Equalizer, the programmer structures the
+source code so that the OpenGL rendering can be executed in parallel,
+potentially using multiple processes for cluster-based execution.
+
+Equalizer provides the domain-specific parallel rendering know-how and
+abstracts configuration, threading, synchronization, windowing and event
+handling. It is a `GLUT on steroids', providing parallel and distributed
+execution, scalable rendering features, network data distribution and
 fully customizable event handling.
@@
 
-\section{Getting Started}
+\subsection{Parallel Rendering}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
+  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}\vspace{-2ex}
+\end{wrapfigure}
+
+\fig{fExecutionFlow} illustrates the basic principle of any parallel
+rendering application. The typical OpenGL application, for example
+GLUT-based applications, has an event loop which redraws the scene,
+updates data based on received events, and eventually redraws a new
+frame.
+
+A parallel rendering application uses the same basic execution model and
+extends it by separating the rendering code from the main event
+loop. The rendering code is then executed in parallel on different
+resources. This model is naturally followed by Equalizer, thus making
+application development as easy as possible.
 
@@
 Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
-to get familiar with the capabilities of the \textsf{eqPly} example.
-
-Compiling Equalizer is as simple as running \textsf{make} on Linux and
-Mac OS X or building the Equalizer Visual Studio 2005 solution on
-Windows. Note that on Mac OS X 10.4 (Tiger), some prerequisites have to
-be installed before running \textsf{make}, as explained in
-\textsf{README.Darwin}.
+to get familiar with the capabilities of Equalizer and the
+\textsf{eqPly} example.
+
+Compiling Equalizer is as simple as running \textsf{make} on Linux or
+Mac OS X, building the provided XCode project on Mac OS X 10.5, or
+building the Equalizer Visual Studio 2005 solution on Windows. Note that
+on Mac OS X 10.4 (Tiger), some prerequisites have to be installed before
+running \textsf{make}, as explained in \textsf{README.Darwin}.
 
@@
 only supports one application per server, but it will provide
-concurrent and efficient multi-application support in future.
+concurrent and efficient multi-application support in the future.
 
@@
 client, which will be controlled by the server. The application reacts
-on events, updates its database and controls the rendering.
+on events, updates its data and controls the rendering.
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/processes.pdf}
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/processes.pdf}
   {\caption{\small\label{fProcesses}Equalizer Processes}}
@@
 also contribute to the rendering. If it does not implement any render
-client-related code, it is reduced to be the application's `master'
-process without any OpenGL windows and rendering code.
+client code, it is reduced to be the application's `master' process
+without any OpenGL windows and rendering code.
 
@@
 takes over the control by calling the render client task
-methods. Complex applications usually implement a separate, light-weight
+methods. Complex applications usually implement a separate, lightweight
 rendering client.
-
-\subsubsection{Parallel Rendering}
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
-  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
-\end{wrapfigure}
-
-\fig{fExecutionFlow} illustrates the basic principle of any parallel
-rendering application. The typical OpenGL application, for example
-GLUT-based applications, has an event loop which redraws the scene,
-updates data based on received events, and eventually redraws a new
-frame.
-
-A parallel rendering application uses the same basic execution model and
-extends it by separating the rendering code from the main event
-loop. The rendering code is then executed in parallel on different
-resources. This model is naturally followed by Equalizer, thus making
-application development as easy as possible.
 
@@
 around the origin. The \textsf{frameDraw} meth\-od provided by the
-\textsf{eq::Channel} can be used as a convience function to setup the
+\textsf{eq::Channel} can be used as a convenience function to setup the
 frustum and other OpenGL state. After setting up some lighting
@@
 algorithms, which require different scalable rendering modes to address
-the right bottlenecks. Equalizer supports the most important algorithms,
-and will continue to add new ones over time to meet application
-requirements.
-
-This section gives an introduction to scalable rendering, in order to
+the bottlenecks correctly. Equalizer supports all important algorithms as
+listed below, and will continue to add new ones over time to meet
+application requirements.
+
+This section gives an introduction to scalable rendering, to
 provide some background for application developers. The scalability
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/DB.pdf}
-  {\caption{\small A database compound}}
-\end{wrapfigure}
-DB\footnote{3D model courtesy of AVS, USA.} decomposes the
-rendered database so that all rendering units process a part of the
-scene in parallel. This mode is also known as sort-last.
+DB, as shown in \fig{fDB}\footnote{3D model courtesy of AVS, USA.},
+decomposes the rendered database so that all rendering units process a
+part of the scene in parallel. This mode is also known as sort-last, and
+is very similar to the data decomposition approach used by HPC
+applications.
 
@@
 
-The application has to partion the database so that the rendering units
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/DB.pdf}
+  {\caption{\label{fDB}\small A database compound}}
+\end{wrapfigure}
+The application has to partition the database so that the rendering units
 render only part of the database. Some OpenGL features do not work
@@
 
-The best use of database compounds is to divide the data to a managable
+The best use of database compounds is to divide the data to a manageable
 size, and then to use other decomposition modes to achieve further
@@
 destination channel. For more than two contributing channels, it is
-recommended to configure parallel direct send compositing, as described
-in \sref{sDirectSend}.
+recommended to configure streaming or parallel direct send compositing,
+as described in \sref{sDirectSend}.
 
@@
 
-Stereo compounds\footnote{3D model courtesy of Stereolithography Archive
-  at Clemson University.} assign each eye pass to individual rendering
-units. The resulting images are copied to the appropriate stereo
-buffer. This mode supports a variety of stereo modes, including active
-(quad-buffered) stereo, anaglyphic stereo and auto-stereo displays with
-multiple eye passes.
+Stereo compounds, as shown in \fig{fStereoCmp}\footnote{3D model
+  courtesy of Stereolithography Archive at Clemson University.}, assign
+each eye pass to individual rendering units. The resulting images are
+copied to the appropriate stereo buffer. This mode supports a variety of
+stereo modes, including active (quad-buffered) stereo, anaglyphic stereo
+and auto-stereo displays with multiple eye passes.
 
@@
   \includegraphics[width=.618\textwidth]{images/EYE.pdf}
-  {\caption{\small A Stereo compound}}
+  {\caption{\label{fStereoCmp}\small A Stereo compound}}
 \end{wrapfigure}
@@
 
-DPlex compounds\footnote{3D model courtesy of Cyberware, Inc.} assign
-full, alternating frames to individual rendering units. The resulting
-images are copied to the destination channel, and Equalizer
-load-balancing is used to ensure a steady framerate on the destination
-window. This mode is also known as time-multiplex or AFR.
+DPlex compounds assign full, alternating frames to individual rendering
+units. The resulting images are copied to the destination channel, and
+Equalizer load-balancing is used to ensure a steady framerate on the
+destination window. This mode is also known as time-multiplex or AFR.
 
@@
 
-Pixel compounds are similar to 2D compounds. The frustra of the
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
+  {\caption{\label{fPixel}\small A pixel compound}}\vspace{-5ex}
+\end{wrapfigure}
+Pixel compounds are similar to 2D compounds. The frusta of the
 source rendering units are modified so that each unit renders an evenly
-distributed subset of pixels.
+distributed subset of pixels, as shown in \fig{fPixel}\footnote{3D model courtesy of AVS, USA.}
 
@@
 the pixel transfers during recomposition.
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
-  {\caption{\small A pixel compound}}
-\end{wrapfigure}
-Pixel compounds work well for purely fill-limited applications.
-Techniques like frustum culling do not reduce the rendered data for the
-source rendering resources.
 
@@
 
-Pixel compounds are ideal for raytracing, which is highly fill-limited
-and needs the full database for rendering anyway. Volume rendering
-applications are also well suited for this mode, and should choose it
-over 2D compounds.
+Pixel compounds work well for purely fill-limited applications.
+Techniques like frustum culling do not reduce the rendered data for the
+source rendering resources. Pixel compounds are ideal for raytracing,
+which is highly fill-limited and needs the full database for rendering
+anyway. Volume rendering applications are also well suited for this
+mode, and should choose it over 2D compounds.
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/lb.pdf}
-  {\caption{\label{fLoadBalancing}\small Static 2D compound vs 2D
-      load-balancing}}
-\end{wrapfigure}
 While pixel and stereo compounds are naturally load-balanced, 2D, DB and
 DPlex compounds might need load-balancing to provide optimal
-performance.
-
-Equalizer provides automatic, application-transparent load-balancing for
-2D and DB compounds. Using an appropriate load-balancer for these
-compounds will adapt the 2D split or database range each frame. The
-current load-balancing algorithm is a technology preview and will
-improve in future versions.
-
-2D load-balancing gives better framerates than static decomposition in
-virtually all cases. DB load-balancing is benefitial for applications
-which can not precisely predict the load for their scene data. Volume
-rendering is a counter-example, where the data is uniform and a static
-DB decomposition typically results in a better performance.
+performance. Using a load-balancer is transparent to the application,
+and can be used on any application for 2D and DPlex, and on most
+applications for DB load-balancing. Some applications supporting DB
+decomposition do not support dynamic updates of the database range, and
+therefore cannot be used with DB load-balancing.
+
+Using a 2D or DB load-balancer will adapt the 2D split or database range
+automatically each frame. The 2D load-balancer exists in three flavors:
+\textsf{2D} using tiles, \textsf{VERTICAL} using columns and
+\textsf{HORIZONTAL} using rows. The current load-balancing algorithm for
+2D and DB is a technology preview and will improve in future versions.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/lb.pdf}
+  {\caption{\label{fLoadBalancing}\small 2D load-balancing}}
+\end{wrapfigure}
+2D load-balancing increases the framerate over a static decomposition in
+virtually all cases. It works best if the application data is relatively
+uniformly distributed in screen-space.
+
+DB load-balancing is beneficial for applications which cannot precisely
+predict the load for their scene data. Volume rendering is a
+counterexample, where the data is uniform and a static DB decomposition
+typically results in a better performance.
 
@@
 slower. Using a DPlex load-balancer will smoothen the swap buffer rate
-on the destination channel for optimal user experience.
+on the destination window for optimal user experience.
 
@@
 OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
-methods are similar in concept to C function callbacks. The
-\textsf{eqPly} section will discuss the most important task methods. A
-full list can be found on the website\footnote{see
+methods are similar in concept to C function callbacks. \sref{sEqPly}
+will discuss the important task methods. A full list can be found on the
+website\footnote{see
   \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.
@@
 thread. If structured correctly, the application rarely has to take care
-of thread synchonization or protection of shared data.
+of thread synchronization or protection of shared data.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/model.pdf}
+  {\caption{\small\label{fModel}Simplified execution model}}
+\end{wrapfigure}
 All rendering threads in the configuration run asynchronously to the
@@
 
-Rendering threads on a single node are by default synchronized. When a
-frame is finished, all local rendering threads are done
-drawing. Therefore the application can safely modify the data between
-the end of a frame and the beginning of a new frame. Furthermore, only
-one instance of the scene data has to be maintained within a process
-since all rendering threads are guaranteed to draw the same frame.
+Rendering threads on a single node are synchronized when using the
+default thread model \textsf{draw\_sync}. When a frame is finished, all
+local rendering threads are done drawing. Therefore the application can
+safely modify the data between the end of a frame and the beginning of a
+new frame. Furthermore, only one instance of the scene data has to be
+maintained within a process, since all rendering threads are guaranteed
+to draw the same frame.
 
@@
 thread. Non-threaded pipes are supported for integrating with
-thread-unsafe libraries, but have various performance caveats. All pipe,
-window and channel task methods are executed from the pipe's thread, or
-in the case of non-threaded pipes from the main application
+thread-unsafe libraries, but have various performance caveats. They
+should only be used if using a different, but synchronized rendering
+thread is not an option.
+
+All pipe, window and channel task methods are executed from the pipe
+thread, or in the case of non-threaded pipes from the main application
 thread\footnote{see
@@
 context. The drawable can be an on-screen window or an off-screen
-PBuffer. Framebuffer object (FBO) support will be implement by a later
+PBuffer. FrameBuffer object (FBO) support will be implement by a later
 version of Equalizer. The window uses an \textsf{eq::OSWindow}, which
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frustra.pdf}
-  {\caption{\small\label{fFrustra}Wall and projection parameters}}
+  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
+  {\caption{\small\label{fFrusta}Wall and projection parameters}}
 \end{wrapfigure}
 The frustum description is inherited by the children, therefore the
-frustum is typically defined on the top-most compound.
-
-The frustrum can be specified as a wall or projection description.
+frustum is typically defined on the topmost compound.
+
+The frustum can be specified as a wall or projection description.
 
@@
 
-\fig{fFrustra} illustrates the wall and projection frustum parameters.
+\fig{fFrusta} illustrates the wall and projection frustum parameters.
 
@@
 The channels of the leaf compounds in the compound tree are designated
-as source channels. The top-most channel in the tree is the destination
-channel. Only source channels execute the draw task. All channels in a
-compound tree work for the destination channel. The destination channel
-defines the 2D pixel viewport rendered by all leaf compounds. The
-destination channel and pixel viewport can not be overridden by child
-compounds.
+as source channels. The topmost channel in the tree is the destination
+channel. One compound tree might have multiple destination channels,
+e.g., for a swap-synchronized immersive installation.
+
+All channels in a compound tree work for the destination channel. The
+destination channel defines the 2D pixel viewport rendered by all leaf
+compounds. The destination channel and pixel viewport cannot be
+overridden by child compounds.
+
+\subsubsection{Tasks}
+Compounds execute a number of tasks: clear, draw, assemble and
+readback. By default, a leaf compound executes all tasks and a non-leaf
+compound assemble and readback. A non-leaf compound never executes the
+draw task.
+
+A compound can be configured to execute a specific set of tasks, for
+example to configure the multiple steps used by binary-swap compositing.
 
@@
 Compounds have attributes which configure the decomposition of the
-destination's channel viewport, frustum and database. A
-\textsf{viewport} decomposes the destination channel and frustum in
-screen space. A \textsf{range} tells the application to render a part of
-its database, and an \textsf{eye} rendering pass can selectively render
-different stereo passes. A \textsf{pixel} parameter adapts the frustum
-so that the source channel renders an evenly subset of the parent's
-pixels. Setting one or multiple attributes causes the parent's view to
-be decomposed accordingly. Attributes are cumulative, that is,
-intermediate compound attributes affect and therefore decompose the
-rendering of all their children.
+destination channel's rendering, which is defined by the viewport,
+frustum and database. A \textsf{viewport} decomposes the destination
+channel and frustum in screen space. A \textsf{range} tells the
+application to render a part of its database, and an \textsf{eye}
+rendering pass can selectively render different stereo passes. A
+\textsf{pixel} parameter adapts the frustum so that the source channel
+renders an evenly subset of the parent's pixels. Setting one or multiple
+attributes causes the parent's view to be decomposed
+accordingly. Attributes are cumulative, that is, intermediate compound
+attributes affect and therefore decompose the rendering of all their
+children.
 
@@
 \sref{sCompositing}. Frame names are only valid within the compound
-tree, that is, an output frame from one compound tree can not be used as
+tree, that is, an output frame from one compound tree cannot be used as
 an input frame of another compound tree.
@@
 hardware synchronization, e.g., nVidia G-Sync cards. It is however
-sufficient for display walls made out of LCD's.
+sufficient for display walls made out of LCD's and for some
+non-immersive projection systems.
 
 Swapbarrier names are only valid within the compound tree, that is, a
-compound from one compound tree can not be synchronized with a compound
+compound from one compound tree cannot be synchronized with a compound
 from another compound tree.
@@
 
+\textsf{eqPly} is a parallel renderer for polygonal data in the
+\textsf{ply} file format. It supports all relevant Equalizer features,
+and can be used to render on large-scale displays, immersive
+environments with head tracking and to render massive data sets using
+all scalable rendering features of Equalizer.
+
 The \textsf{eqPly} example is shipped with the Equalizer distribution
@@
 application of medium complexity. It focuses on the example usage of
-core Equalizer features, not on rendering features or visual quality.
+core Equalizer features, not on advanced rendering features or visual
+quality.
 
@@
 The \textsf{eqPly} classes fall into two categories: Subclasses of the
-rendering entities introduced in \sref{sConfig}, and two other classes
-for distributing data. The function and typical usage for each of the
+rendering entities introduced in \sref{sConfig}, and classes for
+distributing data. The function and typical usage for each of the
 rendering entities is discussed in this section.
@@
 \textsf{eq::net::Object}. They illustrate the typical usage of distributed
-objects for static as well as dynamic, frame-specific data.
+objects for static as well as dynamic, frame-specific data. Furthermore
+they are used for a basic scene graph distribution of the model data.
 
@@
 public:
-    virtual eq::Config*  createConfig()  
-        { return new eqPly::Config; }
+    virtual eq::Config*  createConfig( eq::ServerPtr parent )
+        { return new eqPly::Config( parent ); }
     virtual eq::Node*    createNode( eq::Config* parent )  
@@
 The third step is to create an instance of the application and to
-initialize it locally. The application is an \textsf{eq::Client}, which
-in turn is an \textsf{eq::net::Node}. The underlying network layer in
-Equalizer is a peer-to-peer network of \textsf{eq::net::Node}s. The
-application programmer does not usually have to be aware of the classes
-in the \textsf{eq::net} namespace, but both the \textsf{eq::Client} and
-the server are \textsf{eq::net::Node}s.
-
-The local initialization of a node creates a local listening socket,
-which allows the \textsf{eq::Client} to communicate over the network
-with other nodes, such as the server and the rendering clients.
+initialize it locally. The application is a subclass of
+\textsf{eq::Client}, which in turn is an \textsf{eq::net::Node}. The
+underlying network layer in Equalizer is a peer-to-peer network of
+\textsf{eq::net::Node}s. The client/server concept is implement the
+higher-level \textsf{eq} client namespace.
+
+The application programmer does not usually have to be aware of the
+classes in the \textsf{eq::net} namespace, but both the
+\textsf{eq::Client} and the Equalizer server are
+\textsf{eq::net::Node}s.
+
+The local initialization of a node creates at least one local listening
+socket, which allows the \textsf{eq::Client} to communicate over the
+network with other nodes, such as the server and the rendering
+clients. The listening socket(s) can be configured using the
+\textsf{--eq-listen} command line parameter, or by programmatically
+adding connection descriptions to the client before the local
+initialization.
 
@@
 In the case of \textsf{eqPly}, the application is also the render
-client. The \textsf{eqPly} executable has three run-time behaviours:
+client. The \textsf{eqPly} executable has three run-time behaviors:
 
@@
 \item \textbf{Application}: The executable started by the user, the
-  controlling entity in the rendering session.
+  controlling entity of the rendering session.
 \item \textbf{Auto-launched render client}: The typical render client,
@@
   return. During exit, the server terminates the process. By default,
-  the server starts the render client using \textsf{ssh}.
+  the server starts the render client using \textsf{ssh}. The launch
+  command can be used to configure another program to auto-launch render
+  clients.
 \item \textbf{Resident render client}: Manually pre-started render
@@
     // 1. connect to server
-    RefPtr<eq::Server> server = new eq::Server;
+    eq::ServerPtr server = new eq::Server;
     if( !connectServer( server ))
@@
 
-    EQLOG( eq::LOG_CUSTOM ) << "Config init took " << clock.getTimef() << " ms"
+    EQLOG( eq::LOG_STATS ) << "Config init took " << clock.getTimef() << " ms"
                            << endl;
@@
 completion of all local rendering tasks for the current frame. This
-facilitates porting of exising rendering codes, since the database does
+facilitates porting of existing rendering codes, since the database does
 not have to be multi-buffered. Applications such as \textsf{eqPly}, which
@@
 
+\textsf{eqPly} uses event-driven execution, that is, it only request new
+rendering frames if an event or animation requires an update. The
+\textsf{eqPly::Config} maintains a dirty state, which is cleared after a
+frame has been started, and set when an event causes a
+redraw. Furthermore, when an animation is running or head tracking is
+active, the config always signals the need for a new frame.
+
+If the application detects that it is currently idle, all pending frames
+are flushed. Then it waits and handles one event at a time, until a
+redraw is needed:
+
+{\footnotesize\begin{lstlisting}
+    // 4. run main loop
+    uint32_t maxFrames = _initData.getMaxFrames();
+
+    clock.reset();
+    while( config->isRunning( ) && maxFrames-- )
+    {
+        config->startFrame();
+        // config->renderData(...);
+        config->finishFrame();
+
+        if( !config->needsRedraw( ))
+            config->finishAllFrames(); // flush, TODO flush one task at a time
+
+        while( !config->needsRedraw( )) // wait for an event requiring redraw
+        {
+            const eq::ConfigEvent* event = config->nextEvent();
+            if( !config->handleEvent( event ))
+                EQVERB << "Unhandled " << event << endl;
+        }
+    }
+\end{lstlisting}}%>>
+
 When the main rendering loop has finished,
@@
 {\footnotesize\begin{lstlisting}
-    // 4. run main loop
-    uint32_t maxFrames = _initData.getMaxFrames();
-
-    clock.reset();
-    while( config->isRunning( ) && maxFrames-- )
-    {
-        config->startFrame();
-        // config->renderData(...);
-        config->finishFrame();
-    }
     const uint32_t frame = config->finishAllFrames();
     const float    time  = clock.getTimef();
-    EQLOG( eq::LOG_CUSTOM ) << "Rendering took " << time << " ms (" << frame
+    EQLOG( eq::LOG_STATS ) << "Rendering took " << time << " ms (" << frame
                            << " frames @ " << ( frame / time * 1000.f)
@@
     config->exit();
-    EQLOG( eq::LOG_CUSTOM ) << "Exit took " << clock.getTimef() << " ms" <<endl;
+    EQLOG( eq::LOG_STATS ) << "Exit took " << clock.getTimef() << " ms" <<endl;
 
@@
 Distributed objects are addressed using session-unique identifiers,
-because pointers to other objects can not be distributed directly, since
+because pointers to other objects cannot be distributed directly, since
 they have no meaning on remote nodes. The session with which distributed
-objects are registered is normally the \textsf{eq::Config}.
+objects are registered is normally the \textsf{eq::Config}, which is
+subclassed from \textsf{eq::net::Session}.
 
@@
 identifier is passed in the initial data, and the frame data version is
-passed with each start frame request.
+passed with each new frame request.
 
@@
 
-The way changes are to be handled is determined by calling
+Equalizer determines the way changes are to be handled by calling
 \textsf{Object::getChangeType} during the registration of the master
@@
     serialized and saved to map old versions.
-  \item[DELTA\_UNBUFFERED] The object is versioned, and delta data is
-    used to update slave instances. No data is stored, and no previous
-    versions can be mapped. The instance data is serialized whenever a
-    new slave instance is mapped.
+  \item[UNBUFFERED] The object is versioned. No data is stored, and no
+    previous versions can be mapped. The instance data is serialized
+    whenever a new slave instance is mapped. The application can choose
+    to use a different, more optimal implementation to pack deltas.
 \end{description}
@@
 distributed data. These method provide an output or input stream as a
-parameter, which abstract the data transmission and can be used like a
+parameter, which abstracts the data transmission and can be used like a
 \textsf{std::stream}.
+
+Custom data type serializers can be implemented by providing the
+appropriate serialization functions. No pointers should be directly
+transmitted through the data streams. Typically the corresponding object
+is a distributed object as well, and it's identifier and potentially
+version is transmitted in place of a pointer.
 
@@
 object to a new version. The \textsf{sync} method might block if a
-version has not been committed or is still in transmission.
+version has not yet been committed or is still in transmission.
 
@@
 
-The \textsf{eqPly} does not track changes since it is relatively small
-in size and changes frequently. Therefore, the instance and delta data
-are the same, and \textsf{pack} and \textsf{unpack} are not
+\textsf{eqPly} does not track changes in the \textsf{FrameData} since it
+is relatively small and changes frequently. Therefore, the instance and
+delta data are the same, and \textsf{pack} and \textsf{unpack} are not
 implemented. The change type is \textsf{INSTANCE}. The serialization
-functions exploit the fact that the frame data is layed
-out in one anonymous structure in memory, and transmit all the data in
-one call:
+functions exploit the fact that the frame data is layed out in one
+anonymous structure in memory, and transmit all the data in one call:
 
@@
 
-\textsf{eqPly} uses static distributed objects to distribute the model
-loaded by the application. This model can be easily extended to
-versioned objects, in order to support dynamic data changes.
-
-The kD-tree data structure and rendering code for the model is strongly
-separated from Equalizer, and kept in the separate namespace
-\textsf{mesh}. It could also be used in a GLUT application. To keep this
-separation, an external 'mirror' hierarchy is constructed around the
-tree. This hierarchy of \textsf{VertexBufferDist} nodes is responsible
-for cloning the model data on the remote render clients.
-
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
-The identifer of the root object of this distributed hierarchy is passed
+\textsf{eqPly} uses static distributed objects to distribute the model
+loaded by the application. This model can be easily extended to
+versioned objects to support dynamic data changes.
+
+The kD-tree data structure and rendering code for the model is strongly
+separated from Equalizer, and kept in the separate namespace
+\textsf{mesh}. It could also be used in other rendering software, for
+example in a GLUT application. To keep this separation, an external
+'mirror' hierarchy is constructed around the tree. This hierarchy of
+\textsf{VertexBufferDist} nodes is responsible for cloning the model
+data on the remote render clients.
+
+The identifier of the root object of this distributed hierarchy is passed
 as part of the \textsf{InitData}, and used during remote node
@@
 \begin{wrapfigure}{r}{.618\textwidth}
+%\begin{figure}[ht!]\center
   \includegraphics[width=.618\textwidth]{images/objects.pdf}
@@
 The objects are still created by the application, and then registered or
-mapped with the session in order to distribute them. When mapping
+mapped with the session to distribute them. When mapping
 objects in a hierarchical data structure, their type often has to be
-known in order to create them. Equalizer does not currently provide
+known to create them. Equalizer does not currently provide
 object typing, this has to be done by the application, either implicitly
@@
 
+\begin{figure}[ht!]\center
+%\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
+  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
+\end{figure}
 The config initialization happens in parallel, that is, all config
@@
 the other hand, it means that initialization functions are called even
-if the parent's initialization has failed.
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
-  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
-\end{wrapfigure}
+if the parent's initialization has failed. \fig{fConfigInit} shows a
+sequence diagram of the config initialization.
+
 The \textsf{eqPly::Config} class holds the master versions of the
@@
     _frameData.data.color      = _initData.useColor();
+    _frameData.data.renderMode = _initData.getRenderMode();
     registerObject( &_frameData );
+
     _initData.setFrameDataID( _frameData.getID( ));
-
     registerObject( &_initData );
@@
         {
-            // Set up position of tracking system in world space
+            // Set up position of tracking system wrt world space
             // Note: this depends on the physical installation.
@@
             m.scale( 1.f, 1.f, -1.f );
-            //m.x = .5;
             _tracker.setWorldToEmitter( m );
@@
             _tracker.setSensorToObject( m );
-            EQLOG( eq::LOG_CUSTOM ) << "Tracker initialised" << endl;
+            EQINFO << "Tracker initialized" << endl;
         }
@@
 The rendering frames are issued by the application. The
-\textsf{eqPly::Config} only overrides \textsf{startFrame} in order to
+\textsf{eqPly::Config} only overrides \textsf{startFrame} to
 update its data before forwarding the start frame request to the
@@
 with the wall or projection description to compute the view
-frustra\footnote{see
-  \link{http://www.equalizergraphics.com/documents/design/immersive.html}}.
-
-The camera position is updated and the frame data is commited, which
+frusta, as explained in \sref{sTracking}.
+
+The camera position is updated and the frame data is committed, which
 generates a new version of this object. This version is passed to the
@@
 
-For event-driven execution, the application can override
-\textsf{Config::handleEvents} to blockingly receive events using
-\textsf{Config::nextEvent} until a new frame has to be rendered.
-
-The \textsf{eqPly} example continuously renders new frames. It
-implements \textsf{Config::hand\-le\-Event} to provide the various reactions
-to user input, most importantly camera updates based on mouse
-events. The camera position has to be handled correctly with respect to
-latency, and is therefore saved in the frame data:
+Since \textsf{eqPly} uses event-driven execution, the config maintains a
+dirty state to allow the application to know when a redraw is needed.
+
+The \textsf{eqPly} example implements \textsf{Config::hand\-le\-Event}
+to provide the various reactions to user input, most importantly camera
+updates based on mouse events. The camera position has to be handled
+correctly regarding latency, and is therefore saved in the frame data:
 
@@
                     -0.005f * event->pointerMotion.dy );
+                _redraw = true;
             }
@@
                     .005f * event->pointerMotion.dy;
+                _redraw = true;
             }
@@
                     .0005f * event->pointerMotion.dy;
+                _redraw = true;
             }
@@
 
+All dynamic data is multi-buffered in \textsf{eqPly}. During
+initialization, the node relaxes the thread synchronization between the
+node and pipe threads, unless the configuration file overrides
+this. \sref{sThreads} provides a detailed explanation of thread
+synchronization in Equalizer.
+
 During node initialization the static, per-config data is mapped to a
@@
 \textsf{Config::init}. No pipe, window or channel tasks methods are
-executed before \textsf{Node::configInit} has returned.
+executed before \textsf{Node::configInit} has returned:
 
@@
 
+    // All render data is static or multi-buffered, we can run asynchronously
+    if( getIAttribute( IATTR_THREAD_MODEL ) == eq::UNDEFINED )
+        setIAttribute( IATTR_THREAD_MODEL, eq::ASYNC );
+
     Config*    config = static_cast< Config* >( getConfig( ));
@@
 first retrieves the distributed \textsf{InitData}. The object is
-directly unmapped since it is static, and the data has be retrieved.
+directly unmapped since it is static, and the data has been retrieved.
 
@@
 \end{lstlisting}}%>>
-
-The two remaining functions in the node relax the thread synchronization
-between the node and pipe threads, since all dynamic data is
-multi-buffered in \textsf{eqPly}. \sref{sThreads} provides a detailed
-explanation of thread synchronization in Equalizer:
-
-{\footnotesize\begin{lstlisting}
-void Node::frameDrawFinish( const uint32_t frameID,
-                            const uint32_t frameNumber )
-{ /* nop, see frameStart */ }
-
-void Node::frameStart( const uint32_t frameID, const uint32_t frameNumber )
-{
-    startFrame( frameNumber ); // unlock pipe threads
-
-    // Don't wait for pipes to release frame locally, sync not needed since all
-    // dynamic data is multi-buffered
-    releaseFrameLocal( frameNumber );
-}
-\end{lstlisting}}
 
@@
 
-Pipe threads are not explicitely synchronized with each other, that is,
+Pipe threads are not explicitly synchronized with each other, that is,
 pipes might be rendering different frames at any given time. Therefore
 frame-specific data has to be allocated for each pipe thread, which in
-the \textsf{eqPly} example is the frame data. The frame data is a member
+the case of \textsf{eqPly} is the frame data. The frame data is a member
 variable of the \textsf{eqPly::Pipe}, and is mapped to the identifier
@@
 {
-    // don't wait for node to start frame, local sync not needed
-    // node->waitFrameStarted( frameNumber );
+    node->waitFrameStarted( frameNumber );
+
     _frameData.sync( frameID );
@@
 initialization. The default implementation sets up some typical OpenGL
-state, e.g., it enables the depth test.
+state, e.g., it enables the depth test. Most nontrivial applications
+do override this task method.
 
@@
 initialization only overrides the OpenGL-specific initialization
-function \textsf{configInitGL} in order to initialize a state object and
+function \textsf{configInitGL} to initialize a state object and
 an overlay logo. This function is only called if an OpenGL context was
@@
 The state object is used to handle the creation of OpenGL objects in a
-multipipe, multi-threaded execution environment. It uses the object
+multipipe, multithreaded execution environment. It uses the object
 manager of the \textsf{eq::Window}, which is described in detail in
@@
 The logo texture is loaded from the file system and bound to a texture
-ID used later by the channel for rendering. A code listing is ommitted,
+ID used later by the channel for rendering. A code listing is omitted,
 since the code consists of standard OpenGL calls and is not
@@
 Each \textsf{eq::Window} has an object manager with the key type
-\textsf{const void*} for as long as it has an OpenGL context. The OM is
-shared between all windows of a pipe, if the windows are created using
-the default \textsf{configInit} functions. If the window is created by
-the application, the OM is not shared since no assumption can be made
-about OpenGL context sharing. Later version of Equalizer will introduce
-an API to configure OpenGL context sharing, and therefore OM sharing, for
-application-created windows.
+\textsf{const void*}, for as long as it is initialized. Each window can
+have a shared context window. The OM is shared with this shared context
+window. The shared context window is set by default to the first window
+of each pipe, and therefore the OM will be shared between all windows of
+a pipe. The method \textsf{eq::Window::setSharedContextWindow} can be
+used to set up a different context sharing.
 
@@
 
-The channel is the heart of the application in that it contains the
-actual rendering code. The channel is used to perform the various rendering
-operations for the compounds.
+The channel is the heart of the application, it contains the actual
+rendering code. The channel is used to perform the various rendering
+operations for the compounds. Each channel has a set of task methods to
+execute the clear, draw, readback and assemble stages needed to render a
+frame.
 
@@
 using the \textsf{get} methods, and use them to set up the OpenGL state
-appropriatly, for example by passing the parameters to the renderer used
+appropriately, for example by passing the parameters to the renderer used
 by the application.
@@
 
-The contextual information has to be used in order to render the view as
+The contextual information has to be used to render the view as
 expected by Equalizer. Failure to use certain information will result in
@@
   frustum. This is typically an identity matrix and is used for off-axis
-  frustra in immersive rendering. It is normally used to set up the
+  frusta in immersive rendering. It is normally used to set up the
   `view' part of the modelview matrix, before static light sources are
@@
 convenience functions to setup the OpenGL state for all render context
-information, with the exception of the range which will be used later
-during rendering:
+information, except for the range which will be used later during
+rendering:
 
@@
   \includegraphics[width=.382\textwidth]{images/DB.png}
-  {\caption{\small\label{fDB}Destination view of a DB compound}}
+  {\caption{\small\label{fDBDest}Destination view of a DB compound}}
 \end{wrapfigure}
@@
 Furthermore, a white color is set in case the model does not contain
-color information, or the color information is not used. In
-sort-last rendering, \textsf{eqPly} uses a different color for each
-channel to illustrate the database decomposition, as shown in
-\fig{fDB}. The Equalizer channel provides a method to obtain a random,
-but unique color for all channels in the configuration. This color is
-determined by the server to ensure uniqueness across all channels of the
-configuration:
+color information, or the color information is not used. In sort-last
+rendering, \textsf{eqPly} uses a different color for each channel to
+illustrate the database decomposition, as shown in \fig{fDBDest}. The
+Equalizer channel provides a method to obtain a random, but unique color
+for all channels in the configuration. This color is determined by the
+server to ensure uniqueness across all channels of the configuration:
 
@@
 
-In order to draw the model, a helper class for view frustum culling is
-set up using the view frustum from Equalizer and the camera position
-from the frame data. The frustum helper computes the six frustum planes
-from the projection and modelView matrices. During rendering, the
-bounding spheres of the model are tested against these planes to
-determine the visibility with the frustum.
+To draw the model, a helper class for view frustum culling is set up
+using the view frustum from Equalizer and the camera position from the
+frame data. The frustum helper computes the six frustum planes from the
+projection and modelView matrices. During rendering, the bounding
+spheres of the model are tested against these planes to determine the
+visibility with the frustum.
 
@@
   \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Render Loop}}\vspace{-5ex}
+  {\caption{\small\label{fRender}Main Render Loop}}
 \end{wrapfigure}
-The model data is spatially organized in an 3-dimensional
+The model data is spatially organized in a 3-dimensional
 kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
 for efficient view frustum culling. When the model is loaded by
-\textsf{Node::config\-Init}, it is preprocessed into the kD-tree During
+\textsf{Node::config\-Init}, it is preprocessed into the kD-tree. During
 this preprocessing step, each node of the tree gets a database range
@@
 initially contains the root node. Each candidate of this list is tested
-for full visiblity against the frustum and range, and rendered if
+for full visibility against the frustum and range, and rendered if
 visible. It is dropped if it is fully invisible or fully out of
@@
 
-This section discusses some additional important features not covered
-by the previous \textsf{eqPly} section. Where possible, code examples
-from the Equalizer distribution are used to illustrate one use case of
-the feature.
+This section discusses important features not covered by the previous
+\textsf{eqPly} section. Where possible, code examples from the Equalizer
+distribution are used to illustrate one usage of the feature. Its
+purpose is to provide information on how to address a typical problem or
+use case when developing an Equalizer-based application.
 
@@
 
-Event handling requires a lot of flexibility. On one hand, the
-implementation differs slightly for each operating and window system due
-to conceptual differences in the specific implementation. On the
-other hand, each application and widget set has its own model on how
-events are to be handled. Therefore, event handling in Equalizer is
-customizable at any stage of the processing, to the extreme of making it
-possible to disable all event-related code in Equalizer. In this aspect,
-Equalizer substantially differs from GLUT, which imposes an event model
-and hides most of the event handling in \textsf{glutMainLoop}. More
-information on event handling can be found on the Equalizer
-website\footnote{see
+Event handling requires flexibility. On one hand, the implementation
+differs slightly for each operating and window system due to conceptual
+differences in the specific implementation. On the other hand, each
+application and widget set has its own model on how events are to be
+handled. Therefore, event handling in Equalizer is customizable at any
+stage of the processing, to the extreme of making it possible to disable
+all event-related code in Equalizer. In this aspect, Equalizer
+substantially differs from GLUT, which imposes an event model and hides
+most of the event handling in \textsf{glutMainLoop}. More information on
+event handling can be found on the Equalizer website\footnote{see
   \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.
@@
 \textsf{Config::processEvent} to update its data based on the
-pre-processed, generic keyboard and mouse events. It is very easy to use
-and similar to an GLUT-based implementation.
+preprocessed, generic keyboard and mouse events. It is very easy to use
+and similar to a GLUT-based implementation.
 
@@
 
-In order to dispatch events, Equalizer 'pumps' the native events. On
-WGL and GLX, this happens on each thread with windows, whereas on
-AGL it happens only on the main thread. By default, Equalizer
-pumps these events automatically for the application in-between
-executing task methods.
+To dispatch events, Equalizer 'pumps' the native events. On WGL and GLX,
+this happens on each thread with windows, whereas on AGL it happens only
+on the main thread. By default, Equalizer pumps these events
+automatically for the application in-between executing task methods.
 
@@
 \textsf{aglUpdateContext} whenever a resize event is received. For
-further processing, the \textsf{Event} is passed on to
+further, generic processing, the \textsf{Event} is passed on to
 \textsf{Window::processEvent}. This \textsf{Event} no longer contains
@@
 
-Currently, Equalizer provides an \textsf{AGLEvent\-Handler},
+Equalizer provides an \textsf{AGLEvent\-Handler},
 \textsf{GLXEventHandler} and \textsf{WGL\-Event\-Handler}, which handle
@@
 
-\subsection{\label{sThreads}Multi-Threading and Synchronization}
-
-Equalizer applications use multiple, asynchronous execution threads. The
-default execution model is focused on making the porting of existing
-applications as easy as possible, as described in \sref{sExecution}. The
-default, per-node thread synchronization provided by Equalizer can be
-relaxed by advanced applications to gain better performance through
-higher asynchronicity.
+\subsection{\label{sThreads}Thread Synchronization}
+
+Equalizer applications use multiple, potentially asynchronous execution
+threads. The default execution model is focused on making the porting of
+existing applications as easy as possible, as described in
+\sref{sExecution}. The default, per-node thread synchronization provided
+by Equalizer can be relaxed by advanced applications to gain better
+performance through higher asynchronicity.
 
@@
 
-The command thread processes Equalizer-related request from other nodes,
-for example during \textsf{eq::net::Object} mapping. In some special cases
-the command thread executes application code, for example when a remote
-node maps a static or unbuffered object,
+The command thread processes internal Equalizer request from other
+nodes, for example during \textsf{eq::net::Object} mapping. In some
+special cases the command thread executes application code, for example
+when a remote node maps a static or unbuffered object,
 \textsf{Object::getInstanceData} is called from the command thread.
@@
 \item[ASYNC:] No synchronization happens between render threads, except
-  for synchonizing the finish of frame \textsf{current-latency}. The
+  for synchronizing the finish of frame \textsf{current-latency}. The
   \textsf{eqPly} and \textsf{eVolve} Equalizer examples use this
   threading model, by setting it in \textsf{Node::configInit}. This
-  synchonization model provides the best performance and should be used
+  synchronization model provides the best performance and should be used
   by all applications which multi-buffer all non-static, frame-specific
@@
   which keep one copy of the scene graph per node.
-\item[LOCAL\_SYNC:] In addition to the async threas model, all local
+\item[LOCAL\_SYNC:] In addition to the async thread model, all local
   frame operations, including readback, assemble and swap buffer are
@@
   {\caption{\small\label{fThreadModels}Async, draw\_sync and local\_sync
-    thread sunchronization models}}
+    thread synchronization models}}
 \end{figure}
@@
 any given time, only one frame can be rendered from the same
-process. The speed-up of DPlex relies however on the capability to
+process. The speedup of DPlex relies however on the capability to
 render different frames concurrently.
@@
 \textsf{current-latency}. The \textsf{frameFinish} task methods have to
-call \textsf{releaseFrame} to signal that the entitity is done with the
+call \textsf{releaseFrame} to signal that the entity is done with the
 frame. The release causes the parent's \textsf{frameFinish} to be
@@
 and initialize the \textsf{OSWindow} implementation during config
-initialization. After an successful \textsf{OSWindow} initialization,
+initialization. After a successful \textsf{OSWindow} initialization,
 \textsf{Window::configInitGL} is called for the generic OpenGL state
@@
 OpenGL drawables have a multitude of buffer modes. A drawable might be
-single-buffered, double-buffered or quad-buffered, have auxilary image
+single-buffered, double-buffered or quad-buffered, have auxiliary image
 planes such as stencil, accumulation and depth buffer or multisampling.
@@
 The \textsf{eqPly} example contains rudimentary support for head
-tracking, in order to show how head tracking can be integrated with
-Equalizer. Support for a wide range of tracking devices is not within the
-scope of Equalizer. Other open source and commercial implementations
+tracking to show how head tracking can be integrated with
+Equalizer. Support for a wide range of tracking devices is not within
+the scope of Equalizer. Other open source and commercial implementations
 cover this functionality sufficiently and can easily be integrated with
@@
   {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked
-    frustra}}
+    frusta}}
 \end{figure}
@@
 
-In stereo rendering, the scene is rendered twice, with the two frustra
+In stereo rendering, the scene is rendered twice, with the two frusta
 'moved' by the distance between the eyes, as shown in \fig{fStereo}.
@@
 In immersive visualization, the observer is tracked in and the view
-frustra are adapted to the viewer's position and orientation, as shown
+frusta are adapted to the viewer's position and orientation, as shown
 in \fig{fTracked}. The transformation $origin \rightarrow viewer$ is set by
-the application using \textsf{Config::setHeadMatrix}, which is used by
-the server to compute the frustra. The resulting off-axis frustra are
+the application using \textsf{Config::setHead\-Matrix}, which is used by
+the server to compute the frusta. The resulting off-axis frusta are
 positioned using the channel's head transformation, which can be
@@
 to order the input frames or to optimize the readback. The following
-sections provide a detailed description of the image compositing API in
-Equalizer.
+sections describe the image compositing API in Equalizer.
 
@@
 
-In order to provide a motivation for the design of the image compositing
+To provide a motivation for the design of the image compositing
 API, the direct send parallel compositing algorithm is introduced in this
@@
 
-The following operations have to be possible in order to perform this
+The following operations have to be possible to perform this
 algorithm:
@@
 The frame synchronization through the frame data allows the input frame
-to wait for the pixel data to become ready, which is signalled by the
+to wait for the pixel data to become ready, which is signaled by the
 output frame after readback.
@@
 
+The offset of input and output frames characterizes the position of the
+frame data relative to the framebuffer, that is, the \textbf{window's}
+lower-left corner. For output frames this is the position of the channel
+relative to the window.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The offset of input and output frames characterizes the position of the
-frame data with respect to the framebuffer, that is, the
-\textbf{window's} lower-left corner. For output frames this is the
-position of the channel with respect to the window.
-
 For output frames, the frame data's pixel viewport is the area of the
@@
 view-aligned slices. The slices are rendered back-to-front and blended
-together to produce the final image, as shown in
-\fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382 of the German
-  Research Council (DFG)}.
+to produce the final image, as shown in \fig{fSlices}\footnote{Volume
+  Data Set courtesy of: SFB-382 of the German Research Council (DFG)}.
 
@@
 \end{wrapfigure}
-For recomposition, the 2D frame buffer contents are blended together to
-form a seamless picture. For correct blending, the frames are ordered in
-the same back-to-front order as the slices used for rendering, and use the
+For recomposition, the 2D frame buffer contents are blended to form a
+seamless picture. For correct blending, the frames are ordered in the
+same back-to-front order as the slices used for rendering, and use the
 same blending parameters. Simplified, the frame buffer images are
@@
 
-Finding the correct assembly order for perspective frustra is more
+Finding the correct assembly order for perspective frusta is more
 complex. The perspective distortion invalidates a simple orientation
-criteria like the one used for orthographic frustra. For the view and
+criteria like the one used for orthographic frusta. For the view and
 frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
@@
   {\caption{\small\label{fBlend}Back-to-Front Com\-po\-siting for
-      Orthogonal and Perspective View Frustra}}
+      Orthogonal and Perspective View Frusta}}
 \end{figure}
 
-In order to compute the assembly order, \textsf{eVolve} uses the angle
-between the $origin \rightarrow slice$ vector and the near plane, as shown
-in \fig{fBlendPerspective}. When the angle becomes greater than
+To compute the assembly order, \textsf{eVolve} uses the angle between
+the $origin \rightarrow slice$ vector and the near plane, as shown in
+\fig{fBlendPerspective}. When the angle becomes greater than
 90\textdegree, the compositing order of the remaining frames has to be
@@
 the same time by the server during \textsf{Config::init}. Each statistic
-event records the originator's (channel, window or config) unique
+event records the originator's (channel, window, node or config) unique
 identifier.
@@
 \fig{fStatistics} shows the visualization of statistics events in an
-overlay\footnote{3D model courtesy Stanford University Computer Graphics
-  Laboratory.}.
+overlay\footnote{3D model courtesy Cyberware, Inc.}.
 
@@
 
-The X axis is the time. One pixel on the screen corresponds to one
-millisecond. The right-most pixel is the most current time. On the Y
-axis are the rendering entities: channels, windows and the config. The
-upper channel is the local channel since it executes
-\textsf{frameAssemble}, and the lower channel is the remote channel,
-executing {frameReadback}.
-
-The remote machine is a Mac Mini, the local a MacBook Pro. The draw
-times are unbalanced due to the speed difference of the machines, and
-the transmit time is relatively long since 100 MBit Ethernet was used.
-
-In order to facilitate the understanding, older frames are gradually
-greyed out. The right-most, current frame is brighter than the
+The X axis is the time. One pixel on the screen corresponds to a given
+time unit, typically one millisecond. The scale is adapted dynamically
+to powers-of-ten milliseconds to fit the statistics into the available
+viewport. This allows easy and accurate evaluations of bottlenecks or
+misconfigurations in the rendering pipeline. The scale of the statistics
+is printed.
+
+The rightmost pixel is the most current time. On the Y axis are the
+rendering entities: channels, windows, nodes and the config. The upper
+channel is the local channel since it executes \textsf{frameAssemble},
+and the lower channel is the remote channel, executing {frameReadback}.
+
+The draw time has been artificially inflated by using immediate mode
+rendering to provide more expressive statistics.
+
+To facilitate the understanding, older frames are gradually grayed
+out. The rightmost, current frame is brighter than the
 frame-before-last.
@@
 
-The white \textsf{Window::swapBuffers} task takes also a long time in
-debug builds, since the execution of the swap buffer is locked to the
-vertical retrace of the display, and enforced by \textsf{glFinish}
-before sampling the end time of the event. In release builds, this does
-not happen and all OpenGL commands are pipelined with the execution on
-the CPU.
+The white \textsf{Window::swapBuffers} task might also take also a long
+time, since the execution of the swap buffer is locked to the vertical
+retrace of the display, and enforced by \textsf{glFinish} before
+sampling the end time of the event. In release builds, this does not
+happen and all OpenGL commands are pipelined with the execution on the
+CPU.
 
@@
 
+Note that the remote source window does not execute the swapBuffers in
+this configuration, since it is a single-buffered PBuffer.
+
 \newpage
@@
 
-The file format currently used is the one-to-one representation of the
+The current Equalizer file format is a one-to-one representation of the
 server's internal data structures. Its purpose is intermediate, that is,
-it will gradually be replaced by automatic resource dedection and
-automatic configuration. However the core scalability engine will
-always use a similar structure as currently exposed by the file format.
-
-The file format represents an ascii deserialization of the
+it will gradually be replaced by automatic resource detection and
+automatic configuration. However the core scalability engine will always
+use a similar structure as currently exposed by the file format.
+
+The file format represents an ASCII deserialization of the
 server. Streaming an \textsf{eq::server::Server} to an \textsf{eq::base::Log}
@@
 value: \textsf{S} for strings, \textsf{C} for a character, \textsf{I}
-for an integer and \textsf{F} for floating-point values. Enumation
+for an integer and \textsf{F} for floating-point values. Enumeration
 values are handled as integers. Strings have always to be surrounded by
@@
 Global attribute values have useful default parameters, which can be
-overriden with an environment variable of the same name. For enumeration
+overridden with an environment variable of the same name. For enumeration
 values the corresponding integer value has to be used. The global values
@@
   and then tries to launch the render client executable on this host.\\
+  \textbf{See also} & EQ\_CONNECTION\_SATTR\_LAUNCH\_COMMAND
+  EQ\_CONNECTION\_IATTR\_TCPIP\_PORT EQ\_CONNECTION\_IATTR\_TYPE\\
   \hline
@@
                    & ssh -n \%h \%c [WIN32]\\
-  \textbf{Details} & The command used by the server to launch nodes
+  \textbf{Details} & The command used by the server to auto-launch nodes
   which could not be connected. The launch command is executed from a
@@
   the server at runtime with concrete data: \%h is replaced by the
-  hostname, \%c by the command to launch, including command line
+  hostname, \%c by the render client command to launch, including command line
   arguments and \%n by a node-unique identifier. Each command line
   argument is surrounded by launch command quotes.\\
+  \textbf{See also} & EQ\_CONNECTION\_SATTR\_LAUNCH\_COMMAND\_QUOTE 
+  EQ\_CONNECTION\_IATTR\_LAUNCH\_TIMEOUT\\
   \hline
@@
   arguments might contain white spaces, and therefore have to be
-  surrounded by quotes to
-  identify their borders.\\
+  surrounded by quotes to identify their borders. This option is mostly
+  used on Windows.\\
   \hline
@@
   \textbf{Value} & unsigned\\
-  \textbf{Default} & 60'000\\
+  \textbf{Default} & 60'000 (1 minute)\\
   \textbf{Details} & Defines the timeout in milliseconds to wait for
@@
   frustum during stereo rendering. See \sref{sTracking} for details.\\
+  \textbf{See also} & EQ\_WINDOW\_IATTR\_HINT\_STEREO 
+  EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
   \hline
@@
   \sref{sThreadModel} for details.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_NODE\_IATTR\_HINT\_STATISTICS\\
+  \textbf{Value} & OFF \textbar \ FASTEST [ON] \textbar \ NICEST\\
+  \textbf{Default} & FASTEST [Release Build]\\
+                   & NICEST [Debug Build]\\
+  \textbf{Details} & Determines how statistics are gathered. Fastest and
+  nicest currently gather statistics in the same way.\\
+  \textbf{See also} & EQ\_WINDOW\_IATTR\_HINT\_STATISTICS 
+  EQ\_CHANNEL\_IATTR\_HINT\_STATISTICS\\
   \hline
@@
   \textbf{Details} & Determines if the window selects a quad-buffered
-  stereo visual. When set to AUTO, the window initialization methods try
+  stereo visual. When set to AUTO, the default window initialization methods try
   to allocate a stereo visual for windows, but fall back to a mono
   visual if allocation fails. For PBuffers, AUTO selects a mono visual.\\
+  \textbf{See also} & EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
   \hline
@@
   \textbf{Details} & Determines if the window selects a double-buffered
-  stereo visual. When set to AUTO, the window initialization methods try
+  stereo visual. When set to AUTO, the default window initialization methods try
   to allocate a double-buffered visual for windows, but fall back to a
@@
   \textbf{Details} & When set to OFF, window borders and other
-  decorations are disabled, and typically the window can not be moved or
+  decorations are disabled, and typically the window cannot be moved or
   resized. This option is useful for source windows during
@@
   required. For optimal performance, set swap synchronization to OFF for
-  source-only windows.\\
+  source-only windows. This option has no effect on single-buffered windows.\\
   \hline
@@
   drawable. PBuffers are off-screen drawables created using window
-  system-dependent PBuffer API's. In order to calculate the PBuffer size
-  on unconnected devices, a pipe viewport size of 4096x4096 is assumed,
-  unless specified otherwise.\\
+  system-dependent PBuffer API's. To calculate the PBuffer size on
+  unconnected devices, a pipe viewport size of 4096x4096 is assumed,
+  unless specified otherwise using the pipe's viewport parameter.\\
   \hline
@@
   previous OpenGL commands.\\
+  \textbf{See also} & EQ\_NODE\_IATTR\_HINT\_STATISTICS 
+  EQ\_CHANNEL\_IATTR\_HINT\_STATISTICS\\
   \hline
@@
   \textbf{Details} & See EQ\_WINDOW\_IATTR\_HINT\_STATISTICS.\\
+  \textbf{See also} & EQ\_NODE\_IATTR\_HINT\_STATISTICS 
+  EQ\_WINDOW\_IATTR\_HINT\_STATISTICS\\
   \hline
@@
   \hline
-\end{tabularx}\\\vfill
+\end{tabularx}\\%\vfill
 
@@
   \hline
-\end{tabularx}\\\vfill
-
-\begin{tabularx}{\textwidth}{|l|X|}
-  \hline
-  \textbf{Name} & EQ\_COMPOUND\_IATTR\_UPDATE\_FOV\\
-  \textbf{Value} & HORIZONTAL [ON] \textbar \ VERTICAL \textbar \ OFF\\
-  \textbf{Default} & HORIZONTAL\\
-  \textbf{Details} & Define how the wall or projection description is
-  updated when a window is resized. Horizontal increases or decreases
-  the horizontal field of view, while keeping the vertical size
-  constant. Vertical updates the vertical field of view, while keeping
-  the horizontal size constant. OFF does not update the wall or
-  projection description at all.\\
-  \hline
-\end{tabularx}\\
+\end{tabularx}\\%\vfill
 \end{center}
@@
 is used to enforce a SDP connection. Note that you can also use the
-transparent mode provided by most InfiniBand implementations to used SDP
+transparent mode provided by most InfiniBand implementations to use SDP
 with TCP connections.
@@
 is specified for a server connection description, it listens to
-INADDR\_ANY, and is therefore reachable on all interfaces. If the
-server's hostname is specified, the listening socket is bound to this
-address. If any hostname given is not resolvable, or any port can not be
-used, server initialization will fail.
+INADDR\_ANY, and is therefore reachable on all network interfaces. If
+the server's hostname is specified, the listening socket is bound only
+to this address. If any of the given hostnames is not resolvable, or any
+port cannot be used, server initialization will fail.
 
@@
 all connection descriptions of the node are passed to the launched node
-process, which binds to all provided descriptions.
+process, which will caus it to bind to all provided descriptions.
 
@@
         thread_model    ASYNC | DRAW_SYNC | LOCAL_SYNC
+        hint_statistics FASTEST | NICEST | OFF
     }
@@
 the system. The GPU can be offline, in this case only PBuffer windows
-can be used. In order to restrict the OpenGL output to the GPU, the
+can be used. To restrict the OpenGL output to the GPU, the
 \textsf{WGL\_NV\_gpu\_affinity} extension is used. If the extension is
@@
 the compound. The frustum is inherited and typically only defined on the
-top-most compound. The last specified frustum description is used. Sizes
-are specified in meters. \fig{fFrustra} illustrates the frustum
+topmost compound. The last specified frustum description is used. Sizes
+are specified in meters. \fig{fFrusta} illustrates the frustum
 parameters.
@@
 \pagestyle{empty}
-\begin{textblock}{}(0,5)
+\begin{textblock}{}(0,4.5)
   \hspace{-1cm}\includegraphics[width=22cm]{images/teaserBack.pdf}

### 2008-12-08 Developer/ProgrammingGuide/paper.tex
Programming Guide: updates based on print version
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     parallel rendering framework}\\\vspace{1cm}
-  Version 1.4 for Equalizer 0.6\\[\medskipamount]
+  Version 1.5.1 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
     1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
+    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
     \hline
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/statistics.png}
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/statistics.png}
   {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}

### 2009-01-10 Developer/ProgrammingGuide/paper.tex
Programming Guide: Add compositor documentation
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     parallel rendering framework}\\\vspace{1cm}
-  Version 1.5.1 for Equalizer 0.6+svn\\[\medskipamount]
+  Version 1.5.2 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
 
-\textcopyright 2007-2008 \htmladdnormallink{Eyescale Software GmbH}
+\textcopyright 2007-2009 \htmladdnormallink{Eyescale Software GmbH}
 {http://www.eyescale.ch}. All rights reserved. No permission is
@@
     1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
+    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
     \hline
@@
 
+\subsubsection{\label{sCompositor}The Compositor}
+
+%\begin{wrapfigure}{r}{.618\textwidth}
+\begin{figure}[h!t]\center
+  \includegraphics[width=.9\textwidth]{images/compositor.pdf}
+  {\caption{\small\label{fCompositor}Functional diagram of the compositor}}
+\end{figure}
+The \textsf{Compositor} class gathers a set of static functions which
+implement the various compositing algorithms and low-level
+optimizations. \fig{fCompositor} provides a top-down functional overview
+of the various compositor functions.
+
+On a high level, the compositor combines multiple input frames using 2D
+tiling, depth-compositing for polygonal data or sorted, alpha-blended
+compositing for semi-transparent volumetric data. These operations
+composite either directly all images on the GPU, or use a CPU-based
+compositor and then transfer the pre-integrated result to the GPU. The
+high-level entry points automatically select the best algorithm. The
+CPU-based compositor uses, based on availability, OpenMP and the
+Paracomp compositing library to accelerate its operation.
+
+On the next lower level, the compositor provides functionality to
+composite a single frame, either using 2D tiling (possibly with blending
+for alpha-blended compositing) or depth-based compositing.
+
+The per-frame compositing in turn relies on the per-image compositing
+functionality, which automatically decides on the algorithm to be used
+(2D or depth-based). The concrete per-image assembly operation uses
+OpenGL operations to composite the pixel data into the framebuffer,
+potentially using GLSL for better performance.
+
 \subsubsection{Custom Assembly in eVolve}
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
-\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
@@
 \begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
 \end{wrapfigure}
@@
 
+\begin{wrapfigure}{l}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 Finding the correct assembly order for perspective frusta is more
@@
 1-4-3-2.
+
+To compute the assembly order, \textsf{eVolve} uses the angle between
+the $origin \rightarrow slice$ vector and the near plane, as shown in
+\fig{fBlendPerspective}. When the angle becomes greater than
+90\textdegree, the compositing order of the remaining frames has to be
+changed. The result image of this composition naturally looks the same
+as the volume rendering would when rendered on a single
+channel. \fig{fVolResult} shows the result of the composition from
+\fig{fBlendPerspective}.
 
@@
 
-To compute the assembly order, \textsf{eVolve} uses the angle between
-the $origin \rightarrow slice$ vector and the near plane, as shown in
-\fig{fBlendPerspective}. When the angle becomes greater than
-90\textdegree, the compositing order of the remaining frames has to be
-changed. The result image of this composition naturally looks the same
-as the volume rendering would when rendered on a single
-channel. \fig{fVolResult} shows the result of the composition from
-\fig{fBlendPerspective}.
-
 The assembly algorithm described in this section also works with parallel
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/statistics.png}
-  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
-\end{wrapfigure}
 Statistics are enabled per entity using an attribute hint. The hint
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/statistics.png}
+  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
+\end{wrapfigure}
 The events are processed by the channel's and window's

### 2009-01-13 Developer/ProgrammingGuide/paper.tex
Programming Guide: Add section describing network layer
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     parallel rendering framework}\\\vspace{1cm}
-  Version 1.5.2 for Equalizer 0.6+svn\\[\medskipamount]
+  Version 1.5.3 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
     1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
+    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
     \hline
@@
 
-\subsection{Distributed Objects}
+\subsection{\label{sNetObject}Distributed Objects}
 
@@
 
+\subsection{\label{sNetwork}The Network Layer}
+
+The Equalizer networking layer provides a peer-to-peer communication
+infrastructure. It is used by Equalizer to communicate between the
+application node, the server and the render clients. It can also be
+used by applications to implement distributed processing independently
+or complementary to the core Equalizer functionality.
+
+The network layer is implemented in the \textsf{eq::net} namespace. It
+provides layered functionality, which provide higher level functionality
+for the programmer. The main primitives in the network layer are:
+
+\begin{description}
+\item[Connection] A stream-oriented point-to-point communication
+  line. Different implementations of a connection exist.
+\item[Node] The abstraction of a process in the cluster. Nodes
+  communicate with each other using connections.
+\item[Session] Provides unique identifiers and object mapping to a set
+  of nodes.
+\item[Object] Provides object-oriented data distribution of C++ objects
+  between nodes within a session.
+\end{description}
+
+\subsubsection{Connections}
+
+The \textsf{eq::net::Connection} is the basic primitive used for
+communication between processes in Equalizer. It provides a
+stream-oriented communication between two endpoints.
+
+A connection is either closed, connected or listening. A closed
+connection can not be used for communications. A connected connection
+can be used to read or write data to the communication peer. A listening
+connection can accept connection requests.
+
+An \textsf{eq::net::ConnectionSet} is used to manage multiple
+connections. The typical use case is to have one or more listening
+connections for the local process, and a number of connected connections
+for communicating with other processes.
+
+The connection set is used to select one connection which requires some
+action. This can be a connection request on a listening connection,
+pending data on a connected connection or the notification of a
+disconnect.
+
+The connection and connection set can be used by applications to
+implement other networking-related functionality, e.g., to communicate
+with a sound server on a different machine.
+
+\subsubsection{Packet Handling}
+
+Nodes, sessions and objects communicate using packets derived from
+\textsf{eq::net::Packet}. The basic packet dispatch and handling is
+implemented in the \textsf{eq::net::Base} class, from which
+\textsf{eq::net::Node}, \textsf{eq::net::Session} and
+\textsf{eq::net::Object} is subclassed.
+
+The network base class allows the registration of a packets with a
+dispatch queue and an invocation method. Each packet has a command
+identifier, which is used to identify the registered queue and
+method. The method \textsf{dispatchCommand} pushes the packet to the
+registered queue, whereas the method \textsf{invokeCommand} calls the
+registered command function.
+
+A command function groups the method and \textsf{this} pointer, allowing
+to call a C++ method on a concrete instance. If no queue is registered
+for a certain command, \textsf{dispatchCommand} directly calls the
+registered command function.
+
+This dispatch and invocation functionality is used within Equalizer to
+dispatch commands from the receiver thread to the appropriate node or
+pipe thread, and then to invoke the command when it is processed by
+these threads.
+
+\subsubsection{Nodes}
+
+The \textsf{eq::net::Node} is the abstraction of one process in the
+cluster. Each node has a unique identifier, based on the UUID
+standard\footnote{\link{http://en.wikipedia.org/wiki/Universally\_Unique\_Identifier}}. This
+identifier is used to address nodes, e.g., in order to query connection
+information to connect to the node. Nodes use connections to communicate
+with each other by sending packets subclassed from
+\textsf{eq::net::NodePacket}.
+
+Nodes, like connections, are either in the closed, listening or
+connected state. A listening node uses one or more listening connections
+in conjunction with a connection set to wait for data or connection
+requests. For each connection request, a connection and node are created
+to represent this node on the local machine. This node is in the
+connected state, and its connection is added to the connection set of
+the listening node.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
+  {\caption{\small\label{fNetNode}Communication between two nodes}}
+\end{wrapfigure}
+\fig{fNetNode} shows the communication between two nodes. The nodes in
+the listening state also have a listening connection, which is not shown
+in the Figure for complexity reasons.
+
+When the remote node sends a packet, the listening node receives the
+packet and dispatches it from the receiver thread using the method
+\textsf{dispatchCommand}. The default implementation knows how to
+dispatch packets of type node, session or object. Applications can
+define custom data types for packets, and then have to extend
+\textsf{dispatchCommand} to handle these custom data types.
+
+Node packets are directly dispatched using
+\textsf{Base::dispatchCommand}. For session and object packets, the
+appropriate session is found and \textsf{Session::dispatchCommand} will
+dispatch the packet.
+
+If \textsf{dispatchCommand} returns false, the packet will be
+redispatched at a later time. This is used if an object has not been
+mapped locally, and therefore the packet could not be dispatched.
+
+If an application wants to extend communication on the node level, it
+can either define its own datatype for packets, or to define custom node
+packets. Packets with a custom datatype are handled by overriding
+\textsf{Node::dispatchCommand}. Custom node packets can use any command
+greater than \textsf{CMD\_NODE\_CUSTOM} as the packet command. By
+registering a \textsf{eq::net::CommandFunc} for these commands, the
+\textsf{eq::net::Base} dispatch and invoke mechanism can be used
+automatically.
+
+\subsubsection{Session}
+
+The \textsf{eq::net::Session} gathers provides higher-level
+functionality to a set of nodes. The master instance of a session is
+registered to a node, which makes this node the session server and
+assigns a node-unique identifier to the session. All other nodes map
+their instance of the session to their local listening node using the
+session's identifier. One given node can participate in any number of
+sessions, both as master and slave. This allows the flexible usage of
+sessions to group different aspects of an application.
+
+A session provides the facilitate to allocate session-unique
+identifiers. These identifiers can be used by the application to
+uniquely address data within the cluster. They are also used by the
+session to address distributed objects, which are described in the
+following section.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
+  {\caption{\small\label{fNetSession}Communication between two Sessions}}
+\end{wrapfigure}
+Sessions instances can communicate to each other directly using an
+\textsf{eq::net::Ses\-sion\-Packet}, which is derived from the
+\textsf{NodePacket} and extends it by the session identifier. The
+session ID is used by the node to dispatch the packet to the session.
+
+The session, similar to the node, uses the methods
+\textsf{dispatchCommand} and \textsf{invokeCommand} to dispatch and
+execute packets using either the \textsf{eq::net::Base} class for
+session packets or to dispatch and invoke object packets on the
+appropriate \textsf{eq::net::Object}. Application developers can use
+this mechanism by creating custom session packets with a command greater
+than \textsf{CMD\_SESSION\_CUSTOM} and registering a command function
+for these commands.
+
+\subsubsection{Object}
+
+Distributed objects provide powerful, object-oriented data
+distribution for C++ objects. Their functionality and an example use
+case for parallel rendering has been described in \sref{sNetObject}.
+
+The master instance of an \textsf{eq::net::Object} is registered with a
+session, which assigns a session-unique identifier to the object. Using
+this identifier, other nodes can map their slave instance of the object
+to their local proxy of the session.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
+  {\caption{\small\label{fNetObject}Communication between two objects}}
+\end{wrapfigure}
+Objects can send command packets to each other by subclassing
+\textsf{eq::net::Ob\-ject\-Packet}, which is derived from
+\textsf{SessionPacket}. The object packet extends the session packet by
+the object identifier and an object instance identifier.
+
+Multiple instances of the same object (identifier) can be mapped on the
+same node and session simultaneously. Each object instance has a
+node-unique instance identifier. When the instance ID is set to
+\textsf{EQ\_ID\_ANY} (the default value), a command is dispatch to all
+object instance registered to the the session with the correct object
+ID. Multiple object instance per node are used when different threads on
+the node need to process a different version of the same object.
+
+Distributed objects provide data synchronization functionality, as
+described in \sref{sNetObject}. Additionally, applications can send
+command packets by creating custom object packets with a command greater
+than \textsf{CMD\_OBJECT\_CUSTOM}.
+
+Distributed object communication relies on the lower node and session
+communication layers, as shown in \fig{fNetObject}.
+
+\subsubsection{Barrier}
+
+The \textsf{eq::net::Barrier} provides a networked barrier primitive. It
+is an \textsf{eq::net::Object} used by Equalizer for software swap
+barrier support, but it can be used as a generic barrier in application
+code.
+
+The barrier uses both the data distribution for synchronizing its data,
+as well as custom command packets to implement the barrier logics. A
+barrier is a versioned object. Each version can have a different height,
+and enter request are automatically grouped together by version.
+
 \newpage

### 2009-02-09 Developer/ProgrammingGuide/paper.tex
Programming Guide: Add section on eq::Object
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     parallel rendering framework}\\\vspace{1cm}
-  Version 1.5.3 for Equalizer 0.6+svn\\[\medskipamount]
+  Version 1.5.4 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
     1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
+    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
     \hline
@@
   \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
-  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}\vspace{-2ex}
+  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
+  \vspace{-3ex}
 \end{wrapfigure}
@@
 tasks are executed by calling the appropriate task methods (see
-\sref{ssTaskMethods}) in the correct thread and context. The application
+\sref{sTaskMethods}) in the correct thread and context. The application
 either implements the task methods with application-specific code or
@@
   \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
-  {\caption{\label{fPixel}\small A pixel compound}}\vspace{-5ex}
+  {\caption{\label{fPixel}\small A pixel compound}}
+  \vspace{-6ex}
 \end{wrapfigure}
@@
 
-\subsection{\label{ssTaskMethods}Task Methods}
+\subsection{\label{sTaskMethods}Task Methods}
 
@@
 \subsubsection{\label{sFrustum}Frustum}
-Compounds have a frustum description to define the physical layout of
-the display environment.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
   {\caption{\small\label{fFrusta}Wall and projection parameters}}
+  \vspace{-3ex}
 \end{wrapfigure}
+Compounds have a frustum description to define the physical layout of
+the display environment.
+
 The frustum description is inherited by the children, therefore the
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Render Loop}}
-\end{wrapfigure}
 The model data is spatially organized in a 3-dimensional
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\small\label{fRender}Main Render Loop}}
+\end{wrapfigure}
 The actual rendering uses display lists or vertex buffer objects. These
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
+\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
@@
 visualize with any other approach.
-
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
-\end{wrapfigure}
-For recomposition, the 2D frame buffer contents are blended to form a
-seamless picture. For correct blending, the frames are ordered in the
-same back-to-front order as the slices used for rendering, and use the
-same blending parameters. Simplified, the frame buffer images are
-`thick' slices which are `rendered' by writing their content to the
-destination frame buffer using the correct order.
-
-For orthographic rendering, determining the compositing order of the
-input frames is trivial. The screen-space orientation of the volume
-bricks determines the order in which they have to be composited. The
-bricks in \textsf{eVolve} are created by slicing the volume along one
-dimension. Therefore the range of the resulting frame buffer images,
-together with the sorting order, is used to arrange the frames during
-compositing. \fig{fBlendOrtho} shows this composition for one view.
-
-\begin{wrapfigure}{l}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
-Finding the correct assembly order for perspective frusta is more
-complex. The perspective distortion invalidates a simple orientation
-criteria like the one used for orthographic frusta. For the view and
-frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
-  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
-1-4-3-2.
-
-To compute the assembly order, \textsf{eVolve} uses the angle between
-the $origin \rightarrow slice$ vector and the near plane, as shown in
-\fig{fBlendPerspective}. When the angle becomes greater than
-90\textdegree, the compositing order of the remaining frames has to be
-changed. The result image of this composition naturally looks the same
-as the volume rendering would when rendered on a single
-channel. \fig{fVolResult} shows the result of the composition from
-\fig{fBlendPerspective}.
 
@@
 
+For recomposition, the 2D frame buffer contents are blended to form a
+seamless picture. For correct blending, the frames are ordered in the
+same back-to-front order as the slices used for rendering, and use the
+same blending parameters. Simplified, the frame buffer images are
+`thick' slices which are `rendered' by writing their content to the
+destination frame buffer using the correct order.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
+For orthographic rendering, determining the compositing order of the
+input frames is trivial. The screen-space orientation of the volume
+bricks determines the order in which they have to be composited. The
+bricks in \textsf{eVolve} are created by slicing the volume along one
+dimension. Therefore the range of the resulting frame buffer images,
+together with the sorting order, is used to arrange the frames during
+compositing. \fig{fBlendOrtho} shows this composition for one view.
+
+Finding the correct assembly order for perspective frusta is more
+complex. The perspective distortion invalidates a simple orientation
+criteria like the one used for orthographic frusta. For the view and
+frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
+  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
+1-4-3-2.
+
+To compute the assembly order, \textsf{eVolve} uses the angle between
+the $origin \rightarrow slice$ vector and the near plane, as shown in
+\fig{fBlendPerspective}. When the angle becomes greater than
+90\textdegree, the compositing order of the remaining frames has to be
+changed. The result image of this composition naturally looks the same
+as the volume rendering would when rendered on a single
+channel. \fig{fVolResult} shows the result of the composition from
+\fig{fBlendPerspective}.
+
 The assembly algorithm described in this section also works with parallel
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/statistics.png}
+  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
+\end{wrapfigure}
 Statistics are enabled per entity using an attribute hint. The hint
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/statistics.png}
-  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
-\end{wrapfigure}
 The events are processed by the channel's and window's
@@
 
-\subsection{\label{sNetwork}The Network Layer}
+\section{\label{sNetwork}The Network Layer}
 
@@
 
-\subsubsection{Connections}
+\subsection{Connections}
 
@@
 
-\subsubsection{Packet Handling}
+\subsection{Packet Handling}
 
@@
 
-\subsubsection{Nodes}
+\subsection{Nodes}
 
@@
 
-\subsubsection{Session}
+\subsection{Session}
 
@@
 
-\subsubsection{Object}
+\subsection{\label{sNetObject2}Object}
 
@@
 
+Objects can send command packets to each other by subclassing
+\textsf{eq::net::Ob\-ject\-Packet}, which is derived from
+\textsf{SessionPacket}. The object packet extends the session packet by
+the object identifier and an object instance identifier.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-Objects can send command packets to each other by subclassing
-\textsf{eq::net::Ob\-ject\-Packet}, which is derived from
-\textsf{SessionPacket}. The object packet extends the session packet by
-the object identifier and an object instance identifier.
-
 Multiple instances of the same object (identifier) can be mapped on the
@@
 
-\subsubsection{Barrier}
+\subsection{Barrier}
 
@@
 and enter request are automatically grouped together by version.
+
+\subsection{\label{sNetUsage}Usage in Equalizer}
+
+The Equalizer client library and server are built on top of the network
+layer. They influence the design of it, and serve as a sample
+implementation on how to use classes in the \textsf{eq::net} namespace.
+
+\subsubsection{eq::Object}
+
+Equalizer uses the \textsf{eq::net::Object} data distribution mechanism
+for implementing data distribution between the server, application and
+render clients. To do this, \textsf{eq::Object} implements one usage
+pattern for the \textsf{eq::net::Object}, allowing inheritance with data
+distribution. It is easier to use, but imposes one typical way to implement data
+distribution.
+
+The \textsf{eq::Object} data distribution is centered around the concept
+of dirty bits. Dirty bits form a 64-bit mask which marks which parts of
+the object have to be distributed during the next commit.
+
+For serialization, the default \textsf{eq::net::Object} serialization
+functions are implemented by \textsf{eq::Object}, which (de-)serializes
+and resets the dirty bits, and calls \textsf{serialize} or
+\textsf{deserialize} with the bit mask specifiying which data has to be
+transmitted or received. During a commit or sync, the current dirty bits
+are given, whereas during object mapping all dirty bits are passed to
+the serialization methods.
+
+In order to use \textsf{eq::Object}, the following steps have to be
+taken:
+
+\begin{description}
+\item[Inherit from \textsf{eq::Object}:] The base class will provide the
+  dirty bit management and call serialize and deserialize
+  appropriately. By overriding \textsf{getChangeType}, the default
+  versioning strategy might be changed.
+\item[Define new dirty bits:] Define dirty bits for each data item by
+  starting at \textsf{Object::DIRTY\_CUSTOM}, shifting this value
+  consecutively by each new dirty bit.
+\item[Implement serialize and deserialize:] For each object-specific
+  dirty bit which is set, stream the corresponding data item to or from
+  the provided stream. Call the parent method first in both
+  functions. For application-specific objects, write a (de-)serialization
+  function.
+\item[Mark dirty data:] In each 'setter' method, call \textsf{setDirty}
+  with the corresponding dirty bit.
+\end{description}
+
+The inheritance chain \textsf{eq::View} $\rightarrow$
+\textsf{eq::Frustum} $\rightarrow$ \textsf{eq::Object} $\rightarrow$
+\textsf{eq::net::Object} serves as sample implementation for this
+process. The \textsf{eq::Object} base class provides the facility to set
+a name on the object, which is automatically distributed.
+
+The registration and mapping of \textsf{eq::Object}s is done in the same
+way as for \textsf{eq::net::Object}s, which has been described in
+\sref{sNetObject2}. Equalizer objects derived from \textsf{eq::Object},
+e.g., \textsf{eq::View}, are typically registered and mapped by
+Equalizer.
 

### 2009-03-10 Developer/ProgrammingGuide/paper.tex
Update Programming Guide with new AGL event dispatch
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     parallel rendering framework}\\\vspace{1cm}
-  Version 1.5.4 for Equalizer 0.6+svn\\[\medskipamount]
+  Version 1.5.5 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
     1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
+    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
     \hline
@@
 
-In general, events are received and processed by the pipe thread a
-window belongs to. An exception to this rule is AGL, where all events
-are dispatched from the main node thread. WGL and GLX receive and
-process the events from the pipe threads that created the
-windows. Whenever the term \textbf{event thread} is used, it refers to
-the thread receiving the event, i.e., the pipe thread for WGL and GLX,
-and the main thread for AGL.
+Events are received and processed by the pipe thread a window belongs
+to. For AGL, Equalizer internally forwards the event from the main
+thread, where it is received, to the appropriate pipe thread.
 
@@
 To dispatch events, Equalizer 'pumps' the native events. On WGL and GLX,
-this happens on each thread with windows, whereas on AGL it happens only
-on the main thread. By default, Equalizer pumps these events
-automatically for the application in-between executing task methods.
+this happens on each thread with windows, whereas on AGL it happens on
+the main thread and each pipe thread. By default, Equalizer pumps these
+events automatically for the application in-between executing task
+methods.
 
@@
 determine if the pipe's thread shall automatically dispatch OS
-events. For non-threaded and AGL pipes, the method affects the node
-thread. On WGL and GLX, messages are dispatched in the application and
-each pipe thread. On AGL, events are dispatched only on the application
-thread.
+events. For AGL, this affects the node threads and pipe thread, since
+the node thread message pump needs to dispatch the events to the pipe
+thread event queue.
 
@@
 
-Carbon events issued for AGL windows are processed from the node's main
-thread. One \textsf{AGLEventHandler} per process is used. When a Carbon
-window is set on an \textsf{AGLWindow}, the window is registered with
-the global \textsf{AGLEventHandler}. The event handler installs a Carbon
+Carbon events issued for AGL windows are initially receive by the node
+main thread, and automatically dispatched to the pipe thread. The pipe
+thread will dispatch the event to the appropriate event handler. One
+\textsf{AGLEventHandler} per window is used, which is created during
+\textsf{AGLWindow::configInit}. The event handler installs a Carbon
 event handler for all important events. The event handler uses an
@@
 \textsf{AGLWindowIF::processEvent}. During window exit, the installed
-Carbon handler is removed when the window is deregistered with the
-\textsf{AGLEventHandler}. No event handling is set up for PBuffers.
+Carbon handler is removed when the window's event handler is deleted. No
+event handling is set up for PBuffers.
 
@@
 
+When using 2D (sort-first) or stereo decompositions, no special
+programming is needed to achieve good scalability, as \textsf{eVolve} is
+mostly fill-limited and therefore scales nicely in these modes.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
-When using 2D (sort-first) or stereo decompositions, no special
-programming is needed to achieve good scalability, as \textsf{eVolve} is
-mostly fill-limited and therefore scales nicely in these modes.
-
 The full power of scalable volume rendering is however in DB (sort-last)
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \vspace{-3ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+  \vspace{-5ex}
+\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended to form a
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
+  {\caption{\small\label{fNetObject}Communication between two objects}}
+\end{wrapfigure}
 Distributed objects provide powerful, object-oriented data
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
-  {\caption{\small\label{fNetObject}Communication between two objects}}
-\end{wrapfigure}
 Multiple instances of the same object (identifier) can be mapped on the

### 2009-03-23 Developer/ProgrammingGuide/paper.tex
PG: Introduction to Layout API
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-\usepackage{tabularx,longtable,graphicx,a4,listings,wrapfig,subfigure,textcomp}
+\usepackage{tabularx,longtable,graphicx,a4,listings,wrapfig,subfigure,textcomp,ccaption}
 \usepackage[absolute]{textpos}
@@
 \newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}
+\indentcaption{2em}
 
@@
   \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
-  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
+  {\caption{\label{fExecutionFlow}Parallel Rendering}}
   \vspace{-3ex}
@@
   \includegraphics[width=.618\textwidth]{images/processes.pdf}
-  {\caption{\small\label{fProcesses}Equalizer Processes}}
+  {\caption{\label{fProcesses}Equalizer Processes}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/eqHello.png}
-  {\caption{\small\label{fHello}Hello, World!}}
+  {\caption{\label{fHello}Hello, World!}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/2D.pdf}
-  {\caption{\small A 2D compound}}
+  {\caption{ A 2D compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
-  {\caption{\small A DPlex compound}}
+  {\caption{ A DPlex compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.9\textwidth]{images/pixelKernels.pdf}
-  {\caption{\small\label{fPixelKernels}Example pixel kernels for a
+  {\caption{\label{fPixelKernels}Example pixel kernels for a
       four-to-one pixel compound}}
@@
   \includegraphics[width=.618\textwidth]{images/model.pdf}
-  {\caption{\small\label{fModel}Simplified execution model}}
+  {\caption{\label{fModel}Simplified execution model}}
 \end{wrapfigure}
@@
   \includegraphics[width=\textwidth]{images/cave.pdf}
-  {\caption{\small\label{fConfig}An example configuration}}
+  {\caption{\label{fConfig}An example configuration}}
 \end{figure}
@@
 
+\subsection{Canvases}
+
+Canvases provide a convenient way to configure projection surfaces. A
+canvas uses a layout, which describes the logical views. The usage of
+canvases and layouts is optional. Typically, each desktop window uses
+one canvas.
+
+A canvas represents one logical projection surface, e.g., a powerwall, a
+curved screen or an immersive installation. One configuration might
+drive multiple canvases, for example an immersive installation and an
+operator station.
+
+A canvas consists of one or more segments. A planar canvas typically has
+a frustum description (see \sref{sFrustum}), which is inherited by the
+segments. Non-planar frusta are configured using the segment
+frusta. These frusta typically describe a physically-correct display
+setup for Virtual Reality installations.
+
+\subsubsection{Segments}
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
+  {\caption{\label{fCanvas}A canvas using four, non-consecutive segments}}
+\end{wrapfigure}
+A segment represents one output channel of the canvas, e.g., a projector
+or a LCD. A segment has an output channel, which references the channel to
+which the display device is connected.
+
+A segment covers a part of its parent canvas, which is configured using
+the segment viewport. The viewport is in normalized coordinates with
+respect to the canvas. Segments might overlap (edge-blended projectors)
+or have gaps between each other (\fig{fCanvas}\footnote{Dataset courtesy
+  of VolVis distribution of SUNY Stony Brook, NY, USA.}). The viewport
+is used to configure the segment's default frustum from the canvas
+frustum description, and to place layout views correctly.
+
+Passive stereo installations have one segment per eye pass, where the
+segment for the left and right eye fully overlap.
+
+\subsection{Layouts}
+
+A layout is the grouping of logical views. It is used by one or more
+canvases. For all given layouts, Equalizer creates destination channels
+when the configuration file is loaded. These destination channels can be
+referenced by compounds to configure scalable rendering.
+
+Layouts can be switched at runtime by the application. Switching a
+layout will activate different destination channels for rendering.
+
+\subsubsection{Views}
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/layout.png}
+  {\caption{\label{fLayout}A layout with four views}}
+\end{wrapfigure}
+A view is a logical view of the application data, in the sense used by
+the Model-View-Controller pattern. It can be a scene, viewing mode,
+viewing position, or any other representation of the application's data.
+
+A view has a fractional viewport with respect to its layout.  A layout
+is often fully covered by its views, but this is not a requirement.
+
+Each view can have a frustum description. The view's frustum overrides
+frusta specified at the canvas or segment level. This is typically used
+for non-physically correct rendering, e.g., to compare two models
+side-by-side on a canvas. If the view does not specify a frustum, it
+will use the sub-frustum resulting from the covered area on the canvas.
+
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
+  {\caption{A display wall using one canvas and layout with six segments
+    and two views.}}
+\end{figure}
+
 \subsection{\label{sCompounds}Compounds}
@@
   \includegraphics[width=.618\textwidth]{images/frusta.pdf}
-  {\caption{\small\label{fFrusta}Wall and projection parameters}}
+  {\caption{\label{fFrusta}Wall and projection parameters}}
   \vspace{-3ex}
@@
   \includegraphics[width=\textwidth]{images/uml}
-  {\caption{\small\label{fUml}UML diagram of significant Equalizer and
-      eqPly classes and task methods.}}
+  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes.}}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{images/syncAsync}
-  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
+  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
 \end{figure}
@@
   \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
-  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
+  {\caption{\label{fModelDist}Scene Data in eqPly}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/objects.pdf}
-  {\caption{\small\label{fObjects}Scene Graph Distribution}}
+  {\caption{\label{fObjects}Scene Graph Distribution}}
 \end{wrapfigure}
@@
   \includegraphics[width=.9\textwidth]{images/configInit.pdf}
-  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
+  {\caption{\label{fConfigInit}Config Initialization Sequence}}
 \end{figure}
@@
   \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
-  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
+  {\caption{\label{fOSWindow}eq::OSWindow UML class hierarchy}}
 \end{wrapfigure}
@@
   \includegraphics[width=.382\textwidth]{images/DB.png}
-  {\caption{\small\label{fDBDest}Destination view of a DB compound}}
+  {\caption{\label{fDBDest}Destination view of a DB compound}}
 \end{wrapfigure}
@@
   \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\small\label{fRender}Main Render Loop}}
+  {\caption{\label{fRender}Main Render Loop}}
 \end{wrapfigure}
@@
   \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\small\label{fEventProcessing}Event Processing}}
+  {\caption{\label{fEventProcessing}Event Processing}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
+  {\caption{\label{fEventUML}UML Diagram for Event Handling Classes}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/threads.pdf}
-  {\caption{\small\label{fThreads}Threads within one node process}}
+  {\caption{\label{fThreads}Threads within one node process}}
 \end{wrapfigure}
@@
   \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
-  {\caption{\small\label{fThreadModels}Async, draw\_sync and local\_sync
+  {\caption{\label{fThreadModels}Async, draw\_sync and local\_sync
     thread synchronization models}}
@@
   \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
+  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
 \end{wrapfigure}
@@
   \includegraphics[width=\textwidth]{images/mainloop.pdf}
-  {\caption{\small\label{fMainLoop}Synchronization of frame tasks}}
+  {\caption{\label{fMainLoop}Synchronization of frame tasks}}
 \end{figure}
@@
     \includegraphics[height=3.5cm]{images/tracked}}%
-  {\caption{\small\label{fImmersive}Monoscopic, Stereoscopic and Tracked
+  {\caption{\label{fImmersive}Monoscopic, Stereoscopic and Tracked
     frusta}}
@@
   \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
-  {\caption{\small\label{fViewUpdate}Event Flow during a View Update}}
+  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
+  {\caption{\label{fDirectSend}Parallel Direct Send Compositing}}
   \vspace{-4ex}
@@
   \includegraphics[width=.618\textwidth]{images/assembly.pdf}
-  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
+  {\caption{\label{fAssembly}Hierarchy of assembly classes}}
 \end{wrapfigure}
@@
   \includegraphics[width=.9\textwidth]{images/compositor.pdf}
-  {\caption{\small\label{fCompositor}Functional diagram of the compositor}}
+  {\caption{\label{fCompositor}Functional diagram of the compositor}}
 \end{figure}
@@
   \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
+  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
 \end{wrapfigure}
@@
   }%
-  {\caption{\small\label{fBlend}Back-to-Front Com\-po\-siting for
-      Orthogonal and Perspective View Frusta}}
+  {\caption{\label{fBlend}Back-to-Front Com\-po\-siting for
+      Orthogonal and Perspective Frusta}}
 \end{figure}
@@
   \vspace{-3ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
+  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
   \vspace{-5ex}
@@
   \includegraphics[width=.618\textwidth]{images/statistics.png}
-  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
+  {\caption{\label{fStatistics}Statistics for a two-node 2D compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=\textwidth]{images/statisticsDetail.pdf}
-    {\caption{\small\label{fStatisticsDetail}Annotated Detail of the
+    {\caption{\label{fStatisticsDetail}Annotated Detail of the
         Statistics from \fig{fStatistics}.}}
@@
   \includegraphics[width=.618\textwidth]{images/netNode.pdf}
-  {\caption{\small\label{fNetNode}Communication between two nodes}}
+  {\caption{\label{fNetNode}Communication between two nodes}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/netSession.pdf}
-  {\caption{\small\label{fNetSession}Communication between two Sessions}}
+  {\caption{\label{fNetSession}Communication between two Sessions}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/netObject.pdf}
-  {\caption{\small\label{fNetObject}Communication between two objects}}
+  {\caption{\label{fNetObject}Communication between two objects}}
 \end{wrapfigure}

### 2009-03-23 Developer/ProgrammingGuide/paper.tex
PG: Introduction to Layout API
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   {\caption{\label{fFrusta}Wall and projection parameters}}
-  \vspace{-3ex}
 \end{wrapfigure}
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
   {\caption{\label{fRender}Main Render Loop}}
@@
   {\caption{\label{fDirectSend}Parallel Direct Send Compositing}}
-  \vspace{-4ex}
 \end{wrapfigure}
@@
 \begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-3ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
   {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
-  \vspace{-5ex}
 \end{wrapfigure}

### 2009-03-23 Developer/ProgrammingGuide/paper.tex
PG: Document layout API
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     parallel rendering framework}\\\vspace{1cm}
-  Version 1.5.5 for Equalizer 0.6+svn\\[\medskipamount]
+  Version 1.5.6 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
     1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
+    1.5.6   & Mar 23, 2009 & Documented Layout API\\
     \hline
@@
 
-\subsection{Canvases}
+\subsection{\label{sCanvas}Canvases}
 
@@
 
-\subsection{Layouts}
+\subsection{\label{sLayout}Layouts}
 
@@
 
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/uml}
+  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes.}}
+\end{figure}
+
 In this section the source code of \textsf{eqPly} is explained in
@@
 quality.
-
-\begin{figure}[ht!]\center
-  \includegraphics[width=\textwidth]{images/uml}
-  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes.}}
-\end{figure}
 
@@
         { return new eqPly::Channel( parent ); }
+    virtual eq::View* createView() { return new eqPly::View(); }
 };
@@
 
-Distributed objects are created by subclassing from
-\textsf{eq::net::Object}. Distributed objects can be static (immutable) or
-dynamic. Dynamic objects are versioned.
+Distributed objects are created by subclassing from \textsf{eq::Object}
+or \textsf{eq::net::Object}. Distributed objects can be static
+(immutable) or dynamic. Dynamic objects are versioned.
 
@@
 Equalizer determines the way changes are to be handled by calling
-\textsf{Object::getChangeType} during the registration of the master
+\textsf{Object::get\-Change\-Type} during the registration of the master
 version of a distributed object. The change type determines the memory
@@
 currently implemented have the following characteristics:
+
 \begin{itemize}
@@
 
-\textsf{eqPly} does not track changes in the \textsf{FrameData} since it
-is relatively small and changes frequently. Therefore, the instance and
-delta data are the same, and \textsf{pack} and \textsf{unpack} are not
-implemented. The change type is \textsf{INSTANCE}. The serialization
-functions exploit the fact that the frame data is layed out in one
-anonymous structure in memory, and transmit all the data in one call:
-
-{\footnotesize\begin{lstlisting}
-void FrameData::getInstanceData( eq::net::DataOStream& os )
-{ 
-    os.writeOnce( &data, sizeof( data )); 
-}
-
-void FrameData::applyInstanceData( eq::net::DataIStream& is )
-{
-    memcpy( &data, is.getRemainingBuffer(), sizeof( data ));
-    is.advanceBuffer( sizeof( data ));
+The \textsf{eq::Object} implements one possible way of tracking data
+changes in versioned objects using the concept of dirty bits, which
+facilitates the implementation of the data distribution code.
+
+The \textsf{FrameData} object tracks changes by setting the appropriate
+dirty bit whenever some of its data is changed. The serialization
+methods are called by the \textsf{eq::Object} with the dirty bit mask
+needed to serialize all data or the changes since the last
+\textsf{commit}. The \textsf{FrameData} only defines its own dirty bits
+and serialization code:
+
+{\footnotesize\begin{lstlisting}
+/** The changed parts of the data since the last pack(). */
+enum DirtyBits
+{
+    DIRTY_CAMERA = eq::Object::DIRTY_CUSTOM << 0,
+    DIRTY_FLAGS  = eq::Object::DIRTY_CUSTOM << 1,
+    DIRTY_VIEW   = eq::Object::DIRTY_CUSTOM << 2,
+};
+
+void FrameData::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
+{
+    eq::Object::serialize( os, dirtyBits );
+    if( dirtyBits & DIRTY_CAMERA )
+        os << _translation << _rotation;
+    if( dirtyBits & DIRTY_FLAGS )
+        os << _modelID << _renderMode << _color << _ortho << _statistics
+           << _help << _wireframe;
+    if( dirtyBits & DIRTY_VIEW )
+        os << _currentViewID;
+}
+
+void FrameData::deserialize( eq::net::DataIStream& is,
+                             const uint64_t dirtyBits )
+{
+    eq::Object::deserialize( is, dirtyBits );
+    if( dirtyBits & DIRTY_CAMERA )
+        is >> _translation >> _rotation;
+    if( dirtyBits & DIRTY_FLAGS )
+        is >> _modelID >> _renderMode >> _color >> _ortho >> _statistics
+           >> _help >> _wireframe;
+    if( dirtyBits & DIRTY_VIEW )
+        is >> _currentViewID;
 }
@@
 
+\textsf{eqPly} uses static distributed objects to distribute the model
+loaded by the application. This approach can be easily extended to
+versioned objects to support dynamic data modifications.
+
+The kD-tree data structure and rendering code for the model is strongly
+separated from Equalizer, and kept in the separate namespace
+\textsf{mesh}. It can also be used in other rendering software, for
+example in a GLUT application. To keep this separation, an external
+'mirror' hierarchy is constructed around the tree. This hierarchy of
+\textsf{VertexBufferDist} nodes is responsible for cloning the model
+data on the remote render clients.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
-\textsf{eqPly} uses static distributed objects to distribute the model
-loaded by the application. This model can be easily extended to
-versioned objects to support dynamic data changes.
-
-The kD-tree data structure and rendering code for the model is strongly
-separated from Equalizer, and kept in the separate namespace
-\textsf{mesh}. It could also be used in other rendering software, for
-example in a GLUT application. To keep this separation, an external
-'mirror' hierarchy is constructed around the tree. This hierarchy of
-\textsf{VertexBufferDist} nodes is responsible for cloning the model
-data on the remote render clients.
-
-The identifier of the root object of this distributed hierarchy is passed
-as part of the \textsf{InitData}, and used during remote node
-initialization to map the model. \fig{fModelDist} shows the UML
-hierarchy of the model and distribution classes.
-
-Each \textsf{VertexBufferDist} object has one node of the model's data
-tree. It is serializing the data for this node. Furthermore, it mirrors
-the kD-tree by having a \textsf{VertexBufferDist} child for each child
-of its corresponding tree node. During serialization, the identifier of
-these children is sent to the remote nodes, which reconstruct the mirror
-distribution hierarchy and model data tree based on this data.
+The identifier of the model's root object of this distributed hierarchy
+is passed as part of the \textsf{InitData} for the default model, or as
+part of the \textsf{View} for each logical view. It is used on the
+render clients to map the model when it is needed for
+rendering. \fig{fModelDist} shows the UML hierarchy of the model and
+distribution classes.
+
+Each \textsf{VertexBufferDist} object corresponds to one node of the
+model's data tree. It is serializing the data for this
+node. Furthermore, it mirrors the kD-tree by having a
+\textsf{VertexBufferDist} child for each child of its corresponding tree
+node. During serialization, the identifier of these children is sent to
+the remote nodes, which reconstruct the mirror distribution hierarchy
+and model data tree based on this data.
 
@@
 %\begin{figure}[ht!]\center
+  \vspace{-2ex}
   \includegraphics[width=.618\textwidth]{images/objects.pdf}
@@
 
-\begin{figure}[ht!]\center
-%\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
-  {\caption{\label{fConfigInit}Config Initialization Sequence}}
-\end{figure}
 The config initialization happens in parallel, that is, all config
@@
 
+%\begin{figure}[ht!]\center
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
+  {\caption{\label{fConfigInit}Config Initialization Sequence}}
+\end{wrapfigure}
 The tasks are executed by the node and pipe threads in parallel. The
@@
 
+\subsubsection{Model Handling}
+
+Model data is handled by the config. Models in eqPly are static,
+and therefore the render clients only need to map one instance of the
+model per node.
+
+Multiple models can be loaded in \textsf{eqPly}. A configuration has a
+default model, stored in \textsf{InitData}, and one model per view,
+stored and distributed using \textsf{View}. The
+loaded models are evenly distributed over the available views of the
+configuration, as shown in \fig{fLayout}.
+
+The channel acquires the model during rendering from the config, using
+the model identifier from its current view, or from the frame data if
+no view is configured.
+
+The config maintains the mapped models, and lazily maps new models,
+which are registered by the application process:
+
+{\footnotesize\begin{lstlisting}
+const Model* Config::getModel( const uint32_t modelID )
+{
+    if( modelID == EQ_ID_INVALID ) // no model loaded by application
+        return 0;
+
+    // Accessed concurrently from pipe threads
+    eq::base::ScopedMutex< eq::base::SpinLock > _mutex( _modelLock );
+
+    const size_t nModels = _models.size();
+    EQASSERT( _modelDist.size() == nModels );
+
+    for( size_t i = 0; i < nModels; ++i )
+    {
+        const ModelDist* dist = _modelDist[ i ];
+        if( dist->getID() == modelID )
+            return _models[ i ];
+    }
+
+    _modelDist.push_back( new ModelDist );
+    Model* model = _modelDist.back()->mapModel( this, modelID );
+    EQASSERT( model );
+    _models.push_back( model );
+
+    return model;
+}
+\end{lstlisting}}
+
+\subsubsection{Layout and View Handling}
+
+For layout and model selection, \textsf{eqPly} maintains an active
+view. The identifier of the active view is stored in the frame data,
+which is used by the render client to highlight it using a different
+background color. The active view can be selected by clicking into a
+view, or by cycling through all the views using a keyboard shortcut:
+
+{\footnotesize\begin{lstlisting}
+bool Config::handleEvent( const eq::ConfigEvent* event )
+{
+    switch( event->data.type )
+    {
+        ...
+        case eq::Event::POINTER_BUTTON_PRESS:
+            _frameData.setCurrentViewID( event->data.context.view.id );
+            return true;
+        ...
+    }
+}
+
+bool Config::_handleKeyEvent( const eq::KeyEvent& event )
+{
+    switch( event.key )
+    {
+        ...
+        case 'v':
+        case 'V':
+        {
+            NextViewFinder finder( _frameData.getCurrentViewID( ));
+            accept( finder );
+            const eq::View* view = finder.getResult();
+            if( view )
+                _frameData.setCurrentViewID( view->getID( ));
+            else
+                _frameData.setCurrentViewID( EQ_ID_INVALID );
+            return true;
+        }
+        ...
+    }
+}
+\end{lstlisting}}
+
+The model of the active view can be changed using a keyboard
+shortcut. The model is view-specific, and therefore the model identifier
+for each view is stored on the view, which is used to retrieve the model
+on the render clients. View-specific data is not limited to a
+model. Applications can choose to make any application-specific data
+view-specific, e.g., cameras, rendering modes or annotations. A view is
+a generic concept for an application-specific view on data,
+\textsf{eqPly} is using different models to illustrate the concept:
+
+{\footnotesize\begin{lstlisting}
+bool Config::_handleKeyEvent( const eq::KeyEvent& event )
+{
+    switch( event.key )
+    {
+        ...
+        case 'm':
+        case 'M':
+        {
+            if( _modelDist.empty( )) // no models
+                return true;
+
+            // current model
+            const uint32_t viewID = _frameData.getCurrentViewID();
+            View* view = static_cast< View* >( findView( viewID ));
+            const uint32_t currentID = view ? 
+                view->getModelID() : _frameData.getModelID();
+
+            // next model
+            ModelDistVector::const_iterator i;
+            for( i = _modelDist.begin(); i != _modelDist.end(); ++i )
+            {
+                if( (*i)->getID() != currentID )
+                    continue;
+
+                ++i;
+                break;
+            }
+            if( i == _modelDist.end( ))
+                i = _modelDist.begin(); // wrap around
+
+            // set identifier on view or frame data (default model)
+            const uint32_t modelID = (*i)->getID();
+            if( view )
+                view->setModelID( modelID );
+            else
+                _frameData.setModelID( modelID );
+            return true;
+        }
+        ...
+    }
+}
+\end{lstlisting}}
+
+The layout of the canvas with the active view can also be dynamically
+switched using a keyboard shortcut. The first canvas using the layout is
+found, and then the next layout of the configuration is set on this
+canvas.
+
+Switching a layout causes the initialization and de-initialization task
+methods to be called on the involved channels, and potentially windows,
+pipes and nodes. This operation might fail, which will cause the config
+to stop running.
+
+Layout switching is typically used to change the presentation of views
+at runtime. The source code omitted for brevity.
+
 \subsection{Node}
@@
 created on the appropriate machine. Nodes are only instantiated on their
-render client processes, i.e., each process should only have one
+render client processes, i.e., each process will only have one
 instance of the \textsf{eq::Node} class. The application process might
@@
 The actual mapping of the static data is done by the config. The config
-first retrieves the distributed \textsf{InitData}. The object is
-directly unmapped since it is static, and the data has been retrieved.
-
-The init data contains, among other things, the identifier of the
-root of the model's kD-tree. Using this identifier, the model is mapped
-to the node, as explained in \sref{sSceneData}:
+retrieves the distributed \textsf{InitData}. The object is directly
+unmapped since it is static, and the data has been retrieved:
 
@@
     {
-        const bool mapped = mapObject( &_initData, initDataID );
-        EQASSERT( mapped );
+        EQCHECK( mapObject( &_initData, initDataID ));
         unmapObject( &_initData ); // data was retrieved, unmap immediately
@@
         EQASSERT( _initData.getID() == initDataID );
-
-    const uint32_t modelID = _initData.getModelID();
-    if( modelID == EQ_ID_INVALID ) // no model loaded by application
-        return true;
-
-    if( !_model )
-        _model = ModelDist::mapModel( this, modelID );
-
-    return (_model != 0);
 }
 \end{lstlisting}}
-
-Since the model data is static, the mirror tree for data distribution
-is only created temporarily to retrieve the data. This is done in a
-static function of the \textsf{VertexBufferDist} class. One instance of
-the distribution class is mapped to the model's identifier. During
-application of the instance data, it will map its children recursively:
-
-{\footnotesize\begin{lstlisting}
-mesh::VertexBufferRoot* VertexBufferDist::mapModel(eq::net::Session* session,
-                                                   const uint32_t modelID )
-{
-    VertexBufferDist modelDist( 0, 0 );
-
-    if( !session->mapObject( &modelDist, modelID ))
-    {
-        EQWARN << "Mapping of model failed" << endl;
-        return 0;
-    }
-
-    modelDist._unmapTree(); // No longer needed
-
-    return const_cast< mesh::VertexBufferRoot* >( modelDist._root );
-}
-\end{lstlisting}}%>>
 
@@
 
-The object manager is not strictly a part of the window. It
+The object manager is, strictly speaking, not a part of the window. It
 is mentioned here since the \textsf{eqPly} window uses an object manager.
@@
 obtain the OpenGL objects for a given data item. The address of the data
-item to be rendered is used as the key. All objects managed by the OM
-are reference counted. If an application releases the objects properly,
-they are automatically de-allocated. It is also possible to manually
-manage de-allocation of objects, which might be more convenient in some
-cases.
-
-Currently, support for display lists, VBO's, textures and shaders is
-implemented. For each object, the following functions are available:
+item to be rendered is used as the key.
+
+For the currently supported types of OpenGL objects please refer to the
+API documentation on the Equalizer website. For each object, the
+following functions are available:
 
@@
 \item[getObject( key )] returns the object associated with the given
-  key, or FAILED. Increases the reference count of existing objects.
+  key, or FAILED.
 \item[newObject( key )] allocates a new object for the given
   key. Returns FAILED if the object already exists or if the allocation
-  failed. Sets the reference count of a newly created object to one.
+  failed.
 \item[obtainObject( key )] convenience function which gets or obtains
@@
   object allocation failed.
-\item[releaseObject( key \textbar\ name )] decreases the reference count and
-  deletes the object if the reference count reaches zero.
-\item[deleteObject( key \textbar\ name )] manually deletes the object. To be
-  used if reference counting is not used.
+\item[deleteObject( key )] deletes the object.
 \end{description}
@@
 \begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/DB.png}
+  \vspace{-8ex}\includegraphics[width=.382\textwidth]{images/DB.png}
   {\caption{\label{fDBDest}Destination view of a DB compound}}
@@
     Config*          config = static_cast< Config* >( getConfig( ));
-    const Model*     model  = config->getModel();
+    const Model*     model  = getModel();
     const eq::Range& range  = getRange();
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\label{fRender}Main Render Loop}}
+\end{wrapfigure}
 The rendering loop maintains a list of candidates to render, which
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\label{fRender}Main Render Loop}}
-\end{wrapfigure}
 The actual rendering uses display lists or vertex buffer objects. These
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 The purpose of the \textsf{OSWindow} processing method,
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\label{fEventProcessing}Event Processing}}
+\end{wrapfigure}
 \textsf{Window::processEvent} is responsible for either handling
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
+  {\caption{\label{fEventUML}UML Diagram for Event Handling Classes}}
+\end{wrapfigure}
 The concrete implementation of these window interfaces is responsible
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\label{fEventUML}UML Diagram for Event Handling Classes}}
-\end{wrapfigure}
 The glX event handler iterates over all windows of the registered pipe
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/threads.pdf}
+  {\caption{\label{fThreads}Threads within one node process}}
+\end{wrapfigure}
 During this initialization, Equalizer creates and manages two threads
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/threads.pdf}
-  {\caption{\label{fThreads}Threads within one node process}}
-\end{wrapfigure}
 The receiver and command thread are terminated when the application
@@
 
-\begin{figure}[ht!]\center
-  \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
-  {\caption{\label{fThreadModels}Async, draw\_sync and local\_sync
-    thread synchronization models}}
-\end{figure}
-
 \fig{fThreadModels} illustrates the synchronization and task execution
@@
 
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
+  {\caption{\label{fThreadModels}Async, draw\_sync and local\_sync
+    thread synchronization models}}
+\end{figure}
+
 \subsubsection{\label{sAdvDPlex}DPlex Compounds}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\label{fFrameSync}Per-node frame synchronization}}
+\end{wrapfigure}
 The application has extended control over the task synchronization
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
 The local synchronization consists of \textsf{releaseFrameLocal} to
@@
 \begin{enumerate}
-\item\textsf{getWGLPipeDC} is used to get an affinity device context,
-  which might be needed for window creation. If a context is returned, a
-  function pointer to delete the returned device context is set as
-  well. This function might return 0, which is not an error, but
-  signifies that no affinity context is used.
+\item\textsf{initWGLAffinityDC} is used to set up an affinity device
+  context, which might be needed for window creation. The WGL window
+  tracks potentially two device context handles, one for OpenGL context
+  creation (the affinity DC), and one for \textsf{swapBuffers} (the
+  window's DC).
 \item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
@@
 {
-    PFNEQDELETEDCPROC deleteDCProc = 0;
-    HDC dc = getWGLPipeDC( deleteDCProc );
-    EQASSERT( !dc || deleteDCProc );
-
-    int pixelFormat = chooseWGLPixelFormat( dc );
+    if( !initWGLAffinityDC( ))
+    {
+        _window->setErrorMessage( "Can't create affinity dc" );
+        return false;
+    }
+
+    int pixelFormat = chooseWGLPixelFormat();
     if( pixelFormat == 0 )
     {
-        if( dc )
-            deleteDCProc( dc );
+        exitWGLAffinityDC();
         return false;
@@
 
-    if( !configInitWGLDrawable( dc, pixelFormat ))
-    {
-        if( dc )
-            deleteDCProc( dc );
+    if( !configInitWGLDrawable( pixelFormat ))
+    {
+        exitWGLAffinityDC();
         return false;
@@
     {
-        if( dc )
-            deleteDCProc( dc );
+        exitWGLAffinityDC();
+        setWGLDC( 0, WGL_DC_NONE );
         _window->setErrorMessage(
@@
 
-    HGLRC context = createWGLContext( dc );
+    HGLRC context = createWGLContext();
+    if( !context )
+    {
+        configExit();
+        return false;
+    }
+
     setWGLContext( context );
+    makeCurrent();
+    _initGlew();
 
@@
 
-    if( !context )
-    {
-        if( dc )
-            deleteDCProc( dc );
+    if( !joinNVSwapBarrier( ))
+    {
+        _window->setErrorMessage( "Joining NV_swap_group failed" );
         return false;
@@
 
-    if( dc )
-        deleteDCProc( dc );
+    if( getIAttribute( Window::IATTR_HINT_DRAWABLE ) == FBO )
+        return configInitFBO();
+
     return true;
@@
 
-\subsection{\label{sView}View Management}
-
-A view describes the projection of the scene onto a coherent 2D area,
-typically a display or window rendered by a destination channel. The
-application has read and write access to all the views of a
-configuration.
-
-Changing view parameters is typically done for non-fullscreen
-application windows, where the rendering is not meant to be viewed in
-real-world size in an immersive environment. A typical use case is
-changing the field-of-view of the rendering.
-
-A configuration has one or more views. A view is determined by the
-frustum (wall or projection description) and eye separation. A view is
-uniquely assigned to one compound and possibly one destination
-channel. For a description of wall and projection parameters please
-refer to \sref{sFrustum}.
-
-The application node can query all views of the \textsf{eq::Config}. It
-can change all parameters of each view (wall, projection and eye
-separation). The new view parameters are applied for the next frame.
-
-Destination channels on the render clients have access to a read-only
-version of their view. The view data is automatically kept up-to-date
-with the application's instance by Equalizer.
-
-The view API is not finalized and may change in the future.
-
-\subsubsection{View Update}
+\subsection{\label{sLayoutAPI}Layout API}
+
+The Layout API provides an abstraction for render surfaces (Canvas and
+Segment) and the arrangment of rendering areas on them (Layout and
+View). Its functionality has been described in \sref{sCanvas} and
+\sref{sLayout}. This section focuses on how to use the Layout API
+programmatically.
+
+The application has read access to all canvases, segment, layouts and
+views of the configuration. The render client has access to the current
+view in the channel task methods. The layout entities can be subclassed
+using the \textsf{NodeFactory}. Currently the layout of a canvas and the
+frustum of a view can be changed at run-time.
+
+\subsubsection{Subclassing and Data Distribution}
+
+Layout API entities (Canvas, Segment, Layout, View) are subclassed like
+all other Equalizer entities using the \textsf{NodeFactory}. Equalizer
+registers the master instance of these entities on the application node,
+so that the application can change mutable parameters and distribute its
+own data to the render clients. \fig{fViewUML} shows the UML class
+hierarchy for the \textsf{eqPly::View}.
+
+The render clients can access a slave instance of the view using
+\textsf{Channel::getView}. When called from one of the frame task
+methods, this method will return the view of the current destination
+channel for which the task method is executed. Otherwise it returns the
+channel's native view, if it has one. Only destination channels of an
+active canvas have a native view.
+
+Equalizer commits dirty layout entities at the beginning of each
+\textsf{Config::startFrame}, and synchronizes the slave instances on the
+render clients correctly with the current frame.
+
+The most common entity to subclass is the \textsf{View}, since the
+application amends it by view-specific data. The view, like all other
+layout entities, is indirectly subclassed from \textsf{eq::Object} via
+\textsf{eq::Frustum}. The design and usage of \textsf{eq::Object} is
+explained in \sref{sEqObject}. In \textsf{eqPly}, the
+application-specific data is the model identifier. A new dirty bit for
+the model is defined, and the model identifier is serialized when
+requested:
+
+{\footnotesize\begin{lstlisting}
+/** The changed parts of the view. */
+enum DirtyBits
+{
+    DIRTY_MODEL       = eq::View::DIRTY_CUSTOM << 0,
+};
+
+void View::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
+{
+    eq::View::serialize( os, dirtyBits );
+    if( dirtyBits & DIRTY_MODEL )
+        os << _modelID;
+}
+
+void View::deserialize( eq::net::DataIStream& is, const uint64_t dirtyBits )
+{
+    eq::View::deserialize( is, dirtyBits );
+    if( dirtyBits & DIRTY_MODEL )
+        is >> _modelID;
+}
+
+void View::setModelID( const uint32_t id )
+{
+    _modelID = id;
+    setDirty( DIRTY_MODEL );
+}
+\end{lstlisting}}
+
+\subsubsection{Dynamic Layout Switch}
+
+The application can use a different layout on a canvas. This will cause
+the running entities to be updated on the next frame. At a minimum, this
+means the channels involved in the last layout on the canvas are
+deinitialized, that is \textsf{configExit} and
+\textsf{NodeFactory::releaseEntity} is called, and channels involved in
+the new layout are initialized. If a layout does not cover fully a
+canvas, the layout switch can also cause the (de-)initialization of
+windows, pipes and nodes.
+
+Due to the entity (de-)-initialization and the potential need to
+initialize view-specific data, e.g., a model, a layout switch is
+relatively expensive and will stall the application for about a second.
+
+Initializing an entity can fail. If that happens, the server will exit
+the whole configuration, and send an \textsf{EXIT} event to the
+application node. The exit event will cause the application to exit,
+since it resets the config's running state.
+
+\subsubsection{View Frustum Update}
+
+Changing view frustum parameters is typically done for non-fullscreen
+application windows and multi-view layouts, where the rendering is not
+meant to be viewed in real-world size in an immersive environment. A
+typical use case is changing the field-of-view of the rendering.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
+  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
+\end{wrapfigure}
+A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
+application process can set the wall or projection parameters at
+runtime.For a description of wall and projection parameters please refer
+to \sref{sFrustum}. The new data will be effective for the next frame.
+
+The frustum of a view overrides the underlying frustum of the
+segments. The frustum of a view does not typically correspond to a
+physical projection, and can therefore be changed at run-time. The
+frustum of a segment does however correspond to a physical projection,
+and is therefore immutable.
 
@@
 the aspect ratio of destination channels after a window resize. Without
-updating the wall or projection description, the rendering would become distorted.
-
-\begin{wrapfigure}{r}{.618\textwidth}
+updating the wall or projection description, the rendering would become
+distorted.
+
+\begin{wrapfigure}{l}{.618\textwidth}
   \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
@@
 
-The config event is dispatched to \textsf{Config::handleView\-Resize},
-which uses the view identifier to find the corresponding view. Based on
+The config event is dispatched to \textsf{View::handleEvent}. Based on
 the original size and wall or projection description of the view, a new
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\label{fDirectSend}Parallel Direct Send Compositing}}
+\end{wrapfigure}
 In direct send compositing, each rendering channel is also responsible
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\label{fDirectSend}Parallel Direct Send Compositing}}
-\end{wrapfigure}
 \fig{fDirectSend} illustrates this algorithm for three channels. The
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended to form a
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 Finding the correct assembly order for perspective frusta is more
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
-  {\caption{\label{fNetObject}Communication between two objects}}
-\end{wrapfigure}
 Distributed objects provide powerful, object-oriented data
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
+  {\caption{\label{fNetObject}Communication between two objects}}
+\end{wrapfigure}
 Distributed objects provide data synchronization functionality, as
@@
 
-\subsubsection{eq::Object}
+\subsubsection{\label{sEqObject}eq::Object}
 

### 2009-04-20 Developer/ProgrammingGuide/paper.tex
Programming Guide: update layout API
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     parallel rendering framework}\\\vspace{1cm}
-  Version 1.5.6 for Equalizer 0.6+svn\\[\medskipamount]
+  Version 1.5.7 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
     1.5.6   & Mar 23, 2009 & Documented Layout API\\
+    1.5.7   & Apr 17, 2009 & Updated Layout API (multiple observers)\\
     \hline
@@
 Canvases provide a convenient way to configure projection surfaces. A
-canvas uses a layout, which describes the logical views. The usage of
-canvases and layouts is optional. Typically, each desktop window uses
-one canvas.
+canvas uses layouts, which describe logical views. The usage of canvases
+and layouts is optional. Typically, each desktop window uses one canvas.
 
@@
 setup for Virtual Reality installations.
+
+A canvas has one or more layouts. One of the layouts is the active
+layout, that is, this set of views is currently used for rendering. It
+is possible to specify \textsf{OFF} as a layout, which deactivates the
+canvas. It is possible to use the same layout on different canvases.
 
@@
 
+A view might have an observer, in which case its frustum is tracked by
+this observer.
+
 \begin{figure}[ht!]\center
   \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
-  {\caption{A display wall using one canvas and layout with six segments
-    and two views.}}
+  {\caption{A display wall using a six-segment canvas with a two-view layout.}}
 \end{figure}
+
+\subsection{Observers}
+
+An observer represents an actor looking at multiple views. It has a head
+matrix, defining its position and orientation within the world, and an
+eye separation. Typically, a configuration has one
+observer. Configurations with multiple observers are used if multiple,
+head-tracked users are in the same configuration session, e.g., a
+non-tracked control host with two tracked head-mounted displays.
 
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
+  {\caption{\label{fModelDist}Scene Data in eqPly}}
+\end{wrapfigure}
 \textsf{eqPly} uses static distributed objects to distribute the model
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
-  {\caption{\label{fModelDist}Scene Data in eqPly}}
-\end{wrapfigure}
 The identifier of the model's root object of this distributed hierarchy
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-%\begin{figure}[ht!]\center
-  \vspace{-2ex}
+%\begin{wrapfigure}{r}{.618\textwidth}
+\begin{figure}[ht!]\center
   \includegraphics[width=.618\textwidth]{images/objects.pdf}
   {\caption{\label{fObjects}Scene Graph Distribution}}
-\end{wrapfigure}
+\end{figure}
 Applications distributing a dynamic scene graph use the frame data
@@
 
+%\begin{figure}[ht!]\center
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2ex}
+  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
+  {\caption{\label{fConfigInit}Config Initialization Sequence}}
+  \vspace{-4ex}
+\end{wrapfigure}
 The config initialization happens in parallel, that is, all config
@@
 
-%\begin{figure}[ht!]\center
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
-  {\caption{\label{fConfigInit}Config Initialization Sequence}}
-\end{wrapfigure}
 The tasks are executed by the node and pipe threads in parallel. The
@@
 
-For layout and model selection, \textsf{eqPly} maintains an active
-view. The identifier of the active view is stored in the frame data,
-which is used by the render client to highlight it using a different
-background color. The active view can be selected by clicking into a
-view, or by cycling through all the views using a keyboard shortcut:
-
-{\footnotesize\begin{lstlisting}
-bool Config::handleEvent( const eq::ConfigEvent* event )
-{
-    switch( event->data.type )
-    {
-        ...
-        case eq::Event::POINTER_BUTTON_PRESS:
-            _frameData.setCurrentViewID( event->data.context.view.id );
-            return true;
-        ...
-    }
-}
-
-bool Config::_handleKeyEvent( const eq::KeyEvent& event )
-{
-    switch( event.key )
-    {
-        ...
-        case 'v':
-        case 'V':
-        {
-            NextViewFinder finder( _frameData.getCurrentViewID( ));
-            accept( finder );
-            const eq::View* view = finder.getResult();
-            if( view )
-                _frameData.setCurrentViewID( view->getID( ));
-            else
-                _frameData.setCurrentViewID( EQ_ID_INVALID );
-            return true;
-        }
-        ...
-    }
-}
-\end{lstlisting}}
+For layout and model selection, \textsf{eqPly} maintains an active view
+and canvas. The identifier of the active view is stored in the frame
+data, which is used by the render client to highlight it using a
+different background color. The active view can be selected by clicking
+into a view, or by cycling through all the views using a keyboard
+shortcut.
 
@@
 for each view is stored on the view, which is used to retrieve the model
-on the render clients. View-specific data is not limited to a
-model. Applications can choose to make any application-specific data
-view-specific, e.g., cameras, rendering modes or annotations. A view is
-a generic concept for an application-specific view on data,
-\textsf{eqPly} is using different models to illustrate the concept:
+on the render clients.
+
+View-specific data is not limited to a model. Applications can choose to
+make any application-specific data view-specific, e.g., cameras,
+rendering modes or annotations. A view is a generic concept for an
+application-specific view on data, \textsf{eqPly} is using different
+models to illustrate the concept:
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-8ex}\includegraphics[width=.382\textwidth]{images/DB.png}
-  {\caption{\label{fDBDest}Destination view of a DB compound}}
-\end{wrapfigure}
 After the basic view setup, a directional light is configured, and the
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/DB.png}
+  {\caption{\label{fDBDest}Destination view of a DB compound}}
+\end{wrapfigure}
 To draw the model, a helper class for view frustum culling is set up
@@
 
+An Equalizer configuration has a vector of observers. Each observer has
+its own head matrix and eye separation for tracking. In \textsf{eqPly},
+all observers are updated with the same head matrix. A typical
+application would assign different tracking data to different observers.
+
 \begin{figure}[h!t]
@@
 initialize view-specific data, e.g., a model, a layout switch is
-relatively expensive and will stall the application for about a second.
+relatively expensive and will stall \textsf{eqPly} for about a second.
 

### 2009-05-30 Developer/ProgrammingGuide/paper.tex
PG: Update documentation related Equalizers.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     parallel rendering framework}\\\vspace{1cm}
-  Version 1.5.7 for Equalizer 0.6+svn\\[\medskipamount]
+  Version 1.5.8 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
 \tableofcontents
-\listoffigures
-\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
+\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
     \hline
@@
     1.5.7   & Apr 17, 2009 & Updated Layout API (multiple observers)\\
+    1.5.8   & May 30, 2009 & Added \sref{sEqualizers}, updated \sref{sLoadBalancing} and \sref{sSwapBarrier}\\
     \hline
@@
 
-\subsection{\label{sLoadBalancing}Load-Balancing}
-
-While pixel and stereo compounds are naturally load-balanced, 2D, DB and
-DPlex compounds might need load-balancing to provide optimal
-performance. Using a load-balancer is transparent to the application,
-and can be used on any application for 2D and DPlex, and on most
-applications for DB load-balancing. Some applications supporting DB
-decomposition do not support dynamic updates of the database range, and
-therefore cannot be used with DB load-balancing.
-
-Using a 2D or DB load-balancer will adapt the 2D split or database range
-automatically each frame. The 2D load-balancer exists in three flavors:
-\textsf{2D} using tiles, \textsf{VERTICAL} using columns and
-\textsf{HORIZONTAL} using rows. The current load-balancing algorithm for
-2D and DB is a technology preview and will improve in future versions.
-
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/lb.pdf}
-  {\caption{\label{fLoadBalancing}\small 2D load-balancing}}
-\end{wrapfigure}
-2D load-balancing increases the framerate over a static decomposition in
-virtually all cases. It works best if the application data is relatively
-uniformly distributed in screen-space.
-
-DB load-balancing is beneficial for applications which cannot precisely
-predict the load for their scene data. Volume rendering is a
-counterexample, where the data is uniform and a static DB decomposition
-typically results in a better performance.
-
-DPlex compounds require a smoothing of the framerate at the destination
-channel, otherwise the framerate will become periodically faster and
-slower. Using a DPlex load-balancer will smoothen the swap buffer rate
-on the destination window for optimal user experience.
+\subsection{\label{sLoadBalancing}Automatic Runtime Adjustments}
+
+Some scalable rendering parameters have to be updated at run-time to
+provide optimal performance or influence other scalable rendering
+features. These adjustments are often referred to as load-balancing, but
+are called \textsf{equalizers}, since their functionality is not limited
+to load-balancing in Equalizer.
+
+The server provides a number of equalizers, which automatically update
+certain parameters of compounds based on run-time information. They
+balance the load of 2D and DB compounds, optimally distribute render
+resources for multi-view configurations, adapt the resolution to provide
+a constant framerate or zoom images to allow monitoring of another view.
+
+Equalizer are described in more detail in \sref{sEqualizers}.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/model.pdf}
+  {\caption{\label{fModel}Simplified execution model}}
+\end{wrapfigure}
 The main thread is responsible for maintaining the application logic. It
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/model.pdf}
-  {\caption{\label{fModel}Simplified execution model}}
-\end{wrapfigure}
 All rendering threads in the configuration run asynchronously to the
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/layout.png}
+  {\caption{\label{fLayout}A layout with four views}}
+  \vspace{-4ex}
+\end{wrapfigure}
 A layout is the grouping of logical views. It is used by one or more
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/layout.png}
-  {\caption{\label{fLayout}A layout with four views}}
-\end{wrapfigure}
 A view is a logical view of the application data, in the sense used by
@@
   \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
-  {\caption{A display wall using a six-segment canvas with a two-view layout.}}
+  {\caption{A display wall using a six-segment canvas with a two-view layout}}
 \end{figure}
@@
 
-\subsubsection{Synchronization - Swapbarrier}
-Compounds may have a swapbarrier to synchronize the buffer swap of
-multiple channels. Windows with a swapbarrier of the same name enter a
-barrier before executing the swap buffer task. Before entering the
-barrier, \textsf{Window::finish} is called to ensure that all OpenGL
-commands have been executed.
-
-Note that this software swap barrier is not accurate enough for
+\subsubsection{\label{sSwapBarrier}Synchronization - Swapbarriers}
+Compounds may have a software or hardware swapbarrier to synchronize the
+buffer swap of multiple channels. A software swapbarrier is configured
+by giving it a name. Windows with a swapbarrier of the same name
+synchronize with each other before executing the swap buffer
+task. Before entering the barrier, \textsf{Window::finish} is called to
+ensure that all OpenGL commands have been executed.
+
+Note that a software swap barrier is not accurate enough for
 edge-blended projection system. Such installations typically use a
@@
 
+A hardware swapbarrier is configured by setting the \textsf{NV\_group}
+and \textsf{NV\_barrier} parameters. These parameters follow the
+\textsf{NV\_swap\_group} extension, which synchronizes all windows bound
+to the same group on a single machine, and all groups bound to the same
+barrier across systems.
+
+\subsubsection{\label{sEqualizers}Adaptation - Equalizers}
+
+Equalizers are used to update compound parameters based on run-time
+data. They are attached to a compound (sub-)tree, on which they
+operate. The Equalizer distribution contains numerous example
+configuration files using equalizers.
+
+\paragraph{Load Equalizer}
+\vspace{-2ex}\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/lb.pdf}
+  {\caption{\label{fLoadBalancing}\small 2D load-balancing}}
+\end{wrapfigure}
+While pixel and stereo compounds are naturally load-balanced, 2D and DB
+compounds often need load-balancing for optimal rendering
+performance. Using a load equalizer is transparent to the application,
+and can be used on any application for 2D, and on most applications for
+DB load-balancing. Some applications do not support dynamic updates of
+the database range, and therefore cannot be used with DB load-balancing.
+
+Using a 2D or DB load-balancer will adapt the 2D split or database range
+automatically each frame. The 2D load-balancer exists in three flavors:
+\textsf{2D} using tiles, \textsf{VERTICAL} using columns and
+\textsf{HORIZONTAL} using rows.
+
+2D load-balancing increases the framerate over a static decomposition in
+virtually all cases. It works best if the application data is relatively
+uniformly distributed in screen-space. A damping parameter can be used
+to fine-tune the algorithm.
+
+DB load-balancing is beneficial for applications which cannot precisely
+predict the load for their scene data. Volume rendering is a
+counterexample, where the data is uniform and a static DB decomposition
+typically results in a better performance.
+
+\paragraph{View Equalizer}
+Depending on the model position and data structure, each segment of a
+multi-display system has a different rendering load. The segment with
+the biggest load determines the overall performance when using a static
+assignment of resources to segments. The view equalizer analyzes the
+load of all segments, and adapts the resource usage each frame. It
+equalizes the load on all segments of a view.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
+  {\caption{\label{fViewLoadBalancing}\small Cross-segment
+      load-balancing for two segments using eight GPU's}}
+\end{wrapfigure}
+\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
+  AG, 2008} illustrates this process. On the left side, a static
+assigment of resources to display segments is used. The right-hand
+segment has a higher load than the left-hand segment, causing
+sub-optimal performance. The configuration on the left uses a view
+equalizer, which assigns two GPU's to the left segment and four GPU's to
+the right segment, which leads to optimal performance for this model and
+camera position.
+
+The view equalizer can also use resources from another display resource,
+if this resource has little rendering load by itself. It is therefore
+possible to improve the rendering performance of a multi-display system
+without any additional resources. This is particulary useful for
+installations with a higher number of displays where the rendering load
+is typically in a few segments only, e.g., for a CAVE.
+
+\begin{wrapfigure}{l}{.382\textwidth} 
+    \includegraphics[width=.382\textwidth]{images/viewLB.png}
+    {\caption{\label{fViewLoadBalancing2}\small Cross-segment
+        load-balancing for a CAVE}}
+\end{wrapfigure}
+\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
+by five GPU's. The front and left segments show the model and have a
+significant rendering load. The view equalizer assigns the GPU's from
+the top, bottom and right wall for rendering the left and front wall in
+this configuration.
+
+Cross-segment load-balan\-cing is configured hierarchically. On the top
+compound level, a view equalizer assigns resources to each of its
+children, so that the optimal number of resources is used for each
+segment. On the next level, a load equalizer on each child computes the
+resource distribution within the segment, taking the resource usage
+given by the view equalizer into account.
+
+\paragraph{Framerate Equalizer}
+Certain configurations, in particular DPlex compounds, require a
+smoothing of the framerate at the destination channel, otherwise the
+framerate will become periodically faster and slower. Using a framerate
+equalizer will smoothen the swap buffer rate on the destination window
+for optimal user experience.
+
+\paragraph{DFR Equalizer}
+Dynamic Frame Resolution (DFR) trades rendering performance for visual
+quality. The rendering for a channel is done at a different resolution
+than the native channel resolution to keep the framerate constant. The
+DFR equalizer adapts the zoom of a channel, based on the target and
+current framerate. It is typically used for fill-rate bound
+applications, such as volume rendering and raytracing.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
+  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
+\end{wrapfigure}
+\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
+  Science Institute, University of Freiburg, Germany} shows DFR for
+volume rendering. To achieve 10 frames per second, the model is rendered
+at a lower resolution, and upscaled to the native resolution for
+display. The rendering quality is slightly degraded, while the rendering
+performance remains interactive. When the application is idle, it can
+render a full-resolution view.
+
+The dynamic frame resolution is not limited to down-scaling the
+rendering resolution, it will also up-scale the resolution if the
+parameters allow for it. Upscaled rendering, which will downsample the
+result for display, provides dynamic anti-aliasing at a constant
+framerate.
+
+\paragraph{Monitor Equalizer}
+The monitor equalizer allows the observation of another view,
+potentially made of multiple segments, in a different channel at a
+different resolution. This is typically used to re-use the rendering of
+a large-scale display on an operator station.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/monitorEq.pdf}
+  {\caption{\label{fMonitorEq}\small Monitoring a projection wall}}
+\end{wrapfigure}
+The monitor equalizer adapts the frame zoom of the output frames used to
+observe the rendering, depending on the destination channel size. The
+output frames are downscaled on the GPU before readback, which results
+in optimal performance.
+
+\fig{fMonitorEq} shows a usage of the monitor equalizer. A two-segment
+display wall is driven by a separate control station. The rendering
+happens only on the display wall, and the control window receives the
+correctly downscaled version of the rendering.
+
 \section{\label{sEqPly}The eqPly polygonal renderer}
@@
   \includegraphics[width=.9\textwidth]{images/uml}
-  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes.}}
+  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes}}
 \end{figure}

### 2009-06-11 Developer/ProgrammingGuide/paper.tex
Rework scalability section
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 DB load-balancing is beneficial for applications which cannot precisely
-predict the load for their scene data. Volume rendering is a
-counterexample, where the data is uniform and a static DB decomposition
-typically results in a better performance.
+predict the load for their scene data, e.g., when the data is
+non-uniform. Volume rendering is a counterexample, where the data is
+uniform and a static DB decomposition typically results in a better
+performance.
 
@@
 display. The rendering quality is slightly degraded, while the rendering
-performance remains interactive. When the application is idle, it can
-render a full-resolution view.
+performance remains interactive. When the application is idle, it
+renders a full-resolution view.
 

### 2009-07-27 Developer/ProgrammingGuide/paper.tex
Programming Guide review pass
[ ] Will break Win32 build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
     parallel rendering framework}\\\vspace{1cm}
-  Version 1.5.8 for Equalizer 0.6+svn\\[\medskipamount]
+  Version 1.5.9 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
     1.5.8   & May 30, 2009 & Added \sref{sEqualizers}, updated \sref{sLoadBalancing} and \sref{sSwapBarrier}\\
+    1.5.9.  & Jul 27, 2009 & Major review pass before release\\
     \hline
@@
 
-Some scalable rendering parameters have to be updated at run-time to
-provide optimal performance or influence other scalable rendering
+Some scalable rendering parameters are updated at run-time to provide
+optimal performance or influence other scalable rendering
 features. These adjustments are often referred to as load-balancing, but
@@
 balance the load of 2D and DB compounds, optimally distribute render
-resources for multi-view configurations, adapt the resolution to provide
-a constant framerate or zoom images to allow monitoring of another view.
+resources for multi-segment projections systems, adapt the resolution to
+provide a constant framerate or zoom images to allow monitoring of
+another view.
 
@@
 
-\section{The Programming Interface}
-
-Equalizer uses a C++ programming interface. The API is minimally
-invasive, so Equalizer imposes only a minimal, natural execution
-framework upon the application. It does not provide a scene graph, or
-interfere in any other way with the application's rendering code. The
-restructuring work required for Equalizer is the minimal refactoring
-needed to parallelize the application for rendering.
-
-Methods called by the application have the form \textsf{verb[Noun]},
-whereas methods called by Equalizer (`Task Methods') have the form
-\textsf{nounVerb}. For example, the application calls
-\textsf{Config::startFrame} to render a new frame, which causes --among
-other things-- \textsf{Node::frameStart} to be called in all active
-render clients.
-
-\subsection{\label{sTaskMethods}Task Methods}
-
-The application inherits from Equalizer classes and overrides virtual
-functions to implement certain functionality, e.g., the application's
-OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
-methods are similar in concept to C function callbacks. \sref{sEqPly}
-will discuss the important task methods. A full list can be found on the
-website\footnote{see
-  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.
-
-\subsection{\label{sExecution}Execution Model and Thread Safety}
-
-Using threading correctly in OpenGL-based applications is easy with
-Equalizer. Equalizer creates one rendering thread for each graphics
-card. All task methods for a pipe, and therefore all OpenGL commands,
-are executed from this thread. This threading model is the OpenGL
-`threading model', which maintains a current context for each
-thread. If structured correctly, the application rarely has to take care
-of thread synchronization or protection of shared data.
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/model.pdf}
-  {\caption{\label{fModel}Simplified execution model}}
-\end{wrapfigure}
-The main thread is responsible for maintaining the application logic. It
-reacts on user events, updates the data model and requests new frames to
-be rendered. It drives the whole application, as shown in \fig{fModel}.
-
-The rendering threads concurrently render the application's
-database. The data\-base should be accessed in a read-only fashion
-during rendering to avoid threading problems. This is normally the case,
-for example all modern scene graphs use read-only render traversals.
-
-All rendering threads in the configuration run asynchronously to the
-application's main thread. Depending on the configuration's latency,
-they can fall $n$ frames behind the last frame finished by the
-application thread. A latency of one frame is usually not perceived by
-the user, but can increase rendering performance substantially.
-
-Rendering threads on a single node are synchronized when using the
-default thread model \textsf{draw\_sync}. When a frame is finished, all
-local rendering threads are done drawing. Therefore the application can
-safely modify the data between the end of a frame and the beginning of a
-new frame. Furthermore, only one instance of the scene data has to be
-maintained within a process, since all rendering threads are guaranteed
-to draw the same frame.
-
-This per-node frame synchronization does not inhibit latency across
-rendering nodes. Furthermore, advanced rendering software which
-multi-buffers the dynamic parts of the database can disable the per-node
-frame synchronization, as explained in \sref{sThreadModel}. Some scene
-graphs implement multi-buffered data, and can profit from relaxing the
-frame synchronization.
-
-\subsection{\label{sConfig}Config}
-
-The \textsf{eq::Config} represents the current configuration of the
-application. The configuration is the session in which all render
-clients are registered. A configuration consists of the description of
-the rendering resources and the usage description for these resources.
+\section{\label{sConfig}Writing Configurations}
+
+\subsection{Preparation}
+
+Before writing a configuration, it is useful to assemble the following
+information:
+
+\begin{itemize}
+\item A list of all \textbf{computers} in you rendering cluster,
+  including the \textbf{IP addresses} of all network interfaces to be
+  used.
+\item The number of \textbf{graphics cards} in each computer.
+\item The \textbf{physical dimensions} of the display system, if
+  applicable. These are typically the bottom-left, bottom-right and
+  top-left corner points of each display surfaces in meters.
+\item The relative coordinates of all the \textbf{segments} belonging to
+  each display surface, and the graphics card \textbf{output} used for
+  each segment. For homogenous setups, it is often enough to know the
+  number of rows and columns on each surface, as well as the overlap or
+  underlap percentage, if applicable.
+\item The number of desired \textbf{application windows}. Application
+  windows are typically destination windows for scalable rendering or
+  windows paired with a view on a display system.
+\item \textbf{Characteristics} of the application, e.g., supported
+  scalability modes and features.
+\end{itemize}
+
+\subsection{Overview}
+
+Equalizer applications are configured at run-time by the Equalizer
+server. The server loads its configuration from a text file, which is an
+one-to-one representation of the configuration data structures at
+runtime.
+
+For an extensive documentation of the file format please refer to
+\aref{aFileFormat}. This section gives an introduction on how to write
+configuration files.
+
+A configuration consists of the declaration of the rendering resources
+and the usage description for these resources. It can optionally contain
+the description of the physical layout of the projection system and
+logical layouts on the projection canvases.
 
@@
 which corresponds to the physical and logical resources found in a 3D
-rendering environment.
+rendering environment: nodes (computers), pipes (graphics cards),
+windows, channels.
 
@@
 recomposition across the resources. It is explained in \sref{sCompounds}.
+
+Physical layouts of display systems are configured using canvases with
+segments, which represent 2D rendering areas composed of multiplex
+displays or projectors. Logical layouts are applied to canvases and
+define views on a canvas.
 
@@
 
-\subsubsection{Node}
+\subsection{Node}
+
+For each machine in your cluster, create one \textsf{node}. Create one
+\textsf{appNode} if your application process should also render, which
+is often the case. List all nodes, even if you are not planning to use
+them at first. Equalizer will only instantiate and access used nodes,
+that is, nodes which are referenced by an active compound.
+
+In each node, list all \textsf{connection}s thorugh which this node is
+reachable. Typically a node uses only one connection, but it is possible
+to configure multiple connections if the machine and cluster is set up
+to use multiple, independent network interfaces. Make sure the
+\textsf{hostname} used is reachable from all nodes. An IP address may be
+used as the hostname.
 
@@
 
-\subsubsection{Pipe}
+\subsection{Pipe}
+
+For each node, create on \textsf{pipe} for each graphics card in the
+machine. Set the \textsf{device} number to the correct index. On
+operating systems using X11, e.g., Linux, also set the port number if
+your X-Server is running on a non-standard port.
 
@@
 
-\subsubsection{Window}
+\subsection{Window}
+
+Configure one \textsf{window} for each desired application window on the
+appNode. Configure one full-screen window for each display
+segment. Configure one off-screen window, typically a pbuffer, for each
+graphics card used as a source for scalable rendering.
+
+Sometimes display segments cover only a part of the graphics card
+output. In this case it is adviced to configure a non-fullscreen window
+without window decorations, using the correct window viewport.
 
@@
 context. The drawable can be an on-screen window or an off-screen
-PBuffer. FrameBuffer object (FBO) support will be implement by a later
-version of Equalizer. The window uses an \textsf{eq::OSWindow}, which
-abstracts and manages window-system-specific handles to the drawable and
-context, e.g., an X11 window \textsf{XID} and \textsf{GLXContext} for
-the glX window system.
-
-\subsubsection{Channel}
+PBuffer or FrameBuffer Object (FBO). The window uses an
+\textsf{eq::OSWindow}, which abstracts and manages
+window-system-specific handles to the drawable and context, e.g., an X11
+window \textsf{XID} and \textsf{GLXContext} for the glX window system.
+
+\subsection{Channel}
+
+Configure one \textsf{channel} for each desired rendering in each window
+area. Typically one full-screen channel per window is used. Name the
+channel using a unique, easily-identifiable name.
+
+Multiple channels in application windows may be used to view the model
+from different viewports.
+
+Sometimes, a single window is split across multiple projectors, e.g., by
+using an external splitter such as the Matrox triple-head-to-go. In this
+case configure one channel for each segment, using the channel's
+\textsf{viewport} to configure its position wit respect to the window.
 
@@
 \subsection{\label{sCanvas}Canvases}
+
+If you are writing a configuration for workstation usage you can skip
+the following sections and restart with \sref{sCompounds}.
+
+Configure one \textsf{canvas} for each display surface. For planar
+surfaces, e.g., a display wall, configure a frustum. For non-planar
+surfaces, the frustum will be configured on each display segment.
+
+The frustum can be specified as a wall or projection description. Take
+care to choose your reference system for describing the frustum to be
+the same system as used by the head-tracking matrix calculated by the
+application.
+
+A wall is completely defined by the bottom-left, bottom-right and
+top-left coordinates relative to the origin.
+
+A projection is defined by the position and head-pitch-roll orientation
+of the projector, as well as the horizontal and vertical field-of-view
+and distance of the projection wall.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
+  {\caption{\label{fFrusta}Wall and projection parameters}}
+\end{wrapfigure}
+
+\fig{fFrusta} illustrates the wall and projection frustum parameters.
 
@@
 \end{wrapfigure}
+
+Configure one \textsf{segment} for each display or projector of each
+canvas. Configure the \textsf{viewport} of the segment to match the area
+covered by the segment on the physical canvas. Set the output
+\textsf{channel} to the resource driving the described projector.
+
+For non-planar displays, configure the frustum as described in
+\sref{sCanvas}. For passive stereo installations, configure one segment
+per eye pass, where the segment for the left and right eye fully
+have the same viewport.
+
 A segment represents one output channel of the canvas, e.g., a projector
@@
 
-Passive stereo installations have one segment per eye pass, where the
-segment for the left and right eye fully overlap.
-
 \subsection{\label{sLayout}Layouts}
@@
   {\caption{\label{fLayout}A layout with four views}}
-  \vspace{-4ex}
 \end{wrapfigure}
+
+Configure one \textsf{layout} for each configuration of logical
+views. Name the layout using a unique name. Often only one layout with a
+one view is used for all canvases.
+
+Enable the layout on each desired canvas by adding it to the
+canvas. Since canvases reference layouts by name or index, layouts have
+to be configured before their respective canvases in the configuration
+file.
+
 A layout is the grouping of logical views. It is used by one or more
-canvases. For all given layouts, Equalizer creates destination channels
-when the configuration file is loaded. These destination channels can be
-referenced by compounds to configure scalable rendering.
+canvases. For all given layout/canvas combinations, Equalizer creates
+destination channels when the configuration file is loaded. These
+destination channels can be referenced by compounds to configure
+scalable rendering.
+
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
+  {\caption{A display wall using a six-segment canvas with a two-view layout}}
+\end{figure}
 
@@
 \subsubsection{Views}
+
+Configure one \textsf{view} for each logical view in each layout. Set
+the \textsf{viewport} to position the view.
 
@@
 
-\begin{figure}[ht!]\center
-  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
-  {\caption{A display wall using a six-segment canvas with a two-view layout}}
-\end{figure}
-
 \subsection{Observers}
+
+Unless you have multiple tracked persons, or want to disable tracking on
+certain views, you can skip this section.
+
+Configure one \textsf{observer} for each tracked person in the
+configuration. Most configurations have at most one observer. Assign the
+observer to all views which belong to this observer. Since the observer
+is referenced by its name or index, it has to be specified before the
+layout in the configuration file.
+
+Views with no observer are not tracked. The config file loader will
+create one default observer and assign it to all views if the
+configuration has no observer.
 
@@
 \subsection{\label{sCompounds}Compounds}
+
+Compound trees are used to describe how the rendering resources are
+combined to produce the desired output, and how they are aggregated to
+increase the performance. It is adviced to study and understand the
+basic configuration files shipped with the Equalizer configuration,
+before attempting to write compound configurations.
+
+For configurations using canvases and layouts without scalability, the
+configuration file loader will create the appropriate compounds.
 
@@
 \subsubsection{\label{sFrustum}Frustum}
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
-  {\caption{\label{fFrusta}Wall and projection parameters}}
-\end{wrapfigure}
+
 Compounds have a frustum description to define the physical layout of
@@
 
-The frustum can be specified as a wall or projection description.
-
-A wall is completely defined by the bottom-left, bottom-right and
-top-left coordinates relative to the origin.
-
-The projection is defined by the position and head-pitch-roll
-orientation of the projector, as well as the horizontal and vertical
-field-of-view and distance of the projection wall.
-
-\fig{fFrusta} illustrates the wall and projection frustum parameters.
+The frustum can be specified as described in \sref{sCanvas}.
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
-  {\caption{\label{fViewLoadBalancing}\small Cross-segment
-      load-balancing for two segments using eight GPU's}}
-\end{wrapfigure}
 \fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
+  {\caption{\label{fViewLoadBalancing}\small Cross-segment
+      load-balancing for two segments using eight GPU's}}
+\end{wrapfigure}
 The view equalizer can also use resources from another display resource,
@@
 \paragraph{DFR Equalizer}
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
+  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
+\end{wrapfigure}
 Dynamic Frame Resolution (DFR) trades rendering performance for visual
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
-  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
-\end{wrapfigure}
 \fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
@@
 
+\section{The Programming Interface}
+
+Equalizer uses a C++ programming interface. The API is minimally
+invasive, so Equalizer imposes only a minimal, natural execution
+framework upon the application. It does not provide a scene graph, or
+interferes in any other way with the application's rendering code. The
+restructuring work enforced by Equalizer is the minimal refactoring
+needed to parallelize the application for rendering.
+
+Methods called by the application have the form \textsf{verb[Noun]},
+whereas methods called by Equalizer (`Task Methods') have the form
+\textsf{nounVerb}. For example, the application calls
+\textsf{Config::startFrame} to render a new frame, which causes --among
+other things-- \textsf{Node::frameStart} to be called in all active
+render clients.
+
+\subsection{\label{sTaskMethods}Task Methods}
+
+The application inherits from Equalizer classes and overrides virtual
+functions to implement certain functionality, e.g., the application's
+OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
+methods are similar in concept to C function callbacks. \sref{sEqPly}
+will discuss the important task methods. A full list can be found on the
+website\footnote{see
+  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.
+
+\subsection{\label{sExecution}Execution Model and Thread Safety}
+
+Using threading correctly in OpenGL-based applications is easy with
+Equalizer. Equalizer creates one rendering thread for each graphics
+card. All task methods for a pipe, and therefore all OpenGL commands,
+are executed from this thread. This threading model is the OpenGL
+`threading model', which maintains a current context for each
+thread. If structured correctly, the application rarely has to take care
+of thread synchronization or protection of shared data.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/model.pdf}
+  {\caption{\label{fModel}Simplified execution model}}
+\end{wrapfigure}
+The main thread is responsible for maintaining the application logic. It
+reacts on user events, updates the data model and requests new frames to
+be rendered. It drives the whole application, as shown in \fig{fModel}.
+
+The rendering threads concurrently render the application's
+database. The data\-base should be accessed in a read-only fashion
+during rendering to avoid threading problems. This is normally the case,
+for example all modern scene graphs use read-only render traversals.
+
+All rendering threads in the configuration run asynchronously to the
+application's main thread. Depending on the configuration's latency,
+they can fall $n$ frames behind the last frame finished by the
+application thread. A latency of one frame is usually not perceived by
+the user, but can increase rendering performance substantially since
+operations are pipelined better.
+
+Rendering threads on a single node are synchronized when using the
+default thread model \textsf{draw\_sync}. When a frame is finished, all
+local rendering threads are done drawing. Therefore the application can
+safely modify the data between the end of a frame and the beginning of a
+new frame. Furthermore, only one instance of the scene data has to be
+maintained within a process, since all rendering threads are guaranteed
+to draw the same frame.
+
+This per-node frame synchronization does not inhibit latency across
+rendering nodes. Furthermore, advanced rendering software which
+multi-buffers the dynamic parts of the database can disable the per-node
+frame synchronization, as explained in \sref{sThreadModel}. Some scene
+graphs implement multi-buffered data, and can profit from relaxing the
+frame synchronization.
+
 \section{\label{sEqPly}The eqPly polygonal renderer}
@@
 detail, and relevant design decisions, caveats and other remarks are
-raised.
+discussed.
 
@@
 The distributed data classes are helper classes based on
-\textsf{eq::net::Object}. They illustrate the typical usage of distributed
-objects for static as well as dynamic, frame-specific data. Furthermore
-they are used for a basic scene graph distribution of the model data.
+\textsf{eq::Object} or its parent class \textsf{eq::net::Object}. They
+illustrate the typical usage of distributed objects for static as well
+as dynamic, frame-specific data. Furthermore they are used for a basic
+scene graph distribution of the model data.
 
@@
 
-The second step is to initialize the Equalizer library. The
+The second step is to redirect all Equalizer output to a file if this
+was requested through the command line. This should happen before
+initializing Equalizer, so that output during initialization is
+redirected and threads launched by Equalizer will use the given log
+stream. Any \textsf{std::ostream} can be used as log output:
+
+{\footnotesize\begin{lstlisting}
+    // 2. Redirect log output to file, if requested
+    std::ofstream* logFile = 0;
+    const std::string& logFilename = initData.getLogFilename();
+    if( !logFilename.empty( ))
+    {
+        logFile = new ofstream( logFilename.c_str( ));
+        eq::base::Log::setOutput( *logFile );
+    }
+\end{lstlisting}}
+
+The third step is to initialize the Equalizer library. The
 initialization function of Equalizer also parses the command line, which
@@
 
-The third step is to create an instance of the application and to
+The fourth step is to create an instance of the application and to
 initialize it locally. The application is a subclass of
 \textsf{eq::Client}, which in turn is an \textsf{eq::net::Node}. The
-underlying network layer in Equalizer is a peer-to-peer network of
-\textsf{eq::net::Node}s. The client/server concept is implement the
-higher-level \textsf{eq} client namespace.
-
-The application programmer does not usually have to be aware of the
-classes in the \textsf{eq::net} namespace, but both the
-\textsf{eq::Client} and the Equalizer server are
-\textsf{eq::net::Node}s.
+underlying Equalizer network layer, discussed in \sref{sNetwork}, is a
+peer-to-peer network of \textsf{eq::net::Node}s. The client/server
+concept is implement the higher-level \textsf{eq} client namespace.
 
@@
 {\footnotesize\begin{lstlisting}
-    // 3. initialization of local client node
+    // 4. initialization of local client node
     RefPtr< eqPly::Application > client = new eqPly::Application( initData );
@@
 {\footnotesize\begin{lstlisting}
-    // 4. run client
+    // 5. run client
     const int ret = client->run();
@@
 {\footnotesize\begin{lstlisting}
-    // 5. cleanup and exit
+    // 6. cleanup and exit
     client->exitLocal();
@@
     eq::exit();
+
+    if( logFile )
+        logFile->close();
+    delete logFile;
+
     return ret;
@@
   is selected using the command-line option \textsf{--eq-client} and
-  potentially \textsf{--eq-listen \textless address\textgreater} and
-  \textsf{-r}\footnote{see
+  \textsf{--eq-listen \textless address\textgreater} to specify a
+  well-defined listening address, and potentially \textsf{-r} to keep
+  the client running across multiple runs\footnote{see
     \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
@@
 \textsf{Client::connectServer} tries first to connect to a server on the
-local machine using the default port. If that fails, it will create
-a server running within the application process with a default
-one-channel configuration\footnote{see
+local machine using the default port. If that fails, it will create a
+server running within the application process with a default one-channel
+configuration\footnote{see
   \link{http://www.equalizergraphics.com/documents/design/standalone.html}}.
+
+Another configuration can be selected using the command-line parameter
+\textsf{--eq-config}. This configuration may only contain an
+\textsf{appNode}, i.e., no distributed configuration is possibly since
+the process-local server does not create a listening socket.
 
@@
 
-When the frame is started, the server computes all rendering tasks and
+When a new frame is started, the server computes all rendering tasks and
 sends them to the appropriate render client nodes. The render client
@@
 
-\textsf{Config::finishFrame} synchronizes on the completion of the frame
+\textsf{Config::finishFrame} blocks on the completion of the frame
 \textsf{current - latency}. The latency is specified in the
 configuration file, and allows several outstanding frames. This allows
-overlapping execution in the node and pipe threads and minimizes idle
-times.
+overlapping execution in the node processes and pipe threads and
+minimizes idle times.
 
@@
 not have to be multi-buffered. Applications such as \textsf{eqPly}, which
-do not need this per-node frame synchronization can disable it, as
+do not need this per-node frame synchronization, can disable it as
 explained in \sref{sThreadModel}.
@@
 imbalances in the load distribution, resulting in an improved
-framerate. With \textsf{eqPly}, a speedup of 15\% has been observed on a
+framerate. For example, we have observed a speedup of 15\% on a
 five-node rendering cluster when using a latency of one frame instead of
@@
 
-If the application detects that it is currently idle, all pending frames
-are flushed. Then it waits and handles one event at a time, until a
-redraw is needed:
+If the application detects that it is currently idle, all pending
+commands are gradually flushed, while still looking for a redraw
+event. Then it waits and handles one event at a time, until a redraw is
+needed:
 
@@
 
-        if( !config->needsRedraw( ))
-            config->finishAllFrames(); // flush, TODO flush one task at a time
-
         while( !config->needsRedraw( )) // wait for an event requiring redraw
+        {
+            if( hasCommands( )) // execute non-critical pending commands
+            {
+                processCommand();
+                config->handleEvents(); // non-blocking
+            }
+            else  // no pending commands, block on user event
             {
@@
             }
+        }
+        config->handleEvents(); // process all pending events
     }
@@
 
-Equalizer provides distributed objects which help implementing data
-distribution in a cluster environment. The master version of a
-distributed object is registered with a \textsf{eq::net::Session}, which
-assigns a session-unique identifier to the object. This identifier can
-be used by other nodes to map their instance of the object, thus
-synchronizing the object's data with the remotely registered master
-version.
-
-Distributed objects are created by subclassing from \textsf{eq::Object}
-or \textsf{eq::net::Object}. Distributed objects can be static
-(immutable) or dynamic. Dynamic objects are versioned.
+Equalizer provides distributed objects which facilitate the
+implementation of data distribution in a cluster
+environment. Distributed objects are created by subclassing from
+\textsf{eq::Object} or \textsf{eq::net::Object}. The application
+programmer implements serialization and deserialization of the
+distributed data.
+
+Objects are dynamically assigned to a session, which assigns
+session-unique identifiers to objects and manages the network
+communication between different instances of the same distributed
+object.
+
+The master version of a distributed object is registered with a
+\textsf{eq::net::Session}, which assigns a session-unique identifier to
+the object. This identifier can be used by other nodes to map their
+instance of the object, thus synchronizing the object's data with the
+remotely registered master version.
+
+Distributed objects can be static (immutable) or dynamic. Dynamic
+objects are versioned.
 
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+%\begin{figure}[h!t]
+  \vspace{-2ex}
+  \subfigure[]{\label{fObjectDist1}
+    \includegraphics[width=.382\textwidth]{images/umlObjectSubclass}}\hfil
+  \subfigure[]{\label{fObjectDist2}
+    \includegraphics[width=.382\textwidth]{images/umlObjectProxy}}\hfil
+  \subfigure[]{\label{fObjectDist3}
+    \includegraphics[width=.382\textwidth]{images/umlObjectMI}}%
+  {\caption{\label{fObjectDist}Object distribution using subclassing,
+      proxies or multiple inheritance}}
+  \vspace{-4ex}
+\end{wrapfigure}
 Distributed objects are addressed using session-unique identifiers,
-because pointers to other objects cannot be distributed directly, since
-they have no meaning on remote nodes. The session with which distributed
-objects are registered is normally the \textsf{eq::Config}, which is
+because pointers to other objects cannot be distributed directly; they
+have no meaning on remote nodes. The session used to register
+distributed objects is normally the \textsf{eq::Config}, which is
 subclassed from \textsf{eq::net::Session}.
 
-The shared data entry point for the render clients is the identifier
+The entry point for shared data on a render client is the identifier
 passed by the application to \textsf{Config::init}. This identifier
@@
 passed by Equalizer to all \textsf{configInit} task methods. Normally
-this initial object is mapped by the render clients, and it contains
-identifiers to other shared data objects.
+this initial object is mapped by the render clients in
+\textsf{Node::configInit}. It typically contains identifiers of other
+shared data objects.
 
@@
 frames. Similar to the initial identifier passed to
-\textsf{Config::init}, an object identifier or version can be passed to
-\textsf{Config::startFrame}. Equalizer will pass this identifier to all
-\textsf{frameStart} task methods. In \textsf{eqPly}, the frame-specific
-data, e.g., the global camera data, is versioned. The frame data
-identifier is passed in the initial data, and the frame data version is
-passed with each new frame request.
+\textsf{Config::init}, an object identifier or object version can be
+passed to \textsf{Config::startFrame}. Equalizer will pass this
+identifier to all \textsf{frameStart} task methods. In \textsf{eqPly},
+the frame-specific data, e.g., the global camera data, is versioned. The
+frame data identifier is passed in the initial data, and the frame data
+version is passed with each new frame request.
+
+There are multiple ways of implementing data distribution for an
+existing C++ class hierarchy:
+\begin{description}
+  \item[Subclassing]: The classes to be distributed inherit from
+    \textsf{eq::net::Object} and implement the serialization
+      methods. This requires modification of the existing classes. This
+      is used for \textsf{eqPly::InitData} and
+      \textsf{eqPly::FrameData}. (\fig{fObjectDist1})
+  \item[Proxies]: For each object to be distributed, a proxy object is
+    created which manages data distribution for its associated
+    object. This requires the application to track changes on the object
+    separately from the object itself. The model data distribution of
+    eqPly is using this pattern. (\fig{fObjectDist2})
+  \item[Multiple Inheritance]: A new class inheriting from the class to
+    be distributed and from \textsf{eq::net::Object} implements the data
+    distribution. This requires the application to instantiate a
+    different type of object instead of the existing object, and to
+    create wrapper methods in the superclass calling the original method
+    and setting the appropriate dirty bits. This pattern is not used in
+    eqPly. (\fig{fObjectDist3})
+\end{description}
 
@@
 version of a distributed object. The change type determines the memory
-footprint and the contract for calling the serialization methods. The
-following change types are possible:
+footprint and the contract for the serialization methods. The following
+change types are possible:
 
@@
 
+The data streams implement efficient buffering and data transport
+between nodes. They perform no type checking or transformation on the
+data. It is the application's responsibility to exactly match the order
+and types of variables during serialization and de-serialization.
+
 Custom data type serializers can be implemented by providing the
 appropriate serialization functions. No pointers should be directly
-transmitted through the data streams. Typically the corresponding object
-is a distributed object as well, and it's identifier and potentially
-version is transmitted in place of a pointer.
+transmitted through the data streams. For pointers, the corresponding
+object is typically a distributed object as well, and it's identifier
+and potentially version is transmitted in place of its pointer.
 
@@
 streaming all member variables to or from the provided data
-streams. Efficient buffering and data transport between nodes is
-implemented in the data streams:
+streams:
 
@@
 {
-    os << _frameDataID << _modelID << _windowSystem << _useVBOs << _useGLSL
-       << _invFaces ;
+    os << _frameDataID << _windowSystem << _renderMode << _useGLSL << _invFaces
+       << _logo;
 }
@@
 {
-    is >> _frameDataID >> _modelID >> _windowSystem >> _useVBOs >> _useGLSL
-       >> _invFaces;
+    is >> _frameDataID >> _windowSystem >> _renderMode >> _useGLSL >> _invFaces
+       >> _logo;
 
@@
 
-The data input and output streams perform no type checking on the data.
-It is the application's responsibility to exactly match the order and
-types of variables during serialization and de-serialization.
-
 \subsubsection{FrameData - a Versioned Distributed Object}
@@
 Versioned objects have to override \textsf{getChangeType} to indicate
-how they want to have changes to be handled. All versioned objects
-currently implemented have the following characteristics:
+how they want to have changes to be handled. All types of versioned
+objects currently implemented have the following characteristics:
 
@@
 \item Newly mapped slave instance are mapped to the oldest available
-  version by default, or to the specified version in
+  version by default, or to the version specified when calling
   \textsf{Session::mapObject}.
@@
 version has not yet been committed or is still in transmission.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
+  {\caption{\label{fUMLObject}eq::Object and eq::net::Object}}
+\end{wrapfigure}
+Not syncing a mapped, versioned object creates a memory leak. The method
+\textsf{Object::notifyNewHeadVersion} is called whenever a new version
+is received by the node. The notification is send from the command
+thread, which is different from the node main thread. The object should
+not be synced from this method, but instead a message may be send to the
+application, which then takes the appropriate action. The default
+implementation asserts when too many versions have been queued in order
+to detect memory leaks during development.
 
@@
 
-The \textsf{eq::Object} implements one possible way of tracking data
-changes in versioned objects using the concept of dirty bits, which
-facilitates the implementation of the data distribution code.
-
-The \textsf{FrameData} object tracks changes by setting the appropriate
-dirty bit whenever some of its data is changed. The serialization
-methods are called by the \textsf{eq::Object} with the dirty bit mask
-needed to serialize all data or the changes since the last
+The creation of distributed, versioned objects is further facilitated by
+\textsf{eq::Object}, which implements the most common way of tracking
+data changes in versioned objects. The concept of a dirty bit mask is
+used to mark parts of the object for serialization, while preserving the
+capability to inherit objects. Other ways of implementing change
+tracking, e.g., using incarnation counters, can still be implemented by
+using \textsf{eq::net::Object} which leaves this flexibility to the
+developer. \fig{fUMLObject} shows the relationship between
+\textsf{eq::Object} and \textsf{eq::net::Object}.
+
+The \textsf{FrameData} is subclassed from \textsf{eq::Object}, and
+consequently tracks its changes by setting the appropriate dirty bit
+whenever it is changed. The serialization methods are called by the
+\textsf{eq::Object} with the dirty bit mask needed to serialize all
+data, or with the dirty bit mask of the changes since the last
 \textsf{commit}. The \textsf{FrameData} only defines its own dirty bits
@@
     DIRTY_VIEW    = eq::Object::DIRTY_CUSTOM << 2,
+    DIRTY_MESSAGE = eq::Object::DIRTY_CUSTOM << 3,
 };
@@
     if( dirtyBits & DIRTY_CAMERA )
-        os << _translation << _rotation;
+        os << _translation << _rotation << _modelRotation;
     if( dirtyBits & DIRTY_FLAGS )
-        os << _modelID << _renderMode << _color << _ortho << _statistics
-           << _help << _wireframe;
+        os << _modelID << _renderMode << _colorMode << _ortho << _statistics
+           << _help << _wireframe << _pilotMode;
     if( dirtyBits & DIRTY_VIEW )
         os << _currentViewID;
+    if( dirtyBits & DIRTY_MESSAGE )
+        os << _message;
 }
@@
     if( dirtyBits & DIRTY_CAMERA )
-        is >> _translation >> _rotation;
+        is >> _translation >> _rotation >> _modelRotation;
     if( dirtyBits & DIRTY_FLAGS )
-        is >> _modelID >> _renderMode >> _color >> _ortho >> _statistics
-           >> _help >> _wireframe;
+        is >> _modelID >> _renderMode >> _colorMode >> _ortho >> _statistics
+           >> _help >> _wireframe >> _pilotMode;
     if( dirtyBits & DIRTY_VIEW )
         is >> _currentViewID;
+    if( dirtyBits & DIRTY_MESSAGE )
+        is >> _message;
 }
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
-  {\caption{\label{fModelDist}Scene Data in eqPly}}
-\end{wrapfigure}
 \textsf{eqPly} uses static distributed objects to distribute the model
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
+  {\caption{\label{fModelDist}Scene Data in eqPly}}
+\end{wrapfigure}
 The identifier of the model's root object of this distributed hierarchy
@@
 
-%\begin{wrapfigure}{r}{.618\textwidth}
-\begin{figure}[ht!]\center
+\begin{wrapfigure}{r}{.618\textwidth}
+%\begin{figure}[ht!]\center
   \includegraphics[width=.618\textwidth]{images/objects.pdf}
   {\caption{\label{fObjects}Scene Graph Distribution}}
-\end{figure}
+\end{wrapfigure}
 Applications distributing a dynamic scene graph use the frame data
@@
 
-%\begin{figure}[ht!]\center
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}
-  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
-  {\caption{\label{fConfigInit}Config Initialization Sequence}}
-  \vspace{-4ex}
-\end{wrapfigure}
 The config initialization happens in parallel, that is, all config
@@
 
+\begin{figure}[ht!]\center
+%\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2ex}
+  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
+  {\caption{\label{fConfigInit}Config Initialization Sequence}}
+  \vspace{-4ex}
+\end{figure}
 The tasks are executed by the node and pipe threads in parallel. The
@@
 problem to parallel rendering. Tracking device support has already been
-solved by a number of implementations\footnote{VRCO Trackd VRPN, etc.},
-which can easily be integrated with Equalizer. The example code in
-\textsf{eqPly} provides a reference implementation for the integration
-of such a tracking library. \sref{sTracking} provides more background on
-head tracking.
+solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
+which can and have been easily integrated with Equalizer. The example
+code in \textsf{eqPly} provides a reference implementation for the
+integration of such a tracking library. \sref{sTracking} provides more
+background on head tracking.
 
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/DB.png}
+  {\caption{\label{fDBDest}Destination view of a DB compound using
+      demonstrative coloring}}
+\end{wrapfigure}
 After the basic view setup, a directional light is configured, and the
@@
 
-Furthermore, a white color is set in case the model does not contain
-color information, or the color information is not used. In sort-last
-rendering, \textsf{eqPly} uses a different color for each channel to
-illustrate the database decomposition, as shown in \fig{fDBDest}. The
-Equalizer channel provides a method to obtain a random, but unique color
-for all channels in the configuration. This color is determined by the
-server to ensure uniqueness across all channels of the configuration:
+Three different ways of coloring the object are possible: Using the
+colors of the mode, using an unique per-channel color to demonstrate the
+decomposition as shown in \fig{fDBDest}, or using solid white for
+anaglyphics stereo. The model colors are per-vertex and are set during
+rendering, whereas the unique per-channel color is set in
+\textsf{frameDraw} for the whole model:
 
@@
 
-    const Pipe*      pipe      = static_cast<Pipe*>( getPipe( ));
-    const FrameData& frameData = pipe->getFrameData();
-
-    glTranslatef( frameData.data.translation.x,
-                  frameData.data.translation.y,
-                  frameData.data.translation.z );
-    glMultMatrixf( frameData.data.rotation.ml );
-
-    Config*          config = static_cast< Config* >( getConfig( ));
-    const Model*     model  = getModel();
-    const eq::Range& range  = getRange();
-
-    if( !range.isFull( )) // Color DB-patches
-    {
-        const vmml::Vector3ub color = getUniqueColor();
-        glColor3ub( color.r, color.g, color.b );
-    }
-    else if( !frameData.data.color || (model && !model->hasColors( )) )
-    {
-        glColor3f( 1.0f, 1.0f, 1.0f );
-    }
-
+    const FrameData& frameData = getFrameData();
+
+    glPolygonMode( GL_FRONT_AND_BACK, 
+                   frameData.useWireframe() ? GL_LINE : GL_FILL );
+
+    const eq::Vector3f& translation = frameData.getCameraTranslation();
+
+    glMultMatrixf( frameData.getCameraRotation().array );
+    glTranslatef( translation.x(), translation.y(), translation.z() );
+    glMultMatrixf( frameData.getModelRotation().array );
+
+    if( frameData.getColorMode() == COLOR_DEMO )
+    {
+        const eq::Vector3ub color = getUniqueColor();
+        glColor3ub( color.r(), color.g(), color.b() );
+    }
+    else
+        glColor3f( .75f, .75f, .75f );
 \end{lstlisting}}
 
-Finally the model, which has been loaded by the node, is rendered. If
-the model was not loaded during node initialization, a quad is drawn in
-its place:
-
-{\footnotesize\begin{lstlisting}
+Finally the model is rendered. If the model was not loaded during node
+initialization, a quad is drawn in its place:
+
+{\footnotesize\begin{lstlisting}
+    const Model* model = _getModel();
     if( model )
@@
     {
-        glColor3f( 1.f, 1.f, 0.f );
         glNormal3f( 0.f, -1.f, 0.f );
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/DB.png}
-  {\caption{\label{fDBDest}Destination view of a DB compound}}
-\end{wrapfigure}
 To draw the model, a helper class for view frustum culling is set up
@@
 
-    state.setColors( frameData.data.color && 
-                     range.isFull() && 
-                     model->hasColors() );
+    if( frameData.getColorMode() == COLOR_MODEL && model->hasColors( ))
+        state.setColors( true );
+    else
+        state.setColors( false );
+
     _initFrustum( culler, model->getBoundingSphere( ));
-
     model->beginRendering( state );
-
 \end{lstlisting}}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
+%\begin{wrapfigure}{r}{.618\textwidth}
+\begin{figure}[h!t]\center
   \includegraphics[width=.618\textwidth]{images/render.pdf}
   {\caption{\label{fRender}Main Render Loop}}
-\end{wrapfigure}
+\end{figure}
 The rendering loop maintains a list of candidates to render, which
@@
 
+\subsubsection{Assembly}
+
+Like most applications, eqPly can use the default implementation of the
+\textsf{frameReadback} and \textsf{frameAssemble} task methods. Only
+\textsf{frameReadback} is overwritten for an optimization. Since eqPly
+does not need the alpha channel on the destination view, the output
+frames are flagged to ignore it, which allows the compressor to drop
+25\% of the data during image transfer:
+
+{\footnotesize\begin{lstlisting}
+void Channel::frameReadback( const uint32_t frameID )
+{
+    // OPT: Drop alpha channel from all frames during network transport
+    const eq::FrameVector& frames = getOutputFrames();
+    for( eq::FrameVector::const_iterator i = frames.begin(); 
+         i != frames.end(); ++i )
+    {
+        (*i)->setAlphaUsage( false );
+    }
+
+    eq::Channel::frameReadback( frameID );
+}
+\end{lstlisting}}
+
 \section{Advanced Features}
@@
 \textsf{eqPly} section. Where possible, code examples from the Equalizer
-distribution are used to illustrate one usage of the feature. Its
+distribution are used to illustrate usage of the feature. Its
 purpose is to provide information on how to address a typical problem or
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+\vspace{-2ex}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\label{fEventProcessing}Event Processing}}
+\vspace{-4ex}
+\end{wrapfigure}
 Events are received by an event handler. The event handler finds the
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
-\textsf{Window::processEvent} is responsible for either handling
-the event locally, or for translating it into a generic
+\textsf{Window::processEvent} is responsible for handling
+the event locally and for translating it into a generic
 \textsf{ConfigEvent}. The config events are send to the application
@@
 
-\begin{figure}[ht!]\center
+\begin{figure}[h!t]\center
   \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\label{fFrameSync}Per-node frame synchronization}}
-\end{wrapfigure}
 The application has extended control over the task synchronization
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\label{fFrameSync}Per-node frame synchronization}}
+\end{wrapfigure}
 Each entity uses \textsf{waitFrame\-Started} to block on the parent's
@@
   \subfigure[]{\label{fMono}
-    \includegraphics[height=3.5cm]{images/mono}}\hfil
+    \includegraphics[height=2.5cm]{images/mono}}\hfil
   \subfigure[]{\label{fStereo}
-    \includegraphics[height=3.5cm]{images/stereo}}\hfil
+    \includegraphics[height=2.5cm]{images/stereo}}\hfil
   \subfigure[]{\label{fTracked}
-    \includegraphics[height=3.5cm]{images/tracked}}%
-  {\caption{\label{fImmersive}Monoscopic, Stereoscopic and Tracked
+    \includegraphics[height=2.5cm]{images/tracked}}
+  \subfigure[]{\label{fHMD}
+    \includegraphics[height=2.5cm]{images/hmd}}%
+  {\caption{\label{fImmersive}Monoscopic, Stereoscopic, Tracked and HMD
     frusta}}
@@
 retrieved using \textsf{Channel::getHeadTransform}.
+
+For head-mounted displays (HMD), the tracking information is used to
+move the frusta with the observer, as shown in \fig{fHMF}. This results
+in different projections compared to normal tracking with fixed
+projection screens.
 
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \vspace{-2ex}
+  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
+  {\caption{\label{fViewUML}UML hierarchy of \textsf{eqPly::View}}}
+  \vspace{-5ex}
+\end{wrapfigure}
 Changing view frustum parameters is typically done for non-fullscreen
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
-  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
-\end{wrapfigure}
 A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
@@
 
+\subsubsection{Compression Plugins}
+
+Compression plugins allow the creation of runtime-loadable modules for
+image compression. Equalizer will search pre-defined directories during
+\textsf{eq::init} for dynamic shared objects (DSO) containing compression
+libraries (EqualizerCompressor*.dll on Windows, libeqCompressor*.dylib
+on Mac OS X, libeqCompressor*.so on Linux).
+
+The interface to a compression DSO is a C API, which allows to maintain
+binary compatiblity across Equalizer versions. Furthermore, the
+definition of an interface facilitates the creation of new compression
+codecs for developers.
+
+Please refer to the Equalizer API documentation on the website for the
+full specification for compression plugins. The Equalizer DSO doubles as
+a compression plugin and implements a set of compression engines, which
+can be used as a reference implementation.
+
+Each compression DSO may contain multiple compression engines. The
+number of compressors in the DSO is queried by Equalizer using
+\textsf{EqCompressGetNumCompressors}.
+
+For each compressor, \textsf{EqCompressorGetInfo} is called to retrieve
+the information about the compressor. The information contains the API
+version the DSO was written against, a unique name of the compressor,
+the type of input data accepted as well as information about the
+compressor's speed, quality and compression ratio.
+
+Each image transported over the network allocates its own compressor or
+decompressor instance. This allows compressor implementations to
+maintain information in a thread-safe manner. The handle to a compressor
+or decompressor instance is a void pointer, which typically hides a C++
+object instantiated by the compression DSO.
+
+A unit test is delivered with Equalizer which runs all compressors
+against a set of images and provides performance information to
+calculate the compressor characteristics.
+
 \subsubsection{\label{sDirectSend}Parallel Direct Send Compositing}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\label{fDirectSend}Direct Send Compositing}}
+\end{wrapfigure}
 The main idea behind direct send is to parallelize the costly
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\label{fDirectSend}Parallel Direct Send Compositing}}
-\end{wrapfigure}
 In direct send compositing, each rendering channel is also responsible
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
+  {\caption{\label{fAssembly}Hierarchy of assembly classes}}
+\end{wrapfigure}
 The frame synchronization through the frame data allows the input frame
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
-  {\caption{\label{fAssembly}Hierarchy of assembly classes}}
-\end{wrapfigure}
 For output frames, the frame data's pixel viewport is the area of the
@@
 
-%\begin{wrapfigure}{r}{.618\textwidth}
-\begin{figure}[h!t]\center
-  \includegraphics[width=.9\textwidth]{images/compositor.pdf}
-  {\caption{\label{fCompositor}Functional diagram of the compositor}}
-\end{figure}
 The \textsf{Compositor} class gathers a set of static functions which
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+%\begin{figure}[h!t]\center
+  \includegraphics[width=.618\textwidth]{images/compositor.pdf}
+  {\caption{\label{fCompositor}Functional diagram of the compositor}}
+\end{wrapfigure}
 On a high level, the compositor combines multiple input frames using 2D
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
-\end{wrapfigure}
 The full power of scalable volume rendering is however in DB (sort-last)
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
+\end{wrapfigure}
 Conceptually, the individual volume bricks of each of the source
@@
 visualize with any other approach.
+
+For recomposition, the 2D frame buffer contents are blended to form a
+seamless picture. For correct blending, the frames are ordered in the
+same back-to-front order as the slices used for rendering, and use the
+same blending parameters. Simplified, the frame buffer images are
+`thick' slices which are `rendered' by writing their content to the
+destination frame buffer using the correct order.
+
+For orthographic rendering, determining the compositing order of the
+input frames is trivial. The screen-space orientation of the volume
+bricks determines the order in which they have to be composited. The
+bricks in \textsf{eVolve} are created by slicing the volume along one
+dimension. Therefore the range of the resulting frame buffer images,
+together with the sorting order, is used to arrange the frames during
+compositing. \fig{fBlendOrtho} shows this composition for one view.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
+Finding the correct assembly order for perspective frusta is more
+complex. The perspective distortion invalidates a simple orientation
+criteria like the one used for orthographic frusta. For the view and
+frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
+  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
+1-4-3-2.
 
@@
 
-For recomposition, the 2D frame buffer contents are blended to form a
-seamless picture. For correct blending, the frames are ordered in the
-same back-to-front order as the slices used for rendering, and use the
-same blending parameters. Simplified, the frame buffer images are
-`thick' slices which are `rendered' by writing their content to the
-destination frame buffer using the correct order.
-
-For orthographic rendering, determining the compositing order of the
-input frames is trivial. The screen-space orientation of the volume
-bricks determines the order in which they have to be composited. The
-bricks in \textsf{eVolve} are created by slicing the volume along one
-dimension. Therefore the range of the resulting frame buffer images,
-together with the sorting order, is used to arrange the frames during
-compositing. \fig{fBlendOrtho} shows this composition for one view.
-
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
-Finding the correct assembly order for perspective frusta is more
-complex. The perspective distortion invalidates a simple orientation
-criteria like the one used for orthographic frusta. For the view and
-frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
-  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
-1-4-3-2.
-
 To compute the assembly order, \textsf{eVolve} uses the angle between
@@
 Statistics are measured in milliseconds since the configuration was
-initialized. On each node, the global configuration clock is reset at
-the same time by the server during \textsf{Config::init}. Each statistic
-event records the originator's (channel, window, node or config) unique
-identifier.
+initialized. On each node, the global configuration clock is sychronized
+with the server clock. Each statistic event records the originator's
+(channel, window, frame or config) unique identifier.
 
 \begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/statistics.png}
+  \includegraphics[width=.618\textwidth]{images/statistics}
   {\caption{\label{fStatistics}Statistics for a two-node 2D compound}}
@@
 \textsf{processEvent} method. The default implementation sends these
-events to the config, using \textsf{Config::sendEvent}, as explained in
+events to the config using \textsf{Config::sendEvent}, as explained in
 \sref{sEventHandling}. When the default implementation of
@@
 
+\fig{fStatisticsDetail} shows a detail view of \fig{fStatistics}. The
+statistics shown are for a two-node 2D compound. The destination channel
+is on the \textsf{appNode} and contributes to the rendering.
+
 \begin{figure}[h!t]
-  \includegraphics[width=\textwidth]{images/statisticsDetail.pdf}
-    {\caption{\label{fStatisticsDetail}Annotated Detail of the
-        Statistics from \fig{fStatistics}.}}
+  \includegraphics[width=\textwidth]{images/statisticsDetail}
+  {\caption{\label{fStatisticsDetail}Detail of the Statistics from
+      \fig{fStatistics}.}}
 \end{figure}
 
-\fig{fStatisticsDetail} shows a detail view of \fig{fStatistics},
-together with some annotations for the overlay. The statistics shown are
-for a two-node 2D compound. The destination channel is on the
-\textsf{appNode} and contributes to the rendering.
-
-This compound executes two \textsf{Channel::frameDraw} tasks, one
-\textsf{Channel::frame\-Read\-back} task on the remote node and one
-\textsf{Channel::frameAssemble} task on the local node.
+This configuration executes two \textsf{Channel::frameDraw} tasks, one
+\textsf{Channel::frame\-Read\-back} task on the remote node, one
+\textsf{Channel::frameAssemble} task on the local node, as well as frame
+transmission and compression.
 
 The X axis is the time. One pixel on the screen corresponds to a given
-time unit, typically one millisecond. The scale is adapted dynamically
+time unit, here one millisecond per pixel. The scale is adapted dynamically
 to powers-of-ten milliseconds to fit the statistics into the available
@@
 The rightmost pixel is the most current time. On the Y axis are the
-rendering entities: channels, windows, nodes and the config. The upper
-channel is the local channel since it executes \textsf{frameAssemble},
-and the lower channel is the remote channel, executing {frameReadback}.
+entities: channels, windows, frames and the config. The upper channel is
+the local channel since it executes \textsf{frameAssemble}, and the
+lower channel is the remote channel, executing \textsf{frameReadback}.
 
@@
 
+The frame statistics provide information on the image transmission for
+scalable rendering. They are grouped per frame, not per node, that is,
+the transmit and compress part are executed by one node and the receive
+by another, even though the statistics are drawn in the same line. The
+decompression time is part of the receive statistic. The total
+compression ratio is printed with each compression statistic, in this
+case the image has been compressed to 11\% of its original size.
+
 The white \textsf{Window::swapBuffers} task might also take also a long
@@
 CPU.
-
-The configuration used in this examples does not use a swap
-barrier. Swap barrier tasks are solid red events directly in front of
-the white swap buffer task.
 
@@
 
+\subsection{File Format Version}
+
 \subsection{\label{sGlobal}Global Section}
@@
   \textbf{Name} & EQ\_CONNECTION\_IATTR\_TYPE\\
-  \textbf{Value} & TCPIP \textbar \ SDP\\
+  \textbf{Value} & TCPIP \textbar \ SDP \textbar \ PIPE [Win32 only]\\
   \textbf{Default} & TCPIP\\
-  \textbf{Details} & The protocol for connections. SDP
-  programmatically selects the socket direct protocol (AF\_INET\_SDP)
-  provided by most InfiniBand protocol stacks, TCPIP uses normal TCP
-  sockets (AF\_INET).\\
+  \textbf{Details} & The protocol for connections. SDP programmatically
+  selects the socket direct protocol (AF\_INET\_SDP) provided by most
+  InfiniBand protocol stacks, TCPIP uses normal TCP
+  sockets (AF\_INET). PIPE uses a named pipe to communicate between
+  two processes on the same machine.\\
   \hline
@@
   \hline
+  \textbf{Name} & EQ\_CONNECTION\_IATTR\_PIPE\_FILENAME\\
+  \textbf{Value} & string\\
+  \textbf{Default} & none\\
+  \textbf{Details} & The filename of the named pipe used by the server or
+  node. The filename has to be unique on the local host.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
   \textbf{Name} & EQ\_CONNECTION\_IATTR\_LAUNCH\_TIMEOUT\\
@@
   \textbf{Default} & 0.05\\
-  \textbf{Details} & The distance in meters between the left and the
-  right eye, i.e., the eye separation. The eye base influences the
+  \textbf{Details} & The default distance in meters between the left and
+  the right eye, i.e., the eye separation. The eye base influences the
   frustum during stereo rendering. See \sref{sTracking} for details.\\
@@
   \sref{sThreadModel} for details.\\
-  \hline
-\end{tabularx}\\\vfill
-
-\begin{tabularx}{\textwidth}{|l|X|}
-  \hline
-  \textbf{Name} & EQ\_NODE\_IATTR\_HINT\_STATISTICS\\
-  \textbf{Value} & OFF \textbar \ FASTEST [ON] \textbar \ NICEST\\
-  \textbf{Default} & FASTEST [Release Build]\\
-                   & NICEST [Debug Build]\\
-  \textbf{Details} & Determines how statistics are gathered. Fastest and
-  nicest currently gather statistics in the same way.\\
-  \textbf{See also} & EQ\_WINDOW\_IATTR\_HINT\_STATISTICS 
-  EQ\_CHANNEL\_IATTR\_HINT\_STATISTICS\\
   \hline
@@
   \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_DRAWABLE\\
-  \textbf{Value} & window \textbar \ pbuffer\\
+  \textbf{Value} & window \textbar \ pbuffer \textbar \ FBO\\
   \textbf{Default} & window\\
@@
   drawable. PBuffers are off-screen drawables created using window
-  system-dependent PBuffer API's. To calculate the PBuffer size on
-  unconnected devices, a pipe viewport size of 4096x4096 is assumed,
+  system-dependent PBuffer API's. FBO are off-screen frame buffer
+  objects. To calculate the PBuffer or FBO size on unconnected devices,
+  a pipe viewport size of 4096x4096 is assumed,
   unless specified otherwise using the pipe's viewport parameter.\\
@@
   \textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_COLOR\\
-  \textbf{Value} & unsigned\\
+  \textbf{Value} & unsigned \textbar \ RGBA16F \textbar \ RGBA32F\\
   \textbf{Default} & AUTO\\
@@
   value, and some select a visual with at least the number of color
-  planes specified. AUTO selects a visual with a reasonable quality,
+  planes specified. RGBA16F and RGBA32F select floating point
+  framebuffers with 16 or 32 bit precision per component,
+  respectively. AUTO selects a visual with a reasonable quality,
   typically eight bits per color.\\
@@
 A connection description defines the network parameters of an Equalizer
-process. Currently two connection types, TCP/IP and SDP, are
+process. Currently TCP/IP, SDP and PIPE connection types are
 supported. TCP/IP creates a TCP socket. SDP is very similar, except that
 the address family \textsf{AF\_INET\_SDP} instead of \textsf{AF\_INET}
-is used to enforce a SDP connection. Note that you can also use the
-transparent mode provided by most InfiniBand implementations to use SDP
-with TCP connections.
+is used to enforce a SDP connection. PIPE uses a named pipe for fast
+inter-process communication on Windows.
+
+Note that you can also use the transparent mode provided by most
+InfiniBand implementations to use SDP with TCP connections.
 
@@
 The hostname is the IP address or (resolvable) host name. A server or
-node may have multiple connection descriptions.
+node may have multiple connection descriptions, for example to use a
+named pipe for local communications and TCP/IP for remote nodes.
 
@@
 {
-    connection // 0-n times, listening connections of the server
-    {
-        type       TCPIP | SDP
+    connection # 0-n times, listening connections of the server
+    {
+        type       TCPIP | SDP | PIPE [Win32 only]
         TCPIP_port unsigned
+        PIPE_filename string
         hostname   string
@@
 
-    config {...} // 1-n times, currently only the first one is used by the server
+    config {...} # 1-n times, currently only the first one is used by the server
 }
@@
 
-A configuration has a number of parameters, nodes and compounds.
+A configuration has a number of parameters, nodes, observers, layouts,
+canvases and compounds.
 
 The nodes and their children describe the rendering resources in a
-natural, hierarchical way. Compounds use rendering resources (channels)
-to execute rendering tasks. For an introduction on the concepts of
-nodes, pipes, windows, channels and compounds please refer to
-\sref{sConfig} and \sref{sCompounds}.
+natural, hierarchical way. Observers, layouts and canvases describe the
+properties of the physical projection system. Compounds use rendering
+resources (channels) to execute rendering tasks.
+
+For an introduction to writing configurations and the concepts of the
+configuration entities please refer to \sref{sConfig}.
 
@@
 {
-    latency int    // # of frames nodes may fall behind application
+    latency int    # Nr. of frames nodes may fall behind application, default 1
     attributes
     {
-         eye_base float // distance between left and right eye
-    }
-
-    (node|appNode) {...}// 1-n times, a system in the cluster
-                   // 0|1 appNode: launches render thread within app process
-    compound {...} // 1-n times
+         eye_base float # distance between left and right eye
+    }
+
+    appNode {...}  # 0|1 times, launches render threads within app process
+    node {...}     # 1-n times, a system in the cluster
+    observer {...} # 0-n times, one tracked entity
+    layout {...}   # 0-n times, logical views on canvases
+    canvas {...}   # 0-n times, physical segment layout of a projection surface
+    compound {...} # 0-n times, rendering description
 }
@@
     name     string
-    connection // 0-n times, possible connections to this node
-    {
-        type          TCPIP | SDP
+    connection # 0-n times, possible connections to this node
+    {
+        type          TCPIP | SDP | PIPE
         TCPIP_port    unsigned
+        PIPE_filename string
         hostname      string
-        command       string      // render client launch command
-        command_quote 'character' // launch command argument quote char
-        timeout       unsigned    // timeout in milliseconds for launch
+        command       string      # render client launch command
+        command_quote 'character' # launch command argument quote char
+        timeout       unsigned    # timeout in milliseconds for launch
     }
@@
         thread_model    ASYNC | DRAW_SYNC | LOCAL_SYNC
-        hint_statistics FASTEST | NICEST | OFF
-    }
-    pipe {...} // 1-n times
+    }
+    pipe {...} # 1-n times
 }
@@
     name     string
-    port     unsigned     // X server number or ignored
-    device   unsigned     // graphics adapter number
-    viewport [ viewport ] // default: autodetect
+    port     unsigned     # X server number or ignored
+    device   unsigned     # graphics adapter number
+    viewport [ viewport ] # default: autodetect
     attributes
     {
-        hint_thread   OFF | ON   // default ON
-    }
-
-    window {...} // 1-n times
+        hint_thread   OFF | ON   # default ON
+    }
+
+    window {...} # 1-n times
 }
@@
 has a name, a viewport and attributes. The name of a window can be used
-for debugging, it has no influence on the execution of Equalizer.
+for debugging, it has no influence on the execution of Equalizer, other
+then it being used as the window title by the default window creation
+methods.
 
@@
     name     string
-    viewport [ viewport ] // wrt pipe, default full screen
+    viewport [ viewport ] # wrt pipe, default full screen
 
@@
         hint_fullscreen     OFF | ON
-        hint_swapsync       OFF | ON         // AGL, WGL only
-        hint_drawable       window | pbuffer
+        hint_swapsync       OFF | ON         # AGL, WGL only
+        hint_drawable       window | pbuffer | FBO
         hint_statistics     off | fastest [on] | nicest
-        planes_color        unsigned
+        planes_color        unsigned | RGBA16F | RGBA32F
         planes_alpha        unsigned
@@
 
-    channel {...} // 1-n times
+    channel {...} # 1-n times
 }
@@
 viewport and attributes. The name of the channel is used to identify the
-channel in the respective compounds. It should be unique within the config.
+channel in the respective segments or compounds. It should be unique
+within the config.
 
@@
 
+The channel can have an alternate drawable description. Currently, the
+window's framebuffer can be replaced by framebuffer objects bound to the
+window's OpenGL context. The window's default framebuffer can be
+partially overwritten with framebuffer objects.
+
 For a description of channel attributes please refer to \sref{sGlobal}.
@@
     name     string
-    viewport [ viewport ] //wrt window, default full window
+    viewport [ viewport ] #wrt window, default full window
+    drawable [ FBO_COLOR FBO_DEPTH FBO_STENCIL ]
 
@@
         hint_statistics     OFF | FASTEST [ON] | NICEST
+    }
+}
+\end{lstlisting}}
+
+\subsubsection{Observer Section}
+
+An observer represents a tracked entity, i.e, one user. It has a name
+and an eye separation. The name of an observer can be used for
+debugging, it has no influence on the execution of Equalizer. It can be
+used to reference the observer in views, in which case the name should
+be unique. Not all views have to be tracked by an observer.
+
+{\footnotesize\begin{lstlisting}
+observer # 0...n times
+{
+    name    string
+    eyeBase float
+}
+\end{lstlisting}}
+
+\subsubsection{Layout Section}
+
+A layout represent a set of logical views on one or more canvases. It
+has a name and child views. The name of a layout can be used for
+debugging, it has no influence on the execution of Equalizer. It can be
+used to reference the layout, in which case the name should be unique.
+
+A layout is applied to a canvas. If no layout is applied to a canvas,
+nothing is rendered on this canvas, i.e, the canvas is inactive.
+
+The layout assignment can be changed at run-time by the application. The
+intersection between views and segments defines which output
+(sub-)channels are available. These output channels are typically used
+as destination channels in a compound. They are automatically created
+during configuration loading or creation.
+
+{\footnotesize\begin{lstlisting}
+layout  # 0...n times
+{
+    name string
+    view {...}  # 1...n times
+}
+\end{lstlisting}}
+
+\subsubsection{View Section}
+
+A view represents a 2D area on a canvas. It has a name, viewport,
+observer and frustum. The name of a view can be used for debugging, it
+has no influence on the execution of Equalizer. It can be used to
+reference the view, in which case the name should be unique.
+
+A view can have a frustum description. The view's frustum overrides
+frusta specified at the canvas or segment level. This is typically used
+for non-physically correct rendering, e.g., to compare two models
+side-by-side. If the view does not specify a frustum, the corresponding
+destination channels will use the sub-frustum resulting from the
+view/segment intersection.
+
+A view is a view on the application's model, in the sense used by the
+Model-View-Controller pattern. It can be a scene, viewing mode, viewing
+position, or any other representation of the application's data.
+
+{\footnotesize\begin{lstlisting}
+view  # 1...n times
+{
+    name     string
+    observer observer-ref
+    viewport [ viewport ]
+
+    wall       // frustum description
+    {
+        bottom_left  [ float float float ]
+        bottom_right [ float float float ]
+        top_left     [ float float float ]
+        type         fixed | HMD
+    }
+    projection // alternate frustum description, last one wins
+    {
+        origin       [ float float float ]
+        distance     float
+        fov          [ float float ]
+        hpr          [ float float float ]
+    }
+}
+\end{lstlisting}}
+
+\subsubsection{Canvas Section}
+
+A canvas represents a logical projection surface of multiple
+segments. It has a name, frustum, layouts, and segments. The name of a
+canvas can be used for debugging, it has no influence on the execution
+of Equalizer. It can be used to reference the canvas, in which case the
+name should be unique.
+
+Each canvas is made of one or more segments. Segments can be planar or
+non-planar to each other, and can overlap or have gaps between each
+other. A canvas can define a frustum, which will create default, planar
+sub-frusta for its segments.
+
+The layouts referenced by the canvas can be applied to it by the
+application at run-time. One layout can be referenced by multiple
+canvases. The first layout is the layout active by default.
+
+{\footnotesize\begin{lstlisting}
+canvas  # 0...n times
+{
+    name    string
+    layout  layout-ref   # 1...n times
+
+    wall       
+    {
+            bottom_left  [ float float float ]
+            bottom_right [ float float float ]
+            top_left     [ float float float ]
+            type         fixed | HMD
+    }
+    projection
+    {
+            origin       [ float float float ]
+            distance     float
+            fov          [ float float ]
+            hpr          [ float float float ]
+    }
+
+    segment {...}   # 1...n times
+}
+\end{lstlisting}}
+
+\subsubsection{Segment Section}
+
+A segment represents a single display, i.e., a projector or monitor. It
+references a channel, has a name, viewport and frustum. The name of a
+segment can be used for debugging, it has no influence on the execution
+of Equalizer. It can be used to reference the segment, in which case the
+name should be unique.
+
+The channel referenced by the segment defines the output channel. The
+viewport of the section defines the 2D area covered by the channel on
+the canvas. Segments can overlap each other, e.g., when edge-blended
+projectors or passive stereo is used.
+
+A segment can define a frustum, in which case it overrides the default
+frustum calculated from the canvas frustum and segment viewport.
+
+{\footnotesize\begin{lstlisting}
+segment # 1...n times
+{
+    channel  string
+    name     string
+    viewport [ viewport ]
+
+    wall       // frustum description
+    {
+        bottom_left  [ float float float ]
+        bottom_right [ float float float ]
+        top_left     [ float float float ]
+        type         fixed | HMD
+    }
+    projection // alternate frustum description, last one wins
+    {
+        origin       [ float float float ]
+        distance     float
+        fov          [ float float ]
+        hpr          [ float float float ]
     }
@@
 The name of the compound is used for the default names of swap barriers
-and output frames. The channel references the first channel of the same
-name in the configuration.
+and output frames.
+
+A channel reference is either the name of the channel in the resource
+section if no canvases are used, or the destination channel reference of
+a view/segment intersection. Channel segment references are delimited by
+braces, in which the canvas, segment, layout and view describing the
+channel are named, i.e, 'channel ( canvas "PowerWall" segment 0 layout
+"Simple" view 0 )'.
 
@@
 The buffer defines the default frame buffer attachments read back by
-output frames. The viewport restricts the rendering to the area relative
-to the parent compound. The range restricts the database range, relative
-to the parent. The pixel setting selects the pixel decomposition kernel,
+output frames. Output frames may change the buffer attachments used.
+
+The viewport restricts the rendering to the area relative to the parent
+compound. The range restricts the database range, relative to the
+parent. The pixel setting selects the pixel decomposition kernel,
 relative to the parent. The eye pass select the mono, left or right eye
-passes rendered by the compound. The DPlex period defines that
-$\frac{1}{period}$ frames are rendered, and the phase defines when in the period
-the rendering starts. All these attributes are inherited by the children
-of a compound. Viewport, range, pixel and period parameters are
-cumulative.
-
-A load-balancer is used to automatically optimize the decomposition. A
-2D, horizontal or vertical load-balancer adapts the viewport of all
-direct children of the compound each frame. A DB load-balancer adapts
-the range of all direct children. A DPlex load-balancer smoothens the
-framerate of the compound's window to produce a steady output framerate.
+passes rendered by the compound. The zoom scales the parent pixel
+viewport resolution. The DPlex period defines that $\frac{1}{period}$
+frames are rendered, and the phase defines when in the period the
+rendering starts. All these attributes are inherited by the children of
+a compound. Viewport, range, pixel and period parameters are cumulative.
+
+Equalizers are used to automatically optimize the decomposition. A 2D,
+horizontal or vertical load equalizer adapts the viewport of all direct
+children of the compound each frame. A DB load equalizer adapts the
+range of all direct children. A dynamic framerate (DFR) equalizer adapts
+the zoom for a constant framerate. A framerate equalizer
+smoothens the framerate of the compound's window to produce a steady
+output framerate, typically for DPlex compounds. A monitor equalizer
+adapts the output image zoom to monitor another canvas.
 
@@
 compounds. All swap barriers of the same name are synchronized The
-default name is \textsf{barrier[.root\-Com\-pound\-Name]}.
+default name is \textsf{barrier[.root\-Com\-pound\-Name]}. All windows
+using the same \textsf{NV\_group} on a single system are synchronized
+with each other using hardware synchronization. All groups using the
+same \textsf{NV\_barrier} across systems are synchronized with each
+other using hardware synchronization. When using hardware
+synchronization, the barrier name is ignored.
 
@@
 \textsf{frame.channelName}. The frame buffer attachments to read back
-are inherited from the compound, but can be overridden by output frames.
+are inherited from the compound, but can be overridden by output
+frames. Frames of type \textsf{texture} copy the framebuffer contents to
+a texture, and can only be used to composite frames between windows of
+the same pipe.
 
@@
     name     string
-    channel  string   // where the compound's tasks are executed
-    task     [ CLEAR DRAW READBACK ASSEMBLE ] // CULL later
-    buffer   [ COLOR DEPTH ]         // default COLOR
-    viewport [ viewport ]            // wrt parent compound, sort-first
-    range    [ float float ]         // DB-range for sort-last
-    pixel    [ int int ]             // pixel decomposition (step size)
-    eye      [ CYCLOP LEFT RIGHT ]   // monoscopic or stereo view
-    period   int                     // DPlex period, every nth frame rendered
-    phase    int                     // DPlex phase, when to start rendering
-
-    loadBalancer
-    {
-        mode   2D | HORIZONTAL | VERTICAL | DB | DPLEX
-    }
+    channel  channel-ref  # where the compound tasks are executed
+
+    task     [ CLEAR DRAW READBACK ASSEMBLE ]
+    buffer   [ COLOR DEPTH ]         # default COLOR
+
+    viewport [ viewport ]            # wrt parent compound, sort-first
+    range    [ float float ]         # DB-range for sort-last
+    pixel    [ int int ]             # pixel decomposition (step size)
+    eye      [ CYCLOP LEFT RIGHT ]   # monoscopic or stereo view
+    zoom     [ float float ]         # up/downscale of parent pvp
+    period   int                     # DPlex period, every nth frame rendered
+    phase    int                     # DPlex phase, when to start rendering
+
+    load_equalizer # adapt 2D tiling or DB range of children
+    {
+        mode 2D | DB | VERTICAL | HORIZONTAL
+        damping float # 0: no damping, 1: no changes
+    }
+    DFR_equalizer # adapt ZOOM to achieve constant framerate
+    {
+        framerate float # target framerate
+        damping float   # 0: no damping, 1: no changes
+    }
+    framerate_equalizer {} # smoothen window swapbuffer rate (DPlex)
+    monitor_equalizer {}   # set frame zoom when monitoring other views
 
@@
     {
-        stereo_mode                  QUAD | ANAGLYPH    // default QUAD
-        stereo_anaglyph_left_mask    [ RED GREEN BLUE ] // default red
-        stereo_anaglyph_right_mask   [ RED GREEN BLUE ] // df green blue
-        update_FOV                   HORIZONTAL [ON] | VERTICAL | OFF
-    }
-
-    wall       // frustum description
+        stereo_mode                  QUAD | ANAGLYPH    # default QUAD
+        stereo_anaglyph_left_mask    [ RED GREEN BLUE ] # default red
+        stereo_anaglyph_right_mask   [ RED GREEN BLUE ] # df green blue
+    }
+
+    wall       # frustum description
     {
@@
         top_left     [ float float float ]
-    }
-    projection // alternate frustum description, last one wins
+        type         fixed | HMD
+    }
+    projection # alternate frustum description, last one wins
     {
@@
 
-    swapbarrier  // compounds with the same barriername sync swap
+    swapbarrier  # compounds with the same barriername sync swap
     {
         name string
+        NV_group OFF | ON | unsigned
+        NV_barrier OFF | ON | unsigned
     }
@@
         buffer [ COLOR DEPTH ]
+        type texture | memory
     }
@@
     {
-        name string // corresponding output frame
-    }
-}
+        name string # corresponding output frame
+    }
+}
+
+channel-ref:  'string' | '(' channel-segment-ref ')'
+channel-segment-ref: ( canvas-ref ) segment-ref ( layout-ref ) view-ref
+canvas-ref:   'string' | 'index'
+segment-ref:  'string' | 'index'
+layout-ref:   'string' | 'index'
+view-ref:     'string' | 'index'
+observer-ref: 'string' | 'index'
 \end{lstlisting}}

### 2009-07-28 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] Will break Win32 build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-\title{Equalizer Programming Guide}
+\title{Equalizer\\Programming and User Guide}
 \author{Eyescale Software GmbH}
@@
 \date{
-  \vspace{15cm}
-  {\Large\sffamily The official reference for creating parallel,
-    scalable OpenGL\texttrademark\ applications with the Equalizer
-    parallel rendering framework}\\\vspace{1cm}
-  Version 1.5.9 for Equalizer 0.6+svn\\[\medskipamount]
+  \vspace{14cm}
+\begin{textblock}{}(0,4.5)
+  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
+\end{textblock} {\Large The official reference for developing and
+  deploying parallel, scalable OpenGL\texttrademark\ applications based on
+  the Equalizer parallel rendering framework}\\\vspace{1cm}
+Version 1.5.10 for Equalizer 0.6+svn\\[\medskipamount]
 %  \today
@@
 \thispagestyle{empty}
-\begin{textblock}{}(0,4.5)
-  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
-\end{textblock}
-
-\clearpage
+
+%\clearpage
 
 \lstset{language=C++}
-\section*{Equalizer 0.6 Programming Guide}
+\section*{Equalizer 0.6 Programming and User Guide}
 \today
@@
     1.5.8   & May 30, 2009 & Added \sref{sEqualizers}, updated \sref{sLoadBalancing} and \sref{sSwapBarrier}\\
-    1.5.9.  & Jul 27, 2009 & Major review pass before release\\
+    1.5.9   & Jul 27, 2009 & Major review pass before release\\
+    1.5.10  & Jul 28, 2009 & Split in User and Programming Guide, added \sref{sNamespaces}\\
     \hline
@@
 
+\part{User Guide}
 \section{Introduction}
 
-Equalizer provides a framework for the development of parallel OpenGL
-applications. Equalizer-based applications can run from a single
-shared-memory system with one or multiple graphics cards up to
-large-scale graphics clusters. This Programming Guide introduces the
-programming interface, often using the Equalizer example \textsf{eqPly}
-as a guideline.
-
-Equalizer is used for scientific simulation visualization at various
-research institutes, for creating electronic art exhibitions as well as
-in commercial 3D design applications.
-
-Equalizer is the next step in the evolution of generic parallel programming
-interfaces for OpenGL-based visualization applications. Existing
-solutions, such as OpenGL Multipipe SDK, Cavelib and VRJuggler,
-implement a subset of concepts similar to Equalizer. In other areas,
-e.g., tracking device support, they provide more functionality.
-
-To adapt an application for Equalizer, the programmer structures the
-source code so that the OpenGL rendering can be executed in parallel,
-potentially using multiple processes for cluster-based execution.
+Equalizer is the standard middleware for the development and deployment
+of parallel OpenGL applications. It enables applications to benefit from
+multiple graphics cards, processors and computers to scale rendering
+performance, visual quality and display size. An Equalizer-based
+application runs unmodified on any visualization system, from a simple
+workstation to large scale graphics clusters, multi-GPU workstations and
+Virtual Reality installations.
+
+This User and Programming Guide introduces parallel rendering concepts,
+the configuration of Equalizer-based applications and programming using
+the Equalizer parallel rendering framework.
+
+Equalizer is the next step in the evolution of middleware for scalable
+OpenGL applications. Existing solutions, such as OpenGL Multipipe SDK,
+Cavelib and VRJuggler, implement a subset of Equalizer. In other areas,
+e.g., tracking device support, some of them provide more functionality.
 
@@
 
-If you have any question regarding Equalizer programming, this
-programming guide, or other specific problems you encountered, please
-direct them to the \textsf{eq-dev} mailing
+If you have any question regarding Equalizer programming, this guide, or
+other specific problems you encountered, please direct them to the
+\textsf{eq-dev} mailing
 list\footnote{\link{http://www.equalizergraphics.com/lists.html}}.
@@
 
-\section{Hello, World!}
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eqHello.png}
-  {\caption{\label{fHello}Hello, World!}}
-\end{wrapfigure}
-The \textsf{eqHello} example is a minimal application to illustrate the
-basic principle of an Equalizer application: The application developer
-has to implement the rendering method \textsf{Channel::frameDraw},
-similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
-run as a stand-alone application from the command line.
-
-The \textsf{eqHello} redraw function renders six colored quads, rotating
-around the origin. The \textsf{frameDraw} meth\-od provided by the
-\textsf{eq::Channel} can be used as a convenience function to setup the
-frustum and other OpenGL state. After setting up some lighting
-parameters, \textsf{eqHello} rotates the scene and renders the quads
-using immediate mode:
-
-{\footnotesize\begin{lstlisting}
-void Channel::frameDraw( const uint32_t spin )
-{
-    // setup OpenGL State
-    eq::Channel::frameDraw( spin );
-
-    const float lightPos[] = { 0.0f, 0.0f, 1.0f, 0.0f };
-    glLightfv( GL_LIGHT0, GL_POSITION, lightPos );
-
-    const float lightAmbient[] = { 0.2f, 0.2f, 0.2f, 1.0f };
-    glLightfv( GL_LIGHT0, GL_AMBIENT, lightAmbient );
-
-    // rotate scene around the origin
-    glRotatef( static_cast< float >( spin ) * 0.5f, 1.0f, 0.5f, 0.25f );
-
-    // render six axis-aligned colored quads around the origin
-    [...]
-}
-\end{lstlisting}}
-
-The \textsf{eqHello} main function sets up the communication with the
-server, initializes and drives the rendering. The details of this setup
-are explained in \sref{sEqPly}.
-
 \section{Scalable Rendering}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
-  {\caption{\label{fStereoCmp}\small A Stereo compound}}
-\end{wrapfigure}
 Due to the frame consistency between the eye views, this modes scales
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
+  {\caption{\label{fStereoCmp}\small A stereo compound}}
+\end{wrapfigure}
 Eye compounds in Equalizer are configured using the \textsf{eye}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
-  {\caption{\label{fPixel}\small A pixel compound}}
-  \vspace{-6ex}
-\end{wrapfigure}
 Pixel compounds are similar to 2D compounds. The frusta of the
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
+  {\caption{\label{fPixel}\small A pixel compound}}
+\end{wrapfigure}
 OpenGL functionality influenced by the raster position will not work
@@
 
-Equalizer are described in more detail in \sref{sEqualizers}.
+Equalizers are described in more detail in \sref{sEqualizers}.
 
@@
 
+\part{Programming Guide}
+
+To adapt an application for Equalizer, the programmer structures the
+source code so that the OpenGL rendering can be executed in parallel,
+potentially using multiple processes for cluster-based execution.
+
+\section{Hello, World!}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eqHello.png}
+  {\caption{\label{fHello}Hello, World!}}
+\end{wrapfigure}
+The \textsf{eqHello} example is a minimal application to illustrate the
+basic principle of an Equalizer application: The application developer
+has to implement the rendering method \textsf{Channel::frameDraw},
+similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
+run as a stand-alone application from the command line.
+
+The \textsf{eqHello} redraw function renders six colored quads, rotating
+around the origin. The \textsf{frameDraw} meth\-od provided by the
+\textsf{eq::Channel} can be used as a convenience function to setup the
+frustum and other OpenGL state. After setting up some lighting
+parameters, \textsf{eqHello} rotates the scene and renders the quads
+using immediate mode:
+
+{\footnotesize\begin{lstlisting}
+void Channel::frameDraw( const uint32_t spin )
+{
+    // setup OpenGL State
+    eq::Channel::frameDraw( spin );
+
+    const float lightPos[] = { 0.0f, 0.0f, 1.0f, 0.0f };
+    glLightfv( GL_LIGHT0, GL_POSITION, lightPos );
+
+    const float lightAmbient[] = { 0.2f, 0.2f, 0.2f, 1.0f };
+    glLightfv( GL_LIGHT0, GL_AMBIENT, lightAmbient );
+
+    // rotate scene around the origin
+    glRotatef( static_cast< float >( spin ) * 0.5f, 1.0f, 0.5f, 0.25f );
+
+    // render six axis-aligned colored quads around the origin
+    [...]
+}
+\end{lstlisting}}
+
+The \textsf{eqHello} main function sets up the communication with the
+server, initializes and drives the rendering. The details of this setup
+are explained in \sref{sEqPly}.
+
 \section{The Programming Interface}
@@
 render clients.
+
+\subsection{\label{sNamespaces}Namespaces}
+
+Application developers are exposed to the following namespaces:
+
+\begin{description}
+\item[eq::base] The eq::base namespace provides C++ classes to abstract
+  the underlying operating system and implements common helper
+  functionality. One example is the \textsf{eq::base::Clock} class
+  providing a high-resolution timer. Classes in this namespace are well
+  documented on the API documentation on the Equalizer website, and are
+  not subject of this Programming Guide.
+\item[eq::net] The Equalizer network layer provides basic functionality
+  for network communication, such as \textsf{Connection} and
+  \textsf{ConnectionSet}, as well as higher-level functionality such as
+  \textsf{Node}, \textsf{Session} and \textsf{Object}. Please refer to
+  \sref{sNetwork} for an introduction into the network layer, and to
+  \sref{sNetObject} for distributed objects.
+\item[eq] The core namespace for the Equalizer client library. The class
+  hierarchy in this namespace closely models the configuration file
+  format. The classes in the \textsf{eq} namespace are the main subject
+  of this Programming Guide.
+\end{description}
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/namespaces}
+  {\caption{\label{fNamespaces}Namespace layering for eqPly and eqServer}}
+\end{wrapfigure}
+Examples shipped with Equalizer are implemented in their own namespaces,
+e.g., \textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing
+classes from the \textsf{eq} namespace, with the occasional usage of
+functionality from the \textsf{eq::net} and \textsf{eq::base} namespace.
+
+The functionality of the Equalizer server is implemented in the
+\textsf{eq::server} namespace. This namespace does not yet expose an API
+intended to be used by developers.
+
+\fig{fNamespaces} shows the namespace layering for \textsf{eqPly} and
+\textsf{eqServer}.
 
@@
 %\begin{figure}[h!t]
-  \vspace{-2ex}
   \subfigure[]{\label{fObjectDist1}
@@
       proxies or multiple inheritance}}
-  \vspace{-4ex}
 \end{wrapfigure}
@@
 For head-mounted displays (HMD), the tracking information is used to
-move the frusta with the observer, as shown in \fig{fHMF}. This results
+move the frusta with the observer, as shown in \fig{fHMD}. This results
 in different projections compared to normal tracking with fixed

### 2009-07-31 Developer/ProgrammingGuide/paper.tex
Release Candidate of Programming Guide
[ ] Will break Win32 build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.5.10 for Equalizer 0.6+svn\\[\medskipamount]
+Version 1.6 RC1 for Equalizer 0.9\\[\medskipamount]
 %  \today
@@
 \lstset{language=C++}
-\section*{Equalizer 0.6 Programming and User Guide}
+\section*{Equalizer 0.9 Programming and User Guide}
 \today
@@
 
-OpenGL is a registered trademark, OpenGL Multipipe is a trademark of
-Silicon Graphics, Inc. Linux is a registered trademark of Linus
-Torvalds.  Mac OS is a trademark of Apple Inc. CAVELib is a registered
-trademark of the University of Illinois. The CAVE is a registered
-trademark of the Board of Trustees of the University of Illinois at
-Chicago. Qt is a registered trademark of Trolltech. All other
-trademarks and copyrights herein are the property of their respective
-owners.
+OpenGL is a registered Trademark, OpenGL Multipipe is a Trademark of
+Silicon Graphics, Inc. Linux is a registered Trademark of Linus
+Torvalds.  Mac OS is a Trademark of Apple Inc. CAVELib is a registered
+Trademark of the University of Illinois. The CAVE is a registered
+Trademark of the Board of Trustees of the University of Illinois at
+Chicago. Qt is a registered Trademark of Trolltech. TripleHead2Go is a
+Trademark of Matrox Graphics. PowerWall is a Trademark of Mechdyne
+Corporation. All other trademarks and copyrights herein are the property
+of their respective owners.
 
@@
     1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
-    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
-    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
-    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
-    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
-    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
-    1.5.6   & Mar 23, 2009 & Documented Layout API\\
-    1.5.7   & Apr 17, 2009 & Updated Layout API (multiple observers)\\
-    1.5.8   & May 30, 2009 & Added \sref{sEqualizers}, updated \sref{sLoadBalancing} and \sref{sSwapBarrier}\\
-    1.5.9   & Jul 27, 2009 & Major review pass before release\\
-    1.5.10  & Jul 28, 2009 & Split in User and Programming Guide, added \sref{sNamespaces}\\
+    1.6 RC1 & Jul 31, 2009 & Release Candidate 1 for Equalizer 0.9\\
     \hline
@@
 
-Equalizer is the next step in the evolution of middleware for scalable
-OpenGL applications. Existing solutions, such as OpenGL Multipipe SDK,
-Cavelib and VRJuggler, implement a subset of Equalizer. In other areas,
-e.g., tracking device support, some of them provide more functionality.
+Equalizer is the most advanced middleware for scalable 3D visualization,
+providing the broadest set of parallel rendering features available in
+an open source library to any OpenGL application. Many commercial and
+open source applications in a variety of different markets rely on
+Equalizer for flexibility and scalability.
 
@@
   {\caption{\label{fExecutionFlow}Parallel Rendering}}
-  \vspace{-3ex}
+  \vspace{-4ex}
 \end{wrapfigure}
@@
 \fig{fExecutionFlow} illustrates the basic principle of any parallel
-rendering application. The typical OpenGL application, for example
-GLUT-based applications, has an event loop which redraws the scene,
-updates data based on received events, and eventually redraws a new
-frame.
+rendering application. The typical OpenGL application, for example GLUT,
+has an event loop which redraws the scene, updates data based on
+received events, and eventually redraws a new frame.
 
@@
 loop. The rendering code is then executed in parallel on different
-resources. This model is naturally followed by Equalizer, thus making
-application development as easy as possible.
+resources, depending on the configuration chosen at runtime.
+
+This model is naturally followed by Equalizer, thus making application
+development as easy as possible.
 
@@
 programmer, and provides communication between the different Equalizer
-processes.
+processes. The underlying network layer uses a peer-to-peer structure,
+allowing render clients to communicate directly with each other, e.g.,
+during parallel compositing.
 
@@
 execution is passive, it has no main loop and is completely driven by
-Equalizer, based on the rendering tasks received from the server. The
-tasks are executed by calling the appropriate task methods (see
-\sref{sTaskMethods}) in the correct thread and context. The application
-either implements the task methods with application-specific code or
-uses the default methods provided by Equalizer.
+Equalizer. It executes the rendering tasks received from the server by
+calling the appropriate task methods (see \sref{sTaskMethods}) in the
+correct thread and context. The application either implements the task
+methods with application-specific code or uses the default methods
+provided by Equalizer.
 
@@
 client code, it is reduced to be the application's `master' process
-without any OpenGL windows and rendering code.
+without any OpenGL windows and 3D rendering.
 
@@
 
-This section gives an introduction to scalable rendering, to
-provide some background for application developers. The scalability
+This section gives an introduction to scalable rendering, providing some
+background for end users and application developers. The scalability
 modes offered by Equalizer are discussed, along with their advantages
@@
 possible. In most cases, a \textsf{loadBalancer} should be used to
-automatically adapt the tiling each frame, based on the current
+automatically adjust the tiling each frame, based on the current
 rendering load.
@@
 glBitmap, glPolygonStipple. The application can query the current pixel
-parameters at run-time to adjust the rendering accordingly.
+parameters at runtime to adjust the rendering accordingly.
 
@@
 Techniques like frustum culling do not reduce the rendered data for the
-source rendering resources. Pixel compounds are ideal for raytracing,
+source rendering resources. Pixel compounds are ideal for ray-tracing,
 which is highly fill-limited and needs the full database for rendering
@@
 
-Some scalable rendering parameters are updated at run-time to provide
+Some scalable rendering parameters are updated at runtime to provide
 optimal performance or influence other scalable rendering
@@
 The server provides a number of equalizers, which automatically update
-certain parameters of compounds based on run-time information. They
+certain parameters of compounds based on runtime information. They
 balance the load of 2D and DB compounds, optimally distribute render
-resources for multi-segment projections systems, adapt the resolution to
-provide a constant framerate or zoom images to allow monitoring of
+resources for segment projections systems, adjust the resolution
+to provide a constant framerate or zoom images to allow monitoring of
 another view.
@@
 \begin{itemize}
-\item A list of all \textbf{computers} in you rendering cluster,
+\item A list of all \textbf{computers} in your rendering cluster,
   including the \textbf{IP addresses} of all network interfaces to be
@@
   applicable. These are typically the bottom-left, bottom-right and
-  top-left corner points of each display surfaces in meters.
+  top-left corner points of each display surface in meters.
 \item The relative coordinates of all the \textbf{segments} belonging to
@@
   windows are typically destination windows for scalable rendering or
-  windows paired with a view on a display system.
+  'control' windows paired with a view on a display system.
 \item \textbf{Characteristics} of the application, e.g., supported
@@
 
-Equalizer applications are configured at run-time by the Equalizer
-server. The server loads its configuration from a text file, which is an
+Equalizer applications are configured at runtime by the Equalizer
+server. The server loads its configuration from a text file, which is a
 one-to-one representation of the configuration data structures at
@@
 
-In each node, list all \textsf{connection}s thorugh which this node is
+In each node, list all \textsf{connection}s through which this node is
 reachable. Typically a node uses only one connection, but it is possible
@@
 to use multiple, independent network interfaces. Make sure the
-\textsf{hostname} used is reachable from all nodes. An IP address may be
-used as the hostname.
+configured \textsf{hostname} is reachable from all nodes. An IP address
+may be used as the hostname.
 
@@
 operating systems using X11, e.g., Linux, also set the port number if
-your X-Server is running on a non-standard port.
+your X-Server is running on a nonstandard port.
 
@@
 thread-unsafe libraries, but have various performance caveats. They
-should only be used if using a different, but synchronized rendering
-thread is not an option.
+should only be used if using a different, synchronized rendering thread
+is not an option.
 
@@
 segment. Configure one off-screen window, typically a pbuffer, for each
-graphics card used as a source for scalable rendering.
+graphics card used as a source for scalable rendering. Provide a useful
+name to each on-screen window if you want to easily identify it at runtime.
 
 Sometimes display segments cover only a part of the graphics card
-output. In this case it is adviced to configure a non-fullscreen window
+output. In this case it is advised to configure a non-fullscreen window
 without window decorations, using the correct window viewport.
@@
 
-Configure one \textsf{channel} for each desired rendering in each window
-area. Typically one full-screen channel per window is used. Name the
-channel using a unique, easily-identifiable name.
+Configure one \textsf{channel} for each desired rendering area in each
+window. Typically one full-screen channel per window is used. Name the
+channel using a unique, easily identifiable name, e.g., 'source-1',
+'control-2' or 'segment-2\_3'.
 
@@
 Sometimes, a single window is split across multiple projectors, e.g., by
-using an external splitter such as the Matrox triple-head-to-go. In this
+using an external splitter such as the Matrox TripleHead2Go. In this
 case configure one channel for each segment, using the channel's
-\textsf{viewport} to configure its position wit respect to the window.
+\textsf{viewport} to configure its position relative to the window.
 
@@
 
-A canvas represents one logical projection surface, e.g., a powerwall, a
+A canvas represents one logical projection surface, e.g., a PowerWall, a
 curved screen or an immersive installation. One configuration might
@@
 segments. Non-planar frusta are configured using the segment
-frusta. These frusta typically describe a physically-correct display
+frusta. These frusta typically describe a physically correct display
 setup for Virtual Reality installations.
@@
 
+Configure one \textsf{segment} for each display or projector of each
+canvas. Configure the \textsf{viewport} of the segment to match the area
+covered by the segment on the physical canvas. Set the output
+\textsf{channel} to the resource driving the described projector.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 
-Configure one \textsf{segment} for each display or projector of each
-canvas. Configure the \textsf{viewport} of the segment to match the area
-covered by the segment on the physical canvas. Set the output
-\textsf{channel} to the resource driving the described projector.
-
 For non-planar displays, configure the frustum as described in
 \sref{sCanvas}. For passive stereo installations, configure one segment
-per eye pass, where the segment for the left and right eye fully
-have the same viewport.
+per eye pass, where the segment for the left and right eye have the same
+viewport.
 
 A segment represents one output channel of the canvas, e.g., a projector
-or a LCD. A segment has an output channel, which references the channel to
+or an LCD. A segment has an output channel, which references the channel to
 which the display device is connected.
@@
 respect to the canvas. Segments might overlap (edge-blended projectors)
-or have gaps between each other (\fig{fCanvas}\footnote{Dataset courtesy
-  of VolVis distribution of SUNY Stony Brook, NY, USA.}). The viewport
-is used to configure the segment's default frustum from the canvas
-frustum description, and to place layout views correctly.
+or have gaps between each other (display walls,
+\fig{fCanvas}\footnote{Dataset courtesy of VolVis distribution of SUNY
+  Stony Brook, NY, USA.}). The viewport is used to configure the
+segment's default frustum from the canvas frustum description, and to
+place layout views correctly.
 
 \subsection{\label{sLayout}Layouts}
+
+Configure one \textsf{layout} for each configuration of logical
+views. Name the layout using a unique name. Often only one layout with a
+one view is used for all canvases.
 
@@
 \end{wrapfigure}
-
-Configure one \textsf{layout} for each configuration of logical
-views. Name the layout using a unique name. Often only one layout with a
-one view is used for all canvases.
 
@@
 
-A view has a fractional viewport with respect to its layout.  A layout
+A view has a fractional viewport relative to its layout.  A layout
 is often fully covered by its views, but this is not a requirement.
@@
 combined to produce the desired output, and how they are aggregated to
-increase the performance. It is adviced to study and understand the
-basic configuration files shipped with the Equalizer configuration,
-before attempting to write compound configurations.
+increase the performance.
+
+It is advised to study and understand the basic configuration files
+shipped with the Equalizer configuration, before attempting to write
+compound configurations. The command line program \textsf{configTool},
+shipped with the Equalizer distribution, creates some standard
+configurations automatically.
 
 For configurations using canvases and layouts without scalability, the
-configuration file loader will create the appropriate compounds.
-
-Usage of the rendering resources is configured using a compound
-tree. Although the API does not currently expose compounds, the basic
-design behind compounds is explained here for a better understanding of
-the Equalizer architecture. Further information on the configuration of
-compounds can be found on the Equalizer website\footnote{see
-  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.
+configuration file loader will create the appropriate compounds. It is
+typically not necessary to write compounds for this use case.
+
+The following subsection outlines the basic approach to writing
+compounds. The remaining subsections provide an in-depth explanation of
+the compound structure to give the necessary background for compound
+configuration.
+
+\subsubsection{Writing Compounds}
+The following steps are typically taken when writing compound
+configurations:
+\begin{description}
+\item [Root compound] Define an empty top-level compound when
+  synchronizing multiple destination views. Multiple destination views
+  are used for multi-display systems, e.g., a PowerWall or CAVE. All
+  windows used for one display surface should be swap-locked (see below)
+  to provide a seamless image. A single destination view is typically
+  used for providing scalability to a single workstation window.
+\item [Destination compound(s)] Define one compound for each destination
+  channel, either as a child of the empty group, or as a top-level
+  compound.
+  \begin{description}
+  \item[Standalone channels] Set the channel by using the channel name
+    from the resource section. Define a frustum as explained in
+    \sref{sCanvas}.
+  \item[View/Segment channels] Set the channel by using the canvas,
+    segment, layout and view name. The compound frustum will be
+    calculated automatically based on the segment or view frustum. Note
+    that one segment may created multiple view/segment channels, one for
+    each view intersection of each layout used on the canvas. Only the
+    compounds belonging to the active layout of a canvas are activate at
+    runtime.
+    \begin{description}
+    \item [Software swap lock] When using software swap synchronization,
+      swap-lock the destination compounds using a swapbarrier. All
+      windows with a swapbarrier of the same name synchronize their
+      swapbuffers.
+    \item [Hardware swap lock] When using hardware swap synchronization,
+      use swapbarriers for the destination compounds, setting
+      \textsf{NV\_group} and \textsf{NV\_barrier} appropriately. The
+      swap barrier name is ignored in this case. All windows of the same
+      group on a single node synchronize their swap buffer. All groups
+      of the same barrier synchronize their swap buffer across
+      nodes. Please note that the driver typically limits the number of
+      groups and barriers to one, and that multiple windows per swap
+      group are not supported by all drivers. Please refer to your
+      OpenGL driver documentation for details.
+    \end{description}
+  \end{description}
+\item[Scalability] If desired, define scalability for each of your
+  destination compounds. Add one compound using a source channel for
+  each contributor to the rendering. The destination channel may also be
+  used as a source. 
+  \begin{description}
+  \item[Decomposition] On each child compound, limit the rendering task
+    of that child by setting the \textsf{viewport}, \textsf{range},
+    \textsf{period} and \textsf{phase}, \textsf{pixel}, \textsf{eye} or
+    \textsf{zoom} as desired.
+  \item[Runtime Adjustments] A \textsf{load\_equalizer} may be used on
+    the destination compounds to set the \textsf{viewport} or
+    \textsf{range} of all children each frame, based on the current
+    load. A \textsf{view\_equalizer} may be used on the root compound to
+    assign resources to all destination compounds, which have to use
+    \textsf{load\_equalizer}s. A \textsf{framerate\_equalizer} should be
+    used to smoothen the framerate of DPlex compounds. A
+    \textsf{DFR\_equalizer} may be used to set the zoom of a compound to
+    achieve a constant framerate. One compound may have multiple
+    equalizers, e.g., a \textsf{load\_equalizer} and a
+    \textsf{DFR\_equalizer} for a 2D compound with a constant framerate.
+  \item[Recomposition] For each source compound, define an
+    \textsf{output\_frame} to read back the result. Use this output frame
+    as an \textsf{input\_frame} on the destination compound. The frames
+    are connected with each other by their name, which has to be unique
+    within the root compound tree. For parallel compositing, describe
+    your algorithm by defining multiple input and output frames across
+    all source compounds.
+  \end{description}
+\end{description}
 
@@
 Compounds have a frustum description to define the physical layout of
-the display environment.
+the display environment. The frustum specification is described in
+\sref{sCanvas}.
 
@@
 frustum is typically defined on the topmost compound.
-
-The frustum can be specified as described in \sref{sCanvas}.
 
@@
 rendering pass can selectively render different stereo passes. A
-\textsf{pixel} parameter adapts the frustum so that the source channel
-renders an evenly subset of the parent's pixels. Setting one or multiple
+\textsf{pixel} parameter adjusts the frustum so that the source channel
+renders an even subset of the parent's pixels. Setting one or multiple
 attributes causes the parent's view to be decomposed
@@
 Note that a software swap barrier is not accurate enough for
-edge-blended projection system. Such installations typically use a
-hardware synchronization, e.g., nVidia G-Sync cards. It is however
-sufficient for display walls made out of LCD's and for some
-non-immersive projection systems.
+edge-blended projection system or immersive installations. Such
+installations typically use a hardware synchronization, e.g., nVidia
+G-Sync cards. It is however sufficient for display walls made out of
+LCD's and for some non-immersive projection systems.
 
@@
 
-\subsubsection{\label{sEqualizers}Adaptation - Equalizers}
-
-Equalizers are used to update compound parameters based on run-time
+\subsubsection{\label{sEqualizers}Adjustments - Equalizers}
+
+Equalizers are used to update compound parameters based on runtime
 data. They are attached to a compound (sub-)tree, on which they
@@
 
-Using a 2D or DB load-balancer will adapt the 2D split or database range
+Using a 2D or DB load-balancer will adjust the 2D split or database range
 automatically each frame. The 2D load-balancer exists in three flavors:
@@
 predict the load for their scene data, e.g., when the data is
-non-uniform. Volume rendering is a counterexample, where the data is
+nonuniform. Volume rendering is a counterexample, where the data is
 uniform and a static DB decomposition typically results in a better
@@
 assignment of resources to segments. The view equalizer analyzes the
-load of all segments, and adapts the resource usage each frame. It
+load of all segments, and adjusts the resource usage each frame. It
 equalizes the load on all segments of a view.
@@
   AG, 2008} illustrates this process. On the left side, a static
-assigment of resources to display segments is used. The right-hand
+assignment of resources to display segments is used. The right-hand
 segment has a higher load than the left-hand segment, causing
@@
 possible to improve the rendering performance of a multi-display system
-without any additional resources. This is particulary useful for
+without any additional resources. This is particularly useful for
 installations with a higher number of displays where the rendering load
@@
 \paragraph{DFR Equalizer}
+Dynamic Frame Resolution (DFR) trades rendering performance for visual
+quality. The rendering for a channel is done at a different resolution
+than the native channel resolution to keep the framerate constant. The
+DFR equalizer adjusts the zoom of a channel, based on the target and
+current framerate. It is typically used for fill-rate bound
+applications, such as volume rendering and ray-tracing.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-Dynamic Frame Resolution (DFR) trades rendering performance for visual
-quality. The rendering for a channel is done at a different resolution
-than the native channel resolution to keep the framerate constant. The
-DFR equalizer adapts the zoom of a channel, based on the target and
-current framerate. It is typically used for fill-rate bound
-applications, such as volume rendering and raytracing.
-
 \fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
@@
 
-The dynamic frame resolution is not limited to down-scaling the
-rendering resolution, it will also up-scale the resolution if the
-parameters allow for it. Upscaled rendering, which will downsample the
+The dynamic frame resolution is not limited to downscaling the
+rendering resolution, it will also upscale the resolution if the
+parameters allow for it. Upscaled rendering, which will down-sample the
 result for display, provides dynamic anti-aliasing at a constant
@@
 potentially made of multiple segments, in a different channel at a
-different resolution. This is typically used to re-use the rendering of
+different resolution. This is typically used to reuse the rendering of
 a large-scale display on an operator station.
@@
 \end{wrapfigure}
-The monitor equalizer adapts the frame zoom of the output frames used to
+A monitor equalizer adjusts the frame zoom of the output frames used to
 observe the rendering, depending on the destination channel size. The
@@
 
-To adapt an application for Equalizer, the programmer structures the
+To modify an application for Equalizer, the programmer structures the
 source code so that the OpenGL rendering can be executed in parallel,
@@
 \textsf{nounVerb}. For example, the application calls
-\textsf{Config::startFrame} to render a new frame, which causes --among
-other things-- \textsf{Node::frameStart} to be called in all active
+\textsf{Config::startFrame} to render a new frame, which causes, among
+many other things, \textsf{Node::frameStart} to be called in all active
 render clients.
@@
   functionality. One example is the \textsf{eq::base::Clock} class
-  providing a high-resolution timer. Classes in this namespace are well
-  documented on the API documentation on the Equalizer website, and are
+  providing a high-resolution timer. Classes in this namespace are fully
+  documented in the API documentation on the Equalizer website, and are
   not subject of this Programming Guide.
@@
   \sref{sNetObject} for distributed objects.
-\item[eq] The core namespace for the Equalizer client library. The class
-  hierarchy in this namespace closely models the configuration file
-  format. The classes in the \textsf{eq} namespace are the main subject
-  of this Programming Guide.
+\item[eq] The core namespace for the Equalizer client library. The
+  classes and their relationship in this namespace closely model the
+  configuration file format. The classes in the \textsf{eq} namespace
+  are the main subject of this Programming Guide.
 \end{description}
@@
 In the case of \textsf{eqPly}, the application is also the render
-client. The \textsf{eqPly} executable has three run-time behaviors:
+client. The \textsf{eqPly} executable has three runtime behaviors:
 
@@
 executable is used as a render client, \textsf{Client::initLocal} never
-returns. Therefore the application's main loop is never executed. In
-order to keep the client resident, the \textsf{eqPly} example overrides
-the client loop to keep it running beyond one configuration run:
+returns. Therefore the application's main loop is never executed. To
+keep the client resident, the \textsf{eqPly} example overrides the
+client loop to keep it running beyond one configuration run:
 
@@
 distributed objects is normally the \textsf{eq::Config}, which is
-subclassed from \textsf{eq::net::Session}.
+sub-classed from \textsf{eq::net::Session}.
 
@@
 \begin{description}
-  \item[Subclassing]: The classes to be distributed inherit from
-    \textsf{eq::net::Object} and implement the serialization
-      methods. This requires modification of the existing classes. This
-      is used for \textsf{eqPly::InitData} and
+\item[Subclassing] The classes to be distributed inherit from
+  \textsf{eq::net::Object} and implement the serialization methods. This
+  approach is recommended if the source code of existing classes can be
+  modified. It is used for \textsf{eqPly::InitData} and
   \textsf{eqPly::FrameData}. (\fig{fObjectDist1})
-  \item[Proxies]: For each object to be distributed, a proxy object is
+\item[Proxies] For each object to be distributed, a proxy object is
   created which manages data distribution for its associated
@@
   eqPly is using this pattern. (\fig{fObjectDist2})
-  \item[Multiple Inheritance]: A new class inheriting from the class to
-    be distributed and from \textsf{eq::net::Object} implements the data
-    distribution. This requires the application to instantiate a
-    different type of object instead of the existing object, and to
-    create wrapper methods in the superclass calling the original method
-    and setting the appropriate dirty bits. This pattern is not used in
+\item[Multiple Inheritance] A new class inheriting from the class to be
+  distributed and from \textsf{eq::net::Object} implements the data
+  distribution. This requires the application to instantiate a different
+  type of object instead of the existing object, and to create wrapper
+  methods in the superclass calling the original method and setting the
+  appropriate dirty flags. This pattern is not used in
   eqPly. (\fig{fObjectDist3})
@@
   the current version will fail.
-\item Newly mapped slave instance are mapped to the oldest available
+\item Newly mapped slave instances are mapped to the oldest available
   version by default, or to the version specified when calling
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
-  {\caption{\label{fUMLObject}eq::Object and eq::net::Object}}
-\end{wrapfigure}
 Not syncing a mapped, versioned object creates a memory leak. The method
@@
 application, which then takes the appropriate action. The default
-implementation asserts when too many versions have been queued in order
+implementation asserts when too many versions have been queued
 to detect memory leaks during development.
 
-In addition to the instance data (de-)serialization methods used to map
-an object, versioned objects may implement \textsf{pack} and
+Besides the instance data (de-)serialization methods used to map an
+object, versioned objects may implement \textsf{pack} and
 \textsf{unpack} to serialize or de-serialize the changes since the last
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
+  {\caption{\label{fUMLObject}eq::Object and eq::net::Object}}
+\end{wrapfigure}
 The creation of distributed, versioned objects is further facilitated by
@@
 
-The \textsf{FrameData} is subclassed from \textsf{eq::Object}, and
+The \textsf{FrameData} is sub-classed from \textsf{eq::Object}, and
 consequently tracks its changes by setting the appropriate dirty bit
@@
 
-The initialization sequence uses multiple, overrideable task
+The initialization sequence uses multiple, override-able task
 methods. The main task method \textsf{configInit} calls first
@@
 model is positioned using the camera parameters from the frame data. The
-camera parameters are transported using the the frame data to ensure
+camera parameters are transported using the frame data to ensure
 that all channels render a given frame using the same position.
@@
 Three different ways of coloring the object are possible: Using the
-colors of the mode, using an unique per-channel color to demonstrate the
+colors of the mode, using a unique per-channel color to demonstrate the
 decomposition as shown in \fig{fDBDest}, or using solid white for
-anaglyphics stereo. The model colors are per-vertex and are set during
+anaglyphic stereo. The model colors are per-vertex and are set during
 rendering, whereas the unique per-channel color is set in
@@
   data.
-\item[DRAW\_SYNC:] In addition to async thread model, all local render
-  threads are synchronized, so that the draw operations happen
-  synchronously with the node main loop. This model allows to use the
-  same database for rendering, and safe modifications of this database
-  are possible from the node thread, since the pipe threads do not
-  execute any rendering tasks between frames. This is the default
-  threading model. This threading model should be used by applications
-  which keep one copy of the scene graph per node.
-\item[LOCAL\_SYNC:] In addition to the async thread model, all local
-  frame operations, including readback, assemble and swap buffer are
-  synchronized with the node main loop. This threading model should be
-  used by applications which need to access non-buffered, frame-specific
-  data after rendering, e.g., during \textsf{Channel::frameAssemble}.
+\item[DRAW\_SYNC:] Additionaly to the synchronization of the async
+  thread model, all local render threads are synchronized, so that the
+  draw operations happen synchronously with the node main loop. This
+  model allows to use the same database for rendering, and safe
+  modifications of this database are possible from the node thread,
+  since the pipe threads do not execute any rendering tasks between
+  frames. This is the default threading model. This threading model
+  should be used by applications which keep one copy of the scene graph
+  per node.
+\item[LOCAL\_SYNC:] Additionaly to the synchronization of the async
+  thread model, all local frame operations, including readback, assemble
+  and swap buffer are synchronized with the node main loop. This
+  threading model should be used by applications which need to access
+  non-buffered, frame-specific data after rendering, e.g., during
+  \textsf{Channel::frameAssemble}.
 \end{description}
@@
 unlock all their children by calling \textsf{startFrame}, e.g.,
-\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
+\textsf{Node::frameStart} has to call \textsf{Node::startFrame}
 to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
@@
 \textsf{Config} instance on the application node, where it can be
-queried using \textsf{getErrorMessage}.
+retrieved using \textsf{getErrorMessage}.
 
 The sample implementations \textsf{AGLWindow}, \textsf{GLXWindow} and
-\textsf{WGLWindow} all have similar, overrideable methods for all
+\textsf{WGLWindow} all have similar, override-able methods for all
 sub-tasks. This allows partial customization, without the need of
@@
 The Layout API provides an abstraction for render surfaces (Canvas and
-Segment) and the arrangment of rendering areas on them (Layout and
+Segment) and the arrangement of rendering areas on them (Layout and
 View). Its functionality has been described in \sref{sCanvas} and
@@
 views of the configuration. The render client has access to the current
-view in the channel task methods. The layout entities can be subclassed
+view in the channel task methods. The layout entities can be sub-classed
 using the \textsf{NodeFactory}. Currently the layout of a canvas and the
-frustum of a view can be changed at run-time.
+frustum of a view can be changed at runtime.
 
@@
 
-Layout API entities (Canvas, Segment, Layout, View) are subclassed like
+Layout API entities (Canvas, Segment, Layout, View) are sub-classed like
 all other Equalizer entities using the \textsf{NodeFactory}. Equalizer
@@
 application amends it by view-specific data. The view, like all other
-layout entities, is indirectly subclassed from \textsf{eq::Object} via
-\textsf{eq::Frustum}. The design and usage of \textsf{eq::Object} is
-explained in \sref{sEqObject}. In \textsf{eqPly}, the
-application-specific data is the model identifier. A new dirty bit for
-the model is defined, and the model identifier is serialized when
-requested:
+layout entities, is indirectly sub-classed from \textsf{eq::Object}
+through \textsf{eq::Frustum}. The design and usage of
+\textsf{eq::Object} is explained in \sref{sEqObject}.
+
+In \textsf{eqPly}, the application-specific data is the model
+identifier. A new dirty bit for the model is defined, and the model
+identifier is serialized when requested:
 
@@
 means the channels involved in the last layout on the canvas are
-deinitialized, that is \textsf{configExit} and
+de-initialized, that is \textsf{configExit} and
 \textsf{NodeFactory::releaseEntity} is called, and channels involved in
@@
 application process can set the wall or projection parameters at
-runtime.For a description of wall and projection parameters please refer
+runtime. For a description of wall and projection parameters please refer
 to \sref{sFrustum}. The new data will be effective for the next frame.
@@
 segments. The frustum of a view does not typically correspond to a
-physical projection, and can therefore be changed at run-time. The
+physical projection, and can therefore be changed at runtime. The
 frustum of a segment does however correspond to a physical projection,
@@
 Compression plugins allow the creation of runtime-loadable modules for
-image compression. Equalizer will search pre-defined directories during
+image compression. Equalizer will search predefined directories during
 \textsf{eq::init} for dynamic shared objects (DSO) containing compression
@@
 The interface to a compression DSO is a C API, which allows to maintain
-binary compatiblity across Equalizer versions. Furthermore, the
+binary compatibility across Equalizer versions. Furthermore, the
 definition of an interface facilitates the creation of new compression
@@
 tiling, depth-compositing for polygonal data or sorted, alpha-blended
-compositing for semi-transparent volumetric data. These operations
+compositing for semitransparent volumetric data. These operations
 composite either directly all images on the GPU, or use a CPU-based
@@
 Statistics are measured in milliseconds since the configuration was
-initialized. On each node, the global configuration clock is sychronized
+initialized. On each node, the global configuration clock is synchronized
 with the server clock. Each statistic event records the originator's
@@
 The X axis is the time. One pixel on the screen corresponds to a given
-time unit, here one millisecond per pixel. The scale is adapted dynamically
+time unit, here one millisecond per pixel. The scale is zoomed dynamically
 to powers-of-ten milliseconds to fit the statistics into the available
@@
 A connection is either closed, connected or listening. A closed
-connection can not be used for communications. A connected connection
+connection cannot be used for communications. A connected connection
 can be used to read or write data to the communication peer. A listening
@@
 The connection and connection set can be used by applications to
-implement other networking-related functionality, e.g., to communicate
+implement other network-related functionality, e.g., to communicate
 with a sound server on a different machine.
@@
 \textsf{eq::net::Node}, \textsf{eq::net::Session} and
-\textsf{eq::net::Object} is subclassed.
+\textsf{eq::net::Object} is sub-classed.
 
@@
 standard\footnote{\link{http://en.wikipedia.org/wiki/Universally\_Unique\_Identifier}}. This
-identifier is used to address nodes, e.g., in order to query connection
+identifier is used to address nodes, e.g., to query connection
 information to connect to the node. Nodes use connections to communicate
-with each other by sending packets subclassed from
+with each other by sending packets sub-classed from
 \textsf{eq::net::NodePacket}.
@@
 If \textsf{dispatchCommand} returns false, the packet will be
-redispatched at a later time. This is used if an object has not been
-mapped locally, and therefore the packet could not be dispatched.
+re-dispatched later. This is used if an object has not been mapped
+locally, and therefore the packet could not be dispatched.
 
@@
 \textsf{EQ\_ID\_ANY} (the default value), a command is dispatch to all
-object instance registered to the the session with the correct object
+object instance registered to the session with the correct object
 ID. Multiple object instance per node are used when different threads on
@@
 
-The \textsf{eq::Object} data distribution is centered around the concept
-of dirty bits. Dirty bits form a 64-bit mask which marks which parts of
-the object have to be distributed during the next commit.
+The \textsf{eq::Object} data distribution is based on the concept of
+dirty bits. Dirty bits form a 64-bit mask which marks which parts of the
+object have to be distributed during the next commit.
 
@@
 and resets the dirty bits, and calls \textsf{serialize} or
-\textsf{deserialize} with the bit mask specifiying which data has to be
+\textsf{deserialize} with the bit mask specifying which data has to be
 transmitted or received. During a commit or sync, the current dirty bits
@@
 
-In order to use \textsf{eq::Object}, the following steps have to be
-taken:
+To use \textsf{eq::Object}, the following steps have to be taken:
 
@@
   rendering. QUAD-buffered stereo uses the left and right buffers of a
-  stereo window (active stereo). Anaglyphics stereo uses
+  stereo window (active stereo). Anaglyphic stereo uses
   \textsf{glColorMask} to mask colors for individual eye passes, used in
@@
 is used to enforce a SDP connection. PIPE uses a named pipe for fast
-inter-process communication on Windows.
+interprocess communication on Windows.
 
@@
 all connection descriptions of the node are passed to the launched node
-process, which will caus it to bind to all provided descriptions.
+process, which will cause it to bind to all provided descriptions.
 
@@
 
-The layout assignment can be changed at run-time by the application. The
+The layout assignment can be changed at runtime by the application. The
 intersection between views and segments defines which output
@@
 frusta specified at the canvas or segment level. This is typically used
-for non-physically correct rendering, e.g., to compare two models
+for nonphysically correct rendering, e.g., to compare two models
 side-by-side. If the view does not specify a frustum, the corresponding
@@
 The layouts referenced by the canvas can be applied to it by the
-application at run-time. One layout can be referenced by multiple
+application at runtime. One layout can be referenced by multiple
 canvases. The first layout is the layout active by default.
@@
 
-Compounds are the basic datastructure describing the rendering
+Compounds are the basic data structure describing the rendering
 setup. They use channels for rendering. Please refer to
@@
 Equalizers are used to automatically optimize the decomposition. A 2D,
-horizontal or vertical load equalizer adapts the viewport of all direct
-children of the compound each frame. A DB load equalizer adapts the
-range of all direct children. A dynamic framerate (DFR) equalizer adapts
+horizontal or vertical load equalizer adjusts the viewport of all direct
+children of the compound each frame. A DB load equalizer adjusts the
+range of all direct children. A dynamic framerate (DFR) equalizer adjusts
 the zoom for a constant framerate. A framerate equalizer
@@
 output framerate, typically for DPlex compounds. A monitor equalizer
-adapts the output image zoom to monitor another canvas.
+adjusts the output image zoom to monitor another canvas.
 
@@
 
-    load_equalizer # adapt 2D tiling or DB range of children
+    load_equalizer # adjust 2D tiling or DB range of children
     {
@@
     }
-    DFR_equalizer # adapt ZOOM to achieve constant framerate
+    DFR_equalizer # adjust ZOOM to achieve constant framerate
     {

### 2009-08-08 Developer/ProgrammingGuide/paper.tex
Final version of 0.9 Programming and User Guide
[ ] Will break Win32 build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.6 RC1 for Equalizer 0.9\\[\medskipamount]
+Version 1.6 for Equalizer 0.9\\[\medskipamount]
 %  \today
@@
 Trademark of Matrox Graphics. PowerWall is a Trademark of Mechdyne
-Corporation. All other trademarks and copyrights herein are the property
-of their respective owners.
+Corporation. CUDA is a Trademark of NVIDIA Corporation. All other
+trademarks and copyrights herein are the property of their respective
+owners.
 
@@
 If you have comments about the content, accuracy or comprehensibility of
-this programming guide, please contact
+this Programming and User Guide, please contact
 \htmladdnormallink{eile@eyescale.ch}
@@
 The images on the front page show the following applications build using
-Equalizer: RTT DeltaGen\footnote{Image copyright Realtime Technology AG,
-  2008} [top left], a flow visualization application for climate
-research\footnote{Image courtesy of Computer Graphics and Multimedia
-  Systems, University of Siegen}, the \textsf{eVolve} volume
-renderer\footnote{Data set courtesy of General Electric, USA} [bottom
-left] the \textsf{eqPly} polygonal renderer in a three-sided CAVE
-[bottom center], and a six-node sort-last database decomposition with
-parallel direct-send recomposition\footnote{Data set courtesy of
-  Stanford University Computer Graphics Laboratory} [bottom right].
+Equalizer: 
+RTT DeltaGen\footnote{Image copyright Realtime Technology AG, 2008} [top
+right], 
+\textsf{eqRaster}, a scalable terrain visualization\footnote{Image
+  courtesy of Visualization and Multimedia Lab, University of Zurich}
+[middle center],
+a flow visualization application for climate research\footnote{Image
+  courtesy of Computer Graphics and Multimedia Systems, University of
+  Siegen} [middle right],
+the \textsf{eqPly} polygonal renderer in a three-sided CAVE [bottom
+left],
+the \textsf{eVolve} volume renderer\footnote{Data set courtesy of
+  General Electric, USA} [bottom center],
+and the \textsf{eqNBody} gravitational n-body simulation, based on the
+N-Body CUDA example [bottom right].
 
 The images on the back page show the following scalable rendering modes:
-2D (sort-first, SFR), DB (sort-last), stereo, pixel and DPlex
-(time-multiplex, AFR).
+Combined 2D/DB [top left], DPlex [middle left], Pixel [middle center],
+2D [bottom left], DB [bottom center] and stereo [bottom right].
 
@@
     1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
-    1.6 RC1 & Jul 31, 2009 & Release Candidate 1 for Equalizer 0.9\\
+    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
     \hline

### 2009-09-09 Developer/ProgrammingGuide/paper.tex
Minor PUG updates based on print version
[ ] Will break Win32 build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.6 for Equalizer 0.9\\[\medskipamount]
+Version 1.7.0 for Equalizer 0.9+svn\\[\medskipamount]
 %  \today
@@
     1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
+    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
     \hline
@@
 
-\subsection{Installing Equalizer and running \textsf{eqPly}}
+\subsection{Installing Equalizer and Running \textsf{eqPly}}
 
@@
   \includegraphics[width=.618\textwidth]{images/2D.pdf}
-  {\caption{ A 2D compound}}
+  {\caption{ A 2D Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/DB.pdf}
-  {\caption{\label{fDB}\small A database compound}}
+  {\caption{\label{fDB}\small A Database Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/EYE.pdf}
-  {\caption{\label{fStereoCmp}\small A stereo compound}}
+  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
-  {\caption{ A DPlex compound}}
+  {\caption{ A DPlex Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
-  {\caption{\label{fPixel}\small A pixel compound}}
+  {\caption{\label{fPixel}\small A Pixel Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
-  {\caption{\label{fPixelKernel}\small Pixel compound kernel}}
+  {\caption{\label{fPixelKernel}\small Pixel Compound Kernel}}
 \end{wrapfigure}
@@
   \includegraphics[width=.9\textwidth]{images/pixelKernels.pdf}
-  {\caption{\label{fPixelKernels}Example pixel kernels for a
-      four-to-one pixel compound}}
+  {\caption{\label{fPixelKernels}Example Pixel Kernels for a
+      four-to-one Pixel Compound}}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{images/cave.pdf}
-  {\caption{\label{fConfig}An example configuration}}
+  {\caption{\label{fConfig}An Example Configuration}}
 \end{figure}
@@
   \includegraphics[width=.618\textwidth]{images/frusta.pdf}
-  {\caption{\label{fFrusta}Wall and projection parameters}}
+  {\caption{\label{fFrusta}Wall and Projection Parameters}}
 \end{wrapfigure}
@@
   \includegraphics[width=.382\textwidth]{images/canvas.pdf}
-  {\caption{\label{fCanvas}A canvas using four, non-consecutive segments}}
+  {\caption{\label{fCanvas}A Canvas using four Segments}}
 \end{wrapfigure}
@@
   \includegraphics[width=.382\textwidth]{images/layout.png}
-  {\caption{\label{fLayout}A layout with four views}}
+  {\caption{\label{fLayout}A Layout with four Views}}
 \end{wrapfigure}
@@
   \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
-  {\caption{A display wall using a six-segment canvas with a two-view layout}}
+  {\caption{A Display Wall using a six-Segment Canvas with a two-View Layout}}
 \end{figure}
@@
   \includegraphics[width=.382\textwidth]{images/lb.pdf}
-  {\caption{\label{fLoadBalancing}\small 2D load-balancing}}
+  {\caption{\label{fLoadBalancing}\small 2D Load-Balancing}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
-  {\caption{\label{fViewLoadBalancing}\small Cross-segment
-      load-balancing for two segments using eight GPU's}}
+  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
+      Load-Balancing for two Segments using eight GPU's}}
 \end{wrapfigure}
@@
     \includegraphics[width=.382\textwidth]{images/viewLB.png}
-    {\caption{\label{fViewLoadBalancing2}\small Cross-segment
-        load-balancing for a CAVE}}
+    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
+        Load-Balancing for a CAVE}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/dfr.pdf}
-  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
+  {\caption{\label{fDFR}Dynamic Frame Resolution}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/monitorEq.pdf}
-  {\caption{\label{fMonitorEq}\small Monitoring a projection wall}}
+  {\caption{\label{fMonitorEq}Monitoring a Projection Wall}}
 \end{wrapfigure}
@@
   \includegraphics[width=.382\textwidth]{images/namespaces}
-  {\caption{\label{fNamespaces}Namespace layering for eqPly and eqServer}}
+  {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/model.pdf}
-  {\caption{\label{fModel}Simplified execution model}}
+  {\caption{\label{fModel}Simplified Execution Model}}
 \end{wrapfigure}
@@
 
-\section{\label{sEqPly}The eqPly polygonal renderer}
+\section{\label{sEqPly}The eqPly Polygonal Renderer}
 
@@
   \includegraphics[width=.9\textwidth]{images/uml}
-  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes}}
+  {\caption{\label{fUml}UML Diagram eqPly and relevant Equalizer Classes}}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{images/syncAsync}
-  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
+  {\caption{\label{fSyncAsync}Synchronous and Asynchronous Execution}}
 \end{figure}
@@
     \includegraphics[width=.382\textwidth]{images/umlObjectMI}}%
-  {\caption{\label{fObjectDist}Object distribution using subclassing,
-      proxies or multiple inheritance}}
+  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
+      Proxies or Multiple Inheritance}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
-  {\caption{\label{fOSWindow}eq::OSWindow UML class hierarchy}}
+  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}
 \end{wrapfigure}
@@
   \includegraphics[width=.382\textwidth]{images/DB.png}
-  {\caption{\label{fDBDest}Destination view of a DB compound using
-      demonstrative coloring}}
+  {\caption{\label{fDBDest}Destination View of a DB Compound using
+      Demonstrative Coloring}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/threads.pdf}
-  {\caption{\label{fThreads}Threads within one node process}}
+  {\caption{\label{fThreads}Threads within one Node Process}}
 \end{wrapfigure}
@@
   {\caption{\label{fThreadModels}Async, draw\_sync and local\_sync
-    thread synchronization models}}
+    Thread Synchronization Models}}
 \end{figure}
@@
   \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\label{fFrameSync}Per-node frame synchronization}}
+  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
 \end{wrapfigure}
@@
   \includegraphics[width=\textwidth]{images/mainloop.pdf}
-  {\caption{\label{fMainLoop}Synchronization of frame tasks}}
+  {\caption{\label{fMainLoop}Synchronization of Frame Tasks}}
 \end{figure}
@@
   {\caption{\label{fImmersive}Monoscopic, Stereoscopic, Tracked and HMD
-    frusta}}
+    Frusta}}
 \end{figure}
@@
   \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
-  {\caption{\label{fViewUML}UML hierarchy of \textsf{eqPly::View}}}
+  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
   \vspace{-5ex}
@@
   \includegraphics[width=.618\textwidth]{images/assembly.pdf}
-  {\caption{\label{fAssembly}Hierarchy of assembly classes}}
+  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/compositor.pdf}
-  {\caption{\label{fCompositor}Functional diagram of the compositor}}
+  {\caption{\label{fCompositor}Functional Diagram of the Compositor}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/statistics}
-  {\caption{\label{fStatistics}Statistics for a two-node 2D compound}}
+  {\caption{\label{fStatistics}Statistics for a two-Node 2D Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/netNode.pdf}
-  {\caption{\label{fNetNode}Communication between two nodes}}
+  {\caption{\label{fNetNode}Communication between two Nodes}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/netObject.pdf}
-  {\caption{\label{fNetObject}Communication between two objects}}
+  {\caption{\label{fNetObject}Communication between two Objects}}
 \end{wrapfigure}

### 2009-10-19 Developer/ProgrammingGuide/paper.tex
Updates for the new window system interface (see revision 3463).
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 Equalizer provides a generic interface and implementation for the three
-most common window systems: AGL, WGL and glX. The interfaces define the
+most common window systems through the OpenGL specific 
+\textsf{eq::GLWindow} class: AGL, WGL and glX. The interfaces define the
 minimal functionality needed to reuse other window system specific
-classes, for example the AGL, WGL and glX event handlers. The
-implementation derived from these interfaces provides a sample
-implementation which honors all configurable window attributes.
+classes, for example the AGL, WGL and glX event handlers. The separation 
+of the OpenGL calls from the \textsf{OSWindow} permits to implement 
+window systems which don't use any OpenGL context and therefore to use a 
+different renderer. The implementation derived from these interfaces 
+provides a sample implementation which honors all configurable window 
+attributes.
 
@@
     _state = new VertexBufferState( getObjectManager( ));
-    _loadLogo();
 
@@
     const InitData& initData = config->getInitData();
+
+    if( initData.showLogo( ))
+        _loadLogo();
+
     if( initData.useGLSL() )

### 2009-10-20 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.0 for Equalizer 0.9+svn\\[\medskipamount]
+Version 1.7.1 for Equalizer 0.9+svn\\[\medskipamount]
 %  \today
@@
     1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
+    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
     \hline
@@
 most common window systems through the OpenGL specific 
-\textsf{eq::GLWindow} class: AGL, WGL and glX. The interfaces define the
+\textsf{eq::GL\-Window} class: AGL, WGL and glX. The interfaces define the
 minimal functionality needed to reuse other window system specific

### 2009-11-18 Developer/ProgrammingGuide/paper.tex
FSAA-related documentation update.
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
       four-to-one Pixel Compound}}
+\end{figure}
+
+\subsection{Subpixel Compounds}
+
+Subpixel decomposes the rendering into different jittered frusta so 
+that the recomposition anti-aliases the final rendering.
+The Subpixel compound is also compatible with applications doing their own software 
+anti-aliasing.
+
+As for the DB compound, the subpixel recomposition has linear increasing 
+IO requirements for the pixel transfer with the difference that only 
+color information has to be transmitted.
+
+Subpixel compounds in Equalizer are configured using the \textsf{subpixel}
+parameter, using the values \textsf{[ index size ]}. The \textsf{index} 
+parameter corresponds to the current ressource's index and the \textsf{size} 
+is the total number of ressources needed.
+
+The \textsf{index} has to be smaller than the ressources \textsf{size}. 
+\fig{fSubpixel} illustrates a four-way subpixel decomposition.
+
+The destination channel can also be used as a source channel. 
+As for the pixel compound, it has to use an output and corresponding 
+input frame.
+
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/subpixel.pdf}
+  {\caption{\label{fSubpixel}A four-way Subpixel compound}}
 \end{figure}
@@
 \textsf{pixel} parameter adjusts the frustum so that the source channel
-renders an even subset of the parent's pixels. Setting one or multiple
-attributes causes the parent's view to be decomposed
-accordingly. Attributes are cumulative, that is, intermediate compound
-attributes affect and therefore decompose the rendering of all their
-children.
+renders an even subset of the parent's pixels. A \textsf{subpixel} parameter
+tells the source channels to render different jittered frustum to perform
+anti-aliasing. Setting one or multiple attributes causes the parent's view
+to be decomposed accordingly. Attributes are cumulative, that is,
+intermediate compound attributes affect and therefore decompose the
+rendering of all their children.
 
@@
         os << _modelID << _renderMode << _colorMode << _ortho << _statistics
-           << _help << _wireframe << _pilotMode;
+           << _help << _wireframe << _pilotMode << _idleMode;
     if( dirtyBits & DIRTY_VIEW )
@@
         is >> _modelID >> _renderMode >> _colorMode >> _ortho >> _statistics
-           >> _help >> _wireframe >> _pilotMode;
+           >> _help >> _wireframe >> _pilotMode >> _idleMode;
     if( dirtyBits & DIRTY_VIEW )
@@
         _tracker.update();
-        const vmml::Matrix4f& headMatrix = _tracker.getMatrix();
-        setHeadMatrix( headMatrix );
+        const eq::Matrix4f& headMatrix = _tracker.getMatrix();
+        _setHeadMatrix( headMatrix );
     }
@@
     // update database
-    _frameData.data.rotation.preRotateX( -0.001f * _spinX );
-    _frameData.data.rotation.preRotateY( -0.001f * _spinY );
+    if( _animation.isValid( ))
+    {
+        const eq::Vector3f&  modelRotation = _animation.getModelRotation();
+        const CameraAnimation::Step& curStep = _animation.getNextStep();
+
+        _frameData.setModelRotation( modelRotation        );
+        _frameData.setRotation(     curStep.rotation      );
+        _frameData.setTranslation(  curStep.translation   );
+    }
+    else
+    {
+        if( _frameData.usePilotMode())
+            _frameData.spinCamera( -0.001f * _spinX, -0.001f * _spinY );
+        else
+            _frameData.spinModel( -0.001f * _spinX, -0.001f * _spinY );
+
+        _frameData.moveCamera( 0.0f, 0.0f, 0.001f*_advance );
+    }
+
+    [...]
+
     const uint32_t version = _frameData.commit();
 
+    _redraw = false;
     return eq::Config::startFrame( version );
@@
   channel's view definition, compound viewports, head matrix and the
-  current eye pass.
+  current eye pass. If the anti-aliasing idle mode is set, the frustum
+  will be jittered by a vector. This part is described in \ref{sFSAA}.
 \item[Head Transformation] A transformation matrix positioning the
@@
     }
+
+    [...]
 }
@@
 Like most applications, eqPly can use the default implementation of the
-\textsf{frameReadback} and \textsf{frameAssemble} task methods. Only
-\textsf{frameReadback} is overwritten for an optimization. Since eqPly
-does not need the alpha channel on the destination view, the output
-frames are flagged to ignore it, which allows the compressor to drop
-25\% of the data during image transfer:
+\textsf{frameReadback} and \textsf{frameAssemble} task methods.
+\textsf{frameReadback} is overwritten for an optimization since eqPly
+does not need the alpha channel on the destination view.
+The output frames are flagged to ignore it, which allows the compressor to
+drop 25\% of the data during image transfer:
 
@@
 \end{lstlisting}}
+
+\textsf{frameAssemble} is overwritten for the use of Subpixel compound.
+The whole process is described in \ref{sFSAA}.
 
@@
 
+\subsection{\label{sFSAA}Full-scene and Software Anti-Aliasing}
+
+The eqPly example contains support for software anti-aliasing to show 
+how to combine it with the Equalizer Subpixel compound.
+
+\subsubsection{FSAA or Supersampling algorithm}
+
+The anti-aliasing algorithm uses here the supersampling technique which 
+computes samples within a pixel using the jitter algorithm. The pixel is split
+into several sub-pixels and a sample is taken randomly within the sub-pixel.
+
+The sub-pixel area is a function of the current jitter step and the channel's
+\textsf{subpixel} description. Jitter values for all channels should be evenly 
+distributed over the total sub-pixel grid and we use therefore a precomputed
+prime number table.
+
+{\footnotesize\begin{lstlisting}
+eq::Vector2i Channel::_getJitterStep() const
+{
+    uint32_t channelID  = getSubPixel().index;
+    uint32_t idx = ( channelID * _totalSteps );
+    idx += ( _jitterStep * primeNumberTable[ channelID ] ) \% _totalSteps;
+
+    uint32_t sampleSize = _getSampleSize();
+    const int dx = idx \% sampleSize;
+    const int dy = idx / sampleSize;
+
+    return eq::Vector2i( dx, dy );
+}
+\end{lstlisting}}
+
+Next to that, sizes of a pixel and a sub-pixel on the near plane are computed.
+A random position within the sub-pixel is set as a sample position and the 
+vector values are correctly defined to finally be passed to the frustum.
+
+{\footnotesize\begin{lstlisting}
+eq::Vector2f Channel::_getJitter() const
+{
+    const eq::PixelViewport& pvp = getPixelViewport();
+    float pvp_w = static_cast<float>( pvp.w );
+    float pvp_h = static_cast<float>( pvp.h );
+    float frustum_w = static_cast<float>(( getFrustum().get_width( )));
+    float frustum_h = static_cast<float>(( getFrustum().get_height( )));
+
+    float pixel_w = frustum_w / pvp_w;
+    float pixel_h = frustum_h / pvp_h;
+
+    uint32_t sampleSize = _getSampleSize();
+    float subpixel_w = pixel_w / static_cast<float>( sampleSize );
+    float subpixel_h = pixel_h / static_cast<float>( sampleSize );
+
+    eq::Vector2i jitterStep = _getJitterStep();
+
+    // Sample value randomly computed within the subpixel
+    eq::base::RNG rng;
+    float value_i = rng.get< float >() * subpixel_w
+                    + static_cast<float>( jitterStep.x( )) * subpixel_w;
+
+    float value_j = rng.get< float >() * subpixel_h
+                    + static_cast<float>( jitterStep.y( )) * subpixel_h;
+
+    return eq::Vector2f( value_i, value_j );
+}
+\end{lstlisting}}
+
+\subsubsection{SWAA in eqPly}
+
+This is the \textsf{FrameData} class which will give the information to the
+application about being in idle mode or not. \textsf{Config} will update the 
+idle mode information to the new \textsf{FrameData} version depending of the 
+application state.
+
+Since the application has a notion of idle/non-idle rendering, the 
+\textsf{Channel} will be able to perform anti-aliasing when no user event is
+detected and more frames are requested.
+
+If the application is in idle mode, the \textsf{applyFrustum} operation will 
+apply the computed jitter vector to the current frustum.
+
+{\footnotesize\begin{lstlisting}
+void Channel::applyFrustum() const
+{
+    const FrameData& frameData = _getFrameData();
+
+    if( frameData.isIdle() && _jitterStep > 0 )
+    {
+        eq::Vector2f jitter = _getJitter();
+
+        if( frameData.useOrtho( ))
+        {
+            eq::Frustumf ortho = getOrtho();
+            ortho.apply_jitter( jitter );
+
+            glOrtho( ortho.left(), ortho.right(),
+                     ortho.bottom(), ortho.top(),
+                     ortho.near_plane(), ortho.far_plane( ));
+        }
+        else
+        {
+            eq::Frustumf frustum = getFrustum();
+            frustum.apply_jitter( jitter );
+
+            glFrustum( frustum.left(), frustum.right(),
+                       frustum.bottom(), frustum.top(),
+                       frustum.near_plane(), frustum.far_plane( ));
+        }
+        return;
+    }
+    //else
+
+    eq::Channel::applyFrustum();
+}
+\end{lstlisting}}
+
+Then, \textsf{frameViewFinish} will get the frame buffer and accumulate 
+it into an accumulation buffer. At each jitter step, the average result 
+contained in the accumulation buffer will be displayed until all the steps
+are done.
+If the current pixel viewport is different from the one saved in 
+\textsf{frameViewStart}, the accumulation buffer needs also to be resized 
+and the idle anti-aliasing is reset.
+
+{\footnotesize\begin{lstlisting}
+void Channel::frameViewFinish( const uint32_t frameID )
+{
+    const FrameData& frameData = _getFrameData();
+
+    const eq::PixelViewport& pvp = getPixelViewport();
+    bool isResized = _accum->resize( pvp.w, pvp.h );
+
+    if( isResized )
+    {
+        _accum->clear();
+        _jitterStep = _totalSteps;
+        _subpixelStep = 0;
+        return;
+    }
+    //else
+
+    if( frameData.isIdle() && _needsTransfer )
+    {
+        setupAssemblyState();
+
+        _accum->accum();
+        _accum->display();
+
+        resetAssemblyState();
+    }
+
+    _drawOverlay();
+    _drawHelp();
+
+    if( frameData.useStatistics())
+        EQ_GL_CALL( drawStatistics( ));
+}
+\end{lstlisting}}
+
+At the end, \textsf{frameFinish} will decrement the jitter step and create a
+new event to inform the \textsf{Config} of how much frames which remain to 
+draw.
+
+\subsubsection{SWAA in Equalizer}
+
+As the software anti-aliasing decomposition should also work with the idle 
+anti-aliasing in the application, the accumulation buffer is part of the 
+\textsf{ObjectManager}. \textsf{frameAssemble} and \textsf{frameViewFinish} 
+will then share the same accumulation buffer.
+
+As for the application, the \textsf{eq::Channel::applyFrustum} will apply the
+jitter vector if it exists. Then, if the function is not overwritten in the
+application but if it still uses \textsf{subpixel} compounds, anti-aliasing 
+will be correctly computed in Equalizer.
+
+The main bit here is the assembly operation. \textsf{assembleFramesUnsorted} 
+and \textsf{assembleFramesSorted} are similarly modified to handle the 
+Subpixel decomposition.
+
+For each Subpixel decomposition in input frames, all input frames are 
+assembled with \textsf{subpixel} decomposition. The result is then accumulated
+and displayed.
+
+For example a 2-way subpixel compound configuration like 
+\textsf{[0 2] [0 2] [1 2] [1 2]} will first assemble the both \textsf{[0 2]}
+input frames and accumulate the result of the two assemblies. 
+Similarly, the the both \textsf{[1 2]} frames are assembled and their result 
+accumulated to the previous one.
+In another hand, a 4-way subpixel compound configuration will do four 
+assemblies and four accumulations.
+
+{\footnotesize\begin{lstlisting}
+void Compositor::assembleFramesUnsorted( const FrameVector& frames, 
+                                         Channel* channel, Accum* accum )
+{
+    if( frames.empty( ))
+        return;
+
+    if( _isSubPixelDecomposition( frames ))
+    {
+        if( !accum )
+        {
+            accum = obtainAccum( channel );
+            accum->clear();
+        }
+
+        FrameVector framesLeft = frames;
+    	while( !framesLeft.empty( ))
+    	{
+            // Get the frames with the same subpixel compound
+    	    FrameVector current = _extractOneSubPixel( framesLeft );
+            assembleFramesUnsorted( current, channel, accum );
+            accum->accum();
+            accum->display();
+        }
+        return;
+    }
+
+    [...]
+
+    // assemble frame
+}
+\end{lstlisting}}
+
 \subsection{\label{sStatistics}Statistics}
@@
 parent. The pixel setting selects the pixel decomposition kernel,
-relative to the parent. The eye pass select the mono, left or right eye
-passes rendered by the compound. The zoom scales the parent pixel
-viewport resolution. The DPlex period defines that $\frac{1}{period}$
+relative to the parent. The subpixel defines the jittering applied to
+the frustum, relative to the parent. The eye pass select the mono, left
+or right eye passes rendered by the compound. The zoom scales the parent
+pixel viewport resolution. The DPlex period defines that $\frac{1}{period}$
 frames are rendered, and the phase defines when in the period the
@@
     pixel    [ int int ]             # pixel decomposition (step size)
+    subpixel [ int int ]			 # subpixel decomposition (index size)
     eye      [ CYCLOP LEFT RIGHT ]   # monoscopic or stereo view

### 2009-11-18 Developer/ProgrammingGuide/paper.tex
FSAA-related documentation cleanup.
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
     range    [ float float ]         # DB-range for sort-last
-    pixel    [ int int ]             # pixel decomposition (step size)
+    pixel    [ int int int int ]     # pixel decomposition (x y w h)
     subpixel [ int int ]	     # subpixel decomposition (index size)

### 2009-11-18 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.1 for Equalizer 0.9+svn\\[\medskipamount]
+Version 1.7.2 for Equalizer 0.9+svn\\[\medskipamount]
 %  \today
@@
 Engineering contributions by Maxim Makhinya, Jonas B\"osch, Christian
-Marten and Patrick Bouchaud.
+Marten, Sarah Amsellem and Patrick Bouchaud.
 
@@
     1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
+    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
     \hline
@@
 
-\subsection{Subpixel Compounds}
-
-Subpixel decomposes the rendering into different jittered frusta so 
-that the recomposition anti-aliases the final rendering.
-The Subpixel compound is also compatible with applications doing their own software 
-anti-aliasing.
+\subsection{\label{sSubpixel}Subpixel Compounds}
+
+Subpixel decomposes the rendering of multiple samples for one pixel, e.g., for
+anti-aliasing or depth-of-field rendering. The default implementation provides
+transparent, scalable software idle-antialiasing when configuring a subpixel
+compound.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/Subpixel.pdf}
+  {\caption{\label{fSubpixel}\small A Subpixel Compound}}
+\end{wrapfigure}
+Applications can use subpixel compounds to accelerate depth-of-field effects and
+software anti-aliasing, with potentially a different number of idle or non-idle
+samples per pixel.
 
@@
 
-Subpixel compounds in Equalizer are configured using the \textsf{subpixel}
-parameter, using the values \textsf{[ index size ]}. The \textsf{index} 
-parameter corresponds to the current ressource's index and the \textsf{size} 
-is the total number of ressources needed.
-
-The \textsf{index} has to be smaller than the ressources \textsf{size}. 
-\fig{fSubpixel} illustrates a four-way subpixel decomposition.
+Subpixel compounds are configured using the \textsf{subpixel} parameter, using
+the values \textsf{[ index size ]}. The \textsf{index} parameter corresponds to
+the current resource index and the \textsf{size} is the total number of
+resources used.
+
+The \textsf{index} has to be smaller than the \textsf{size}. 
+\fig{fSubpixel} illustrates a three-way subpixel decomposition.
 
@@
 input frame.
-
-\begin{figure}[ht!]\center
-  \includegraphics[width=.9\textwidth]{images/subpixel.pdf}
-  {\caption{\label{fSubpixel}A four-way Subpixel compound}}
-\end{figure}
 
@@
 
-Configure one \textsf{layout} for each configuration of logical
-views. Name the layout using a unique name. Often only one layout with a
-one view is used for all canvases.
-
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
+
+Configure one \textsf{layout} for each configuration of logical
+views. Name the layout using a unique name. Often only one layout with a
+one view is used for all canvases.
 
@@
   \begin{description}
-  \item[Decomposition] On each child compound, limit the rendering task
-    of that child by setting the \textsf{viewport}, \textsf{range},
-    \textsf{period} and \textsf{phase}, \textsf{pixel}, \textsf{eye} or
+  \item[Decomposition] On each child compound, limit the rendering task of that
+    child by setting the \textsf{viewport}, \textsf{range}, \textsf{period} and
+    \textsf{phase}, \textsf{pixel}, \textsf{subpixel}, \textsf{eye} or
     \textsf{zoom} as desired.
@@
 \subsubsection{Decomposition - Attributes}
-Compounds have attributes which configure the decomposition of the
-destination channel's rendering, which is defined by the viewport,
-frustum and database. A \textsf{viewport} decomposes the destination
-channel and frustum in screen space. A \textsf{range} tells the
-application to render a part of its database, and an \textsf{eye}
-rendering pass can selectively render different stereo passes. A
-\textsf{pixel} parameter adjusts the frustum so that the source channel
-renders an even subset of the parent's pixels. A \textsf{subpixel} parameter
-tells the source channels to render different jittered frustum to perform
-anti-aliasing. Setting one or multiple attributes causes the parent's view
-to be decomposed accordingly. Attributes are cumulative, that is,
-intermediate compound attributes affect and therefore decompose the
-rendering of all their children.
+Compounds have attributes which configure the decomposition of the destination
+channel's rendering, which is defined by the viewport, frustum and database. A
+\textsf{viewport} decomposes the destination channel and frustum in screen
+space. A \textsf{range} tells the application to render a part of its database,
+and an \textsf{eye} rendering pass can selectively render different stereo
+passes. A \textsf{pixel} parameter adjusts the frustum so that the source
+channel renders an even subset of the parent's pixels. A \textsf{subpixel}
+parameter tells the source channels to render different samples for one pixel to
+perform anti-aliasing or depth-of-field rendering. Setting one or multiple
+attributes causes the parent's view to be decomposed accordingly. Attributes are
+cumulative, that is, intermediate compound attributes affect and therefore
+decompose the rendering of all their children.
 
@@
 \end{wrapfigure}
-While pixel and stereo compounds are naturally load-balanced, 2D and DB
-compounds often need load-balancing for optimal rendering
-performance. Using a load equalizer is transparent to the application,
-and can be used on any application for 2D, and on most applications for
-DB load-balancing. Some applications do not support dynamic updates of
-the database range, and therefore cannot be used with DB load-balancing.
+While pixel, subpixel and stereo compounds are naturally load-balanced, 2D and
+DB compounds often need load-balancing for optimal rendering performance. Using
+a load equalizer is transparent to the application, and can be used on any
+application for 2D, and on most applications for DB load-balancing. Some
+applications do not support dynamic updates of the database range, and therefore
+cannot be used with DB load-balancing.
 
@@
 \paragraph{View Equalizer}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
+  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
+      Load-Balancing for two Segments using eight GPU's}}
+\end{wrapfigure}
 Depending on the model position and data structure, each segment of a
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
-  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
-      Load-Balancing for two Segments using eight GPU's}}
+\begin{wrapfigure}{r}{.382\textwidth} 
+    \includegraphics[width=.382\textwidth]{images/viewLB.png}
+    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
+        Load-Balancing for a CAVE}}
 \end{wrapfigure}
@@
 
-\begin{wrapfigure}{l}{.382\textwidth} 
-    \includegraphics[width=.382\textwidth]{images/viewLB.png}
-    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
-        Load-Balancing for a CAVE}}
-\end{wrapfigure}
 \fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
@@
 \paragraph{DFR Equalizer}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
+  {\caption{\label{fDFR}Dynamic Frame Resolution}}
+\end{wrapfigure}
 Dynamic Frame Resolution (DFR) trades rendering performance for visual
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
-  {\caption{\label{fDFR}Dynamic Frame Resolution}}
-\end{wrapfigure}
 \fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
@@
 
+\clearpage
 \part{Programming Guide}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/osWindow.pdf}
+  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}\vspace{-4ex}
+\end{wrapfigure}
 The particulars of creating a window and OpenGL context depend on the
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
-  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}
-\end{wrapfigure}
 All window-system specific functionality is implemented by a
@@
 \item[Frustum] The same frustum parameters as defined by
-  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL
-  projection matrix. The frustum is influenced by the destination
-  channel's view definition, compound viewports, head matrix and the
-  current eye pass. If the anti-aliasing idle mode is set, the frustum
-  will be jittered by a vector. This part is described in \ref{sFSAA}.
+  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL projection
+  matrix. The frustum is influenced by the destination channel's view
+  definition, compound viewports, head matrix and the current eye pass. If the
+  channel has a subpixel parameter, the frustum will be jittered before it is
+  applied. Please refer to \ref{sFSAA} for more information.
 \item[Head Transformation] A transformation matrix positioning the
@@
 
-%\begin{wrapfigure}{r}{.618\textwidth}
-\begin{figure}[h!t]\center
+\begin{wrapfigure}{r}{.618\textwidth}
+%\begin{figure}[h!t]\center
   \includegraphics[width=.618\textwidth]{images/render.pdf}
   {\caption{\label{fRender}Main Render Loop}}
-\end{figure}
+\end{wrapfigure}
 The rendering loop maintains a list of candidates to render, which
@@
 
-\textsf{frameAssemble} is overwritten for the use of Subpixel compound.
-The whole process is described in \ref{sFSAA}.
+The \textsf{frameAssemble} method is overwritten for the use of Subpixel
+compound with idle software anti-aliasing, as described in \ref{sFSAA}.
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-\vspace{-2ex}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}
-\vspace{-4ex}
-\end{wrapfigure}
 Events are received by an event handler. The event handler finds the
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\label{fEventProcessing}Event Processing}}
+\end{wrapfigure}
 The purpose of the \textsf{OSWindow} processing method,
@@
 
-Equalizer provides an \textsf{AGLEvent\-Handler},
-\textsf{GLXEventHandler} and \textsf{WGL\-Event\-Handler}, which handle
-events for an \textsf{AGLWindowIF}, \textsf{GLXWindowIF} and
-\textsf{WGLWindowIF}, respectively. \fig{fEventUML} illustrates the
-class hierarchy for event processing.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+Equalizer provides an \textsf{AGLEvent\-Handler},
+\textsf{GLXEventHandler} and \textsf{WGL\-Event\-Handler}, which handle
+events for an \textsf{AGLWindowIF}, \textsf{GLXWindowIF} and
+\textsf{WGL\-Win\-dowIF}, respectively. \fig{fEventUML} illustrates the
+class hierarchy for event processing.
+
 The concrete implementation of these window interfaces is responsible
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/threads.pdf}
+  {\caption{\label{fThreads}Threads within one Node Process}}
+\end{wrapfigure}
 The application or node main thread is the primary thread of each
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/threads.pdf}
-  {\caption{\label{fThreads}Threads within one Node Process}}
-\end{wrapfigure}
 During this initialization, Equalizer creates and manages two threads
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
 The application has extended control over the task synchronization
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
 Each entity uses \textsf{waitFrame\-Started} to block on the parent's
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-2ex}
-  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
-  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
-  \vspace{-5ex}
-\end{wrapfigure}
 Changing view frustum parameters is typically done for non-fullscreen
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
+  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
+\end{wrapfigure}
 A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\label{fDirectSend}Direct Send Compositing}}
+  \vspace{-3ex}
+\end{wrapfigure}
 To provide a motivation for the design of the image compositing
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\label{fDirectSend}Direct Send Compositing}}
-\end{wrapfigure}
 The main idea behind direct send is to parallelize the costly
@@
 
-The frame data is a holder for images and additional information, such
-as output frame attributes and pixel data availability.
-
-An \textsf{eq::Image} holds a
-two-dimensional snapshot of the framebuffer and can contain color and/or
-depth information.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+The frame data is a holder for images and additional information, such
+as output frame attributes and pixel data availability.
+
+An \textsf{eq::Image} holds a
+two-dimensional snapshot of the framebuffer and can contain color and/or
+depth information.
+
 The frame synchronization through the frame data allows the input frame
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
+\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
-\end{wrapfigure}
 Conceptually, the individual volume bricks of each of the source
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended to form a
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 Finding the correct assembly order for perspective frusta is more
@@
 
-\subsection{\label{sFSAA}Full-scene and Software Anti-Aliasing}
-
-The eqPly example contains support for software anti-aliasing to show 
-how to combine it with the Equalizer Subpixel compound.
-
-\subsubsection{FSAA or Supersampling algorithm}
-
-The anti-aliasing algorithm uses here the supersampling technique which 
-computes samples within a pixel using the jitter algorithm. The pixel is split
-into several sub-pixels and a sample is taken randomly within the sub-pixel.
-
-The sub-pixel area is a function of the current jitter step and the channel's
-\textsf{subpixel} description. Jitter values for all channels should be evenly 
-distributed over the total sub-pixel grid and we use therefore a precomputed
-prime number table.
+\subsection{\label{sFSAA}Custom Subpixel Processing}
+
+The eqPly example contains support for software anti-aliasing when being idle,
+showing how to combine it with subpixel compounds.
+
+\subsubsection{Supersampling Algorithm}
+
+The supersampling algorith used by eqPly performs anti-aliasing, using a
+technique which computes samples within a pixel by jittering the frustum by a
+subpixel amount. The pixel is split into several, regular sub-pixels and a
+sample is taken randomly within the sub-pixel.
+
+The subpixel area is a function of the current jitter step and the channel's
+\textsf{subpixel} description. To evenly distribute jitter values for all
+channels over the total sub-pixel grid a precomputed prime number table is used
+to pseudo-randomly select the subpixel:
 
@@
     uint32_t idx = ( channelID * _totalSteps );
-    idx += ( _jitterStep * primeNumberTable[ channelID ] ) \% _totalSteps;
+    idx += ( _jitterStep * primeNumberTable[ channelID ] ) % _totalSteps;
 
     uint32_t sampleSize = _getSampleSize();
-    const int dx = idx \% sampleSize;
+    const int dx = idx % sampleSize;
     const int dy = idx / sampleSize;
@@
 
-Next to that, sizes of a pixel and a sub-pixel on the near plane are computed.
-A random position within the sub-pixel is set as a sample position and the 
-vector values are correctly defined to finally be passed to the frustum.
+Using this normalized subpixel center point and the size of a pixel on the near
+plane, the jitter for the frustum is computed. A random position within the
+sub-pixel is set as a sample position, which will be used to move the frustum:
 
@@
 
-\subsubsection{SWAA in eqPly}
-
-This is the \textsf{FrameData} class which will give the information to the
-application about being in idle mode or not. \textsf{Config} will update the 
-idle mode information to the new \textsf{FrameData} version depending of the 
-application state.
-
-Since the application has a notion of idle/non-idle rendering, the 
-\textsf{Channel} will be able to perform anti-aliasing when no user event is
-detected and more frames are requested.
+\subsubsection{Idle Anti-Aliasing in eqPly}
+
+The \textsf{FrameData} class holds the information whether or not the
+application is in idle mode. The \textsf{Config} updates the idle mode
+information depending on the application's state.  Each \textsf{Channel} will
+perform anti-aliasing when no user event requires a redraw.
 
@@
 
-Then, \textsf{frameViewFinish} will get the frame buffer and accumulate 
-it into an accumulation buffer. At each jitter step, the average result 
-contained in the accumulation buffer will be displayed until all the steps
-are done.
+At the end of the frame, \textsf{frameViewFinish} accumulates the frame buffer
+into an accumulation buffer. Since \textsf{frameViewFinish} is only called on
+destination channels this operation is done only on the final rendering
+result. After each frame, the result accumulated so far will be displayed. When
+all the steps are done the config will stop rendering new frames.
+
 If the current pixel viewport is different from the one saved in
-\textsf{frameViewStart}, the accumulation buffer needs also to be resized 
-and the idle anti-aliasing is reset.
+\textsf{frameViewStart}, the accumulation buffer needs also to be resized and
+the idle anti-aliasing is reset.
 
@@
 
-At the end, \textsf{frameFinish} will decrement the jitter step and create a
-new event to inform the \textsf{Config} of how much frames which remain to 
-draw.
-
-\subsubsection{SWAA in Equalizer}
-
-As the software anti-aliasing decomposition should also work with the idle 
-anti-aliasing in the application, the accumulation buffer is part of the 
-\textsf{ObjectManager}. \textsf{frameAssemble} and \textsf{frameViewFinish} 
-will then share the same accumulation buffer.
-
-As for the application, the \textsf{eq::Channel::applyFrustum} will apply the
-jitter vector if it exists. Then, if the function is not overwritten in the
-application but if it still uses \textsf{subpixel} compounds, anti-aliasing 
-will be correctly computed in Equalizer.
-
-The main bit here is the assembly operation. \textsf{assembleFramesUnsorted} 
-and \textsf{assembleFramesSorted} are similarly modified to handle the 
-Subpixel decomposition.
-
-For each Subpixel decomposition in input frames, all input frames are 
-assembled with \textsf{subpixel} decomposition. The result is then accumulated
-and displayed.
-
-For example a 2-way subpixel compound configuration like 
-\textsf{[0 2] [0 2] [1 2] [1 2]} will first assemble the both \textsf{[0 2]}
-input frames and accumulate the result of the two assemblies. 
-Similarly, the the both \textsf{[1 2]} frames are assembled and their result 
-accumulated to the previous one.
-In another hand, a 4-way subpixel compound configuration will do four 
-assemblies and four accumulations.
-
-{\footnotesize\begin{lstlisting}
-void Compositor::assembleFramesUnsorted( const FrameVector& frames, 
-                                         Channel* channel, Accum* accum )
-{
-    if( frames.empty( ))
-        return;
-
-    if( _isSubPixelDecomposition( frames ))
-    {
-        if( !accum )
-        {
-            accum = obtainAccum( channel );
-            accum->clear();
-        }
-
-        FrameVector framesLeft = frames;
-    	while( !framesLeft.empty( ))
-    	{
-            // Get the frames with the same subpixel compound
-    	    FrameVector current = _extractOneSubPixel( framesLeft );
-            assembleFramesUnsorted( current, channel, accum );
-            accum->accum();
-            accum->display();
-        }
-        return;
-    }
-
-    [...]
-
-    // assemble frame
-}
-\end{lstlisting}}
+At the end of the frame for each channel, \textsf{frameFinish} will decrement
+the jitter step and create a new event to inform the \textsf{Config} of how many
+frames remain to draw before the idle anti-aliasing is complete.
+
+\subsubsection{Transparent Software Anti-Aliasing}
+
+Any Equalizer applications can benefit without modification from subpixel
+compounds. Applications performing their own, configurable idle or non-idle
+anti-aliasing can easily be integrated with subpixel compounds, as described in
+the next section.
+
+The default implementation of \textsf{applyFrustum} and \textsf{applyOrtho} in
+\textsf{eq::Channel} apply jittering when a subpixel parameter is set.
+Furthermore, the \textsf{Compositor} uses an \textsf{eq::Accum} buffer to
+accumulate and display the results of a subpixel decomposition in \textsf{assembleFramesUnsorted} and \textsf{assembleFramesSorted}.
+
+\subsubsection{Subpixel Compounds in eqPly}
+
+Subpixel compounds accelerate the idle anti-aliasing computation in eqPly. The
+accumulation buffer used in the \textsf{Compositor} in \textsf{frameAssemble} is
+the same accumulation buffer used by eqPly in \textsf{frameViewFinish} for
+collecting the idle anti-aliasing results.
+
+When an application is passing an \textsf{eq::Accum} object to the compositor,
+the compositor adds subpixel decomposition result into this accumulation
+buffer. Otherwise it uses its own accumulation buffer, which it clears in the
+beginning of the compositing process.
+
+Since \textsf{eqPly} accumulates the results over multiple rendering frames,
+each \textsf{Channel} manages its own accumulation buffer, which is passed to
+the compositor in \textsf{frameAssemble} and used in \textsf{frameViewFinish} to
+display the results accumulated so far.
+
+The number of performed jitter steps is tracked in \textsf{frameDraw} and
+\textsf{frameAssemble}, and is used in \textsf{frameFinish} to decrement the
+outstanding anti-aliasing samples. Therefore, the idle anti-aliasing is finished
+much faster if the current configuration uses a subpixel compound.
 

### 2009-12-23 Developer/ProgrammingGuide/paper.tex
CUDA section added, including hint_cuda_GL_interop. More to come...

@@
 Engineering contributions by Maxim Makhinya, Jonas B\"osch, Christian
-Marten, Sarah Amsellem and Patrick Bouchaud.
+Marten, Sarah Amsellem, Patrick Bouchaud and Philippe Robert.
 
@@
 
+\subsection{\label{sCUDA}GPU Computing with CUDA}
+
+The Equalizer parallel rendering framework can be used to scale GPU 
+computing applications (also known as GPGPU) based on CUDA. The necessary
+steps to write a distributed GPGPU application using Equalizer are described
+in the following.
+
+Please note that the support for CUDA is only enabled if the CUDA libraries are found
+in their default locations. Otherwise please adapt the makefiles accordingly.
+
+\subsubsection{CUDA Initialization}
+
+Equalizer automatically initializes a CUDA context on each pipe (using the specified
+device number) whenever \textsf{hint\_cuda\_GL\_interop} is set. The initialization is 
+currently done using the CUDA runtime API in order to support device emulation. 
+On the other side, if this hint is not set the OpenGL interop feature is not enabled, and 
+furthermore, the GPU which executes CUDA code defaults to device 0.
+
+Once CUDA is initialized correctly, C for CUDA can be used as any other GPU 
+programming interface.
+
+\subsubsection{CUDA Memory Management}
+
+Equalizer does not provide any facility to perform memory transfer from and to the CUDA
+devices. This is entirely left to the programmer.
+
 \section{\label{sNetwork}The Network Layer}
@@
   \hline
+  \textbf{Name} & EQ\_PIPE\_IATTR\_HINT\_CUDA\_GL\_INTEROP\\
+  \textbf{Value} & OFF \textbar \ ON\\
+  \textbf{Default} & OFF\\
+  \textbf{Details} & Determines if the pipe thread enables interoperability between CUDA
+  and OpenGL or not. When set to ON, the CUDA device is configured on which the 
+  thread executes the CUDA device code. If no graphics adapter is specified in the config 
+  file, the one providing the highest compute performances on the particular node is chosen. 
+  When set to OFF, no CUDA device is configured at all.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
   \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_STEREO\\
@@
         hint_thread            OFF | ON   # default ON
+        hint_cuda_GL_interop   OFF | ON   # default OFF
     }

### 2010-02-02 Developer/ProgrammingGuide/paper.tex
Programming Guide: fix typo
[ ] May break build
[ ] May break existing applications
[x] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[ ] Documentation
Update Programming Guide and 0.9.1-related web pages
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.2 for Equalizer 0.9+svn\\[\medskipamount]
+Version 1.7.3 for Equalizer 0.9.1\\[\medskipamount]
 %  \today
@@
 \lstset{language=C++}
-\section*{Equalizer 0.9 Programming and User Guide}
+\section*{Equalizer 0.9.1 Programming and User Guide}
 \today
@@
 
-\textcopyright 2007-2009 \htmladdnormallink{Eyescale Software GmbH}
+\textcopyright 2007-2010 \htmladdnormallink{Eyescale Software GmbH}
 {http://www.eyescale.ch}. All rights reserved. No permission is
@@
 The images on the back page show the following scalable rendering modes:
-Combined 2D/DB [top left], DPlex [middle left], Pixel [middle center],
+Subpixel (FSAA) [top left], DPlex [middle left], Pixel [middle center],
 2D [bottom left], DB [bottom center] and stereo [bottom right].
@@
 \tableofcontents
-\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
+\clearpage
+\listoffigures \vfill
+\section*{Revision History}
+{\center\begin{tabularx}{\textwidth}{|l|l|X|}
     \hline
@@
     1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
+    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
     \hline
@@
 
-For each node, create on \textsf{pipe} for each graphics card in the
+For each node, create one \textsf{pipe} for each graphics card in the
 machine. Set the \textsf{device} number to the correct index. On
@@
 
-\begin{figure}[h!t]
+\begin{figure}[h!t]\center
   \subfigure[]{\label{fMono}
-    \includegraphics[height=2.5cm]{images/mono}}\hfil
+    \includegraphics[height=4cm]{images/mono}}\hfil
   \subfigure[]{\label{fStereo}
-    \includegraphics[height=2.5cm]{images/stereo}}\hfil
+    \includegraphics[height=4cm]{images/stereo}}
   \subfigure[]{\label{fTracked}
-    \includegraphics[height=2.5cm]{images/tracked}}
+    \includegraphics[height=4cm]{images/tracked}}\hfil
   \subfigure[]{\label{fHMD}
-    \includegraphics[height=2.5cm]{images/hmd}}%
-  {\caption{\label{fImmersive}Monoscopic, Stereoscopic, Tracked and HMD
-    Frusta}}
+    \includegraphics[height=4cm]{images/hmd}}
+  {\caption{\label{fImmersive}Monoscopic (a), Stereoscopic (b), Tracked (c) and
+      HMD (d) Frusta}}
 \end{figure}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\label{fDirectSend}Direct Send Compositing}}
-  \vspace{-3ex}
-\end{wrapfigure}
 To provide a motivation for the design of the image compositing
@@
 
+The frame data is a holder for images and additional information, such
+as output frame attributes and pixel data availability.
+
+An \textsf{eq::Image} holds a
+two-dimensional snapshot of the framebuffer and can contain color and/or
+depth information.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The frame data is a holder for images and additional information, such
-as output frame attributes and pixel data availability.
-
-An \textsf{eq::Image} holds a
-two-dimensional snapshot of the framebuffer and can contain color and/or
-depth information.
-
 The frame synchronization through the frame data allows the input frame
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-%\begin{figure}[h!t]\center
+%\begin{wrapfigure}{r}{.618\textwidth}
+\begin{figure}[h!t]\center
   \includegraphics[width=.618\textwidth]{images/compositor.pdf}
   {\caption{\label{fCompositor}Functional Diagram of the Compositor}}
-\end{wrapfigure}
+\end{figure}
 On a high level, the compositor combines multiple input frames using 2D
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
+\end{wrapfigure}
 The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
-\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
@@
   and then tries to launch the render client executable on this host.\\
-  \textbf{See also} & EQ\_CONNECTION\_SATTR\_LAUNCH\_COMMAND
-  EQ\_CONNECTION\_IATTR\_TCPIP\_PORT EQ\_CONNECTION\_IATTR\_TYPE\\
+  \textbf{See also} & EQ\_NODE\_SATTR\_LAUNCH\_COMMAND
+  EQ\_CONNECTION\_IATTR\_PORT
+  EQ\_CONNECTION\_IATTR\_TYPE\\
   \hline
@@
   \hline
-  \textbf{Name} & EQ\_CONNECTION\_SATTR\_LAUNCH\_COMMAND\\
+  \textbf{Name} & EQ\_CONNECTION\_IATTR\_TYPE\\
+  \textbf{Value} & TCPIP \textbar \ SDP \textbar \ RSP \textbar \ 
+  PIPE [Win32 only]\\
+  \textbf{Default} & TCPIP\\
+  \textbf{Details} & The protocol for connections. SDP programmatically
+  selects the socket direct protocol (AF\_INET\_SDP) provided by most
+  InfiniBand protocol stacks, TCPIP uses normal TCP
+  sockets (AF\_INET). RSP provides reliable UDP multicast. PIPE uses a named
+  pipe to communicate between two processes on the same machine.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONNECTION\_IATTR\_PORT\\
+  \textbf{Value} & unsigned\\
+  \textbf{Default} & 0\\
+  \textbf{Details} & The listening port used by the server or
+  node. For nodes, the port can be used to contact pre-started, resident
+  render client nodes or to use a specific port for the node. If 0 is
+  specified, a random port is chosen. Note that a server with no
+  connections automatically creates a default connection using the
+  server's default port.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONNECTION\_IATTR\_FILENAME\\
+  \textbf{Value} & string\\
+  \textbf{Default} & none\\
+  \textbf{Details} & The filename of the named pipe used by the server or
+  node. The filename has to be unique on the local host.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_NODE\_SATTR\_LAUNCH\_COMMAND\\
   \textbf{Value} & string\\
@@
   argument is surrounded by launch command quotes.\\
-  \textbf{See also} & EQ\_CONNECTION\_SATTR\_LAUNCH\_COMMAND\_QUOTE 
-  EQ\_CONNECTION\_IATTR\_LAUNCH\_TIMEOUT\\
+  \textbf{See also} & EQ\_NODE\_SATTR\_LAUNCH\_COMMAND\_QUOTE 
+  EQ\_NODE\_IATTR\_LAUNCH\_TIMEOUT\\
   \hline
@@
   \hline
-  \textbf{Name} & EQ\_CONNECTION\_CATTR\_LAUNCH\_COMMAND\_QUOTE\\
+  \textbf{Name} & EQ\_NODE\_CATTR\_LAUNCH\_COMMAND\_QUOTE\\
   \textbf{Value} & character\\
@@
   \hline
-  \textbf{Name} & EQ\_CONNECTION\_IATTR\_TYPE\\
-  \textbf{Value} & TCPIP \textbar \ SDP \textbar \ PIPE [Win32 only]\\
-  \textbf{Default} & TCPIP\\
-  \textbf{Details} & The protocol for connections. SDP programmatically
-  selects the socket direct protocol (AF\_INET\_SDP) provided by most
-  InfiniBand protocol stacks, TCPIP uses normal TCP
-  sockets (AF\_INET). PIPE uses a named pipe to communicate between
-  two processes on the same machine.\\
-  \hline
-\end{tabularx}\\\vfill
-
-\begin{tabularx}{\textwidth}{|l|X|}
-  \hline
-  \textbf{Name} & EQ\_CONNECTION\_IATTR\_TCPIP\_PORT\\
-  \textbf{Value} & unsigned\\
-  \textbf{Default} & 0\\
-  \textbf{Details} & The listening port used by the server or
-  node. For nodes, the port can be used to contact pre-started, resident
-  render client nodes or to use a specific port for the node. If 0 is
-  specified, a random port is chosen. Note that a server with no
-  connections automatically creates a default connection using the
-  server's default port.\\
-  \hline
-\end{tabularx}\\\vfill
-
-\begin{tabularx}{\textwidth}{|l|X|}
-  \hline
-  \textbf{Name} & EQ\_CONNECTION\_IATTR\_PIPE\_FILENAME\\
-  \textbf{Value} & string\\
-  \textbf{Default} & none\\
-  \textbf{Details} & The filename of the named pipe used by the server or
-  node. The filename has to be unique on the local host.\\
-  \hline
-\end{tabularx}\\\vfill
-
-\begin{tabularx}{\textwidth}{|l|X|}
-  \hline
-  \textbf{Name} & EQ\_CONNECTION\_IATTR\_LAUNCH\_TIMEOUT\\
+  \textbf{Name} & EQ\_NODE\_IATTR\_LAUNCH\_TIMEOUT\\
   \textbf{Value} & unsigned\\
@@
 A connection description defines the network parameters of an Equalizer
-process. Currently TCP/IP, SDP and PIPE connection types are
-supported. TCP/IP creates a TCP socket. SDP is very similar, except that
-the address family \textsf{AF\_INET\_SDP} instead of \textsf{AF\_INET}
-is used to enforce a SDP connection. PIPE uses a named pipe for fast
-interprocess communication on Windows.
+process. Currently TCP/IP, SDP, RSP and PIPE connection types are
+supported. TCP/IP creates a TCP socket. SDP is very similar, except that the
+address family \textsf{AF\_INET\_SDP} instead of \textsf{AF\_INET} is used to
+enforce a SDP connection. RSP provides reliable UDP multicast. PIPE uses a named
+pipe for fast interprocess communication on Windows.
 
@@
 
-The hostname is the IP address or (resolvable) host name. A server or
+The hostname is the IP address or resolvable host name. A server or
 node may have multiple connection descriptions, for example to use a
 named pipe for local communications and TCP/IP for remote nodes.
+
+The interface is the IP address or resolvable host name of the adapter to which
+multicast traffic is send.
 
@@
     {
-        type       TCPIP | SDP | PIPE [Win32 only]
-        TCPIP_port unsigned
-        PIPE_filename string
+        type       TCPIP | SDP | RSP | PIPE [Win32 only]
+        port       unsigned
+        filename   string
         hostname   string
+        interface  string
     }
@@
         damping float # 0: no damping, 1: no changes
+        boundary [ x y ] | float  # x,y: tile boundary for 2D, float: range
     }

### 2010-02-22 Developer/ProgrammingGuide/paper.tex
Programming Guide: add slave object commit section
[ ] May break build
[ ] May break existing applications
[x] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[ ] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.3 for Equalizer 0.9.1\\[\medskipamount]
+Version 1.7.4 for Equalizer 0.9.1+svn\\[\medskipamount]
 %  \today
@@
 Engineering contributions by Maxim Makhinya, Jonas B\"osch, Christian
-Marten, Sarah Amsellem, Patrick Bouchaud and Philippe Robert.
+Marten, Sarah Amsellem, Patrick Bouchaud, Philippe Robert and Robert Hauck.
 
@@
     1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
+    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
     \hline
@@
 programmer implements serialization and deserialization of the
-distributed data.
-
-Objects are dynamically assigned to a session, which assigns
-session-unique identifiers to objects and manages the network
-communication between different instances of the same distributed
-object.
-
-The master version of a distributed object is registered with a
-\textsf{eq::net::Session}, which assigns a session-unique identifier to
-the object. This identifier can be used by other nodes to map their
-instance of the object, thus synchronizing the object's data with the
-remotely registered master version.
-
-Distributed objects can be static (immutable) or dynamic. Dynamic
-objects are versioned.
-
-The \textsf{eqPly} example uses static distributed objects to provide
-initial data and the model to all rendering nodes, as well as a
-versioned object to provide frame-specific data such as the camera
-position to the rendering methods.
-
-\subsubsection{Common Usage}
-
-\begin{wrapfigure}{r}{.382\textwidth}
-%\begin{figure}[h!t]
-  \subfigure[]{\label{fObjectDist1}
-    \includegraphics[width=.382\textwidth]{images/umlObjectSubclass}}\hfil
-  \subfigure[]{\label{fObjectDist2}
-    \includegraphics[width=.382\textwidth]{images/umlObjectProxy}}\hfil
-  \subfigure[]{\label{fObjectDist3}
-    \includegraphics[width=.382\textwidth]{images/umlObjectMI}}%
-  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
-      Proxies or Multiple Inheritance}}
-\end{wrapfigure}
-Distributed objects are addressed using session-unique identifiers,
-because pointers to other objects cannot be distributed directly; they
-have no meaning on remote nodes. The session used to register
-distributed objects is normally the \textsf{eq::Config}, which is
-sub-classed from \textsf{eq::net::Session}.
-
-The entry point for shared data on a render client is the identifier
-passed by the application to \textsf{Config::init}. This identifier
-typically contains the identifier of a static distributed object, and is
-passed by Equalizer to all \textsf{configInit} task methods. Normally
-this initial object is mapped by the render clients in
-\textsf{Node::configInit}. It typically contains identifiers of other
-shared data objects.
-
-The distributed data objects referenced by the initial data object are
-often versioned objects, to keep them in sync with the rendered
-frames. Similar to the initial identifier passed to
-\textsf{Config::init}, an object identifier or object version can be
-passed to \textsf{Config::startFrame}. Equalizer will pass this
-identifier to all \textsf{frameStart} task methods. In \textsf{eqPly},
-the frame-specific data, e.g., the global camera data, is versioned. The
-frame data identifier is passed in the initial data, and the frame data
-version is passed with each new frame request.
-
-There are multiple ways of implementing data distribution for an
-existing C++ class hierarchy:
-\begin{description}
-\item[Subclassing] The classes to be distributed inherit from
-  \textsf{eq::net::Object} and implement the serialization methods. This
-  approach is recommended if the source code of existing classes can be
-  modified. It is used for \textsf{eqPly::InitData} and
-  \textsf{eqPly::FrameData}. (\fig{fObjectDist1})
-\item[Proxies] For each object to be distributed, a proxy object is
-  created which manages data distribution for its associated
-  object. This requires the application to track changes on the object
-  separately from the object itself. The model data distribution of
-  eqPly is using this pattern. (\fig{fObjectDist2})
-\item[Multiple Inheritance] A new class inheriting from the class to be
-  distributed and from \textsf{eq::net::Object} implements the data
-  distribution. This requires the application to instantiate a different
-  type of object instead of the existing object, and to create wrapper
-  methods in the superclass calling the original method and setting the
-  appropriate dirty flags. This pattern is not used in
-  eqPly. (\fig{fObjectDist3})
-\end{description}
-
-\subsubsection{Change Handling}
-
-Equalizer determines the way changes are to be handled by calling
-\textsf{Object::get\-Change\-Type} during the registration of the master
-version of a distributed object. The change type determines the memory
-footprint and the contract for the serialization methods. The following
-change types are possible:
-
-\begin{description}
-  \item[STATIC] The object is not versioned. The instance data is
-    serialized whenever a new slave instance is mapped. No additional
-    data is stored.
-  \item[INSTANCE] The object is versioned, and the instance and delta
-    data is identical, that is, only instance data is
-    serialized. Previous instance data is saved to be able to map old
-    versions.
-  \item[DELTA] The object is versioned, and the delta data is typically
-    smaller than the instance data. Both the delta and instance data are
-    serialized and saved to map old versions.
-  \item[UNBUFFERED] The object is versioned. No data is stored, and no
-    previous versions can be mapped. The instance data is serialized
-    whenever a new slave instance is mapped. The application can choose
-    to use a different, more optimal implementation to pack deltas.
-\end{description}
+distributed data. \sref{sNetObject2} covers distributed objects in detail.
+
+Distributed objects can be static (immutable) or dynamic. Dynamic objects are
+versioned. The \textsf{eqPly} example uses static distributed objects to provide
+initial data and the model to all rendering nodes, as well as a versioned object
+to provide frame-specific data such as the camera position to the rendering
+methods.
 
@@
 \begin{itemize}
-\item Only the master instance of the object is writable, that is,
-  \textsf{eq::net::Object::com\-mit} can be called only on the master
-  instance to generate a new version.
+\item Only the master instance of the object generates new versions, that is,
+  \textsf{eq::net::Object::com\-mit} should only be called only on the master
+  instance. It is possible to commit on slave instances, but special care has to
+  be taken to handle possible conflicts. \sref{sSlaveCommit} covers slave object
+  commits in detail.
 \item Slave instance versions can only be advanced, that is,
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-%\begin{figure}[ht!]\center
+%\begin{wrapfigure}{r}{.618\textwidth}
+\begin{figure}[ht!]\center
   \includegraphics[width=.618\textwidth]{images/objects.pdf}
   {\caption{\label{fObjects}Scene Graph Distribution}}
-\end{wrapfigure}
+\end{figure}
 Applications distributing a dynamic scene graph use the frame data
@@
   \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/osWindow.pdf}
-  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}\vspace{-4ex}
+  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}
 \end{wrapfigure}
@@
     {
-        READBACK = eq::ConfigEvent::USER,
+        READBACK = eq::Event::USER,
         ASSEMBLE
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
+  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
+\end{wrapfigure}
 Changing view frustum parameters is typically done for non-fullscreen
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\label{fDirectSend}Direct Send Compositing}}
+\end{wrapfigure}
 In direct send compositing, each rendering channel is also responsible
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
-  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
-\end{wrapfigure}
 The frame synchronization through the frame data allows the input frame
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
+  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
+\end{wrapfigure}
 The offset of input and output frames characterizes the position of the
@@
   \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
+  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}\vspace{-2ex}
 \end{wrapfigure}
@@
 
-\subsection{Session}
+\subsection{Sessions}
 
@@
 
-\subsection{\label{sNetObject2}Object}
-
-Distributed objects provide powerful, object-oriented data
-distribution for C++ objects. Their functionality and an example use
-case for parallel rendering has been described in \sref{sNetObject}.
-
-The master instance of an \textsf{eq::net::Object} is registered with a
-session, which assigns a session-unique identifier to the object. Using
-this identifier, other nodes can map their slave instance of the object
-to their local proxy of the session.
+\subsection{\label{sNetObject2}Objects}
+
+Distributed objects provide powerful, object-oriented data distribution for C++
+objects. They facilitate the implementation of data distribution in a cluster
+environment. Their functionality and an example use case for parallel rendering
+has been described in \sref{sNetObject}.
+
+Distributed objects are created by subclassing from \textsf{eq::Object} or
+\textsf{eq::net::Object}. The application programmer implements serialization
+and deserialization of the distributed data. Objects are dynamically assigned to
+a session, which assigns session-unique identifiers to objects and manages the
+network communication between different instances of the same distributed
+object.
+
+The master version of a distributed object is registered with a
+\textsf{eq::net::Session}, which assigns a session-unique identifier to
+the object. This identifier can be used by other nodes to map their
+instance of the object, thus synchronizing the object's data with the
+remotely registered master version.
+
+Distributed objects can be static (immutable) or dynamic. Dynamic objects are
+versioned. New versions are commited from the master instance, which sends
+the delta between the previous and current version to all mapped slave
+objects. The slave objects sync the queued deltas when they need a version.
+
+\subsubsection{Common Usage for Parallel Rendering}
+
+Distributed objects are addressed using session-unique identifiers,
+because pointers to other objects cannot be distributed directly; they
+have no meaning on remote nodes. The session used to register
+distributed objects is normally the \textsf{eq::Config}, which is
+sub-classed from \textsf{eq::net::Session}.
+
+The entry point for shared data on a render client is the identifier
+passed by the application to \textsf{Config::init}. This identifier
+typically contains the identifier of a static distributed object, and is
+passed by Equalizer to all \textsf{configInit} task methods. Normally
+this initial object is mapped by the render clients in
+\textsf{Node::configInit}. It normally contains identifiers of other
+shared data objects.
+
+The distributed data objects referenced by the initial data object are
+often versioned objects, to keep them in sync with the rendered
+frames. Similar to the initial identifier passed to
+\textsf{Config::init}, an object identifier or object version can be
+passed to \textsf{Config::startFrame}. Equalizer will pass this
+identifier to all \textsf{frameStart} task methods. In \textsf{eqPly},
+the frame-specific data, e.g., the global camera data, is versioned. The
+frame data identifier is passed in the initial data, and the frame data
+version is passed with each new frame request.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+%\begin{figure}[h!t]
+  \subfigure[]{\label{fObjectDist1}
+    \includegraphics[width=.382\textwidth]{images/umlObjectSubclass}}\hfil
+  \subfigure[]{\label{fObjectDist2}
+    \includegraphics[width=.382\textwidth]{images/umlObjectProxy}}\hfil
+  \subfigure[]{\label{fObjectDist3}
+    \includegraphics[width=.382\textwidth]{images/umlObjectMI}}%
+  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
+      Proxies or Multiple Inheritance}}
+\end{wrapfigure}
+There are multiple ways of implementing data distribution for an
+existing C++ class hierarchy, such as a scene graph:
+\begin{description}
+\item[Subclassing] The classes to be distributed inherit from
+  \textsf{eq::net::Object} and implement the serialization methods. This
+  approach is recommended if the source code of existing classes can be
+  modified. It is used for \textsf{eqPly::InitData} and
+  \textsf{eqPly::FrameData}. (\fig{fObjectDist1})
+\item[Proxies] For each object to be distributed, a proxy object is
+  created which manages data distribution for its associated
+  object. This requires the application to track changes on the object
+  separately from the object itself. The model data distribution of
+  eqPly is using this pattern. (\fig{fObjectDist2})
+\item[Multiple Inheritance] A new class inheriting from the class to be
+  distributed and from \textsf{eq::net::Object} implements the data
+  distribution. This requires the application to instantiate a different
+  type of object instead of the existing object, and to create wrapper
+  methods in the superclass calling the original method and setting the
+  appropriate dirty flags. This pattern is not used in
+  eqPly. (\fig{fObjectDist3})
+\end{description}
+
+\subsubsection{Change Handling and Serialization}
+
+Equalizer determines the way changes are to be handled by calling
+\textsf{Object::get\-Change\-Type} during the registration of the master version
+of a distributed object. The change type determines the memory footprint and the
+contract for the serialization methods. The deserialization always happens in
+the application thread causing the deserialization. The following change types
+are possible:
+
+\begin{description}
+  \item[STATIC] The object is not versioned. The instance data is serialized
+    whenever a new slave instance is mapped. The serialization happens in the
+    command thread on the master node, i.e., in parallel to the application
+    threads. Since the object's distributed data is supposed to be static, this
+    should not create any race conditions. No additional data is stored.
+  \item[INSTANCE] The object is versioned, and the instance and delta data is
+    identical, that is, only instance data is serialized. The serialization
+    always happens before \textsf{Session::registerObject} or
+    \textsf{Object::commit} returns. Previous instance data is saved to be able
+    to map old versions.
+  \item[DELTA] The object is versioned, and the delta data is typically smaller
+    than the instance data. Both the delta and instance data are serialized
+    before \textsf{Object::commit} returns. Previous instance data is saved to
+    be able to map old versions.
+  \item[UNBUFFERED] The object is versioned. No data is stored, and no previous
+    versions can be mapped. The instance data is serialized from the command
+    thread whenever a new slave instance is mapped, i.e., in parallel to
+    application threads. The application has to ensure that this does not create
+    any thread conflicts. The delta data is serialized before
+    \textsf{Object::commit} returns. The application may choose to use a
+    different, more optimal implementation to pack deltas by using a different
+    implementation for \textsf{getInstanceData} and \textsf{pack}.
+\end{description}
+
+\subsubsection{\label{sSlaveCommit}Slave Object Commit}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
+  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
+\end{wrapfigure}
+Versioned slave objects may also commit data to the master instance. However,
+this requires the implementation to take special care to resolve conflicts
+during deserialization. The algorithm and constraints for slave object commits
+are explained in this section.
+
+When using \textsf{commit} on slave object instances, multiple modifications on
+the same database may happen simultaneously. In order to avoid excessive
+synchronization in the network layer, Equalizer a light-weight approach which
+requires the application to avoid or handle conflicting changes on the same
+object.
+
+When a slave object instance commits data, \textsf{pack} is called on the
+object. The data serialized by \textsf{pack} is send to the master instance of
+the object. On the master instance, \textsf{notifyNewVersion} is called from the
+command thread to signal the data availability. No new version is generated on
+the slave instance.
+
+The master instance can \textsf{sync} delta data received from slaves. The data
+is synchronized in the order it has been received from the slaves. If a specific
+order is needed, it is the application's responsibity to serialize the slave
+commits accordingly. When using \textsf{VERSION\_NEXT} as the version parameter
+to \textsf{sync}, one slave commit is unpacked. This call may block if no data
+is pending. When using \textsf{VERSION\_HEAD} all pending versions are
+unpacked. This call will never block.
+
+Synchronizing slave data on a master instance does not create a new version. The
+application has to explicitly call \textsf{commit} to pack the changed data,
+which will create a new version send to all slaves.
+
+This approach requires the application to handle or avoid, if possible, the
+following situations:
+\begin{itemize}
+\item A slave instance will always receive its own changes when the master
+  commits the slave change.
+\item A slave instance commits data which does not correspond to the last
+  version of the master instance.
+\item Two different slave instances commit changes simultaneously which may
+  result in conflicts.
+\item A master instance applies data to an already changed database.
+\item The master instance commits a new version while a slave instance is
+  modifying its instance. 
+\end{itemize}
+
+\fig{fSlaveCommit} depicts one possible communication sequence. In general, it
+is best to adhere to the following rules:
+\begin{itemize}
+\item Handle the case that a slave instance will receive its own changes again.
+\item Write your application code so that only one object instance is modified
+  at any given time. 
+\item When multiple slave instances have to modify the same object at the same
+  time, try to modify and serialize disjunct data of the object.
+\item Before modifying a slave instance, \textsf{sync} it to the head version.
+\item Write serialization code defensively, e.g.:
+  \begin{itemize}
+  \item When adding a new child to an object, check that it has not been added
+    already.
+    \item When removing a child, transmit the child's identifier instead of its
+      index for removal.
+    \item Modify a variable by serializing its value (instead of an operation on
+      it) so that the last change will win.
+  \end{itemize}
+\end{itemize}
+
+\subsubsection{Communication between Object Instances}
 
@@
 \textsf{EQ\_ID\_ANY} (the default value), a command is dispatch to all
-object instance registered to the session with the correct object
+object instances registered to the session with the correct object
 ID. Multiple object instance per node are used when different threads on
@@
 \end{wrapfigure}
-Distributed objects provide data synchronization functionality, as
-described in \sref{sNetObject}. Additionally, applications can send
-command packets by creating custom object packets with a command greater
-than \textsf{CMD\_OBJECT\_CUSTOM}.
+Distributed objects provide data synchronization functionality, as described
+above. Additionally, applications can send command packets by creating custom
+object packets with a command greater than \textsf{CMD\_OBJECT\_CUSTOM}.
 

### 2010-02-22 Developer/ProgrammingGuide/paper.tex
Programming Guide: update slave object commit section Doxygen warning fixes
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 \item Handle the case that a slave instance will receive its own changes again.
+\item Set the appropriate dirty bits when unpacking slave delta data on the
+  master object instance.
 \item Write your application code so that only one object instance is modified
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
+  {\caption{\label{fNetObject}Communication between two Objects}}
+\end{wrapfigure}
 Multiple instances of the same object (identifier) can be mapped on the
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
-  {\caption{\label{fNetObject}Communication between two Objects}}
-\end{wrapfigure}
 Distributed objects provide data synchronization functionality, as described

### 2010-06-03 Developer/ProgrammingGuide/paper.tex
PG update
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.4 for Equalizer 0.9.1+svn\\[\medskipamount]
+Version 1.7.5 for Equalizer 0.9.1+svn\\[\medskipamount]
 %  \today
@@
 \tableofcontents
-\clearpage
+%\clearpage
 \listoffigures \vfill
@@
     1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
+    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
     \hline
@@
 
+\section{\label{sClusterSetup}Setting up a Visualization Cluster}
+This section covers the setup of a visualization cluster to run Equalizer
+applications. It does not cover basic cluster management and driver
+installation. A prerequisite to the following steps is a pre-installed cluster,
+including network and graphics card drivers.
+
+\subsection{Name Resolution}
+Pay attention that your name resolution works properly, that is, the output of
+the hostname command should resolve to the same, non-local IP address on all
+machines. If this is not the case, you will have to provide the public IP
+address for all processes, in the following way:
+\begin{description}
+\item[Server] Specify the server IP as the hostname field in the connection
+  description in the server section.
+\item[Application] Start the application using the IP as the listening port,
+  e.g., \textsf{./build/Linux/bin/eqPly -- --eq-listen 192.168.0.1}
+\item[Render Clients] Specify the client IP's as the hostname field in the
+  connection description of each node section.
+\end{description}
+
+\subsection{The Server}
+The server may be started as a separate process or within the application
+process. If it is started separately, simply provide the desired configuration
+file as a parameter. It will listen on all network addresses, unless other
+connection parameters are specified in the configuration file. If the
+server is started within the application process, using the \textsf{--eq-config}
+parameter, you will have to specify a connection description for the server in
+the configuration file. Application-local servers do not create a listening
+socket by default for security reasons.
+
+\subsection{Starting Render Clients}
+Cluster configurations use multiple Equalizer nodes, where each node represents
+a separate process, typically on a different machine. These node processes have
+to be started for a visualization session, and need to communicate with each
+other.
+
+Equalizer supports pre-launched render clients and automatic render client
+launching, if configured properly. Pre-launched render clients are started
+manually or by an external script on a pre-defined address. Auto-launched render
+client are started and stopped by the Equalizer server on demand.
+
+The two mechanism can co-exist. The server will first try to connect a
+pre-launched render client on the given connection descriptions for each
+node. If this fails, he will try to auto-launch the render client. If the render
+client does not connect back to the server within a certain timeout (default one
+minute), a failure to initialize the configuration is reported back to the
+application.
+
+\subsubsection{Pre-launched Render Clients}
+Pre-launched render clients are useful if setting up auto-launching is too
+time-consuming, e.g., on Windows which does not have a standard remote login
+procedure. The following steps are to be taken to use pre-launched render
+clients:
+
+\begin{itemize}
+\item Set the \textbf{connection} hostname and port parameters of each node in
+  the configuration file.
+\item Start the \textbf{render clients} using the parameters
+  \textsf{--eq-client} and \textsf{--eq-listen}, e.g.,
+  \textsf{./build/Linux/bin/eqPly -- --eq-client --eq-listen
+  192.168.0.2:1234}. Pay attention to use the same connection parameters as in
+    the configuration file.
+\item Start the \textbf{application}. If the server is running on the same
+  machine and user as the application, the application will connect to it
+  automatically. Otherwise use the \textsf{--eq-server} parameter to specify the
+  server address.
+\end{itemize}
+
+The render clients will automatically exit when the config is closed. The
+\textsf{eqPly} example application implements the option \textsf{-r} to keep the
+render client processes resident.
+
+\subsubsection{Auto-launched Render Clients}
+To automatically launch the render clients, the server needs to know the name of
+the render client and the command to launch them without user interaction.
+
+The name of the render client is automatically set to the name of the
+application executable. This may be changed programmatically by the
+application. Normally it suffices to install the application in the same
+directory on all machines, ideally using a shared file system.
+
+The default launch command is set to ssh, which is the most common solution for
+remote logins. To allow the server to launch the render clients without user
+interaction, password-less ssh needs to be set up. Please refer to the ssh
+documentation (cf. \textsf{ssh-keygen} and \textsf{~/.ssh/authorised\_keys}) and
+verify the functionality by logging in to each machine from the host running the
+server.
+
+\subsection{Debugging}
+If your configuration does not work, simplify it as much as possible
+first. Normally this means that there is one server, application and render
+client. Failure to launch a cluster configuration often is caused by one of the
+following reasons:
+\begin{itemize}
+\item A firewall is blocking network connections.
+\item The render client can't access the GPU's on the remote host. Set up your
+  X-Server or application rights correctly. Log into the remote machine using
+  the launch command and try to execute a simple application, e.g.,
+  \textsf{glxinfo -display :0.1}.
+\item The server does not find the pre-launched render client. Verify that the
+  client is listening on the correct IP and port, and that this IP and port are
+  reachable from the server host.
+\item The server can not launch a render client. Check the server log for the
+  launch command used, and try to execute a simple application from the server
+  host using this launch command. It should run without user interaction. Check
+  that the render client is installed in the correct path. Pay attention to the
+  launch command quotes used to separate arguments on Windows. Check that the
+  same software version, including Equalizer, are installed on all machines.
+\item A client can't connect back to the application. Check the client log, this
+  is typically caused by a misconfigured host name resolution.
+\end{itemize}
+
 \clearpage

### 2010-08-21 Developer/ProgrammingGuide/paper.tex
Finalize window API for 1.0, related doxygen and code cleanups: - Make GLEWContext const - mark eq and fabric window methods in doxygen
[x] May break build
[x] May break existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
 Equalizer uses a C++ programming interface. The API is minimally
-invasive, so Equalizer imposes only a minimal, natural execution
-framework upon the application. It does not provide a scene graph, or
-interferes in any other way with the application's rendering code. The
-restructuring work enforced by Equalizer is the minimal refactoring
-needed to parallelize the application for rendering.
+invasive. Equalizer imposes only a minimal, natural execution framework upon the
+application. It does not provide a scene graph, or interferes in any other way
+with the application's rendering code. The restructuring work enforced by
+Equalizer is the minimal refactoring needed to parallelize the application for
+scalable, distributed rendering.
 
@@
 render clients.
+
+The API documentation on the website or in the header files provides a
+comprehensive documentation on individual methods, types and other elements of
+the API. Methods marked with a specific version are part of the official, public
+API and have been introduced by this Equalizer version. Reasonable care is taken
+to not break API compatibility or change the semantics of these methods within
+future Equalizer versions of the same major revision. Any eventual changes to
+the public is documented in the release notes.
+
+In addition the the official, public API Equalizer exposeses a number of
+unstable methods and, where unavoidable, internal APIs. These are clearly marked
+in the API documentation. Unstable methods may be used by the programmer, but
+their interface or functionality may change in any future Equalizer version. The
+usage of internal methods is discouraged. Undocumented or unversioned methods
+should be considered as part of the unstable API.
 
@@
   \sref{sNetObject} for distributed objects.
+\item[eq::util] The eq::util namespace provides common utility classes, which
+  often the usage of OpenGL functions. Most classes in this namespace are used
+  by the the Equalizer client library, but are usable independently from
+  Equalizer for application development.
 \item[eq] The core namespace for the Equalizer client library. The
@@
   {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
+  \vspace{-3ex}
 \end{wrapfigure}
@@
 intended to be used by developers.
-
-\fig{fNamespaces} shows the namespace layering for \textsf{eqPly} and
-\textsf{eqServer}.
 
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
+  {\caption{\label{fModelDist}Scene Data in eqPly}}
+\end{wrapfigure}
 Some applications might rely on a shared filesystem to access the data,

### 2010-08-23 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.5 for Equalizer 0.9.1+svn\\[\medskipamount]
+Version 1.7.6 for Equalizer 0.9.1+svn\\[\medskipamount]
 %  \today
@@
     1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
+    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
     \hline
@@
 \begin{description}
+\item[eq] The core namespace for the Equalizer client library. The
+  classes and their relationship in this namespace closely model the
+  configuration file format. The classes in the \textsf{eq} namespace
+  are the main subject of this Programming Guide.
+\item[eq::util] The eq::util namespace provides common utility classes, which
+  often simplify the usage of OpenGL functions. Most classes in this namespace
+  are used by the the Equalizer client library, but are usable independently
+  from Equalizer for application development.
+\item[eq::fabric] The eq::fabric namespace is the common data management and
+  transport layer between the client library, the server and the upcoming
+  administrative API. Most Equalizer client classes inherit from base classes in
+  this namespace and have all their data access methods in these base
+  classes. Furthermore, this namespace contains a \textsf{Serializable} class
+  which implements one common use pattern of object-based data distribution
+  using \textsf{eq::net::Object}.
+\item[eq::net] The Equalizer network layer provides basic functionality
+  for network communication, such as \textsf{Connection} and
+  \textsf{ConnectionSet}, as well as higher-level functionality such as
+  \textsf{Node}, \textsf{Session} and \textsf{Object}. Please refer to
+  \sref{sNetwork} for an introduction into the network layer, and to
+  \sref{sNetObject} for distributed objects.
 \item[eq::base] The eq::base namespace provides C++ classes to abstract
@@
   not subject of this Programming Guide.
-\item[eq::net] The Equalizer network layer provides basic functionality
-  for network communication, such as \textsf{Connection} and
-  \textsf{ConnectionSet}, as well as higher-level functionality such as
-  \textsf{Node}, \textsf{Session} and \textsf{Object}. Please refer to
-  \sref{sNetwork} for an introduction into the network layer, and to
-  \sref{sNetObject} for distributed objects.
-\item[eq::util] The eq::util namespace provides common utility classes, which
-  often the usage of OpenGL functions. Most classes in this namespace are used
-  by the the Equalizer client library, but are usable independently from
-  Equalizer for application development.
-\item[eq] The core namespace for the Equalizer client library. The
-  classes and their relationship in this namespace closely model the
-  configuration file format. The classes in the \textsf{eq} namespace
-  are the main subject of this Programming Guide.
 \end{description}
@@
   {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
-  \vspace{-3ex}
 \end{wrapfigure}
-Examples shipped with Equalizer are implemented in their own namespaces,
-e.g., \textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing
-classes from the \textsf{eq} namespace, with the occasional usage of
-functionality from the \textsf{eq::net} and \textsf{eq::base} namespace.
+Examples shipped with Equalizer are implemented in their own namespaces, e.g.,
+\textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing classes from
+the \textsf{eq} namespace, with the occasional usage of functionality from the
+\textsf{eq::util}, \textsf{eq::fabric}, \textsf{eq::net} and \textsf{eq::base}
+namespace.
 
 The functionality of the Equalizer server is implemented in the
-\textsf{eq::server} namespace. This namespace does not yet expose an API
-intended to be used by developers.
+\textsf{eq::server} namespace. The \textsf{eq::admin} namespace enable run-time
+administration of Equalizer servers. These namespace does not yet expose an API
+intended to be used by developers. \fig{fClientUml} provides an overview map of
+the most important classes in the Equalizer namespace, grouped by functionality.
+
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/clientUML}
+  {\caption{\label{fClientUml}Equalizer client UML map}}
+\end{figure}
 
@@
 card. All task methods for a pipe, and therefore all OpenGL commands,
-are executed from this thread. This threading model is the OpenGL
+are executed from this thread. This threading model follows the OpenGL
 `threading model', which maintains a current context for each
@@
 
-\begin{figure}[ht!]\center
-  \includegraphics[width=.9\textwidth]{images/uml}
-  {\caption{\label{fUml}UML Diagram eqPly and relevant Equalizer Classes}}
-\end{figure}
-
 In this section the source code of \textsf{eqPly} is explained in
@@
 
-All classes in the example are in the \textsf{eqPly} namespace to avoid
-type name ambiguities, in particular for the \textsf{Window} class which
-is frequently used as a type in the global namespace by windowing
-systems. \fig{fUml} shows how the most important Equalizer classes are
-used through inheritance by the \textsf{eqPly} example.
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/uml}
+  {\caption{\label{fUml}UML Diagram eqPly and relevant Equalizer Classes}}
+\end{figure}
+
+\fig{fUml} shows how the most important Equalizer classes are used through
+inheritance by the \textsf{eqPly} example. All classes in the example are in the
+\textsf{eqPly} namespace to avoid type name ambiguities, in particular for the
+\textsf{Window} class which is frequently used as a type in the global namespace
+by windowing systems.
 
@@
 rendering entities introduced in \sref{sConfig}, and classes for
-distributing data. The function and typical usage for each of the
-rendering entities is discussed in this section.
+distributing data.
+
+The function and typical usage for each of the rendering entities is discussed
+in this section. Each of these classes inherits from a base class in the
+\textsf{fabric} name\-space, which implements data distribution for the
+entity. The fabric base classes are omitted in \fig{fUml}.
 
 The distributed data classes are helper classes based on
-\textsf{eq::Object} or its parent class \textsf{eq::net::Object}. They
-illustrate the typical usage of distributed objects for static as well
-as dynamic, frame-specific data. Furthermore they are used for a basic
-scene graph distribution of the model data.
+\textsf{eq::fabric::Serializable} or its base class
+\textsf{eq::net::Object}. They illustrate the typical usage of distributed
+objects for static as well as dynamic, frame-specific data. Furthermore they are
+used for a basic scene graph distribution of the model data.
 
@@
 
-Equalizer provides distributed objects which facilitate the
-implementation of data distribution in a cluster
-environment. Distributed objects are created by subclassing from
-\textsf{eq::Object} or \textsf{eq::net::Object}. The application
-programmer implements serialization and deserialization of the
-distributed data. \sref{sNetObject2} covers distributed objects in detail.
+Equalizer provides distributed objects which facilitate the implementation of
+data distribution in a cluster environment. Distributed objects are created by
+subclassing from \textsf{eq::fabric::Serializable} or
+\textsf{eq::net::Object}. The application programmer implements serialization
+and deserialization of the distributed data. \sref{sNetObject2} covers
+distributed objects in detail.
 
@@
 \item Only the master instance of the object generates new versions, that is,
-  \textsf{eq::net::Object::com\-mit} should only be called only on the master
-  instance. It is possible to commit on slave instances, but special care has to
-  be taken to handle possible conflicts. \sref{sSlaveCommit} covers slave object
-  commits in detail.
-\item Slave instance versions can only be advanced, that is,
-  \textsf{eq::net::Object::sync( version )} with a version smaller than
-  the current version will fail.
+  \textsf{commit} should only be called only on the master instance. It is
+  possible to commit on slave instances, but special care has to be taken to
+  handle possible conflicts. \sref{sSlaveCommit} covers slave object commits in
+  detail.
+\item Slave instance versions can only be advanced, that is, \textsf{sync(
+  version )} with a version smaller than the current version will fail.
 \item Newly mapped slave instances are mapped to the oldest available
@@
   \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
-  {\caption{\label{fUMLObject}eq::Object and eq::net::Object}}
+  {\caption{\label{fUMLObject}eq::fabric::Serializable and eq::net::Object}}
 \end{wrapfigure}
-The creation of distributed, versioned objects is further facilitated by
-\textsf{eq::Object}, which implements the most common way of tracking
-data changes in versioned objects. The concept of a dirty bit mask is
-used to mark parts of the object for serialization, while preserving the
-capability to inherit objects. Other ways of implementing change
-tracking, e.g., using incarnation counters, can still be implemented by
-using \textsf{eq::net::Object} which leaves this flexibility to the
-developer. \fig{fUMLObject} shows the relationship between
-\textsf{eq::Object} and \textsf{eq::net::Object}.
-
-The \textsf{FrameData} is sub-classed from \textsf{eq::Object}, and
+The creation of distributed, versioned objects is simplified when using
+\textsf{eq::fabric::Serializable}, which implements one common way of tracking
+data changes in versioned objects. The concept of a dirty bit mask is used to
+mark parts of the object for serialization, while preserving the capability to
+inherit objects. Other ways of implementing change tracking, e.g., using
+incarnation counters, can still be implemented by using \textsf{eq::net::Object}
+which leaves all flexibility to the developer. \fig{fUMLObject} shows the
+relationship between \textsf{eq::fabric::Serializable} and
+\textsf{eq::net::Object}.
+
+The \textsf{FrameData} is sub-classed from \textsf{Serializable}, and
 consequently tracks its changes by setting the appropriate dirty bit
 whenever it is changed. The serialization methods are called by the
-\textsf{eq::Object} with the dirty bit mask needed to serialize all
+\textsf{eq::fabric::Serializable} with the dirty bit mask needed to serialize all
 data, or with the dirty bit mask of the changes since the last
@@
 {
-    DIRTY_CAMERA  = eq::Object::DIRTY_CUSTOM << 0,
-    DIRTY_FLAGS   = eq::Object::DIRTY_CUSTOM << 1,
-    DIRTY_VIEW    = eq::Object::DIRTY_CUSTOM << 2,
-    DIRTY_MESSAGE = eq::Object::DIRTY_CUSTOM << 3,
+    DIRTY_CAMERA  = eq::fabric::Serializable::DIRTY_CUSTOM << 0,
+    DIRTY_FLAGS   = eq::fabric::Serializable::DIRTY_CUSTOM << 1,
+    DIRTY_VIEW    = eq::fabric::Serializable::DIRTY_CUSTOM << 2,
+    DIRTY_MESSAGE = eq::fabric::Serializable::DIRTY_CUSTOM << 3,
 };
@@
 {
-    eq::Object::serialize( os, dirtyBits );
+    eq::fabric::Serializable::serialize( os, dirtyBits );
     if( dirtyBits & DIRTY_CAMERA )
@@
 {
-    eq::Object::deserialize( is, dirtyBits );
+    eq::fabric::Serializable::deserialize( is, dirtyBits );
     if( dirtyBits & DIRTY_CAMERA )
@@
 
-To dispatch events, Equalizer 'pumps' the native events. On WGL and GLX,
-this happens on each thread with windows, whereas on AGL it happens on
-the main thread and each pipe thread. By default, Equalizer pumps these
-events automatically for the application in-between executing task
-methods.
-
-The method \textsf{Pipe::useMessagePump} can be overridden to return
-\textsf{false} to not enable message dispatch for the respective thread.
-
-The method \textsf{useMessagePump} is called by Equalizer during
-application and pipe thread initialization (before configInit!) to
-determine if the pipe's thread shall automatically dispatch OS
-events. For AGL, this affects the node threads and pipe thread, since
-the node thread message pump needs to dispatch the events to the pipe
-thread event queue.
+To dispatch events, Equalizer 'pumps' the native events. On WGL and GLX, this
+happens on each thread with windows, whereas on AGL it happens on the main
+thread and on each pipe thread. By default, Equalizer pumps these events
+automatically for the application in-between executing task methods.
+
+The method \textsf{Pipe::createMessagePump} is called by Equalizer during
+application and pipe thread initialization (before \textsf{Pipe::configInit}!)
+to create a message pump for the given pipe. For AGL, this affects the node
+thread and pipe threads, since the node thread message pump needs to dispatch
+the events to the pipe thread event queue. Custom message pumps may be
+implemented, and it is valid to return no message pump to disable message
+dispatch for the respective pipe.
 
@@
 
-Layout API entities (Canvas, Segment, Layout, View) are sub-classed like
-all other Equalizer entities using the \textsf{NodeFactory}. Equalizer
-registers the master instance of these entities on the application node,
-so that the application can change mutable parameters and distribute its
-own data to the render clients. \fig{fViewUML} shows the UML class
-hierarchy for the \textsf{eqPly::View}.
-
+Layout API entities (Canvas, Segment, Layout, View) are sub-classed like all
+other Equalizer entities using the \textsf{NodeFactory}. Equalizer registers the
+master instance of these entities on the server node. Mutable parameters, e.g.,
+the active layout of a canvas, are distributed using slave object commits
+(cf. \sref{sSlaveCommit}). Application-specific data can be attached to a view
+using a distributable \textsf{UserData} object. \fig{fViewUML} shows the UML
+class hierarchy for the \textsf{eqPly::View}.
+
+\begin{wrapfigure}{r}{.612\textwidth}
+  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
+  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
+\end{wrapfigure}
 The render clients can access a slave instance of the view using
@@
 
-The most common entity to subclass is the \textsf{View}, since the
-application amends it by view-specific data. The view, like all other
-layout entities, is indirectly sub-classed from \textsf{eq::Object}
-through \textsf{eq::Frustum}. The design and usage of
-\textsf{eq::Object} is explained in \sref{sEqObject}.
-
-In \textsf{eqPly}, the application-specific data is the model
-identifier. A new dirty bit for the model is defined, and the model
-identifier is serialized when requested:
+The most common entity to subclass is the \textsf{View}, since the application
+often amends with view-specific application data. A view might have a
+distributable user data object, which has to inherit from
+\textsf{eq::net::Object} to be committed and synchronized with the associated
+object.
+
+Externalizing the data distribution in a \textsf{UserData} object is necessary
+since the server holds the master instance of a view. The server contains no
+application code, and would not be able to (de-)serialize the application data
+of a view. Instead, the user data object has its master and slave instances only
+in application processes, and the server only handles the identifier and version
+of this object.
+
+In \textsf{eqPly}, the application-specific data is the model identifier and the
+number of anti-aliasing steps to be rendered when idle. This data is distributed
+by a \textsf{View::Proxy} object which serves as the view's user data. The proxy
+defines the necessary dirty bits and serializes the data:
 
@@
 {
-    DIRTY_MODEL       = eq::View::DIRTY_CUSTOM << 0,
+    DIRTY_MODEL       = eq::fabric::Serializable::DIRTY_CUSTOM << 0,
+    DIRTY_IDLE        = eq::fabric::Serializable::DIRTY_CUSTOM << 1
 };
 
-void View::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
-{
-    eq::View::serialize( os, dirtyBits );
+void View::Proxy::serialize( eq::net::DataOStream& os,
+                             const uint64_t dirtyBits )
+{
     if( dirtyBits & DIRTY_MODEL )
-        os << _modelID;
-}
-
-void View::deserialize( eq::net::DataIStream& is, const uint64_t dirtyBits )
-{
-    eq::View::deserialize( is, dirtyBits );
+        os << _view->_modelID;
+    if( dirtyBits & DIRTY_IDLE )
+        os << _view->_idleSteps;
+}
+
+void View::Proxy::deserialize( eq::net::DataIStream& is,
+                               const uint64_t dirtyBits )
+{
     if( dirtyBits & DIRTY_MODEL )
-        is >> _modelID;
-}
-
+        is >> _view->_modelID;
+    if( dirtyBits & DIRTY_IDLE )
+    {
+        is >> _view->_idleSteps;
+        if( isMaster( ))
+            setDirty( DIRTY_IDLE ); // redistribute slave settings
+    }
+}
+\end{lstlisting}}
+
+The \textsf{eqPly::View} sets its \textsf{Proxy} as the user data object in the
+constructor. By default, the master instance of the view's user data is on the
+application instance of the view. This may be changed by overriding
+\textsf{hasMasterUserData}. The proxy object registration, mapping and
+synchronization is fully handled be the \textsf{fabric} layer, no further
+handling has to be done by the application:
+
+{\footnotesize\begin{lstlisting}
+View::View( eq::Layout* parent )
+    : [...]
+{
+    setUserData( &_proxy );
+}
 void View::setModelID( const uint32_t id )
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
-  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
-\end{wrapfigure}
 Changing view frustum parameters is typically done for non-fullscreen
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
+  {\caption{\label{fNetSession}Communication between two Sessions}}
+\end{wrapfigure}
 The \textsf{eq::net::Session} gathers provides higher-level
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
-  {\caption{\label{fNetSession}Communication between two Sessions}}
-\end{wrapfigure}
 Sessions instances can communicate to each other directly using an
@@
 
-Distributed objects are created by subclassing from \textsf{eq::Object} or
+Distributed objects subclass from \textsf{eq::fabric::Serializable} or
 \textsf{eq::net::Object}. The application programmer implements serialization
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-%\begin{figure}[h!t]
+%\begin{wrapfigure}{r}{.382\textwidth}
+\begin{figure}[h!t]
   \subfigure[]{\label{fObjectDist1}
-    \includegraphics[width=.382\textwidth]{images/umlObjectSubclass}}\hfil
+    \includegraphics[width=.3\textwidth]{images/umlObjectSubclass}}\hfil
   \subfigure[]{\label{fObjectDist2}
-    \includegraphics[width=.382\textwidth]{images/umlObjectProxy}}\hfil
+    \includegraphics[width=.3\textwidth]{images/umlObjectProxy}}\hfil
   \subfigure[]{\label{fObjectDist3}
-    \includegraphics[width=.382\textwidth]{images/umlObjectMI}}%
+    \includegraphics[width=.3\textwidth]{images/umlObjectMI}}%
   {\caption{\label{fObjectDist}Object Distribution using Subclassing,
       Proxies or Multiple Inheritance}}
-\end{wrapfigure}
+\end{figure}
 There are multiple ways of implementing data distribution for an
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
-  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
-\end{wrapfigure}
 Versioned slave objects may also commit data to the master instance. However,
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
+  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
+\end{wrapfigure}
 When using \textsf{commit} on slave object instances, multiple modifications on
@@
 
-\subsubsection{\label{sEqObject}eq::Object}
-
-Equalizer uses the \textsf{eq::net::Object} data distribution mechanism
-for implementing data distribution between the server, application and
-render clients. To do this, \textsf{eq::Object} implements one usage
-pattern for the \textsf{eq::net::Object}, allowing inheritance with data
+\subsubsection{\label{sEqObject}eq::fabric::Serializable}
+
+Equalizer uses the \textsf{eq::net::Object} data distribution mechanism for
+implementing data distribution between the server, application and render
+clients. Based on dirty bits, \textsf{eq::fabric::Serializable} implements one
+usage pattern for the \textsf{eq::net::Object}, allowing inheritance with data
 distribution. It is easier to use, but imposes one typical way to implement data
@@
 
-The \textsf{eq::Object} data distribution is based on the concept of
-dirty bits. Dirty bits form a 64-bit mask which marks which parts of the
+The \textsf{eq::fabric::Serializable} data distribution is based on the concept
+of dirty bits. Dirty bits form a 64-bit mask which marks which parts of the
 object have to be distributed during the next commit.
 
-For serialization, the default \textsf{eq::net::Object} serialization
-functions are implemented by \textsf{eq::Object}, which (de-)serializes
-and resets the dirty bits, and calls \textsf{serialize} or
-\textsf{deserialize} with the bit mask specifying which data has to be
-transmitted or received. During a commit or sync, the current dirty bits
-are given, whereas during object mapping all dirty bits are passed to
-the serialization methods.
-
-To use \textsf{eq::Object}, the following steps have to be taken:
+For serialization, the default \textsf{eq::net::Object} serialization functions
+are implemented by \textsf{eq::fabric::Serializable}, which (de-)serializes and
+resets the dirty bits, and calls \textsf{serialize} or \textsf{deserialize} with
+the bit mask specifying which data has to be transmitted or received. During a
+commit or sync, the current dirty bits are given, whereas during object mapping
+all dirty bits are passed to the serialization methods.
+
+To use \textsf{eq::fabric::Serializable}, the following steps have to be taken:
 
 \begin{description}
-\item[Inherit from \textsf{eq::Object}:] The base class will provide the
-  dirty bit management and call serialize and deserialize
-  appropriately. By overriding \textsf{getChangeType}, the default
+\item[Inherit from \textsf{eq::fabric::Serializable}:] The base class will
+  provide the dirty bit management and call serialize and deserialize
+  appropriately. By overriding \textsf{get\-Chan\-ge\-Type}, the default
   versioning strategy might be changed.
 \item[Define new dirty bits:] Define dirty bits for each data item by
-  starting at \textsf{Object::DIRTY\_CUSTOM}, shifting this value
+  starting at \textsf{Serializable::DIRTY\_CUSTOM}, shifting this value
   consecutively by each new dirty bit.
@@
 
-The inheritance chain \textsf{eq::View} $\rightarrow$
-\textsf{eq::Frustum} $\rightarrow$ \textsf{eq::Object} $\rightarrow$
-\textsf{eq::net::Object} serves as sample implementation for this
-process. The \textsf{eq::Object} base class provides the facility to set
-a name on the object, which is automatically distributed.
-
-The registration and mapping of \textsf{eq::Object}s is done in the same
-way as for \textsf{eq::net::Object}s, which has been described in
-\sref{sNetObject2}. Equalizer objects derived from \textsf{eq::Object},
-e.g., \textsf{eq::View}, are typically registered and mapped by
-Equalizer.
+The Equalizer entities use the common base class
+\textsf{eq::fabric::Object}. This class manages common entity data, and is not
+intended to by subclassed directly.  The inheritance chain \textsf{eq::View}
+$\rightarrow$ \textsf{eq::fabric::View} $\rightarrow$ \textsf{eq::Object}
+$\rightarrow$ \textsf{eq::fabric::Se\-rial\-izable} $\rightarrow$
+\textsf{eq::net::Object} serves as sample implementation for this process.
+
+The registration and mapping of \textsf{eq::fabric::Serializable}s is done in
+the same way as for \textsf{eq::net::Object}s, which has been described in
+\sref{sNetObject2}. Equalizer objects derived from
+\textsf{eq::fabric::Serializable}, e.g., \textsf{eq::View}, are typically
+registered, mapped and synchronized by Equalizer.
 

### 2010-10-01 Developer/ProgrammingGuide/paper.tex
Update Programming Guide with SystemPipe and SystemWindow renaming.
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.6 for Equalizer 0.9.1+svn\\[\medskipamount]
+Version 1.7.7 for Equalizer 0.9.1+svn\\[\medskipamount]
 %  \today
@@
     1.7.6   & Aug 23, 2010 & General update for fabric changes\\
+    1.7.7   & Oct 01, 2010 & SystemPipe and SystemWindow renaming\\
     \hline
@@
 context. The drawable can be an on-screen window or an off-screen
-PBuffer or FrameBuffer Object (FBO). The window uses an
-\textsf{eq::OSWindow}, which abstracts and manages
-window-system-specific handles to the drawable and context, e.g., an X11
-window \textsf{XID} and \textsf{GLXContext} for the glX window system.
+PBuffer or FrameBuffer Object (FBO).
 
@@
 
+\fig{fFrusta} illustrates the wall and projection frustum parameters.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-
-\fig{fFrusta} illustrates the wall and projection frustum parameters.
 
@@
 
+Configure one \textsf{layout} for each configuration of logical
+views. Name the layout using a unique name. Often only one layout with a
+one view is used for all canvases.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
-
-Configure one \textsf{layout} for each configuration of logical
-views. Name the layout using a unique name. Often only one layout with a
-one view is used for all canvases.
-
 Enable the layout on each desired canvas by adding it to the
@@
 
+Layouts can be switched at runtime by the application. Switching a
+layout will activate different destination channels for rendering.
+
 \begin{figure}[ht!]\center
@@
 \end{figure}
-
-Layouts can be switched at runtime by the application. Switching a
-layout will activate different destination channels for rendering.
 
@@
   \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
-  {\caption{\label{fUMLObject}eq::fabric::Serializable and eq::net::Object}}
+  {\caption{\label{fUMLObject}fabric::Serializable and net::Object}}
 \end{wrapfigure}
@@
 
+The pipe uses an \textsf{eq::SystemPipe}, which abstracts and manages
+window-system-specific code for the GPU, e.g., an X11 \textsf{Display}
+connection for the glX pipe system.
+
 \subsubsection{Initialization and Exit}
@@
 this version to update its instance of the frame data to the current
-version, and then unlocks its child entities by calling
+version, and unlocks its child entities by calling
 \textsf{startFrame}:
@@
 {
-    node->waitFrameStarted( frameNumber );
-
+    startFrame( frameNumber );
     _frameData.sync( frameID );
-    startFrame( frameNumber );
 }
@@
 
+The window uses an \textsf{eq::SystemWindow}, which abstracts and manages
+window-system-specific handles to the drawable and context, e.g., an X11 window
+\textsf{XID} and \textsf{GLXContext} for the glX window system.
+
 The window class is the natural place for the application to maintain
@@
   \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/osWindow.pdf}
-  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}
+  {\caption{\label{fOSWindow}SystemWindow UML Class Hierarchy}}
 \end{wrapfigure}
@@
 All window-system specific functionality is implemented by a
-specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
+specialization of \textsf{eq::System\-Window}. The \textsf{SystemWindow} class
 defines the minimal interface to be implemented for a new window
-system. Each \textsf{Window} uses one \textsf{OSWindow} during
+system. Each \textsf{Window} uses one \textsf{SystemWindow} during
 execution. This separation allows an easy implementation and adaption to
@@
 classes, for example the AGL, WGL and glX event handlers. The separation 
-of the OpenGL calls from the \textsf{OSWindow} permits to implement 
+of the OpenGL calls from the \textsf{SystemWindow} permits to implement 
 window systems which don't use any OpenGL context and therefore to use a 
@@
 methods. The main task method \textsf{configInit} calls first
-\textsf{configInitOSWindow}, which creates and initializes the
-\textsf{OSWindow} for this window. The \textsf{OSWindow} initialization
-code is implementation specific. If the \textsf{OSWindow} was
+\textsf{configInitSystemWindow}, which creates and initializes the
+\textsf{SystemWindow} for this window. The \textsf{SystemWindow} initialization
+code is implementation specific. If the \textsf{SystemWindow} was
 initialized successfully, \textsf{configInit} calls
@@
 
-The \textsf{OSWindow} initialization takes into account various
+The \textsf{SystemWindow} initialization takes into account various
 attributes set in the configuration file. Attributes include the size of
@@
 initialization. First, \textsf{configExitGL} is called to de-initialize
-OpenGL, followed by \textsf{configExitOSWindow} which de-initializes the
-drawable and context and deletes the \textsf{OSWindow} allocated in
-\textsf{configInitOSWindow}.
+OpenGL, followed by \textsf{configExitSystemWindow} which de-initializes the
+drawable and context and deletes the \textsf{SystemWindow} allocated in
+\textsf{configInitSystemWindow}.
 
@@
 Events are received by an event handler. The event handler finds the
-\textsf{eq::OSWindow} associated to the event. It then creates a generic
+\textsf{eq::SystemWindow} associated to the event. It then creates a generic
 \textsf{Event}, which holds the event data in an independent
 format. The original native event and this generic \textsf{Event}
-form the \textsf{OSWindowEvent}, which is passed to the concrete
-\textsf{OSWindow} for processing.
+form the \textsf{SystemWindowEvent}, which is passed to the concrete
+\textsf{SystemWindow} for processing.
 
@@
 \end{wrapfigure}
-The purpose of the \textsf{OSWindow} processing method,
+The purpose of the \textsf{SystemWindow} processing method,
 \textsf{processEvent}, is to perform window-system specific event
@@
 \textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
-resolution changes. It is also likely that an \textsf{OSPipe}
-abstraction, similar to the \textsf{OSWindow}, will be created to
-separate the window system specific code from the \textsf{eq::Pipe}.
+resolution changes.
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\label{fDirectSend}Direct Send Compositing}}
-\end{wrapfigure}
 In direct send compositing, each rendering channel is also responsible
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
+  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
+\end{wrapfigure}
 Readback and assemble operations on frames and images are designed to be
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
-  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
-\end{wrapfigure}
 The offset of input and output frames characterizes the position of the
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 DB compounds have the advantage of scaling any part of the volume
@@
 
-Please note that the support for CUDA is only enabled if the CUDA libraries are found
-in their default locations. Otherwise please adapt the makefiles accordingly.
+Please note that the support for CUDA is only enabled if the CUDA libraries are
+found in their default locations. Otherwise please adapt the build system
+accordingly.
 
@@
 
-Equalizer automatically initializes a CUDA context on each pipe (using the specified
-device number) whenever \textsf{hint\_cuda\_GL\_interop} is set. The initialization is 
-currently done using the CUDA runtime API in order to support device emulation. 
-On the other side, if this hint is not set the OpenGL interop feature is not enabled, and 
-furthermore, the GPU which executes CUDA code defaults to device 0.
+Equalizer automatically initializes a CUDA context on each pipe (using the
+specified device number) whenever \textsf{hint\_cuda\_GL\_interop} is set. The
+initialization is currently done using the CUDA runtime API to support device
+emulation. On the other side, if this hint is not set the OpenGL interop feature
+is not enabled, and furthermore, the GPU which executes CUDA code defaults to
+device 0.
 

### 2010-10-04 Developer/ProgrammingGuide/paper.tex
affinity WAR documentation update
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.7 for Equalizer 0.9.1+svn\\[\medskipamount]
+Version 1.7.8 for Equalizer 0.9.1+svn\\[\medskipamount]
 %  \today
@@
     1.7.7   & Oct 01, 2010 & SystemPipe and SystemWindow renaming\\
+    1.7.8   & Oct 04, 2010 & Document GPU affinity workaround hint\\
     \hline
@@
   \hline
+  \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_AFFINITY\\
+  \textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\textbar \ EXTERN\\
+  \textbf{Default} & AUTO\\
+  \textbf{Details} & Determines how the window handles the affinity context (WGL
+  only). This hint is a workaround for NVidia drivers greater than 200, which
+  contain a faulty implementation of the WGL\_NV\_gpu\_affinity extension. When
+  set to AUTO, the hint is interpreted like EXTERN for windows and pbuffers with
+  these drivers, in all other cases it is interpreted like ON. When set to ON,
+  an affinity DC is used for all drawable types. When set to EXTERN, a normal DC
+  is used to create the drawable, but an affinity DC is created and can be used
+  by the application to render into a FBO. This hint may be ignored in later
+  versions of Equalizer, once this workaround has become obsolete.\\
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
   \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_STEREO\\
@@
     {
+        hint_affinity       OFF | ON | AUTO | EXTERN
         hint_stereo         OFF | ON | AUTO

### 2010-10-04 Developer/ProgrammingGuide/paper.tex
Programming Guide: minor update
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 \end{textblock} {\Large The official reference for developing and
-  deploying parallel, scalable OpenGL\texttrademark\ applications based on
+  deploying parallel, scalable OpenGL\texttrademark\ applications using
   the Equalizer parallel rendering framework}\\\vspace{1cm}
@@
   versions of Equalizer, once this workaround has become obsolete.\\
+  \hline
 \end{tabularx}\\\vfill
@@
   \hline
-\end{tabularx}\\%\vfill
+\end{tabularx}\\\vfill
 
@@
   \textbf{Name} & EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
-  \textbf{Value} & QUAD \textbar \ ANAGLYPH\\
-  \textbf{Default} & QUAD\\
+  \textbf{Value} & AUTO \textbar QUAD \textbar \ ANAGLYPH\\
+  \textbf{Default} & AUTO\\
   \textbf{Details} & Selects the algorithm used for stereo
@@
   \textsf{glColorMask} to mask colors for individual eye passes, used in
-  conjunction with colored glasses.\\
+  conjunction with colored glasses. AUTO selects QUAD if the drawable is capable
+  of active stereo rendering, otherwise ANAGLYPH is selected\\
   \hline
-\end{tabularx}\\%\vfill
+\end{tabularx}\\\vfill
 
@@
     {
-        stereo_mode                  QUAD | ANAGLYPH    # default QUAD
+        stereo_mode                  AUTO | QUAD | ANAGLYPH    # default AUTO
         stereo_anaglyph_left_mask    [ RED GREEN BLUE ] # default red

### 2010-11-03 Developer/ProgrammingGuide/paper.tex
Error codes: Programming Guide update
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.8 for Equalizer 0.9.1+svn\\[\medskipamount]
+Version 1.7.9 for Equalizer 0.9.1+svn\\[\medskipamount]
 %  \today
@@
     1.7.8   & Oct 04, 2010 & Document GPU affinity workaround hint\\
+    1.7.9   & Nov 03, 2010 & Add \sref{sErrorHandling}\\
     \hline
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/processes.pdf}
-  {\caption{\label{fProcesses}Equalizer Processes}}
-\end{wrapfigure}
 The render client implements the rendering part of an application. Its
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/processes.pdf}
+  {\caption{\label{fProcesses}Equalizer Processes}}
+\end{wrapfigure}
 The application can also be a rendering client, in which case it can
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
+  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
+\end{wrapfigure}
 Stereo compounds, as shown in \fig{fStereoCmp}\footnote{3D model
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
-  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
-\end{wrapfigure}
 Eye compounds in Equalizer are configured using the \textsf{eye}
@@
 
-Pixel compounds are similar to 2D compounds. The frusta of the
-source rendering units are modified so that each unit renders an evenly
-distributed subset of pixels, as shown in \fig{fPixel}\footnote{3D model courtesy of AVS, USA.}
+Pixel compounds are similar to 2D compounds. While 2D compounds decompose the
+screen space into contiguous regions, pixel compounds assign one pixel of a
+regular kernel to each resource. The frusta of the source rendering units are
+modified so that each unit renders an evenly distributed subset of pixels, as
+shown in \fig{fPixel}\footnote{3D model courtesy of AVS, USA.}
 
 As 2D compounds, pixels compounds have low, constant IO requirements for
-the pixel transfers during recomposition.
+the pixel transfers during recomposition. They are naturally load-balanced for
+fill-limited operations, while they do not decompose the geometry processing at
+all.
 
@@
     {
-        EQERROR << "Config initialization failed: " 
-                << config->getErrorMessage() << endl;
         server->releaseConfig( config );
@@
 
+\subsection{\label{sErrorHandling}Error Handling}
+
+All Equalizer entities use an error code to report error conditions during
+various operations. This error code is reported back from the render processes
+to the application config instance. The last error is propagated from the failed
+resource to the config object, where it can be queried by the application using
+\textsf{Config::getError}.
+
+Each error emitted by Equalizer code has a textual description. This string is
+automatically used for printing an \textsf{Error} on an \textsf{std::ostream},
+and can be queried from the \textsf{base::ErrorRegistry}, which is accessible
+from \textsf{base::Global}.
+
+Applications can register additional error codes and strings. The
+\textsf{eVolve} example uses this to report if required OpenGL extensions are
+missing. The registration of new error codes is not thread-safe, that is, no
+other thread should use the error registry when it is modified. It is therefore
+strongly advised to register application-specific errors before
+\textsf{eq::init} and clear them after \textsf{eq::exit}:
+
+{\footnotesize\begin{lstlisting}
+int main( const int argc, char** argv )
+{
+    ...
+    // 2. Equalizer initialization
+    NodeFactory nodeFactory;
+    eVolve::initErrors();
+
+    if( !eq::init( argc, argv, &nodeFactory ))
+    ...
+
+    eq::exit();
+    eVolve::exitErrors();
+    return ret;
+}
+\end{lstlisting}}
+
+When the application returns false from a \textsf{configInit} task method due to
+an application-specific error, it should set an error code using
+\textsf{setError}. Application-specific errors can have any value equal or
+greater than \textsf{eq::ERROR\_CUSTOM}:
+
+{\footnotesize\begin{lstlisting}
+/** Defines errors produced by eVolve. */
+enum Error
+{
+    ERROR_EVOLVE_ARB_SHADER_OBJECTS_MISSING = eq::ERROR_CUSTOM,
+    ERROR_EVOLVE_EXT_BLEND_FUNC_SEPARATE_MISSING,
+    ERROR_EVOLVE_ARB_MULTITEXTURE_MISSING,
+    ERROR_EVOLVE_LOADSHADERS_FAILED,
+    ERROR_EVOLVE_LOADMODEL_FAILED
+};
+
+bool Window::configInitGL( const uint32_t initID )
+{
+    ...
+    if( !GLEW_ARB_shader_objects )
+    {
+        setError( ERROR_EVOLVE_ARB_SHADER_OBJECTS_MISSING );
+        return false;
+    }
+    ...
+}
+\end{lstlisting}}
+
+It is not required to register an error string for each application-specific
+error. Registering a string will however cause error printing to use this string
+instead of using only the numerical error value. Applications may also redefine
+existing error strings to their convenience, e.g., for internationalization
+purposes:
+
+{\footnotesize\begin{lstlisting}
+namespace
+{
+struct ErrorData
+{
+    const uint32_t code;
+    const std::string text;
+};
+
+ErrorData _errors[] = {
+    { ERROR_EVOLVE_ARB_SHADER_OBJECTS_MISSING,
+      "GL_ARB_shader_objects extension missing" },
+    { ERROR_EVOLVE_EXT_BLEND_FUNC_SEPARATE_MISSING,
+      "GL_ARB_shader_objects extension missing" },
+    { ERROR_EVOLVE_ARB_MULTITEXTURE_MISSING,
+      "GL_ARB_shader_objects extension missing" },
+    { ERROR_EVOLVE_LOADSHADERS_FAILED, "Can't load shaders" },
+    { ERROR_EVOLVE_LOADMODEL_FAILED, "Can't load model" },
+
+    { 0, "" } // last!
+};
+}
+
+void initErrors()
+{
+    eq::base::ErrorRegistry& registry = eq::base::Global::getErrorRegistry();
+
+    for( size_t i=0; _errors[i].code != 0; ++i )
+        registry.setString( _errors[i].code, _errors[i].text );
+}
+
+void exitErrors()
+{
+    eq::base::ErrorRegistry& registry = eq::base::Global::getErrorRegistry();
+
+    for( size_t i=0; _errors[i].code != 0; ++i )
+        registry.eraseString( _errors[i].code );
+}
+\end{lstlisting}}
+
 \subsection{\label{sThreads}Thread Synchronization}
@@
 
+The application or node main thread is the primary thread of each
+process and executes the \textsf{main} function. The application and
+render clients initialize the local node for communications with other
+nodes, including the server, using \textsf{Client::initLocal}.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The application or node main thread is the primary thread of each
-process and executes the \textsf{main} function. The application and
-render clients initialize the local node for communications with other
-nodes, including the server, using \textsf{Client::initLocal}.
-
 During this initialization, Equalizer creates and manages two threads
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
 The application has extended control over the task synchronization
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
+  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
+\end{wrapfigure}
 \fig{fFrameSync} illustrates the local frame synchronization.
@@
 
-Since window initialization is notoriously error-prone and hard to
-debug, the sample implementation propagates the reason for errors
-from the render clients back to the application. The \textsf{Pipe} and
-\textsf{Window} classes have a \textsf{setErrorMessage} method, which is
-used to set an error string. This string is passed to the
-\textsf{Config} instance on the application node, where it can be
-retrieved using \textsf{getErrorMessage}.
+Since window initialization is notoriously error-prone and hard to debug, the
+sample implementation propagates the reason for errors from the render clients
+back to the application. The \textsf{Pipe} and \textsf{Window} classes have a
+\textsf{setError} method, which is used to set an error code. This string is
+passed to the \textsf{Config} instance on the application node, where it can be
+retrieved using \textsf{getError}. The \sref{sErrorHandling} explains error
+handling in more detail.
 
@@
 
+    // Early glxew init to have function pointers for init code
+    glXMakeCurrent( _xDisplay, None, 0 );
+    glxewInit();
+
     if( !context )
@@
 
-    if( success && !_xDrawable )
-    {
-        _window->setErrorMessage( 
-            "configInitGLXDrawable did not set a X11 drawable");
+    if( !success || !_xDrawable )
+    {
+        if( getError() == ERROR_NONE )
+            setError( ERROR_GLXWINDOW_NO_DRAWABLE );
         return false;
     }
+
+    makeCurrent();
+    initGLEW();
+
+    if( getIAttribute( Window::IATTR_HINT_DRAWABLE ) == FBO )
+        configInitFBO();
 
@@
     {
-        _window->setErrorMessage( "Can't create affinity dc" );
+        setError( ERROR_WGL_CREATEAFFINITYDC_FAILED );
         return false;
@@
         setWGLDC( 0, WGL_DC_NONE );
-        _window->setErrorMessage(
-            "configInitWGLDrawable did not set a WGL drawable" );
+        setError( ERROR_WGLWINDOW_NO_DRAWABLE );
         return false;
@@
     makeCurrent();
-    _initGlew();
+    initGLEW();
 
@@
                    << "swapsync hint" << std::endl;
-    }
-
-    if( !joinNVSwapBarrier( ))
-    {
-        _window->setErrorMessage( "Joining NV_swap_group failed" );
-        return false;
     }
@@
 
-\begin{wrapfigure}{r}{.612\textwidth}
-  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
-  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
-\end{wrapfigure}
 The render clients can access a slave instance of the view using
@@
 
+\begin{wrapfigure}{r}{.612\textwidth}
+  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
+  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
+\end{wrapfigure}
 The most common entity to subclass is the \textsf{View}, since the application
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
+  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
+\end{wrapfigure}
 The frustum of a view overrides the underlying frustum of the
@@
 
-\begin{wrapfigure}{l}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
-  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
-\end{wrapfigure}
 When a window is resized, a \textsf{CHANNEL\_RESIZE} event is eventually
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
-  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
-\end{wrapfigure}
 Readback and assemble operations on frames and images are designed to be
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}\vspace{-2ex}
-\end{wrapfigure}
 The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
@@
 \begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
 \end{wrapfigure}
@@
 
+\begin{wrapfigure}{l}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 For orthographic rendering, determining the compositing order of the
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
-  {\caption{\label{fNetSession}Communication between two Sessions}}
-\end{wrapfigure}
 The \textsf{eq::net::Session} gathers provides higher-level
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
+  {\caption{\label{fNetSession}Communication between two Sessions}}
+\end{wrapfigure}
 A session provides the facilitate to allocate session-unique
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
-  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
-\end{wrapfigure}
 When using \textsf{commit} on slave object instances, multiple modifications on
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
+  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
+\end{wrapfigure}
 The master instance can \textsf{sync} delta data received from slaves. The data

### 2010-12-10 Developer/ProgrammingGuide/paper.tex
Implement RFE 3132979: Add passive stereo detection to compound auto stereo mode
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[x] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
   \textbf{Name} & EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
-  \textbf{Value} & AUTO \textbar QUAD \textbar \ ANAGLYPH\\
+  \textbf{Value} & AUTO \textbar QUAD \textbar \ ANAGLYPH \textbar \ PASSIVE\\
   \textbf{Default} & AUTO\\
   \textbf{Details} & Selects the algorithm used for stereo
-  rendering. QUAD-buffered stereo uses the left and right buffers of a
-  stereo window (active stereo). Anaglyphic stereo uses
-  \textsf{glColorMask} to mask colors for individual eye passes, used in
-  conjunction with colored glasses. AUTO selects QUAD if the drawable is capable
-  of active stereo rendering, otherwise ANAGLYPH is selected\\
+  rendering. QUAD-buffered stereo uses the left and right buffers of a stereo
+  window (active stereo). Anaglyphic stereo uses \textsf{glColorMask} to mask
+  colors for individual eye passes, used in conjunction with colored
+  glasses. PASSIVE stereo never selects a stereo buffer of a quad-buffered
+  drawable. AUTO selects PASSIVE if the left or right eye pass is not used, QUAD
+  if the drawable is capable of active stereo rendering, and ANAGLYPH in all
+  other cases.
   \hline

### 2010-12-20 Developer/ProgrammingGuide/paper.tex
Fix XCode Makefile rule Remove now-unneeded affinity WAR
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[x] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
   \hline
-  \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_AFFINITY\\
-  \textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\textbar \ EXTERN\\
-  \textbf{Default} & AUTO\\
-  \textbf{Details} & Determines how the window handles the affinity context (WGL
-  only). This hint is a workaround for NVidia drivers greater than 200, which
-  contain a faulty implementation of the WGL\_NV\_gpu\_affinity extension. When
-  set to AUTO, the hint is interpreted like EXTERN for windows and pbuffers with
-  these drivers, in all other cases it is interpreted like ON. When set to ON,
-  an affinity DC is used for all drawable types. When set to EXTERN, a normal DC
-  is used to create the drawable, but an affinity DC is created and can be used
-  by the application to render into a FBO. This hint may be ignored in later
-  versions of Equalizer, once this workaround has become obsolete.\\
-  \hline
-\end{tabularx}\\\vfill
-
-\begin{tabularx}{\textwidth}{|l|X|}
-  \hline
   \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_STEREO\\
@@
 not present, the window is opened on the \textit{n}th monitor, but
-OpenGL commands are sent to all GPU's.
+OpenGL commands are sent to a different or all GPU's.
 
@@
     {
-        hint_affinity       OFF | ON | AUTO | EXTERN
         hint_stereo         OFF | ON | AUTO

### 2010-12-21 Developer/ProgrammingGuide/paper.tex
Update Quickstart Guide Update Programming Guide: - Collage-related changes - Inline code from Eq directly, instead of cpy-pasting it - General review for 1.0a - related code cleanups
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.9 for Equalizer 0.9.1+svn\\[\medskipamount]
+Version 1.7.10 for Equalizer 1.0$\alpha$ \\[\medskipamount]
 %  \today
@@
 \lstset{language=C++}
-\section*{Equalizer 0.9.1 Programming and User Guide}
+\section*{Equalizer 1.0$\alpha$ Programming and User Guide}
 \today
@@
 right], 
-\textsf{eqRaster}, a scalable terrain visualization\footnote{Image
-  courtesy of Visualization and Multimedia Lab, University of Zurich}
+\textsf{Bino}, a stereo-capable, multi-display video player\footnote{\link{http://bino.nongnu.org/}}
 [middle center],
@@
     1.7.9   & Nov 03, 2010 & Add \sref{sErrorHandling}\\
+    1.7.10  & Dec 21, 2010 & Review for Equalizer 1.0$\alpha$\\
     \hline
@@
 
-Equalizer is the standard middleware for the development and deployment
-of parallel OpenGL applications. It enables applications to benefit from
-multiple graphics cards, processors and computers to scale rendering
-performance, visual quality and display size. An Equalizer-based
-application runs unmodified on any visualization system, from a simple
-workstation to large scale graphics clusters, multi-GPU workstations and
-Virtual Reality installations.
+Equalizer is the standard middleware for the development and deployment of
+parallel OpenGL applications. It enables applications to benefit from multiple
+graphics cards, processors and computers to improve the rendering performance,
+visual quality and display size. An Equalizer-based application runs unmodified
+on any visualization system, from a simple workstation to large scale graphics
+clusters, multi-GPU workstations and Virtual Reality installations.
 
@@
 
-Equalizer provides the domain-specific parallel rendering know-how and
-abstracts configuration, threading, synchronization, windowing and event
-handling. It is a `GLUT on steroids', providing parallel and distributed
-execution, scalable rendering features, network data distribution and
-fully customizable event handling.
+Equalizer provides the domain-specific parallel rendering know-how and abstracts
+configuration, threading, synchronization, windowing and event handling. It is a
+`GLUT on steroids', providing parallel and distributed execution, scalable
+rendering features, an advanced network library and fully customizable event
+handling.
 
@@
 
-Compiling Equalizer is as simple as running \textsf{make} on Linux or
-Mac OS X, building the provided XCode project on Mac OS X 10.5, or
-building the Equalizer Visual Studio 2005 solution on Windows. Note that
-on Mac OS X 10.4 (Tiger), some prerequisites have to be installed before
-running \textsf{make}, as explained in \textsf{README.Darwin}.
+Equalizer uses CMake to generate platform-specific build files. Compiling
+Equalizer is as simple as running \textsf{make} on Linux or Mac OS X, running
+\textsf{make xcode} and building the generated XCode project on Mac OS X, or
+creating a Equalizer Visual Studio solution using \textsf{CMake} on Windows.
 
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/processes.pdf}
+  {\caption{\label{fProcesses}Equalizer Processes}}
+\end{wrapfigure}
 The Equalizer architecture is based on a client-server model. The client
@@
 programmer, and provides communication between the different Equalizer
-processes. The underlying network layer uses a peer-to-peer structure,
-allowing render clients to communicate directly with each other, e.g.,
-during parallel compositing.
+processes.
+
+Collage, the underlying network library, uses a peer-to-peer
+structure. Equalizer utilizes this to provide optimal performance scalable
+rendering, e.g., by setting up direct communication between two nodes when
+needed for image compositing or software swap barriers.
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/processes.pdf}
-  {\caption{\label{fProcesses}Equalizer Processes}}
-\end{wrapfigure}
 The application can also be a rendering client, in which case it can
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
-  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
-\end{wrapfigure}
 Stereo compounds, as shown in \fig{fStereoCmp}\footnote{3D model
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
+  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
+\end{wrapfigure}
 Stereo compounds are used by all applications when rendering in stereo,
@@
 
+Applications can use subpixel compounds to accelerate depth-of-field effects and
+software anti-aliasing, with potentially a different number of idle or non-idle
+samples per pixel.
+
+As for the DB compound, the subpixel recomposition has linear increasing 
+IO requirements for the pixel transfer with the difference that only 
+color information has to be transmitted.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-Applications can use subpixel compounds to accelerate depth-of-field effects and
-software anti-aliasing, with potentially a different number of idle or non-idle
-samples per pixel.
-
-As for the DB compound, the subpixel recomposition has linear increasing 
-IO requirements for the pixel transfer with the difference that only 
-color information has to be transmitted.
-
 Subpixel compounds are configured using the \textsf{subpixel} parameter, using
@@
 balance the load of 2D and DB compounds, optimally distribute render
-resources for segment projections systems, adjust the resolution
+resources for multi-display projection systems, adjust the resolution
 to provide a constant framerate or zoom images to allow monitoring of
@@
   description in the server section.
-\item[Application] Start the application using the IP as the listening port,
-  e.g., \textsf{./build/Linux/bin/eqPly -- --eq-listen 192.168.0.1}
+\item[Application] Specify the application IP as the hostname field in the
+  connection description in the appNode section. If the application should not
+  contribute to the rendering, set up an appNode without a pipe section.
 \item[Render Clients] Specify the client IP's as the hostname field in the
@@
   \textsf{--eq-client} and \textsf{--eq-listen}, e.g.,
-  \textsf{./build/Linux/bin/eqPly -- --eq-client --eq-listen
+  \textsf{./build/Linux/bin/eqPly --eq-client --eq-listen
   192.168.0.2:1234}. Pay attention to use the same connection parameters as in
@@
 
-{\footnotesize\begin{lstlisting}
-void Channel::frameDraw( const uint32_t spin )
-{
-    // setup OpenGL State
-    eq::Channel::frameDraw( spin );
-
-    const float lightPos[] = { 0.0f, 0.0f, 1.0f, 0.0f };
-    glLightfv( GL_LIGHT0, GL_POSITION, lightPos );
-
-    const float lightAmbient[] = { 0.2f, 0.2f, 0.2f, 1.0f };
-    glLightfv( GL_LIGHT0, GL_AMBIENT, lightAmbient );
-
-    // rotate scene around the origin
-    glRotatef( static_cast< float >( spin ) * 0.5f, 1.0f, 0.5f, 0.25f );
-
-    // render six axis-aligned colored quads around the origin
-    [...]
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=107, lastline=123]
+  {../../../src/examples/eqHello/hello.cpp}}
 
@@
 future Equalizer versions of the same major revision. Any eventual changes to
-the public is documented in the release notes.
+the public API is documented in the release notes and the file
+\textsf{CHANGES.txt}.
 
@@
 \item[eq::fabric] The eq::fabric namespace is the common data management and
-  transport layer between the client library, the server and the upcoming
-  administrative API. Most Equalizer client classes inherit from base classes in
-  this namespace and have all their data access methods in these base
-  classes. Furthermore, this namespace contains a \textsf{Serializable} class
-  which implements one common use pattern of object-based data distribution
-  using \textsf{eq::net::Object}.
-\item[eq::net] The Equalizer network layer provides basic functionality
-  for network communication, such as \textsf{Connection} and
+  transport layer between the client library, the server and the administrative
+  API. Most Equalizer client classes inherit from base classes in this namespace
+  and have all their data access methods in these base classes. Furthermore,
+  this namespace contains a \textsf{Serializable} class which implements one
+  common use pattern of object-based data distribution using
+  \textsf{co::Object}.
+\item[co] Collage is the network library used by Equalizer. It provides basic
+  functionality for network communication, such as \textsf{Connection} and
   \textsf{ConnectionSet}, as well as higher-level functionality such as
-  \textsf{Node}, \textsf{Session} and \textsf{Object}. Please refer to
+  \textsf{Node}, \textsf{LocalNode} and \textsf{Object}. Please refer to
   \sref{sNetwork} for an introduction into the network layer, and to
   \sref{sNetObject} for distributed objects.
-\item[eq::base] The eq::base namespace provides C++ classes to abstract
+\item[co::base] The co::base namespace provides C++ classes to abstract
   the underlying operating system and implements common helper
-  functionality. One example is the \textsf{eq::base::Clock} class
+  functionality. One example is the \textsf{co::base::Clock} class
   providing a high-resolution timer. Classes in this namespace are fully
@@
 the \textsf{eq} namespace, with the occasional usage of functionality from the
-\textsf{eq::util}, \textsf{eq::fabric}, \textsf{eq::net} and \textsf{eq::base}
+\textsf{eq::util}, \textsf{eq::fabric}, \textsf{co} and \textsf{co::base}
 namespace.
@@
 The functionality of the Equalizer server is implemented in the
-\textsf{eq::server} namespace. The \textsf{eq::admin} namespace enable run-time
-administration of Equalizer servers. These namespace does not yet expose an API
-intended to be used by developers. \fig{fClientUml} provides an overview map of
-the most important classes in the Equalizer namespace, grouped by functionality.
+\textsf{eq::server} namespace. The \textsf{eq::admin} namespace enables run-time
+administration of Equalizer servers. These namespaces do not yet expose a stable
+API. \fig{fClientUml} provides an overview map of the most important classes in
+the Equalizer namespace, grouped by functionality.
 
 \begin{figure}[ht!]\center
-  \includegraphics[width=.9\textwidth]{images/clientUML}
+  \includegraphics[width=\textwidth]{images/clientUML}
   {\caption{\label{fClientUml}Equalizer client UML map}}
@@
 The distributed data classes are helper classes based on
-\textsf{eq::fabric::Serializable} or its base class
-\textsf{eq::net::Object}. They illustrate the typical usage of distributed
-objects for static as well as dynamic, frame-specific data. Furthermore they are
-used for a basic scene graph distribution of the model data.
+\textsf{eq::fabric::Serializable} or its base class \textsf{co::Object}. They
+illustrate the typical usage of distributed objects for static as well as
+dynamic, frame-specific data. Furthermore they are used for a basic scene graph
+distribution of the model data.
 
@@
 
-The main function starts off with parsing the command line into the
-\textsf{LocalInitData} data structure. A part of it, the base class
-\textsf{InitData}, will be distributed to all render client nodes. The
-command line parsing is done by the \textsf{LocalInitData} class, which
-is discussed in \sref{sInitData}:
-
-{\footnotesize\begin{lstlisting}
-int main( int argc, char** argv )
-{
-    // 1. parse arguments
-    eqPly::LocalInitData initData;
-    initData.parseArguments( argc, argv );
-\end{lstlisting}}
-
-The second step is to redirect all Equalizer output to a file if this
-was requested through the command line. This should happen before
-initializing Equalizer, so that output during initialization is
-redirected and threads launched by Equalizer will use the given log
-stream. Any \textsf{std::ostream} can be used as log output:
-
-{\footnotesize\begin{lstlisting}
-    // 2. Redirect log output to file, if requested
-    std::ofstream* logFile = 0;
-    const std::string& logFilename = initData.getLogFilename();
-    if( !logFilename.empty( ))
-    {
-        logFile = new ofstream( logFilename.c_str( ));
-        eq::base::Log::setOutput( *logFile );
-    }
-\end{lstlisting}}
-
-The third step is to initialize the Equalizer library. The
-initialization function of Equalizer also parses the command line, which
-is used to set certain default values based on Equalizer-specific
-options\footnote{Equalizer-specific options always start with -\,-eq-},
-e.g., the default server address. Furthermore, a \textsf{NodeFactory} is
-provided. The \textsf{EQERROR} macro, and its counterparts
-\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
-debugging outputs with various logging levels:
-
-{\footnotesize\begin{lstlisting}
-    // 2. Equalizer initialization
-    NodeFactory nodeFactory;
-    if( !eq::init( argc, argv, &nodeFactory ))
-    {
-        EQERROR << "Equalizer init failed" << endl;
-        return EXIT_FAILURE;
-    }
-\end{lstlisting}}%>>
+The main function starts off with initializing the Equalizer library. The
+command line arguments are passed on to Equalizer. The are used to set certain
+default values based on Equalizer-specific options\footnote{Equalizer-specific
+  options always start with -\,-eq-}, e.g., the default server
+address. Furthermore, a \textsf{NodeFactory} is provided. The \textsf{EQERROR}
+macro, and its counterparts \textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB}
+allow selective debugging outputs with various logging levels:
+
+{\footnotesize\lstinputlisting[firstline=57, lastline=66]
+  {../../../src/examples/eqPly/main.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-class NodeFactory : public eq::NodeFactory
-{
-public:
-    virtual eq::Config*  createConfig( eq::ServerPtr parent )
-        { return new eqPly::Config( parent ); }
-    virtual eq::Node*    createNode( eq::Config* parent )  
-        { return new eqPly::Node( parent ); }
-    virtual eq::Pipe*    createPipe( eq::Node* parent )
-        { return new eqPly::Pipe( parent ); }
-    virtual eq::Window*  createWindow( eq::Pipe* parent )
-        { return new eqPly::Window( parent ); }
-    virtual eq::Channel* createChannel( eq::Window* parent )
-        { return new eqPly::Channel( parent ); }
-    virtual eq::View* createView() { return new eqPly::View(); }
-};
-\end{lstlisting}}
-
-The fourth step is to create an instance of the application and to
+{\footnotesize\lstinputlisting[firstline=40, lastline=55]
+  {../../../src/examples/eqPly/main.cpp}}
+
+The second step is to parse the command line into the
+\textsf{LocalInitData} data structure. A part of it, the base class
+\textsf{InitData}, will be distributed to all render client nodes. The
+command line parsing is done by the \textsf{LocalInitData} class, which
+is discussed in \sref{sInitData}:
+
+{\footnotesize\lstinputlisting[firstline=67, lastline=70]
+  {../../../src/examples/eqPly/main.cpp}}
+
+The third step is to create an instance of the application and to
 initialize it locally. The application is a subclass of
-\textsf{eq::Client}, which in turn is an \textsf{eq::net::Node}. The
-underlying Equalizer network layer, discussed in \sref{sNetwork}, is a
-peer-to-peer network of \textsf{eq::net::Node}s. The client/server
+\textsf{eq::Client}, which in turn is an \textsf{co::LocalNode}. The
+underlying Collage network library, discussed in \sref{sNetwork}, is a
+peer-to-peer network of \textsf{co::LocalNode}s. The client/server
 concept is implement the higher-level \textsf{eq} client namespace.
 
-The local initialization of a node creates at least one local listening
-socket, which allows the \textsf{eq::Client} to communicate over the
-network with other nodes, such as the server and the rendering
-clients. The listening socket(s) can be configured using the
-\textsf{--eq-listen} command line parameter, or by programmatically
-adding connection descriptions to the client before the local
-initialization.
-
-{\footnotesize\begin{lstlisting}
-    // 4. initialization of local client node
-    RefPtr< eqPly::Application > client = new eqPly::Application( initData );
-    if( !client->initLocal( argc, argv ))
-    {
-        EQERROR << "Can't init client" << endl;
-        eq::exit();
-        return EXIT_FAILURE;
-    }
-\end{lstlisting}}%>>
+The local initialization of a node creates at least one local listening socket,
+which allows the \textsf{eq::Client} to communicate over the network with other
+nodes, such as the server and the rendering clients. The listening socket(s) can
+be configured using the \textsf{--eq-listen} command line parameter, by adding
+connections to the \textsf{appNode} in the configuration file, or by
+programmatically adding connection descriptions to the client before the local
+initialization:
+
+{\footnotesize\lstinputlisting[firstline=71, lastline=78]
+  {../../../src/examples/eqPly/main.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-    // 5. run client
-    const int ret = client->run();
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=80, lastline=81]
+  {../../../src/examples/eqPly/main.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-    // 6. cleanup and exit
-    client->exitLocal();
-    client = 0;
-
-    eq::exit();
-
-    if( logFile )
-        logFile->close();
-    delete logFile;
-
-    return ret;
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=83]
+  {../../../src/examples/eqPly/main.cpp}}
 
@@
 Another configuration can be selected using the command-line parameter
-\textsf{--eq-config}. This configuration may only contain an
-\textsf{appNode}, i.e., no distributed configuration is possibly since
-the process-local server does not create a listening socket.
-
-{\footnotesize\begin{lstlisting}
-int Application::run()
-{
-    // 1. connect to server
-    eq::ServerPtr server = new eq::Server;
-    if( !connectServer( server ))
-    {
-        EQERROR << "Can't open server" << endl;
-        return EXIT_FAILURE;
-    }
-\end{lstlisting}}%>>
+\textsf{--eq-config}. If this configuration contains additional nodes, the
+\textsf{server} and \textsf{appNode} sections need a globally reachable
+\textsf{connection} configured for the render clients to connect to.
+
+{\footnotesize\lstinputlisting[firstline=80, lastline=88]
+  {../../../src/examples/eqPly/eqPly.cpp}}
 
@@
 \textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
-making the cast necessary:
-
-{\footnotesize\begin{lstlisting}
-    // 2. choose config
-    eq::ConfigParams configParams;
-    Config* config = static_cast<Config*>(server->chooseConfig( configParams ));
-
-    if( !config )
-    {
-        EQERROR << "No matching config on server" << endl;
-        disconnectServer( server );
-        return EXIT_FAILURE;
-    }
-\end{lstlisting}}%>>
+making the static\_cast necessary:
+
+{\footnotesize\lstinputlisting[firstline=90, lastline=101]
+  {../../../src/examples/eqPly/eqPly.cpp}}
 
@@
 sections. \textsf{Config::init} returns after all nodes, pipes,
-windows and channels are initialized. It returns \textsf{true} only if
-all initialization task methods were successful.
+windows and channels are initialized.
+
+The return value of \textsf{Config::init} depends on the configuration
+robustness attribute. This attribute is by default activated, allowing
+configurations to launch even when some entities failed to initialize. If
+activated, \textsf{Config::init} always returns true. If deactivated, it returns
+\textsf{true} only if all initialization task methods were successful. In any
+case, \textsf{Config::getError} only returns \textsf{ERROR\_NONE} if all
+entities have initialized successfully.
 
@@
 
-{\footnotesize\begin{lstlisting}
-    // 3. init config
-    eq::base::Clock clock;
-
-    config->setInitData( _initData );
-    if( !config->init( ))
-    {
-        server->releaseConfig( config );
-        disconnectServer( server );
-        return EXIT_FAILURE;
-    }
-
-    EQLOG( eq::LOG_STATS ) << "Config init took " << clock.getTimef() << " ms"
-                           << endl;
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=103, lastline=118]
+  {../../../src/examples/eqPly/eqPly.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-    // 4. run main loop
-    uint32_t maxFrames = _initData.getMaxFrames();
-
-    clock.reset();
-    while( config->isRunning( ) && maxFrames-- )
-    {
-        config->startFrame();
-        // config->renderData(...);
-        config->finishFrame();
-
-        while( !config->needsRedraw( )) // wait for an event requiring redraw
-        {
-            if( hasCommands( )) // execute non-critical pending commands
-            {
-                processCommand();
-                config->handleEvents(); // non-blocking
-            }
-            else  // no pending commands, block on user event
-            {
-                const eq::ConfigEvent* event = config->nextEvent();
-                if( !config->handleEvent( event ))
-                    EQVERB << "Unhandled " << event << endl;
-            }
-        }
-        config->handleEvents(); // process all pending events
-    }
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=120, lastline=147]
+  {../../../src/examples/eqPly/eqPly.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-    const uint32_t frame = config->finishAllFrames();
-    const float    time  = clock.getTimef();
-    EQLOG( eq::LOG_STATS ) << "Rendering took " << time << " ms (" << frame
-                           << " frames @ " << ( frame / time * 1000.f)
-                           << " FPS)" << endl;
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=148, lastline=152]
+  {../../../src/examples/eqPly/eqPly.cpp}}
 
 The remainder of the application code cleans up in the reverse order of
-initialization. The config is exited, released and the connection to
-the server is closed:
-
-{\footnotesize\begin{lstlisting}
-    // 5. exit config
-    clock.reset();
-    config->exit();
-    EQLOG( eq::LOG_STATS ) << "Exit took " << clock.getTimef() << " ms" <<endl;
-
-    // 6. cleanup and exit
-    server->releaseConfig( config );
-    if( !disconnectServer( server ))
-        EQERROR << "Client::disconnectServer failed" << endl;
-    server = 0;
-    return EXIT_SUCCESS;
-}
-\end{lstlisting}}%>>
+initialization. The config is exited, released and the connection to the server
+is closed:
+
+{\footnotesize\lstinputlisting[firstline=154, lastline=169]
+  {../../../src/examples/eqPly/eqPly.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool Application::clientLoop()
-{
-    if( !_initData.isResident( )) // execute only one config run
-        return eq::Client::clientLoop();
-
-    // else execute client loops 'forever'
-    while( true ) // TODO: implement SIGHUP handler to exit?
-    {
-        if( !eq::Client::clientLoop( ))
-            return false;
-        EQINFO << "One configuration run successfully executed" << endl;
-    }
-    return true;
-}
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=171, lastline=179]
+  {../../../src/examples/eqPly/eqPly.cpp}}
 
@@
 data distribution in a cluster environment. Distributed objects are created by
-subclassing from \textsf{eq::fabric::Serializable} or
-\textsf{eq::net::Object}. The application programmer implements serialization
-and deserialization of the distributed data. \sref{sNetObject2} covers
-distributed objects in detail.
+subclassing from \textsf{eq::fabric::Serializable} or \textsf{co::Object}. The
+application programmer implements serialization and deserialization of the
+distributed data. \sref{sNetObject2} covers distributed objects in detail.
 
@@
 
-The data streams implement efficient buffering and data transport
-between nodes. They perform no type checking or transformation on the
-data. It is the application's responsibility to exactly match the order
-and types of variables during serialization and de-serialization.
+The data streams implement efficient buffering and compression, and
+automatically select the best connection, i.e., multicast where available, for
+data transport. They perform no type checking or transformation on the data. It
+is the application's responsibility to exactly match the order and types of
+variables during serialization and de-serialization.
 
@@
 
-{\footnotesize\begin{lstlisting}
-void InitData::getInstanceData( eq::net::DataOStream& os )
-{
-    os << _frameDataID << _windowSystem << _renderMode << _useGLSL << _invFaces
-       << _logo;
-}
-
-void InitData::applyInstanceData( eq::net::DataIStream& is )
-{
-    is >> _frameDataID >> _windowSystem >> _renderMode >> _useGLSL >> _invFaces
-       >> _logo;
-
-    EQASSERT( _frameDataID != EQ_ID_INVALID );
-    EQINFO << "New InitData instance" << endl;
-}
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=61, lastline=74]
+  {../../../src/examples/eqPly/initData.cpp}}
 
@@
 \begin{itemize}
-\item Only the master instance of the object generates new versions, that is,
-  \textsf{commit} should only be called only on the master instance. It is
-  possible to commit on slave instances, but special care has to be taken to
-  handle possible conflicts. \sref{sSlaveCommit} covers slave object commits in
-  detail.
+\item The master instance of the object generates new versions for all
+  slaves. These versions are continous, starting at
+  \textsf{co::VERSION\_FIRST}. It is possible to commit on slave instances, but
+  special care has to be taken to handle possible conflicts. \sref{sSlaveCommit}
+  covers slave object commits in detail.
 \item Slave instance versions can only be advanced, that is, \textsf{sync(
@@
   \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
-  {\caption{\label{fUMLObject}fabric::Serializable and net::Object}}
+  {\caption{\label{fUMLObject}fabric::Serializable and co::Object}}
 \end{wrapfigure}
@@
 inherit objects. Other ways of implementing change tracking, e.g., using
-incarnation counters, can still be implemented by using \textsf{eq::net::Object}
+incarnation counters, can still be implemented by using \textsf{co::Object}
 which leaves all flexibility to the developer. \fig{fUMLObject} shows the
-relationship between \textsf{eq::fabric::Serializable} and
-\textsf{eq::net::Object}.
+relationship between \textsf{eq::fabric::Serializable} and \textsf{co::Object}.
 
 The \textsf{FrameData} is sub-classed from \textsf{Serializable}, and
-consequently tracks its changes by setting the appropriate dirty bit
-whenever it is changed. The serialization methods are called by the
-\textsf{eq::fabric::Serializable} with the dirty bit mask needed to serialize all
-data, or with the dirty bit mask of the changes since the last
-\textsf{commit}. The \textsf{FrameData} only defines its own dirty bits
-and serialization code:
-
-{\footnotesize\begin{lstlisting}
-/** The changed parts of the data since the last pack(). */
-enum DirtyBits
-{
-    DIRTY_CAMERA  = eq::fabric::Serializable::DIRTY_CUSTOM << 0,
-    DIRTY_FLAGS   = eq::fabric::Serializable::DIRTY_CUSTOM << 1,
-    DIRTY_VIEW    = eq::fabric::Serializable::DIRTY_CUSTOM << 2,
-    DIRTY_MESSAGE = eq::fabric::Serializable::DIRTY_CUSTOM << 3,
-};
-
-void FrameData::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
-{
-    eq::fabric::Serializable::serialize( os, dirtyBits );
-    if( dirtyBits & DIRTY_CAMERA )
-        os << _translation << _rotation << _modelRotation;
-    if( dirtyBits & DIRTY_FLAGS )
-        os << _modelID << _renderMode << _colorMode << _ortho << _statistics
-           << _help << _wireframe << _pilotMode << _idleMode;
-    if( dirtyBits & DIRTY_VIEW )
-        os << _currentViewID;
-    if( dirtyBits & DIRTY_MESSAGE )
-        os << _message;
-}
-
-void FrameData::deserialize( eq::net::DataIStream& is,
-                             const uint64_t dirtyBits )
-{
-    eq::fabric::Serializable::deserialize( is, dirtyBits );
-    if( dirtyBits & DIRTY_CAMERA )
-        is >> _translation >> _rotation >> _modelRotation;
-    if( dirtyBits & DIRTY_FLAGS )
-        is >> _modelID >> _renderMode >> _colorMode >> _ortho >> _statistics
-           >> _help >> _wireframe >> _pilotMode >> _idleMode;
-    if( dirtyBits & DIRTY_VIEW )
-        is >> _currentViewID;
-    if( dirtyBits & DIRTY_MESSAGE )
-        is >> _message;
-}
-\end{lstlisting}}%>>
+consequently tracks its changes by setting the appropriate dirty bit whenever it
+is changed. The serialization methods are called by the
+\textsf{eq::fabric::Serializable} with the dirty bit mask needed to serialize
+all data, or with the dirty bit mask of the changes since the last
+\textsf{commit}. The \textsf{FrameData} only defines its own dirty bits and
+serialization code:
+
+{\footnotesize\lstinputlisting[firstline=122, lastline=129]
+  {../../../src/examples/eqPly/frameData.h}}
+{\footnotesize\lstinputlisting[firstline=51, lastline=77]
+  {../../../src/examples/eqPly/frameData.cpp}}
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
-  {\caption{\label{fModelDist}Scene Data in eqPly}}
-\end{wrapfigure}
 Some applications might rely on a shared filesystem to access the data,
@@
 
-{\footnotesize\begin{lstlisting}
-void VertexBufferDist::getInstanceData( eq::net::DataOStream& os )
-{
-    EQASSERT( _node );
-    os << _isRoot;
-
-    if( _left && _right )
-    {
-        os << _left->getID() << _right->getID();
-
-        if( _isRoot )
-        {
-            EQASSERT( _root );
-            const mesh::VertexBufferData& data = _root->_data;
-
-            os << data.vertices << data.colors << data.normals << data.indices;
-        }
-    }
-    else
-    {
-        os << EQ_ID_INVALID << EQ_ID_INVALID;
-
-        EQASSERT( dynamic_cast< const mesh::VertexBufferLeaf* >( _node ));
-        const mesh::VertexBufferLeaf* leaf = 
-            static_cast< const mesh::VertexBufferLeaf* >( _node );
-
-        os << leaf->_vertexStart << leaf->_vertexLength << leaf->_indexStart
-           << leaf->_indexLength;
-    }
-
-    os << _node->_boundingSphere << _node->_range;
-}
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=143, lastline=174]
+  {../../../src/examples/eqPly/vertexBufferDist.cpp}}
 
@@
 \end{figure}
-The tasks are executed by the node and pipe threads in parallel. The
-parent's initialization methods are always executed before any child
-initialization method. This parallelization is necessary to allow a
-speedy startup of the configuration on large-scale graphics clusters. On
-the other hand, it means that initialization functions are called even
-if the parent's initialization has failed. \fig{fConfigInit} shows a
-sequence diagram of the config initialization.
-
-The \textsf{eqPly::Config} class holds the master versions of the
-initialization and frame data. Both objects are registered with the
-\textsf{eq::Config}, which is the \textsf{eq::net::Session} used for
-rendering. Equalizer takes care of the session setup and exit in
-\textsf{Client::choose\-Config} and \textsf{Client::releaseConfig},
-respectively.
-
-The frame data is registered before the initialization data, since its
-identifier is transmitted using the \textsf{InitData}. The identifier of
-the initialization data is transmitted to the render client nodes using
-the \textsf{initID} parameter of \textsf{eq::Config::init}.
+The tasks are executed by the node and pipe threads in parallel. The parent's
+initialization methods are always executed before any child initialization
+method. This parallelization is necessary to allow a speedy startup of the
+configuration on large-scale graphics clusters. On the other hand, it means that
+initialization functions are called even if the parent's initialization has
+failed. \fig{fConfigInit} shows a sequence diagram of the config initialization.
+
+The \textsf{eqPly::Config} class holds the master versions of the initialization
+and frame data. Both objects are registered with the \textsf{eq::Config}. The
+configuration forwards the registration to the local client node, and retains
+buffered, deregistered object data for $latency$ frames.
+
+The identifier of the initialization data is transmitted to the render client
+nodes using the \textsf{initID} parameter of \textsf{eq::Config::init}. The
+identifier of the frame data is transmitted using the \textsf{InitData}.
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool Config::init()
-{
-    // init distributed objects
-    _frameData.data.color      = _initData.useColor();
-    _frameData.data.renderMode = _initData.getRenderMode();
-    registerObject( &_frameData );
-
-    _initData.setFrameDataID( _frameData.getID( ));
-    registerObject( &_initData );
-
-    // init config
-    _running = eq::Config::init( _initData.getID( ));
-    if( !_running )
-        return false;
-\end{lstlisting}}
-
-If the configuration was initialized correctly, the configuration tries
-to set up a tracking device for head tracking. Equalizer does not
-provide extensive support for tracking devices, as this is an orthogonal
-problem to parallel rendering. Tracking device support has already been
-solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
-which can and have been easily integrated with Equalizer. The example
-code in \textsf{eqPly} provides a reference implementation for the
-integration of such a tracking library. \sref{sTracking} provides more
-background on head tracking.
-
-{\footnotesize\begin{lstlisting}
-    // init tracker
-    if( !_initData.getTrackerPort().empty( ))
-    {
-        if( !_tracker.init( _initData.getTrackerPort() ))
-            EQWARN << "Failed to initialise tracker" << endl;
-        else
-        {
-            // Set up position of tracking system wrt world space
-            // Note: this depends on the physical installation.
-            vmml::Matrix4f m( vmml::Matrix4f::IDENTITY );
-            m.scale( 1.f, 1.f, -1.f );
-            _tracker.setWorldToEmitter( m );
-
-            m = vmml::Matrix4f::IDENTITY;
-            m.rotateZ( -M_PI_2 );
-            _tracker.setSensorToObject( m );
-            EQINFO << "Tracker initialized" << endl;
-        }
-    }
-
-    return true;
-}
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=74, lastline=95]
+  {../../../src/examples/eqPly/config.cpp}}
+
+After a successful initialization, the models are loaded and registered for data
+distribution. When idle, Equalizer will pre-distribute object data during
+registration to accelerate the mapping of slave instances. Registering the
+models after \textsf{Config::init} ensures that the render clients are running
+and can cache the data:
+
+{\footnotesize\lstinputlisting[firstline=97, lastline=99]
+  {../../../src/examples/eqPly/config.cpp}}
+
+The configuration tries to set up a tracking device for head tracking. Equalizer
+does not provide extensive support for tracking devices, as this is an
+orthogonal problem to parallel rendering. Tracking device support has already
+been solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
+which are easily integrated with Equalizer. The example code in \textsf{eqPly}
+provides a reference implementation for the integration of such a tracking
+library. \sref{sTracking} provides more background on head tracking.
+
+{\footnotesize\lstinputlisting[firstline=100, lastline=118]
+  {../../../src/examples/eqPly/config.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool Config::exit()
-{
-    _running = false;
-    const bool ret = eq::Config::exit();
-
-    _initData.setFrameDataID( EQ_ID_INVALID );
-    deregisterObject( &_initData );
-    deregisterObject( &_frameData );
-
-    return ret;
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=130, lastline=139]
+  {../../../src/examples/eqPly/config.cpp}}
 
@@
 
-The rendering frames are issued by the application. The
-\textsf{eqPly::Config} only overrides \textsf{startFrame} to
-update its data before forwarding the start frame request to the
-\textsf{eq::Config}.
-
-If a tracker is used, the current head position and orientation is
-retrieved and passed to Equalizer, which uses the head matrix together
-with the wall or projection description to compute the view
-frusta, as explained in \sref{sTracking}.
-
-The camera position is updated and the frame data is committed, which
-generates a new version of this object. This version is passed to the
-rendering callbacks and will be used by the rendering threads to
-synchronize the frame data to the state belonging to the current frame:
-
-{\footnotesize\begin{lstlisting}
-uint32_t Config::startFrame()
-{
-    // update head position
-    if( _tracker.isRunning() )
-    {
-        _tracker.update();
-        const eq::Matrix4f& headMatrix = _tracker.getMatrix();
-        _setHeadMatrix( headMatrix );
-    }
-
-    // update database
-    if( _animation.isValid( ))
-    {
-        const eq::Vector3f&  modelRotation = _animation.getModelRotation();
-        const CameraAnimation::Step& curStep = _animation.getNextStep();
-
-        _frameData.setModelRotation( modelRotation        );
-        _frameData.setRotation(     curStep.rotation      );
-        _frameData.setTranslation(  curStep.translation   );
-    }
-    else
-    {
-        if( _frameData.usePilotMode())
-            _frameData.spinCamera( -0.001f * _spinX, -0.001f * _spinY );
-        else
-            _frameData.spinModel( -0.001f * _spinX, -0.001f * _spinY );
-
-        _frameData.moveCamera( 0.0f, 0.0f, 0.001f*_advance );
-    }
-
-    [...]
-
-    const uint32_t version = _frameData.commit();
-
-    _redraw = false;
-    return eq::Config::startFrame( version );
-}
-\end{lstlisting}}
+The rendering frames are issued by the application. The \textsf{eqPly::Config}
+overrides \textsf{startFrame} to update its data, commit a new version of the
+frame data object, and then requests the rendering of a new frame using the
+current frame data version. This version is passed to the rendering callbacks
+and will be used by the rendering threads to synchronize the frame data to the
+state belonging to the current frame:
+
+{\footnotesize\lstinputlisting[firstline=314, lastline=321]
+  {../../../src/examples/eqPly/config.cpp}}
+
+Updating the shared data is decomposed in different steps. If a tracker is used,
+the current head position and orientation is retrieved and passed to Equalizer,
+which uses the head matrix together with the wall or projection description to
+compute the view frusta, as explained in \sref{sTracking}:
+
+{\footnotesize\lstinputlisting[firstline=323, lastline=331]
+  {../../../src/examples/eqPly/config.cpp}}
+
+The camera position is recomputed based on the current navigation mode, and the
+idle state for rendering is set. When idle, \textsf{eqPly} performs
+anti-aliasing to gradually reduce aliasing effects in the rendering. The idle
+state is tracked by the application and used by the rendering callbacks:
+
+{\footnotesize\lstinputlisting[firstline=333, lastline=363]
+  {../../../src/examples/eqPly/config.cpp}}
 
@@
 updates based on mouse events. The camera position has to be handled
-correctly regarding latency, and is therefore saved in the frame data:
-
-{\footnotesize\begin{lstlisting}
-bool Config::handleEvent( const eq::ConfigEvent* event )
-{
-    switch( event->type )
-    {
-        [...]
-        case eq::ConfigEvent::POINTER_MOTION:
-            if( event->pointerMotion.buttons == eq::PTR_BUTTON_NONE )
-                return true;
-
-            if( event->pointerMotion.buttons == eq::PTR_BUTTON1 )
-            {
-                _spinX = 0;
-                _spinY = 0;
-
-                _frameData.data.rotation.preRotateX( 
-                    -0.005f * event->pointerMotion.dx );
-                _frameData.data.rotation.preRotateY(
-                    -0.005f * event->pointerMotion.dy );
-                _redraw = true;
-            }
-            else if( event->pointerMotion.buttons == eq::PTR_BUTTON2 ||
-                     event->pointerMotion.buttons == ( eq::PTR_BUTTON1 |
-                                                       eq::PTR_BUTTON3 ))
-            {
-                _frameData.data.translation.z +=
-                    .005f * event->pointerMotion.dy;
-                _redraw = true;
-            }
-            else if( event->pointerMotion.buttons == eq::PTR_BUTTON3 )
-            {
-                _frameData.data.translation.x += 
-                    .0005f * event->pointerMotion.dx;
-                _frameData.data.translation.y -= 
-                    .0005f * event->pointerMotion.dy;
-                _redraw = true;
-            }
-            return true;
-
-        default:
-            break;
-    }
-    return eq::Config::handleEvent( event );
-}
-\end{lstlisting}}
+correctly regarding latency, and is therefore saved in the frame data.
+
+The event handling code is shortened here. A detailed description on how to
+customize event handling can be found in \sref{sEventHandling}:
+
+{\footnotesize\lstinputlisting[firstline=391, lastline=394]
+  {../../../src/examples/eqPly/config.cpp}}
+{\footnotesize\lstinputlisting[firstline=454, lastline=485]
+  {../../../src/examples/eqPly/config.cpp}}
+{\footnotesize\lstinputlisting[firstline=532, lastline=538]
+  {../../../src/examples/eqPly/config.cpp}}
 
@@
 
-Multiple models can be loaded in \textsf{eqPly}. A configuration has a
-default model, stored in \textsf{InitData}, and one model per view,
-stored and distributed using \textsf{View}. The
-loaded models are evenly distributed over the available views of the
-configuration, as shown in \fig{fLayout}.
+Multiple models can be loaded in \textsf{eqPly}. A configuration has a default
+model, stored in \textsf{InitData}, and one model per view, stored and
+distributed using the \textsf{View}. The loaded models are evenly distributed
+over the available views of the configuration, as shown in \fig{fLayout}.
 
@@
 
-The config maintains the mapped models, and lazily maps new models,
-which are registered by the application process:
-
-{\footnotesize\begin{lstlisting}
-const Model* Config::getModel( const uint32_t modelID )
-{
-    if( modelID == EQ_ID_INVALID ) // no model loaded by application
-        return 0;
-
-    // Accessed concurrently from pipe threads
-    eq::base::ScopedMutex< eq::base::SpinLock > _mutex( _modelLock );
-
-    const size_t nModels = _models.size();
-    EQASSERT( _modelDist.size() == nModels );
-
-    for( size_t i = 0; i < nModels; ++i )
-    {
-        const ModelDist* dist = _modelDist[ i ];
-        if( dist->getID() == modelID )
-            return _models[ i ];
-    }
-
-    _modelDist.push_back( new ModelDist );
-    Model* model = _modelDist.back()->mapModel( this, modelID );
-    EQASSERT( model );
-    _models.push_back( model );
-
-    return model;
-}
-\end{lstlisting}}
+The pre-process config instance maintains the mapped models, and lazily maps new
+models, which are registered by the application process:
+
+{\footnotesize\lstinputlisting[firstline=288, lastline=312]
+  {../../../src/examples/eqPly/config.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool Config::_handleKeyEvent( const eq::KeyEvent& event )
-{
-    switch( event.key )
-    {
-        ...
-        case 'm':
-        case 'M':
-        {
-            if( _modelDist.empty( )) // no models
-                return true;
-
-            // current model
-            const uint32_t viewID = _frameData.getCurrentViewID();
-            View* view = static_cast< View* >( findView( viewID ));
-            const uint32_t currentID = view ? 
-                view->getModelID() : _frameData.getModelID();
-
-            // next model
-            ModelDistVector::const_iterator i;
-            for( i = _modelDist.begin(); i != _modelDist.end(); ++i )
-            {
-                if( (*i)->getID() != currentID )
-                    continue;
-
-                ++i;
-                break;
-            }
-            if( i == _modelDist.end( ))
-                i = _modelDist.begin(); // wrap around
-
-            // set identifier on view or frame data (default model)
-            const uint32_t modelID = (*i)->getID();
-            if( view )
-                view->setModelID( modelID );
-            else
-                _frameData.setModelID( modelID );
-            return true;
-        }
-        ...
-    }
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=833, lastline=870]
+  {../../../src/examples/eqPly/config.cpp}}
 
@@
 methods to be called on the involved channels, and potentially windows,
-pipes and nodes. This operation might fail, which will cause the config
+pipes and nodes. This operation might fail, which may cause the config
 to stop running.
@@
 
-For each active render client, one \textsf{eq::Node} instance is
-created on the appropriate machine. Nodes are only instantiated on their
-render client processes, i.e., each process will only have one
-instance of the \textsf{eq::Node} class. The application process might
-also have a node class, which is handled in exactly the same way as the
-render client nodes. The application and render clients might use a
-different node factory, instantiating a different type of \textsf{eq::Node}.
+For each active render client, one \textsf{eq::Node} instance is created on the
+appropriate machine. Nodes are only instantiated on their render client
+processes, i.e., each process will only have one instance of the
+\textsf{eq::Node} class. The application process might also have a node class,
+which is handled in exactly the same way as the render client nodes. The
+application and render clients might use a different node factory, instantiating
+a different type of \textsf{eq::Node} and children.
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool Node::configInit( const uint32_t initID )
-{
-    if( !eq::Node::configInit( initID ))
-        return false;
-
-    // All render data is static or multi-buffered, we can run asynchronously
-    if( getIAttribute( IATTR_THREAD_MODEL ) == eq::UNDEFINED )
-        setIAttribute( IATTR_THREAD_MODEL, eq::ASYNC );
-
-    Config*    config = static_cast< Config* >( getConfig( ));
-    return config->mapData( initID );
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=35, lastline=48]
+  {../../../src/examples/eqPly/node.cpp}}
 
 The actual mapping of the static data is done by the config. The config
-retrieves the distributed \textsf{InitData}. The object is directly
-unmapped since it is static, and the data has been retrieved:
-
-{\footnotesize\begin{lstlisting}
-bool Config::mapData( const uint32_t initDataID )
-{
-    if( _initData.getID() == EQ_ID_INVALID )
-    {
-        EQCHECK( mapObject( &_initData, initDataID ));
-        unmapObject( &_initData ); // data was retrieved, unmap immediately
-    }
-    else  // appNode, _initData is registered already
-        EQASSERT( _initData.getID() == initDataID );
-}
-\end{lstlisting}}
+retrieves the distributed \textsf{InitData}. The object is directly unmapped
+since it is static, and therefore all data has been retrieved:
+
+{\footnotesize\lstinputlisting[firstline=261, lastline=272]
+  {../../../src/examples/eqPly/config.cpp}}
 
@@
 
-Pipe threads are not explicitly synchronized with each other, that is,
-pipes might be rendering different frames at any given time. Therefore
-frame-specific data has to be allocated for each pipe thread, which in
-the case of \textsf{eqPly} is the frame data. The frame data is a member
-variable of the \textsf{eqPly::Pipe}, and is mapped to the identifier
-provided by the initialization data. The initialization in
+Pipe threads are not explicitly synchronized with each other in \textsf{eqPly}
+due to the use of the async thread model. Pipes might be rendering different
+frames at any given time. Therefore frame-specific data has to be allocated for
+each pipe thread, which applies only to the frame data.
+
+The frame data is a member variable of the \textsf{eqPly::Pipe}, and is mapped
+to the identifier provided by the initialization data. The initialization in
 \textsf{eq::Pipe} does the GPU-specific initialization, which is
-window-system-dependent. On AGL the display ID is determined, and on glX
-the display connection is opened.
-
-{\footnotesize\begin{lstlisting}
-bool Pipe::configInit( const uint32_t initID )
-{
-    const Config*     config    = static_cast<Config*>( getConfig( ));
-    const InitData& initData    = config->getInitData();
-    const uint32_t  frameDataID = initData.getFrameDataID();
-    eq::Config*     config      = getConfig();
-
-    const bool mapped = config->mapObject( &_frameData, frameDataID );
-    EQASSERT( mapped );
-
-    return eq::Pipe::configInit( initID );
-}
-\end{lstlisting}}
+window-system-dependent. On AGL the display ID is determined, and on glX the
+display connection is opened:
+
+{\footnotesize\lstinputlisting[firstline=57, lastline=70]
+  {../../../src/examples/eqPly/pipe.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool Pipe::configExit()
-{
-    eq::Config* config = getConfig();
-    config->unmapObject( &_frameData );
-
-    return eq::Pipe::configExit();
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=72, lastline=78]
+  {../../../src/examples/eqPly/pipe.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-eq::WindowSystem Pipe::selectWindowSystem() const
-{
-    const Config*            config = static_cast<Config*>( getConfig( ));
-    const InitData&        initData = config->getInitData();
-    const eq::WindowSystem ws       = initData.getWindowSystem();
-
-    if( ws == eq::WINDOW_SYSTEM_NONE )
-        return eq::Pipe::selectWindowSystem();
-    if( !supportsWindowSystem( ws ))
-    {
-        EQWARN << "Window system " << ws 
-               << " not supported, using default window system" << endl;
-        return eq::Pipe::selectWindowSystem();
-    }
-
-    return ws;
-}
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=39, lastline=55]
+  {../../../src/examples/eqPly/pipe.cpp}}
 
@@
 
-Parts of the Carbon API used for window and event handling in the AGL
-window system are not thread safe. The application has to call
-\textsf{eq::Global::enterCarbon} before any thread-unsafe Carbon call,
-and \textsf{eq::Global::leaveCarbon} afterwards. These functions should
-be used only during window initialization and exit, not during
-rendering. For various reasons \textsf{enterCarbon} might block up to 50
-milliseconds. Carbon calls in the window event handling routine
-\textsf{Window::processEvent} are thread-safe, since the global carbon
-lock is set in this method. Please contact the Equalizer developer
-mailing list if you need to use Carbon calls on a per-frame basis.
+Parts of the Carbon API used for window and event handling in the AGL window
+system are not thread safe. The application has to call
+\textsf{eq::Global::enterCarbon} before any thread-unsafe Carbon call, and
+\textsf{eq::Global::leaveCarbon} afterwards. These functions should be used only
+during window initialization and exit, not during rendering. For implementation
+reasons \textsf{enterCarbon} might block up to 50 milliseconds. Carbon calls in
+the window event handling routine \textsf{Window::processEvent} are thread-safe,
+since the global carbon lock is set in this method. Please contact the Equalizer
+developer mailing list if you need to use Carbon calls on a per-frame basis.
 
@@
 
-{\footnotesize\begin{lstlisting}
-void Pipe::frameStart( const uint32_t frameID, const uint32_t frameNumber )
-{
-    startFrame( frameNumber );
-    _frameData.sync( frameID );
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=80, lastline=84]
+  {../../../src/examples/eqPly/pipe.cpp}}
 
@@
 
+The particulars of creating a window and OpenGL context depend on the
+window system used. One can either use the implementation provided by
+the operating system, e.g., AGL, WGL or glX, or some higher-level
+toolkit, e.g., Qt.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The particulars of creating a window and OpenGL context depend on the
-window system used. One can either use the implementation provided by
-the operating system, e.g., AGL, WGL or glX, or some higher-level
-toolkit, e.g., Qt.
-
-All window-system specific functionality is implemented by a
-specialization of \textsf{eq::System\-Window}. The \textsf{SystemWindow} class
-defines the minimal interface to be implemented for a new window
-system. Each \textsf{Window} uses one \textsf{SystemWindow} during
-execution. This separation allows an easy implementation and adaption to
-another window system or application.
+All window-system specific functionality is implemented by a specialization of
+\textsf{eq::Sys\-tem\-Window}. The \textsf{SystemWindow} class defines the
+minimal interface to be implemented for a new window system. Each
+\textsf{Window} uses one \textsf{SystemWindow} during execution. This separation
+allows an easy implementation and adaption to another window system or
+application.
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool Window::configInitGL( const uint32_t initID )
-{
-    if( !eq::Window::configInitGL( initID ))
-        return false;
-
-    glLightModeli( GL_LIGHT_MODEL_LOCAL_VIEWER, 1 );
-    glEnable( GL_CULL_FACE ); // OPT - produces sparser images in DB mode
-    glCullFace( GL_BACK );
-
-    EQASSERT( !_state );
-    _state = new VertexBufferState( getObjectManager( ));
-
-    const Config*   config   = static_cast< const Config* >( getConfig( ));
-    const InitData& initData = config->getInitData();
-
-    if( initData.showLogo( ))
-        _loadLogo();
-
-    if( initData.useGLSL() )
-        _loadShaders();
-
-    return true;
-}
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=76, lastline=98]
+  {../../../src/examples/eqPly/window.cpp}}
 
 The state object is used to handle the creation of OpenGL objects in a
-multipipe, multithreaded execution environment. It uses the object
-manager of the \textsf{eq::Window}, which is described in detail in
-\sref{sObjectManager}.
-
-The logo texture is loaded from the file system and bound to a texture
-ID used later by the channel for rendering. A code listing is omitted,
-since the code consists of standard OpenGL calls and is not
-Equalizer-specific.
-
-The window exit happens in the reverse order of the
-initialization. First, \textsf{configExitGL} is called to de-initialize
-OpenGL, followed by \textsf{configExitSystemWindow} which de-initializes the
-drawable and context and deletes the \textsf{SystemWindow} allocated in
+multipipe, multithreaded execution environment. It uses the object manager of
+the \textsf{eq::Window}, which is described in detail in \sref{sObjectManager}.
+
+The logo texture is loaded from the file system and bound to a texture ID used
+later by the channel for rendering. A code listing is omitted, since the code
+consists of standard OpenGL calls and is not Equalizer-specific.
+
+The window exit happens in the reverse order of the initialization. First,
+\textsf{configExitGL} is called to de-initialize OpenGL, followed by
+\textsf{configExitSystemWindow} which de-initializes the drawable and context
+and deletes the \textsf{SystemWindow} allocated in
 \textsf{configInitSystemWindow}.
 
-The window OpenGL exit function of \textsf{eqPly} de-allocates all
-OpenGL objects. The object manager does not delete the object in its
-destructor, since it does not know if an OpenGL context is still
-current.
-
-{\footnotesize\begin{lstlisting}
-bool Window::configExitGL()
-{
-    if( _state )
-        _state->deleteAll();
-
-    delete _state;
-    _state = 0;
-
-    return eq::Window::configExitGL();
-}
-\end{lstlisting}}
+The window OpenGL exit function of \textsf{eqPly} de-allocates all OpenGL
+objects. The object manager does not delete the object in its destructor, since
+it does not know if an OpenGL context is still current.
+
+{\footnotesize\lstinputlisting[firstline=100, lastline=109]
+  {../../../src/examples/eqPly/window.cpp}}
 
@@
 
-The channel is the heart of the application, it contains the actual
-rendering code. The channel is used to perform the various rendering
-operations for the compounds. Each channel has a set of task methods to
-execute the clear, draw, readback and assemble stages needed to render a
-frame.
+The channel is the heart of the application's rendering code, it executes all
+task methods needed to produce the configured rendering. It is used to perform
+the various rendering operations for the compounds. Each channel has a set of
+task methods to execute the clear, draw, readback and assemble stages needed to
+render a frame.
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool Channel::configInit( const uint32_t initID )
-{
-    setNearFar( 0.1f, 10.0f );
-    return true;
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=69, lastline=78]
+  {../../../src/examples/eqPly/channel.cpp}}
 
@@
   channel has a subpixel parameter, the frustum will be jittered before it is
-  applied. Please refer to \ref{sFSAA} for more information.
+  applied. Please refer to \sref{sFSAA} for more information.
 \item[Head Transformation] A transformation matrix positioning the
@@
 
-The \textsf{frameDraw} method in \textsf{eqPly} calls the
-\textsf{frameDraw} method from the parent class, the Equalizer
-channel. The default \textsf{frameDraw} method uses the apply
-convenience functions to setup the OpenGL state for all render context
-information, except for the range which will be used later during
-rendering:
+The rendering first checks a number of preconditions, such as if the rendering
+was interrupted by a reset and if the idle anti-aliasing is finished. Then the
+near and far planes are re-computed, before the rendering context is applied:
+
+{\footnotesize\lstinputlisting[firstline=123, lastline=137]
+  {../../../src/examples/eqPly/channel.cpp}}
+
+The \textsf{frameDraw} method in \textsf{eqPly} calls the \textsf{frameDraw}
+method from the parent class, the Equalizer channel. The default
+\textsf{frameDraw} method uses the apply convenience functions to setup the
+OpenGL state for all render context information, except for the range which will
+be used later during rendering:
 
 {\footnotesize\begin{lstlisting}
-void eq::Channel::frameDraw( const uint32_t frameID )
+void eq::Channel::frameDraw( const uin128_t& frameID )
 {
@@
 }
-
-void Channel::frameDraw( const uint32_t frameID )
-{
-    // Setup OpenGL state
-    eq::Channel::frameDraw( frameID );
 \end{lstlisting}}
@@
 
-{\footnotesize\begin{lstlisting}
-    glLightfv( GL_LIGHT0, GL_POSITION, lightPosition );
-    glLightfv( GL_LIGHT0, GL_AMBIENT,  lightAmbient  );
-    glLightfv( GL_LIGHT0, GL_DIFFUSE,  lightDiffuse  );
-    glLightfv( GL_LIGHT0, GL_SPECULAR, lightSpecular );
-
-    glMaterialfv( GL_FRONT, GL_AMBIENT,   materialAmbient );
-    glMaterialfv( GL_FRONT, GL_DIFFUSE,   materialDiffuse );
-    glMaterialfv( GL_FRONT, GL_SPECULAR,  materialSpecular );
-    glMateriali(  GL_FRONT, GL_SHININESS, materialShininess );
-
-    const FrameData& frameData = getFrameData();
-
-    glPolygonMode( GL_FRONT_AND_BACK, 
-                   frameData.useWireframe() ? GL_LINE : GL_FILL );
-
-    const eq::Vector3f& translation = frameData.getCameraTranslation();
-
-    glMultMatrixf( frameData.getCameraRotation().array );
-    glTranslatef( translation.x(), translation.y(), translation.z() );
-    glMultMatrixf( frameData.getModelRotation().array );
-
-    if( frameData.getColorMode() == COLOR_DEMO )
-    {
-        const eq::Vector3ub color = getUniqueColor();
-        glColor3ub( color.r(), color.g(), color.b() );
-    }
-    else
-        glColor3f( .75f, .75f, .75f );
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=139, lastline=165]
+  {../../../src/examples/eqPly/channel.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-    const Model* model = _getModel();
-    if( model )
-    {
-        _drawModel( model );
-    }
-    else
-    {
-        glNormal3f( 0.f, -1.f, 0.f );
-        glBegin( GL_TRIANGLE_STRIP );
-        glVertex3f(  .25f, 0.f,  .25f );
-        glVertex3f(  .25f, 0.f, -.25f );
-        glVertex3f( -.25f, 0.f,  .25f );
-        glVertex3f( -.25f, 0.f, -.25f );
-        glEnd();
-    }
-
-    [...]
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=167, lastline=178]
+  {../../../src/examples/eqPly/channel.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-void Channel::_drawModel( const Model* model )
-{
-    Window*                  window    = static_cast<Window*>( getWindow() );
-    mesh::VertexBufferState& state     = window->getState();
-
-    const Pipe*              pipe      = static_cast<Pipe*>( getPipe( ));
-    const FrameData&         frameData = pipe->getFrameData();
-
-    const eq::Range&         range     = getRange();
-    vmml::FrustumCullerf     culler;
-
-    if( frameData.getColorMode() == COLOR_MODEL && model->hasColors( ))
-        state.setColors( true );
-    else
-        state.setColors( false );
-
-    _initFrustum( culler, model->getBoundingSphere( ));
-    model->beginRendering( state );
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=600, lastline=621]
+  {../../../src/examples/eqPly/channel.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-    model->beginRendering( state );
-
-    // start with root node
-    vector< const VertexBufferBase* > candidates;
-    candidates.push_back( model );
-
-    while( !candidates.empty() )
-    {
-        const VertexBufferBase* treeNode = candidates.back();
-        candidates.pop_back();
-
-        // completely out of range check
-        if( treeNode->getRange()[0] >= range.end || 
-            treeNode->getRange()[1] < range.start )
-            continue;
-
-        // bounding sphere view frustum culling
-        switch( culler.testSphere( treeNode->getBoundingSphere() ) )
-        {
-            case vmml::VISIBILITY_FULL:
-                // if fully visible and fully in range, render it
-                if( treeNode->getRange()[0] >= range.start && 
-                    treeNode->getRange()[1] < range.end )
-                {
-                    treeNode->render( state );
-                    break;
-                }
-                // partial range, fall through to partial visibility
-            case vmml::VISIBILITY_PARTIAL:
-            {
-                const VertexBufferBase* left = treeNode->getLeft();
-                const VertexBufferBase* right = treeNode->getRight();
-
-                if( !left && !right )
-                {
-                    if( treeNode->getRange()[0] >= range.start )
-                        treeNode->render( state );
-                    // else drop, to be drawn by 'previous' channel
-                }
-                else
-                {
-                    if( left )
-                        candidates.push_back( left );
-                    if( right )
-                        candidates.push_back( right );
-                }
-                break;
-            }
-            case vmml::VISIBILITY_NONE:
-                // do nothing
-                break;
-        }
-    }
-
-    model->endRendering( state );
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=628, lastline=698]
+  {../../../src/examples/eqPly/channel.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-void Channel::frameReadback( const uint32_t frameID )
-{
-    // OPT: Drop alpha channel from all frames during network transport
-    const eq::FrameVector& frames = getOutputFrames();
-    for( eq::FrameVector::const_iterator i = frames.begin(); 
-         i != frames.end(); ++i )
-    {
-        (*i)->setAlphaUsage( false );
-    }
-
-    eq::Channel::frameReadback( frameID );
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=238, lastline=261]
+  {../../../src/examples/eqPly/channel.cpp}}
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\label{fEventUML}UML Diagram for Event Handling Classes}}
-\end{wrapfigure}
 Equalizer provides an \textsf{AGLEvent\-Handler},
@@
 
-GLX uses one X11 Display connection for all windows on a GPU, which is
-managed by the \textsf{eq::Pipe}. For each thread, one
-\textsf{GLXEventHandler} is allocated. The pipe registers the display
-connection with this per-thread event handler. Events for windows
-created using this display connection are automatically received by the
-event handler. No per-window event handling setup needs to be
-done. PBuffers are handled in the same way as window drawables.
-
-The glX event handler iterates over all windows of the registered pipe
-to find the originating \textsf{GLXWindowIF}, which is then used to
-process the event in a fashion similar to AGL. The
-\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
-\textsf{GLXWindowIF::\-process\-Event}.
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
+  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
+\end{wrapfigure}
+For each window, one \textsf{GLXEventHandler} is allocated. FBO's and PBuffers
+are handled in the same way as window drawables. The event dispatch finds the
+corresponding event handler for each received event, which is then used to
+process the event in a fashion similar to AGL. The \textsf{GLXWindowEvent}
+passes the \textsf{XEvent} to \textsf{GLXWindowIF::\-process\-Event}.
 
@@
 
-Later Equalizer versions will introduce \textsf{Pipe::pro\-cess\-Event} and
-\textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
-resolution changes.
-
 \subsubsection{Custom Events in eqPixelBench}
@@
 
-{\footnotesize\begin{lstlisting}
-struct ConfigEvent : public eq::ConfigEvent
-{
-public:
-    enum Type
-    {
-        READBACK = eq::Event::USER,
-        ASSEMBLE
-    };
-
-    ConfigEvent()
-        {
-            size = sizeof( ConfigEvent );
-        }
-
-    // channel name is in user event data
-    char           formatType[64];
-    vmml::Vector2i area;
-    float          msec;
-};
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=36, lastline=57]
+  {../../../src/examples/eqPixelBench/configEvent.h}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-std::ostream& operator << ( std::ostream& os, const ConfigEvent* event );
-...
-bool Config::handleEvent( const eq::ConfigEvent* event )
-{
-    switch( event->type )
-    {
-        case ConfigEvent::READBACK:
-        case ConfigEvent::ASSEMBLE:
-            cout << static_cast< const ConfigEvent* >( event ) << endl;
-            return true;
-
-        default:
-            return eq::Config::handleEvent( event );
-    }
-}
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=58, lastline=71]
+  {../../../src/examples/eqPixelBench/config.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-int main( const int argc, char** argv )
-{
+{\footnotesize\lstinputlisting[firstline=55, lastline=61]
+  {../../../src/examples/eVolve/main.cpp}}
 ...
-    // 2. Equalizer initialization
-    NodeFactory nodeFactory;
-    eVolve::initErrors();
-
-    if( !eq::init( argc, argv, &nodeFactory ))
-    ...
-
-    eq::exit();
-    eVolve::exitErrors();
-    return ret;
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=92, lastline=95]
+  {../../../src/examples/eVolve/main.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-/** Defines errors produced by eVolve. */
-enum Error
-{
-    ERROR_EVOLVE_ARB_SHADER_OBJECTS_MISSING = eq::ERROR_CUSTOM,
-    ERROR_EVOLVE_EXT_BLEND_FUNC_SEPARATE_MISSING,
-    ERROR_EVOLVE_ARB_MULTITEXTURE_MISSING,
-    ERROR_EVOLVE_LOADSHADERS_FAILED,
-    ERROR_EVOLVE_LOADMODEL_FAILED
-};
-
-bool Window::configInitGL( const uint32_t initID )
-{
-    ...
-    if( !GLEW_ARB_shader_objects )
-    {
-        setError( ERROR_EVOLVE_ARB_SHADER_OBJECTS_MISSING );
-        return false;
-    }
-    ...
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=36, lastline=44]
+  {../../../src/examples/eVolve/error.h}}
+
+During initialization, eVolve may use these error codes:
+
+{\footnotesize\lstinputlisting[firstline=45, lastline=57]
+  {../../../src/examples/eVolve/window.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-namespace
-{
-struct ErrorData
-{
-    const uint32_t code;
-    const std::string text;
-};
-
-ErrorData _errors[] = {
-    { ERROR_EVOLVE_ARB_SHADER_OBJECTS_MISSING,
-      "GL_ARB_shader_objects extension missing" },
-    { ERROR_EVOLVE_EXT_BLEND_FUNC_SEPARATE_MISSING,
-      "GL_ARB_shader_objects extension missing" },
-    { ERROR_EVOLVE_ARB_MULTITEXTURE_MISSING,
-      "GL_ARB_shader_objects extension missing" },
-    { ERROR_EVOLVE_LOADSHADERS_FAILED, "Can't load shaders" },
-    { ERROR_EVOLVE_LOADMODEL_FAILED, "Can't load model" },
-
-    { 0, "" } // last!
-};
-}
-
-void initErrors()
-{
-    eq::base::ErrorRegistry& registry = eq::base::Global::getErrorRegistry();
-
-    for( size_t i=0; _errors[i].code != 0; ++i )
-        registry.setString( _errors[i].code, _errors[i].text );
-}
-
-void exitErrors()
-{
-    eq::base::ErrorRegistry& registry = eq::base::Global::getErrorRegistry();
-
-    for( size_t i=0; _errors[i].code != 0; ++i )
-        registry.eraseString( _errors[i].code );
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=37, lastline=73]
+  {../../../src/examples/eVolve/error.cpp}}
 
@@
 The command thread processes internal Equalizer request from other
-nodes, for example during \textsf{eq::net::Object} mapping. In some
+nodes, for example during \textsf{co::Object} mapping. In some
 special cases the command thread executes application code, for example
@@
 
-Equalizer uses GLEW\footnote{\link{http://glew.sourceforge.net}} for
-OpenGL extension handling, particularly the GLEW MX implementation
-providing multi-context support.
+Equalizer uses GLEW\footnote{\link{http://glew.sourceforge.net}} for OpenGL
+extension handling, particularly the GLEW\_MX implementation providing
+multi-context support.
 
 Each \textsf{eq::Window} has a \textsf{GLEWContext}. This context can be
-obtained by using \textsf{glewGetContext} on the window or channel. GLEW
-MX uses this function to dispatch the functions to the correct
-context. Equalizer (re-)initializes the GLEW context whenever a new
-OpenGL context is set on the window.
+obtained by using \textsf{glewGetContext} on the window or channel. GLEW\_MX
+uses this function to dispatch the functions to the correct context. Equalizer
+(re-)initializes the GLEW context whenever a new OpenGL context is set on the
+window.
 
@@
 
+The WGL and GLX pipe manage a WGLEWContext and GLXEWContext, respectively. These
+context are useful if extended wgl or glX functions are used during window
+initialization. The GLEW context structures are initialized using a temporary
+OpenGL context, created using the proper display device of the pipe.
+
 \subsection{Advanced Window Initialization}
@@
 detail the handling of the different window systems. Window systems are
-integrated through the \textsf{OSWindow} interface, which externalizes
+integrated through the \textsf{SystemWindow} interface, which externalizes
 the details of the windowing API from the \textsf{eq::Window}
@@
 window systems, it is often easier to subclass directly from
-\textsf{OSWindow} and implement the abstract methods of this interface
+\textsf{SystemWindow} and implement the abstract methods of this interface
 class.
 
-The method \textsf{Window::configInitOSWindow} is used to instantiate
-and initialize the \textsf{OSWindow} implementation during config
-initialization. After a successful \textsf{OSWindow} initialization,
+The method \textsf{Window::configInitSystemWindow} is used to instantiate
+and initialize the \textsf{SystemWindow} implementation during config
+initialization. After a successful \textsf{SystemWindow} initialization,
 \textsf{Window::configInitGL} is called for the generic OpenGL state
@@
 
-{\footnotesize\begin{lstlisting}
-bool eVolve::Window::configInit( const uint32_t initID )
-{
-    // Enforce alpha channel, since we need one for rendering
-    setIAttribute( IATTR_PLANES_ALPHA, 8 );
-
-    return eq::Window::configInit( initID );
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=37, lastline=43]
+  {../../../src/examples/eVolve/window.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool AGLWindow::configInit( )
-{
-    AGLPixelFormat pixelFormat = chooseAGLPixelFormat();
-    if( !pixelFormat )
-        return false;
-
-    AGLContext context = createAGLContext( pixelFormat );
-    destroyAGLPixelFormat ( pixelFormat );
-    setAGLContext( context );
-
-    if( !context )
-        return false;
-
-    return configInitAGLDrawable();
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=124, lastline=140]
+  {../../../src/libs/client/aglWindow.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool GLXWindow::configInit( )
-{
-    XVisualInfo* visualInfo = chooseXVisualInfo();
-    if( !visualInfo )
-        return false;
-
-    GLXContext context = createGLXContext( visualInfo );
-    setGLXContext( context );
-
-    // Early glxew init to have function pointers for init code
-    glXMakeCurrent( _xDisplay, None, 0 );
-    glxewInit();
-
-    if( !context )
-        return false;
-
-    const bool success = configInitGLXDrawable( visualInfo );
-    XFree( visualInfo );
-
-    if( !success || !_xDrawable )
-    {
-        if( getError() == ERROR_NONE )
-            setError( ERROR_GLXWINDOW_NO_DRAWABLE );
-        return false;
-    }
-
-    makeCurrent();
-    initGLEW();
-
-    if( getIAttribute( Window::IATTR_HINT_DRAWABLE ) == FBO )
-        configInitFBO();
-
-    return success;
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=67, lastline=95]
+  {../../../src/libs/client/glXWindow.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-bool WGLWindow::configInit()
-{
-    if( !initWGLAffinityDC( ))
-    {
-        setError( ERROR_WGL_CREATEAFFINITYDC_FAILED );
-        return false;
-    }
-
-    int pixelFormat = chooseWGLPixelFormat();
-    if( pixelFormat == 0 )
-    {
-        exitWGLAffinityDC();
-        return false;
-    }
-
-    if( !configInitWGLDrawable( pixelFormat ))
-    {
-        exitWGLAffinityDC();
-        return false;
-    }
-
-    if( !_wglDC )
-    {
-        exitWGLAffinityDC();
-        setWGLDC( 0, WGL_DC_NONE );
-        setError( ERROR_WGLWINDOW_NO_DRAWABLE );
-        return false;
-    }
-
-    HGLRC context = createWGLContext();
-    if( !context )
-    {
-        configExit();
-        return false;
-    }
-
-    setWGLContext( context );
-    makeCurrent();
-    initGLEW();
-
-    if( getIAttribute( Window::IATTR_HINT_SWAPSYNC ) != AUTO )
-    {
-        if( WGLEW_EXT_swap_control )
-        {
-            // set vsync on/off
-            const GLint vsync = 
-                ( getIAttribute( Window::IATTR_HINT_SWAPSYNC )==OFF ) ? 0 : 1;
-            wglSwapIntervalEXT( vsync );
-        }
-        else
-            EQWARN << "WGLEW_EXT_swap_control not supported, ignoring window "
-                   << "swapsync hint" << std::endl;
-    }
-
-    if( getIAttribute( Window::IATTR_HINT_DRAWABLE ) == FBO )
-        return configInitFBO();
-
-    return true;
-}
-\end{lstlisting}}%>>
+{\footnotesize\lstinputlisting[firstline=178, lastline=224]
+  {../../../src/libs/client/wglWindow.cpp}}
 
@@
 often amends with view-specific application data. A view might have a
-distributable user data object, which has to inherit from
-\textsf{eq::net::Object} to be committed and synchronized with the associated
-object.
+distributable user data object, which has to inherit from \textsf{co::Object} to
+be committed and synchronized with the associated object.
 
@@
 
-{\footnotesize\begin{lstlisting}
-/** The changed parts of the view. */
-enum DirtyBits
-{
-    DIRTY_MODEL       = eq::fabric::Serializable::DIRTY_CUSTOM << 0,
-    DIRTY_IDLE        = eq::fabric::Serializable::DIRTY_CUSTOM << 1
-};
-
-void View::Proxy::serialize( eq::net::DataOStream& os,
-                             const uint64_t dirtyBits )
-{
-    if( dirtyBits & DIRTY_MODEL )
-        os << _view->_modelID;
-    if( dirtyBits & DIRTY_IDLE )
-        os << _view->_idleSteps;
-}
-
-void View::Proxy::deserialize( eq::net::DataIStream& is,
-                               const uint64_t dirtyBits )
-{
-    if( dirtyBits & DIRTY_MODEL )
-        is >> _view->_modelID;
-    if( dirtyBits & DIRTY_IDLE )
-    {
-        is >> _view->_idleSteps;
-        if( isMaster( ))
-            setDirty( DIRTY_IDLE ); // redistribute slave settings
-    }
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=59, lastline=64, xleftmargin=-7em]
+  {../../../src/examples/eqPly/view.h}}
+{\footnotesize\lstinputlisting[firstline=58, lastline=76]
+  {../../../src/examples/eqPly/view.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-View::View( eq::Layout* parent )
-    : [...]
-{
-    setUserData( &_proxy );
-}
-void View::setModelID( const uint32_t id )
-{
-    _modelID = id;
-    setDirty( DIRTY_MODEL );
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=40, lastline=47]
+  {../../../src/examples/eqPly/view.cpp}}
+{\footnotesize\lstinputlisting[firstline=78, lastline=85]
+  {../../../src/examples/eqPly/view.cpp}}
 
@@
 \end{wrapfigure}
-DB compounds have the advantage of scaling any part of the volume
-rendering pipeline: texture and main memory (smaller bricks for each
-channel), fill rate (less samples per channel) and IO bandwidth for
-time-dependent data (less data per time step and channel). Since the
-amount of texture memory needed for each node decreases linearly, they
-make it possible to render data sets which are not feasible to
-visualize with any other approach.
+Database decomposition has the advantage of scaling any part of the volume
+rendering pipeline: texture and main memory (smaller bricks for each channel),
+fill rate (less samples per channel) and IO bandwidth for time-dependent data
+(less data per time step and channel). Since the amount of texture memory needed
+for each node decreases linearly, they make it possible to render data sets
+which are not feasible to visualize with any other approach.
 
@@
 The eqPly example contains support for software anti-aliasing when being idle,
-showing how to combine it with subpixel compounds.
+showing how to combine it with subpixel compounds. The same basic algorithm
+might be applied to ohter multi-sampling effects, e.g., depth-of-field.
 
@@
 
-{\footnotesize\begin{lstlisting}
-eq::Vector2i Channel::_getJitterStep() const
-{
-    uint32_t channelID  = getSubPixel().index;
-    uint32_t idx = ( channelID * _totalSteps );
-    idx += ( _jitterStep * primeNumberTable[ channelID ] ) % _totalSteps;
-
-    uint32_t sampleSize = _getSampleSize();
-    const int dx = idx % sampleSize;
-    const int dy = idx / sampleSize;
-
-    return eq::Vector2i( dx, dy );
-}
-\end{lstlisting}}
-
-Using this normalized subpixel center point and the size of a pixel on the near
-plane, the jitter for the frustum is computed. A random position within the
-sub-pixel is set as a sample position, which will be used to move the frustum:
-
-{\footnotesize\begin{lstlisting}
-eq::Vector2f Channel::_getJitter() const
-{
-    const eq::PixelViewport& pvp = getPixelViewport();
-    float pvp_w = static_cast<float>( pvp.w );
-    float pvp_h = static_cast<float>( pvp.h );
-    float frustum_w = static_cast<float>(( getFrustum().get_width( )));
-    float frustum_h = static_cast<float>(( getFrustum().get_height( )));
-
-    float pixel_w = frustum_w / pvp_w;
-    float pixel_h = frustum_h / pvp_h;
-
-    uint32_t sampleSize = _getSampleSize();
-    float subpixel_w = pixel_w / static_cast<float>( sampleSize );
-    float subpixel_h = pixel_h / static_cast<float>( sampleSize );
-
-    eq::Vector2i jitterStep = _getJitterStep();
-
-    // Sample value randomly computed within the subpixel
-    eq::base::RNG rng;
-    float value_i = rng.get< float >() * subpixel_w
-                    + static_cast<float>( jitterStep.x( )) * subpixel_w;
-
-    float value_j = rng.get< float >() * subpixel_h
-                    + static_cast<float>( jitterStep.y( )) * subpixel_h;
-
-    return eq::Vector2f( value_i, value_j );
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=556, lastline=579]
+  {../../../src/examples/eqPly/channel.cpp}}
+
+When the rendering is not in idle mode, the jitter is queried from
+Equalizer which returns an optimal subpixel offset for the given subpixel
+decomposition. This is used during normal rendering of subpixel compounds.
+
+During idle rendering of any decomposition, the jitter for the frustum is
+computed using the normalized subpixel center point and the size of a pixel on
+the near plane. A random position within the sub-pixel is set as a sample
+position, which will be used to move the frustum:
+
+{\footnotesize\lstinputlisting[firstline=512, lastline=554]
+  {../../../src/examples/eqPly/channel.cpp}}
 
@@
 If the application is in idle mode, the \textsf{applyFrustum} operation will 
-apply the computed jitter vector to the current frustum.
-
-{\footnotesize\begin{lstlisting}
-void Channel::applyFrustum() const
-{
-    const FrameData& frameData = _getFrameData();
-
-    if( frameData.isIdle() && _jitterStep > 0 )
-    {
-        eq::Vector2f jitter = _getJitter();
-
-        if( frameData.useOrtho( ))
-        {
-            eq::Frustumf ortho = getOrtho();
-            ortho.apply_jitter( jitter );
-
-            glOrtho( ortho.left(), ortho.right(),
-                     ortho.bottom(), ortho.top(),
-                     ortho.near_plane(), ortho.far_plane( ));
-        }
-        else
-        {
-            eq::Frustumf frustum = getFrustum();
-            frustum.apply_jitter( jitter );
-
-            glFrustum( frustum.left(), frustum.right(),
-                       frustum.bottom(), frustum.top(),
-                       frustum.near_plane(), frustum.far_plane( ));
-        }
-        return;
-    }
-    //else
-
-    eq::Channel::applyFrustum();
-}
-\end{lstlisting}}
+apply the computed jitter vector to the current frustum:
+
+{\footnotesize\lstinputlisting[firstline=377, lastline=400]
+  {../../../src/examples/eqPly/channel.cpp}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-void Channel::frameViewFinish( const uint32_t frameID )
-{
-    const FrameData& frameData = _getFrameData();
-
-    const eq::PixelViewport& pvp = getPixelViewport();
-    bool isResized = _accum->resize( pvp.w, pvp.h );
-
-    if( isResized )
-    {
-        _accum->clear();
-        _jitterStep = _totalSteps;
-        _subpixelStep = 0;
-        return;
-    }
-    //else
-
-    if( frameData.isIdle() && _needsTransfer )
-    {
-        setupAssemblyState();
-
-        _accum->accum();
-        _accum->display();
-
-        resetAssemblyState();
-    }
-
-    _drawOverlay();
-    _drawHelp();
-
-    if( frameData.useStatistics())
-        EQ_GL_CALL( drawStatistics( ));
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=305, lastline=337]
+  {../../../src/examples/eqPly/channel.cpp}}
 
 At the end of the frame for each channel, \textsf{frameFinish} will decrement
-the jitter step and create a new event to inform the \textsf{Config} of how many
-frames remain to draw before the idle anti-aliasing is complete.
+the jitter step and create a new event to inform the \textsf{Config} using an
+event of how many frames remain to draw before the idle anti-aliasing is
+complete:
+
+{\footnotesize\lstinputlisting[firstline=346, lastline=371]
+  {../../../src/examples/eqPly/channel.cpp}}
 
@@
 
-\section{\label{sNetwork}The Network Layer}
-
-The Equalizer networking layer provides a peer-to-peer communication
-infrastructure. It is used by Equalizer to communicate between the
-application node, the server and the render clients. It can also be
-used by applications to implement distributed processing independently
-or complementary to the core Equalizer functionality.
-
-The network layer is implemented in the \textsf{eq::net} namespace. It
-provides layered functionality, which provide higher level functionality
-for the programmer. The main primitives in the network layer are:
+\section{\label{sNetwork}The Collage Network Library}
+
+The Collage networking library provides a peer-to-peer communication
+infrastructure. It is used by Equalizer to communicate between the application
+node, the server and the render clients. It can also be used by applications to
+implement distributed processing independently or complementary to the Equalizer
+functionality.
+
+The network library is implemented in the \textsf{co} namespace. It provides
+networking functionality of different abstraction layers, gradually providing
+higher level functionality for the programmer. The main primitives in Collage
+are:
 
@@
   line. Different implementations of a connection exist.
-\item[Node] The abstraction of a process in the cluster. Nodes
-  communicate with each other using connections.
-\item[Session] Provides unique identifiers and object mapping to a set
-  of nodes.
-\item[Object] Provides object-oriented data distribution of C++ objects
-  between nodes within a session.
+\item[Node/LocalNode] The abstraction of a process in the cluster. Nodes
+  communicate with each other using connections. A LocalNode listens on various
+  various connections and processes requests for a given process. A Node is a
+  proxy for a remote LocalNode.
+\item[Object] Provides object-oriented data distribution of C++ objects between
+  nodes within a session. Objects are registered or mapped on a LocalNode.
 \end{description}
 
+\begin{figure}[ht!]\center
+  \includegraphics[width=\textwidth]{images/collageUML}
+  {\caption{\label{fCollageUML}UML class diagram of the major Collage classes}}
+\end{figure}
+\fig{fCollageUML} provides an overview of the major collage classes and their
+relationship, as discussed in the following sections.
+
 \subsection{Connections}
 
-The \textsf{eq::net::Connection} is the basic primitive used for
-communication between processes in Equalizer. It provides a
-stream-oriented communication between two endpoints.
-
-A connection is either closed, connected or listening. A closed
-connection cannot be used for communications. A connected connection
-can be used to read or write data to the communication peer. A listening
-connection can accept connection requests.
-
-An \textsf{eq::net::ConnectionSet} is used to manage multiple
-connections. The typical use case is to have one or more listening
-connections for the local process, and a number of connected connections
-for communicating with other processes.
+The \textsf{co::Connection} is the basic primitive used for communication
+between processes in Equalizer. It provides a stream-oriented communication
+between two endpoints.
+
+A connection is either closed, connected or listening. A closed connection
+cannot be used for communications. A connected connection can be used to read or
+write data to the communication peer. A listening connection can accept
+connection requests.
+
+An \textsf{co::ConnectionSet} is used to manage multiple connections. The
+typical use case is to have one or more listening connections for the local
+process, and a number of connected connections for communicating with other
+processes.
 
 The connection set is used to select one connection which requires some
-action. This can be a connection request on a listening connection,
-pending data on a connected connection or the notification of a
-disconnect.
-
-The connection and connection set can be used by applications to
-implement other network-related functionality, e.g., to communicate
-with a sound server on a different machine.
+action. This can be a connection request on a listening connection, pending data
+on a connected connection or the notification of a disconnect.
+
+The connection and connection set can be used by applications to implement other
+network-related functionality, e.g., to communicate with a sound server on a
+different machine.
 
@@
 
-Nodes, sessions and objects communicate using packets derived from
-\textsf{eq::net::Packet}. The basic packet dispatch and handling is
-implemented in the \textsf{eq::net::Base} class, from which
-\textsf{eq::net::Node}, \textsf{eq::net::Session} and
-\textsf{eq::net::Object} is sub-classed.
-
-The network base class allows the registration of a packets with a
-dispatch queue and an invocation method. Each packet has a command
-identifier, which is used to identify the registered queue and
-method. The method \textsf{dispatchCommand} pushes the packet to the
-registered queue, whereas the method \textsf{invokeCommand} calls the
-registered command function.
+Nodes and objects communicate using packets derived from
+\textsf{co::Packet}. The basic packet dispatch is implemented in the
+\textsf{co::Dispatcher} class, from which \textsf{co::Node} and \textsf{co::Object} is sub-classed.
+
+The dispatcher allows the registration of a packets with a dispatch queue and an
+invocation method. Each packet has a type and command identifier, which is used
+to identify the receiver, registered queue and method. The method
+\textsf{dispatchCommand} pushes the packet to the registered queue and sets the
+registered command function on the packet command. When the
+commands are dequeued by the processing thread the registered method is executed
+using \textsf{Command::invoke}.
 
@@
 
-The \textsf{eq::net::Node} is the abstraction of one process in the
-cluster. Each node has a unique identifier, based on the UUID
-standard\footnote{\link{http://en.wikipedia.org/wiki/Universally\_Unique\_Identifier}}. This
-identifier is used to address nodes, e.g., to query connection
-information to connect to the node. Nodes use connections to communicate
-with each other by sending packets sub-classed from
-\textsf{eq::net::NodePacket}.
-
-Nodes, like connections, are either in the closed, listening or
-connected state. A listening node uses one or more listening connections
-in conjunction with a connection set to wait for data or connection
-requests. For each connection request, a connection and node are created
-to represent this node on the local machine. This node is in the
-connected state, and its connection is added to the connection set of
-the listening node.
+The \textsf{co::Node} is the abstraction of one process in the cluster. Each
+node has a universally unique identifier. This identifier is used to address
+nodes, e.g., to query connection information to connect to the node. Nodes use
+connections to communicate with each other by sending packets sub-classed from
+\textsf{co::NodePacket}.
+
+The \textsf{co::LocalNode} is the specialization of the node for the given
+process. It encapsulates the communication logic for connecting remote nodes, as
+well as object registration and mapping. Local nodes are set up in the listening
+state during initialization.
+
+A remote \textsf{Node} can be either be connected explicitly be the application
+or due to a connection from the given remote node. The explicit connection can
+be done by programmatically creating a node, adding the necessary
+\textsf{ConnectionDescription}s and connecting it to the local node. It may also
+be done by connecting the remote node to the local node by using its
+\textsf{NodeID}. This will cause Collage to query connection information for
+this node from the already-connected nodes, instantiating the node and
+connecting it. Both operations may fail.
+
+\fig{fNetNode} shows the communication between two nodes. The nodes in
+the listening state also have a listening connection, which is not shown
+in the Figure for complexity reasons.
 
@@
 \end{wrapfigure}
-\fig{fNetNode} shows the communication between two nodes. The nodes in
-the listening state also have a listening connection, which is not shown
-in the Figure for complexity reasons.
-
 When the remote node sends a packet, the listening node receives the
@@
 Node packets are directly dispatched using
-\textsf{Base::dispatchCommand}. For session and object packets, the
-appropriate session is found and \textsf{Session::dispatchCommand} will
-dispatch the packet.
+\textsf{Dispatcher::dispatchCommand}. For object packets, the appropriate object
+is found on the local node and \textsf{LocalNode::dispatchCommand} will dispatch
+the packet.
 
@@
 
-If an application wants to extend communication on the node level, it
-can either define its own datatype for packets, or to define custom node
-packets. Packets with a custom datatype are handled by overriding
-\textsf{Node::dispatchCommand}. Custom node packets can use any command
-greater than \textsf{CMD\_NODE\_CUSTOM} as the packet command. By
-registering a \textsf{eq::net::CommandFunc} for these commands, the
-\textsf{eq::net::Base} dispatch and invoke mechanism can be used
-automatically.
-
-\subsection{Sessions}
-
-The \textsf{eq::net::Session} gathers provides higher-level
-functionality to a set of nodes. The master instance of a session is
-registered to a node, which makes this node the session server and
-assigns a node-unique identifier to the session. All other nodes map
-their instance of the session to their local listening node using the
-session's identifier. One given node can participate in any number of
-sessions, both as master and slave. This allows the flexible usage of
-sessions to group different aspects of an application.
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
-  {\caption{\label{fNetSession}Communication between two Sessions}}
-\end{wrapfigure}
-A session provides the facilitate to allocate session-unique
-identifiers. These identifiers can be used by the application to
-uniquely address data within the cluster. They are also used by the
-session to address distributed objects, which are described in the
-following section.
-
-Sessions instances can communicate to each other directly using an
-\textsf{eq::net::Ses\-sion\-Packet}, which is derived from the
-\textsf{NodePacket} and extends it by the session identifier. The
-session ID is used by the node to dispatch the packet to the session.
-
-The session, similar to the node, uses the methods
-\textsf{dispatchCommand} and \textsf{invokeCommand} to dispatch and
-execute packets using either the \textsf{eq::net::Base} class for
-session packets or to dispatch and invoke object packets on the
-appropriate \textsf{eq::net::Object}. Application developers can use
-this mechanism by creating custom session packets with a command greater
-than \textsf{CMD\_SESSION\_CUSTOM} and registering a command function
-for these commands.
+If an application wants to extend communication on the node level, it can either
+define its own datatype for packets, or to define custom node packets. Packets
+with a custom datatype are handled by overriding
+\textsf{LocalNode::dispatchCommand}. Custom node packets can use any command
+greater than \textsf{CMD\_NODE\_CUSTOM} as the packet command. By registering a
+\textsf{co::CommandFunc} for these commands, the \textsf{co::Dispatcher}
+dispatch and invoke mechanism can be used automatically.
 
@@
 Distributed objects subclass from \textsf{eq::fabric::Serializable} or
-\textsf{eq::net::Object}. The application programmer implements serialization
-and deserialization of the distributed data. Objects are dynamically assigned to
-a session, which assigns session-unique identifiers to objects and manages the
-network communication between different instances of the same distributed
-object.
-
-The master version of a distributed object is registered with a
-\textsf{eq::net::Session}, which assigns a session-unique identifier to
-the object. This identifier can be used by other nodes to map their
-instance of the object, thus synchronizing the object's data with the
-remotely registered master version.
+\textsf{co::Object}. The application programmer implements serialization and
+deserialization of the distributed data. Objects are dynamically attached to a
+listening local node, which manages the network communication and packet
+dispatch between different instances of the same distributed object.
+
+Objects are addresses using a universally unique identifier. The identifier is
+automatically created in the object constructor. The master version of a
+distributed object is registered with the \textsf{co::LocalNode}. The identifier
+of the master instance can be used by other nodes to map their instance of the
+object, thus synchronizing the object's data and identifier with the remotely
+registered master version.
 
@@
 
-Distributed objects are addressed using session-unique identifiers,
-because pointers to other objects cannot be distributed directly; they
-have no meaning on remote nodes. The session used to register
-distributed objects is normally the \textsf{eq::Config}, which is
-sub-classed from \textsf{eq::net::Session}.
-
-The entry point for shared data on a render client is the identifier
-passed by the application to \textsf{Config::init}. This identifier
-typically contains the identifier of a static distributed object, and is
-passed by Equalizer to all \textsf{configInit} task methods. Normally
-this initial object is mapped by the render clients in
-\textsf{Node::configInit}. It normally contains identifiers of other
-shared data objects.
+Distributed objects are addressed using universally unique identifiers, because
+pointers to other objects cannot be distributed directly; they have no meaning
+on remote nodes.
+
+The entry point for shared data on a render client is the identifier passed by
+the application to \textsf{Config::init}. This identifier typically contains the
+identifier of a static distributed object, and is passed by Equalizer to all
+\textsf{configInit} task methods. Normally this initial object is mapped by the
+render clients in \textsf{Node::configInit}. It normally contains identifiers of
+other shared data objects.
 
@@
 
-%\begin{wrapfigure}{r}{.382\textwidth}
+There are multiple ways of implementing data distribution for an
+existing C++ class hierarchy, such as a scene graph:
+\begin{description}
+\item[Subclassing] The classes to be distributed inherit from
+  \textsf{co::Object} and implement the serialization methods. This approach is
+  recommended if the source code of existing classes can be modified. It is used
+  for \textsf{eqPly::InitData} and
+  \textsf{eqPly::FrameData}. (\fig{fObjectDist1})
+\item[Proxies] For each object to be distributed, a proxy object is
+  created which manages data distribution for its associated
+  object. This requires the application to track changes on the object
+  separately from the object itself. The model data distribution of
+  eqPly is using this pattern. (\fig{fObjectDist2})
+\item[Multiple Inheritance] A new class inheriting from the class to be
+  distributed and from \textsf{co::Object} implements the data
+  distribution. This requires the application to instantiate a different type of
+  object instead of the existing object, and to create wrapper methods in the
+  superclass calling the original method and setting the appropriate dirty
+  flags. This pattern is not used in eqPly. (\fig{fObjectDist3})
+\end{description}
+
 \begin{figure}[h!t]
@@
 \end{figure}
-There are multiple ways of implementing data distribution for an
-existing C++ class hierarchy, such as a scene graph:
-\begin{description}
-\item[Subclassing] The classes to be distributed inherit from
-  \textsf{eq::net::Object} and implement the serialization methods. This
-  approach is recommended if the source code of existing classes can be
-  modified. It is used for \textsf{eqPly::InitData} and
-  \textsf{eqPly::FrameData}. (\fig{fObjectDist1})
-\item[Proxies] For each object to be distributed, a proxy object is
-  created which manages data distribution for its associated
-  object. This requires the application to track changes on the object
-  separately from the object itself. The model data distribution of
-  eqPly is using this pattern. (\fig{fObjectDist2})
-\item[Multiple Inheritance] A new class inheriting from the class to be
-  distributed and from \textsf{eq::net::Object} implements the data
-  distribution. This requires the application to instantiate a different
-  type of object instead of the existing object, and to create wrapper
-  methods in the superclass calling the original method and setting the
-  appropriate dirty flags. This pattern is not used in
-  eqPly. (\fig{fObjectDist3})
-\end{description}
 
@@
     identical, that is, only instance data is serialized. The serialization
-    always happens before \textsf{Session::registerObject} or
+    always happens before \textsf{LocalNode::registerObject} or
     \textsf{Object::commit} returns. Previous instance data is saved to be able
@@
 
-When a slave object instance commits data, \textsf{pack} is called on the
-object. The data serialized by \textsf{pack} is send to the master instance of
-the object. On the master instance, \textsf{notifyNewVersion} is called from the
-command thread to signal the data availability. No new version is generated on
-the slave instance.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The master instance can \textsf{sync} delta data received from slaves. The data
-is synchronized in the order it has been received from the slaves. If a specific
-order is needed, it is the application's responsibity to serialize the slave
-commits accordingly. When using \textsf{VERSION\_NEXT} as the version parameter
+When a slave object instance commits data, \textsf{pack} is called on the
+object. The data serialized by \textsf{pack} is send to the master instance of
+the object. On the master instance, \textsf{notifyNewVersion} is called from the
+command thread to signal the data availability. The slave instance generates a
+unique, random version. When no data was serialized, \textsf{VERSION\_NONE} is
+returned.
+
+The master instance can \textsf{sync} delta data received from slaves. When
+using the version generated by the slave object commit, only the delta for this
+version is applied. When using \textsf{VERSION\_NEXT} as the version parameter
 to \textsf{sync}, one slave commit is unpacked. This call may block if no data
 is pending. When using \textsf{VERSION\_HEAD} all pending versions are
-unpacked. This call will never block.
+unpacked. The data is synchronized in the order it has been received from the
+slaves. If a specific order is needed, it is the application's responsibity to
+serialize the slave commits accordingly.
 
@@
 \item Set the appropriate dirty bits when unpacking slave delta data on the
-  master object instance.
+  master object instance to redistribute the changes.
 \item Write your application code so that only one object instance is modified
@@
 Objects can send command packets to each other by subclassing
-\textsf{eq::net::Ob\-ject\-Packet}, which is derived from
-\textsf{SessionPacket}. The object packet extends the session packet by
-the object identifier and an object instance identifier.
+\textsf{co::Ob\-ject\-Packet}, which is derived from \textsf{NodePacket}. The
+object packet extends the node packet by the object identifier and an object
+instance identifier.
+
+Multiple instances of the same object (identifier) can be mapped on the same
+node and session simultaneously. Each object instance has a node-unique instance
+identifier. When the instance ID is set to \textsf{EQ\_INSTANCE\_ALL} (the
+default value), a command is dispatch to all object instances registered to the
+session with the correct object ID. Multiple object instance per node are used
+when different threads on the node need to process a different version of the
+same object.
 
@@
 \end{wrapfigure}
-Multiple instances of the same object (identifier) can be mapped on the
-same node and session simultaneously. Each object instance has a
-node-unique instance identifier. When the instance ID is set to
-\textsf{EQ\_ID\_ANY} (the default value), a command is dispatch to all
-object instances registered to the session with the correct object
-ID. Multiple object instance per node are used when different threads on
-the node need to process a different version of the same object.
+When the packet is intended for a specific instance, set the instance identifier
+of the object packet to the corresponding value.
 
@@
 
-Distributed object communication relies on the lower node and session
-communication layers, as shown in \fig{fNetObject}.
+Distributed object communication relies on the lower node communication layers,
+as shown in \fig{fNetObject}.
 
@@
 
-The \textsf{eq::net::Barrier} provides a networked barrier primitive. It
-is an \textsf{eq::net::Object} used by Equalizer for software swap
-barrier support, but it can be used as a generic barrier in application
-code.
+The \textsf{co::Barrier} provides a networked barrier primitive. It is an
+\textsf{co::Object} used by Collage for software swap barrier support, but it
+can be used as a generic barrier in application code.
 
@@
 layer. They influence the design of it, and serve as a sample
-implementation on how to use classes in the \textsf{eq::net} namespace.
+implementation on how to use classes in the \textsf{co} namespace.
 
@@
 
-Equalizer uses the \textsf{eq::net::Object} data distribution mechanism for
+Equalizer uses the \textsf{co::Object} data distribution mechanism for
 implementing data distribution between the server, application and render
 clients. Based on dirty bits, \textsf{eq::fabric::Serializable} implements one
-usage pattern for the \textsf{eq::net::Object}, allowing inheritance with data
+usage pattern for the \textsf{co::Object}, allowing inheritance with data
 distribution. It is easier to use, but imposes one typical way to implement data
@@
 
-For serialization, the default \textsf{eq::net::Object} serialization functions
-are implemented by \textsf{eq::fabric::Serializable}, which (de-)serializes and
+For serialization, the default \textsf{co::Object} serialization functions are
+implemented by \textsf{eq::fabric::Serializable}, which (de-)serializes and
 resets the dirty bits, and calls \textsf{serialize} or \textsf{deserialize} with
@@
 intended to by subclassed directly.  The inheritance chain \textsf{eq::View}
-$\rightarrow$ \textsf{eq::fabric::View} $\rightarrow$ \textsf{eq::Object}
-$\rightarrow$ \textsf{eq::fabric::Se\-rial\-izable} $\rightarrow$
-\textsf{eq::net::Object} serves as sample implementation for this process.
+$\rightarrow$ \textsf{eq::fabric::View} $\rightarrow$
+\textsf{eq::fabric::Object} $\rightarrow$ \textsf{eq::fabric::Se\-rial\-izable}
+$\rightarrow$ \textsf{co::Object} serves as sample implementation for this
+process.
 
 The registration and mapping of \textsf{eq::fabric::Serializable}s is done in
-the same way as for \textsf{eq::net::Object}s, which has been described in
+the same way as for \textsf{co::Object}s, which has been described in
 \sref{sNetObject2}. Equalizer objects derived from
-\textsf{eq::fabric::Serializable}, e.g., \textsf{eq::View}, are typically
-registered, mapped and synchronized by Equalizer.
+\textsf{eq::fabric::Serializable}, e.g., \textsf{eq::View}, are registered,
+mapped and synchronized by Equalizer.
 
@@
   if the drawable is capable of active stereo rendering, and ANAGLYPH in all
-  other cases.
+  other cases.\\
   \hline

### 2010-12-22 Developer/ProgrammingGuide/paper.tex
Fix link in presentation
[ ] May break build
[ ] Breaks existing applications
[x] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
 
-{\footnotesize\lstinputlisting[firstline=154, lastline=169]
+{\footnotesize\lstinputlisting[firstline=154, lastline=165]
   {../../../src/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=171, lastline=179]
+{\footnotesize\lstinputlisting[firstline=167, lastline=175]
   {../../../src/examples/eqPly/eqPly.cpp}}

### 2011-02-11 Developer/ProgrammingGuide/paper.tex
Minor cleanup
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[ ] Documentation

@@
 
-Written by Stefan Eilemann.\\
-Engineering contributions by Maxim Makhinya, Jonas B\"osch, Christian
-Marten, Sarah Amsellem, Patrick Bouchaud, Philippe Robert and Robert Hauck.
+Written by Stefan Eilemann.\\ Contributions by Maxim Makhinya, Jonas B\"osch,
+Christian Marten, Sarah Amsellem, Patrick Bouchaud, Philippe Robert, Robert
+Hauck and Lucas Peetz Dulley.
 
@@
 segment has a higher load than the left-hand segment, causing
-sub-optimal performance. The configuration on the left uses a view
+sub-optimal performance. The configuration on the right uses a view
 equalizer, which assigns two GPU's to the left segment and four GPU's to

### 2011-02-14 Developer/ProgrammingGuide/paper.tex
Improve statistics overlay: - Improve readability of compression ratios - Add downloader and compressor plugin name (uploader missing for now) Add return value to image readback Documentation updates
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[x] New Feature
[x] Cleanup
[ ] Optimization
[ ] Documentation

@@
 
-{\footnotesize\lstinputlisting[firstline=90, lastline=101]
+{\footnotesize\lstinputlisting[firstline=90, lastline=100]
   {../../../src/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=103, lastline=118]
+{\footnotesize\lstinputlisting[firstline=101, lastline=118]
   {../../../src/examples/eqPly/eqPly.cpp}}

### 2011-02-22 Developer/ProgrammingGuide/paper.tex
Programming and User Guide Update.
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.7.10 for Equalizer 1.0$\alpha$ \\[\medskipamount]
+Version 1.8.RC1 for Equalizer 1.0$\beta$ \\[\medskipamount]
 %  \today
@@
 \lstset{language=C++}
-\section*{Equalizer 1.0$\alpha$ Programming and User Guide}
+\section*{Equalizer 1.0$\beta$ Programming and User Guide}
 \today
@@
     1.7.10  & Dec 21, 2010 & Review for Equalizer 1.0$\alpha$\\
+    1.8.RC1  & Feb 22, 2010 & Review for Equalizer 1.0$\beta$\\
     \hline
@@
 For non-planar displays, configure the frustum as described in
-\sref{sCanvas}. For passive stereo installations, configure one segment
-per eye pass, where the segment for the left and right eye have the same
-viewport.
+\sref{sCanvas}. For passive stereo installations, configure one segment per eye
+pass, where the segment for the left and right eye have the same viewport. Set
+the eyes displayed by the segment, i.e., left or right and potentially cyclop.
 
@@
 
-Configure one \textsf{view} for each logical view in each layout. Set
-the \textsf{viewport} to position the view.
+Configure one \textsf{view} for each logical view in each layout. Set the
+\textsf{viewport} to position the view. Set the mode to stereo if appropriate.
 
@@
 latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
-latency of one or two frames is normally not perceived by the user.
+latency of one or two frames is normally not perceived by the user. The
+statistics overlay is explained in detail in \sref{sStatisticsOverlay}.
 
@@
   {../../../src/examples/eqPly/frameData.h}}
-{\footnotesize\lstinputlisting[firstline=51, lastline=77]
+{\footnotesize\lstinputlisting[firstline=54, lastline=80]
   {../../../src/examples/eqPly/frameData.cpp}}
@@
 
-\begin{figure}[ht!]\center
-%\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}
-  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
-  {\caption{\label{fConfigInit}Config Initialization Sequence}}
-  \vspace{-4ex}
-\end{figure}
 The tasks are executed by the node and pipe threads in parallel. The parent's
@@
 failed. \fig{fConfigInit} shows a sequence diagram of the config initialization.
+\begin{figure}[ht!]\center
+%\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
+  {\caption{\label{fConfigInit}Config Initialization Sequence}}
+\end{figure}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=314, lastline=321]
+{\footnotesize\lstinputlisting[firstline=310, lastline=317]
   {../../../src/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=323, lastline=331]
+{\footnotesize\lstinputlisting[firstline=319, lastline=327]
   {../../../src/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=333, lastline=363]
+{\footnotesize\lstinputlisting[firstline=329, lastline=359]
   {../../../src/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=391, lastline=394]
+{\footnotesize\lstinputlisting[firstline=387, lastline=390]
   {../../../src/examples/eqPly/config.cpp}}
-{\footnotesize\lstinputlisting[firstline=454, lastline=485]
+{\footnotesize\lstinputlisting[firstline=450, lastline=481]
   {../../../src/examples/eqPly/config.cpp}}
-{\footnotesize\lstinputlisting[firstline=532, lastline=538]
+{\footnotesize\lstinputlisting[firstline=528, lastline=534]
   {../../../src/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=288, lastline=312]
+{\footnotesize\lstinputlisting[firstline=284, lastline=308]
   {../../../src/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=833, lastline=870]
+{\footnotesize\lstinputlisting[firstline=790, lastline=859]
   {../../../src/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=261, lastline=272]
+{\footnotesize\lstinputlisting[firstline=257, lastline=268]
   {../../../src/examples/eqPly/config.cpp}}
@@
 \begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/DB.png}
+  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/DB.png}
   {\caption{\label{fDBDest}Destination View of a DB Compound using
@@
 
-{\footnotesize\lstinputlisting[firstline=600, lastline=621]
+{\footnotesize\lstinputlisting[firstline=591, lastline=610]
   {../../../src/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=628, lastline=698]
+{\footnotesize\lstinputlisting[firstline=619, lastline=689]
   {../../../src/examples/eqPly/channel.cpp}}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
+  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
+\end{wrapfigure}
 The concrete implementation of these window interfaces is responsible
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
-\end{wrapfigure}
 For each window, one \textsf{GLXEventHandler} is allocated. FBO's and PBuffers
@@
 
+\fig{fFrameSync} illustrates the local frame synchronization.
 Each entity uses \textsf{waitFrame\-Started} to block on the parent's
@@
 thread until the current frame has been drawn.
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
-  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
-\end{wrapfigure}
-\fig{fFrameSync} illustrates the local frame synchronization.
 
@@
 
-GLX initialization is very similar to AGL initialization. Again the
-steps are: choose visual (pixel format), create OpenGL context and then
+GLX initialization is very similar to AGL initialization. Again the steps are:
+choose frame buffer configuration (pixel format), create OpenGL context and then
 create drawable. The only difference is that the data returned by
-\textsf{chooseXVisualInfo} has to be freed using \textsf{XFree}:
-
-{\footnotesize\lstinputlisting[firstline=67, lastline=95]
+\textsf{chooseGLXFBConfig} has to be freed using \textsf{XFree}:
+
+{\footnotesize\lstinputlisting[firstline=67, lastline=101]
   {../../../src/libs/client/glXWindow.cpp}}
@@
 
+\begin{wrapfigure}{r}{.612\textwidth}
+  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
+  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
+\end{wrapfigure}
 The render clients can access a slave instance of the view using
@@
 
-\begin{wrapfigure}{r}{.612\textwidth}
-  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
-  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
-\end{wrapfigure}
 The most common entity to subclass is the \textsf{View}, since the application
@@
 
-\subsubsection{Dynamic Layout Switch}
+\subsubsection{Run-time Layout Switch}
 
@@
 
+\subsubsection{Run-time Stereo Switch}
+
+Similar to switch a layout, the stereo mode of each view can be switched at
+runtime. This causes all destination channels of the view to update the cyclop
+eye or the left and right eye.
+
+The configuration contains stereo information in three places: the view, the
+segment and the compound.
+
+The view defines which stereo mode is active: mono or stereo. This setting can
+be done in the configuration file or programmatically.
+
+The segment has the \textsf{eye} attribute in the configuration file which
+defines which eyes are displayed by this segment. This is typically all eyes for
+active stereo projections and the left or right eye for passive stereo. The
+cyclop eye for passive stereo can either be set on a single segment or both,
+depending if the second projector is active during monoscopic rendering. The
+default setting enables all eyes for a segment.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/stereoSwitch.pdf}
+  {\caption{\label{fStereoSwitch}Using two different decompositions during
+      stereo and mono rendering}}
+\end{wrapfigure}
+The compound has an \textsf{eye} attribute defining which eyes it is capable of
+updating. This allows to specify different compounds for the same channel,
+depending on the stereo mode. One use case is to use one GPU each for updating a
+stereo destination channel in stereo mode, and using a 2D compound with the two
+GPUs in mono mode. \fig{fStereoSwitch} illustrates this example. The coloring in
+mono mode is to illustrate the 2D decomposition. The default setting for the
+compound eye attributes is all eyes.
+
 \subsubsection{View Frustum Update}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
+  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
+\end{wrapfigure}
 A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
-  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
-\end{wrapfigure}
 The frustum of a view overrides the underlying frustum of the
@@
 
-\begin{wrapfigure}{l}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 For orthographic rendering, determining the compositing order of the
@@
 
+\begin{wrapfigure}{l}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 Finding the correct assembly order for perspective frusta is more
@@
 showing how to combine it with subpixel compounds. The same basic algorithm
-might be applied to ohter multi-sampling effects, e.g., depth-of-field.
-
-\subsubsection{Supersampling Algorithm}
+might be applied to other multi-sampling effects, e.g., depth-of-field.
+
+\subsubsection{Software Anti-Aliasing in eqPly}
 
@@
 
-The subpixel area is a function of the current jitter step and the channel's
-\textsf{subpixel} description. To evenly distribute jitter values for all
-channels over the total sub-pixel grid a precomputed prime number table is used
-to pseudo-randomly select the subpixel:
-
-{\footnotesize\lstinputlisting[firstline=556, lastline=579]
+The subpixel area is a function of the current jitter step, the channel's
+\textsf{subpixel} description and the idle state. To evenly distribute jitter
+values for all channels over the total sub-pixel grid a precomputed prime number
+table is used to pseudo-randomly select the subpixel:
+
+{\footnotesize\lstinputlisting[firstline=547, lastline=570]
   {../../../src/examples/eqPly/channel.cpp}}
@@
 decomposition. This is used during normal rendering of subpixel compounds.
-
-During idle rendering of any decomposition, the jitter for the frustum is
-computed using the normalized subpixel center point and the size of a pixel on
-the near plane. A random position within the sub-pixel is set as a sample
-position, which will be used to move the frustum:
-
-{\footnotesize\lstinputlisting[firstline=512, lastline=554]
-  {../../../src/examples/eqPly/channel.cpp}}
-
-\subsubsection{Idle Anti-Aliasing in eqPly}
 
@@
 
-If the application is in idle mode, the \textsf{applyFrustum} operation will 
-apply the computed jitter vector to the current frustum:
-
-{\footnotesize\lstinputlisting[firstline=377, lastline=400]
+During idle rendering of any decomposition, the jitter for the frustum is
+computed using the normalized subpixel center point and the size of a pixel on
+the near plane. A random position within the sub-pixel is set as a sample
+position, which will be used to move the frustum. The \textsf{getJitter} method
+will return the the computed jitter vector for the current frustum. This method
+has a default implementation in \textsf{eq::Channel} for subpixel compounds, but
+is overwritten in \textsf{eqPly} to perform idle anti-aliasing:
+
+{\footnotesize\lstinputlisting[firstline=491, lastline=533]
   {../../../src/examples/eqPly/channel.cpp}}
@@
 Statistics are measured in milliseconds since the configuration was
-initialized. On each node, the global configuration clock is synchronized
-with the server clock. Each statistic event records the originator's
-(channel, window, frame or config) unique identifier.
+initialized. The server synchronizes the per-configuration clock on each node
+automatically. Each statistic event records the originator's (channel, window,
+pipe, node, view or config) unique identifier.
 
@@
   \includegraphics[width=.618\textwidth]{images/statistics}
-  {\caption{\label{fStatistics}Statistics for a two-Node 2D Compound}}
+  {\caption{\label{fStatistics}Statistics for a two node 2D compound}}
 \end{wrapfigure}
-Statistics are enabled per entity using an attribute hint. The hint
-determines how precise the gathered statistics are. When set to
-\textsf{fastest}, the per-frame clock is sampled directly when the event
-occurs. When set to \textsf{nicest}, all OpenGL commands will be
-finished before sampling the event. This incurs a performance penalty,
-but gives more correct results. The default setting is fastest in
-release builds, and nicest in debug builds.
-
-The events are processed by the channel's and window's
-\textsf{processEvent} method. The default implementation sends these
-events to the config using \textsf{Config::sendEvent}, as explained in
-\sref{sEventHandling}. When the default implementation of
-\textsf{Config::handleEvent} receives the statistics event, it sorts the
-event per frame and per originator. When a frame has been finished, the
-events are pushed to the local (app-)node for visualization.
+Statistics are enabled per entity using an attribute hint. The hint determines
+how precise the gathered statistics are. When set to \textsf{fastest}, the
+per-frame clock is sampled directly when the event occurs. When set to
+\textsf{nicest}, all OpenGL commands will be finished before sampling the
+event. This may incur a performance penalty, but gives more correct results. The
+default setting is fastest in release builds, and nicest in debug builds. The
+fastest setting often attributes times to the operation causing an OpenGL
+synchronization instead of the operation submitting the OpenGL commands, e.g.,
+the readback time contains operations from the preceding draw operation.
+
+The events are processed by the channel's and window's \textsf{processEvent}
+method. The default implementation sends these events to the config using
+\textsf{Config::sendEvent}, as explained in \sref{sEventHandling}. When the
+default implementation of \textsf{Config::handleEvent} receives the statistics
+event, it sorts the event per frame and per originator. When a frame has been
+finished, the events are pushed to the local (app-)node for visualization.
 
@@
 
-\subsubsection{Statistics Overlay}
-
-The Equalizer examples render a statistics overlay using the gathered
-statistics events directly before
-\textsf{Window::swapBuffers}. Statistics are toggled on and off using
-the 's' key. The overlay rendering method,
-\textsf{Channel::drawStatistics}, can be called directly by any
-application to render the overlay. It can also be overwritten by a
-custom implementation.
-
-\fig{fStatisticsDetail} shows a detail view of \fig{fStatistics}. The
+\subsubsection{\label{sStatisticsOverlay}Statistics Overlay}
+
+The \textsf{eq::Channel} provides the method \textsf{drawStatistics} which
+renders a statistics overlay using the gathered statistics events. Statistics
+rendering is toggled on and off using the 's' key in the shipped Equalizer
+examples.
+
+\fig{fStatisticsDetail} shows a detailed view of \fig{fStatistics}. The
 statistics shown are for a two-node 2D compound. The destination channel
@@
 
-The X axis is the time. One pixel on the screen corresponds to a given
-time unit, here one millisecond per pixel. The scale is zoomed dynamically
-to powers-of-ten milliseconds to fit the statistics into the available
-viewport. This allows easy and accurate evaluations of bottlenecks or
-misconfigurations in the rendering pipeline. The scale of the statistics
-is printed.
-
-The rightmost pixel is the most current time. On the Y axis are the
-entities: channels, windows, frames and the config. The upper channel is
-the local channel since it executes \textsf{frameAssemble}, and the
-lower channel is the remote channel, executing \textsf{frameReadback}.
-
-The draw time has been artificially inflated by using immediate mode
-rendering to provide more expressive statistics.
-
-To facilitate the understanding, older frames are gradually grayed
-out. The rightmost, current frame is brighter than the
-frame-before-last.
-
-The configuration used has a latency of one frame. Consequently, the
-execution of two frames overlaps. This can be observed in the early
-execution of the remote channel's \textsf{frameDraw}, which starts while
-the local channel is still assembling the previous frame.
-
-The beginning of a frame is marked by a vertical green line, and the end
-of a frame by a vertical gray line. These lines are also attenuated. The
-brightness and color matches the event for \textsf{Config::startFrame}
-and \textsf{Config::finishFrame}, respectively. The event for
-\textsf{startFrame} is typically not visible, since it takes less than
-one millisecond to execute. If no idle processing is done by the
-application, the event for \textsf{finishFrame} occupies a full frame,
-since the config is blocked here waiting for the frame to complete.
-
-In the above example, the local channel finishes drawing the frame
-early, and therefore spends a considerable amount of time waiting for
-the input frame from the remote channel. These wait events, rendered
-red, are a sub-event of the yellow \textsf{frameAssemble} task.
-
-The frame statistics provide information on the image transmission for
-scalable rendering. They are grouped per frame, not per node, that is,
-the transmit and compress part are executed by one node and the receive
-by another, even though the statistics are drawn in the same line. The
-decompression time is part of the receive statistic. The total
-compression ratio is printed with each compression statistic, in this
-case the image has been compressed to 11\% of its original size.
-
-The white \textsf{Window::swapBuffers} task might also take also a long
-time, since the execution of the swap buffer is locked to the vertical
-retrace of the display, and enforced by \textsf{glFinish} before
-sampling the end time of the event. In release builds, this does not
-happen and all OpenGL commands are pipelined with the execution on the
-CPU.
-
-Note that the remote source window does not execute the swapBuffers in
-this configuration, since it is a single-buffered PBuffer.
+The X axis is the time, the right-most pixel is the most current time. One
+pixel on the screen corresponds to a given time unit, here one millisecond per
+pixel. The scale is zoomed dynamically to powers-of-ten milliseconds to fit the
+statistics into the available viewport. This allows easy and accurate
+evaluations of bottlenecks or misconfigurations in the rendering pipeline. The
+scale of the statistics is printed directly above the legend.
+
+On the Y axis are the entities: channels, windows, nodes and the config. The
+top-most channel is the local channel since it executes \textsf{frameAssemble},
+and the lower channel is the remote channel, executing \textsf{frameReadback}.
+
+To facilitate the understanding, older frames are gradually grayed out. The
+rightmost, current frame is brighter than the frame before it.
+
+The configuration uses the default latency of one frame. Consequently, the
+execution of two frames overlaps. This can be observed in the early execution of
+the remote channel's \textsf{frameDraw}, which starts while the local channel is
+still drawing and assembling the previous frame.
+
+The asynchronous execution allows operations to be pipelined, i.e., the
+compression, network transfer and assembly with the actual rendering and
+readback. This increases performance by minimizing idle and wait times. In this
+example, the remote \textsf{channel2} has no idle times, and executes the
+compression and network transfer of its output frame in parallel with rendering
+and readback. Likewise, the application node receives and decompresses the frame
+in parallel to its rendering thread.
+
+In the above example, the local channel finishes drawing the frame early, and
+therefore spends a couple of milliseconds waiting for the input frame from the
+remote channel. These wait events, rendered red, are a sub-event of the yellow
+\textsf{frameAssemble} task. Using a \textsf{load\_equalizer} instead of a
+static 2D decomposition would balance the rendering in this example better and
+minimize or even eliminate this wait time.
+
+The white \textsf{Window::swapBuffers} task might take a longer time, since the
+execution of the swap buffer is locked to the vertical retrace of the
+display. Note that the remote source window does not execute swapBuffers in
+this configuration, since it is a single-buffered FBO.
+
+The beginning of a frame is marked by a vertical green line, and the end of a
+frame by a vertical gray line. These lines are also attenuated. The brightness
+and color matches the event for \textsf{Config::startFrame} and
+\textsf{Config::finishFrame}, respectively. The event for \textsf{startFrame} is
+typically not visible, since it takes less than one millisecond to execute. If
+no idle processing is done by the application, the event for
+\textsf{finishFrame} occupies a full frame, since the config is blocked here
+waiting for the frame \textsf{current - latency} to complete.
+
+A legend directly below the statistics facilitates understanding. It lists the
+per-entity operations with its associated color. Furthermore, some other textual
+information is overlayed with the statistics. The total compression ratio is
+printed with each readback and compression statistic. In this case the image has
+been not been compressed during download. For network transfer it has been
+compressed to 1\% of its original size since it contains a lot of black
+background pixels. For readback the plugin with the name \textsf{0x101,
+  EQ\_COMPRESSOR\_TRANS\-FER\_RGBA\_TO\_BGRA} has been used, and for compression
+the plugin \textsf{0x11, EQ\_COM\-PRESSOR\_RLE\_DIFF\_BGRA} was used. The
+compressor names are listed in the associated plugin header
+\textsf{compressorTypes.h}.
 
@@
 
-\fig{fNetNode} shows the communication between two nodes. The nodes in
-the listening state also have a listening connection, which is not shown
-in the Figure for complexity reasons.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+\fig{fNetNode} shows the communication between two nodes. The nodes in
+the listening state also have a listening connection, which is not shown
+in the Figure for complexity reasons.
+
 When the remote node sends a packet, the listening node receives the
@@
 version is passed with each new frame request.
+
+\begin{figure}[h!t]
+  \subfigure[]{\label{fObjectDist1}
+    \includegraphics[width=.3\textwidth]{images/umlObjectSubclass}}\hfil
+  \subfigure[]{\label{fObjectDist2}
+    \includegraphics[width=.3\textwidth]{images/umlObjectProxy}}\hfil
+  \subfigure[]{\label{fObjectDist3}
+    \includegraphics[width=.3\textwidth]{images/umlObjectMI}}%
+  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
+      Proxies or Multiple Inheritance}}
+\end{figure}
 
@@
 \end{description}
-
-\begin{figure}[h!t]
-  \subfigure[]{\label{fObjectDist1}
-    \includegraphics[width=.3\textwidth]{images/umlObjectSubclass}}\hfil
-  \subfigure[]{\label{fObjectDist2}
-    \includegraphics[width=.3\textwidth]{images/umlObjectProxy}}\hfil
-  \subfigure[]{\label{fObjectDist3}
-    \includegraphics[width=.3\textwidth]{images/umlObjectMI}}%
-  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
-      Proxies or Multiple Inheritance}}
-\end{figure}
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
+  {\caption{\label{fNetObject}Communication between two Objects}}
+\end{wrapfigure}
 Multiple instances of the same object (identifier) can be mapped on the same
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
-  {\caption{\label{fNetObject}Communication between two Objects}}
-\end{wrapfigure}
 When the packet is intended for a specific instance, set the instance identifier
@@
   \hline
+  \textbf{Name} & EQ\_CONFIG\_IATTR\_ROBUSTNESS\\
+  \textbf{Value} & ON \textbar \ OFF\\
+  \textbf{Default} & ON\\
+  \textbf{Details} & Handle initialization failures robustly by deactivating the
+  failed resources but keeping the configuration running.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONFIG\_FATTR\_EYE\_BASE\\
+  \textbf{Value} & float\\
+  \textbf{Default} & 0.05\\
+  \textbf{Details} & The default distance in meters between the left and
+  the right eye, i.e., the eye separation. The eye base influences the
+  frustum during stereo rendering. See \sref{sTracking} for details.\\
+  \textbf{See also} & EQ\_WINDOW\_IATTR\_HINT\_STEREO
+  EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
   \textbf{Name} & EQ\_NODE\_SATTR\_LAUNCH\_COMMAND\\
@@
   and the initialization of the configuration fails.\\
-  \hline
-\end{tabularx}\\\vfill
-
-\begin{tabularx}{\textwidth}{|l|X|}
-  \hline
-  \textbf{Name} & EQ\_CONFIG\_FATTR\_EYE\_BASE\\
-  \textbf{Value} & float\\
-  \textbf{Default} & 0.05\\
-  \textbf{Details} & The default distance in meters between the left and
-  the right eye, i.e., the eye separation. The eye base influences the
-  frustum during stereo rendering. See \sref{sTracking} for details.\\
-  \textbf{See also} & EQ\_WINDOW\_IATTR\_HINT\_STEREO
-  EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
   \hline
@@
   \textbf{Name} & EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
-  \textbf{Value} & AUTO \textbar QUAD \textbar \ ANAGLYPH \textbar \ PASSIVE\\
+  \textbf{Value} & AUTO \textbar \ QUAD \textbar \ ANAGLYPH \textbar \ PASSIVE\\
   \textbf{Default} & AUTO\\
@@
   \hline
-\end{tabularx}\\%\vfill
+\end{tabularx}\\\vfill
 \end{center}
@@
 
-{\footnotesize\begin{lstlisting}
-server
-{
-    connection # 0-n times, listening connections of the server
-    {
-        type       TCPIP | SDP | RSP | PIPE [Win32 only]
-        port       unsigned
-        filename   string
-        hostname   string
-        interface  string
-    }
-
-    config {...} # 1-n times, currently only the first one is used by the server
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=87, lastline=98]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-config
-{
-    latency int    # Nr. of frames nodes may fall behind application, default 1
-    attributes
-    {
-         eye_base float # distance between left and right eye
-    }
-
-    appNode {...}  # 0|1 times, launches render threads within app process
-    node {...}     # 1-n times, a system in the cluster
-    observer {...} # 0-n times, one tracked entity
-    layout {...}   # 0-n times, logical views on canvases
-    canvas {...}   # 0-n times, physical segment layout of a projection surface
-    compound {...} # 0-n times, rendering description
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=98, lastline=108]
+  {../../../website/documents/design/fileFormat.shtml}}
+{\footnotesize\lstinputlisting[firstline=172, lastline=172]
+  {../../../website/documents/design/fileFormat.shtml}}
+{\footnotesize\lstinputlisting[firstline=178, lastline=178]
+  {../../../website/documents/design/fileFormat.shtml}}
+{\footnotesize\lstinputlisting[firstline=205, lastline=205]
+  {../../../website/documents/design/fileFormat.shtml}}
+{\footnotesize\lstinputlisting[firstline=249, lastline=249]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-node
-{
-    name     string
-    connection # 0-n times, possible connections to this node
-    {
-        type          TCPIP | SDP | PIPE
-        TCPIP_port    unsigned
-        PIPE_filename string
-        hostname      string
-        command       string      # render client launch command
-        command_quote 'character' # launch command argument quote char
-        timeout       unsigned    # timeout in milliseconds for launch
-    }
-    attributes
-    {
-        thread_model    ASYNC | DRAW_SYNC | LOCAL_SYNC
-    }
-    pipe {...} # 1-n times
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=107, lastline=125]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-pipe
-{
-    name     string
-    port     unsigned     # X server number or ignored
-    device   unsigned     # graphics adapter number
-    viewport [ viewport ] # default: autodetect
-    attributes
-    {
-        hint_thread            OFF | ON   # default ON
-        hint_cuda_GL_interop   OFF | ON   # default OFF
-    }
-
-    window {...} # 1-n times
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=125, lastline=136]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-window
-{
-    name     string
-    viewport [ viewport ] # wrt pipe, default full screen
-
-    attributes
-    {
-        hint_stereo         OFF | ON | AUTO
-        hint_doublebuffer   OFF | ON | AUTO
-        hint_decoration     OFF | ON
-        hint_fullscreen     OFF | ON
-        hint_swapsync       OFF | ON         # AGL, WGL only
-        hint_drawable       window | pbuffer | FBO
-        hint_statistics     off | fastest [on] | nicest
-        planes_color        unsigned | RGBA16F | RGBA32F
-        planes_alpha        unsigned
-        planes_depth        unsigned
-        planes_stencil      unsigned
-        planes_accum        unsigned
-        planes_accum_alpha  unsigned
-        planes_samples      unsigned
-    }
-
-    channel {...} # 1-n times
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=136, lastline=158]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-channel
-{
-    name     string
-    viewport [ viewport ] #wrt window, default full window
-    drawable [ FBO_COLOR FBO_DEPTH FBO_STENCIL ]
-
-    attributes
-    {
-        hint_statistics     OFF | FASTEST [ON] | NICEST
-    }
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=158, lastline=167]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-observer # 0...n times
-{
-    name    string
-    eyeBase float
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=172, lastline=176]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 
-{\footnotesize\begin{lstlisting}
-layout  # 0...n times
-{
-    name string
-    view {...}  # 1...n times
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=178, lastline=181]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 has no influence on the execution of Equalizer. It can be used to
-reference the view, in which case the name should be unique.
+reference the view, in which case the name should be unique. The viewport
+specifies which 2D area of the parent layout is covered by this view in
+normalized coordinates.
 
@@
 
+A view has a stereo mode, which defines if the corresponding destination channel
+update the cyclop or left and right eye.
+
 A view is a view on the application's model, in the sense used by the
@@
 
-{\footnotesize\begin{lstlisting}
-view  # 1...n times
-{
-    name     string
-    observer observer-ref
-    viewport [ viewport ]
-
-    wall       // frustum description
-    {
-        bottom_left  [ float float float ]
-        bottom_right [ float float float ]
-        top_left     [ float float float ]
-        type         fixed | HMD
-    }
-    projection // alternate frustum description, last one wins
-    {
-        origin       [ float float float ]
-        distance     float
-        fov          [ float float ]
-        hpr          [ float float float ]
-    }
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=181, lastline=202]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 non-planar to each other, and can overlap or have gaps between each
-other. A canvas can define a frustum, which will create default, planar
+other. A canvas can define a frustum, which will create default planar
 sub-frusta for its segments.
@@
 
-{\footnotesize\begin{lstlisting}
-canvas  # 0...n times
-{
-    name    string
-    layout  layout-ref   # 1...n times
-
-    wall       
-    {
-            bottom_left  [ float float float ]
-            bottom_right [ float float float ]
-            top_left     [ float float float ]
-            type         fixed | HMD
-    }
-    projection
-    {
-            origin       [ float float float ]
-            distance     float
-            fov          [ float float ]
-            hpr          [ float float float ]
-    }
-
-    segment {...}   # 1...n times
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=205, lastline=225]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 
+The segment eyes define which eyes are displayed by this segment. For active
+stereo outputs the default setting 'all' is normally used, while passive stereo
+segments define the left or right eye, and potentially the cyclop eye.
+
 A segment can define a frustum, in which case it overrides the default
@@
 
-{\footnotesize\begin{lstlisting}
-segment # 1...n times
-{
-    channel  string
-    name     string
-    viewport [ viewport ]
-
-    wall       // frustum description
-    {
-        bottom_left  [ float float float ]
-        bottom_right [ float float float ]
-        top_left     [ float float float ]
-        type         fixed | HMD
-    }
-    projection // alternate frustum description, last one wins
-    {
-        origin       [ float float float ]
-        distance     float
-        fov          [ float float ]
-        hpr          [ float float float ]
-    }
-}
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=225, lastline=246]
+  {../../../website/documents/design/fileFormat.shtml}}
 
@@
 
+The eye attribute defines which eyes are handled by this compound. This
+attribute can be used to write one compound for monoscopic rendering and another
+for stereoscopic rendering, as illustrated in \fig{fStereoSwitch}.
+
 The viewport restricts the rendering to the area relative to the parent
@@
 relative to the parent. The subpixel defines the jittering applied to
-the frustum, relative to the parent. The eye pass select the mono, left
-or right eye passes rendered by the compound. The zoom scales the parent
+the frustum, relative to the parent. The zoom scales the parent
 pixel viewport resolution. The DPlex period defines that $\frac{1}{period}$
@@
 are specified in meters. \fig{fFrusta} illustrates the frustum
-parameters.
+parameters. Setting a frustum on a compound is discouraged, a proper view and
+segment description should be used instead. View frusta override segment frusta
+which override compound frusta.
 
@@
 
-{\footnotesize\begin{lstlisting}
-compound
-{
-    name     string
-    channel  channel-ref  # where the compound tasks are executed
-
-    task     [ CLEAR DRAW READBACK ASSEMBLE ]
-    buffer   [ COLOR DEPTH ]         # default COLOR
-
-    viewport [ viewport ]            # wrt parent compound, sort-first
-    range    [ float float ]         # DB-range for sort-last
-    pixel    [ int int int int ]     # pixel decomposition (x y w h)
-    subpixel [ int int ]	     # subpixel decomposition (index size)
-    eye      [ CYCLOP LEFT RIGHT ]   # monoscopic or stereo view
-    zoom     [ float float ]         # up/downscale of parent pvp
-    period   int                     # DPlex period, every nth frame rendered
-    phase    int                     # DPlex phase, when to start rendering
-
-    load_equalizer # adjust 2D tiling or DB range of children
-    {
-        mode 2D | DB | VERTICAL | HORIZONTAL
-        damping float # 0: no damping, 1: no changes
-        boundary [ x y ] | float  # x,y: tile boundary for 2D, float: range
-    }
-    DFR_equalizer # adjust ZOOM to achieve constant framerate
-    {
-        framerate float # target framerate
-        damping float   # 0: no damping, 1: no changes
-    }
-    framerate_equalizer {} # smoothen window swapbuffer rate (DPlex)
-    monitor_equalizer {}   # set frame zoom when monitoring other views
-
-    attributes
-    {
-        stereo_mode                  AUTO | QUAD | ANAGLYPH    # default AUTO
-        stereo_anaglyph_left_mask    [ RED GREEN BLUE ] # default red
-        stereo_anaglyph_right_mask   [ RED GREEN BLUE ] # df green blue
-    }
-
-    wall       # frustum description
-    {
-        bottom_left  [ float float float ]
-        bottom_right [ float float float ]
-        top_left     [ float float float ]
-        type         fixed | HMD
-    }
-    projection # alternate frustum description, last one wins
-    {
-        origin       [ float float float ]
-        distance     float
-        fov          [ float float ]
-        hpr          [ float float float ]
-    }
-
-    child-compounds
-
-    swapbarrier  # compounds with the same barriername sync swap
-    {
-        name string
-        NV_group OFF | ON | unsigned
-        NV_barrier OFF | ON | unsigned
-    }
-    outputframe
-    {
-        name   string
-        buffer [ COLOR DEPTH ]
-        type texture | memory
-    }
-    inputframe
-    {
-        name string # corresponding output frame
-    }
-}
-
-channel-ref:  'string' | '(' channel-segment-ref ')'
-channel-segment-ref: ( canvas-ref ) segment-ref ( layout-ref ) view-ref
-canvas-ref:   'string' | 'index'
-segment-ref:  'string' | 'index'
-layout-ref:   'string' | 'index'
-view-ref:     'string' | 'index'
-observer-ref: 'string' | 'index'
-\end{lstlisting}}
+{\footnotesize\lstinputlisting[firstline=249, lastline=333]
+  {../../../website/documents/design/fileFormat.shtml}}
 

### 2011-03-21 Developer/ProgrammingGuide/paper.tex
Programming Guide: Fix typos & grammar, final 1.0 version.
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.8.RC1 for Equalizer 1.0$\beta$ \\[\medskipamount]
+Version 1.8 for Equalizer 1.0 \\[\medskipamount]
 %  \today
@@
 \lstset{language=C++}
-\section*{Equalizer 1.0$\beta$ Programming and User Guide}
+\section*{Equalizer 1.0 Programming and User Guide}
 \today
@@
 
-\textcopyright 2007-2010 \htmladdnormallink{Eyescale Software GmbH}
+\textcopyright 2007-2011 \htmladdnormallink{Eyescale Software GmbH}
 {http://www.eyescale.ch}. All rights reserved. No permission is
@@
     1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
-    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
-    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
-    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
-    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
-    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
-    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
-    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
-    1.7.7   & Oct 01, 2010 & SystemPipe and SystemWindow renaming\\
-    1.7.8   & Oct 04, 2010 & Document GPU affinity workaround hint\\
-    1.7.9   & Nov 03, 2010 & Add \sref{sErrorHandling}\\
-    1.7.10  & Dec 21, 2010 & Review for Equalizer 1.0$\alpha$\\
-    1.8.RC1  & Feb 22, 2010 & Review for Equalizer 1.0$\beta$\\
+    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
     \hline
@@
 
-Equalizer provides the domain-specific parallel rendering know-how and abstracts
-configuration, threading, synchronization, windowing and event handling. It is a
-`GLUT on steroids', providing parallel and distributed execution, scalable
-rendering features, an advanced network library and fully customizable event
-handling.
+Equalizer provides the domain-specific parallel rendering expertise and
+abstracts configuration, threading, synchronization, windowing and event
+handling. It is a `GLUT on steroids', providing parallel and distributed
+execution, scalable rendering features, an advanced network library and fully
+customizable event handling.
 
@@
 
-Collage, the underlying network library, uses a peer-to-peer
-structure. Equalizer utilizes this to provide optimal performance scalable
+Collage, the underlying network library, uses a peer-to-peer communication
+structure. Equalizer uses Collage to provide optimal performance scalable
 rendering, e.g., by setting up direct communication between two nodes when
@@
 
-\subsection{Overview}
+\subsection{Summary}
 
@@
 
-Further versions of Equalizer might introduce threaded windows, where
-all window-related task methods are executed in a separate operating
-system thread.
-
 \subsection{Window}
@@
 
+Canvases provide a convenient way to configure projection surfaces. A
+canvas uses layouts, which describe logical views. The usage of canvases
+and layouts is optional. Typically, each desktop window uses one canvas.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-
-Canvases provide a convenient way to configure projection surfaces. A
-canvas uses layouts, which describe logical views. The usage of canvases
-and layouts is optional. Typically, each desktop window uses one canvas.
-
 A canvas represents one logical projection surface, e.g., a PowerWall, a
@@
 
-A segment covers a part of its parent canvas, which is configured using
-the segment viewport. The viewport is in normalized coordinates with
-respect to the canvas. Segments might overlap (edge-blended projectors)
-or have gaps between each other (display walls,
-\fig{fCanvas}\footnote{Dataset courtesy of VolVis distribution of SUNY
-  Stony Brook, NY, USA.}). The viewport is used to configure the
-segment's default frustum from the canvas frustum description, and to
-place layout views correctly.
+A segment covers a part of its parent canvas, which is configured using the
+segment viewport. The viewport is in normalized coordinates relative to the
+canvas. Segments might overlap (edge-blended projectors) or have gaps between
+each other (display walls, \fig{fCanvas}\footnote{Dataset courtesy of VolVis
+  distribution of SUNY Stony Brook, NY, USA.}). The viewport is used to
+configure the segment's default frustum from the canvas frustum description, and
+to place layout views correctly.
 
@@
     \sref{sCanvas}.
-  \item[View/Segment channels] Set the channel by using the canvas,
-    segment, layout and view name. The compound frustum will be
-    calculated automatically based on the segment or view frustum. Note
-    that one segment may created multiple view/segment channels, one for
-    each view intersection of each layout used on the canvas. Only the
-    compounds belonging to the active layout of a canvas are activate at
-    runtime.
+  \item[View/Segment channels] Set the channel by using the canvas, segment,
+    layout and view name. The compound frustum will be calculated automatically
+    based on the segment or view frustum. Note that one segment may created
+    multiple view/segment channels, one for each view intersection of each
+    layout used on the canvas. Only the compounds belonging to the active layout
+    of a canvas are active at runtime.
     \begin{description}
     \item [Software swap lock] When using software swap synchronization,
-      swap-lock the destination compounds using a swapbarrier. All
-      windows with a swapbarrier of the same name synchronize their
+      swap-lock the destination compounds using a swap barrier. All
+      windows with a swap barrier of the same name synchronize their
       swapbuffers.
     \item [Hardware swap lock] When using hardware swap synchronization,
-      use swapbarriers for the destination compounds, setting
+      use swap barriers for the destination compounds, setting
       \textsf{NV\_group} and \textsf{NV\_barrier} appropriately. The
@@
 
-\subsubsection{\label{sSwapBarrier}Synchronization - Swapbarriers}
-Compounds may have a software or hardware swapbarrier to synchronize the
-buffer swap of multiple channels. A software swapbarrier is configured
-by giving it a name. Windows with a swapbarrier of the same name
+\subsubsection{\label{sSwapBarrier}Synchronization - Swap Barriers}
+Compounds may have a software or hardware swap barrier to synchronize the
+buffer swap of multiple channels. A software swap barrier is configured
+by giving it a name. Windows with a swap barrier of the same name
 synchronize with each other before executing the swap buffer
@@
 
-Swapbarrier names are only valid within the compound tree, that is, a
+Swap barrier names are only valid within the compound tree, that is, a
 compound from one compound tree cannot be synchronized with a compound
@@
 
-A hardware swapbarrier is configured by setting the \textsf{NV\_group}
-and \textsf{NV\_barrier} parameters. These parameters follow the
-\textsf{NV\_swap\_group} extension, which synchronizes all windows bound
-to the same group on a single machine, and all groups bound to the same
-barrier across systems.
+A hardware swap barrier is configured by setting the \textsf{NV\_group} and
+\textsf{NV\_barrier} parameters. These parameters follow the
+\textsf{NV\_swap\_group} extension, which synchronizes all windows bound to the
+same group on a single machine, and all groups bound to the same barrier across
+systems.
 
@@
 
-The dynamic frame resolution is not limited to downscaling the
-rendering resolution, it will also upscale the resolution if the
-parameters allow for it. Upscaled rendering, which will down-sample the
-result for display, provides dynamic anti-aliasing at a constant
-framerate.
+The dynamic frame resolution is not limited to subsampling the rendering
+resolution, it will also supersample the image if the source buffer is big
+enough. Upscaled rendering, which will down-sample the result for display,
+provides dynamic anti-aliasing at a constant framerate.
 
@@
 applications. It does not cover basic cluster management and driver
-installation. A prerequisite to the following steps is a pre-installed cluster,
+installation. A prerequisite to the following steps is a preinstalled cluster,
 including network and graphics card drivers.
@@
 
-Equalizer supports pre-launched render clients and automatic render client
-launching, if configured properly. Pre-launched render clients are started
-manually or by an external script on a pre-defined address. Auto-launched render
+Equalizer supports prelaunched render clients and automatic render client
+launching, if configured properly. Prelaunched render clients are started
+manually or by an external script on a predefined address. Auto-launched render
 client are started and stopped by the Equalizer server on demand.
 
-The two mechanism can co-exist. The server will first try to connect a
-pre-launched render client on the given connection descriptions for each
+The two mechanism can coexist. The server will first try to connect a
+prelaunched render client on the given connection descriptions for each
 node. If this fails, he will try to auto-launch the render client. If the render
@@
 
-\subsubsection{Pre-launched Render Clients}
-Pre-launched render clients are useful if setting up auto-launching is too
+\subsubsection{Prelaunched Render Clients}
+Prelaunched render clients are useful if setting up auto-launching is too
 time-consuming, e.g., on Windows which does not have a standard remote login
-procedure. The following steps are to be taken to use pre-launched render
+procedure. The following steps are to be taken to use prelaunched render
 clients:
@@
   \textsf{glxinfo -display :0.1}.
-\item The server does not find the pre-launched render client. Verify that the
+\item The server does not find the prelaunched render client. Verify that the
   client is listening on the correct IP and port, and that this IP and port are
   reachable from the server host.
-\item The server can not launch a render client. Check the server log for the
+\item The server cannot launch a render client. Check the server log for the
   launch command used, and try to execute a simple application from the server
@@
 
-In addition the the official, public API Equalizer exposeses a number of
+In addition the official, public API Equalizer exposes a number of
 unstable methods and, where unavoidable, internal APIs. These are clearly marked
@@
   often simplify the usage of OpenGL functions. Most classes in this namespace
-  are used by the the Equalizer client library, but are usable independently
+  are used by the Equalizer client library, but are usable independently
   from Equalizer for application development.
+\item[eq::admin] The eq::admin namespace implements an administrative API to
+  change the configurations of a running server. This admin API is not yet
+  finalized and will very likely change in the future.
 \item[eq::fabric] The eq::fabric namespace is the common data management and
@@
 
-The third step is to create an instance of the application and to
-initialize it locally. The application is a subclass of
-\textsf{eq::Client}, which in turn is an \textsf{co::LocalNode}. The
-underlying Collage network library, discussed in \sref{sNetwork}, is a
-peer-to-peer network of \textsf{co::LocalNode}s. The client/server
-concept is implement the higher-level \textsf{eq} client namespace.
+The third step is to create an instance of the application and to initialize it
+locally. The application is a subclass of \textsf{eq::Client}, which in turn is
+an \textsf{co::LocalNode}. The underlying Collage network library, discussed in
+\sref{sNetwork}, is a peer-to-peer network of \textsf{co::LocalNode}s. The
+client-server concept is implement the higher-level \textsf{eq} client
+namespace.
 
@@
   clients.
-\item \textbf{Resident render client}: Manually pre-started render
+\item \textbf{Resident render client}: Manually prestarted render
   client, listening on a specified port for server commands. This mode
@@
 The return value of \textsf{Config::init} depends on the configuration
-robustness attribute. This attribute is by default activated, allowing
-configurations to launch even when some entities failed to initialize. If
-activated, \textsf{Config::init} always returns true. If deactivated, it returns
+robustness attribute. This attribute is set by default, allowing configurations
+to launch even when some entities failed to initialize. If set,
+\textsf{Config::init} always returns true. If deactivated, it returns
 \textsf{true} only if all initialization task methods were successful. In any
@@
 \item The master instance of the object generates new versions for all
-  slaves. These versions are continous, starting at
+  slaves. These versions are continuous, starting at
   \textsf{co::VERSION\_FIRST}. It is possible to commit on slave instances, but
@@
 After a successful initialization, the models are loaded and registered for data
-distribution. When idle, Equalizer will pre-distribute object data during
+distribution. When idle, Equalizer will predistribute object data during
 registration to accelerate the mapping of slave instances. Registering the
@@
 
-The pre-process config instance maintains the mapped models, and lazily maps new
+The preprocess config instance maintains the mapped models, and lazily maps new
 models, which are registered by the application process:
@@
 stage of the processing, to the extreme of making it possible to disable
-all event-related code in Equalizer. In this aspect, Equalizer
+all event handling code in Equalizer. In this aspect, Equalizer
 substantially differs from GLUT, which imposes an event model and hides
@@
   data.
-\item[DRAW\_SYNC:] Additionaly to the synchronization of the async
+\item[DRAW\_SYNC:] Additionally to the synchronization of the async
   thread model, all local render threads are synchronized, so that the
@@
   per node.
-\item[LOCAL\_SYNC:] Additionaly to the synchronization of the async
+\item[LOCAL\_SYNC:] Additionally to the synchronization of the async
   thread model, all local frame operations, including readback, assemble
@@
 Non-threaded pipes should not be used for DPlex source and destination
-channels. Applications using the \textsf{local\_sync} thread model can
-not benefit from DPlex if the application node uses a DPlex source or
-destination channel.
+channels. Applications using the \textsf{local\_sync} thread model cannot
+benefit from DPlex if the application node uses a DPlex source or destination
+channel.
 
@@
 
-The config event is dispatched to \textsf{View::handleEvent}. Based on
-the original size and wall or projection description of the view, a new
-wall or projection is computed, keeping the aspect ratio and the height
-of the frustum constant. This new frustum is automatically applied by
-Equalizer at the next config frame.
+The config event is dispatched to \textsf{View::handleEvent}. Using the original
+size and wall or projection description of the view, a new wall or projection is
+computed, keeping the aspect ratio and the height of the frustum constant. This
+new frustum is automatically applied by Equalizer at the next config frame.
 
@@
 
-Furthermore it should be possible for the application to implement a
-read back of only the relevant region of interest, that is, the 2D area
-of the framebuffer actually updated during rendering. This optimization
-will be fully supported by later versions of Equalizer.
+Furthermore it has to be possible for the application to implement a read back
+of only the relevant region of interest, that is, the 2D area of the framebuffer
+updated during rendering. This optimization will be fully supported by later
+versions of Equalizer.
 
@@
 composite either directly all images on the GPU, or use a CPU-based
-compositor and then transfer the pre-integrated result to the GPU. The
+compositor and then transfer the preintegrated result to the GPU. The
 high-level entry points automatically select the best algorithm. The
@@
 showing how to combine it with subpixel compounds. The same basic algorithm
-might be applied to other multi-sampling effects, e.g., depth-of-field.
+might be applied to other multisampling effects, e.g., depth-of-field.
 
@@
 
-The supersampling algorith used by eqPly performs anti-aliasing, using a
+The supersampling algorithm used by eqPly performs anti-aliasing, using a
 technique which computes samples within a pixel by jittering the frustum by a
@@
 
-The \textsf{FrameData} class holds the information whether or not the
-application is in idle mode. The \textsf{Config} updates the idle mode
-information depending on the application's state.  Each \textsf{Channel} will
-perform anti-aliasing when no user event requires a redraw.
+The \textsf{FrameData} class holds the information whether the application is in
+idle mode. The \textsf{Config} updates the idle mode information depending on
+the application's state.  Each \textsf{Channel} will perform anti-aliasing when
+no user event requires a redraw.
 
@@
 position, which will be used to move the frustum. The \textsf{getJitter} method
-will return the the computed jitter vector for the current frustum. This method
-has a default implementation in \textsf{eq::Channel} for subpixel compounds, but
-is overwritten in \textsf{eqPly} to perform idle anti-aliasing:
+will return the computed jitter vector for the current frustum. This method has
+a default implementation in \textsf{eq::Channel} for subpixel compounds, but is
+overwritten in \textsf{eqPly} to perform idle anti-aliasing:
 
@@
 been not been compressed during download. For network transfer it has been
-compressed to 1\% of its original size since it contains a lot of black
-background pixels. For readback the plugin with the name \textsf{0x101,
+compressed to 1\% of its original size since it contains many black background
+pixels. For readback the plugin with the name \textsf{0x101,
   EQ\_COMPRESSOR\_TRANS\-FER\_RGBA\_TO\_BGRA} has been used, and for compression
@@
   communicate with each other using connections. A LocalNode listens on various
-  various connections and processes requests for a given process. A Node is a
-  proxy for a remote LocalNode.
+  connections and processes requests for a given process. A Node is a proxy for
+  a remote LocalNode.
 \item[Object] Provides object-oriented data distribution of C++ objects between
@@
 Distributed objects can be static (immutable) or dynamic. Dynamic objects are
-versioned. New versions are commited from the master instance, which sends
+versioned. New versions are committed from the master instance, which sends
 the delta between the previous and current version to all mapped slave
@@
 When using \textsf{commit} on slave object instances, multiple modifications on
-the same database may happen simultaneously. In order to avoid excessive
-synchronization in the network layer, Equalizer a light-weight approach which
-requires the application to avoid or handle conflicting changes on the same
-object.
+the same database may happen simultaneously. To avoid excessive synchronization
+in the network layer, Equalizer a lightweight approach which requires the
+application to avoid or handle conflicting changes on the same object.
 
@@
 unpacked. The data is synchronized in the order it has been received from the
-slaves. If a specific order is needed, it is the application's responsibity to
+slaves. If a specific order is needed, it is the application's responsibility to
 serialize the slave commits accordingly.
@@
 
-\fig{fSlaveCommit} depicts one possible communication sequence. In general, it
-is best to adhere to the following rules:
+\fig{fSlaveCommit} depicts one possible communication sequence. Where possible,
+it is best to adhere to the following rules:
 \begin{itemize}
@@
 implementing data distribution between the server, application and render
-clients. Based on dirty bits, \textsf{eq::fabric::Serializable} implements one
+clients. Using dirty bits, \textsf{eq::fabric::Serializable} implements one
 usage pattern for the \textsf{co::Object}, allowing inheritance with data
@@
   \textbf{Details} & The listening port used by the server or
-  node. For nodes, the port can be used to contact pre-started, resident
+  node. For nodes, the port can be used to contact prestarted, resident
   render client nodes or to use a specific port for the node. If 0 is
@@
 
-Furthermore, a port for the socket can be specified. When no port is
-specified for the server, the default port 4242 (+UID on Posix systems)
-is used. When no port is specified for a node, a random port will be
-chosen by the operating system. For pre-launched render clients, a port
-has to be specified for the server to find the client node.
+Furthermore, a port for the socket can be specified. When no port is specified
+for the server, the default port 4242 (+UID on Posix systems) is used. When no
+port is specified for a node, a random port will be chosen by the operating
+system. For prelaunched render clients, a port has to be specified for the
+server to find the client node.
 
@@
 
-The viewport of the window is relative to the pipe. It can be specified
-in relative or absolute coordinates. Relative coordinates are normalized
-coordinates with respect to the pipe, e.g., a viewport of \textsf{[ 0.25
-  0.25 0.5 0.5 ]} creates a window in the middle of the screen, using
-50\% of the pipe's size. Absolute coordinates are integer pixel values,
-e.g., a viewport of \textsf{[ 50 50 800 600 ]} creates a window 50 pixels
-from the upper-left corner, sized 800x600 pixels, regardless of the
-pipe's size. The default viewport is \textsf{[ 0 0 1 1 ]}, i.e., a
-full-screen window.
+The viewport of the window is relative to the pipe. It can be specified in
+relative or absolute coordinates. Relative coordinates are normalized
+coordinates relative to the pipe, e.g., a viewport of \textsf{[ 0.25 0.25 0.5
+    0.5 ]} creates a window in the middle of the screen, using 50\% of the
+pipe's size. Absolute coordinates are integer pixel values, e.g., a viewport of
+\textsf{[ 50 50 800 600 ]} creates a window 50 pixels from the upper-left
+corner, sized 800x600 pixels, regardless of the pipe's size. The default
+viewport is \textsf{[ 0 0 1 1 ]}, i.e., a full-screen window.
 

### 2011-03-25 Developer/ProgrammingGuide/paper.tex
delete empty dir
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[ ] Documentation
Programming Guide: Minor revision
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.8 for Equalizer 1.0 \\[\medskipamount]
+Version 1.8.1 for Equalizer 1.0 \\[\medskipamount]
 %  \today
@@
     1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
+    1.8.1   & Mar 25, 2011 & Minor Revision\\
     \hline
@@
 The main function starts off with initializing the Equalizer library. The
-command line arguments are passed on to Equalizer. The are used to set certain
+command line arguments are passed on to Equalizer. They are used to set certain
 default values based on Equalizer-specific options\footnote{Equalizer-specific
@@
 and \textsf{applyInstanceData} to serialize and deserialize the object's
-distributed data. These method provide an output or input stream as a
+distributed data. These methods provide an output or input stream as a
 parameter, which abstracts the data transmission and can be used like a
@@
 
-An \textsf{co::ConnectionSet} is used to manage multiple connections. The
+A \textsf{co::ConnectionSet} is used to manage multiple connections. The
 typical use case is to have one or more listening connections for the local
@@
 
-A remote \textsf{Node} can be either be connected explicitly be the application
-or due to a connection from the given remote node. The explicit connection can
-be done by programmatically creating a node, adding the necessary
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
+  {\caption{\label{fNetNode}Communication between two Nodes}}
+\end{wrapfigure}
+A remote \textsf{Node} can either be connected explicitly by the application or
+due to a connection from the given remote node. The explicit connection can be
+done by programmatically creating a node, adding the necessary
 \textsf{ConnectionDescription}s and connecting it to the local node. It may also
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
-  {\caption{\label{fNetNode}Communication between two Nodes}}
-\end{wrapfigure}
 \fig{fNetNode} shows the communication between two nodes. The nodes in
@@
 
-\if 0
-{\footnotesize\begin{lstlisting}
-\end{lstlisting}}
-\fi
-
 \cleardoublepage

### 2011-04-04 Developer/ProgrammingGuide/paper.tex
Programming Guide: add focus distance section
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.8.1 for Equalizer 1.0 \\[\medskipamount]
+Version 1.8.2 for Equalizer 1.0+git \\[\medskipamount]
 %  \today
@@
     1.8.1   & Mar 25, 2011 & Minor Revision\\
+    1.8.2   & Apr 04, 2011 & Added \sref{sFocusDistance}\\
     \hline
@@
 {\footnotesize\lstinputlisting[firstline=107, lastline=123]
-  {../../../src/examples/eqHello/hello.cpp}}
+  {../../../Equalizer/examples/eqHello/hello.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=57, lastline=66]
-  {../../../src/examples/eqPly/main.cpp}}
+  {../../../Equalizer/examples/eqPly/main.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=40, lastline=55]
-  {../../../src/examples/eqPly/main.cpp}}
+  {../../../Equalizer/examples/eqPly/main.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=67, lastline=70]
-  {../../../src/examples/eqPly/main.cpp}}
+  {../../../Equalizer/examples/eqPly/main.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=71, lastline=78]
-  {../../../src/examples/eqPly/main.cpp}}
+  {../../../Equalizer/examples/eqPly/main.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=80, lastline=81]
-  {../../../src/examples/eqPly/main.cpp}}
+  {../../../Equalizer/examples/eqPly/main.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=83]
-  {../../../src/examples/eqPly/main.cpp}}
+  {../../../Equalizer/examples/eqPly/main.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=80, lastline=88]
-  {../../../src/examples/eqPly/eqPly.cpp}}
+  {../../../Equalizer/examples/eqPly/eqPly.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=90, lastline=100]
-  {../../../src/examples/eqPly/eqPly.cpp}}
+  {../../../Equalizer/examples/eqPly/eqPly.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=101, lastline=118]
-  {../../../src/examples/eqPly/eqPly.cpp}}
+  {../../../Equalizer/examples/eqPly/eqPly.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=120, lastline=147]
-  {../../../src/examples/eqPly/eqPly.cpp}}
+  {../../../Equalizer/examples/eqPly/eqPly.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=148, lastline=152]
-  {../../../src/examples/eqPly/eqPly.cpp}}
+  {../../../Equalizer/examples/eqPly/eqPly.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=154, lastline=165]
-  {../../../src/examples/eqPly/eqPly.cpp}}
+  {../../../Equalizer/examples/eqPly/eqPly.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=167, lastline=175]
-  {../../../src/examples/eqPly/eqPly.cpp}}
+  {../../../Equalizer/examples/eqPly/eqPly.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=61, lastline=74]
-  {../../../src/examples/eqPly/initData.cpp}}
+  {../../../Equalizer/examples/eqPly/initData.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=122, lastline=129]
-  {../../../src/examples/eqPly/frameData.h}}
+  {../../../Equalizer/examples/eqPly/frameData.h}}
 {\footnotesize\lstinputlisting[firstline=54, lastline=80]
-  {../../../src/examples/eqPly/frameData.cpp}}
+  {../../../Equalizer/examples/eqPly/frameData.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=143, lastline=174]
-  {../../../src/examples/eqPly/vertexBufferDist.cpp}}
+  {../../../Equalizer/examples/eqPly/vertexBufferDist.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=74, lastline=95]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=97, lastline=99]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=100, lastline=118]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=130, lastline=139]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=310, lastline=317]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=319, lastline=327]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=329, lastline=359]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=387, lastline=390]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 {\footnotesize\lstinputlisting[firstline=450, lastline=481]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 {\footnotesize\lstinputlisting[firstline=528, lastline=534]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=284, lastline=308]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=790, lastline=859]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=35, lastline=48]
-  {../../../src/examples/eqPly/node.cpp}}
+  {../../../Equalizer/examples/eqPly/node.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=257, lastline=268]
-  {../../../src/examples/eqPly/config.cpp}}
+  {../../../Equalizer/examples/eqPly/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=57, lastline=70]
-  {../../../src/examples/eqPly/pipe.cpp}}
+  {../../../Equalizer/examples/eqPly/pipe.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=72, lastline=78]
-  {../../../src/examples/eqPly/pipe.cpp}}
+  {../../../Equalizer/examples/eqPly/pipe.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=39, lastline=55]
-  {../../../src/examples/eqPly/pipe.cpp}}
+  {../../../Equalizer/examples/eqPly/pipe.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=80, lastline=84]
-  {../../../src/examples/eqPly/pipe.cpp}}
+  {../../../Equalizer/examples/eqPly/pipe.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=76, lastline=98]
-  {../../../src/examples/eqPly/window.cpp}}
+  {../../../Equalizer/examples/eqPly/window.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=100, lastline=109]
-  {../../../src/examples/eqPly/window.cpp}}
+  {../../../Equalizer/examples/eqPly/window.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=69, lastline=78]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=123, lastline=137]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=139, lastline=165]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=167, lastline=178]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=591, lastline=610]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=619, lastline=689]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=238, lastline=261]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=36, lastline=57]
-  {../../../src/examples/eqPixelBench/configEvent.h}}
+  {../../../Equalizer/examples/eqPixelBench/configEvent.h}}
 
@@
 {\footnotesize\lstinputlisting[firstline=58, lastline=71]
-  {../../../src/examples/eqPixelBench/config.cpp}}
+  {../../../Equalizer/examples/eqPixelBench/config.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=55, lastline=61]
-  {../../../src/examples/eVolve/main.cpp}}
+  {../../../Equalizer/examples/eVolve/main.cpp}}
 ...
 {\footnotesize\lstinputlisting[firstline=92, lastline=95]
-  {../../../src/examples/eVolve/main.cpp}}
+  {../../../Equalizer/examples/eVolve/main.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=36, lastline=44]
-  {../../../src/examples/eVolve/error.h}}
+  {../../../Equalizer/examples/eVolve/error.h}}
 
@@
 {\footnotesize\lstinputlisting[firstline=45, lastline=57]
-  {../../../src/examples/eVolve/window.cpp}}
+  {../../../Equalizer/examples/eVolve/window.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=37, lastline=73]
-  {../../../src/examples/eVolve/error.cpp}}
+  {../../../Equalizer/examples/eVolve/error.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=37, lastline=43]
-  {../../../src/examples/eVolve/window.cpp}}
+  {../../../Equalizer/examples/eVolve/window.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=124, lastline=140]
-  {../../../src/libs/client/aglWindow.cpp}}
+  {../../../Equalizer/libs/client/aglWindow.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=67, lastline=101]
-  {../../../src/libs/client/glXWindow.cpp}}
+  {../../../Equalizer/libs/client/glXWindow.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=178, lastline=224]
-  {../../../src/libs/client/wglWindow.cpp}}
-
-\subsection{\label{sTracking}Head Tracking}
-
-The \textsf{eqPly} example contains rudimentary support for head
-tracking to show how head tracking can be integrated with
-Equalizer. Support for a wide range of tracking devices is not within
-the scope of Equalizer. Other open source and commercial implementations
-cover this functionality sufficiently and can easily be integrated with
-Equalizer.
-
-An Equalizer configuration has a vector of observers. Each observer has
-its own head matrix and eye separation for tracking. In \textsf{eqPly},
-all observers are updated with the same head matrix. A typical
-application would assign different tracking data to different observers.
-
-\begin{figure}[h!t]\center
-  \subfigure[]{\label{fMono}
-    \includegraphics[height=4cm]{images/mono}}\hfil
-  \subfigure[]{\label{fStereo}
-    \includegraphics[height=4cm]{images/stereo}}
-  \subfigure[]{\label{fTracked}
-    \includegraphics[height=4cm]{images/tracked}}\hfil
-  \subfigure[]{\label{fHMD}
-    \includegraphics[height=4cm]{images/hmd}}
-  {\caption{\label{fImmersive}Monoscopic (a), Stereoscopic (b), Tracked (c) and
-      HMD (d) Frusta}}
-\end{figure}
+  {../../../Equalizer/libs/client/wglWindow.cpp}}
+
+\subsection{\label{sTracking}Stereo and Immersive Rendering}
+
+Equalizer fully supports immersive rendering for Virtual Reality, supporting
+multiple tracked observers, head-mounted displays (HMD) and flexible focus
+distance support. Equalizer does not support tracking devices directly. Other
+open source and commercial implementations cover this functionality well
+and can easily be integrated with Equalizer.
+
+An Equalizer configuration contains a number of observers. Each observer
+represents on tracked entity. Most immersive configurations only have one
+observer. It is possible to support multiple tracked viewers in the same
+configuration, e.g., to use two HMDs.
+
+Each observer has its own head matrix, eye separation and focus information for
+tracking. In \textsf{eqPly}, all observers are updated with the same data. A
+typical application would assign different tracking data to different observers.
+
+The \textsf{eqPly} example contains basic support for head tracking to
+illustrate how applications use these features.
+
+\subsubsection{Stereo Rendering}
 
@@
 
+\begin{figure}[h!t]\center
+  \subfigure[]{\label{fMono}
+    \includegraphics[width=.382\textwidth]{images/mono}}\hfil
+  \subfigure[]{\label{fStereo}
+    \includegraphics[width=.382\textwidth]{images/stereo}}
+  {\caption{Monoscopic (a) and Stereoscopic (b) Frusta}}
+\end{figure}
+
 In stereo rendering, the scene is rendered twice, with the two frusta
-'moved' by the distance between the eyes, as shown in \fig{fStereo}.
+'moved' by the distance between the eyes, as shown in \fig{fStereo}. The stereo
+frusta are asymmetric, offset by half of the eye base. The stereo convergence
+plane is the same as the projection surface, unless specified otherwise using
+the focus distance API (see \sref{sFocusDistance}).
+
+Note that while stereo rendering is often implemented the simpler toe-in method,
+Equalizer uses the correct approach using asymmetric frusta.
+
+\subsubsection{Immersive Rendering}
 
@@
 the application using \textsf{Config::setHead\-Matrix}, which is used by
-the server to compute the frusta. The resulting off-axis frusta are
-positioned using the channel's head transformation, which can be
-retrieved using \textsf{Channel::getHeadTransform}.
+the server to compute the eye positions, and consequently the view frusta.
+
+\begin{figure}[h!t]\center
+  \subfigure[]{\label{fTracked}
+    \includegraphics[width=.382\textwidth]{images/tracked}}\hfil
+  \subfigure[]{\label{fHMD}
+    \includegraphics[width=.382\textwidth]{images/hmd}}
+  {\caption{\label{fImmersive}Tracked (a) and HMD (b) Immersive Frusta}}
+\end{figure}
+
+Projection surfaces which are not X/Y-planar create frusta which are not
+oriented along the Z axis, as shown in \fig{fFixed}. These frusta are positioned
+using the channel's head transformation, which can be retrieved using
+\textsf{Channel::getHeadTransform}.
 
@@
 in different projections compared to normal tracking with fixed
-projection screens.
+projection screens. These difference is transparent to Equalizer applications,
+only the configuration file uses a different wall type for HMD's.
+
+\subsubsection{\label{sFocusDistance}Focus Distance}
+
+The focus distance, also called stereo convergence, is the Z distance in which
+the left and right eye frusta converge. A plane parallel to the near plane
+positioned in the focus distance creates the same 2D image for both eyes.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/focusFixed}
+  {\caption{\label{fFixed}Fixed focus distance}}
+\end{wrapfigure}
+The frustum calculation up to Equalizer version 1.0 places the stereo
+convergence on the projection surface, as shown in \fig{fFixed}. The focus
+distance is therefore the distance between the origin and the middle of the
+projection surface.
+
+In Equalizer 1.1 and later the focus distance and mode can be configured and
+changed at runtime for each \textsf{observer} and on the \textsf{config} for
+untracked views.
+
+The default focus mode \textsf{fixed} implements the algorithm used by Equalizer
+1.0. This mode ignores the focus distance setting. The focus modes
+\textsf{relative\_to\_origin} and \textsf{relative\_to\_observer} use the focus
+distance to dynamically change the stereo convergence.
+
+\begin{figure}[h!t]\center
+  \subfigure[]{\label{fOrigin}
+    \includegraphics[width=.49\textwidth]{images/focusOrigin}}\hfil
+  \subfigure[]{\label{fObserver}
+    \includegraphics[width=.49\textwidth]{images/focusObserver}}
+  {\caption{\label{fImmersive}Dynamic focus distance relative to origin
+      (a) and observer (b)}}
+\end{figure}
+
+The focus distance calculation relative to the origin allows to change the
+focus independent of the observer position by separating the projection surface
+from the stereo convergence plane upon which the observer focuses itself. The
+convergence plane of the first wall in the negative Z direction is moved to be
+at the given focus distance, as shown in \fig{fOrigin}. All other walls are
+moved by the same relative amount. The movement is made from the view of the
+central eye, thus leaving the mono frustum unchanged. \fig{fOrigin} shows the
+new logical 'walls' used for frustum calculations in white, while the physical
+projection from \fig{fFixed} is still visible in the background.
+
+The focus distance calculation relative to the observer is similar to the origin
+algorithm, but it keeps the closest wall in the observer's view direction at the
+given focus distance, as shown in \fig{fObserver}. When the observer moves
+forward, the focal plane moves forward as well. Consequently, when the observer
+looks in a different direction, a different object in the scene is focused, as
+indictated by the dotted circle in \fig{fObserver}.
+
+\begin{figure}[h!t]\center
+  \subfigure[]{\label{fFixedEx}
+    \includegraphics[height=6.3cm]{images/focusFixedEx}}\hfil
+  \subfigure[]{\label{fOriginEx}
+    \includegraphics[height=6.3cm]{images/focusOriginEx}}\hfil
+  \subfigure[]{\label{fObserverEx}
+    \includegraphics[height=6.3cm]{images/focusObserverEx}}
+            {\caption{\label{fImmersiveEx}Fixed(a), relative to origin(b) and
+                observer(c) focus distance examples}}
+\end{figure}
+
+\fig{fImmersiveEx} shows an example for the three focus distance modes. The
+configured wall is one meter behind the origin, the model two meters behind and
+the observer is half a meter in front of the origin. The focus distance was set
+to two meters.
 
@@
 {\footnotesize\lstinputlisting[firstline=59, lastline=64, xleftmargin=-7em]
-  {../../../src/examples/eqPly/view.h}}
+  {../../../Equalizer/examples/eqPly/view.h}}
 {\footnotesize\lstinputlisting[firstline=58, lastline=76]
-  {../../../src/examples/eqPly/view.cpp}}
+  {../../../Equalizer/examples/eqPly/view.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=40, lastline=47]
-  {../../../src/examples/eqPly/view.cpp}}
+  {../../../Equalizer/examples/eqPly/view.cpp}}
 {\footnotesize\lstinputlisting[firstline=78, lastline=85]
-  {../../../src/examples/eqPly/view.cpp}}
+  {../../../Equalizer/examples/eqPly/view.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=547, lastline=570]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=491, lastline=533]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=305, lastline=337]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=346, lastline=371]
-  {../../../src/examples/eqPly/channel.cpp}}
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=87, lastline=98]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=98, lastline=108]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 {\footnotesize\lstinputlisting[firstline=172, lastline=172]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 {\footnotesize\lstinputlisting[firstline=178, lastline=178]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 {\footnotesize\lstinputlisting[firstline=205, lastline=205]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 {\footnotesize\lstinputlisting[firstline=249, lastline=249]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=107, lastline=125]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=125, lastline=136]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=136, lastline=158]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=158, lastline=167]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=172, lastline=176]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=178, lastline=181]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=181, lastline=202]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=205, lastline=225]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=225, lastline=246]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=249, lastline=333]
-  {../../../website/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
 

### 2011-05-26 Developer/ProgrammingGuide/paper.tex
Minor spelling fix in programming guide
Small wording fix
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[x] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[ ] Documentation

@@
 the same database may happen simultaneously. To avoid excessive synchronization
-in the network layer, Equalizer a lightweight approach which requires the
+in the network layer, Equalizer implements a lightweight approach which requires the
 application to avoid or handle conflicting changes on the same object.
@@
 barrier is a versioned object. Each version can have a different height,
-and enter request are automatically grouped together by version.
+and enter requests are automatically grouped together by version.
 
@@
 \textsf{eq::fabric::Object}. This class manages common entity data, and is not
-intended to by subclassed directly.  The inheritance chain \textsf{eq::View}
+intended to be subclassed directly.  The inheritance chain \textsf{eq::View}
 $\rightarrow$ \textsf{eq::fabric::View} $\rightarrow$

### 2011-08-10 Developer/ProgrammingGuide/paper.tex
Programming Guide: add focus distance section
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.8.2 for Equalizer 1.0+git \\[\medskipamount]
+Version 1.8.3 for Equalizer 1.1.3+git \\[\medskipamount]
 %  \today
@@
     1.8.2   & Apr 04, 2011 & Added \sref{sFocusDistance}\\
+    1.8.3   & Aug 10, 2011 & Rewrote \sref{sHello}\\
     \hline
@@
 
-\section{Hello, World!}
+\section{\label{sHello}Hello, World!}
 
@@
 \end{wrapfigure}
-The \textsf{eqHello} example is a minimal application to illustrate the
-basic principle of an Equalizer application: The application developer
-has to implement the rendering method \textsf{Channel::frameDraw},
-similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
-run as a stand-alone application from the command line.
-
-The \textsf{eqHello} redraw function renders six colored quads, rotating
-around the origin. The \textsf{frameDraw} meth\-od provided by the
-\textsf{eq::Channel} can be used as a convenience function to setup the
-frustum and other OpenGL state. After setting up some lighting
-parameters, \textsf{eqHello} rotates the scene and renders the quads
-using immediate mode:
-
-{\footnotesize\lstinputlisting[firstline=107, lastline=123]
+The \textsf{eqHello} example is a minimal application to illustrate the basic
+principle of any Equalizer application: The application developer has to
+implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
+applications.
+
+It can be run as a stand-alone application from the command line or using an
+explicit configuration file for a visualization cluster. In the stand-alone
+case, any Equalizer application including eqHello, will automatically
+configure itself to use all graphics cards found on the local system for
+scalability.
+
+\subsection{Sequel}
+
+Sequel is a thin layer on top of the canonical Equalizer programming
+interface. It substantially simplifies the creation of parallel rendering
+applications by implementing proven design patterns of Equalizer
+applications. Sequel does not prohibit the usage of Equalizer or Collage
+functionality, it simply makes it easier to use for the common use cases.
+
+\subsection{The main Function}
+
+The main function instantiates an application object, initializes it, starts the
+main loop and finally de-initializes the application:
+
+{\footnotesize\lstinputlisting[firstline=58, lastline=66]
   {../../../Equalizer/examples/eqHello/hello.cpp}}
 
-The \textsf{eqHello} main function sets up the communication with the
-server, initializes and drives the rendering. The details of this setup
-are explained in \sref{sEqPly}.
+\subsection{The Application}
+
+The application object in Sequel represents one process in the cluster. The main
+application instance has the rendering loop and controls all execution. All
+other instances used for render client processes are passive and driven by
+Equalizer.
+
+The application is responsible for creating renderers. Sequel will request one
+renderer for each GPU rendering thread. Sequel also provides automatic mapping
+and synchronization of distributed objects, for which the application has to
+provide a creation callback. Since \textsf{eqHello} does not need any
+distributed objects it provides an empty implementation of
+\textsf{createObject}:
+
+{\footnotesize\lstinputlisting[firstline=47, lastline=55]
+  {../../../Equalizer/examples/eqHello/hello.cpp}}
+
+\subsection{The Renderer}
+
+The renderer is responsible for executing the application's rendering code. One
+instance for each GPU is used. All calls to a single renderer are executed
+serially and therefore thread-safe.
+
+In \textsf{eqHello}, the renderer draws six colored quads. The only change from
+a standard OpenGL application is the usage of the rendering context provided by
+Equalizer, most notably the frustum and viewport. The rendering context is
+described in detail in \sref{sRenderContext}, and \textsf{eqHello} simply calls
+\textsf{applyRenderContext} which will execute the appropriate OpenGL calls.
+
+After setting up lighting, the model is positioned using
+\textsf{applyModelMatrix}. For convenience, Sequel maintains one camera per
+view. The usage of this camera is purely optional, an application can implement
+its own camera model.
+
+The actual OpenGL code, rendering six colored quads, is omitted here for
+brevity:
+
+{\footnotesize\lstinputlisting[firstline=69, lastline=81]
+  {../../../Equalizer/examples/eqHello/hello.cpp}}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=57, lastline=66]
+{\footnotesize\lstinputlisting[firstline=57, lastline=68]
   {../../../Equalizer/examples/eqPly/main.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=40, lastline=55]
+{\footnotesize\lstinputlisting[firstline=40, lastline=57]
   {../../../Equalizer/examples/eqPly/main.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=67, lastline=70]
+{\footnotesize\lstinputlisting[firstline=70, lastline=73]
   {../../../Equalizer/examples/eqPly/main.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=71, lastline=78]
+{\footnotesize\lstinputlisting[firstline=74, lastline=81]
   {../../../Equalizer/examples/eqPly/main.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=80, lastline=81]
+{\footnotesize\lstinputlisting[firstline=83, lastline=84]
   {../../../Equalizer/examples/eqPly/main.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=83]
+{\footnotesize\lstinputlisting[firstline=85]
   {../../../Equalizer/examples/eqPly/main.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=61, lastline=74]
+{\footnotesize\lstinputlisting[firstline=60, lastline=73]
   {../../../Equalizer/examples/eqPly/initData.cpp}}
@@
 
+\label{sRenderContext}
 The contextual information has to be used to render the view as
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
-\end{wrapfigure}
 The concrete implementation of these window interfaces is responsible
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
+  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
+\end{wrapfigure}
 For each window, one \textsf{GLXEventHandler} is allocated. FBO's and PBuffers
@@
 
-{\footnotesize\lstinputlisting[firstline=124, lastline=140]
-  {../../../Equalizer/libs/client/aglWindow.cpp}}
+{\footnotesize\lstinputlisting[firstline=138, lastline=154]
+  {../../../Equalizer/libs/eq/client/agl/window.cpp}}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=67, lastline=101]
-  {../../../Equalizer/libs/client/glXWindow.cpp}}
+{\footnotesize\lstinputlisting[firstline=70, lastline=106]
+  {../../../Equalizer/libs/eq/client/glx/window.cpp}}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=178, lastline=224]
-  {../../../Equalizer/libs/client/wglWindow.cpp}}
+{\footnotesize\lstinputlisting[firstline=180, lastline=226]
+  {../../../Equalizer/libs/eq/client/wgl/window.cpp}}
 
@@
 
+\begin{wrapfigure}{r}{.612\textwidth}
+  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
+  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
+\end{wrapfigure}
 Equalizer commits dirty layout entities at the beginning of each
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/stereoSwitch.pdf}
+  {\caption{\label{fStereoSwitch}Using two different decompositions during
+      stereo and mono rendering}}
+\end{wrapfigure}
 The configuration contains stereo information in three places: the view, the
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/stereoSwitch.pdf}
-  {\caption{\label{fStereoSwitch}Using two different decompositions during
-      stereo and mono rendering}}
-\end{wrapfigure}
 The compound has an \textsf{eye} attribute defining which eyes it is capable of
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\label{fDirectSend}Direct Send Compositing}}
+\end{wrapfigure}
 After performing the sort-last compositing, the color information is
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
-\end{wrapfigure}
 Database decomposition has the advantage of scaling any part of the volume
@@
 
-\begin{wrapfigure}{l}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 Finding the correct assembly order for perspective frusta is more

### 2011-11-18 Developer/ProgrammingGuide/paper.tex
Add auto-config section to User Guide

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.8.3 for Equalizer 1.1.3+git \\[\medskipamount]
+Version 1.8.4 for Equalizer 1.1.5+git \\[\medskipamount]
 %  \today
@@
     1.8.3   & Aug 10, 2011 & Rewrote \sref{sHello}\\
+    1.8.4   & Nov 18, 2011 & Added \sref{sAutoConfig}\\
     \hline
@@
 
-\section{\label{sConfig}Writing Configurations}
+\section{\label{sConfig}Configuring Equalizer Clusters}
+
+\subsection{\label{sAutoConfig}Auto-Configuration}
+
+\subsubsection{GPU Service Discovery}
+Equalizer uses the GPU-SD library for discovering local and remote GPUs. Based
+on this information, a typical configuration is dynamically compiled at
+application start. The auto-configuration uses the same format as the static
+configuration files, and can therefore be used to create template configurations
+by running a test application and copying the generated configuration.
+
+Please note that the ZeroConf module is optional and requires a ZeroConf
+implementation when Equalizer is compiled.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/gpu-sd.pdf}
+  {\caption{\label{fGPUSD}GPU discovery for auto-configuration}}
+\end{wrapfigure}
+GPU-SD contains three components: a core library, local and remote discovery
+modules as well as a ZeroConf daemon. The daemon uses the appropriate module to
+query local GPUs and announces them using ZeroConf on the local network.
+
+Equalizer uses the library with the dns\_sd discovery module to gather the
+information announced by all daemons on the local network, as well as the local
+discovery modules for standalone auto-configuration. \fig{fGPUSD} shows how
+GPU-SD is used to discover four remote GPUs on two different machines to use
+them to render a single view on a laptop.
+
+\subsubsection{Usage}
+
+On each node contributing to the configuration, install and start the GPU-SD
+daemon. If multiple, disjoint configurations are used on the same network,
+provide the session name as a parameter when starting the daemon. Verify that
+all GPUs are visible on the application host using the \textsf{gpu\_sd\_list}
+tool. When starting the application, use the command-line parameter
+\textsf{--eq-config}:
+
+\begin{itemize}
+\item The default value is \textsf{local} which uses all local GPUs queried
+  using the cgl, glx or wgl GPU-SD module.
+\item \textsf{--eq-config sessionname} uses the dns\_sd ZeroConf module of
+  GPU-SD to query all GPUs in the subnet for the given session. The default
+  session of the GPU-SD daemon is \textsf{default}. The ZeroConf hostname is
+  used to connect the nodes.
+\item \textsf{--eq-config filename.eqc} loads a static configuration from the
+  given ascii file. The following sections describe how to write configuration
+  files.
+\end{itemize}
+
+The auto-configuration creates one display window on the local machine, and one
+off-screen channel for each GPU. The display window has one full-window channel
+used as an output channel for a single segment. It combines all GPUs into a
+scalability config with different layouts for each of the following scalability
+modes:
+
+\begin{description}
+\item [2D] A dynamically load-balanced sort-first configuration using all GPUs
+  as sources.
+\item [Simple] A no-scalability configuration using only the display GPU for
+  rendering.
+\item [Static DB] A static sort-last configuration distributing the range evenly
+  across all GPUs.
+\item [Dynamic DB] A dynamically load-balanced sort-last configuration using all
+  GPUs as sources.
+\end{description}
+
+The ZeroConf hostname is used to configure the node, that is, the launch command
+has to be able to resolve this hostname for starting the render client
+processes. Given a properly set up ZeroConf installation, the default launch
+command \textsf{ssh} should be able to start the render client correctly.
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
-  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
-      Load-Balancing for two Segments using eight GPU's}}
-\end{wrapfigure}
 Depending on the model position and data structure, each segment of a
@@
 
-\begin{wrapfigure}{r}{.382\textwidth} 
-    \includegraphics[width=.382\textwidth]{images/viewLB.png}
-    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
-        Load-Balancing for a CAVE}}
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
+  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
+      Load-Balancing for two Segments using eight GPU's}}
 \end{wrapfigure}
@@
 
+\begin{wrapfigure}{l}{.382\textwidth} 
+    \includegraphics[width=.382\textwidth]{images/viewLB.png}
+    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
+        Load-Balancing for a CAVE}}
+\end{wrapfigure}
 \fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
+  \vspace{-3ex}\includegraphics[width=.618\textwidth]{images/dfr.pdf}
   {\caption{\label{fDFR}Dynamic Frame Resolution}}
@@
   launch command quotes used to separate arguments on Windows. Check that the
-  same software version, including Equalizer, are installed on all machines.
+  same software versions, including Equalizer, are installed on all machines.
 \item A client can't connect back to the application. Check the client log, this

### 2012-01-23 Developer/ProgrammingGuide/paper.tex
Partial update for 1.2 Programming Guide

@@
 \lstset{language=C++}
-\section*{Equalizer 1.0 Programming and User Guide}
+\section*{Equalizer 1.2 Programming and User Guide}
 \today
@@
 
-Written by Stefan Eilemann.\\ Contributions by Maxim Makhinya, Jonas B\"osch,
-Christian Marten, Sarah Amsellem, Patrick Bouchaud, Philippe Robert, Robert
-Hauck and Lucas Peetz Dulley.
+Written by Stefan Eilemann.\\ 
+Contributions by Maxim Makhinya, Jonas B\"osch, Christian Marten, Sarah
+Amsellem, Patrick Bouchaud, Philippe Robert, Robert Hauck and Lucas Peetz
+Dulley.
 
@@
 
-\textcopyright 2007-2011 \htmladdnormallink{Eyescale Software GmbH}
+\textcopyright 2007-2012 \htmladdnormallink{Eyescale Software GmbH}
 {http://www.eyescale.ch}. All rights reserved. No permission is
@@
 providing the broadest set of parallel rendering features available in
-an open source library to any OpenGL application. Many commercial and
+an open source library to any visualization application. Many commercial and
 open source applications in a variety of different markets rely on
@@
 
-\fig{fExecutionFlow} illustrates the basic principle of any parallel
-rendering application. The typical OpenGL application, for example GLUT,
-has an event loop which redraws the scene, updates data based on
-received events, and eventually redraws a new frame.
-
-A parallel rendering application uses the same basic execution model and
-extends it by separating the rendering code from the main event
-loop. The rendering code is then executed in parallel on different
-resources, depending on the configuration chosen at runtime.
+\fig{fExecutionFlow} illustrates the basic principle of any parallel rendering
+application. The typical OpenGL application, for example using GLUT, has an
+event loop which redraws the scene, updates application data based on received
+events, and eventually renders a new frame.
+
+A parallel rendering application uses the same basic execution model, extending
+it by separating the rendering code from the main event loop. The rendering code
+is then executed in parallel on different resources, depending on the
+configuration chosen at runtime.
 
@@
 
-Equalizer can be installed by downloading the
-distribution\footnote{\link{http://www.equalizergraphics.com/downloads.html}}
-and compiling the source code. After installing Equalizer, please take a
-look at the Quickstart
+Equalizer can be installed by downloading a binary installer or source
+code\footnote{\link{http://www.equalizergraphics.com/downloads.html}}. After
+installing Equalizer, please take a look at the Quickstart
 Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
-to get familiar with the capabilities of Equalizer and the
-\textsf{eqPly} example.
+to get familiar with the capabilities of Equalizer and the \textsf{eqPly}
+example.
 
 Equalizer uses CMake to generate platform-specific build files. Compiling
-Equalizer is as simple as running \textsf{make} on Linux or Mac OS X, running
-\textsf{make xcode} and building the generated XCode project on Mac OS X, or
-creating a Equalizer Visual Studio solution using \textsf{CMake} on Windows.
+Equalizer from source is as simple as running \textsf{make} on Linux or Mac OS
+X, running \textsf{make xcode} and building the generated XCode project on Mac
+OS X, or creating a Equalizer Visual Studio solution using \textsf{CMake} on
+Windows.
 
 \subsection{Equalizer Processes}
+
+The Equalizer architecture is based on a client-server model. The client
+library exposes all functionality discussed in this document to the
+programmer, and provides communication between the different Equalizer
+processes.
 
@@
 \end{wrapfigure}
-The Equalizer architecture is based on a client-server model. The client
-library exposes all functionality discussed in this document to the
-programmer, and provides communication between the different Equalizer
-processes.
-
-Collage, the underlying network library, uses a peer-to-peer communication
-structure. Equalizer uses Collage to provide optimal performance scalable
-rendering, e.g., by setting up direct communication between two nodes when
-needed for image compositing or software swap barriers.
+Collage is a cross-platform C++ library for building heterogenous, distributed
+applications. Collage provides an abstraction of different network connections,
+peer-to-peer messaging, discovery and synchronization as well as
+high-performance, object-oriented, versioned data distribution. Collage is
+designed for low-overhead multi-threaded execution which allows applications to
+easily exploit multi-core architectures. Equalizer uses Collage as the cluster
+backend, e.g., by setting up direct communication between two nodes when needed
+for image compositing or software swap barriers.
+
+\fig{fProcesses} depicts the relationship between the server, application,
+render client and adminstrative processes, which are explained below.
 
 \subsubsection{Server}
-Each Equalizer server is responsible for managing one visualization
-system, i.e., a shared memory system or graphics cluster.  It controls
-and launches the application's rendering clients.  Currently, Equalizer
-only supports one application per server, but it will provide
-concurrent and efficient multi-application support in the future.
+The Equalizer server is responsible for managing one visualization session on a
+shared memory system or graphics cluster. Based on its configuration and
+controlling input from the application, it computes the active resources,
+updates the configuration and generates tasks for all processes. Furthermore it
+controls and launches the application's rendering client processes. The
+Equalizer server is the entity in charge of the configuration, and all other
+processes receive their configuration from the server.
 
@@
 
-The application connects to an Equalizer server and receives a
-configuration.  Furthermore, the application also provides its render
-client, which will be controlled by the server. The application reacts
-on events, updates its data and controls the rendering.
+The application connects to an Equalizer server and receives a configuration.
+Furthermore, the application also provides its render client, which will be
+controlled by the server. The application and render client may use the same
+executable. The application has a main loop, which reacts on events, updates its
+data and controls the rendering.
 
@@
 
-The render client implements the rendering part of an application. Its
-execution is passive, it has no main loop and is completely driven by
-Equalizer. It executes the rendering tasks received from the server by
-calling the appropriate task methods (see \sref{sTaskMethods}) in the
-correct thread and context. The application either implements the task
-methods with application-specific code or uses the default methods
-provided by Equalizer.
+The render client implements the rendering part of an application. Its execution
+is passive, it has no main loop and is completely driven by Equalizer server. It
+executes the rendering tasks received from the server by the calling the
+appropriate task methods (see \sref{sTaskMethods}) in the correct thread and
+context. The application either implements the task methods with
+application-specific code or uses the default methods provided by Equalizer.
 
@@
 
+\subsubsection{Administration Programs}
+
+Equalizer 1.0 introduced the admin library, which can be used to modify a
+running Equalizer server. The admin library is still in development, but already
+allows limited modifications such as adding new windows and changing
+layouts. The admin library may be used to create standalone administration tools
+or from within the application code. In any case, it has an indepedent view of
+the server's configuration. Documentation for admin library is not yet part of
+this Programming and User Guide.
+
 \section{Scalable Rendering}
 
-Real-time visualization is an inherently parallel
-problem. Unfortunately, different applications have different rendering
-algorithms, which require different scalable rendering modes to address
-the bottlenecks correctly. Equalizer supports all important algorithms as
-listed below, and will continue to add new ones over time to meet
-application requirements.
+Scalable rendering is a subset of parallel rendering, where more multiple
+resources are used to update a view.
+
+Real-time visualization is an inherently parallel problem. Different
+applications have different rendering algorithms, which require different
+scalable rendering modes to address the bottlenecks correctly. Equalizer
+supports all important algorithms as listed below, and will continue to add new
+ones over time to meet application requirements.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/2D.pdf}
+  {\caption{A 2D Compound}}
+\end{wrapfigure}
 2D decomposes the rendering in screen-space, that is, each contributing
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/2D.pdf}
-  {\caption{ A 2D Compound}}
-\end{wrapfigure}
 The advantage of this mode is a low, constant IO overhead for the pixel
@@
 be combined with DB compounds to reduce the data per node, if
-possible. In most cases, a \textsf{loadBalancer} should be used to
+possible. In most cases, a \textsf{load\_equalizer} should be used to
 automatically adjust the tiling each frame, based on the current
@@
 
-DB, as shown in \fig{fDB}\footnote{3D model courtesy of AVS, USA.},
-decomposes the rendered database so that all rendering units process a
-part of the scene in parallel. This mode is also known as sort-last, and
-is very similar to the data decomposition approach used by HPC
-applications.
-
+DB, as shown in \fig{fDB}\footnote{3D model courtesy of AVS, USA.}, decomposes
+the rendered database so that all rendering units process a part of the scene in
+parallel. This mode is also known as sort-last, and is very similar to the data
+decomposition approach used by HPC applications.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/DB.pdf}
+  {\caption{\label{fDB}\small A Database Compound}}
+\end{wrapfigure}
 Volume rendering applications use an ordered alpha-based blending to
@@
 Unfortunately, the database recomposition has linear increasing IO
-requirements for the pixel transfer. Parallel recomposition algorithms,
-such as direct-send address this problem by keeping the per-node IO
+requirements for the pixel transfer. Parallel compositing algorithms,
+such as direct-send, address this problem by keeping the per-node IO
 constant (see \fig{fDirectSend}).
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/DB.pdf}
-  {\caption{\label{fDB}\small A Database Compound}}
-\end{wrapfigure}
-The application has to partition the database so that the rendering units
-render only part of the database. Some OpenGL features do not work
-correctly (anti-aliasing) or need special attention (transparency).
+The application has to partition the database so that the rendering units render
+only part of the database. Some OpenGL features do not work correctly
+(anti-aliasing) or need special attention (transparency, shadows).
 
@@
 
-Due to the frame consistency between the eye views, this modes scales
-very well. The IO requirements for pixel transfer are small and
-constant.
-
-The number of rendering resources used by stereo compounds is limited by
-the number of eye passes, typically two.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+Due to the frame consistency between the eye views, this modes scales very
+well. The IO requirements for pixel transfer are small and constant. The number
+of rendering resources used by stereo compounds is limited by the number of eye
+passes, typically two.
+
 Stereo compounds are used by all applications when rendering in stereo,
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2em}
+  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
+  {\caption{ A DPlex Compound}}
+\end{wrapfigure}
 DPlex compounds assign full, alternating frames to individual rendering
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
-  {\caption{ A DPlex Compound}}
-\end{wrapfigure}
-
 Due to the frame consistency between consecutive frames, this mode scales
@@
 to an input frame on the destination channel. The destination channel
-uses a DPlex load-balancer to smoothen the framerate.
+uses a \textsf{framerate\_equalizer} to smoothen the framerate.
 
@@
 
-As 2D compounds, pixels compounds have low, constant IO requirements for
-the pixel transfers during recomposition. They are naturally load-balanced for
-fill-limited operations, while they do not decompose the geometry processing at
-all.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+As 2D compounds, pixel compounds have low, constant IO requirements for the
+pixel transfers during recomposition. They are naturally load-balanced for
+fill-limited operations, but do not scale geometry processing at all.
+
 OpenGL functionality influenced by the raster position will not work
@@
 
-Pixel compounds work well for purely fill-limited applications.
-Techniques like frustum culling do not reduce the rendered data for the
-source rendering resources. Pixel compounds are ideal for ray-tracing,
-which is highly fill-limited and needs the full database for rendering
-anyway. Volume rendering applications are also well suited for this
-mode, and should choose it over 2D compounds.
-
-\begin{wrapfigure}{r}{.382\textwidth}
+\begin{wrapfigure}{l}{.382\textwidth}
   \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
@@
 \end{wrapfigure}
+Pixel compounds work well for purely fill-limited applications. Techniques like
+view frustum culling do not reduce the rendered data, since each resource has
+approximately the same frustum. Pixel compounds are ideal for ray-tracing, which
+is highly fill-limited and needs the full database for rendering anyway. Volume
+rendering applications are also well suited for this mode, and should choose it
+over 2D compounds.
+
 Pixel compounds in Equalizer are configured using the \textsf{pixel}
@@
 
-Applications can use subpixel compounds to accelerate depth-of-field effects and
-software anti-aliasing, with potentially a different number of idle or non-idle
-samples per pixel.
-
-As for the DB compound, the subpixel recomposition has linear increasing 
-IO requirements for the pixel transfer with the difference that only 
-color information has to be transmitted.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+Applications can use subpixel compounds to accelerate depth-of-field effects and
+software anti-aliasing, with potentially a different number of idle or non-idle
+samples per pixel.
+
+As for the DB compound, the subpixel recomposition has linear increasing 
+IO requirements for the pixel transfer, with the difference that only 
+color information has to be transmitted.
+
 Subpixel compounds are configured using the \textsf{subpixel} parameter, using
-the values \textsf{[ index size ]}. The \textsf{index} parameter corresponds to
+the values \textsf{[~index size ]}. The \textsf{index} parameter corresponds to
 the current resource index and the \textsf{size} is the total number of
@@
 
-Please note that the ZeroConf module is optional and requires a ZeroConf
-implementation when Equalizer is compiled.
+Please note that both GPU-SD and the GPU-SD ZeroConf module are optional and
+requires GPU-SD and a ZeroConf implementation when Equalizer is compiled.
 
@@
 
-\subsubsection{Usage}
+\subsubsection{\label{sGPUSDUsage}Usage}
 
@@
   GPU-SD to query all GPUs in the subnet for the given session. The default
-  session of the GPU-SD daemon is \textsf{default}. The ZeroConf hostname is
-  used to connect the nodes.
+  session of the GPU-SD daemon is \textsf{default}. The hostname provided by
+  ZeroConf is used to connect the nodes.
 \item \textsf{--eq-config filename.eqc} loads a static configuration from the
@@
 
-A configuration consists of the declaration of the rendering resources
-and the usage description for these resources. It can optionally contain
-the description of the physical layout of the projection system and
-logical layouts on the projection canvases.
+A configuration consists of the declaration of the rendering resources, the
+description of the physical layout of the projection system, logical layouts on
+the projection canvases and an optional decomposition description using the
+aforementioned resources.
 
@@
 
-The resource usage is configured using a compound tree, which is a
-hierarchical representation of the rendering decomposition and
-recomposition across the resources. It is explained in \sref{sCompounds}.
-
 Physical layouts of display systems are configured using canvases with
-segments, which represent 2D rendering areas composed of multiplex
+segments, which represent 2D rendering areas composed of multiple
 displays or projectors. Logical layouts are applied to canvases and
 define views on a canvas.
+
+Scalable resource usage is configured using a compound tree, which is a
+hierarchical representation of the rendering decomposition and
+recomposition across the resources.
 
@@
 cards (pipes) with one window each to render to the four output channels
-connected to the projectors for each of the walls. The compound
-description is only used by the server to compute the rendering
-tasks. The application is not aware of compounds, and does not need to
-concern itself with the parallel rendering logics of a configuration.
+connected to the projectors for each of the walls.
 
@@
 For each machine in your cluster, create one \textsf{node}. Create one
-\textsf{appNode} if your application process should also render, which
-is often the case. List all nodes, even if you are not planning to use
-them at first. Equalizer will only instantiate and access used nodes,
-that is, nodes which are referenced by an active compound.
+\textsf{appNode} for your application process. List all nodes, even if you are
+not planning to use them at first. Equalizer will only instantiate and access
+used nodes, that is, nodes which are referenced by an active compound.
 
@@
 
+For cluster configurations with multiple nodes, configure at least one
+connection for the server. All render clients connect back to the server, for
+which this connection is needed.
+
 The \textsf{eq::Node} class is the representation of a single computer
@@
 
-The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU).
-In the current implementation it is also one operating system
-thread. Non-threaded pipes are supported for integrating with
-thread-unsafe libraries, but have various performance caveats. They
-should only be used if using a different, synchronized rendering thread
-is not an option.
+The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU), and
+uses one independent operating system thread for rendering. Non-threaded pipes
+are supported for integrating with thread-unsafe libraries, but have various
+performance caveats. They should only be used if using a different, synchronized
+rendering thread is not an option.
 
@@
 
-\fig{fFrusta} illustrates the wall and projection frustum parameters.
-
-Canvases provide a convenient way to configure projection surfaces. A
-canvas uses layouts, which describe logical views. The usage of canvases
-and layouts is optional. Typically, each desktop window uses one canvas.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-A canvas represents one logical projection surface, e.g., a PowerWall, a
+\fig{fFrusta} illustrates the wall and projection frustum parameters.
+
+A canvas represents one physical projection surface, e.g., a PowerWall, a
 curved screen or an immersive installation. One configuration might
@@
 
+A canvas may have a swap barrier, which is used as the default swap barrier by
+all its segments to synchronize the video output.
+
+Canvases provide a convenient way to configure projection surfaces. A canvas
+uses layouts, which describe logical views. Typically, each desktop window uses
+one canvas, one segment, one layout and one view.
+
 \subsubsection{Segments}
@@
 
+For non-planar displays, configure the frustum as described in
+\sref{sCanvas}. For passive stereo installations, configure one segment per eye
+pass, where the segment for the left and right eye have the same viewport. Set
+the eyes displayed by the segment, i.e., left or right and potentially cyclop.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
-
-For non-planar displays, configure the frustum as described in
-\sref{sCanvas}. For passive stereo installations, configure one segment per eye
-pass, where the segment for the left and right eye have the same viewport. Set
-the eyes displayed by the segment, i.e., left or right and potentially cyclop.
+To synchronize the video output, configure either a canvas swap barrier or a
+swap barrier on each segment to be synchronized.
+
+When using software swap synchronization, swap-lock all segments using a swap
+barrier. All windows with a swap barrier of the same name synchronize their
+swapbuffers. Software swap synchronization uses a distributed barrier, and works
+on all hardware.
+
+When using hardware swap synchronization, use swap barriers for all segment to
+be synchronized, setting \textsf{NV\_group} and \textsf{NV\_barrier}
+appropriately. The swap barrier name is ignored in this case. All windows of the
+same group on a single node synchronize their swap buffer. All groups of the
+same barrier synchronize their swap buffer across nodes. Please note that the
+driver typically limits the number of groups and barriers to one, and that
+multiple windows per swap group are not supported by all drivers. Hardware swap
+barriers require support from the OpenGL driver, and has been tested on nVidia
+Quadro GPUs with the G-Sync option. Please refer to your OpenGL driver
+documentation for details.
 
@@
 
-An observer represents an actor looking at multiple views. It has a head
-matrix, defining its position and orientation within the world, and an
-eye separation. Typically, a configuration has one
+An observer represents an actor looking at multiple views. It has a head matrix,
+defining its position and orientation within the world, an eye separation and
+focus distance parameters. Typically, a configuration has one
 observer. Configurations with multiple observers are used if multiple,
-head-tracked users are in the same configuration session, e.g., a
-non-tracked control host with two tracked head-mounted displays.
+head-tracked users are in the same configuration session, e.g., a non-tracked
+control host with two tracked head-mounted displays.
 
@@
 
-Compound trees are used to describe how the rendering resources are
-combined to produce the desired output, and how they are aggregated to
-increase the performance.
-
-It is advised to study and understand the basic configuration files
-shipped with the Equalizer configuration, before attempting to write
-compound configurations. The command line program \textsf{configTool},
-shipped with the Equalizer distribution, creates some standard
-configurations automatically.
+Compound trees are used to describe how multiple rendering resources are
+combined to produce the desired output, especially how multiple GPUs are
+aggregated to increase the performance.
+
+It is advised to study and understand the basic configuration files shipped with
+the Equalizer configuration, before attempting to write compound
+configurations. The auto-configuration code and command line program
+\textsf{configTool}, shipped with the Equalizer distribution, creates some
+standard configurations automatically. These are typically used as templates for
+custom configuration files.
 
@@
 \item [Destination compound(s)] Define one compound for each destination
-  channel, either as a child of the empty group, or as a top-level
-  compound.
-  \begin{description}
-  \item[Standalone channels] Set the channel by using the channel name
-    from the resource section. Define a frustum as explained in
-    \sref{sCanvas}.
-  \item[View/Segment channels] Set the channel by using the canvas, segment,
-    layout and view name. The compound frustum will be calculated automatically
-    based on the segment or view frustum. Note that one segment may created
-    multiple view/segment channels, one for each view intersection of each
-    layout used on the canvas. Only the compounds belonging to the active layout
-    of a canvas are active at runtime.
-    \begin{description}
-    \item [Software swap lock] When using software swap synchronization,
-      swap-lock the destination compounds using a swap barrier. All
-      windows with a swap barrier of the same name synchronize their
-      swapbuffers.
-    \item [Hardware swap lock] When using hardware swap synchronization,
-      use swap barriers for the destination compounds, setting
-      \textsf{NV\_group} and \textsf{NV\_barrier} appropriately. The
-      swap barrier name is ignored in this case. All windows of the same
-      group on a single node synchronize their swap buffer. All groups
-      of the same barrier synchronize their swap buffer across
-      nodes. Please note that the driver typically limits the number of
-      groups and barriers to one, and that multiple windows per swap
-      group are not supported by all drivers. Please refer to your
-      OpenGL driver documentation for details.
-    \end{description}
-  \end{description}
+  channel, either as a child of the empty group, or as a top-level compound. Set
+  the destination channel by using the canvas, segment, layout and view name or
+  index. The compound frustum will be calculated automatically based on the
+  segment or view frustum. Note that one segment may created multiple
+  view/segment channels, one for each view intersection of each layout used on
+  the canvas. Only the compounds belonging to the active layout of a canvas are
+  active at runtime.
 \item[Scalability] If desired, define scalability for each of your
@@
 \subsubsection{Compound Channels}
-Each compound has a channel, which is used by the compound to execute
-the rendering tasks. One channel might be used by multiple
-compounds. Unused channels, windows, pipes and nodes are not
-instantiated during config initialization. The rendering tasks for the
-channels are computed by the server and send to the appropriate render
-client nodes at the beginning of each frame.
+Each compound has a channel, which is used by the compound to execute the
+rendering tasks. One channel might be used by multiple compounds. Compounds are
+only active if their corresponding destination channel is active, that is, if
+the parent layout of the view which created the destination channel is active on
+at least one canvas.
+
+Unused channels, windows, pipes and nodes are not instantiated during
+initialization.  Switching an active layout may cause rendering resources to be
+stopped and started. The rendering tasks for the channels are computed by the
+server and send to the appropriate render client nodes at the beginning of each
+frame.
 
@@
 
-Compounds have a frustum description to define the physical layout of
-the display environment. The frustum specification is described in
-\sref{sCanvas}.
-
-The frustum description is inherited by the children, therefore the
-frustum is typically defined on the topmost compound.
+Compounds have a frustum description to define the physical layout of the
+display environment. The frustum specification is described in
+\sref{sCanvas}. The frustum description is inherited by the children, therefore
+the frustum is defined on the topmost compound, typically by the corresponding
+segment.
 
@@
 \subsubsection{Tasks}
-Compounds execute a number of tasks: clear, draw, assemble and
-readback. By default, a leaf compound executes all tasks and a non-leaf
-compound assemble and readback. A non-leaf compound never executes the
-draw task.
+Compounds execute a number of tasks: clear, draw, assemble and readback. By
+default, a leaf compound executes all tasks and a non-leaf compound assemble and
+readback. A non-leaf compound will never execute the draw task.
 
@@
 \subsubsection{\label{sSwapBarrier}Synchronization - Swap Barriers}
-Compounds may have a software or hardware swap barrier to synchronize the
-buffer swap of multiple channels. A software swap barrier is configured
-by giving it a name. Windows with a swap barrier of the same name
-synchronize with each other before executing the swap buffer
-task. Before entering the barrier, \textsf{Window::finish} is called to
-ensure that all OpenGL commands have been executed.
-
-Note that a software swap barrier is not accurate enough for
-edge-blended projection system or immersive installations. Such
-installations typically use a hardware synchronization, e.g., nVidia
-G-Sync cards. It is however sufficient for display walls made out of
-LCD's and for some non-immersive projection systems.
-
-Swap barrier names are only valid within the compound tree, that is, a
-compound from one compound tree cannot be synchronized with a compound
-from another compound tree.
-
-A hardware swap barrier is configured by setting the \textsf{NV\_group} and
-\textsf{NV\_barrier} parameters. These parameters follow the
-\textsf{NV\_swap\_group} extension, which synchronizes all windows bound to the
-same group on a single machine, and all groups bound to the same barrier across
-systems.
+Compounds may have a software or hardware swap barrier to synchronize the buffer
+swap of multiple channels. Swap barriers are defined by the segment, which will
+set the swap barrier on all its destination channel compounds.
+
+A software swap barrier is configured by giving it a name. Windows with a swap
+barrier of the same name synchronize with each other before executing the swap
+buffer task. Before entering the barrier, \textsf{Window::finish} is called to
+ensure that all OpenGL commands have been executed.  Swap barrier names are only
+valid within the compound tree, that is, a compound from one compound tree
+cannot be synchronized with a compound from another compound tree.
+
+A hardware swap barrier uses a hardware component to synchronize the buffer swap
+of multiple windows. It guarantees that the swap happens at the same vertical
+retrace of all corresponding video outputs.  It is configured by setting the
+\textsf{NV\_group} and \textsf{NV\_barrier} parameters. These parameters follow
+the \textsf{NV\_swap\_group} extension, which synchronizes all windows bound to
+the same group on a single machine, and all groups bound to the same barrier
+across systems.
+
+Display synchronization uses different algorithms. Framelock synchronizes each
+vertical retrace of multiple graphic outputs using a hardware component. This is
+configured in the driver, independently of the application. Genlock synchronizes
+each horizontal and vertical retrace of multiple graphic outputs, but is not
+used anymore for 3D graphics. Swap lock synchronizes the front and back buffer
+swap of multiple windows, either using a software solution based on network
+communication or a hardware solution based on a physical cable. It is
+independent of, but often used in conjunction with framelock.
+
+Framelock is used to synchronize the vertical retrace in multi-display active
+stereo installations, e.g., for edge-blended projection systems or immersive
+installations. It is combined with a software or hardware swap barrier. Software
+barriers in this case can not guarantee that the buffer swap of all outputs
+always happens with the same retrace. The time window for an output to miss the
+retrace for the buffer swap is however very small, and the output will simply
+swap on the next retrace, typically in 16 milliseconds.
+
+Display walls made out of LCD's, monoscopic or passive stereo projection systems
+often only use a software swap barrier and no framelock. The display bezels make
+it very hard to notice the missing synchronization.
 
@@
 \paragraph{Load Equalizer}
-\vspace{-2ex}\begin{wrapfigure}{r}{.382\textwidth}
+While pixel, subpixel and stereo compounds are naturally load-balanced, 2D and
+DB compounds often need load-balancing for optimal rendering performance.
+
+\begin{wrapfigure}{r}{.382\textwidth}
   \includegraphics[width=.382\textwidth]{images/lb.pdf}
@@
 \end{wrapfigure}
-While pixel, subpixel and stereo compounds are naturally load-balanced, 2D and
-DB compounds often need load-balancing for optimal rendering performance. Using
-a load equalizer is transparent to the application, and can be used on any
-application for 2D, and on most applications for DB load-balancing. Some
+Using a load equalizer is transparent to the application, and can be used with
+any application for 2D, and with most applications for DB load-balancing. Some
 applications do not support dynamic updates of the database range, and therefore
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
+  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
+      Load-Balancing for two Segments using eight GPU's}}
+\end{wrapfigure}
 \fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
-  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
-      Load-Balancing for two Segments using eight GPU's}}
-\end{wrapfigure}
 The view equalizer can also use resources from another display resource,
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-3ex}\includegraphics[width=.618\textwidth]{images/dfr.pdf}
-  {\caption{\label{fDFR}Dynamic Frame Resolution}}
-\end{wrapfigure}
 Dynamic Frame Resolution (DFR) trades rendering performance for visual
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
+  {\caption{\label{fDFR}Dynamic Frame Resolution}}
+\end{wrapfigure}
 \fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
@@
 
-\subsubsection{Prelaunched Render Clients}
+\subsubsection{\label{sResident}Prelaunched Render Clients}
 Prelaunched render clients are useful if setting up auto-launching is too
@@
   API. Most Equalizer client classes inherit from base classes in this namespace
-  and have all their data access methods in these base classes. Furthermore,
-  this namespace contains a \textsf{Serializable} class which implements one
-  common use pattern of object-based data distribution using
-  \textsf{co::Object}.
+  and have all their data access methods in these base classes.
 \item[co] Collage is the network library used by Equalizer. It provides basic
@@
   \textsf{ConnectionSet}, as well as higher-level functionality such as
-  \textsf{Node}, \textsf{LocalNode} and \textsf{Object}. Please refer to
-  \sref{sNetwork} for an introduction into the network layer, and to
-  \sref{sNetObject} for distributed objects.
+  \textsf{Node}, \textsf{LocalNode}, \textsf{Object} and
+  \textsf{Serializable}. Please refer to \sref{sNetwork} for an introduction
+  into the network layer, and to \sref{sNetObject} for distributed objects.
 \item[co::base] The co::base namespace provides C++ classes to abstract
@@
 
+Examples shipped with Equalizer are implemented in their own namespaces, e.g.,
+\textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing from the
+\textsf{eq} namespace, with the occasional usage of functionality from the
+\textsf{eq::util}, \textsf{eq::fabric}, \textsf{co} and \textsf{co::base}
+namespace.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
-Examples shipped with Equalizer are implemented in their own namespaces, e.g.,
-\textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing classes from
-the \textsf{eq} namespace, with the occasional usage of functionality from the
-\textsf{eq::util}, \textsf{eq::fabric}, \textsf{co} and \textsf{co::base}
-namespace.
-
 The functionality of the Equalizer server is implemented in the
-\textsf{eq::server} namespace. The \textsf{eq::admin} namespace enables run-time
+\textsf{eq::server} namespace, which is an internal namespace not to be used by
+application developers. The \textsf{eq::admin} namespace enables run-time
 administration of Equalizer servers. These namespaces do not yet expose a stable
@@
 during rendering to avoid threading problems. This is normally the case,
-for example all modern scene graphs use read-only render traversals.
+for example all modern scene graphs use read-only render traversals, writing the
+GPU-specific information into a separate per-GPU data structure.
 
@@
 the user, but can increase rendering performance substantially since
-operations are pipelined better.
+operations can be better pipelined.
 
@@
 
-This per-node frame synchronization does not inhibit latency across
-rendering nodes. Furthermore, advanced rendering software which
-multi-buffers the dynamic parts of the database can disable the per-node
-frame synchronization, as explained in \sref{sThreadModel}. Some scene
-graphs implement multi-buffered data, and can profit from relaxing the
-frame synchronization.
+This per-node frame synchronization does not inhibit latency across rendering
+nodes. Furthermore, advanced rendering software which multi-buffers the dynamic
+parts of the database can disable the per-node frame synchronization, as
+explained in \sref{sThreadModel}. Some scene graphs implement multi-buffered
+data, and can profit from relaxing the local frame synchronization.
 
@@
 in this section. Each of these classes inherits from a base class in the
-\textsf{fabric} name\-space, which implements data distribution for the
+\textsf{eq::fabric} name\-space, which implements data distribution for the
 entity. The fabric base classes are omitted in \fig{fUml}.
@@
 The distributed data classes are helper classes based on
-\textsf{eq::fabric::Serializable} or its base class \textsf{co::Object}. They
-illustrate the typical usage of distributed objects for static as well as
-dynamic, frame-specific data. Furthermore they are used for a basic scene graph
-distribution of the model data.
+\textsf{co::Serializable} or its base class \textsf{co::Object}. They illustrate
+the typical usage of distributed objects for static as well as dynamic,
+frame-specific data. Furthermore they are used for a basic scene graph
+distribution of the model data to the render client processes.
 
@@
 
-The application's main loop starts by connecting the application to an
-Equalizer server. The command line parameter \textsf{--eq-server}
-explicitly specifies a server address. If no server was specified,
-\textsf{Client::connectServer} tries first to connect to a server on the
-local machine using the default port. If that fails, it will create a
-server running within the application process with a default one-channel
-configuration\footnote{see
-  \link{http://www.equalizergraphics.com/documents/design/standalone.html}}.
-
-Another configuration can be selected using the command-line parameter
-\textsf{--eq-config}. If this configuration contains additional nodes, the
-\textsf{server} and \textsf{appNode} sections need a globally reachable
-\textsf{connection} configured for the render clients to connect to.
+The application's main loop starts by connecting the application to an Equalizer
+server. If no server is specified, \textsf{Client::connectServer} tries first to
+connect to a server on the local machine using the default port. If that fails,
+it will create a server running within the application process using
+auto-configuration as described in \sref{sGPUSDUsage}. The command line
+parameter \textsf{--eq-config} can be used to specify a gpu-sd session or
+configuration file, and \textsf{--eq-server} to explicitly specify a server
+address.
 
@@
 The second step is to ask the server for a configuration. The
-\textsf{ConfigParams} are a placeholder for later Equalizer
-implementations to provide additional hints and information to the
-server for choosing the configuration. The configuration chosen by the
-server is created locally using
-\textsf{NodeFactory::createConfig}. Therefore it is of type
-\textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
-making the static\_cast necessary:
+\textsf{ConfigParams} are a placeholder for later Equalizer implementations to
+provide additional hints and information to the server for
+auto-configuration. The configuration chosen by the server is created locally
+using \textsf{NodeFactory::createConfig}. Therefore it is of type
+\textsf{eqPly::Config}, but the return value is \textsf{eq::Config}, making the
+static\_cast necessary:
 
@@
 do not need this per-node frame synchronization, can disable it as
-explained in \sref{sThreadModel}.
+explained in \sref{sThreadModel}:
+
+{\footnotesize\lstinputlisting[firstline=121, lastline=132]
+  {../../../Equalizer/examples/eqPly/eqPly.cpp}}
+
+\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node 2D
+compound without latency and with a latency of one frame. The asynchronous
+execution pipelines certain rendering operations and hides imbalances in the
+load distribution, resulting in an improved framerate. For example, we have
+observed a speedup of 15\% on a five-node rendering cluster when using a latency
+of one frame instead of no
+latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
+latency of one or two frames is normally not perceived by the user. The
+statistics overlay is explained in detail in \sref{sStatisticsOverlay}.
 
@@
 \end{figure}
-\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
-2D compound without latency and with a latency of one frame. The
-asynchronous execution pipelines certain rendering operations and hides
-imbalances in the load distribution, resulting in an improved
-framerate. For example, we have observed a speedup of 15\% on a
-five-node rendering cluster when using a latency of one frame instead of
-no
-latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
-latency of one or two frames is normally not perceived by the user. The
-statistics overlay is explained in detail in \sref{sStatisticsOverlay}.
+
+When playing a camera animation, \textsf{eqPly} prints the rendering performance
+once per animation loop for benchmarking purposes:
+
+{\footnotesize\lstinputlisting[firstline=132, lastline=143]
+  {../../../Equalizer/examples/eqPly/eqPly.cpp}}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=120, lastline=147]
+{\footnotesize\lstinputlisting[firstline=161, lastline=165]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=154, lastline=165]
+{\footnotesize\lstinputlisting[firstline=167, lastline=178]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=167, lastline=175]
+{\footnotesize\lstinputlisting[firstline=180, lastline=188]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 data distribution in a cluster environment. Distributed objects are created by
-subclassing from \textsf{eq::fabric::Serializable} or \textsf{co::Object}. The
+subclassing from \textsf{co::Serializable} or \textsf{co::Object}. The
 application programmer implements serialization and deserialization of the
@@
 
-{\footnotesize\lstinputlisting[firstline=60, lastline=73]
+{\footnotesize\lstinputlisting[firstline=60, lastline=75]
   {../../../Equalizer/examples/eqPly/initData.cpp}}
@@
   version by default, or to the version specified when calling
-  \textsf{Session::mapObject}.
+  \textsf{mapObject}.
 \end{itemize}
@@
   \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
-  {\caption{\label{fUMLObject}fabric::Serializable and co::Object}}
+  {\caption{\label{fUMLObject}co::Serializable and co::Object}}
 \end{wrapfigure}
 The creation of distributed, versioned objects is simplified when using
-\textsf{eq::fabric::Serializable}, which implements one common way of tracking
+\textsf{co::Serializable}, which implements one common way of tracking
 data changes in versioned objects. The concept of a dirty bit mask is used to
@@
 which leaves all flexibility to the developer. \fig{fUMLObject} shows the
-relationship between \textsf{eq::fabric::Serializable} and \textsf{co::Object}.
+relationship between \textsf{co::Serializable} and \textsf{co::Object}.
 
@@
 is changed. The serialization methods are called by the
-\textsf{eq::fabric::Serializable} with the dirty bit mask needed to serialize
+\textsf{co::Serializable} with the dirty bit mask needed to serialize
 all data, or with the dirty bit mask of the changes since the last
@@
 
-Some applications might rely on a shared filesystem to access the data,
-for example when out-of-core algorithms are used. Other applications
-prefer to load the data only on the application process, and use
-distributed objects to synchronize the scene data with the render
-clients.
-
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
-\textsf{eqPly} uses static distributed objects to distribute the model
-loaded by the application. This approach can be easily extended to
+
+Some applications rely on a shared filesystem to access the data, for example
+when out-of-core algorithms are used. Other applications prefer to load the data
+only on the application process, and use distributed objects to synchronize the
+scene data with the render clients.
+
+\textsf{eqPly} chooses the second approach, using static distributed objects to
+distribute the model loaded by the application. It can be easily extended to
 versioned objects to support dynamic data modifications.
@@
 The kD-tree data structure and rendering code for the model is strongly
-separated from Equalizer, and kept in the separate namespace
-\textsf{mesh}. It can also be used in other rendering software, for
-example in a GLUT application. To keep this separation, an external
-'mirror' hierarchy is constructed around the tree. This hierarchy of
-\textsf{VertexBufferDist} nodes is responsible for cloning the model
-data on the remote render clients.
+separated from Equalizer, and kept in the separate namespace \textsf{mesh}. It
+can also be used in other rendering software, for example in a GLUT
+application. To keep this separation while implementing data distribution, an
+external 'mirror' hierarchy is constructed aside the kD-tree. This hierarchy of
+\textsf{VertexBufferDist} nodes is responsible for cloning the model data on the
+remote render clients.
 
@@
 render clients to map the model when it is needed for
-rendering. \fig{fModelDist} shows the UML hierarchy of the model and
-distribution classes.
+rendering.
+
+\fig{fModelDist} shows the UML hierarchy of the model and distribution
+classes. \sref{sNetObject2} illustrates other approaches to employ distributed
+objects for data distribution and synchronization.
 
@@
 
-The configuration is driving the application's rendering, that is, it is
-responsible for updating the data based on received events, requesting
-new frames to be rendered and to provide the render clients with the
-necessary data.
+The \textsf{eq::Config} class is driving the application's rendering, that is,
+it is responsible for updating the data based on received events, requesting new
+frames to be rendered and to provide the render clients with the necessary data.
 
@@
 initialization methods are always executed before any child initialization
-method. This parallelization is necessary to allow a speedy startup of the
-configuration on large-scale graphics clusters. On the other hand, it means that
-initialization functions are called even if the parent's initialization has
-failed. \fig{fConfigInit} shows a sequence diagram of the config initialization.
+method. This parallelization allows a speedy startup on large-scale graphics
+clusters. On the other hand, it means that initialization functions are called
+even if the parent's initialization has failed. \fig{fConfigInit} shows a
+sequence diagram of the config initialization.
+
 \begin{figure}[ht!]\center
@@
 The \textsf{eqPly::Config} class holds the master versions of the initialization
-and frame data. Both objects are registered with the \textsf{eq::Config}. The
-configuration forwards the registration to the local client node, and retains
-buffered, deregistered object data for $latency$ frames.
+and frame data objects. Both are registered with the \textsf{eq::Config}. The
+configuration forwards the registration to the local client node and augments
+the object registration for buffered objects.
+
+First, it configures the objects to retain its data \textsf{latency+1} commits,
+which corresponds to the typical use case where objects are commited once per
+frame. This allows render clients, which often are behind the application
+process, to map objects with an old version. This does not necessarily translate
+into increased memory usage, since new versions are only created when the object
+was dirty during commit.
+
+Second, it retains the data for buffered objects data for \textsf{latency}
+frames after their deregistration. This allows to map the object on a render
+client, even after it has been deregistered on the application node. This does
+delay the deallocation of the buffered object data by \textsf{latency} frames.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=74, lastline=95]
+{\footnotesize\lstinputlisting[firstline=76, lastline=93]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=97, lastline=99]
+{\footnotesize\lstinputlisting[firstline=94, lastline=96]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 orthogonal problem to parallel rendering. Tracking device support has already
-been solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
-which are easily integrated with Equalizer. The example code in \textsf{eqPly}
-provides a reference implementation for the integration of such a tracking
-library. \sref{sTracking} provides more background on head tracking.
-
-{\footnotesize\lstinputlisting[firstline=100, lastline=118]
+been solved by a number of implementations, such as VRCO Trackd, VRPN and
+others, which have proven to be easily integrated with Equalizer. The example
+code in \textsf{eqPly} provides a reference implementation for the integration
+of such a tracking library. \sref{sTracking} provides more background on head
+tracking.
+
+{\footnotesize\lstinputlisting[firstline=98, lastline=115]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=130, lastline=139]
+{\footnotesize\lstinputlisting[firstline=128, lastline=135]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=310, lastline=317]
+{\footnotesize\lstinputlisting[firstline=306, lastline=313]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=319, lastline=327]
+{\footnotesize\lstinputlisting[firstline=315, lastline=323]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 anti-aliasing to gradually reduce aliasing effects in the rendering. The idle
-state is tracked by the application and used by the rendering callbacks:
-
-{\footnotesize\lstinputlisting[firstline=329, lastline=359]
+state is tracked by the application and used by the rendering callbacks to
+jitter the frusta, accumulate and display the results:
+
+{\footnotesize\lstinputlisting[firstline=325, lastline=355]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-The preprocess config instance maintains the mapped models, and lazily maps new
-models, which are registered by the application process:
-
-{\footnotesize\lstinputlisting[firstline=284, lastline=308]
+The per-process config instance maintains the mapped models, and lazily maps new
+models, which are registered by the application process. Since the model loading
+may be called concurrently from different pipe render threads, it is protected
+by a \textsf{mutex}:
+
+{\footnotesize\lstinputlisting[firstline=280, lastline=304]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-View-specific data is not limited to a model. Applications can choose to
-make any application-specific data view-specific, e.g., cameras,
-rendering modes or annotations. A view is a generic concept for an
-application-specific view on data, \textsf{eqPly} is using different
-models to illustrate the concept:
-
-{\footnotesize\lstinputlisting[firstline=790, lastline=859]
+View-specific data is not limited to a model. Applications can choose to make
+any application-specific data view-specific, e.g., cameras, rendering modes or
+annotations. A view is a generic concept for an application-specific view on
+data, \textsf{eqPly} is simply using different models to illustrate the concept:
+
+{\footnotesize\lstinputlisting[firstline=805, lastline=860]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=35, lastline=48]
+{\footnotesize\lstinputlisting[firstline=36, lastline=52]
   {../../../Equalizer/examples/eqPly/node.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=257, lastline=268]
+{\footnotesize\lstinputlisting[firstline=249, lastline=265B]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 All task methods for a pipe and its children are executed in a separate
-thread. This approach optimizes usage of the GPU, since all
-tasks are executed serially and therefore do not compete for resources
-or cause OpenGL context switches. Later versions of Equalizer might
-introduce threaded windows to allow the parallel and independent
-execution of rendering tasks on a single pipe.
+thread. This approach optimizes GPU usage, since all tasks are executed serially
+and therefore do not compete for resources or cause OpenGL context
+switches. Multiple GPU run in parallel with each other.
 
@@
 frames at any given time. Therefore frame-specific data has to be allocated for
-each pipe thread, which applies only to the frame data.
+each pipe thread, which affect only to the frame data.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=57, lastline=70]
+{\footnotesize\lstinputlisting[firstline=45, lastline=55]
   {../../../Equalizer/examples/eqPly/pipe.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=72, lastline=78]
+{\footnotesize\lstinputlisting[firstline=57, lastline=63]
   {../../../Equalizer/examples/eqPly/pipe.cpp}}
@@
 
-Equalizer supports multiple window system interfaces, at the moment
-glX/X11, WGL and AGL/Carbon. Some operating systems, and therefore some
-Equalizer versions, support multiple window systems
-concurrently\footnote{see
-  \link{http://www.equalizergraphics.com/compatibility.html}}.
-
-Each pipe might use a different window system for rendering, which is
-determined before \textsf{Pipe::configInit} by
-\textsf{Pipe::selectWindowSystem}. The default implementation of
-\textsf{selectWindowSystem} loops over all window systems and returns
-the first supported window system, determined by using
-\textsf{supportsWindowSystem}.
+Equalizer supports multiple window system interfaces, at the moment glX/X11, WGL
+and AGL/Carbon. Some operating systems, and therefore some Equalizer versions,
+support multiple window systems concurrently.
+
+Each pipe might use a different window system for rendering, which is determined
+before \textsf{Pipe::configInit} by \textsf{Pipe::selectWindowSystem}. The
+default implementation of \textsf{selectWindowSystem} uses the first supported
+window system.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=39, lastline=55]
+{\footnotesize\lstinputlisting[firstline=39, lastline=43]
   {../../../Equalizer/examples/eqPly/pipe.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=80, lastline=84]
+{\footnotesize\lstinputlisting[firstline=65, lastline=69]
   {../../../Equalizer/examples/eqPly/pipe.cpp}}
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/osWindow.pdf}
+  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
   {\caption{\label{fOSWindow}SystemWindow UML Class Hierarchy}}
@@
 
-{\footnotesize\lstinputlisting[firstline=76, lastline=98]
+{\footnotesize\lstinputlisting[firstline=77, lastline=99]
   {../../../Equalizer/examples/eqPly/window.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=100, lastline=109]
+{\footnotesize\lstinputlisting[firstline=101, lastline=110]
   {../../../Equalizer/examples/eqPly/window.cpp}}
@@
 
-The object manager (OM) is a utility class and can be used to manage
-OpenGL objects across shared contexts. Typically one OM is used for each
-set of shared contexts and spawns all contexts of a single
+The object manager (OM) is a utility class and can be used to manage OpenGL
+objects across shared contexts. Typically one OM is used for each set of shared
+contexts of a single
 GPU\footnote{\link{http://www.equalizergraphics.com/documents/design/objectManager.html}}.
 
-The OM is a template class. The template type is the key used to
-identify objects. The same key is used by all contexts to get the
-OpenGL name of an object.
-
-Each \textsf{eq::Window} has an object manager with the key type
-\textsf{const void*}, for as long as it is initialized. Each window can
-have a shared context window. The OM is shared with this shared context
-window. The shared context window is set by default to the first window
-of each pipe, and therefore the OM will be shared between all windows of
-a pipe. The method \textsf{eq::Window::setSharedContextWindow} can be
-used to set up a different context sharing.
+Each \textsf{eq::Window} has an object manager with the key type \textsf{const
+  void*}, for as long as it is initialized. Each window can have a shared
+context window. The OM is shared with this shared context window. The shared
+context window is set by default to the first window of each pipe, and therefore
+the OM will be shared between all windows of a pipe. The same key is used by all
+contexts to get the OpenGL name of an object, thus reusing of the same object
+within the same share group.  The method
+\textsf{eq::Window::setSharedContextWindow} can be used to set up a different
+context sharing.
 
@@
 The channel is the heart of the application's rendering code, it executes all
-task methods needed to produce the configured rendering. It is used to perform
-the various rendering operations for the compounds. Each channel has a set of
-task methods to execute the clear, draw, readback and assemble stages needed to
-render a frame.
+task methods needed to update the configured views. It performs the various
+rendering operations for the compounds. Each channel has a set of task methods
+to execute the clear, draw, readback and assemble stages needed to render a
+frame.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=69, lastline=78]
+{\footnotesize\lstinputlisting[firstline=68, lastline=77]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-The central rendering routine is \textsf{Channel::frameDraw}. This
-routine contains the application's OpenGL rendering code, which is being
-rendered using the contextual information provided by Equalizer. As most
-of the other task methods, \textsf{frameDraw} is called in parallel by
-Equalizer on all pipe threads in the configuration. Therefore the
-rendering must not write to shared data, which is the case for all major
-scene graph implementations.
+The central rendering routine is \textsf{Channel::frameDraw}. This routine
+contains the application's OpenGL rendering code, which uses the contextual
+information provided by Equalizer. As most of the other task methods,
+\textsf{frameDraw} is called in parallel by Equalizer on all pipe threads in the
+configuration. Therefore the rendering must not write to shared data, which is
+the case for all major scene graph implementations.
 
 In \textsf{eqPly}, the OpenGL context is first set up using various
-\textsf{apply} convenience methods from the base Equalizer channel
-class. Each of the \textsf{apply} methods uses the corresponding
-\textsf{get} method(s) and then calls the appropriate OpenGL
-function(s). It is also possible to just query the values from Equalizer
-using the \textsf{get} methods, and use them to set up the OpenGL state
-appropriately, for example by passing the parameters to the renderer used
-by the application.
+\textsf{apply} convenience methods from the base Equalizer channel class. Each
+of the \textsf{apply} methods uses the corresponding \textsf{get} methods and
+then calls the appropriate OpenGL functions. It is also possible to just query
+the values from Equalizer using the \textsf{get} methods, and use them to set up
+the OpenGL state appropriately, for example by passing the parameters to the
+renderer used by the application.
 
@@
   defined.
-\item[Range] A one-dimensional range with the interval [0..1]. This
-  parameter is optional and should be used by the application to render
-  only the appropriate subset of its data. It is influenced by the
-  compound range attribute.
+\item[Range] A one-dimensional range with the interval [0..1]. This parameter is
+  optional and should be used by the application to render only the appropriate
+  subset of its data for sort-last rendering. It is influenced by the compound
+  range attribute.
 \end{description}
@@
 
-{\footnotesize\lstinputlisting[firstline=123, lastline=137]
+{\footnotesize\lstinputlisting[firstline=122, lastline=137]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/DB.png}
-  {\caption{\label{fDBDest}Destination View of a DB Compound using
-      Demonstrative Coloring}}
-\end{wrapfigure}
 After the basic view setup, a directional light is configured, and the
@@
 
-{\footnotesize\lstinputlisting[firstline=139, lastline=165]
+{\footnotesize\lstinputlisting[firstline=138, lastline=164]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=167, lastline=178]
+{\footnotesize\lstinputlisting[firstline=166, lastline=177]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
 
-To draw the model, a helper class for view frustum culling is set up
-using the view frustum from Equalizer and the camera position from the
-frame data. The frustum helper computes the six frustum planes from the
-projection and modelView matrices. During rendering, the bounding
-spheres of the model are tested against these planes to determine the
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/DB.png}
+  {\caption{\label{fDBDest}Destination View of a DB Compound using
+      Demonstrative Coloring}}
+\end{wrapfigure}
+To draw the model, a helper class for view frustum culling is set up, using the
+view frustum from Equalizer (projection and view matrix) and the camera position
+(model matrix) from the frame data. The frustum helper computes the six frustum
+planes from the projection and modelView matrices. During rendering, the
+bounding spheres of the model are tested against these planes to determine the
 visibility with the frustum.
@@
 
-{\footnotesize\lstinputlisting[firstline=591, lastline=610]
+{\footnotesize\lstinputlisting[firstline=588, lastline=625]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
 
+\begin{wrapfigure}{r}{.618\textwidth}
+%\begin{figure}[h!t]\center
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\label{fRender}Main Render Loop}}
+\end{wrapfigure}
 The model data is spatially organized in a 3-dimensional
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-%\begin{figure}[h!t]\center
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\label{fRender}Main Render Loop}}
-\end{wrapfigure}
 The rendering loop maintains a list of candidates to render, which
@@
 range. If it is partially visible or partially in range, the children of
-the node are added to the candidate list. \fig{fRender} shows a flow
-chart of the rendering algorithm, which performs efficient view frustum
-and range culling.
+the node are added to the candidate list.
+
+\fig{fRender} shows a flow chart of the rendering algorithm, which performs
+efficient view frustum and range culling.
 
@@
 
-The main rendering loop in \textsf{eqPly} looks like this:
-
-{\footnotesize\lstinputlisting[firstline=619, lastline=689]
+The main rendering loop is implemented in \textsf{VertexBufferRoot::cullDraw()},
+and not duplicated here.
+
+\subsubsection{Assembly}
+
+Like most applications, eqPly uses most of the default implementation of the
+\textsf{frameReadback} and \textsf{frameAssemble} task methods. To implement an
+optimization and various customizations, \textsf{frameReadback} is
+overwritten. eqPly does not need the alpha channel on the destination view. The
+output frames are flagged to ignore alpha, which allows the compressor to drop
+25\% of the data during image transfer. Furthermore, compression can be
+disabled and the compression quality can be changed at runtime to demonstrate
+the impact of compression on scalable rendering:
+
+{\footnotesize\lstinputlisting[firstline=242, lastline=267]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
 
-\subsubsection{Assembly}
-
-Like most applications, eqPly can use the default implementation of the
-\textsf{frameReadback} and \textsf{frameAssemble} task methods.
-\textsf{frameReadback} is overwritten for an optimization since eqPly
-does not need the alpha channel on the destination view.
-The output frames are flagged to ignore it, which allows the compressor to
-drop 25\% of the data during image transfer:
-
-{\footnotesize\lstinputlisting[firstline=238, lastline=261]
-  {../../../Equalizer/examples/eqPly/channel.cpp}}
-
 The \textsf{frameAssemble} method is overwritten for the use of Subpixel
-compound with idle software anti-aliasing, as described in \ref{sFSAA}.
+compound with idle software anti-aliasing, as described in \sref{sFSAA}.
 
@@
 \textsf{eqPly} section. Where possible, code examples from the Equalizer
-distribution are used to illustrate usage of the feature. Its
+distribution are used to illustrate usage of the specific feature. Its
 purpose is to provide information on how to address a typical problem or
@@
 substantially differs from GLUT, which imposes an event model and hides
-most of the event handling in \textsf{glutMainLoop}. More information on
-event handling can be found on the Equalizer website\footnote{see
-  \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.
+most of the event handling in \textsf{glutMainLoop}.
 
@@
 to. For AGL, Equalizer internally forwards the event from the main
-thread, where it is received, to the appropriate pipe thread.
+thread, where it is received, to the appropriate pipe thread. This model allows
+window and channel modifications which are naturally thread-safe, since they are
+executed from the corresponding render thread.
 
@@
 Events are received by an event handler. The event handler finds the
-\textsf{eq::SystemWindow} associated to the event. It then creates a generic
-\textsf{Event}, which holds the event data in an independent
-format. The original native event and this generic \textsf{Event}
-form the \textsf{SystemWindowEvent}, which is passed to the concrete
+\textsf{eq::Sys\-tem\-Win\-dow} associated to the event. It then creates a
+generic \textsf{Event}, which holds the event data in an independent format. The
+original native event and this generic \textsf{Event} form the
+\textsf{SystemWindowEvent}, which is passed to the concrete
 \textsf{SystemWindow} for processing.
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 The purpose of the \textsf{SystemWindow} processing method,
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\label{fEventProcessing}Event Processing}}
+\end{wrapfigure}
 \textsf{Window::processEvent} is responsible for handling
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
-\end{wrapfigure}
-For each window, one \textsf{GLXEventHandler} is allocated. FBO's and PBuffers
-are handled in the same way as window drawables. The event dispatch finds the
-corresponding event handler for each received event, which is then used to
-process the event in a fashion similar to AGL. The \textsf{GLXWindowEvent}
-passes the \textsf{XEvent} to \textsf{GLXWindowIF::\-process\-Event}.
+For each GLX window, one \textsf{GLXEventHandler} is allocated. FBO's and
+PBuffers are handled in the same way as window drawables. The event dispatch
+finds the corresponding event handler for each received event, which is then
+used to process the event in a fashion similar to AGL. The
+\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
+\textsf{GLXWindowIF::\-process\-Event}.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=37, lastline=73]
+{\footnotesize\lstinputlisting[firstline=37, lastline=75]
   {../../../Equalizer/examples/eVolve/error.cpp}}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/threads.pdf}
+  {\caption{\label{fThreads}Threads within one Node Process}}
+\end{wrapfigure}
 The receiver and command thread are terminated when the application
@@
 
-During config initialization, one pipe thread is created for each
-pipe. The pipe threads execute all render task methods for this pipe,
-and therefore executes the application's rendering code. The pipe
-threads are terminated during \textsf{Config::exit}.
-
-The rest of this section discusses the thread synchronization between
-the main thread and the pipe threads.
+During config initialization, one pipe thread is created for each pipe. The pipe
+threads execute all render task methods for this pipe, and therefore executes
+the application's rendering code. All pipe threads are terminated during
+\textsf{Config::exit}.
+
+Since a layout switch on a canvas may involve different resources, pipe threads
+may be started and stopped dynamically at runtime. Before such an update,
+Equalizer will always finish all pending rendering operations to ensure that all
+resources are idle. The rest of this section discusses the thread
+synchronization between the main thread and the pipe threads.
 
@@
   data.
-\item[DRAW\_SYNC:] Additionally to the synchronization of the async
-  thread model, all local render threads are synchronized, so that the
-  draw operations happen synchronously with the node main loop. This
-  model allows to use the same database for rendering, and safe
-  modifications of this database are possible from the node thread,
-  since the pipe threads do not execute any rendering tasks between
-  frames. This is the default threading model. This threading model
-  should be used by applications which keep one copy of the scene graph
-  per node.
+\item[DRAW\_SYNC:] Additionally to the synchronization of the async thread
+  model, all local render threads are synchronized, so that the draw operations
+  happen synchronously with the node main loop. Compositing, swap barriers and
+  buffer swaps happen asynchronously. This model allows to use the same database
+  for rendering, and safe modifications of this database are possible from the
+  node thread, since the pipe threads do not execute any rendering tasks between
+  frames. This is the default threading model. This threading model should be
+  used by applications which keep one copy of the scene graph per node.
 \item[LOCAL\_SYNC:] Additionally to the synchronization of the async
@@
 
-Each \textsf{eq::Window} has a \textsf{GLEWContext}. This context can be
-obtained by using \textsf{glewGetContext} on the window or channel. GLEW\_MX
-uses this function to dispatch the functions to the correct context. Equalizer
-(re-)initializes the GLEW context whenever a new OpenGL context is set on the
-window.
-
-Extended OpenGL functions called from a window or channel instance can
-be called directly. GLEW will call the object's \textsf{glewGetContext}
-to obtain the correct context:
+GLEW uses a function table called \textsf{GLEWContext} to store function
+pointers for non-standard OpenGL functions. This table is initialized once per
+OpenGL context by retrieving the pointers through a system-specific
+method. OpenGL functions are then redefined using a macro which retrieves the
+\textsf{GLEWContext} using \textsf{glewGetContext}, looking up the corresponding
+function pointer in this table and calling the function. Equalizer applications
+can disable these GLEW macros by defining \textsf{EQ\_IGNORE\_GLEW} before
+including \textsf{eq/eq.h}.
+
+Each \textsf{eq::GLWindow}, which is the \textsf{SystemWindow} base class for
+all OpenGL system windows, has a \textsf{GLEWContext}. This context can be
+obtained by using \textsf{glewGetContext} on the \textsf{eq::Window} or
+\textsf{eq::Channel}. Equalizer (re-)initializes the GLEW context whenever a new
+OpenGL context is set.
+
+Extended OpenGL functions called from a GLWindow, window or channel instance can
+be called directly. GLEW will call the object's \textsf{glewGetContext} to
+obtain the correct context:
 
@@
 
-An application typically choses to subclass the sample implementation if
-only minor tweaks are needed for integration. For major changes or new
-window systems, it is often easier to subclass directly from
-\textsf{SystemWindow} and implement the abstract methods of this interface
-class.
+An application typically choses to subclass the sample implementation if only
+minor tweaks are needed for integration. For major changes or new window
+systems, it is often easier to subclass directly from \textsf{SystemWindow} or
+\textsf{GLWindow} and implement the abstract methods of this interface class.
 
@@
 
-Since window initialization is notoriously error-prone and hard to debug, the
-sample implementation propagates the reason for errors from the render clients
-back to the application. The \textsf{Pipe} and \textsf{Window} classes have a
-\textsf{setError} method, which is used to set an error code. This string is
-passed to the \textsf{Config} instance on the application node, where it can be
-retrieved using \textsf{getError}. The \sref{sErrorHandling} explains error
-handling in more detail.
-
-The sample implementations \textsf{AGLWindow}, \textsf{GLXWindow} and
-\textsf{WGLWindow} all have similar, override-able methods for all
+Since window initialization is notoriously error-prone and hard to debug in an
+distributed application, the sample implementation propagates the reason for
+errors from the render clients back to the application. The \textsf{Pipe} and
+\textsf{Window} classes have a \textsf{setError} method, which is used to set an
+error code. This string is passed to the \textsf{Config} instance on the
+application node, where it can be retrieved using \textsf{getError}. The
+\sref{sErrorHandling} explains error handling in more detail.
+
+The sample implementations \textsf{agl::Window}, \textsf{glx::Window} and
+\textsf{wgl::Window} all have similar, override-able methods for all
 sub-tasks. This allows partial customization, without the need of
@@
 open source and commercial implementations cover this functionality well
-and can easily be integrated with Equalizer.
+and are easily integrated with Equalizer.
 
 An Equalizer configuration contains a number of observers. Each observer
-represents on tracked entity. Most immersive configurations only have one
+represents one tracked entity. Most immersive configurations only have one
 observer. It is possible to support multiple tracked viewers in the same
@@
 
-Note that while stereo rendering is often implemented the simpler toe-in method,
-Equalizer uses the correct approach using asymmetric frusta.
+Note that while stereo rendering is often implemented using the simpler toe-in
+method, Equalizer implements the correct approach using asymmetric frusta.
 
@@
 
-For head-mounted displays (HMD), the tracking information is used to
-move the frusta with the observer, as shown in \fig{fHMD}. This results
-in different projections compared to normal tracking with fixed
-projection screens. These difference is transparent to Equalizer applications,
-only the configuration file uses a different wall type for HMD's.
+For head-mounted displays (HMD), the tracking information is used to move the
+frusta with the observer, as shown in \fig{fHMD}. This results in different
+projections compared to normal tracking with fixed projection screens. This
+difference is transparent to Equalizer applications, only the configuration file
+uses a different wall type for HMD's.
 
@@
 \begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/focusFixed}
-  {\caption{\label{fFixed}Fixed focus distance}}
+  \subfigure[]{\label{fFixed}
+    \includegraphics[width=.382\textwidth]{images/focusFixed}}\hfil
+  \subfigure[]{\label{fOrigin}
+    \includegraphics[width=.382\textwidth]{images/focusOrigin}}\hfil
+  \subfigure[]{\label{fObserver}
+    \includegraphics[width=.382\textwidth]{images/focusObserver}}
+  {\caption{Fixed(a) and dynamic focus distance relative to origin(b) and observer(c)}}
 \end{wrapfigure}
+
 The frustum calculation up to Equalizer version 1.0 places the stereo
@@
 distance is therefore the distance between the origin and the middle of the
-projection surface.
-
-In Equalizer 1.1 and later the focus distance and mode can be configured and
+projection surface. This is how almost all Virtual Reality software handles the
+focal plane.
+
+In Equalizer 1.2 and later the focus distance and mode can be configured and
 changed at runtime for each \textsf{observer} and on the \textsf{config} for
-untracked views.
+untracked views. This allows applications to expose the focal plane to the user,
+or to automatically calculate it based on the scene and view direction.
 
@@
 distance to dynamically change the stereo convergence.
-
-\begin{figure}[h!t]\center
-  \subfigure[]{\label{fOrigin}
-    \includegraphics[width=.49\textwidth]{images/focusOrigin}}\hfil
-  \subfigure[]{\label{fObserver}
-    \includegraphics[width=.49\textwidth]{images/focusObserver}}
-  {\caption{\label{fImmersive}Dynamic focus distance relative to origin
-      (a) and observer (b)}}
-\end{figure}
 
@@
 new logical 'walls' used for frustum calculations in white, while the physical
-projection from \fig{fFixed} is still visible in the background.
+projection from \fig{fFixed} is still visible.
 
@@
 
-The application has read access to all canvases, segment, layouts and
-views of the configuration. The render client has access to the current
-view in the channel task methods. The layout entities can be sub-classed
-using the \textsf{NodeFactory}. Currently the layout of a canvas and the
-frustum of a view can be changed at runtime.
+The application has read access to all canvases, segment, layouts and views of
+the configuration. The render client has access to the current view in the
+channel task methods. The layout entities can be sub-classed using the
+\textsf{NodeFactory}. Currently the layout of a canvas, the frustum and stereo
+mode of a view as well as the view's user data object can be changed at runtime.
 
@@
 
+\begin{wrapfigure}{r}{.612\textwidth}
+  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
+  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
+\end{wrapfigure}
 Layout API entities (Canvas, Segment, Layout, View) are sub-classed like all
@@
 
-\begin{wrapfigure}{r}{.612\textwidth}
-  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
-  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
-\end{wrapfigure}
 Equalizer commits dirty layout entities at the beginning of each
@@
 The most common entity to subclass is the \textsf{View}, since the application
-often amends with view-specific application data. A view might have a
+often amends it with view-specific application data. A view might have a
 distributable user data object, which has to inherit from \textsf{co::Object} to
-be committed and synchronized with the associated object.
+be committed and synchronized with the associated view.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=59, lastline=64, xleftmargin=-7em]
+{\footnotesize\lstinputlisting[firstline=61, lastline=66, xleftmargin=-7em]
   {../../../Equalizer/examples/eqPly/view.h}}
@@
 de-initialized, that is \textsf{configExit} and
-\textsf{NodeFactory::releaseEntity} is called, and channels involved in
+\textsf{NodeFactory::releaseChannel} is called, and channels involved in
 the new layout are initialized. If a layout does not cover fully a
@@
 eye or the left and right eye.
+
+The configuration contains stereo information in three places: the view, the
+segment and the compound.
+
+The view defines which stereo mode is active: mono or stereo. This setting can
+be done in the configuration file or programmatically.
+
+The segment has an \textsf{eye} attribute in which defines which eyes are
+displayed by this segment. This is typically all eyes for active stereo
+projections and the left or right eye for passive stereo. The cyclop eye for
+passive stereo can either be set on a single segment or both, depending if the
+second projector is active during monoscopic rendering. The default setting
+enables all eyes for a segment.
 
@@
 \end{wrapfigure}
-The configuration contains stereo information in three places: the view, the
-segment and the compound.
-
-The view defines which stereo mode is active: mono or stereo. This setting can
-be done in the configuration file or programmatically.
-
-The segment has the \textsf{eye} attribute in the configuration file which
-defines which eyes are displayed by this segment. This is typically all eyes for
-active stereo projections and the left or right eye for passive stereo. The
-cyclop eye for passive stereo can either be set on a single segment or both,
-depending if the second projector is active during monoscopic rendering. The
-default setting enables all eyes for a segment.
-
 The compound has an \textsf{eye} attribute defining which eyes it is capable of
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
-  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
-\end{wrapfigure}
 A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
+  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
+\end{wrapfigure}
 When a window is resized, a \textsf{CHANNEL\_RESIZE} event is eventually
@@
 of parallel rendering, when multiple channels are contributing to a single
-view.
-
-The source channels producing one or more \textsf{outputFrame}s use
+view. This requires reading back the pixel data from the source GPU and
+assembling it on the destination GPU.
+
+Channels producing one or more \textsf{outputFrame}s use
 \textsf{Channel::frame\-Read\-back} to read the pixel data from the frame
@@
 \textsf{Channel::frameAssemb\-le} to assemble the pixel data into the
-framebuffer. Equalizer takes care of the network transport of frame
-buffer data between nodes.
+framebuffer. Equalizer takes care of the network transport of frame buffer data
+between nodes.
 
@@
 \textsf{eq::init} for dynamic shared objects (DSO) containing compression
-libraries (EqualizerCompressor*.dll on Windows, libeqCompressor*.dylib
-on Mac OS X, libeqCompressor*.so on Linux).
+libraries (EqualizerCompressor*.dll on Windows, libEqualizerCompressor*.dylib
+on Mac OS X, libEqualizerCompressor*.so on Linux).
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\label{fDirectSend}Direct Send Compositing}}
+\end{wrapfigure}
 To provide a motivation for the design of the image compositing
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\label{fDirectSend}Direct Send Compositing}}
-\end{wrapfigure}
 After performing the sort-last compositing, the color information is
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
+  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
+\end{wrapfigure}
 The frame data is a holder for images and additional information, such
@@
 
-An \textsf{eq::Image} holds a
-two-dimensional snapshot of the framebuffer and can contain color and/or
-depth information.
+An \textsf{eq::Image} holds a two-dimensional snapshot of the framebuffer and
+can contain color and/or depth information.
 
@@
 
-%\begin{wrapfigure}{r}{.618\textwidth}
-\begin{figure}[h!t]\center
+\begin{wrapfigure}{r}{.618\textwidth}
+%\begin{figure}[h!t]\center
   \includegraphics[width=.618\textwidth]{images/compositor.pdf}
   {\caption{\label{fCompositor}Functional Diagram of the Compositor}}
-\end{figure}
+\end{wrapfigure}
 On a high level, the compositor combines multiple input frames using 2D
@@
 
+\begin{wrapfigure}{l}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended to form a
@@
 
-\subsection{\label{sFSAA}Custom Subpixel Processing}
-
-The eqPly example contains support for software anti-aliasing when being idle,
-showing how to combine it with subpixel compounds. The same basic algorithm
-might be applied to other multisampling effects, e.g., depth-of-field.
+\subsection{\label{sFSAA}Subpixel Processing}
+
+In eqPly, the image quality is gradually refined when the camera is not
+moving. Up to 256 samples per pixel are accumulated to implement idle
+antialiasing. This feature is compatible with subpixel compounds, which use a
+fixed non-idle anti-aliasing and then accelerate the generation of the remaining
+samples. The same basic algorithm might be applied to other multisampling
+effects, e.g., depth-of-field.
+
+\subsubsection{Transparent Software Anti-Aliasing}
+
+Any Equalizer applications can benefit without modification from subpixel
+compounds. Applications performing their own, configurable idle or non-idle
+anti-aliasing can easily be integrated with subpixel compounds, as described in
+the next sections.
+
+The default implementation of \textsf{applyFrustum} and \textsf{applyOrtho} in
+\textsf{eq::Channel} apply jittering when a subpixel parameter is set.
+Furthermore, the \textsf{Compositor} uses an \textsf{eq::Accum} buffer to
+accumulate and display the results of a subpixel decomposition in \textsf{assembleFramesUnsorted} and \textsf{assembleFramesSorted}.
 
@@
 technique which computes samples within a pixel by jittering the frustum by a
-subpixel amount. The pixel is split into several, regular sub-pixels and a
-sample is taken randomly within the sub-pixel.
+subpixel amount. The pixel is split into an 16x16 subpixel grid, and a sample is
+taken randomly within each sub-pixel. The samples are accumulated and displayed
+after each step.
+
+To accumulate the result, each \textsf{eqPly::Channel} maintains an accumulation
+buffer for each rendered eye pass. This accumulation buffer is lazily allocated
+and resized at the beginning of each frame:
+
+{\footnotesize\lstinputlisting[firstline=460, lastline=483]
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
+
+In idle mode, the results are accumulated at the end of the frame and displayed
+after each iteration by \textsf{frameViewFinish}. Since \textsf{frameViewFinish}
+is only called on destination channels this operation is done only on the final
+rendering result. When all the steps are done the config will stop rendering new
+frames.
+
+If the current pixel viewport is different from the one saved in
+\textsf{frameViewStart}, the accumulation buffer needs also to be resized and
+the idle anti-aliasing is reset:
+
+{\footnotesize\lstinputlisting[firstline=317, lastline=344]
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
 The subpixel area is a function of the current jitter step, the channel's
-\textsf{subpixel} description and the idle state. To evenly distribute jitter
-values for all channels over the total sub-pixel grid a precomputed prime number
-table is used to pseudo-randomly select the subpixel:
-
-{\footnotesize\lstinputlisting[firstline=547, lastline=570]
+\textsf{subpixel} description and the idle state. Each source channel is
+responsible for filling a subset of the sampling grid. To quickly converge to a
+good anti-aliasing, each channel selects its samples using a pseudo-random
+approach, using a precomputed prime number table to find the subpixel for the
+current step:
+
+{\footnotesize\lstinputlisting[firstline=544, lastline=567]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
-
-When the rendering is not in idle mode, the jitter is queried from
-Equalizer which returns an optimal subpixel offset for the given subpixel
-decomposition. This is used during normal rendering of subpixel compounds.
 
@@
 no user event requires a redraw.
+
+When the rendering is not in idle mode, the jitter is queried from
+Equalizer which returns an optimal subpixel offset for the given subpixel
+decomposition. This is used during normal rendering of subpixel compounds:
+
+{\footnotesize\lstinputlisting[firstline=491, lastline=497]
+  {../../../Equalizer/examples/eqPly/channel.cpp}}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=491, lastline=533]
+{\footnotesize\lstinputlisting[firstline=503, lastline=529]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
-
-At the end of the frame, \textsf{frameViewFinish} accumulates the frame buffer
-into an accumulation buffer. Since \textsf{frameViewFinish} is only called on
-destination channels this operation is done only on the final rendering
-result. After each frame, the result accumulated so far will be displayed. When
-all the steps are done the config will stop rendering new frames.
-
-If the current pixel viewport is different from the one saved in
-\textsf{frameViewStart}, the accumulation buffer needs also to be resized and
-the idle anti-aliasing is reset.
-
-{\footnotesize\lstinputlisting[firstline=305, lastline=337]
-  {../../../Equalizer/examples/eqPly/channel.cpp}}
-
-At the end of the frame for each channel, \textsf{frameFinish} will decrement
-the jitter step and create a new event to inform the \textsf{Config} using an
-event of how many frames remain to draw before the idle anti-aliasing is
-complete:
-
-{\footnotesize\lstinputlisting[firstline=346, lastline=371]
-  {../../../Equalizer/examples/eqPly/channel.cpp}}
-
-\subsubsection{Transparent Software Anti-Aliasing}
-
-Any Equalizer applications can benefit without modification from subpixel
-compounds. Applications performing their own, configurable idle or non-idle
-anti-aliasing can easily be integrated with subpixel compounds, as described in
-the next section.
-
-The default implementation of \textsf{applyFrustum} and \textsf{applyOrtho} in
-\textsf{eq::Channel} apply jittering when a subpixel parameter is set.
-Furthermore, the \textsf{Compositor} uses an \textsf{eq::Accum} buffer to
-accumulate and display the results of a subpixel decomposition in \textsf{assembleFramesUnsorted} and \textsf{assembleFramesSorted}.
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/statistics}
-  {\caption{\label{fStatistics}Statistics for a two node 2D compound}}
-\end{wrapfigure}
 Statistics are enabled per entity using an attribute hint. The hint determines
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/statistics}
+  {\caption{\label{fStatistics}Statistics for a two node 2D compound}}
+\end{wrapfigure}
 The events are processed by the channel's and window's \textsf{processEvent}
@@
 event, it sorts the event per frame and per originator. When a frame has been
-finished, the events are pushed to the local (app-)node for visualization.
+finished, the events are pushed to the local (app-)node for
+visualization. Furthermore, the server also receives these event which will be
+used to implement the various runtime adjustments by the equalizers.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
+  {\caption{\label{fNetNode}Communication between two Nodes}}
+\end{wrapfigure}
 The \textsf{co::Node} is the abstraction of one process in the cluster. Each
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
-  {\caption{\label{fNetNode}Communication between two Nodes}}
-\end{wrapfigure}
 A remote \textsf{Node} can either be connected explicitly by the application or
@@
 
-Distributed objects subclass from \textsf{eq::fabric::Serializable} or
+Distributed objects subclass from \textsf{co::Serializable} or
 \textsf{co::Object}. The application programmer implements serialization and
@@
 
-When using \textsf{commit} on slave object instances, multiple modifications on
-the same database may happen simultaneously. To avoid excessive synchronization
-in the network layer, Equalizer implements a lightweight approach which requires the
-application to avoid or handle conflicting changes on the same object.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+When using \textsf{commit} on slave object instances, multiple modifications on
+the same database may happen simultaneously. To avoid excessive synchronization
+in the network layer, Equalizer implements a lightweight approach which requires
+the application to avoid or handle conflicting changes on the same object.
+
 When a slave object instance commits data, \textsf{pack} is called on the
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
-  {\caption{\label{fNetObject}Communication between two Objects}}
-\end{wrapfigure}
 Multiple instances of the same object (identifier) can be mapped on the same
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
+  {\caption{\label{fNetObject}Communication between two Objects}}
+\end{wrapfigure}
 When the packet is intended for a specific instance, set the instance identifier
@@
 
-\subsubsection{\label{sEqObject}eq::fabric::Serializable}
+\subsubsection{\label{sEqObject}co::Serializable}
 
@@
 implementing data distribution between the server, application and render
-clients. Using dirty bits, \textsf{eq::fabric::Serializable} implements one
+clients. Using dirty bits, \textsf{co::Serializable} implements one
 usage pattern for the \textsf{co::Object}, allowing inheritance with data
@@
 
-The \textsf{eq::fabric::Serializable} data distribution is based on the concept
+The \textsf{co::Serializable} data distribution is based on the concept
 of dirty bits. Dirty bits form a 64-bit mask which marks which parts of the
@@
 For serialization, the default \textsf{co::Object} serialization functions are
-implemented by \textsf{eq::fabric::Serializable}, which (de-)serializes and
+implemented by \textsf{co::Serializable}, which (de-)serializes and
 resets the dirty bits, and calls \textsf{serialize} or \textsf{deserialize} with
@@
 
-To use \textsf{eq::fabric::Serializable}, the following steps have to be taken:
+To use \textsf{co::Serializable}, the following steps have to be taken:
 
 \begin{description}
-\item[Inherit from \textsf{eq::fabric::Serializable}:] The base class will
+\item[Inherit from \textsf{co::Serializable}:] The base class will
   provide the dirty bit management and call serialize and deserialize
@@
 
-The registration and mapping of \textsf{eq::fabric::Serializable}s is done in
+The registration and mapping of \textsf{co::Serializable}s is done in
 the same way as for \textsf{co::Object}s, which has been described in
 \sref{sNetObject2}. Equalizer objects derived from
-\textsf{eq::fabric::Serializable}, e.g., \textsf{eq::View}, are registered,
+\textsf{co::Serializable}, e.g., \textsf{eq::View}, are registered,
 mapped and synchronized by Equalizer.
@@
 \appendix
+\section{Command Line Options}
+
+Equalizer recognizes a number of command line options. To not pollute the
+application namespace for argument, all Equalizer command line options start
+with \textsf{--eq-}. The following options are recognized:
+
+\begin{description}
+\item[--eq-client] is used to start render client processes. Starts a resident
+  render client when used without additional arguments, as described in
+  \sref{sResident}. The server appends an undocumented argument to this option
+  which is used by the \textsf{eq::Client} to connect to the server and identify
+  itself.
+\item[--eq-config \textless sessionName\textbar filename.eqc\textgreater] is
+  used to configure an application-local server, as described in
+  \sref{sGPUSDUsage}.
+\item[--eq-layout \textless layoutName\textgreater] activates all layouts of the
+  given name on all respective canvases during initialization. This option may
+  be used multiple times.
+\item[--eq-listen \textless connection\textgreater] configures the
+  local node to listen on the given connections. The formation of the connection
+  argument is documented in
+  \textsf{co::Con\-nec\-tionDescription::fromString()}. This option may be used
+  multiple times.
+\item[--eq-logfile \textless filename\textgreater] redirects all Equalizer
+  output to the given file.
+\item[--eq-render-client \textless filename\textgreater] provides an alternate
+  executable name for starting render clients. This option is useful if the
+  client machines use a different directory layout from the application machine.
+\item[--eq-server \textless connection\textgreater] provides a connection
+  description when using a separate server process.
+\end{description}
+
 \section{\label{aFileFormat}File Format}
@@
   \hline
-\end{tabularx}\\\vfill
+\end{tabularx}\\
 
@@
   \hline
-\end{tabularx}\\\vfill
+\end{tabularx}\\
 \end{center}
@@
 A connection description defines the network parameters of an Equalizer
-process. Currently TCP/IP, SDP, RSP and PIPE connection types are
+process. Currently TCP/IP, SDP, RDMA, RSP and PIPE connection types are
 supported. TCP/IP creates a TCP socket. SDP is very similar, except that the
 address family \textsf{AF\_INET\_SDP} instead of \textsf{AF\_INET} is used to
-enforce a SDP connection. RSP provides reliable UDP multicast. PIPE uses a named
+enforce a SDP connection. RDMA implements native connections over the InfiniBand
+verbs protocol on Linux. RSP provides reliable UDP multicast. PIPE uses a named
 pipe for fast interprocess communication on Windows.
 
-Note that you can also use the transparent mode provided by most
-InfiniBand implementations to use SDP with TCP connections.
+On Linux, RDMA connections provide the fastest implementation for InfiniBand
+fabrics. SDP is slower than RDMA and faster than IP over InfiniBand, and is also
+supported on Windows. Note that you can also use the transparent mode provided
+by most InfiniBand implementations to use SDP with TCP connections by preloading
+the SDP shared library.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=87, lastline=98]
+{\footnotesize\lstinputlisting[firstline=87, lastline=96]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=98, lastline=108]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
-{\footnotesize\lstinputlisting[firstline=172, lastline=172]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
-{\footnotesize\lstinputlisting[firstline=178, lastline=178]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
-{\footnotesize\lstinputlisting[firstline=205, lastline=205]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
-{\footnotesize\lstinputlisting[firstline=249, lastline=249]
+{\footnotesize\lstinputlisting[firstline=98, lastline=105]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

### 2012-01-27 Developer/ProgrammingGuide/paper.tex
Finish PG review for Eq 1.2

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.8.4 for Equalizer 1.1.5+git \\[\medskipamount]
+Version 1.8.5 for Equalizer 1.2 \\[\medskipamount]
 %  \today
@@
     1.8.4   & Nov 18, 2011 & Added \sref{sAutoConfig}\\
+    1.8.5   & Jan 27, 2012 & RC review for Equalizer 1.2\\
     \hline
@@
   \textbf{See also} & EQ\_WINDOW\_IATTR\_HINT\_STEREO
-  EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
+  EQ\_COMPOUND\_IATTR\_STEREO\_MODE EQ\_CONFIG\_FATTR\_FOCUS\_DISTANCE
+  EQ\_CONFIG\_IATTR\_FOCUS\_MODE\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONFIG\_FATTR\_FOCUS\_DISTANCE\\
+  \textbf{Value} & float\\
+  \textbf{Default} & 1.0\\
+  \textbf{Details} & The default distance in meters to the focal plane. The
+  focus distance and mode influence the calculation of stereo frusta. See
+  \sref{sFocusDistance} for details.\\
+  \textbf{See also} & EQ\_CONFIG\_IATTR\_FOCUS\_MODE
+  EQ\_CONFIG\_FATTR\_EYE\_BASE\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
+  \textbf{Name} & EQ\_CONFIG\_IATTR\_FOCUS\_MODE\\
+  \textbf{Value} & fixed \textbar relative\_to\_origin \textbar
+  relative\_to\_observer\\
+  \textbf{Default} & fixed\\
+  \textbf{Details} & The mode how to use the focus distance. The
+  focus maode and distance influence the calculation of stereo frusta. See
+  \sref{sFocusDistance} for details.\\
+  \textbf{See also} & EQ\_CONFIG\_IATTR\_FOCUS\_DISTANCE
+  EQ\_CONFIG\_FATTR\_EYE\_BASE\\
   \hline
@@
   \hline
-\end{tabularx}\\
+\end{tabularx}\\\vfill
 
@@
 
-{\footnotesize\lstinputlisting[firstline=87, lastline=96]
+{\footnotesize\lstinputlisting[firstline=89, lastline=98]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=98, lastline=105]
+{\footnotesize\lstinputlisting[firstline=100, lastline=109]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-A node represents a machine in the cluster, and is one process. It has a
-name, a number of connection descriptions and at least one pipe. The
-name of the node can be used for debugging, it has no influence on the
-execution of Equalizer. For a description of node and connection
-attributes please refer to \sref{sGlobal}.
-
-{\footnotesize\lstinputlisting[firstline=107, lastline=125]
+A node represents a machine in the cluster, and is one process. It has a name, a
+hostname, a number of connection descriptions and at least one pipe. The name of
+the node can be used for debugging, it has no influence on the execution of
+Equalizer. The host is used to automatically launch remote render clients. For a
+description of node and connection attributes please refer to \sref{sGlobal}.
+
+{\footnotesize\lstinputlisting[firstline=111, lastline=129]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-For the WGL window system, the device selects the \textit{n}th GPU in
-the system. The GPU can be offline, in this case only PBuffer windows
-can be used. To restrict the OpenGL output to the GPU, the
-\textsf{WGL\_NV\_gpu\_affinity} extension is used. If the extension is
-not present, the window is opened on the \textit{n}th monitor, but
-OpenGL commands are sent to a different or all GPU's.
+For the WGL window system, the device selects the \textit{n}th GPU in the
+system. The GPU can be offline, in this case only PBuffer windows can be
+used. To restrict the OpenGL output to the GPU, the
+\textsf{WGL\_NV\_gpu\_affinity} extension is used when available. If the
+extension is not present, the window is opened on the \textit{n}th monitor, but
+OpenGL commands may be sent to a different or all GPU's, depending on the driver
+implementation.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=125, lastline=136]
+{\footnotesize\lstinputlisting[firstline=130, lastline=139]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=136, lastline=158]
+{\footnotesize\lstinputlisting[firstline=141, lastline=162]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
+Output channels are referenced in their respective segments. Source channels are
+directly referenced by their respective source compounds.
+
 The viewport of the channel is relative to the window. As for windows,
@@
 
-{\footnotesize\lstinputlisting[firstline=158, lastline=167]
+{\footnotesize\lstinputlisting[firstline=163, lastline=172]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-An observer represents a tracked entity, i.e, one user. It has a name
-and an eye separation. The name of an observer can be used for
-debugging, it has no influence on the execution of Equalizer. It can be
-used to reference the observer in views, in which case the name should
-be unique. Not all views have to be tracked by an observer.
-
-{\footnotesize\lstinputlisting[firstline=172, lastline=176]
+An observer represents a tracked entity, i.e, one user. It has a name, eye
+separation and focal plane paramters. The name of an observer can be used for
+debugging, it has no influence on the execution of Equalizer. It can be used to
+reference the observer in views, in which case the name should be unique. Not
+all views have to be tracked by an observer.
+
+{\footnotesize\lstinputlisting[firstline=176, lastline=182]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=178, lastline=181]
+{\footnotesize\lstinputlisting[firstline=184, lastline=187]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 A view has a stereo mode, which defines if the corresponding destination channel
-update the cyclop or left and right eye.
+update the cyclop or left and right eye. The stereo mode can be changed at
+runtime by the application.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=181, lastline=202]
+{\footnotesize\lstinputlisting[firstline=187, lastline=208]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-Each canvas is made of one or more segments. Segments can be planar or
+Each canvas consists of one or more segments. Segments can be planar or
 non-planar to each other, and can overlap or have gaps between each
@@
 
-The layouts referenced by the canvas can be applied to it by the
-application at runtime. One layout can be referenced by multiple
-canvases. The first layout is the layout active by default.
-
-{\footnotesize\lstinputlisting[firstline=205, lastline=225]
+The layouts referenced by the canvas can be applied to it by the application at
+runtime. One layout can be referenced by multiple canvases. The first layout is
+the layout active by default, unless the command line option
+\textsf{--eq-layout} was used to select another default layout.
+
+A canvas may have a swap barrier, which becomes the default swap barrier for all
+its subsequent segments. A swap barrier is used to synchronize the output of
+multiple windows. For software swap synchronization, all windows using a swap
+barriers of the same name are synchronized. Hardware swap synchronization is
+used when a \textsf{NV\_group} is specified. All windows using the same
+\textsf{NV\_group} on a single system are synchronized with each other using
+hardware synchronization. All groups using the same \textsf{NV\_barrier} across
+systems are synchronized with each other using hardware synchronization. When
+using hardware synchronization, the barrier name is ignored.
+
+{\footnotesize\lstinputlisting[firstline=211, lastline=236]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 A segment represents a single display, i.e., a projector or monitor. It
-references a channel, has a name, viewport and frustum. The name of a
-segment can be used for debugging, it has no influence on the execution
-of Equalizer. It can be used to reference the segment, in which case the
-name should be unique.
-
-The channel referenced by the segment defines the output channel. The
-viewport of the section defines the 2D area covered by the channel on
-the canvas. Segments can overlap each other, e.g., when edge-blended
-projectors or passive stereo is used.
+references a channel, has a name, viewport, frustum and potentially a swap
+barrier. The name of a segment can be used for debugging, it has no influence on
+the execution of Equalizer. It can be used to reference the segment, in which
+case the name should be unique.
+
+The channel referenced by the segment defines the output channel. The viewport
+of the segment defines the 2D area covered by the channel on the
+canvas. Segments can overlap each other, e.g., when edge-blended projectors or
+passive stereo is used. The intersections of a segment with all views of all
+layouts create destination channels for rendering. The destination channels are
+copies of the segment's output channel with a viewport smaller or equal to the
+output channel viewport.
 
@@
 A segment can define a frustum, in which case it overrides the default
-frustum calculated from the canvas frustum and segment viewport.
-
-{\footnotesize\lstinputlisting[firstline=225, lastline=246]
+frustum calculated from the canvas frustum and segment viewport. A segment can
+have a swap barrier, which is used as the swap barrier on the destination
+compounds of all its destincation channels.
+
+{\footnotesize\lstinputlisting[firstline=238, lastline=259]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-Compound tasks describe the operations the compound executes. The
-default is all tasks for compounds with no children (leaf compounds) and
-\textsf{READBACK ASSEMBLE} for all others. The readback and assemble
-tasks are only executed if the compound has output frames or input
-frames, respectively. Tasks are not inherited by the children of a
-compound.
+Compound tasks describe the operations the compound executes. The default is all
+tasks for compounds with no children (leaf compounds) and \textsf{CLEAR READBACK
+  ASSEMBLE} for all others. A leaf compound using the same channel as its parent
+compound does not have a default clear task, since this has been executed by the
+parent already. The readback and assemble tasks are only executed if the
+compound has output frames or input frames, respectively. Tasks are not
+inherited by the children of a compound.
 
@@
 
-A swap barrier is used to synchronize the output of multiple
-compounds. All swap barriers of the same name are synchronized The
-default name is \textsf{barrier[.root\-Com\-pound\-Name]}. All windows
-using the same \textsf{NV\_group} on a single system are synchronized
-with each other using hardware synchronization. All groups using the
-same \textsf{NV\_barrier} across systems are synchronized with each
-other using hardware synchronization. When using hardware
-synchronization, the barrier name is ignored.
-
 Output frames transport frame buffer contents to input frames of the
@@
 
-{\footnotesize\lstinputlisting[firstline=249, lastline=333]
+{\footnotesize\lstinputlisting[firstline=262, lastline=333]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

### 2012-01-27 Developer/ProgrammingGuide/paper.tex
Minor fix

@@
 in \fig{fTracked}. The transformation $origin \rightarrow viewer$ is set by
-the application using \textsf{Config::setHead\-Matrix}, which is used by
+the application using \textsf{Observer::setHead\-Matrix}, which is used by
 the server to compute the eye positions, and consequently the view frusta.

### 2012-02-01 Developer/ProgrammingGuide/paper.tex
Update 1.2 PG

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.8.5 for Equalizer 1.2 \\[\medskipamount]
+Version 1.8.6 for Equalizer 1.2\\[\medskipamount]
 %  \today
@@
     1.8.5   & Jan 27, 2012 & RC review for Equalizer 1.2\\
+    1.8.6   & Feb 01, 2012 & RC2, added to \sref{sThreadModel}\\
     \hline
@@
 
+An implication of the draw\_sync and local\_sync models is that the application
+node can not issue a new frame until it has completed all its draw tasks. For
+larger cluster configurations it is therefore advisable to assign only assemble
+operations to the application node. If the machine running the application
+process should also contribute for rendering, a second node may be used to
+perform off-screen draw and readback operations for the application node
+process.
+
 \subsubsection{\label{sAdvDPlex}DPlex Compounds}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/statistics}
+  {\caption{\label{fStatistics}Statistics for a two node 2D compound}}
+\end{wrapfigure}
 Statistics are measured in milliseconds since the configuration was
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/statistics}
-  {\caption{\label{fStatistics}Statistics for a two node 2D compound}}
-\end{wrapfigure}
 The events are processed by the channel's and window's \textsf{processEvent}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
-  {\caption{\label{fNetNode}Communication between two Nodes}}
-\end{wrapfigure}
 The \textsf{co::Node} is the abstraction of one process in the cluster. Each
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
+  {\caption{\label{fNetNode}Communication between two Nodes}}
+\end{wrapfigure}
 The \textsf{co::LocalNode} is the specialization of the node for the given

### 2012-02-03 Developer/ProgrammingGuide/paper.tex

@@
 The images on the front page show the following applications build using
-Equalizer: 
-RTT DeltaGen\footnote{Image copyright Realtime Technology AG, 2008} [top
-right], 
-\textsf{Bino}, a stereo-capable, multi-display video player\footnote{\link{http://bino.nongnu.org/}}
-[middle center],
-a flow visualization application for climate research\footnote{Image
-  courtesy of Computer Graphics and Multimedia Systems, University of
-  Siegen} [middle right],
-the \textsf{eqPly} polygonal renderer in a three-sided CAVE [bottom
-left],
-the \textsf{eVolve} volume renderer\footnote{Data set courtesy of
-  General Electric, USA} [bottom center],
-and the \textsf{eqNBody} gravitational n-body simulation, based on the
-N-Body CUDA example [bottom right].
+Equalizer: RTT DeltaGen\footnote{Image copyright Realtime Technology AG, 2008}
+[top right], \textsf{Bino}, a stereo-capable, multi-display video
+player\footnote{\link{http://bino.nongnu.org/}} [middle center], a flow
+visualization application for climate research\footnote{Image courtesy of
+  Computer Graphics and Multimedia Systems, University of Siegen} [middle
+  right], the \textsf{eqPly} polygonal renderer in a three-sided CAVE [bottom
+  left], the \textsf{eVolve} volume renderer\footnote{Data set courtesy of
+  General Electric, USA} [bottom center], and the
+\textsf{RTNeuron}\footnote{\link{http://bluebrain.epfl.ch/}} application to
+  visualize cortical circuit simulations [bottom right].
 

### 2012-02-09 Developer/ProgrammingGuide/paper.tex
Finalize Eq 1.2 PUG

@@
   the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.8.6 for Equalizer 1.2\\[\medskipamount]
+Version 1.10 for Equalizer 1.2\\[\medskipamount]
 %  \today
@@
     1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
-    1.8.1   & Mar 25, 2011 & Minor Revision\\
-    1.8.2   & Apr 04, 2011 & Added \sref{sFocusDistance}\\
-    1.8.3   & Aug 10, 2011 & Rewrote \sref{sHello}\\
-    1.8.4   & Nov 18, 2011 & Added \sref{sAutoConfig}\\
-    1.8.5   & Jan 27, 2012 & RC review for Equalizer 1.2\\
-    1.8.6   & Feb 01, 2012 & RC2, added to \sref{sThreadModel}\\
+    1.10    & Feb 09, 2012 & Revision for Equalizer 1.2\\
     \hline
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2em}
+  \vspace{-2ex}
   \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
@@
 
+\subsection{\label{sTileCompunds}Tile Compounds}
+
+Tile compounds are similar to 2D compounds. They decompose the rendering in
+screen-space, where each rendering unit pulls and processes regular tiles of the
+final view. Tile compounds are ideal for purely fill-limited applications such
+as volume rendering and raytracing.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/tile.pdf}
+  {\caption{\label{fPixel}\small A Tile Compound}}
+\end{wrapfigure}
+Tile compounds have a low, constant IO overhead for the image transfers and can
+provide good scalability when used with fill-limited applications. Tile
+decomposition works transparently for all Equalizer-based applications.
+
+Contrary to all other compounds, the work distribution is not decided
+before-hand using a push model, but each resource pulls work from a central work
+queue, until all tiles have been processed. Therefore the rendering is naturally
+load-balanced since all sources pull data on an as-needed basis. The queue
+implementation uses prefetching to hide the communication overhead of the
+central, distributed tile queue.
+
 \subsection{Pixel Compounds}
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
+  {\caption{\label{fPixel}\small A Pixel Compound}}\vspace{-8ex}
+\end{wrapfigure}
 Pixel compounds are similar to 2D compounds. While 2D compounds decompose the
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
-  {\caption{\label{fPixel}\small A Pixel Compound}}
-\end{wrapfigure}
 As 2D compounds, pixel compounds have low, constant IO requirements for the
@@
 
+Applications can use subpixel compounds to accelerate depth-of-field effects and
+software anti-aliasing, with potentially a different number of idle or non-idle
+samples per pixel.
+
+As for the DB compound, the subpixel recomposition has linear increasing 
+IO requirements for the pixel transfer, with the difference that only 
+color information has to be transmitted.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-Applications can use subpixel compounds to accelerate depth-of-field effects and
-software anti-aliasing, with potentially a different number of idle or non-idle
-samples per pixel.
-
-As for the DB compound, the subpixel recomposition has linear increasing 
-IO requirements for the pixel transfer, with the difference that only 
-color information has to be transmitted.
-
 Subpixel compounds are configured using the \textsf{subpixel} parameter, using
@@
 \subsubsection{GPU Service Discovery}
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/gpu-sd.pdf}
+  {\caption{\label{fGPUSD}GPU discovery for auto-configuration}}
+\end{wrapfigure}
 Equalizer uses the GPU-SD library for discovering local and remote GPUs. Based
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/gpu-sd.pdf}
-  {\caption{\label{fGPUSD}GPU discovery for auto-configuration}}
-\end{wrapfigure}
 GPU-SD contains three components: a core library, local and remote discovery
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
-  {\caption{\label{fFrusta}Wall and Projection Parameters}}
-\end{wrapfigure}
 \fig{fFrusta} illustrates the wall and projection frustum parameters.
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
+  {\caption{\label{fFrusta}Wall and Projection Parameters}}
+\end{wrapfigure}
 A canvas has one or more layouts. One of the layouts is the active

### 2012-02-27 Developer/ProgrammingGuide/paper.tex
1.2 PG work

@@
 
-\title{Equalizer\\Programming and User Guide}
+\title{\vspace{7cm}
+Equalizer 1.2\\Programming and User Guide}
 \author{Eyescale Software GmbH}
 
-\date{
-  \vspace{14cm}
-\begin{textblock}{}(0,4.5)
-  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
-\end{textblock} {\Large The official reference for developing and
-  deploying parallel, scalable OpenGL\texttrademark\ applications using
-  the Equalizer parallel rendering framework}\\\vspace{1cm}
-Version 1.10 for Equalizer 1.2\\[\medskipamount]
+%\date{
+%  \vspace{14cm}
+%  \begin{textblock}{}(0,4.5)
+%    \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
+%  \end{textblock}
+%      {\Large The official reference for developing and
+%        deploying parallel, scalable OpenGL\texttrademark\ applications using
+%        the Equalizer parallel rendering framework}\\\vspace{1cm}
+%      Version 1.10 for Equalizer 1.2\\[\medskipamount]
 %  \today
-}
+%}
 
@@
 
-%\clearpage
+\clearpage
 
@@
   General Electric, USA} [bottom center], and the
-\textsf{RTNeuron}\footnote{\link{http://bluebrain.epfl.ch/}} application to
-  visualize cortical circuit simulations [bottom right].
+\textsf{RTNeuron}\footnote{Image courtesy of
+  \htmladdnormallink{Cajal Blue Brain}{http://cajalbbp.cesvima.upm.es/} /
+  \htmladdnormallink{Blue Brain Project}{http://bluebrain.epfl.ch}}
+  application to visualize cortical circuit simulations [bottom right].
 
@@
     1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
-    1.10    & Feb 09, 2012 & Revision for Equalizer 1.2\\
+    1.10    & Feb 17, 2012 & Revision for Equalizer 1.2\\
     \hline
@@
 \pagenumbering{arabic}
-\pagestyle{headings}
 
@@
   \includegraphics[width=.618\textwidth]{images/tile.pdf}
-  {\caption{\label{fPixel}\small A Tile Compound}}
+  {\caption{\label{fTile}\small A Tile Compound}}
 \end{wrapfigure}
@@
 
-{\footnotesize\lstinputlisting[firstline=57, lastline=68]
+{\footnotesize\lstinputlisting[firstline=58, lastline=68]
   {../../../Equalizer/examples/eqPly/main.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=80, lastline=88]
+{\footnotesize\lstinputlisting[firstline=81, lastline=89]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=90, lastline=100]
+{\footnotesize\lstinputlisting[firstline=91, lastline=101]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=101, lastline=118]
+{\footnotesize\lstinputlisting[firstline=102, lastline=119]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=132, lastline=143]
+{\footnotesize\lstinputlisting[firstline=134, lastline=143]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=161, lastline=165]
+{\footnotesize\lstinputlisting[firstline=145, lastline=159]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=148, lastline=152]
+{\footnotesize\lstinputlisting[firstline=161, lastline=165]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=60, lastline=75]
+{\footnotesize\lstinputlisting[firstline=65, lastline=76]
   {../../../Equalizer/examples/eqPly/initData.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=122, lastline=129]
+{\footnotesize\lstinputlisting[firstline=123, lastline=130]
   {../../../Equalizer/examples/eqPly/frameData.h}}
-{\footnotesize\lstinputlisting[firstline=54, lastline=80]
+{\footnotesize\lstinputlisting[firstline=55, lastline=84]
   {../../../Equalizer/examples/eqPly/frameData.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=143, lastline=174]
+{\footnotesize\lstinputlisting[firstline=142, lastline=174]
   {../../../Equalizer/examples/eqPly/vertexBufferDist.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=128, lastline=135]
+{\footnotesize\lstinputlisting[firstline=127, lastline=135]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=249, lastline=265B]
+{\footnotesize\lstinputlisting[firstline=249, lastline=265]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=588, lastline=625]
+{\footnotesize\lstinputlisting[firstline=584, lastline=621]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
 
-\begin{wrapfigure}{r}{.618\textwidth}
-%\begin{figure}[h!t]\center
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
+%\begin{wrapfigure}{r}{.618\textwidth}
+\begin{figure}[h!t]\center
+  \includegraphics[width=.9\textwidth]{images/render.pdf}
   {\caption{\label{fRender}Main Render Loop}}
-\end{wrapfigure}
+\end{figure}
 The model data is spatially organized in a 3-dimensional
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\label{fEventProcessing}Event Processing}}\vspace{-4ex}
+\end{wrapfigure}
 Events are received by an event handler. The event handler finds the
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 \textsf{Window::processEvent} is responsible for handling
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
+  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
+\end{wrapfigure}
 For each GLX window, one \textsf{GLXEventHandler} is allocated. FBO's and
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/threads.pdf}
+  {\caption{\label{fThreads}Threads within one Node Process}}
+\end{wrapfigure}
 During this initialization, Equalizer creates and manages two threads
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/threads.pdf}
-  {\caption{\label{fThreads}Threads within one Node Process}}
-\end{wrapfigure}
 The receiver and command thread are terminated when the application
@@
 
-{\footnotesize\lstinputlisting[firstline=70, lastline=106]
+{\footnotesize\lstinputlisting[firstline=70, lastline=104]
   {../../../Equalizer/libs/eq/client/glx/window.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=180, lastline=226]
+{\footnotesize\lstinputlisting[firstline=180, lastline=224]
   {../../../Equalizer/libs/eq/client/wgl/window.cpp}}
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \subfigure[]{\label{fFixed}
-    \includegraphics[width=.382\textwidth]{images/focusFixed}}\hfil
-  \subfigure[]{\label{fOrigin}
-    \includegraphics[width=.382\textwidth]{images/focusOrigin}}\hfil
-  \subfigure[]{\label{fObserver}
-    \includegraphics[width=.382\textwidth]{images/focusObserver}}
-  {\caption{Fixed(a) and dynamic focus distance relative to origin(b) and observer(c)}}
-\end{wrapfigure}
-
 The frustum calculation up to Equalizer version 1.0 places the stereo
@@
 focal plane.
+
+\begin{figure}[h!t]\center
+%\begin{wrapfigure}{r}{.382\textwidth}
+  \subfigure[]{\label{fFixed}
+    \includegraphics[width=.32\textwidth]{images/focusFixed}}\hfil
+  \subfigure[]{\label{fOrigin}
+    \includegraphics[width=.32\textwidth]{images/focusOrigin}}\hfil
+  \subfigure[]{\label{fObserver}
+    \includegraphics[width=.32\textwidth]{images/focusObserver}}
+  {\caption{Fixed(a) and dynamic focus distance relative to origin(b) and observer(c)}}
+\end{figure}
 
@@
 
-\begin{wrapfigure}{r}{.612\textwidth}
-  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
-  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
-\end{wrapfigure}
 Layout API entities (Canvas, Segment, Layout, View) are sub-classed like all
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
+  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
+\end{wrapfigure}
 The default Equalizer event handling is using the view API to maintain
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
-  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
-\end{wrapfigure}
 When a window is resized, a \textsf{CHANNEL\_RESIZE} event is eventually
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
-  {\caption{\label{fDirectSend}Direct Send Compositing}}
-\end{wrapfigure}
 To provide a motivation for the design of the image compositing
@@
 
+The frame data is a holder for images and additional information, such
+as output frame attributes and pixel data availability.
+
+An \textsf{eq::Image} holds a two-dimensional snapshot of the framebuffer and
+can contain color and/or depth information.
+
+The frame synchronization through the frame data allows the input frame
+to wait for the pixel data to become ready, which is signaled by the
+output frame after readback.
+
+Furthermore, the frame data transports the inherited range of the output
+frame's compound. The range can be used to compute the assembly order of
+multiple input frames, e.g., for sorted-blend compositing in volume
+rendering applications.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The frame data is a holder for images and additional information, such
-as output frame attributes and pixel data availability.
-
-An \textsf{eq::Image} holds a two-dimensional snapshot of the framebuffer and
-can contain color and/or depth information.
-
-The frame synchronization through the frame data allows the input frame
-to wait for the pixel data to become ready, which is signaled by the
-output frame after readback.
-
-Furthermore, the frame data transports the inherited range of the output
-frame's compound. The range can be used to compute the assembly order of
-multiple input frames, e.g., for sorted-blend compositing in volume
-rendering applications.
-
 Readback and assemble operations on frames and images are designed to be
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-%\begin{figure}[h!t]\center
-  \includegraphics[width=.618\textwidth]{images/compositor.pdf}
+%\begin{wrapfigure}{r}{.618\textwidth}
+\begin{figure}[h!t]\center
+  \includegraphics[width=.9\textwidth]{images/compositor.pdf}
   {\caption{\label{fCompositor}Functional Diagram of the Compositor}}
-\end{wrapfigure}
+\end{figure}
 On a high level, the compositor combines multiple input frames using 2D
@@
 
-{\footnotesize\lstinputlisting[firstline=317, lastline=344]
+{\footnotesize\lstinputlisting[firstline=320, lastline=344]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=544, lastline=567]
+{\footnotesize\lstinputlisting[firstline=542, lastline=563]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
+Statistics are measured in milliseconds since the configuration was
+initialized. The server synchronizes the per-configuration clock on each node
+automatically. Each statistic event records the originator's (channel, window,
+pipe, node, view or config) unique identifier.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-Statistics are measured in milliseconds since the configuration was
-initialized. The server synchronizes the per-configuration clock on each node
-automatically. Each statistic event records the originator's (channel, window,
-pipe, node, view or config) unique identifier.
-
 Statistics are enabled per entity using an attribute hint. The hint determines
@@
 
-{\footnotesize\lstinputlisting[firstline=187, lastline=208]
+{\footnotesize\lstinputlisting[firstline=188, lastline=209]
   {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
@@
 
-\cleardoublepage
-\pagestyle{empty}
-\begin{textblock}{}(0,4.5)
-  \hspace{-1cm}\includegraphics[width=22cm]{images/teaserBack.pdf}
-\end{textblock}
+%\cleardoublepage
+%\pagestyle{empty}
+%\begin{textblock}{}(0,4.5)
+%  \hspace{-1cm}\includegraphics[width=22cm]{images/teaserBack.pdf}
+%\end{textblock}
 

### 2012-03-16 Developer/ProgrammingGuide/paper.tex
Post-release cleanup

@@
 
-\title{\vspace{7cm}
-Equalizer 1.2\\Programming and User Guide}
+\title{
+Equalizer\\Programming and User Guide}
 \author{Eyescale Software GmbH}
 
-%\date{
-%  \vspace{14cm}
-%  \begin{textblock}{}(0,4.5)
-%    \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
-%  \end{textblock}
-%      {\Large The official reference for developing and
-%        deploying parallel, scalable OpenGL\texttrademark\ applications using
-%        the Equalizer parallel rendering framework}\\\vspace{1cm}
-%      Version 1.10 for Equalizer 1.2\\[\medskipamount]
-%  \today
-%}
+\date{
+  \vspace{14cm}
+  \begin{textblock}{}(0,4.5)
+    \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
+  \end{textblock}
+      {\Large The official reference for developing and
+        deploying parallel, scalable OpenGL\texttrademark\ applications using
+        the Equalizer parallel rendering framework}\\\vspace{1cm}
+      Version 1.10 for Equalizer 1.2\\[\medskipamount]
+  \today
+}
 
@@
 \lstset{language=C++}
-\section*{Equalizer 1.2 Programming and User Guide}
+\section*{Equalizer Programming and User Guide}
 \today
@@
   \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}\vspace{-4ex}
+  {\caption{\label{fEventProcessing}Event Processing}}
 \end{wrapfigure}
@@
 
-%\cleardoublepage
-%\pagestyle{empty}
-%\begin{textblock}{}(0,4.5)
-%  \hspace{-1cm}\includegraphics[width=22cm]{images/teaserBack.pdf}
-%\end{textblock}
+\cleardoublepage
+\pagestyle{empty}
+\begin{textblock}{}(0,4.5)
+  \hspace{-1cm}\includegraphics[width=22cm]{images/teaserBack.pdf}
+\end{textblock}
 

### 2012-04-10 Developer/ProgrammingGuide/paper.tex
Minor PG revision

@@
         the Equalizer parallel rendering framework}\\\vspace{1cm}
-      Version 1.10 for Equalizer 1.2\\[\medskipamount]
+      Version 1.11.1 for Equalizer 1.2+git\\[\medskipamount]
   \today
@@
     1.10    & Feb 17, 2012 & Revision for Equalizer 1.2\\
+    1.11.1  & Apr 10, 2012 & Minor revision\\
     \hline
@@
   \includegraphics[width=.618\textwidth]{images/2D.pdf}
-  {\caption{A 2D Compound}}
+  {\caption{2D Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/DB.pdf}
-  {\caption{\label{fDB}\small A Database Compound}}
+  {\caption{\label{fDB}Database Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/EYE.pdf}
-  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
+  {\caption{\label{fStereoCmp}Stereo Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
-  {\caption{ A DPlex Compound}}
+  {\caption{A DPlex Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/tile.pdf}
-  {\caption{\label{fTile}\small A Tile Compound}}
+  {\caption{\label{fTile}Tile Compound}}
 \end{wrapfigure}
@@
   \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
-  {\caption{\label{fPixel}\small A Pixel Compound}}\vspace{-8ex}
+  {\caption{\label{fPixel}Pixel Compound}}\vspace{-8ex}
 \end{wrapfigure}
@@
   \includegraphics[width=.618\textwidth]{images/Subpixel.pdf}
-  {\caption{\label{fSubpixel}\small A Subpixel Compound}}
+  {\caption{\label{fSubpixel}Subpixel Compound}}
 \end{wrapfigure}
@@
   \includegraphics[width=.382\textwidth]{images/layout.png}
-  {\caption{\label{fLayout}A Layout with four Views}}
+  {\caption{\label{fLayout}Layout with four Views}}
 \end{wrapfigure}
@@
   \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
-  {\caption{A Display Wall using a six-Segment Canvas with a two-View Layout}}
+  {\caption{Display Wall using a six-Segment Canvas with a two-View Layout}}
 \end{figure}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
+\begin{wrapfigure}{l}{.618\textwidth}
   \includegraphics[width=.618\textwidth]{images/monitorEq.pdf}
@@
 
-\subsection{The Server}
+\subsection{Server}
 The server may be started as a separate process or within the application
@@
 
-\subsection{The main Function}
+\subsection{main Function}
 
@@
 
-\subsection{The Application}
+\subsection{Application}
 
@@
 
-\subsection{The Renderer}
+\subsection{Renderer}
 
@@
 
-\section{The Programming Interface}
+\section{Programming Interface}
 
@@
   into the network layer, and to \sref{sNetObject} for distributed objects.
-\item[co::base] The co::base namespace provides C++ classes to abstract
-  the underlying operating system and implements common helper
-  functionality. One example is the \textsf{co::base::Clock} class
-  providing a high-resolution timer. Classes in this namespace are fully
-  documented in the API documentation on the Equalizer website, and are
-  not subject of this Programming Guide.
+\item[lunchbox] The lunchbox library provides C++ classes to abstract the
+  underlying operating system and to implement common helper functionality for
+  multi-threaded applications. Examples are \textsf{lunchbox::Clock} providing a
+  high-resolution timer, or \textsf{lunchbox::MTQueue} providing a thread-safe,
+  blocking fifo. Classes in this namespace are fully documented in the API
+  documentation on the Equalizer website, and are not subject of this
+  Programming Guide.
 \end{description}
@@
 \textsf{eq} namespace, with the occasional usage of functionality from the
-\textsf{eq::util}, \textsf{eq::fabric}, \textsf{co} and \textsf{co::base}
+\textsf{eq::util}, \textsf{eq::fabric}, \textsf{co} and \textsf{lunchbox}
 namespace.
@@
 {\footnotesize\lstinputlisting[firstline=89, lastline=98]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=100, lastline=109]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=111, lastline=129]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=130, lastline=139]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=141, lastline=162]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=163, lastline=172]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=176, lastline=182]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=184, lastline=187]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=188, lastline=209]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=211, lastline=236]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=238, lastline=259]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 
@@
 {\footnotesize\lstinputlisting[firstline=262, lastline=333]
-  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
 

### 2012-04-11 Developer/ProgrammingGuide/paper.tex
Minor PG revision

@@
   \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_DRAWABLE\\
-  \textbf{Value} & window \textbar \ pbuffer \textbar \ FBO\\
+  \textbf{Value} & window \textbar \ pbuffer \textbar \ FBO \textbar \ OFF\\
   \textbf{Default} & window\\
@@
   system-dependent PBuffer API's. FBO are off-screen frame buffer
-  objects. To calculate the PBuffer or FBO size on unconnected devices,
-  a pipe viewport size of 4096x4096 is assumed,
-  unless specified otherwise using the pipe's viewport parameter.\\
+  objects. A disabled drawable creates a system window without a frame buffer
+  for context sharing, e.g., for asynchronous operations in a separate
+  thread. To calculate the PBuffer or FBO size on unconnected devices, a pipe
+  viewport size of 4096x4096 is assumed, unless specified otherwise using the
+  pipe's viewport parameter.\\
   \hline

### 2012-06-25 Developer/ProgrammingGuide/paper.tex
Programming Guide update for 1.4, part I

@@
         the Equalizer parallel rendering framework}\\\vspace{1cm}
-      Version 1.11.1 for Equalizer 1.2+git\\[\medskipamount]
+      Version 1.11.2 for Equalizer 1.4-beta\\[\medskipamount]
   \today
@@
 \indentcaption{2em}
+\setcounter{tocdepth}{2}
 
@@
     1.11.1  & Apr 10, 2012 & Minor revision\\
+    1.11.2  & Jun 25, 2012 & Added \sref{sScaling} and \sref{sROI}\\
     \hline
@@
 
+\subsubsection{\label{sScaling}Application-specific Scaling}
+
+All Equalizer units in the configuration file and API are configured in
+meters. This convention allows the reuse of configurations across multiple
+applications. When visualizing real-world objects, e.g., for architectural
+visualizations, it guarantees that they appear realistic and full immersion
+into the visualization can be achieved.
+
+Certain applications want to visualize objects in immersive environments in a
+scale different from their natural size, e.g., astronomical simulations. Each
+model, and therefore each view, might have a different scale. Applications can
+declare this scale as part of the \textsf{eq::View}, which will be applied to
+the virtual environment by Equalizer. Common metric model scales are provided as
+constants.
+
 \subsection{\label{sLayoutAPI}Layout API}
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
-  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
-\end{wrapfigure}
 The default Equalizer event handling is using the view API to maintain
@@
 \fig{fViewUpdate} shows a sequence diagram of such a view update.
+
+\subsection{\label{sROI}Region of Interest}
+
+\subsubsection{Motivation}
+Regions of interest (ROI) define which part of the framebuffer was updated by
+the application. They allow Equalizer to perform important optimizations during
+scalable rendering.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \subfigure[]{\label{fROIDB2Way}
+    \includegraphics[width=.29\textwidth]{images/roiDB2way.png}
+  }\hfil
+  \subfigure[]{\label{fROIDB4Way}
+    \includegraphics[width=.29\textwidth]{images/roiDB4way.png}
+  }%
+  {\caption{\label{fROIDB}ROI for a two-way (a) and four-way DB (b) compound}}
+\end{wrapfigure}
+The first optimization concerns the compositing operation. The declared ROI
+allows Equalizer to restrict the readback, compression, network transmission and
+assembly to only the relevant pixels.
+
+This is particularly important for DB compounds, which typically cover less
+screen space on each channel as resources are added to the
+decomposition. \fig{fROIDB} illustrates this for a two-way and four-way
+decomposition. Declaring the ROI alleviates the increasing compositing cost as
+resources are added to a DB compound.
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/roi}
+  {\caption{\label{fROI}ROI for 2D load balancing}}
+\end{wrapfigure}
+For 2D compounds, the same optimization is applied, but has a smaller impact on
+the compositing performance. \fig{fROI} illustrates the ROI for a four-way 2D
+compound.
+
+The second optimization is important for load-balanced 2D compounds. The 2D
+load equalizer uses internal timing statistics to form a 2D load grid for
+computing the split of the next frame. Without ROI, an even load within each
+channel has to be assumed. With ROI, an even load within the declared, smaller
+region of interest can be used. This causes a smaller error in the load
+estimation, and therefore a more accurate load prediction.
+
+\fig{fROI} illustrates this load grid for a four-way 2D compound with ROI (top)
+and without ROI (bottom).
+
+\subsubsection{ROI in eqPly}
+The application should declare its regions of interest during rendering. This
+declaration can be very fine-grained, e.g., on each leaf node of a scene
+graph. Equalizer will track and optimize multiple regions automatically. The
+current implementation of Equalizer merges all regions of a single channel into
+one region for compositing and load-balancing. Later versions may use different,
+more optimal, heuristics based on the application-declared regions.
+
+Each leaf node of the kd-tree used in eqPly declares its region of interest when
+it is rendered. The region is calculated by projecting the bounding box into
+screen space and normalizing the resulting screen space rectangle:
+
+{\footnotesize
+  \lstinputlisting[firstline=264, lastline=269]
+                  {../../../Equalizer/examples/eqPly/vertexBufferLeaf.cpp}
+                  ...
+  \lstinputlisting[firstline=70, lastline=101]
+                  {../../../Equalizer/examples/eqPly/vertexBufferState.cpp}
+                  ...
+}
+
+The \textsf{declareRegion} is forwarded to \textsf{eq::Channel::declareRegion}.
 

### 2012-06-26 Developer/ProgrammingGuide/paper.tex
Fix layout and included files

@@
 
+Equalizer uses the GPU-SD library for discovering local and remote GPUs. Based
+on this information, a typical configuration is dynamically compiled during the
+initialization of the application. The auto-configuration uses the same format
+as the static configuration files. It can be used to create template
+configurations by running a test application or the eqServer binary and using
+the generated configuration, saved as \textless session\textgreater.auto.eqc.
+
+Please note that both GPU-SD and the GPU-SD ZeroConf module are optional and
+requires GPU-SD and a ZeroConf implementation when Equalizer is compiled.
+
+GPU-SD contains three components: a core library, local and remote discovery
+modules as well as a ZeroConf daemon. The daemon uses the appropriate module to
+query local GPUs and announces them using ZeroConf on the local network.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-Equalizer uses the GPU-SD library for discovering local and remote GPUs. Based
-on this information, a typical configuration is dynamically compiled at
-application start. The auto-configuration uses the same format as the static
-configuration files, and can therefore be used to create template configurations
-by running a test application and copying the generated configuration.
-
-Please note that both GPU-SD and the GPU-SD ZeroConf module are optional and
-requires GPU-SD and a ZeroConf implementation when Equalizer is compiled.
-
-GPU-SD contains three components: a core library, local and remote discovery
-modules as well as a ZeroConf daemon. The daemon uses the appropriate module to
-query local GPUs and announces them using ZeroConf on the local network.
-
 Equalizer uses the library with the dns\_sd discovery module to gather the
 information announced by all daemons on the local network, as well as the local
-discovery modules for standalone auto-configuration. \fig{fGPUSD} shows how
-GPU-SD is used to discover four remote GPUs on two different machines to use
-them to render a single view on a laptop.
+discovery modules for standalone auto-configuration. The default configuration
+is using all locally discovered GPUs.
+
+\fig{fGPUSD} shows how GPU-SD is used to discover four remote GPUs on two
+different machines to use them to render a single view on a laptop.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
+  {\caption{\label{fFrusta}Wall and Projection Parameters}}
+\end{wrapfigure}
 If you are writing a configuration for workstation usage you can skip
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
-  {\caption{\label{fFrusta}Wall and Projection Parameters}}
-\end{wrapfigure}
 A canvas has one or more layouts. One of the layouts is the active
@@
 
+\subsubsection{\label{sSwapBarrier}Swap and Frame Synchronization}
+Canvases and segments may have a software or hardware swap barrier to
+synchronize the buffer swap of multiple channels. The swap barriers are
+inherited from the canvas to the segment ultimately to all compounds using a
+destination channel of the segment.
+
+A software swap barrier is configured by giving it a name. Windows with a swap
+barrier of the same name synchronize with each other before executing the swap
+buffer task. Before entering the barrier, \textsf{Window::finish} is called to
+ensure that all OpenGL commands have been executed.  Swap barrier names are only
+valid within the compound tree, that is, a compound from one compound tree
+cannot be synchronized with a compound from another ompound tree.
+
+A hardware swap barrier uses a hardware component to synchronize the buffer swap
+of multiple windows. It guarantees that the swap happens at the same vertical
+retrace of all corresponding video outputs.  It is configured by setting the
+\textsf{NV\_group} and \textsf{NV\_barrier} parameters. These parameters follow
+the \textsf{NV\_swap\_group} extension, which synchronizes all windows bound to
+the same group on a single machine, and all groups bound to the same barrier
+across systems.
+
+Display synchronization uses different algorithms. Framelock synchronizes each
+vertical retrace of multiple graphic outputs using a hardware component. This is
+configured in the driver, independently of the application. Genlock synchronizes
+each horizontal and vertical retrace of multiple graphic outputs, but is not
+commonly used anymore for 3D graphics. Swap lock synchronizes the front and back
+buffer swap of multiple windows, either using a software solution based on
+network communication or a hardware solution based on a physical cable. It is
+independent of, but often used in conjunction with framelock.
+
+Framelock is used to synchronize the vertical retrace in multi-display active
+stereo installations, e.g., for edge-blended projection systems or immersive
+installations. It is combined with a software or hardware swap barrier. Software
+barriers in this case cannot guarantee that the buffer swap of all outputs
+always happens with the same retrace. The time window for an output to miss the
+retrace for the buffer swap is however very small, and the output will simply
+swap on the next retrace, typically in 16 milliseconds.
+
+Display walls made out of LCD's, monoscopic or passive stereo projection systems
+often only use a software swap barrier and no framelock. The display bezels make
+it very hard to notice the missing synchronization.
+
 \subsection{\label{sLayout}Layouts}
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/layout.png}
-  {\caption{\label{fLayout}Layout with four Views}}
-\end{wrapfigure}
 Enable the layout on each desired canvas by adding it to the
@@
 
-\begin{figure}[ht!]\center
-  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
-  {\caption{Display Wall using a six-Segment Canvas with a two-View Layout}}
-\end{figure}
-
 \subsubsection{Views}
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/layout.png}
+  {\caption{\label{fLayout}Layout with four Views}}
+\end{wrapfigure}
 A view has a fractional viewport relative to its layout.  A layout
@@
 this observer.
+
+\fig{fLayout} shows an example layout using four views on a single
+segment. \fig{fDisplay} shows a real-world setup of a single canvas with six
+segments using underlap, with a two-view layout activated. This configuration
+generates eight destination channels.
+
+\begin{figure}[ht!]\center
+  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
+  {\label{fDisplay}\caption{Display Wall using a six-Segment Canvas with a two-View Layout}}
+\end{figure}
 
@@
 
-\subsubsection{\label{sSwapBarrier}Synchronization - Swap Barriers}
-Compounds may have a software or hardware swap barrier to synchronize the buffer
-swap of multiple channels. Swap barriers are defined by the segment, which will
-set the swap barrier on all its destination channel compounds.
-
-A software swap barrier is configured by giving it a name. Windows with a swap
-barrier of the same name synchronize with each other before executing the swap
-buffer task. Before entering the barrier, \textsf{Window::finish} is called to
-ensure that all OpenGL commands have been executed.  Swap barrier names are only
-valid within the compound tree, that is, a compound from one compound tree
-cannot be synchronized with a compound from another compound tree.
-
-A hardware swap barrier uses a hardware component to synchronize the buffer swap
-of multiple windows. It guarantees that the swap happens at the same vertical
-retrace of all corresponding video outputs.  It is configured by setting the
-\textsf{NV\_group} and \textsf{NV\_barrier} parameters. These parameters follow
-the \textsf{NV\_swap\_group} extension, which synchronizes all windows bound to
-the same group on a single machine, and all groups bound to the same barrier
-across systems.
-
-Display synchronization uses different algorithms. Framelock synchronizes each
-vertical retrace of multiple graphic outputs using a hardware component. This is
-configured in the driver, independently of the application. Genlock synchronizes
-each horizontal and vertical retrace of multiple graphic outputs, but is not
-used anymore for 3D graphics. Swap lock synchronizes the front and back buffer
-swap of multiple windows, either using a software solution based on network
-communication or a hardware solution based on a physical cable. It is
-independent of, but often used in conjunction with framelock.
-
-Framelock is used to synchronize the vertical retrace in multi-display active
-stereo installations, e.g., for edge-blended projection systems or immersive
-installations. It is combined with a software or hardware swap barrier. Software
-barriers in this case can not guarantee that the buffer swap of all outputs
-always happens with the same retrace. The time window for an output to miss the
-retrace for the buffer swap is however very small, and the output will simply
-swap on the next retrace, typically in 16 milliseconds.
-
-Display walls made out of LCD's, monoscopic or passive stereo projection systems
-often only use a software swap barrier and no framelock. The display bezels make
-it very hard to notice the missing synchronization.
-
 \subsubsection{\label{sEqualizers}Adjustments - Equalizers}
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/lb.pdf}
+  {\caption{\label{fLoadBalancing}\small 2D Load-Balancing}}
+\end{wrapfigure}
 Equalizers are used to update compound parameters based on runtime
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/lb.pdf}
-  {\caption{\label{fLoadBalancing}\small 2D Load-Balancing}}
-\end{wrapfigure}
 Using a load equalizer is transparent to the application, and can be used with
@@
 
-\begin{wrapfigure}{l}{.618\textwidth}
+\begin{wrapfigure}{r}{.618\textwidth}
   \includegraphics[width=.618\textwidth]{images/monitorEq.pdf}
@@
 
-{\footnotesize\lstinputlisting[firstline=81, lastline=89]
+{\footnotesize\lstinputlisting[firstline=83, lastline=91]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=91, lastline=101]
+{\footnotesize\lstinputlisting[firstline=93, lastline=103]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=102, lastline=119]
+{\footnotesize\lstinputlisting[firstline=104, lastline=121]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=121, lastline=132]
+{\footnotesize\lstinputlisting[firstline=123, lastline=134]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=134, lastline=143]
+{\footnotesize\lstinputlisting[firstline=136, lastline=145]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=145, lastline=159]
+{\footnotesize\lstinputlisting[firstline=147, lastline=161]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=161, lastline=165]
+{\footnotesize\lstinputlisting[firstline=163, lastline=167]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=167, lastline=178]
+{\footnotesize\lstinputlisting[firstline=169, lastline=180]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=180, lastline=188]
+{\footnotesize\lstinputlisting[firstline=182, lastline=190]
   {../../../Equalizer/examples/eqPly/eqPly.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=65, lastline=76]
+{\footnotesize\lstinputlisting[firstline=66, lastline=77]
   {../../../Equalizer/examples/eqPly/initData.cpp}}
@@
   {../../../Equalizer/examples/eqPly/frameData.h}}
-{\footnotesize\lstinputlisting[firstline=55, lastline=84]
+{\footnotesize\lstinputlisting[firstline=54, lastline=82]
   {../../../Equalizer/examples/eqPly/frameData.cpp}}
@@
 \subsubsection{\label{sSceneData}Scene Data}
-
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
-  {\caption{\label{fModelDist}Scene Data in eqPly}}
-\end{wrapfigure}
 
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
+  {\caption{\label{fModelDist}Scene Data in eqPly}}
+\end{wrapfigure}
 \fig{fModelDist} shows the UML hierarchy of the model and distribution
@@
 
-{\footnotesize\lstinputlisting[firstline=142, lastline=174]
+{\footnotesize\lstinputlisting[firstline=144, lastline=176]
   {../../../Equalizer/examples/eqPly/vertexBufferDist.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=76, lastline=93]
+{\footnotesize\lstinputlisting[firstline=69, lastline=86]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=94, lastline=96]
+{\footnotesize\lstinputlisting[firstline=92, lastline=93]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=98, lastline=115]
+{\footnotesize\lstinputlisting[firstline=96, lastline=113]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=127, lastline=135]
+{\footnotesize\lstinputlisting[firstline=125, lastline=133]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=306, lastline=313]
+{\footnotesize\lstinputlisting[firstline=294, lastline=301]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=315, lastline=323]
+{\footnotesize\lstinputlisting[firstline=303, lastline=311]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=325, lastline=355]
+{\footnotesize\lstinputlisting[firstline=313, lastline=343]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-The event handling code is shortened here. A detailed description on how to
-customize event handling can be found in \sref{sEventHandling}:
-
-{\footnotesize\lstinputlisting[firstline=387, lastline=390]
-  {../../../Equalizer/examples/eqPly/config.cpp}}
-{\footnotesize\lstinputlisting[firstline=450, lastline=481]
-  {../../../Equalizer/examples/eqPly/config.cpp}}
-{\footnotesize\lstinputlisting[firstline=528, lastline=534]
+The event handling code repdroduced here is just showing the handling of one
+type of event. A detailed description on how to customize event handling can be
+found in \sref{sEventHandling}:
+
+{\footnotesize\lstinputlisting[firstline=471, lastline=476]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=280, lastline=304]
+{\footnotesize\lstinputlisting[firstline=265, lastline=292]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=249, lastline=265]
+{\footnotesize\lstinputlisting[firstline=247, lastline=263]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 objects across shared contexts. Typically one OM is used for each set of shared
-contexts of a single
-GPU\footnote{\link{http://www.equalizergraphics.com/documents/design/objectManager.html}}.
+contexts of a single GPU.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=122, lastline=137]
+{\footnotesize\lstinputlisting[firstline=124, lastline=144]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=138, lastline=164]
+{\footnotesize\lstinputlisting[firstline=146, lastline=172]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=166, lastline=177]
+{\footnotesize\lstinputlisting[firstline=174, lastline=185]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=584, lastline=621]
+{\footnotesize\lstinputlisting[firstline=593, lastline=626]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
 
-%\begin{wrapfigure}{r}{.618\textwidth}
-\begin{figure}[h!t]\center
-  \includegraphics[width=.9\textwidth]{images/render.pdf}
-  {\caption{\label{fRender}Main Render Loop}}
-\end{figure}
 The model data is spatially organized in a 3-dimensional
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+%\begin{figure}[h!t]\center
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\label{fRender}Main Render Loop}}
+\end{wrapfigure}
+%\end{figure}
 The rendering loop maintains a list of candidates to render, which
@@
 \textsf{frameReadback} and \textsf{frameAssemble} task methods. To implement an
-optimization and various customizations, \textsf{frameReadback} is
+optimization and various customizations, \textsf{frame\-Read\-back} is
 overwritten. eqPly does not need the alpha channel on the destination view. The
@@
 
-{\footnotesize\lstinputlisting[firstline=242, lastline=267]
+{\footnotesize\lstinputlisting[firstline=256, lastline=275]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 Events are received by an event handler. The event handler finds the
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\label{fEventProcessing}Event Processing}}
+\end{wrapfigure}
 \textsf{Window::processEvent} is responsible for handling
@@
 
-{\footnotesize\lstinputlisting[firstline=36, lastline=44]
+{\footnotesize\lstinputlisting[firstline=36, lastline=45]
   {../../../Equalizer/examples/eVolve/error.h}}
@@
 An implication of the draw\_sync and local\_sync models is that the application
-node can not issue a new frame until it has completed all its draw tasks. For
+node cannot issue a new frame until it has completed all its draw tasks. For
 larger cluster configurations it is therefore advisable to assign only assemble
@@
 
-{\footnotesize\lstinputlisting[firstline=138, lastline=154]
+{\footnotesize\lstinputlisting[firstline=142, lastline=158]
   {../../../Equalizer/libs/eq/client/agl/window.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=180, lastline=224]
+{\footnotesize\lstinputlisting[firstline=185, lastline=229]
   {../../../Equalizer/libs/eq/client/wgl/window.cpp}}
@@
 
-The \textsf{declareRegion} is forwarded to \textsf{eq::Channel::declareRegion}.
+The \textsf{declareRegion} is eventually forwarded to
+\textsf{eq::Channel::declareRegion}.
 
@@
 The supersampling algorithm used by eqPly performs anti-aliasing, using a
-technique which computes samples within a pixel by jittering the frustum by a
-subpixel amount. The pixel is split into an 16x16 subpixel grid, and a sample is
-taken randomly within each sub-pixel. The samples are accumulated and displayed
-after each step.
+technique which computes randomized samples within a pixel by jittering the
+frustum by a subpixel amount. The pixel is split into an 16x16 subpixel grid,
+and a sample is taken randomly within each sub-pixel. The samples are
+accumulated and displayed after each step.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=460, lastline=483]
+{\footnotesize\lstinputlisting[firstline=469, lastline=493]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=320, lastline=344]
+{\footnotesize\lstinputlisting[firstline=329, lastline=353]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=542, lastline=563]
+{\footnotesize\lstinputlisting[firstline=551, lastline=572]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=491, lastline=497]
+{\footnotesize\lstinputlisting[firstline=500, lastline=506]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=503, lastline=529]
+{\footnotesize\lstinputlisting[firstline=512, lastline=538]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=176, lastline=182]
+{\footnotesize\lstinputlisting[firstline=177, lastline=183]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=184, lastline=187]
+{\footnotesize\lstinputlisting[firstline=185, lastline=188]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=211, lastline=236]
+{\footnotesize\lstinputlisting[firstline=212, lastline=237]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=238, lastline=259]
+{\footnotesize\lstinputlisting[firstline=239, lastline=260]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}

### 2012-06-26 Developer/ProgrammingGuide/paper.tex
Update appendix with new file format tokens

@@
   \hline
+  \textbf{Name} & EQ\_PIPE\_IATTR\_HINT\_AFFINITY\\
+  \textbf{Value} & OFF \textbar \ AUTO \textbar \ CORE unsigned  \textbar
+  \ SOCKET unsigned\\
+  \textbf{Default} & AUTO\\
+  \textbf{Details} & Determines how all pipe threads are bound to processing
+  cores. When set to OFF, threads are unbound and will be scheduled by the
+  operation system on any core. If set to AUTO, threads are bound to all cores
+  of the processor which is closest to the GPU used by the pipe, that is, which
+  have the lowest latency to access the GPU device. This feature requires that
+  Lunchbox and Equalizer have been compiled with the hwloc library version 1.5
+  or later. When set to CORE, the pipe threads will be bound to the given
+  core. When set to SOCKET, the pipe threads will be bound to all cores of the
+  given processor (socket). The CORE and SOCKET settings require that Lunchbox
+  and Equalizer have been compiled with the hwloc library version 1.2 or
+  later.\\ \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
   \textbf{Name} & EQ\_PIPE\_IATTR\_HINT\_CUDA\_GL\_INTEROP\\
@@
 
-{\footnotesize\lstinputlisting[firstline=89, lastline=98]
+{\footnotesize\lstinputlisting[firstline=90, lastline=99]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=100, lastline=109]
+{\footnotesize\lstinputlisting[firstline=101, lastline=110]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=111, lastline=129]
+{\footnotesize\lstinputlisting[firstline=112, lastline=130]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=130, lastline=139]
+{\footnotesize\lstinputlisting[firstline=131, lastline=141]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=141, lastline=162]
+{\footnotesize\lstinputlisting[firstline=143, lastline=163]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=163, lastline=172]
+{\footnotesize\lstinputlisting[firstline=165, lastline=174]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=177, lastline=183]
+{\footnotesize\lstinputlisting[firstline=179, lastline=185]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=185, lastline=188]
+{\footnotesize\lstinputlisting[firstline=187, lastline=190]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=188, lastline=209]
+{\footnotesize\lstinputlisting[firstline=190, lastline=211]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=212, lastline=237]
+{\footnotesize\lstinputlisting[firstline=214, lastline=239]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=239, lastline=260]
+{\footnotesize\lstinputlisting[firstline=240, lastline=262]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=262, lastline=333]
+{\footnotesize\lstinputlisting[firstline=264, lastline=337]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}

### 2012-06-26 Developer/ProgrammingGuide/paper.tex
Bump version

@@
         the Equalizer parallel rendering framework}\\\vspace{1cm}
-      Version 1.11.2 for Equalizer 1.4-beta\\[\medskipamount]
+      Version 1.11.3 for Equalizer 1.4-beta\\[\medskipamount]
   \today
@@
     1.11.2  & Jun 25, 2012 & Added \sref{sScaling} and \sref{sROI}\\
+    1.11.3  & Jun 26, 2012 & Update \aref{aFileFormat}\\
     \hline

### 2012-06-26 Developer/ProgrammingGuide/paper.tex
Collage update

@@
     1.11.1  & Apr 10, 2012 & Minor revision\\
-    1.11.2  & Jun 25, 2012 & Added \sref{sScaling} and \sref{sROI}\\
-    1.11.3  & Jun 26, 2012 & Update \aref{aFileFormat}\\
+    1.11.2  & Jun 25, 2012 & Add \sref{sScaling} and \sref{sROI}\\
+    1.11.3  & Jun 26, 2012 & Add \sref{sZeroconf}, update \sref{sNetObject2} and \aref{aFileFormat}\\
     \hline
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\label{fEventProcessing}Event Processing}}
+\end{wrapfigure}
 The purpose of the \textsf{SystemWindow} processing method,
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 \textsf{Window::processEvent} is responsible for handling
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
-  {\caption{\label{fNetNode}Communication between two Nodes}}
-\end{wrapfigure}
 The \textsf{co::LocalNode} is the specialization of the node for the given
@@
 
+\subsubsection{\label{sZeroconf}Zeroconf Discovery}
+
+Each \textsf{LocalNode} provides a \textsf{co::Zeroconf} communicator, which
+allows node and resource discovery. This requires compiling using a Lunchbox
+library supporting dns\_sd. The service "\_collage.\_tcp" is used to announce
+the presence of a listening \textsf{LocalNode} using the zeroconf protocol to
+the network, unless the local node has no listening connections. The node
+identifier and all listening connection descriptions are announced using keys
+starting with "co\_". Internally Collage uses this information to connect
+unknown nodes by using the node identifier alone.
+
+Applications may use the zeroconf communicator to add additional key-value pairs
+to announce application-specific data, and to retrieve a snapshot of all
+key-value pairs of all discovered nodes on the network.
+
+\subsubsection{Communication between Nodes}
+
 \fig{fNetNode} shows the communication between two nodes. The nodes in
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
+  {\caption{\label{fNetNode}Communication between two Nodes}}
+\end{wrapfigure}
 Node packets are directly dispatched using
-\textsf{Dispatcher::dispatchCommand}. For object packets, the appropriate object
-is found on the local node and \textsf{LocalNode::dispatchCommand} will dispatch
-the packet.
+\textsf{Dispat\-cher::\-dis\-patchCommand}. For object packets, the appropriate
+object is found on the local node and \textsf{LocalNode::dis\-patchCommand} will
+dispatch the packet.
 
@@
 the delta between the previous and current version to all mapped slave
-objects. The slave objects sync the queued deltas when they need a version.
+objects. The slave objects sync the queued deltas when they need a
+version.
+
+Objects may have a maximum number of unapplied versions, which will cause the
+commit on the master instance to block if any slave instance has reached the
+maximum number of queued versions. By default, slave instances can queue an
+unlimited amount of unapplied versions.
+
+Objects use compression plugins to reduce the amount of data sent over the
+network. By default, the plugin with the highest compression ratio and lossless
+compression for \textsf{EQ\_COMPRESSOR\_DATATYPE\_BYTE} tokens is chosen. The
+application may override \textsf{Object::chooseCompressor} to deactivate
+compression by returning \textsf{EQ\_COMPRESSOR\_NONE} or to select
+object-specific compressors.
 
@@
 
+\subsubsection{\label{sSerializable}co::Serializable}
+
+\textsf{co::Serializable} implements one typical usage pattern for data
+distribution bsed on \textsf{co::Object}. The \textsf{co::Serializable} data
+distribution is founded on the concept of dirty bits, allowing inheritance with
+data distribution. Dirty bits form a 64-bit mask which marks which parts of the
+object have to be distributed during the next commit. It is easier to use, but
+imposes one typical way to implement data distribution.
+
+For serialization, the default \textsf{co::Object} serialization functions are
+implemented by \textsf{co::Serializable}, which (de-)serializes and
+resets the dirty bits, and calls \textsf{serialize} or \textsf{deserialize} with
+the bit mask specifying which data has to be transmitted or received. During a
+commit or sync, the current dirty bits are given, whereas during object mapping
+all dirty bits are passed to the serialization methods.
+
+To use \textsf{co::Serializable}, the following steps have to be taken:
+
+\begin{description}
+\item[Inherit from \textsf{co::Serializable}:] The base class will
+  provide the dirty bit management and call serialize and deserialize
+  appropriately. By overriding \textsf{get\-Chan\-ge\-Type}, the default
+  versioning strategy might be changed.
+\item[Define new dirty bits:] Define dirty bits for each data item by
+  starting at \textsf{Serializable::DIRTY\_CUSTOM}, shifting this value
+  consecutively for each new dirty bit.
+\item[Implement serialize and deserialize:] For each object-specific
+  dirty bit which is set, stream the corresponding data item to or from
+  the provided stream. Call the parent method first in both
+  functions. For application-specific objects, write a (de-)serialization
+  function.
+\item[Mark dirty data:] In each 'setter' method, call \textsf{setDirty}
+  with the corresponding dirty bit.
+\end{description}
+
+The registration and mapping of \textsf{co::Serializable}s is done in the same
+way as for \textsf{co::Object}s, which has been described in
+\sref{sNetObject2}.
+
 \subsubsection{\label{sSlaveCommit}Slave Object Commit}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
+  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
+\end{wrapfigure}
 The master instance can \textsf{sync} delta data received from slaves. When
@@
 
+\subsubsection{\label{sNetUsage}Usage in Equalizer}
+
+The Equalizer client library and server are built on top of the network
+layer. They influence the design of it, and serve as a sample
+implementation on how to use classes in the \textsf{co} namespace.
+
+Equalizer uses the common base class \textsf{eq::fabric::Object}, which derives
+from \textsf{co::Se\-ria\-lizable} for implementing data distribution between
+the server, application and render clients. This class manages common entity
+data, and is not intended to be subclassed directly.  The inheritance chain
+\textsf{eq::View} $\rightarrow$ \textsf{eq::fabric::View} $\rightarrow$
+\textsf{eq::fabric::Ob\-ject} $\rightarrow$ \textsf{co::Se\-rial\-izable}
+$\rightarrow$ \textsf{co::Object} serves as sample implementation for this
+process.
+
+Equalizer objects derived from \textsf{co::Serializable}, e.g.,
+\textsf{eq::View}, are registered, mapped and synchronized by Equalizer. To
+execute the rendering tasks, the server send the task command packet to the
+render clients, which dispatch and execute them in the received order.
+
 \subsection{Barrier}
@@
 The \textsf{co::Barrier} provides a networked barrier primitive. It is an
-\textsf{co::Object} used by Collage for software swap barrier support, but it
-can be used as a generic barrier in application code.
+\textsf{co::Object} used by Equalizer to implement software swap barriers, but
+it can be used as a generic barrier in application code.
 
@@
 and enter requests are automatically grouped together by version.
-
-\subsection{\label{sNetUsage}Usage in Equalizer}
-
-The Equalizer client library and server are built on top of the network
-layer. They influence the design of it, and serve as a sample
-implementation on how to use classes in the \textsf{co} namespace.
-
-\subsubsection{\label{sEqObject}co::Serializable}
-
-Equalizer uses the \textsf{co::Object} data distribution mechanism for
-implementing data distribution between the server, application and render
-clients. Using dirty bits, \textsf{co::Serializable} implements one
-usage pattern for the \textsf{co::Object}, allowing inheritance with data
-distribution. It is easier to use, but imposes one typical way to implement data
-distribution.
-
-The \textsf{co::Serializable} data distribution is based on the concept
-of dirty bits. Dirty bits form a 64-bit mask which marks which parts of the
-object have to be distributed during the next commit.
-
-For serialization, the default \textsf{co::Object} serialization functions are
-implemented by \textsf{co::Serializable}, which (de-)serializes and
-resets the dirty bits, and calls \textsf{serialize} or \textsf{deserialize} with
-the bit mask specifying which data has to be transmitted or received. During a
-commit or sync, the current dirty bits are given, whereas during object mapping
-all dirty bits are passed to the serialization methods.
-
-To use \textsf{co::Serializable}, the following steps have to be taken:
-
-\begin{description}
-\item[Inherit from \textsf{co::Serializable}:] The base class will
-  provide the dirty bit management and call serialize and deserialize
-  appropriately. By overriding \textsf{get\-Chan\-ge\-Type}, the default
-  versioning strategy might be changed.
-\item[Define new dirty bits:] Define dirty bits for each data item by
-  starting at \textsf{Serializable::DIRTY\_CUSTOM}, shifting this value
-  consecutively by each new dirty bit.
-\item[Implement serialize and deserialize:] For each object-specific
-  dirty bit which is set, stream the corresponding data item to or from
-  the provided stream. Call the parent method first in both
-  functions. For application-specific objects, write a (de-)serialization
-  function.
-\item[Mark dirty data:] In each 'setter' method, call \textsf{setDirty}
-  with the corresponding dirty bit.
-\end{description}
-
-The Equalizer entities use the common base class
-\textsf{eq::fabric::Object}. This class manages common entity data, and is not
-intended to be subclassed directly.  The inheritance chain \textsf{eq::View}
-$\rightarrow$ \textsf{eq::fabric::View} $\rightarrow$
-\textsf{eq::fabric::Object} $\rightarrow$ \textsf{eq::fabric::Se\-rial\-izable}
-$\rightarrow$ \textsf{co::Object} serves as sample implementation for this
-process.
-
-The registration and mapping of \textsf{co::Serializable}s is done in
-the same way as for \textsf{co::Object}s, which has been described in
-\sref{sNetObject2}. Equalizer objects derived from
-\textsf{co::Serializable}, e.g., \textsf{eq::View}, are registered,
-mapped and synchronized by Equalizer.
 

### 2012-06-29 Developer/ProgrammingGuide/paper.tex
1.4 additions

@@
         the Equalizer parallel rendering framework}\\\vspace{1cm}
-      Version 1.11.3 for Equalizer 1.4-beta\\[\medskipamount]
+      Version 1.11.4 for Equalizer 1.4 beta\\[\medskipamount]
   \today
@@
     1.11.3  & Jun 26, 2012 & Add \sref{sZeroconf}, update \sref{sNetObject2} and \aref{aFileFormat}\\
+    1.11.4  & Jun 29, 2012 & Add \sref{sSeqPly}\\
     \hline
@@
 
+The \textsf{eqHello} example is a minimal application to illustrate the basic
+principle of any Equalizer application: The application developer has to
+implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
+applications.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The \textsf{eqHello} example is a minimal application to illustrate the basic
-principle of any Equalizer application: The application developer has to
-implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
-applications.
-
 It can be run as a stand-alone application from the command line or using an
 explicit configuration file for a visualization cluster. In the stand-alone
-case, any Equalizer application including eqHello, will automatically
-configure itself to use all graphics cards found on the local system for
-scalability.
-
-\subsection{Sequel}
-
-Sequel is a thin layer on top of the canonical Equalizer programming
-interface. It substantially simplifies the creation of parallel rendering
-applications by implementing proven design patterns of Equalizer
-applications. Sequel does not prohibit the usage of Equalizer or Collage
-functionality, it simply makes it easier to use for the common use cases.
-
-\subsection{main Function}
-
-The main function instantiates an application object, initializes it, starts the
-main loop and finally de-initializes the application:
-
-{\footnotesize\lstinputlisting[firstline=58, lastline=66]
+case, any Equalizer application, including eqHello, will automatically configure
+itself to use all graphics cards found on the local system for scalability.
+
+The \textsf{eqHello} example uses Sequel, a thin layer on top of the canonical
+Equalizer programming interface. \sref{sSeqPly} introduces Sequel in detail, and
+\sref{sEqPly} introduces the full scope of the Equalizer API.
+
+The main \textsf{eqHello} function instantiates an application object,
+initializes it, starts the main loop and finally de-initializes the application:
+
+{\footnotesize\lstinputlisting[firstline=56, lastline=64]
   {../../../Equalizer/examples/eqHello/hello.cpp}}
 
-\subsection{Application}
-
-The application object in Sequel represents one process in the cluster. The main
+The application object represents one process in the cluster. The primary
 application instance has the rendering loop and controls all execution. All
 other instances used for render client processes are passive and driven by
-Equalizer.
-
-The application is responsible for creating renderers. Sequel will request one
-renderer for each GPU rendering thread. Sequel also provides automatic mapping
-and synchronization of distributed objects, for which the application has to
-provide a creation callback. Since \textsf{eqHello} does not need any
-distributed objects it provides an empty implementation of
-\textsf{createObject}:
-
-{\footnotesize\lstinputlisting[firstline=47, lastline=55]
+Equalizer. The application is responsible for creating the renderers, of which
+one per GPU is used:
+
+{\footnotesize\lstinputlisting[firstline=47, lastline=53]
   {../../../Equalizer/examples/eqHello/hello.cpp}}
-
-\subsection{Renderer}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=69, lastline=81]
+{\footnotesize\lstinputlisting[firstline=67, lastline=79]
   {../../../Equalizer/examples/eqHello/hello.cpp}}
@@
 
-Methods called by the application have the form \textsf{verb[Noun]},
-whereas methods called by Equalizer (`Task Methods') have the form
-\textsf{nounVerb}. For example, the application calls
-\textsf{Config::startFrame} to render a new frame, which causes, among
-many other things, \textsf{Node::frameStart} to be called in all active
-render clients.
-
-The API documentation on the website or in the header files provides a
-comprehensive documentation on individual methods, types and other elements of
-the API. Methods marked with a specific version are part of the official, public
-API and have been introduced by this Equalizer version. Reasonable care is taken
-to not break API compatibility or change the semantics of these methods within
-future Equalizer versions of the same major revision. Any eventual changes to
-the public API is documented in the release notes and the file
-\textsf{CHANGES.txt}.
+The API documentation is available on the website or in the header files, and
+provides a comprehensive documentation on individual methods, types and other
+elements of the API. Methods marked with a specific version are part of the
+official, public API and have been introduced by this Equalizer
+version. Reasonable care is taken to not break API compatibility or change the
+semantics of these methods within future Equalizer versions of the same major
+revision. Any eventual changes to the public API are documented in the release
+notes and the file \textsf{CHANGES.txt}.
 
@@
 
-Application developers are exposed to the following namespaces:
+The Equalizer software stack is modularized, layering gradually more powerful,
+but less generic APIs on to of each other. It furthermore relies on a number of
+required and optional libraries. Application developers are exposed to the
+following name\-spa\-ces:
 
 \begin{description}
-\item[eq] The core namespace for the Equalizer client library. The
-  classes and their relationship in this namespace closely model the
-  configuration file format. The classes in the \textsf{eq} namespace
-  are the main subject of this Programming Guide.
+\item[seq] The core namespace for Sequel, the simple interface to the Equalizer
+  client library.
+\item[eq] The core namespace for the Equalizer client library. The classes and
+  their relationship in this namespace closely model the configuration file
+  format. The classes in the \textsf{eq} namespace are the main subject of this
+  Programming Guide.\fig{fClientUml} provides an overview map of the most
+  important classes in the Equalizer namespace, grouped by functionality.
 \item[eq::util] The eq::util namespace provides common utility classes, which
@@
   Programming Guide.
+\item[hwloc, boost, vmmlib, gpu-sd] External libraries providing manual and
+  automatic thread affinity, serialization and the foundation for RSP multicast,
+  vector and matrix mathematics as well as local and remote GPU discovery,
+  respectively. The hwloc and gpu-sd libraries are used only internally and not
+  exposed through the API.
+\item[eq::server] The server namespace, implementing the functionality of the
+  Equalizer server, which is an internal namespace not to be used by application
+  developers. The \textsf{eq::admin} namespace enables run-time administration
+  of Equalizer servers. The server does not yet expose a stable
+  API.
 \end{description}
 
-Examples shipped with Equalizer are implemented in their own namespaces, e.g.,
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/namespaces}
+  {\caption{\label{fNamespaces}Namespaces}}
+\end{wrapfigure}
+The Equalizer examples are implemented in their own namespaces, e.g.,
 \textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing from the
@@
 \textsf{eq::util}, \textsf{eq::fabric}, \textsf{co} and \textsf{lunchbox}
-namespace.
-
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/namespaces}
-  {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
-\end{wrapfigure}
-The functionality of the Equalizer server is implemented in the
-\textsf{eq::server} namespace, which is an internal namespace not to be used by
-application developers. The \textsf{eq::admin} namespace enables run-time
-administration of Equalizer servers. These namespaces do not yet expose a stable
-API. \fig{fClientUml} provides an overview map of the most important classes in
-the Equalizer namespace, grouped by functionality.
+namespace. \fig{fNamespaces} shows the namespaces and their layering.
 
@@
 \subsection{\label{sTaskMethods}Task Methods}
+
+Methods called by the application have the form \textsf{verb[Noun]},
+whereas methods called by Equalizer (`Task Methods') have the form
+\textsf{nounVerb}. For example, the application calls
+\textsf{Config::startFrame} to render a new frame, which causes, among
+many other things, \textsf{Node::frameStart} to be called in all active
+render clients.
 
@@
 
+The main thread is responsible for maintaining the application logic. It
+reacts on user events, updates the data model and requests new frames to
+be rendered. It drives the whole application, as shown in \fig{fModel}.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-The main thread is responsible for maintaining the application logic. It
-reacts on user events, updates the data model and requests new frames to
-be rendered. It drives the whole application, as shown in \fig{fModel}.
-
 The rendering threads concurrently render the application's
@@
 
-\section{\label{sEqPly}The eqPly Polygonal Renderer}
-
-In this section the source code of \textsf{eqPly} is explained in
-detail, and relevant design decisions, caveats and other remarks are
-discussed.
+\section{\label{sSeqPly}Sequel: The seqPly Polygonal Renderer}
+
+In this section the source code of \textsf{seqPly} is used to introduce the
+Sequel API in detail, and relevant design decisions, caveats and other remarks
+are discussed. Sequel is conceived to lower the entry barrier into creating
+parallel rendering applications without sacrificing major functionality. It
+implements proven design patterns of Equalizer applications. Sequel does not
+prohibit the usage of Equalizer or Collage functionality, it simply makes it
+easier to use for the common use cases.
+
+The \textsf{seqPly} parallel renderer provides a subset of features found in its
+cousin application, \textsf{eqPly}, introduced in the next section. It is more
+approachable, using approximately one tenth of the source code needed by
+\textsf{eqPly} due to the use of the Sequel abstraction layer and reduced
+functionality.
+
+\subsection{main Function}
+
+The main function is almost identical to \textsf{eqHello}. It instantiates an
+application object, initializes it, starts the main loop and finally
+de-initializes the application. The source code is not reproduced here due to
+its similarity with \sref{sHello}.
+
+\subsection{Application}
+
+The application object in Sequel represents one process in the cluster. The main
+application instance has the rendering loop and controls all execution. All
+other instances used for render client processes are passive and driven by
+Equalizer.
+
+Sequel applications derive their application object from
+\textsf{seq::Application} and selectively override functionality. They have to
+implement \textsf{createRenderer}, as explained below. The \textsf{seqPly}
+application overrides init, exit, run and implements createRenderer.
+
+The initialization and exit routines are overwritten to parse
+\textsf{seqPly}-specific command line options and to load and unload the
+requested model:
+
+{\footnotesize\lstinputlisting[firstline=41, lastline=49]
+  {../../../Equalizer/examples/seqPly/application.cpp}}
+{\footnotesize\lstinputlisting[firstline=56, lastline=60]
+  {../../../Equalizer/examples/seqPly/application.cpp}}
+
+Sequel manages distributed objects, simplifying their use. This includes
+registration, automatic creation and mapping as well as commit and
+synchronization of objects. One special object for initialization (not used in
+\textsf{seqPly}) and one for per-frame data are managed by Sequel, in addition
+to an arbitrary number of application-specific objects.
+
+The objects passed to \textsf{seq::Application::init} and
+\textsf{seq::Application::run} are automatically distributed and instantiated on
+the render clients, and then passed to the respective task callback methods. The
+application may pass a 0 pointer if it does not need an object for
+initialization or per-frame data. Objects are registered with a type, and when
+automatically created, the \textsf{createObject} method on the application or
+renderer is used to create an instance based on this type.
+
+The run method is overloaded to pass the frame data object to the Sequel
+application run loop. The object will be distributed and synchronized to all
+renderers:
+
+{\footnotesize\lstinputlisting[firstline=51, lastline=55]
+  {../../../Equalizer/examples/seqPly/application.cpp}}
+
+The application is responsible for creating renderers. Sequel will request one
+renderer for each GPU rendering thread. Sequel also provides automatic mapping
+and synchronization of distributed objects, for which the application has to
+provide a creation callback:
+
+{\footnotesize\lstinputlisting[firstline=62, lastline=77]
+  {../../../Equalizer/examples/seqPly/application.cpp}}
+
+\subsection{Renderer}
+
+The renderer is responsible for executing the application's rendering code. One
+instance for each GPU is used. All calls to a single renderer are executed
+serially and therefore thread-safe.
+
+The \textsf{seqPly} rendering code uses the same data structure and algorithm as
+\textsf{eqPly}, described in \sref{sRendering}. This renderer captures the
+GPU-specific data in a \textsf{State} object, which is created and destroyed
+during init and exit. The state captures also the OpenGL function table, which
+is available when init is called, but not yet during the constructor of the
+renderer:
+
+{\footnotesize\lstinputlisting[firstline=46, lastline=59]
+  {../../../Equalizer/examples/seqPly/renderer.cpp}}
+
+The rendering code is similar to the typical OpenGL rendering code, except for a
+few modifications to configure the rendering. First, the render context is
+applied and lighting is set up. The render context, described in detail in
+\sref{sRenderContext}, sets up the stereo buffer, 3D viewport as well as the
+projection and view matrices using the appropriate OpenGL calls. Applications
+can also retrieve the render context and apply the settings themselves:
+
+{\footnotesize\lstinputlisting[firstline=69, lastline=80]
+  {../../../Equalizer/examples/seqPly/renderer.cpp}}
+
+After the static light setup, the model matrix is applied to the existing view
+matrix, completing the modelview matrix and positioning the model. Sequel
+maintains a per-view camera, which is modified through mouse and keyboard events
+and determines the model matrix. Applications can overwrite this event handling
+and maintain their own camera model. Afterwards, the state is set up with the
+projection-modelview matrix for view frustum culling, the DB range for sort-last
+rendering and the cull-draw traversal is executed, as described in
+\sref{sRendering}:
+
+{\footnotesize\lstinputlisting[firstline=81, lastline=97]
+  {../../../Equalizer/examples/seqPly/renderer.cpp}}
+
+\section{\label{sEqPly}Equalizer: The eqPly Polygonal Renderer}
+
+In this section the source code of \textsf{eqPly} is used to introduce the
+Equalizer API in detail, and relevant design decisions, caveats and other
+remarks are discussed.
 
@@
 environments with head tracking and to render massive data sets using
-all scalable rendering features of Equalizer.
+all scalable rendering features of Equalizer. It is a superset of
+\textsf{seqPly}, introduced in \sref{sSeqPly}.
 
@@
 
-%\begin{wrapfigure}{r}{.618\textwidth}
-\begin{figure}[ht!]\center
+\begin{wrapfigure}{r}{.618\textwidth}
+%\begin{figure}[ht!]\center
   \includegraphics[width=.618\textwidth]{images/objects.pdf}
   {\caption{\label{fObjects}Scene Graph Distribution}}
-\end{figure}
+%\end{figure}
+\end{wrapfigure}
 Applications distributing a dynamic scene graph use the frame data
@@
 
-\subsubsection{Rendering}
+\subsubsection{\label{sRendering}Rendering}
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+%\begin{figure}[h!t]\center
+  \includegraphics[width=.618\textwidth]{images/render.pdf}
+  {\caption{\label{fRender}Main Render Loop}}
+\end{wrapfigure}
+%\end{figure}
 The model data is spatially organized in a 3-dimensional
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-%\begin{figure}[h!t]\center
-  \includegraphics[width=.618\textwidth]{images/render.pdf}
-  {\caption{\label{fRender}Main Render Loop}}
-\end{wrapfigure}
-%\end{figure}
 The rendering loop maintains a list of candidates to render, which
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}
-\end{wrapfigure}
 The purpose of the \textsf{SystemWindow} processing method,
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
-  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
-\end{wrapfigure}
 For each GLX window, one \textsf{GLXEventHandler} is allocated. FBO's and
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
+The first optimization concerns the compositing operation. The declared ROI
+allows Equalizer to restrict the readback, compression, network transmission and
+assembly to only the relevant pixels.
+
+\begin{wrapfigure}{l}{.618\textwidth}
   \subfigure[]{\label{fROIDB2Way}
@@
 \end{wrapfigure}
-The first optimization concerns the compositing operation. The declared ROI
-allows Equalizer to restrict the readback, compression, network transmission and
-assembly to only the relevant pixels.
-
 This is particularly important for DB compounds, which typically cover less
@@
 
+Direct send has two important properties: an algorithmic complexity of O(1) for
+each node, that is, the compositing cost per node is constant as resources are
+added, and the capability to perform total ordering during compositing, e.g, to
+back-to-front sort all contributions of a 3D volume rendering correctly.
+
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\label{fDirectSend}Direct Send Compositing}}
+\end{wrapfigure}
 The main idea behind direct send is to parallelize the costly
@@
 
-\subsubsection{\label{sCompositor}The Compositor}
-
-The \textsf{Compositor} class gathers a set of static functions which
-implement the various compositing algorithms and low-level
-optimizations. \fig{fCompositor} provides a top-down functional overview
-of the various compositor functions.
-
 %\begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{figure}
+
+\subsubsection{\label{sCompositor}The Compositor}
+
+The \textsf{Compositor} class gathers a set of static functions which
+implement the various compositing algorithms and low-level
+optimizations. \fig{fCompositor} provides a top-down functional overview
+of the various compositor functions.
+
 On a high level, the compositor combines multiple input frames using 2D
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/slices.pdf}
+  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}\vspace{-2ex}
+\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
@@
 
-\begin{wrapfigure}{l}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 For recomposition, the 2D frame buffer contents are blended to form a
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/volResult.png}
+  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
+\end{wrapfigure}
 For orthographic rendering, determining the compositing order of the
@@
 \end{figure}
-
-To compute the assembly order, \textsf{eVolve} uses the angle between
-the $origin \rightarrow slice$ vector and the near plane, as shown in
-\fig{fBlendPerspective}. When the angle becomes greater than
-90\textdegree, the compositing order of the remaining frames has to be
-changed. The result image of this composition naturally looks the same
-as the volume rendering would when rendered on a single
-channel. \fig{fVolResult} shows the result of the composition from
-\fig{fBlendPerspective}.
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
-  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
-\end{wrapfigure}
 The master instance can \textsf{sync} delta data received from slaves. When
@@
 following situations:
+\begin{wrapfigure}{r}{.618\textwidth}
+  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
+  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
+  \vspace{-10ex}
+\end{wrapfigure}
 \begin{itemize}

### 2012-09-07 Developer/ProgrammingGuide/paper.tex
PG 1.4 release

@@
         the Equalizer parallel rendering framework}\\\vspace{1cm}
-      Version 1.11.4 for Equalizer 1.4 beta\\[\medskipamount]
+      Version 1.12 for Equalizer 1.4\\[\medskipamount]
   \today
@@
     1.10    & Feb 17, 2012 & Revision for Equalizer 1.2\\
-    1.11.1  & Apr 10, 2012 & Minor revision\\
-    1.11.2  & Jun 25, 2012 & Add \sref{sScaling} and \sref{sROI}\\
-    1.11.3  & Jun 26, 2012 & Add \sref{sZeroconf}, update \sref{sNetObject2} and \aref{aFileFormat}\\
-    1.11.4  & Jun 29, 2012 & Add \sref{sSeqPly}\\
+    1.12    & Jul 20, 2012 & Revision for Equalizer 1.4\\
     \hline

### 2013-01-11 Developer/ProgrammingGuide/paper.tex
Fix file paths

@@
 {\footnotesize\lstinputlisting[firstline=142, lastline=158]
-  {../../../Equalizer/libs/eq/client/agl/window.cpp}}
+  {../../../Equalizer/eq/client/agl/window.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=70, lastline=104]
-  {../../../Equalizer/libs/eq/client/glx/window.cpp}}
+  {../../../Equalizer/eq/client/glx/window.cpp}}
 
@@
 {\footnotesize\lstinputlisting[firstline=185, lastline=229]
-  {../../../Equalizer/libs/eq/client/wgl/window.cpp}}
+  {../../../Equalizer/eq/client/wgl/window.cpp}}
 
@@
 
-The network library is implemented in the \textsf{co} namespace. It provides
-networking functionality of different abstraction layers, gradually providing
-higher level functionality for the programmer. The main primitives in Collage
-are:
+The network library is a separate project on github, implemented in the
+\textsf{co} namespace. It provides networking functionality of different
+abstraction layers, gradually providing higher level functionality for the
+programmer. The main primitives in Collage are:
 
@@
 \item[Connection] A stream-oriented point-to-point communication
-  line. Different implementations of a connection exist.
+  line. Different implementations of a connection exist. The connections
+  transmit raw data reliably between two endpoints for unicast connections, and
+  between a set of endpoints for multicast connections.
+\item[DataOStream] Abstracts the output of C++ data types onto a
+  set of connections by implementing output stream operators. Uses buffering to
+  aggregate data for network transmission.
+\item[OCommand] Extends DataOStream, implements the protocol between Collage
+  nodes.
+\item[DataIStream] Decodes a buffer of received data into C++ objects and PODs
+  by implementing input stream operators. Performs endian swapping if requested.
+\item[ICommand]
 \item[Node/LocalNode] The abstraction of a process in the cluster. Nodes

### 2013-01-12 Developer/ProgrammingGuide/paper.tex
Fix typos

@@
 \end{wrapfigure}
-Collage is a cross-platform C++ library for building heterogenous, distributed
+Collage is a cross-platform C++ library for building heterogeneous, distributed
 applications. Collage provides an abstraction of different network connections,
@@
 \fig{fProcesses} depicts the relationship between the server, application,
-render client and adminstrative processes, which are explained below.
+render client and administrative processes, which are explained below.
 
@@
 layouts. The admin library may be used to create standalone administration tools
-or from within the application code. In any case, it has an indepedent view of
+or from within the application code. In any case, it has an independent view of
 the server's configuration. Documentation for admin library is not yet part of
@@
 \item \textsf{--eq-config filename.eqc} loads a static configuration from the
-  given ascii file. The following sections describe how to write configuration
+  given ASCII file. The following sections describe how to write configuration
   files.
@@
   each display surface, and the graphics card \textbf{output} used for
-  each segment. For homogenous setups, it is often enough to know the
+  each segment. For homogeneous setups, it is often enough to know the
   number of rows and columns on each surface, as well as the overlap or
@@
 context. The drawable can be an on-screen window or an off-screen
-PBuffer or FrameBuffer Object (FBO).
+pbuffer or framebuffer Object (FBO).
 
@@
 multiple windows per swap group are not supported by all drivers. Hardware swap
-barriers require support from the OpenGL driver, and has been tested on nVidia
+barriers require support from the OpenGL driver, and has been tested on NVIDIA
 Quadro GPUs with the G-Sync option. Please refer to your OpenGL driver
@@
 valid within the compound tree, that is, a compound from one compound tree
-cannot be synchronized with a compound from another ompound tree.
+cannot be synchronized with a compound from another compound tree.
 
@@
 
-Display walls made out of LCD's, monoscopic or passive stereo projection systems
+Display walls made out of LCDs, monoscopic or passive stereo projection systems
 often only use a software swap barrier and no framelock. The display bezels make
@@
   {\caption{\label{fViewLoadBalancing}\small Cross-Segment
-      Load-Balancing for two Segments using eight GPU's}}
+      Load-Balancing for two Segments using eight GPUs}}
 \end{wrapfigure}
@@
 sub-optimal performance. The configuration on the right uses a view
-equalizer, which assigns two GPU's to the left segment and four GPU's to
+equalizer, which assigns two GPUs to the left segment and four GPUs to
 the right segment, which leads to optimal performance for this model and
@@
 \fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
-by five GPU's. The front and left segments show the model and have a
-significant rendering load. The view equalizer assigns the GPU's from
+by five GPUs. The front and left segments show the model and have a
+significant rendering load. The view equalizer assigns the GPUs from
 the top, bottom and right wall for rendering the left and front wall in
@@
   contribute to the rendering, set up an appNode without a pipe section.
-\item[Render Clients] Specify the client IP's as the hostname field in the
+\item[Render Clients] Specify the client IPs as the hostname field in the
   connection description of each node section.
@@
 \item A firewall is blocking network connections.
-\item The render client can't access the GPU's on the remote host. Set up your
+\item The render client can't access the GPUs on the remote host. Set up your
   X-Server or application rights correctly. Log into the remote machine using
@@
   high-resolution timer, or \textsf{lunchbox::MTQueue} providing a thread-safe,
-  blocking fifo. Classes in this namespace are fully documented in the API
+  blocking FIFO. Classes in this namespace are fully documented in the API
   documentation on the Equalizer website, and are not subject of this
@@
 transmitted through the data streams. For pointers, the corresponding
-object is typically a distributed object as well, and it's identifier
+object is typically a distributed object as well, and its identifier
 and potentially version is transmitted in place of its pointer.
@@
 First, it configures the objects to retain its data \textsf{latency+1} commits,
-which corresponds to the typical use case where objects are commited once per
+which corresponds to the typical use case where objects are committed once per
 frame. This allows render clients, which often are behind the application
@@
 
-The event handling code repdroduced here is just showing the handling of one
+The event handling code reproduced here is just showing the handling of one
 type of event. A detailed description on how to customize event handling can be
@@
 {\footnotesize\begin{lstlisting}
-void eq::Channel::frameDraw( const uin128_t& frameID )
+void eq::Channel::frameDraw( const uint128_t& frameID )
 {
@@
 done by the leaf nodes, which are small enough to store the vertex
-indices in a \textsf{short} value for optimal performance with VBO's.
+indices in a \textsf{short} value for optimal performance with VBOs.
 The leaf nodes reuse the objects stored in the object manager, or create
@@
 manager is used per thread (pipe), this allows a thread-safe sharing of
-the compiled display lists or VBO's across all windows of a pipe.
+the compiled display lists or VBOs across all windows of a pipe.
 
@@
 Carbon handler is removed when the window's event handler is deleted. No
-event handling is set up for PBuffers.
-
-For each GLX window, one \textsf{GLXEventHandler} is allocated. FBO's and
-PBuffers are handled in the same way as window drawables. The event dispatch
+event handling is set up for pbuffers.
+
+For each GLX window, one \textsf{GLXEventHandler} is allocated. FBOs and
+pbuffers are handled in the same way as window drawables. The event dispatch
 finds the corresponding event handler for each received event, which is then
@@
 \textsf{WGL\-Win\-dowIF::processEvent} as part of the
-\textsf{WGLWindowEvent}. No event handling is set up for PBuffers.
+\textsf{WGLWindowEvent}. No event handling is set up for pbuffers.
 
@@
 the details of the windowing API from the \textsf{eq::Window}
-implementation and facilitates the integration with new windowing API's
+implementation and facilitates the integration with new windowing APIs
 and custom applications.
@@
 
-An application typically choses to subclass the sample implementation if only
+An application typically chooses to subclass the sample implementation if only
 minor tweaks are needed for integration. For major changes or new window
@@
   pixel format and is used as the device context for creating a
-  PBuffer. If no device context is given, the display device context is
+  pbuffer. If no device context is given, the display device context is
   used. On success, it sets the window handle. Setting a window handle
@@
 difference is transparent to Equalizer applications, only the configuration file
-uses a different wall type for HMD's.
+uses a different wall type for HMDs.
 
@@
 looks in a different direction, a different object in the scene is focused, as
-indictated by the dotted circle in \fig{fObserver}.
+indicated by the dotted circle in \fig{fObserver}.
 
@@
 tiling, depth-compositing for polygonal data or sorted, alpha-blended
-compositing for semitransparent volumetric data. These operations
+compositing for semi-transparent volumetric data. These operations
 composite either directly all images on the GPU, or use a CPU-based
@@
 library supporting dns\_sd. The service "\_collage.\_tcp" is used to announce
-the presence of a listening \textsf{LocalNode} using the zeroconf protocol to
+the presence of a listening \textsf{LocalNode} using the ZeroConf protocol to
 the network, unless the local node has no listening connections. The node
@@
 
-Applications may use the zeroconf communicator to add additional key-value pairs
+Applications may use the ZeroConf communicator to add additional key-value pairs
 to announce application-specific data, and to retrieve a snapshot of all
@@
 \textsf{co::Serializable} implements one typical usage pattern for data
-distribution bsed on \textsf{co::Object}. The \textsf{co::Serializable} data
+distribution based on \textsf{co::Object}. The \textsf{co::Serializable} data
 distribution is founded on the concept of dirty bits, allowing inheritance with
@@
   \textbf{Details} & The mode how to use the focus distance. The
-  focus maode and distance influence the calculation of stereo frusta. See
+  focus mode and distance influence the calculation of stereo frusta. See
   \sref{sFocusDistance} for details.\\
@@
   to allocate a stereo visual for windows, but fall back to a mono
-  visual if allocation fails. For PBuffers, AUTO selects a mono visual.\\
+  visual if allocation fails. For pbuffers, AUTO selects a mono visual.\\
   \textbf{See also} & EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
@@
   to allocate a double-buffered visual for windows, but fall back to a
-  single-buffered visual if allocation fails. For PBuffers, AUTO selects
+  single-buffered visual if allocation fails. For pbuffers, AUTO selects
   a single-buffered visual.\\
@@
   on-screen, window system-dependent window with a full-window OpenGL
-  drawable. PBuffers are off-screen drawables created using window
-  system-dependent PBuffer API's. FBO are off-screen frame buffer
+  drawable. Pbuffers are off-screen drawables created using window
+  system-dependent pbuffer APIs. FBO are off-screen frame buffer
   objects. A disabled drawable creates a system window without a frame buffer
   for context sharing, e.g., for asynchronous operations in a separate
-  thread. To calculate the PBuffer or FBO size on unconnected devices, a pipe
+  thread. To calculate the pbuffer or FBO size on unconnected devices, a pipe
   viewport size of 4096x4096 is assumed, unless specified otherwise using the
@@
 For the WGL window system, the device selects the \textit{n}th GPU in the
-system. The GPU can be offline, in this case only PBuffer windows can be
+system. The GPU can be offline, in this case only pbuffer windows can be
 used. To restrict the OpenGL output to the GPU, the
@@
 extension is not present, the window is opened on the \textit{n}th monitor, but
-OpenGL commands may be sent to a different or all GPU's, depending on the driver
+OpenGL commands may be sent to a different or all GPUs, depending on the driver
 implementation.
@@
 the viewport are currently ignored. The default viewport is
-automatically detected. For offline GPU's, a default of 4096x4096 is
+automatically detected. For offline GPUs, a default of 4096x4096 is
 used.
@@
 An observer represents a tracked entity, i.e, one user. It has a name, eye
-separation and focal plane paramters. The name of an observer can be used for
+separation and focal plane parameters. The name of an observer can be used for
 debugging, it has no influence on the execution of Equalizer. It can be used to
@@
 frusta specified at the canvas or segment level. This is typically used
-for nonphysically correct rendering, e.g., to compare two models
+for non-physically correct rendering, e.g., to compare two models
 side-by-side. If the view does not specify a frustum, the corresponding
@@
 have a swap barrier, which is used as the swap barrier on the destination
-compounds of all its destincation channels.
+compounds of all its destination channels.
 

### 2013-01-19 Developer/ProgrammingGuide/paper.tex

@@
 \begin{description}
+\item[--eq-help] shows all available library command line options and their
+  usage.
 \item[--eq-client] is used to start render client processes. Starts a resident
@@
   itself.
+\item[--eq-layout \textless layoutName\textgreater] activates all layouts of the
+  given name on all respective canvases during initialization. This option may
+  be used multiple times.
+\item[--eq-modelunit \textless unitValue\textgreater] is used for scaling the
+  rendered models in all views. The model unit defines the size of the model wrt
+  the virtual room unit which is always in meter. The default unit is 1 (1 meter
+  or EQ\_M).
+\item[--eq-logfile \textless filename\textgreater] redirects all Equalizer
+  output to the given file.
+\item[--eq-server \textless connection\textgreater] provides a connection
+  description when using a separate server process.
 \item[--eq-config \textless sessionName\textbar filename.eqc\textgreater] is
@@
   \sref{sGPUSDUsage}.
-\item[--eq-layout \textless layoutName\textgreater] activates all layouts of the
-  given name on all respective canvases during initialization. This option may
-  be used multiple times.
+\item[--eq-config-flags \textless multiprocess\textbar multiprocess\_db\textbar
+  ethernet\textbar infiniband\textbar 2D\_horizontal\textbar
+  2D\_vertical\textbar 2D\_tiles\textgreater] is used as input for the
+  autoconfiguration to tweak the configuration generation.
+\item[--eq-config-prefixes \textless CIDR-prefixes\textgreater] is used as input
+  for the autoconfiguration to filter network interfaces.
+\item[--eq-render-client \textless filename\textgreater] provides an alternate
+  executable name for starting render clients. This option is useful if the
+  client machines use a different directory layout from the application machine.
 \item[--eq-listen \textless connection\textgreater] configures the
@@
   multiple times.
-\item[--eq-logfile \textless filename\textgreater] redirects all Equalizer
-  output to the given file.
-\item[--eq-render-client \textless filename\textgreater] provides an alternate
-  executable name for starting render clients. This option is useful if the
-  client machines use a different directory layout from the application machine.
-\item[--eq-server \textless connection\textgreater] provides a connection
-  description when using a separate server process.
+
 \end{description}

### 2013-01-21 Developer/ProgrammingGuide/paper.tex
Merge cleanup

@@
 Written by Stefan Eilemann.\\
-Contributions by Maxim Makhinya, Jonas B\"osch, Christian Marten, Sarah
-Amsellem, Patrick Bouchaud, Philippe Robert, Robert Hauck and Lucas Peetz
-Dulley.
+Contributions by Daniel Nachbaur, Maxim Makhinya, Jonas B\"osch, Christian
+Marten, Sarah Amsellem, Patrick Bouchaud, Philippe Robert, Robert Hauck and
+Lucas Peetz Dulley.
 
@@
 context. The drawable can be an on-screen window or an off-screen
-pbuffer or framebuffer Object (FBO).
+pbuffer or framebuffer object (FBO).
 

### 2013-01-22 Developer/ProgrammingGuide/paper.tex
Update autoconfig section

--- 2013-01-19 Developer/ProgrammingGuide/paper.tex
@@
 
-\subsubsection{GPU Service Discovery}
-
-Equalizer uses the GPU-SD library for discovering local and remote GPUs. Based
-on this information, a typical configuration is dynamically compiled during the
-initialization of the application. The auto-configuration uses the same format
-as the static configuration files. It can be used to create template
-configurations by running a test application or the eqServer binary and using
-the generated configuration, saved as \textless session\textgreater.auto.eqc.
-
-Please note that both GPU-SD and the GPU-SD ZeroConf module are optional and
-requires GPU-SD and a ZeroConf implementation when Equalizer is compiled.
-
-GPU-SD contains three components: a core library, local and remote discovery
+\subsubsection{Hardware Service Discovery}
+
+Equalizer uses the hwsd library for discovering local and remote GPUs and
+network interfaces. Based on this information, a typical configuration is
+dynamically compiled during the initialization of the application. The
+auto-configuration uses the same format as the static configuration files. It
+can be used to create template configurations by running a test application or
+the eqServer binary and using the generated configuration, saved as
+\textless session\textgreater.auto.eqc.
+
+Please note that both hwsd and the hwsd ZeroConf module are optional and
+requires hwsd and a ZeroConf implementation when Equalizer is compiled.
+
+hwsd contains three components: a core library, local and remote discovery
 modules as well as a ZeroConf daemon. The daemon uses the appropriate module to
-query local GPUs and announces them using ZeroConf on the local network.
+query local GPUs and network interfaces and announces them using ZeroConf on the
+local network.
 
 \begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/gpu-sd.pdf}
-  {\caption{\label{fGPUSD}GPU discovery for auto-configuration}}
+  \includegraphics[width=.618\textwidth]{images/gpu-sd.pdf} % TODO: update image (hwsd)
+  {\caption{\label{fHWSD}GPU discovery for auto-configuration}}
 \end{wrapfigure}
@@
 discovery modules for standalone auto-configuration. The default configuration
-is using all locally discovered GPUs.
-
-\fig{fGPUSD} shows how GPU-SD is used to discover four remote GPUs on two
+is using all locally discovered GPUs and network interfaces.
+
+\fig{fHWSD} shows how hwsd is used to discover four remote GPUs on two
 different machines to use them to render a single view on a laptop.
 
-\subsubsection{\label{sGPUSDUsage}Usage}
-
-On each node contributing to the configuration, install and start the GPU-SD
+\subsubsection{\label{sHWSDUsage}Usage}
+
+On each node contributing to the configuration, install and start the hwsd
 daemon. If multiple, disjoint configurations are used on the same network,
 provide the session name as a parameter when starting the daemon. Verify that
-all GPUs are visible on the application host using the \textsf{gpu\_sd\_list}
-tool. When starting the application, use the command-line parameter
-\textsf{--eq-config}:
+all GPUs and network interfaces are visible on the application host using
+the \textsf{hw\_sd\_list} tool. When starting the application, use the
+command-line parameter \textsf{--eq-config}:
 
 \begin{itemize}
-\item The default value is \textsf{local} which uses all local GPUs queried
-  using the cgl, glx or wgl GPU-SD module.
+\item The default value is \textsf{local} which uses all local GPUs and network
+  interfaces queried using the cgl, glx, or wgl GPU module and the sys network
+  module.
 \item \textsf{--eq-config sessionname} uses the dns\_sd ZeroConf module of
-  GPU-SD to query all GPUs in the subnet for the given session. The default
-  session of the GPU-SD daemon is \textsf{default}. The hostname provided by
-  ZeroConf is used to connect the nodes.
+  hwsd to query all GPUs and network interfaces in the subnet for the given
+  session. The default session of the hwsd daemon is \textsf{default}. The
+  found network interfaces are used to connect the nodes.
 \item \textsf{--eq-config filename.eqc} loads a static configuration from the
@@
   GPUs as sources.
+\item [DB Direct Send] A direct send sort-last configuration using all GPUs as
+  sources.
+\item [DB 2D] A direct send sort-last configuration using all nodes together
+  with a dynamically load-balanced sort-first configuration using local GPUs.
 \end{description}
 
-The ZeroConf hostname is used to configure the node, that is, the launch command
-has to be able to resolve this hostname for starting the render client
-processes. Given a properly set up ZeroConf installation, the default launch
-command \textsf{ssh} should be able to start the render client correctly.
+All suitable network interfaces are used to configure the node, that is, the
+launch command has to be able to resolve one hostname for starting the render
+client processes. Suitable interfaces have to be up and match optional given
+values which can be specified by the following command-line parameters:
+
+\begin{description}
+\item[--eq-config-flags \textless ethernet\textbar infiniband\textgreater]
+  Limit the selection of network interfaces to one of the those types.
+\item[--eq-config-prefixes \textless CIDR-prefixes\textgreater] Limit the
+  selection of network interfaces that match the given prefix.
+\end{description}
 
@@
   Programming Guide.
-\item[hwloc, boost, vmmlib, gpu-sd] External libraries providing manual and
+\item[hwloc, boost, vmmlib, hwsd] External libraries providing manual and
   automatic thread affinity, serialization and the foundation for RSP multicast,
-  vector and matrix mathematics as well as local and remote GPU discovery,
-  respectively. The hwloc and gpu-sd libraries are used only internally and not
-  exposed through the API.
+  vector and matrix mathematics as well as local and remote hardware (GPU,
+  network interfaces) discovery, respectively. The hwloc and hwsd libraries are
+  used only internally and not exposed through the API.
 \item[eq::server] The server namespace, implementing the functionality of the
@@
 \begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/namespaces}
+  \includegraphics[width=.618\textwidth]{images/namespaces} % TODO: update image (hwsd)
   {\caption{\label{fNamespaces}Namespaces}}
@@
 it will create a server running within the application process using
-auto-configuration as described in \sref{sGPUSDUsage}. The command line
-parameter \textsf{--eq-config} can be used to specify a gpu-sd session or
+auto-configuration as described in \sref{sHWSDUsage}. The command line
+parameter \textsf{--eq-config} can be used to specify a hwsd session or
 configuration file, and \textsf{--eq-server} to explicitly specify a server
@@
   used to configure an application-local server, as described in
-  \sref{sGPUSDUsage}.
+  \sref{sHWSDUsage}.
 \item[--eq-config-flags \textless multiprocess\textbar multiprocess\_db\textbar

### 2013-02-02 Developer/ProgrammingGuide/paper.tex
Update Developer/ProgrammingGuide/paper.tex

@@
 
-All suitable network interfaces are used to configure the node, that is, the
+All suitable network interfaces are used to configure the nodes, that is, the
 launch command has to be able to resolve one hostname for starting the render
@@
 \item[--eq-config-flags \textless ethernet\textbar infiniband\textgreater]
-  Limit the selection of network interfaces to one of the those types.
+  Limit the selection of network interfaces to one of those types.
 \item[--eq-config-prefixes \textless CIDR-prefixes\textgreater] Limit the
@@
   network interfaces) discovery, respectively. The hwloc and hwsd libraries are
-  used only internally and not exposed through the API.
+  used only internally and are not exposed through the API.
 \item[eq::server] The server namespace, implementing the functionality of the

### 2013-03-07 Developer/ProgrammingGuide/paper.tex
Minor cleanups

--- 2013-01-25 Developer/ProgrammingGuide/paper.tex
@@
 
-Equalizer does not provide any facility to perform memory transfer from and to the CUDA
-devices. This is entirely left to the programmer.
+Equalizer does not provide any facility to perform memory transfer from and to
+the CUDA devices. This is entirely left to the programmer.
 
@@
 node, the server and the render clients. It can also be used by applications to
-implement distributed processing independently or complementary to the Equalizer
-functionality.
+implement distributed processing independently or complementary to the
+Equalizer functionality.
 
@@
 \item[Connection] A stream-oriented point-to-point communication
-  line. Different implementations of a connection exist. The connections
+  line. Different implementations of a connection exists. The connections
   transmit raw data reliably between two endpoints for unicast connections, and
   between a set of endpoints for multicast connections.
-\item[DataOStream] Abstracts the output of C++ data types onto a
-  set of connections by implementing output stream operators. Uses buffering to
+\item[DataOStream] Abstracts the output of C++ data types onto a set of
+  connections by implementing output stream operators. Uses buffering to
   aggregate data for network transmission.
@@
   by implementing input stream operators. Performs endian swapping if requested.
-\item[ICommand]
 \item[Node/LocalNode] The abstraction of a process in the cluster. Nodes
@@
   nodes within a session. Objects are registered or mapped on a LocalNode.
+\item[ICommand] The other side of OCommand. Is dispatched to Object or Node to
+  execute a specific task.
 \end{description}
@@
 A connection is either closed, connected or listening. A closed connection
-cannot be used for communications. A connected connection can be used to read or
-write data to the communication peer. A listening connection can accept
+cannot be used for communications. A connected connection can be used to read
+or write data to the communication peer. A listening connection can accept
 connection requests.
@@
 
-The connection set is used to select one connection which requires some
-action. This can be a connection request on a listening connection, pending data
-on a connected connection or the notification of a disconnect.
-
-The connection and connection set can be used by applications to implement other
-network-related functionality, e.g., to communicate with a sound server on a
-different machine.
+The connection set is used to select one connection which requires some action.
+This can be a connection request on a listening connection, pending data on a
+connected connection or the notification of a disconnect.
+
+The connection and connection set can be used by applications to implement
+other network-related functionality, e.g., to communicate with a sound server
+on a different machine.
 
@@
 
-Nodes and objects communicate using packets derived from
-\textsf{co::Packet}. The basic packet dispatch is implemented in the
-\textsf{co::Dispatcher} class, from which \textsf{co::Node} and \textsf{co::Object} is sub-classed.
+Nodes and objects communicate using packets derived from \textsf{co::Packet}.
+The basic packet dispatch is implemented in the \textsf{co::Dispatcher} class,
+from which \textsf{co::Node} and \textsf{co::Object} is sub-classed.
 

### 2013-03-07 Developer/ProgrammingGuide/paper.tex
Add documentation for co::ObjectMap (Eyescale/Collage#17)

@@
 
+\subsection{ObjectMap}
+
+The \textsf{co::ObjectMap} is a specialized \textsf{co::Serializable} which
+provides automatic distribution and synchronization of \textsf{co::Object}s.
+All changed objects are committed when \textsf{co::ObjectMap} is committed, and
+sync() on slave instances updates mapped objects to their respective versions.
+
+The objects on slave ObjectMap instances are mapped explicitly, either by
+providing an already constructed instance or using an implicit object creation
+using \textsf{co::ObjectFactory}. The factory needs to be supplied upon
+construction of \textsf{co::ObjectMap} and the implementation for creating the
+desired object types. Implicitly created objects are owned by the object which
+affects the lifetime of those objects considering unmapping operations caused
+by deregister(), explicit unmapping or destruction of the slave instance
+ObjectMap.
+
+In Sequel, the \textsf{co::ObjectMap} is used to distribute and synchronize the
+InitData and FrameData objects. Sequel adds the object IDs for both objects in
+addition to the default ObjectMap for automatic mapping during first access.
+The main advantage comes from the one point commit and synchronization which
+automatically updates changed objects to their correct version.
+
 \newpage

### 2013-03-12 Developer/ProgrammingGuide/paper.tex
Merge cleanups

@@
 
-Equalizer uses the hwsd library for discovering local and remote GPUs and
+Equalizer uses the hwsd library for discovering local and remote GPUs as well as
 network interfaces. Based on this information, a typical configuration is
@@
 can be used to create template configurations by running a test application or
-the eqServer binary and using the generated configuration, saved as
-\textless session\textgreater.auto.eqc.
+the eqServer binary and using the generated configuration, saved as \textless
+session\textgreater.auto.eqc.
 
 Please note that both hwsd and the hwsd ZeroConf module are optional and
-requires hwsd and a ZeroConf implementation when Equalizer is compiled.
+require hwsd and a ZeroConf implementation when Equalizer is compiled.
 
@@
 modules as well as a ZeroConf daemon. The daemon uses the appropriate module to
-query local GPUs and network interfaces and announces them using ZeroConf on the
+query local GPUs and network interfaces to announce them using ZeroConf on the
 local network.
@@
 \item The default value is \textsf{local} which uses all local GPUs and network
-  interfaces queried using the cgl, glx, or wgl GPU module and the sys network
+  interfaces queried using the cgl, glx, or wgl GPU modules and the sys network
   module.
@@
 The objects on slave ObjectMap instances are mapped explicitly, either by
-providing an already constructed instance or using an implicit object creation
-using \textsf{co::ObjectFactory}. The factory needs to be supplied upon
-construction of \textsf{co::ObjectMap} and the implementation for creating the
-desired object types. Implicitly created objects are owned by the object which
-affects the lifetime of those objects considering unmapping operations caused
-by deregister(), explicit unmapping or destruction of the slave instance
-ObjectMap.
-
-In Sequel, the \textsf{co::ObjectMap} is used to distribute and synchronize the
-InitData and FrameData objects. Sequel adds the object IDs for both objects in
-addition to the default ObjectMap for automatic mapping during first access.
-The main advantage comes from the one point commit and synchronization which
-automatically updates changed objects to their correct version.
+providing an already constructed instance or using implicit object creation
+through the \textsf{co::ObjectFactory}. The factory needs to be supplied upon
+construction of the \textsf{co::ObjectMap} and implement creation of the desired
+object types. Implicitly created objects are owned by the object map which
+limits their lifetime. They are released upon unmapping caused by deregister(),
+explicit unmapping or destruction of the slave instance ObjectMap.
+
+In Sequel, the \textsf{co::ObjectMap} is used to distribute and synchronize all
+objects to the rendering frames. The InitData and FrameData objects are
+implicitely known and managed by Sequel, and other objects might be used in
+addition. The commit and synchronization of the object map, and consequently all
+its objects, is automatically performed in the correct places, most importantly
+the commit at the beginning of the frame, and the sync when accessing an object
+from the renderer.
 

### 2013-03-20 Developer/ProgrammingGuide/paper.tex
Update for Eyescale/Equalizer#204

@@
 
-\textcopyright 2007-2012 \htmladdnormallink{Eyescale Software GmbH}
+\textcopyright 2007-2013 \htmladdnormallink{Eyescale Software GmbH}
 {http://www.eyescale.ch}. All rights reserved. No permission is
@@
     1.12    & Jul 20, 2012 & Revision for Equalizer 1.4\\
+    1.11.1  & Mar 20, 2013 & Adapt to new head tracking event\\
     \hline
@@
 
-The configuration tries to set up a tracking device for head tracking. Equalizer
-does not provide extensive support for tracking devices, as this is an
-orthogonal problem to parallel rendering. Tracking device support has already
-been solved by a number of implementations, such as VRCO Trackd, VRPN and
-others, which have proven to be easily integrated with Equalizer. The example
-code in \textsf{eqPly} provides a reference implementation for the integration
-of such a tracking library. \sref{sTracking} provides more background on head
-tracking.
-
-{\footnotesize\lstinputlisting[firstline=96, lastline=113]
-  {../../../Equalizer/examples/eqPly/config.cpp}}
-
 The exit function of the configuration stops the render clients by calling
@@
 
-{\footnotesize\lstinputlisting[firstline=125, lastline=133]
+{\footnotesize\lstinputlisting[firstline=106, lastline=114]
   {../../../Equalizer/examples/eqPly/config.cpp}}
@@
 \end{figure}
+
+Tracking events are often delivered asynchronously by an external source. The
+recommended way to integrate them into Equalizer is to emit an
+\textsf{OBSERVER\_MOTION} event, which has to contain the observer's object
+identifier and a \textsf{Matrix4f} head matrix. Equalizer optionally implements
+camera-based head tracking using OpenCV, and uses this mechanism to inject the
+asynchronously computed tracking matrix:
+
+{\footnotesize\lstinputlisting[firstline=194, lastline=195]
+  {../../../Equalizer/eq/client/cvTracker.cpp}}
+
+This event will be dispatched to the given observer instance in the application
+process during the next \textsf{Config::handleEvents}. The observer will update
+its head matrix based on the event data:
+
+{\footnotesize\lstinputlisting[firstline=92, lastline=97]
+  {../../../Equalizer/eq/client/observer.cpp}}
 

### 2013-05-09 Developer/VR/paper.tex
Draft paper for VR project

\NeedsTeXFormat{LaTeX2e}
\documentclass[10pt,a4]{scrartcl}

\usepackage{tabularx,longtable,graphicx,a4,listings,wrapfig,subfigure,textcomp,ccaption,latexsym,setspace,algorithm,algorithmic,esvect,color}
\usepackage[absolute]{textpos}

\ifx\pdfoutput\undefined
  % We're not running pdftex
  % european (better) fonts -- does not look good with pdflatex
  \usepackage[T1]{fontenc}
  \newcommand{\href}[2]{#2\\{\hspace*{5mm}\scriptsize <#1>}\\}
\else
  \pdfcompresslevel=9
  \def\pdfBorderAttrs{/Border [0 0 0] } % No border around Links
  \usepackage{hyperref}
\fi

\title{Project, Virtual Reality}
\author{Stefan Eilemann, Benjamin Long}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\sref}[1]{Section~\ref{#1}}
\newcommand{\aref}[1]{Appendix~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}
\newcommand{\fix}[1]{\textbf{\color{red}{#1}}}
\indentcaption{2em}

% suppress  single floating lines on top (widow) and bottom(club)
%  10000 is infinity
%  tradeoff: maybe underfull vboxes
\clubpenalty=10000
\widowpenalty=10000

\begin{document}
\maketitle

\section{Project Description}

The project builds upon the Equalizer parallel rendering framework and one of
its example programs, eqPly. It adds the following features to either the
core framework, where it is generic, or to the application, where it is not:
\begin{itemize}
\item Render any 3D object in the ply file format on any display system
  supported by Equalizer (LCD, stereo screen, CAVE, display wall) using
  monoscopic and stereoscopic (active, passive and anaglyphic) rendering.
\item Generic, transparent head tracking in the core framework, based on
  OpenCV. The tracking will provide X,Y translation, an estimation of the
  observer's Z distance, and when possible, roll estimation using the eye
  positions.
\item Object manipulation using the Logitech 3D Spacemouse. The Spacemouse
  delivers relative sensor data for six degrees of freedom (rotation and
  translation), which will be used to move an object in the scene accordingly.
\item create a waterfall modeled using a particle system based on a physics
  library, which will flow around the displayed model. A spatial data structure
  is used to ply model the interaction between the model and the particle
  system.
\end{itemize}

\fix{screenshots}

\section{Tools and Methods}

\fix{How the 3D scene was built and which tools were used. Libraries used for
  the head tracking, fiducial marker tracking, physics simulation and any
  additional feature.}

\section{Installation}
\fix{Share your experience. What were the difficulties? What you liked and/or
  disliked while developing this project? What you have learned?}

\section{Conclusion}

\end{document}

### 2013-05-14 Developer/VR/paper.tex
Fill some meat into VR project article

@@
 
-\fix{How the 3D scene was built and which tools were used. Libraries used for
-  the head tracking, fiducial marker tracking, physics simulation and any
-  additional feature.}
+The project is integrated into the Equalizer parallel rendering framework, where
+possible in the core library. The non-generic parts of the work are build into
+\textsf{eqPly}, the polygonal rendering example shipped with Equalizer. The
+Equalizer programming and user guide has a detailed description on the structure
+of \textsf{eqPly}
+
+OpenCV is used for head tracking. The support of OpenCV is generic in Equalizer,
+a new field configures the camera index used for each observer. The support is
+conditional on the availability of the OpenCV library and headers at compile
+time, as well as on the presence of a camera at runtime. A face and eye
+detection filter continuously estimates the 3D position in space, and if eyes
+are detected, the roll of the observer. Head and pitch are not estimated due to
+the limited amount of freedom given by a single camera input. Since the computer
+vision is computationally expensive, the OpenCV head tracker runs asynchronous
+in a separate thread from the main application loop. After each detection, it
+injects an observer event into Equalizer, which is processed by the normal event
+handling flow.
+
+For interaction, a 3D space mouse is used. This device reports relative values
+for six degrees of freedom. Again, the spacemouse support is fully integrated
+into the Equalizer event handling code, and the events are processed by
+\textsf{eqPly} to manipulate the model matrix applying the six delta values in a
+one-to-one mapping onto the model. Three different backend implementations
+receive and convert the event from the device into a generic representation: a
+\textsf{spnav} implementation for GLX (X11, Linux), an implementation based on
+the official driver for AGL (Carbon, Mac OS X) and a raw message-based
+implemenation for WGL (Windows).
+
+\fix{Physics:}
 
 \section{Installation}
+
+\section{Discussion}
+
+This project contributed new functionality into an existing open source
+project. While this added requirements to the project, this work immediately
+benefits the existing users. Furthermore, building unto an existing framework
+also facilitates the implementation due to the reuse of existing infrastructure
+and functionality, e.g., the already-build rendering application. The resulting
+demo is also rich in features and can easily be deployed on virtually all types
+of display systems.
+
+Due to the previous knowledge with Equalizer, the integration of the spacemouse
+and OpenCV was relatively straightforward. The correct design of integrating
+camera-based tracking took some time to figure o out. At first, a synchronous
+approach polling the camera each frame was used, where the actual detection was
+used to another thread. This proved too complex and hard to integrate cleanly
+into the existing design, which was finally solved by an asynchronous approach
+injecting events into the already distributed event processing. The chosen
+design is simple in its implementation and robust in execution.
+
 \fix{Share your experience. What were the difficulties? What you liked and/or
@@
 
-\section{Conclusion}
-
 \end{document}

### 2013-05-21 Developer/VR/paper.tex
Submitted VR paper

@@
 
-\title{Project, Virtual Reality}
+\title{Project, Virtual Reality\\Group 8}
 \author{Stefan Eilemann, Benjamin Long}
@@
 
-\fix{screenshots}
-
 \section{Tools and Methods}
@@
 
-\fix{Physics:}
+%\fix{Physics:}
 
 \section{Installation}
+
+Since our project is embedded in a larger software, building it may be a bit
+more complex. The build and functionality has been test on Mac OS X 10.8 using
+the command line XCode utilities and MacPorts. Equalizer and all its
+dependencies is build using the Buildyard multi-project CMake build tool, which
+will download, configure, compile and install all projects in the correct
+order. The only other dependencies to be installed are MacPort and the official
+Mac OS X drivers for the SpaceMouse, and then the example can be build as
+follows:
+
+\begin{lstlisting}
+sudo port install bullet +universal
+sudo port install boost +universal
+sudo port install opencv +universal
+sudo port install cmake
+cd Buildyard
+make Equalizer
+\end{lstlisting}
+
+Afterwards everything is installed in \textsf{Build}, and the \textsf{eqPly}
+example application can be launched:
+
+\begin{lstlisting}
+./Build/Equalizer/bin/eqPly.app/Contents/MacOS/eqPly
+\end{lstlisting}
 
@@
 This project contributed new functionality into an existing open source
-project. While this added requirements to the project, this work immediately
-benefits the existing users. Furthermore, building unto an existing framework
-also facilitates the implementation due to the reuse of existing infrastructure
-and functionality, e.g., the already-build rendering application. The resulting
-demo is also rich in features and can easily be deployed on virtually all types
-of display systems.
+project. While this added additional requirements to the project, this work
+immediately benefits the existing users of the Equalizer parallel rendering
+framework. Furthermore, building unto an existing framework also facilitates the
+implementation due to the reuse of existing infrastructure and functionality,
+e.g., the already-build rendering application. The resulting demo is rich in
+features and can easily be deployed on virtually all types of display systems.
 
+\begin{wrapfigure}{r}{.312\textwidth}
+  \includegraphics[width=.312\textwidth]{screenshot}
+  {\caption{\label{fScreenshot}eqPly running in a five-sided cave emulation}}
+\end{wrapfigure}
 Due to the previous knowledge with Equalizer, the integration of the spacemouse
@@
 
-\fix{Share your experience. What were the difficulties? What you liked and/or
-  disliked while developing this project? What you have learned?}
-
 \end{document}

### 2013-07-25 Developer/ProgrammingGuide/paper.tex
Programming Guide for Equalizer 1.6

@@
         the Equalizer parallel rendering framework}\\\vspace{1cm}
-      Version 1.12 for Equalizer 1.4\\[\medskipamount]
+      Version 1.14 for Equalizer 1.6\\[\medskipamount]
   \today
@@
 \indentcaption{2em}
-\setcounter{tocdepth}{2}
+\setcounter{tocdepth}{3}
 
@@
     1.12    & Jul 20, 2012 & Revision for Equalizer 1.4\\
-    1.11.1  & Mar 20, 2013 & Adapt to new head tracking event\\
+    1.14    & Jul 25, 2013 & Revision for Equalizer 1.6\\
     \hline
@@
 
-Equalizer can be installed by downloading a binary installer or source
+Equalizer can be installed by using a package or building the source
 code\footnote{\link{http://www.equalizergraphics.com/downloads.html}}. After
@@
 to get familiar with the capabilities of Equalizer and the \textsf{eqPly}
-example.
-
-Equalizer uses CMake to generate platform-specific build files. Compiling
-Equalizer from source is as simple as running \textsf{make} on Linux or Mac OS
-X, running \textsf{make xcode} and building the generated XCode project on Mac
-OS X, or creating a Equalizer Visual Studio solution using \textsf{CMake} on
-Windows.
+example. Currently we provide Ubuntu packages on
+launchpad\footnote{\link{https://launchpad.net/~eilemann/+archive/equalizer}}
+and MacPorts portfiles\footnote{\link{https://github.com/Eyescale/portfiles}}.
+
+Equalizer uses Buildyard and cmake to generate platform-specific build
+files. Compiling Equalizer from source consist of cloning Buildyard, the
+Eyescale configurations and then building Equalizer:
+
+{\footnotesize\begin{lstlisting}
+  git clone https://github.com/Eyescale/Buildyard.git
+  cd Buildyard
+  git clone https://github.com/Eyescale/config.git config.eyescale
+  make Equalizer
+\end{lstlisting}}
+
+The Windows build is similar, except that CMake will generate a Visual Studio
+solution which is used to build Equalizer.
 
@@
 Equalizer server is the entity in charge of the configuration, and all other
-processes receive their configuration from the server.
+processes receive their configuration from the server. It typically runs as a
+separate entity within separate threads in the application process.
 
@@
 \end{wrapfigure}
-2D decomposes the rendering in screen-space, that is, each contributing
-rendering unit processes a tile of the final view. The recomposition
-simply assembles the tiles side-by-side on the destination view. This
-mode is also known as sort-first or SFR.
-
 The advantage of this mode is a low, constant IO overhead for the pixel
@@
 
-DB, as shown in \fig{fDB}\footnote{3D model courtesy of AVS, USA.}, decomposes
-the rendered database so that all rendering units process a part of the scene in
-parallel. This mode is also known as sort-last, and is very similar to the data
-decomposition approach used by HPC applications.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+DB, as shown in \fig{fDB}\footnote{3D model courtesy of AVS, USA.}, decomposes
+the rendered data\-base so that all rendering units process a part of the scene
+in parallel. This mode is also known as sort-last, and is very similar to the
+data decomposition approach used by HPC applications.
+
 Volume rendering applications use an ordered alpha-based blending to
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}
-  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
-  {\caption{A DPlex Compound}}
-\end{wrapfigure}
 DPlex compounds assign full, alternating frames to individual rendering
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
+  {\caption{A DPlex Compound}}
+\end{wrapfigure}
 Due to the frame consistency between consecutive frames, this mode scales
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
-  {\caption{\label{fPixel}Pixel Compound}}\vspace{-8ex}
-\end{wrapfigure}
 Pixel compounds are similar to 2D compounds. While 2D compounds decompose the
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
+  {\caption{\label{fPixel}Pixel Compound}}
+\end{wrapfigure}
 As 2D compounds, pixel compounds have low, constant IO requirements for the
@@
 
-\begin{wrapfigure}{l}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
-  {\caption{\label{fPixelKernel}\small Pixel Compound Kernel}}
-\end{wrapfigure}
 Pixel compounds work well for purely fill-limited applications. Techniques like
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
+  {\caption{\label{fPixelKernel}\small Pixel Compound Kernel}}
+\end{wrapfigure}
 Pixel compounds in Equalizer are configured using the \textsf{pixel}
@@
 
-Applications can use subpixel compounds to accelerate depth-of-field effects and
-software anti-aliasing, with potentially a different number of idle or non-idle
-samples per pixel.
-
-As for the DB compound, the subpixel recomposition has linear increasing
-IO requirements for the pixel transfer, with the difference that only
-color information has to be transmitted.
-
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
+Applications can use subpixel compounds to accelerate depth-of-field effects and
+software anti-aliasing, with potentially a different number of idle or non-idle
+samples per pixel.
+
+As for the DB compound, the subpixel recomposition has linear increasing
+IO requirements for the pixel transfer, with the difference that only
+color information has to be transmitted.
+
 Subpixel compounds are configured using the \textsf{subpixel} parameter, using
@@
 
-Multiple channels in application windows may be used to view the model
-from different viewports.
-
-Sometimes, a single window is split across multiple projectors, e.g., by
-using an external splitter such as the Matrox TripleHead2Go. In this
-case configure one channel for each segment, using the channel's
-\textsf{viewport} to configure its position relative to the window.
+Multiple channels in application windows may be used to view the model from
+different viewports. Sometimes, a single window is split across multiple
+projectors, e.g., by using an external splitter such as the Matrox
+TripleHead2Go. In this case configure one channel for each segment, using the
+channel's \textsf{viewport} to configure its position relative to the window.
 
@@
 
+If you are writing a configuration for workstation usage you can skip
+the following sections and restart with \sref{sCompounds}.
+
+Configure one \textsf{canvas} for each display surface. For planar
+surfaces, e.g., a display wall, configure a frustum. For non-planar
+surfaces, the frustum will be configured on each display segment.
+
+The frustum can be specified as a wall or projection description. Take care to
+choose your reference system for describing the frustum to be the same system as
+used by the head-tracking matrix calculated by the application.  A wall is
+completely defined by the bottom-left, bottom-right and top-left coordinates
+relative to the origin.  A projection is defined by the position and
+head-pitch-roll orientation of the projector, as well as the horizontal and
+vertical field-of-view and distance of the projection wall.
+
 \begin{wrapfigure}{r}{.618\textwidth}
@@
 \end{wrapfigure}
-If you are writing a configuration for workstation usage you can skip
-the following sections and restart with \sref{sCompounds}.
-
-Configure one \textsf{canvas} for each display surface. For planar
-surfaces, e.g., a display wall, configure a frustum. For non-planar
-surfaces, the frustum will be configured on each display segment.
-
-The frustum can be specified as a wall or projection description. Take
-care to choose your reference system for describing the frustum to be
-the same system as used by the head-tracking matrix calculated by the
-application.
-
-A wall is completely defined by the bottom-left, bottom-right and
-top-left coordinates relative to the origin.
-
-A projection is defined by the position and head-pitch-roll orientation
-of the projector, as well as the horizontal and vertical field-of-view
-and distance of the projection wall.
-
 \fig{fFrusta} illustrates the wall and projection frustum parameters.
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
-  {\caption{\label{fCanvas}A Canvas using four Segments}}
-\end{wrapfigure}
 To synchronize the video output, configure either a canvas swap barrier or a
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
+  {\caption{\label{fCanvas}A Canvas using four Segments}}
+\end{wrapfigure}
 A segment represents one output channel of the canvas, e.g., a projector
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/layout.png}
+  {\caption{\label{fLayout}Layout with four Views}}\vspace{-6ex}
+\end{wrapfigure}
 Configure one \textsf{view} for each logical view in each layout. Set the
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/layout.png}
-  {\caption{\label{fLayout}Layout with four Views}}
-\end{wrapfigure}
 A view has a fractional viewport relative to its layout.  A layout
@@
 
-A view might have an observer, in which case its frustum is tracked by
-this observer.
-
-\fig{fLayout} shows an example layout using four views on a single
+A view might have an observer, in which case its frustum is tracked by this
+observer. \fig{fLayout} shows an example layout using four views on a single
 segment. \fig{fDisplay} shows a real-world setup of a single canvas with six
@@
 
+Equalizers are used to update compound parameters based on runtime
+data. They are attached to a compound (sub-)tree, on which they
+operate. The Equalizer distribution contains numerous example
+configuration files using equalizers.
+
+\paragraph{Load Equalizer}
+While pixel, subpixel and stereo compounds are naturally load-balanced, 2D and
+DB compounds often need load-balancing for optimal rendering performance.
+
 \begin{wrapfigure}{r}{.382\textwidth}
@@
 \end{wrapfigure}
-Equalizers are used to update compound parameters based on runtime
-data. They are attached to a compound (sub-)tree, on which they
-operate. The Equalizer distribution contains numerous example
-configuration files using equalizers.
-
-\paragraph{Load Equalizer}
-While pixel, subpixel and stereo compounds are naturally load-balanced, 2D and
-DB compounds often need load-balancing for optimal rendering performance.
-
 Using a load equalizer is transparent to the application, and can be used with
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
+  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
+      Load-Balancing for two Segments using eight GPUs}}
+\end{wrapfigure}
 Depending on the model position and data structure, each segment of a
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
-  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
-      Load-Balancing for two Segments using eight GPUs}}
-\end{wrapfigure}
 \fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
@@
 
+\begin{wrapfigure}{r}{.382\textwidth}
+    \includegraphics[width=.382\textwidth]{images/viewLB.png}
+    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
+        Load-Balancing for a CAVE}}
+\end{wrapfigure}
 The view equalizer can also use resources from another display resource,
@@
 
-\begin{wrapfigure}{l}{.382\textwidth}
-    \includegraphics[width=.382\textwidth]{images/viewLB.png}
-    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
-        Load-Balancing for a CAVE}}
-\end{wrapfigure}
 \fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
+  {\caption{\label{fDFR}Dynamic Frame Resolution}}
+\end{wrapfigure}
 Dynamic Frame Resolution (DFR) trades rendering performance for visual
@@
 
-To modify an application for Equalizer, the programmer structures the
-source code so that the OpenGL rendering can be executed in parallel,
-potentially using multiple processes for cluster-based execution.
-
-\section{\label{sHello}Hello, World!}
-
-The \textsf{eqHello} example is a minimal application to illustrate the basic
-principle of any Equalizer application: The application developer has to
-implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
-applications.
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/eqHello.png}
-  {\caption{\label{fHello}Hello, World!}}
-\end{wrapfigure}
-It can be run as a stand-alone application from the command line or using an
-explicit configuration file for a visualization cluster. In the stand-alone
-case, any Equalizer application, including eqHello, will automatically configure
-itself to use all graphics cards found on the local system for scalability.
-
-The \textsf{eqHello} example uses Sequel, a thin layer on top of the canonical
-Equalizer programming interface. \sref{sSeqPly} introduces Sequel in detail, and
-\sref{sEqPly} introduces the full scope of the Equalizer API.
-
-The main \textsf{eqHello} function instantiates an application object,
-initializes it, starts the main loop and finally de-initializes the application:
-
-{\footnotesize\lstinputlisting[firstline=56, lastline=64]
-  {../../../Equalizer/examples/eqHello/hello.cpp}}
-
-The application object represents one process in the cluster. The primary
-application instance has the rendering loop and controls all execution. All
-other instances used for render client processes are passive and driven by
-Equalizer. The application is responsible for creating the renderers, of which
-one per GPU is used:
-
-{\footnotesize\lstinputlisting[firstline=47, lastline=53]
-  {../../../Equalizer/examples/eqHello/hello.cpp}}
-
-The renderer is responsible for executing the application's rendering code. One
-instance for each GPU is used. All calls to a single renderer are executed
-serially and therefore thread-safe.
-
-In \textsf{eqHello}, the renderer draws six colored quads. The only change from
-a standard OpenGL application is the usage of the rendering context provided by
-Equalizer, most notably the frustum and viewport. The rendering context is
-described in detail in \sref{sRenderContext}, and \textsf{eqHello} simply calls
-\textsf{applyRenderContext} which will execute the appropriate OpenGL calls.
-
-After setting up lighting, the model is positioned using
-\textsf{applyModelMatrix}. For convenience, Sequel maintains one camera per
-view. The usage of this camera is purely optional, an application can implement
-its own camera model.
-
-The actual OpenGL code, rendering six colored quads, is omitted here for
-brevity:
-
-{\footnotesize\lstinputlisting[firstline=67, lastline=79]
-  {../../../Equalizer/examples/eqHello/hello.cpp}}
-
-\section{Programming Interface}
+This Programming Guide introduces Equalizer using a top-down approach, starting
+with a general introduction of the API in \sref{sAPI}, followed by the
+simplified Sequel API in \sref{sSeqPly} which implements common use cases to
+deliver a large subset of the canonical Equalizer API introduced in
+\sref{sEqualizer}. \sref{sNetwork} introduces the separate Collage network
+library used as the foundation for the distributed execution and data
+synchronizing throughout Sequel and Equalizer.
+
+\section{\label{sAPI}Programming Interface}
+
+To modify an application for Equalizer, the programmer structures the source
+code so that the OpenGL rendering can be executed in parallel, potentially using
+multiple processes for cluster-based execution.
 
@@
 semantics of these methods within future Equalizer versions of the same major
-revision. Any eventual changes to the public API are documented in the release
-notes and the file \textsf{CHANGES.txt}.
+revision. Any changes to the public API are documented in the release notes and
+the file \textsf{CHANGES.txt}.
 
@@
 
+\subsection{\label{sHello}Hello, World!}
+
+\begin{wrapfigure}{r}{.382\textwidth}
+  \includegraphics[width=.382\textwidth]{images/eqHello.png}
+  {\caption{\label{fHello}Hello, World!}}
+\end{wrapfigure}
+The \textsf{eqHello} example is a minimal application to illustrate the basic
+principle of any Equalizer application: The application developer has to
+implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
+applications.
+
+It can be run as a stand-alone application from the command line or using an
+explicit configuration file for a visualization cluster. In the stand-alone
+case, any Equalizer application, including eqHello, will automatically configure
+itself to use all graphics cards found on the local system for scalability.
+
+The \textsf{eqHello} example uses Sequel, a thin layer on top of the canonical
+Equalizer programming interface. \sref{sSeqPly} introduces Sequel in detail, and
+\sref{sEqPly} introduces the full scope of the Equalizer API.
+
+The main \textsf{eqHello} function instantiates an application object,
+initializes it, starts the main loop and finally de-initializes the application:
+
+{\footnotesize\lstinputlisting[firstline=56, lastline=64]
+  {../../../Equalizer/examples/eqHello/hello.cpp}}
+
+The application object represents one process in the cluster. The primary
+application instance has the rendering loop and controls all execution. All
+other instances used for render client processes are passive and driven by
+Equalizer. The application is responsible for creating the renderers, of which
+one per GPU is used:
+
+{\footnotesize\lstinputlisting[firstline=47, lastline=53]
+  {../../../Equalizer/examples/eqHello/hello.cpp}}
+
+The renderer is responsible for executing the application's rendering code. One
+instance for each GPU is used. All calls to a single renderer are executed
+serially and therefore thread-safe.
+
+In \textsf{eqHello}, the renderer draws six colored quads. The only change from
+a standard OpenGL application is the usage of the rendering context provided by
+Equalizer, most notably the frustum and viewport. The rendering context is
+described in detail in \sref{sRenderContext}, and \textsf{eqHello} simply calls
+\textsf{applyRenderContext} which will execute the appropriate OpenGL calls.
+
+After setting up lighting, the model is positioned using
+\textsf{applyModelMatrix}. For convenience, Sequel maintains one camera per
+view. The usage of this camera is purely optional, an application can implement
+its own camera model.
+
+The actual OpenGL code, rendering six colored quads, is omitted here for
+brevity:
+
+{\footnotesize\lstinputlisting[firstline=67, lastline=79]
+  {../../../Equalizer/examples/eqHello/hello.cpp}}
+
 \subsection{\label{sNamespaces}Namespaces}
@@
 The Equalizer software stack is modularized, layering gradually more powerful,
-but less generic APIs on to of each other. It furthermore relies on a number of
+but less generic APIs on top of each other. It furthermore relies on a number of
 required and optional libraries. Application developers are exposed to the
@@
 
-\section{\label{sSeqPly}Sequel: The seqPly Polygonal Renderer}
+\section{\label{sSeqPly}The Sequel Simple Equalizer API}
 
@@
 
-\section{\label{sEqPly}Equalizer: The eqPly Polygonal Renderer}
-
+\section{\label{sEqualizer}The Equalizer Parallel Rendering Framework}
+
+The core Equalizer client library exposes the full feature set and flexibility
+to the application developer. New application development is encouraged to use
+the Sequel library, and inquire on the \textsf{eq-dev} mailing list when
+specific functionality is not available in Sequel. Only advanced and complex
+applications should be implemented using the Equalizer client library directly.
+
+\subsection{\label{sEqPly}The eqPly Polygonal Renderer}
 In this section the source code of \textsf{eqPly} is used to introduce the
@@
 
-\textsf{eqPly} is a parallel renderer for polygonal data in the
-\textsf{ply} file format. It supports all relevant Equalizer features,
-and can be used to render on large-scale displays, immersive
-environments with head tracking and to render massive data sets using
-all scalable rendering features of Equalizer. It is a superset of
-\textsf{seqPly}, introduced in \sref{sSeqPly}.
+\textsf{eqPly} is a parallel renderer for polygonal data in the \textsf{ply}
+file format. It supports nearly all Equalizer features, and can be used to
+render on large-scale displays, immersive environments with head tracking and to
+render massive data sets using all scalable rendering features of Equalizer. It
+is a superset of \textsf{seqPly}, introduced in \sref{sSeqPly}.
 
@@
 frame-specific data. Furthermore they are used for a basic scene graph
-distribution of the model data to the render client processes.
-
-\subsection{main Function}
+distribution of the model data to the render client processes. \sref{sNetwork}
+provides more background on the Collage network library.
+
+\subsubsection{main Function}
 
@@
 
-\subsection{Application}
+\subsubsection{Application}
 
@@
 
-\subsubsection{Main Loop}
+\paragraph{Main Loop}
 
@@
 
-\subsubsection{Render Clients}
+\paragraph{Render Clients}
 
@@
 
-\subsection{\label{sNetObject}Distributed Objects}
+\subsubsection{\label{sNetObject}Distributed Objects}
 
@@
 
-\subsubsection{\label{sInitData}InitData - a Static Distributed Object}
+\paragraph{\label{sInitData}InitData - a Static Distributed Object}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=66, lastline=77]
+{\footnotesize\lstinputlisting[firstline=62, lastline=73]
   {../../../Equalizer/examples/eqPly/initData.cpp}}
 
-\subsubsection{FrameData - a Versioned Distributed Object}
+\paragraph{FrameData - a Versioned Distributed Object}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=123, lastline=130]
+{\footnotesize\lstinputlisting[firstline=124, lastline=131]
   {../../../Equalizer/examples/eqPly/frameData.h}}
-{\footnotesize\lstinputlisting[firstline=54, lastline=82]
+{\footnotesize\lstinputlisting[firstline=53, lastline=81]
   {../../../Equalizer/examples/eqPly/frameData.cpp}}
 
-\subsubsection{\label{sSceneData}Scene Data}
+\paragraph{\label{sSceneData}Scene Data}
 
@@
 
-\subsection{Config}
+\subsubsection{Config}
 
@@
 
-\subsubsection{Initialization and Exit}
+\paragraph{Initialization and Exit}
 
@@
 \textsf{eq::Con\-fig::exit}, and then de-registers the initialization and
-frame data objects with the session:
+frame data objects:
 
@@
 
-\subsubsection{Frame Control}
-
-The rendering frames are issued by the application. The \textsf{eqPly::Config}
-overrides \textsf{startFrame} to update its data, commit a new version of the
-frame data object, and then requests the rendering of a new frame using the
-current frame data version. This version is passed to the rendering callbacks
-and will be used by the rendering threads to synchronize the frame data to the
-state belonging to the current frame:
-
-{\footnotesize\lstinputlisting[firstline=294, lastline=301]
+\paragraph{Frame Control}
+
+The rendering frames are issued by the application main loop. The
+\textsf{eqPly::Config} overrides \textsf{startFrame} to update its data, commit
+a new version of the frame data object, and then requests the rendering of a new
+frame using the current frame data version. This version is passed to the
+rendering callbacks and will be used by the rendering threads to synchronize the
+frame data to the state belonging to the current frame. This ensures that all
+frame-specific data, e.g., the camera position, is used consistently to generate
+the frame:
+
+{\footnotesize\lstinputlisting[firstline=274, lastline=281]
   {../../../Equalizer/examples/eqPly/config.cpp}}
 
-Updating the shared data is decomposed in different steps. If a tracker is used,
-the current head position and orientation is retrieved and passed to Equalizer,
-which uses the head matrix together with the wall or projection description to
-compute the view frusta, as explained in \sref{sTracking}:
-
-{\footnotesize\lstinputlisting[firstline=303, lastline=311]
+The update of the per-frame shared data consist of calculating the camera
+position based on the current navigation mode, and determining the idle state
+for rendering. When idle, \textsf{eqPly} performs anti-aliasing to gradually
+reduce aliasing effects in the rendering. The idle state is tracked by the
+application and used by the rendering callbacks to jitter the frusta, accumulate
+and display the results, as described in \sref{sFSAA}:
+
+{\footnotesize\lstinputlisting[firstline=283, lastline=315]
   {../../../Equalizer/examples/eqPly/config.cpp}}
 
-The camera position is recomputed based on the current navigation mode, and the
-idle state for rendering is set. When idle, \textsf{eqPly} performs
-anti-aliasing to gradually reduce aliasing effects in the rendering. The idle
-state is tracked by the application and used by the rendering callbacks to
-jitter the frusta, accumulate and display the results:
-
-{\footnotesize\lstinputlisting[firstline=313, lastline=343]
-  {../../../Equalizer/examples/eqPly/config.cpp}}
-
-\subsubsection{Event Handling}
+\paragraph{Event Handling}
 
@@
 \textsf{eq::Config::sendEvent}. At the end of the frame,
-\textsf{Config::finishFrame} calls \textsf{Config::handleEvents} to do
-the event handling. The default implementation processes all pending
-events by calling \textsf{Config::handleEvent} for each of them.
-
-Since \textsf{eqPly} uses event-driven execution, the config maintains a
-dirty state to allow the application to know when a redraw is needed.
+\textsf{Config::finishFrame} calls \textsf{Config::handle\-Events} to perform
+the event handling. The default implementation processes all pending events by
+calling \textsf{Config::handleEvent} for each of them.
+
+Since \textsf{eqPly} uses event-driven execution, the config maintains a dirty
+state to know when a redraw is needed.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=471, lastline=476]
+{\footnotesize\lstinputlisting[firstline=446, lastline=453]
   {../../../Equalizer/examples/eqPly/config.cpp}}
 
-\subsubsection{Model Handling}
-
-Model data is handled by the config. Models in eqPly are static,
-and therefore the render clients only need to map one instance of the
-model per node.
+\paragraph{Model Handling}
+
+Models in eqPly are static, and therefore the render clients only need to map
+one instance of the model per node. The mapped models are shared by all pipe
+render threads, which access them read-only.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=265, lastline=292]
+{\footnotesize\lstinputlisting[firstline=245, lastline=272]
   {../../../Equalizer/examples/eqPly/config.cpp}}
 
-\subsubsection{Layout and View Handling}
-
-For layout and model selection, \textsf{eqPly} maintains an active view
-and canvas. The identifier of the active view is stored in the frame
-data, which is used by the render client to highlight it using a
-different background color. The active view can be selected by clicking
-into a view, or by cycling through all the views using a keyboard
-shortcut.
+\paragraph{Layout and View Handling}
+
+For layout and model selection, \textsf{eqPly} maintains an active view and
+canvas. The identifier of the active view is stored in the frame data, which is
+used by the render client to highlight it using a different background
+color. The active view can be selected by clicking into a view, or by cycling
+through all views using a keyboard shortcut.
 
@@
 application and render clients might use a different node factory, instantiating
-a different type of \textsf{eq::Node} and children.
-
-All dynamic data is multi-buffered in \textsf{eqPly}. During
-initialization, the node relaxes the thread synchronization between the
-node and pipe threads, unless the configuration file overrides
-this. \sref{sThreads} provides a detailed explanation of thread
-synchronization in Equalizer.
+a different types of \textsf{eq::Config}\dots\textsf{eq::Channel}.
+
+All dynamic data is multi-buffered in \textsf{eqPly}. During initialization, the
+\textsf{eqPly::Node} relaxes the thread synchronization between the node and
+pipe threads, unless the configuration file overrides this. \sref{sThreads}
+provides a detailed explanation of thread synchronization modes in Equalizer.
 
@@
 retrieves the distributed \textsf{InitData}. The object is directly unmapped
-since it is static, and therefore all data has been retrieved:
-
-{\footnotesize\lstinputlisting[firstline=247, lastline=263]
+since it is static, and therefore all data has been retrieved during :
+
+{\footnotesize\lstinputlisting[firstline=227, lastline=243]
   {../../../Equalizer/examples/eqPly/config.cpp}}
 
-\subsection{Pipe}
+\subsubsection{Pipe}
 
@@
 and therefore do not compete for resources or cause OpenGL context
-switches. Multiple GPU run in parallel with each other.
+switches. Multiple GPU threads run in parallel with each other.
 
@@
 frames at any given time. Therefore frame-specific data has to be allocated for
-each pipe thread, which affect only to the frame data.
-
-The frame data is a member variable of the \textsf{eqPly::Pipe}, and is mapped
-to the identifier provided by the initialization data. The initialization in
-\textsf{eq::Pipe} does the GPU-specific initialization, which is
-window-system-dependent. On AGL the display ID is determined, and on glX the
-display connection is opened:
+each pipe thread, which is only the frame data in \textsf{eqPly}. The frame data is a member variable of the \textsf{eqPly::Pipe}, and is mapped
+to the identifier provided by the initialization data:
 
@@
   {../../../Equalizer/examples/eqPly/pipe.cpp}}
+
+The initialization in \textsf{eq::Pipe} does the GPU-specific initialization by
+calling \textsf{configInitSystemPipe}.
+which is window-system-dependent. On AGL the display ID is determined, and on
+glX the display connection is opened.
 
@@
 
-\subsubsection{Window System}
+\paragraph{Window System}
 
@@
 
-\subsubsection{\label{sCarbonThread}Carbon/AGL Thread Safety}
+\paragraph{\label{sCarbonThread}Carbon/AGL Thread Safety}
 
@@
 
-\subsubsection{Frame Control}
+\paragraph{Frame Control}
 
@@
 
-\subsection{\label{sEqplyWIndow}Window}
+\subsubsection{\label{sEqplyWIndow}Window}
 
@@
 
-\subsubsection{Window System Interface}
+\paragraph{Window System Interface}
 
@@
 
-\subsubsection{\label{sObjectManager}Object Manager}
+\paragraph{\label{sObjectManager}Object Manager}
 
@@
 
-\subsubsection{\label{sRendering}Rendering}
+\paragraph{\label{sRendering}Rendering}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=124, lastline=144]
+{\footnotesize\lstinputlisting[firstline=122, lastline=142]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=146, lastline=172]
+{\footnotesize\lstinputlisting[firstline=144, lastline=170]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=174, lastline=185]
+{\footnotesize\lstinputlisting[firstline=172, lastline=183]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=593, lastline=626]
+{\footnotesize\lstinputlisting[firstline=589, lastline=621]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 The model data is spatially organized in a 3-dimensional
-kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
+kD-tree\footnote{\link{http://en.wikipedia.org/wiki/Kd-tree}}
 for efficient view frustum culling. When the model is loaded by
@@
 
-\subsubsection{Assembly}
+\paragraph{Assembly}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=256, lastline=275]
+{\footnotesize\lstinputlisting[firstline=249, lastline=275]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-\section{Advanced Features}
+\subsection{Advanced Features}
 
@@
 
-\subsection{\label{sEventHandling}Event Handling}
+\subsubsection{\label{sEventHandling}Event Handling}
 
@@
 
-\subsubsection{Threading}
-
-Events are received and processed by the pipe thread a window belongs
-to. For AGL, Equalizer internally forwards the event from the main
-thread, where it is received, to the appropriate pipe thread. This model allows
-window and channel modifications which are naturally thread-safe, since they are
-executed from the corresponding render thread.
-
-\subsubsection{\label{sMessagePump}Message Pump}
+\paragraph{Threading}
+
+Events are received and processed by the pipe thread a window belongs to. For
+AGL, Equalizer internally forwards the events from the main thread, where it is
+received, to the appropriate pipe thread. This model allows window and channel
+modifications which are naturally thread-safe, since they are executed from the
+corresponding render thread and therefore cannot interfere with rendering
+operations.
+
+\paragraph{\label{sMessagePump}Message Pump}
 
@@
 The method \textsf{Pipe::createMessagePump} is called by Equalizer during
-application and pipe thread initialization (before \textsf{Pipe::configInit}!)
+application and pipe thread initialization before \textsf{Pipe::configInit}
 to create a message pump for the given pipe. For AGL, this affects the node
@@
 
-If the application disables message pumping in Equalizer, it has to make
-sure the events are pumped externally, as it often done within external
-widget sets such as Qt.
-
-\subsubsection{Event Data Flow}
+If the application disables message pumping in Equalizer, it has to make sure
+the events are processed, as it often done by external widget sets such as Qt.
+
+\paragraph{Event Data Flow}
 
 \begin{wrapfigure}{r}{.382\textwidth}
-  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
-  {\caption{\label{fEventProcessing}Event Processing}}\vspace{-4ex}
+  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
+  {\caption{\label{fEventProcessing}Event Processing}}\vspace{-2ex}
 \end{wrapfigure}
 Events are received by an event handler. The event handler finds the
-\textsf{eq::Sys\-tem\-Win\-dow} associated to the event. It then creates a
-generic \textsf{Event}, which holds the event data in an independent format. The
+\textsf{eq::Sys\-tem\-Win\-dow} for the event. It then creates a generic
+\textsf{Event}, which holds the event data in an independent format. The
 original native event and this generic \textsf{Event} form the
@@
 
-\textsf{Window::processEvent} is responsible for handling
-the event locally and for translating it into a generic
-\textsf{ConfigEvent}. The config events are send to the application
-thread using \textsf{Config::sendEvent}.
-
-Events send using \textsf{Config::sendEvent} are queued in the
-application thread. After a frame has been finished,
-\textsf{Config::finishFrame} calls \textsf{Config::handleEvents}. The
-default implementation of this method provides non-blocking event
-processing, that is, it calls \textsf{Config::handleEvent} for each
-queued event. By overriding \textsf{handleEvents}, event-driven
-execution can easily be implemented.
-
-If the event was processed, \textsf{processEvent} has to return
-\textsf{true}. If \textsf{false} is returned to the event handler, the
-event will be passed to the previously installed, window-system-specific
-event handling function.
+\textsf{Window::processEvent} is responsible for handling the event locally and
+for translating it into a generic \textsf{ConfigEvent}. Pointer events are
+translated and dispatched to the channel under the mouse pointer. The window or
+channel \textsf{processEvent} method perform local updates such as setting the
+pixel viewport before forwarding the event to the application main loop. If the
+event was processed, \textsf{processEvent} has to return \textsf{true}. If
+\textsf{false} is returned to the event handler, the event will be passed to the
+previously installed, window-system-specific event handling function.
+
+After local processing, events are send using \textsf{Config::sendEvent} to the
+application node. On reception, they are queued in the application thread. After
+a frame has been finished, \textsf{Config::finishFrame} calls
+\textsf{Config::handleEvents}. The default implementation of this method
+provides non-blocking event processing, that is, it calls
+\textsf{Config::handleEvent} for each queued event. By overriding
+\textsf{handleEvents}, event-driven execution can easily be implemented.
 
@@
 
-\subsubsection{Default Implementation}
+\paragraph{Default Implementation}
 
@@
 
-Carbon events issued for AGL windows are initially receive by the node
+Carbon events issued for AGL windows are initially received by the node
 main thread, and automatically dispatched to the pipe thread. The pipe
@@
 
-\subsubsection{Custom Events in eqPixelBench}
+\paragraph{Custom Events in eqPixelBench}
 
@@
 
-The rendering routine of \textsf{eqPixelBench} in
-\textsf{Channel::frameDraw} loops through a number of pixel formats and
-types. For each of them, it measures the time to readback and assemble a
-full-channel image. The format, type, size and time is recorded in a
-config event, which is sent to the application.
-
-The \textsf{ConfigEvent} derives from the \textsf{eq::ConfigEvent}
-structure and has the following definition:
-
-{\footnotesize\lstinputlisting[firstline=36, lastline=57]
+The rendering routine of \textsf{eqPixelBench} in \textsf{Channel::frameDraw}
+loops through a number of pixel formats and types. For each of them, it measures
+the time to readback and assemble a full-channel image. The format, type, size
+and time is recorded in a config event, which is sent to the application. The
+new custom event types are defined as user events:
+
+{\footnotesize\lstinputlisting[firstline=38, lastline=44]
   {../../../Equalizer/examples/eqPixelBench/configEvent.h}}
 
-The \textsf{Config::sendEvent} method transmits an
-\textsf{eq::ConfigEvent} or derived class to the application. The
-ConfigEvent has to be a C-type structure, and its \textsf{size}
-member has to be set to the full size of the event to be transmitted.
-Each event has a type which is used to identify it by the config
-processing function.
-
-User-defined types start at \textsf{eq::ConfigEvent::USER}, and the
-member variable \textsf{ConfigEvent::user} can be used to store up to
-\textsf{EQ\_USER\_EVENT\_SIZE}\footnote{currently 32 bytes} bytes. In
-this space, the channel's name is stored. Additional variables are used
-to transport the pixel format and type, the size and the time it took
-for rendering.
-
-On the application end, \textsf{Config::handleEvent} uses the
-\textsf{ostream} operator for the derived config event to output these
-events in a nicely formatted way:
-
-{\footnotesize\lstinputlisting[firstline=58, lastline=71]
+The \textsf{Config::sendEvent} method provides an \textsf{eq::EventOCommand} to
+which additional data can be appended. The event output command is derived from
+the Collage \textsf{co::DataOStream} which provides a convenient
+\textsf{std::ostream}-like interface to send data between nodes. The event is
+sent when the command looses its scope, that is, immediately after the following
+call in \textsf{eqPixelBench::Channel}:
+
+{\footnotesize\lstinputlisting[firstline=477, lastline=478]
+  {../../../Equalizer/examples/eqPixelBench/channel.cpp}}
+
+Each event has a type which is used to identify it by the config processing
+function. On the application end, \textsf{Config::handleEvent} receives an
+\textsf{eq::EventICommand}, which provides deserialization capabilities of the
+received data. The underlying \textsf{co::DataIStream} does perform endian
+conversion if the endianess of the sending and receiving nodes does not
+match. The event data is decoded and printed in a nicely formatted way:
+
+{\footnotesize\lstinputlisting[firstline=57, lastline=87]
   {../../../Equalizer/examples/eqPixelBench/config.cpp}}
 
-\subsection{\label{sErrorHandling}Error Handling}
+\subsubsection{\label{sErrorHandling}Error Handling}
 
@@
 automatically used for printing an \textsf{Error} on an \textsf{std::ostream},
-and can be queried from the \textsf{base::ErrorRegistry}, which is accessible
-from \textsf{base::Global}.
+and can be queried from the \textsf{fabric::ErrorRegistry}, which is accessible
+from \textsf{fabric::Global}.
 
@@
 
-\subsection{\label{sThreads}Thread Synchronization}
+\subsubsection{\label{sThreads}Thread Synchronization}
 
 Equalizer applications use multiple, potentially asynchronous execution
-threads. The default execution model is focused on making the porting of
+threads. The default execution model is designed on making the porting of
 existing applications as easy as possible, as described in
@@
 
-\subsubsection{Threads}
-
-The application or node main thread is the primary thread of each
-process and executes the \textsf{main} function. The application and
-render clients initialize the local node for communications with other
-nodes, including the server, using \textsf{Client::initLocal}.
-
-During this initialization, Equalizer creates and manages two threads
+\paragraph{Threads}
+
+The application or node main thread is the primary thread of each process and
+executes the \textsf{main} function. The application and render clients
+initialize the local node for communications with other nodes, including the
+server, using \textsf{Client::initLocal}. The client derives from
+\textsf{co::LocalNode} which does provide most of the communication logic`.
+
+During this initialization, Collage creates and manages two threads
 for communication, the receiver thread and the command thread. Normally
@@
 
-The command thread processes internal Equalizer request from other
-nodes, for example during \textsf{co::Object} mapping. In some
-special cases the command thread executes application code, for example
-when a remote node maps a static or unbuffered object,
-\textsf{Object::getInstanceData} is called from the command thread.
-
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/threads.pdf}
-  {\caption{\label{fThreads}Threads within one Node Process}}
-\end{wrapfigure}
+The command thread processes internal requests from other nodes, for example
+during \textsf{co::Object} mapping. In some special cases the command thread
+executes application code, for example when a remote node maps a static or
+unbuffered object, \textsf{Object::getInstanceData} is called from the command
+thread.
+
 The receiver and command thread are terminated when the application
@@
 the application's rendering code. All pipe threads are terminated during
-\textsf{Config::exit}.
-
-Since a layout switch on a canvas may involve different resources, pipe threads
-may be started and stopped dynamically at runtime. Before such an update,
-Equalizer will always finish all pending rendering operations to ensure that all
-resources are idle. The rest of this section discusses the thread
-synchronization between the main thread and the pipe threads.
-
-\subsubsection{\label{sThreadModel}Thread Synchronization Models}
+\textsf{Config::exit}. Since a layout switch on a canvas may involve different
+resources, pipe threads may be started and stopped dynamically at
+runtime. Before such an update, Equalizer will always finish all pending
+rendering operations to ensure that all resources are idle.
+
+During compositing, readback and tranmission threads are created lazy on the
+first image readback and transmission. Performing readback, compression and
+transmission of images asynchronously pipelines these operations with subsequent
+rendering commands, which increases overall performance. Equalizer creates one
+thread per pipe to finalize an asynchronous readback and one thread per node to
+compress and transmit the images to other nodes. Both threads are only visible
+to transfer and compression plugins, not to any other application code.
+
+\begin{figure}[h!t]\center
+  \includegraphics[width=.9\textwidth]{images/threads.pdf}
+  {\caption{\label{fThreads}Threads within one Node Process}}
+\end{figure}
+
+\fig{fThreads} provides an overview on all the threads used by Equalizer and
+Collage. The rest of this section discusses the thread synchronization between
+the main thread and the pipe threads.
+
+\paragraph{\label{sThreadModel}Thread Synchronization Models}
 Equalizer supports three threading models, which can be set
@@
 format is commonly used to change the threading model for benchmarking
-and experimentation.
+and experimentation using test applications.
 
@@
 \begin{description}
-\item[ASYNC:] No synchronization happens between render threads, except
+\item[ASYNC:] No synchronization happens between the pipe render threads, except
   for synchronizing the finish of frame \textsf{current-latency}. The
-  \textsf{eqPly} and \textsf{eVolve} Equalizer examples use this
-  threading model, by setting it in \textsf{Node::configInit}. This
-  synchronization model provides the best performance and should be used
-  by all applications which multi-buffer all non-static, frame-specific
-  data.
+  \textsf{eqPly} and \textsf{eVolve} Equalizer examples activate this threading
+  model by setting it in \textsf{Node::configInit}. This synchronization model
+  provides the best performance and should be used by all applications which
+  multi-buffer all dynamic, frame-specific data.
 \item[DRAW\_SYNC:] Additionally to the synchronization of the async thread
@@
 larger cluster configurations it is therefore advisable to assign only assemble
-operations to the application node. If the machine running the application
-process should also contribute for rendering, a second node may be used to
-perform off-screen draw and readback operations for the application node
-process.
-
-\subsubsection{\label{sAdvDPlex}DPlex Compounds}
+operations to the application node, allowing it to run asynchronous to the
+rendering nodes. If the machine running the application process should also
+contribute to rendering, a second node on the same host can be used to perform
+off-screen draw and readback operations for the application node process.
+
+\paragraph{\label{sAdvDPlex}DPlex Compounds}
 
@@
 
-\subsubsection{Thread Synchronization in Detail}
+\paragraph{Thread Synchronization in Detail}
 
@@
 
-\subsection{OpenGL Extension Handling}
+\subsubsection{OpenGL Extension Handling}
 
@@
 extension handling, particularly the GLEW\_MX implementation providing
-multi-context support.
+multi-context support. The build system favors an existing GLEW\_MX
+installation, but will fall back to a builtin implementation. Applications
+should use \textsf{eq/client/gl.h} to include the proper GLEW headers. If the
+preprocessor define \textsf{EQ\_IGNORE\_GLEW} is set, no GLEW headers will be
+included into application code by any Equalizer header file.
 
@@
 \textsf{GLEWContext} using \textsf{glewGetContext}, looking up the corresponding
-function pointer in this table and calling the function. Equalizer applications
-can disable these GLEW macros by defining \textsf{EQ\_IGNORE\_GLEW} before
-including \textsf{eq/eq.h}.
+function pointer in this table and calling the function.
 
@@
 Functions called from another place need to define a macro or function
-\textsf{glewGetContext} that returns the pointer to the GLEWContext of
-the appropriate window:
+\textsf{glewGetContext} that returns the pointer to the GLEWContext of the
+appropriate window, e.g., as done by the \textsf{eqPly} kd-tree rendering
+classes:
 
@@
 
-\subsection{Advanced Window Initialization}
-
-This section explains window initialization in detail. It discusses in
-detail the handling of the different window systems. Window systems are
-integrated through the \textsf{SystemWindow} interface, which externalizes
-the details of the windowing API from the \textsf{eq::Window}
-implementation and facilitates the integration with new windowing APIs
-and custom applications.
-
-Equalizer provides sample implementations for the supported window
-systems. These sample implementations are intended to be sub-classed and
-various steps in the window initialization can be overwritten and
-customized.
+\subsubsection{Window System Integration}
+
+Communicating with GPUs, creating windows and handling events is operating
+system dependent in OpenGL. Equalizer abstracts the specifities behind the
+\textsf{eq::Win\-dow\-Sys\-temIF} interface and provides an implemenation for
+glX/X11, AGL/Carbon and WGL. This interface class defines a factory to
+instantiate \textsf{SystemPipe}, \textsf{SystemWindow} and
+\textsf{MessagePump}. Its constructor registers the available instances used at
+runtime, that is, the class should be instantiated exactly once per process by
+the implementation.
+
+The \textsf{SystemPipe} abstracts GPU-specific handling, most notably the
+implementation does query the display resolution and sets it on its
+\textsf{eq::Pipe} during initialization. It also contains a
+\textsf{GLXEWContext} and \textsf{WGLEWContext} dispatch table for the glX and
+WGL window system, respectively.
+
+The \textsf{SystemWindow} interface externalizes the details of the windowing
+API from the \textsf{eq::Window} implementation and facilitates the integration
+with new windowing APIs and custom applications. The system window implements
+the core functionality for Equalizer to interface with the underlying window
+system and is described in detail below.
+
+The \textsf{MessagePump} ensures that OS events are processed and is called by
+Equalizer from the main and render threads regularly. Typically it only needs to
+be customized when integrating a totally new window system.
+
+Equalizer provides sample implementations for the supported window systems in
+the \textsf{agl}, \textsf{glx} and \textsf{wgl} subfolders. These sample
+implementations are intended to be sub-classed and various steps in the window
+initialization can be overwritten and customized.
 
@@
 rewriting tedious window initialization code, e.g., the OpenGL pixel
-format selection.
-
-\subsubsection{\label{sDrawableConfig}Drawable Configuration}
+format selection. \fig{fOSWindow} shows the UML class hierarchy for the system
+window implementations.
+
+\paragraph{\label{sDrawableConfig}Drawable Configuration}
 
@@
 
-\subsubsection{AGL Window Initialization}
+\paragraph{AGL Window Initialization}
 
@@
 
-\subsubsection{GLX Window Initialization}
+\paragraph{GLX Window Initialization}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=70, lastline=104]
+{\footnotesize\lstinputlisting[firstline=71, lastline=105]
   {../../../Equalizer/eq/client/glx/window.cpp}}
 
-\subsubsection{WGL Window Initialization}
+\paragraph{WGL Window Initialization}
 
@@
 
-\subsection{\label{sTracking}Stereo and Immersive Rendering}
-
-Equalizer fully supports immersive rendering for Virtual Reality, supporting
-multiple tracked observers, head-mounted displays (HMD) and flexible focus
-distance support. Equalizer does not support tracking devices directly. Other
-open source and commercial implementations cover this functionality well
-and are easily integrated with Equalizer.
+\subsubsection{\label{sTracking}Stereo and Immersive Rendering}
+
+Equalizer fully supports immersive rendering for Virtual Reality by implementing
+multiple tracked observers, head-mounted displays (HMD), flexible focus distance
+and asymmetric eye positions. Equalizer optionally supports tracking devices
+using the VRPN or OpenCV libraries.
 
 An Equalizer configuration contains a number of observers. Each observer
-represents one tracked entity. Most immersive configurations only have one
-observer. It is possible to support multiple tracked viewers in the same
+represents one tracked entity. Most immersive configurations have one observer,
+but it is possible to support multiple tracked viewers in the same
 configuration, e.g., to use two HMDs.
 
-Each observer has its own head matrix, eye separation and focus information for
-tracking. In \textsf{eqPly}, all observers are updated with the same data. A
-typical application would assign different tracking data to different observers.
-
-The \textsf{eqPly} example contains basic support for head tracking to
-illustrate how applications use these features.
-
-\subsubsection{Stereo Rendering}
+Each observer has its own head matrix, eye positions and focus information for
+tracking. Typically each observer receives tracking data from a different
+device. The observer may set the name of a VRPN tracking device or the index of
+the OpenCV camera to be used to track this observer. The VRPN tracker is
+expected to deliver tracking data in the correct coordinate system, that is,
+using meter units with the same origin as the frustum descriptions in the
+Equalizer configuration file. The OpenCV tracker is limited to 2.5D tracking due
+to the limited amount of information gained by face detection.
+
+\paragraph{Stereo Rendering}
 
@@
     \includegraphics[width=.382\textwidth]{images/stereo}}
-  {\caption{Monoscopic (a) and Stereoscopic (b) Frusta}}
+  {\caption{Monoscopic(a) and Stereoscopic(b) Frusta}}
 \end{figure}
 
-In stereo rendering, the scene is rendered twice, with the two frusta
-'moved' by the distance between the eyes, as shown in \fig{fStereo}. The stereo
-frusta are asymmetric, offset by half of the eye base. The stereo convergence
-plane is the same as the projection surface, unless specified otherwise using
-the focus distance API (see \sref{sFocusDistance}).
+In stereo rendering, the scene is rendered twice, with the two frustum origins
+'moved' to the position of the left and right eye, as shown in
+\fig{fStereo}. The stereo frusta are asymmetric. The stereo convergence plane is
+the same as the projection surface, unless specified otherwise using the focus
+distance API (see \sref{sFocusDistance}).
 
@@
 
-\subsubsection{Immersive Rendering}
-
-In immersive visualization, the observer is tracked in and the view
-frusta are adapted to the viewer's position and orientation, as shown
-in \fig{fTracked}. The transformation $origin \rightarrow viewer$ is set by
-the application using \textsf{Observer::setHead\-Matrix}, which is used by
-the server to compute the eye positions, and consequently the view frusta.
+\paragraph{Immersive Rendering}
+
+In immersive visualization, the observer is tracked in and the view frusta are
+adapted to the viewer's position and orientation, as shown in
+\fig{fTracked}. The transformation $origin \rightarrow viewer$ is set by the
+integrated tracking code or by the application using
+\textsf{Observer::setHead\-Matrix}, which is used by the server to compute the
+absolute eye positions, and consequently the view frusta.
 
@@
     \includegraphics[width=.382\textwidth]{images/hmd}}
-  {\caption{\label{fImmersive}Tracked (a) and HMD (b) Immersive Frusta}}
+  {\caption{\label{fImmersive}Tracked(a) and HMD(b) Immersive Frusta}}
 \end{figure}
@@
 Tracking events are often delivered asynchronously by an external source. The
-recommended way to integrate them into Equalizer is to emit an
-\textsf{OBSERVER\_MOTION} event, which has to contain the observer's object
-identifier and a \textsf{Matrix4f} head matrix. Equalizer optionally implements
-camera-based head tracking using OpenCV, and uses this mechanism to inject the
-asynchronously computed tracking matrix:
-
-{\footnotesize\lstinputlisting[firstline=194, lastline=195]
-  {../../../Equalizer/eq/client/cvTracker.cpp}}
+recommended way is to emit an \textsf{OBSERVER\_MOTION} event, which has to
+contain the observer identifier and a \textsf{Matrix4f} head matrix. Equalizer
+optionally implements head tracking using VRPN or OpenCV, and uses this
+mechanism to inject the asynchronously computed tracking matrix:
+
+{\footnotesize\lstinputlisting[firstline=194, lastline=194]
+  {../../../Equalizer/eq/client/detail/cvTracker.cpp}}
 
@@
 
-{\footnotesize\lstinputlisting[firstline=92, lastline=97]
+{\footnotesize\lstinputlisting[firstline=167, lastline=171]
   {../../../Equalizer/eq/client/observer.cpp}}
@@
 
-\subsubsection{\label{sFocusDistance}Focus Distance}
+\paragraph{\label{sFocusDistance}Focus Distance}
 
@@
   \subfigure[]{\label{fFixed}
-    \includegraphics[width=.32\textwidth]{images/focusFixed}}\hfil
+    \includegraphics[scale=.27]{images/focusFixed}}\hfil
   \subfigure[]{\label{fOrigin}
-    \includegraphics[width=.32\textwidth]{images/focusOrigin}}\hfil
+    \includegraphics[scale=.27]{images/focusOrigin}}\hfil
   \subfigure[]{\label{fObserver}
-    \includegraphics[width=.32\textwidth]{images/focusObserver}}
+    \includegraphics[scale=.27]{images/focusObserver}}
   {\caption{Fixed(a) and dynamic focus distance relative to origin(b) and observer(c)}}
@@
 In Equalizer 1.2 and later the focus distance and mode can be configured and
-changed at runtime for each \textsf{observer} and on the \textsf{config} for
-untracked views. This allows applications to expose the focal plane to the user,
-or to automatically calculate it based on the scene and view direction.
+changed at runtime for each \textsf{observer}. This allows applications to
+expose the focal plane in their user interface, or to automatically calculate it
+based on the scene and view direction.
 
@@
 \textsf{relative\_to\_origin} and \textsf{relative\_to\_observer} use the focus
-distance to dynamically change the stereo convergence.
-
-The focus distance calculation relative to the origin allows to change the
-focus independent of the observer position by separating the projection surface
-from the stereo convergence plane upon which the observer focuses itself. The
-convergence plane of the first wall in the negative Z direction is moved to be
-at the given focus distance, as shown in \fig{fOrigin}. All other walls are
-moved by the same relative amount. The movement is made from the view of the
-central eye, thus leaving the mono frustum unchanged. \fig{fOrigin} shows the
-new logical 'walls' used for frustum calculations in white, while the physical
-projection from \fig{fFixed} is still visible.
+distance parameter to dynamically change the stereo convergence.
+
+The focus distance calculation relative to the origin allows to change the focus
+independent of the observer position by separating the projection surface from
+the stereo convergence plane. The convergence plane of the first wall in the
+negative Z direction is moved to be at the given focus distance, as shown in
+\fig{fOrigin}. All other walls are moved by the same relative amount. The
+movement is made from the view of the central eye, thus leaving the mono frustum
+unchanged. \fig{fOrigin} shows the new logical 'walls' used for frustum
+calculations in white, while the physical projection from \fig{fFixed} is still
+visible.
 
@@
 
-\subsubsection{\label{sScaling}Application-specific Scaling}
+\paragraph{\label{sScaling}Application-specific Scaling}
 
@@
 declare this scale as part of the \textsf{eq::View}, which will be applied to
-the virtual environment by Equalizer. Common metric model scales are provided as
-constants.
-
-\subsection{\label{sLayoutAPI}Layout API}
+the virtual environment by Equalizer. Common metric scale factors are provided
+as constants.
+
+\subsubsection{\label{sLayoutAPI}Layout API}
 
@@
 
-\subsubsection{Subclassing and Data Distribution}
+\paragraph{Subclassing and Data Distribution}
 
@@
 
-The render clients can access a slave instance of the view using
-\textsf{Channel::getView}. When called from one of the frame task
-methods, this method will return the view of the current destination
-channel for which the task method is executed. Otherwise it returns the
-channel's native view, if it has one. Only destination channels of an
-active canvas have a native view.
-
 \begin{wrapfigure}{r}{.612\textwidth}
@@
 
+The render clients can access a slave instance of the view using
+\textsf{Channel::getView}. When called from one of the frame task
+methods, this method will return the view of the current destination
+channel for which the task method is executed. Otherwise it returns the
+channel's native view, if it has one. Only destination channels of an
+active canvas have a native view.
+
 The most common entity to subclass is the \textsf{View}, since the application
@@
 
-{\footnotesize\lstinputlisting[firstline=40, lastline=47]
+{\footnotesize\lstinputlisting[firstline=39, lastline=46]
   {../../../Equalizer/examples/eqPly/view.cpp}}
-{\footnotesize\lstinputlisting[firstline=78, lastline=85]
+{\footnotesize\lstinputlisting[firstline=76, lastline=83]
   {../../../Equalizer/examples/eqPly/view.cpp}}
 
-\subsubsection{Run-time Layout Switch}
+\paragraph{Run-time Layout Switch}
 
@@
 
-Initializing an entity can fail. If that happens, the server will exit
-the whole configuration, and send an \textsf{EXIT} event to the
-application node. The exit event will cause the application to exit,
-since it resets the config's running state.
-
-\subsubsection{Run-time Stereo Switch}
+Initializing an entity can fail. If a failure occurs and runtime reliability is
+not active, the server will exit the whole configuration, and send an
+\textsf{EXIT} event to the application node. The exit event will cause the
+application to exit, since it resets the config's running state.
+
+\paragraph{Run-time Stereo Switch}
 
@@
 
-\begin{wrapfigure}{r}{.618\textwidth}
-  \includegraphics[width=.618\textwidth]{images/stereoSwitch.pdf}
-  {\caption{\label{fStereoSwitch}Using two different decompositions during
-      stereo and mono rendering}}
-\end{wrapfigure}
 The configuration contains stereo information in three places: the view, the
-segment and the compound.
-
-The view defines which stereo mode is active: mono or stereo. This setting can
-be done in the configuration file or programmatically.
+segment and the compound.  The view defines which stereo mode is active: mono or
+stereo. This setting can be done in the configuration file or programmatically.
 
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/stereoSwitch.pdf}
+  {\caption{\label{fStereoSwitch}Using two different decompositions during
+      stereo and mono rendering}}
+\end{wrapfigure}
 The compound has an \textsf{eye} attribute defining which eyes it is capable of
@@
 
-\subsubsection{View Frustum Update}
-
-Changing view frustum parameters is typically done for non-fullscreen
+\paragraph{Frustum Updates}
+
+Frustum parameters can be changed at runtime for views and segments by the
+application. View frusta are typically changed for non-fullscreen
 application windows and multi-view layouts, where the rendering is not
@@
 typical use case is changing the field-of-view of the rendering.
+
+Segment frusta are changed when the display system changes at runtime, for
+example by moving a tracked LCD through a virtual world.
+
+The view and segment are derived from \textsf{eq::Frustum} (\fig{fViewUML}), and
+the application process can set the wall or projection parameters at
+runtime. For a description of wall and projection parameters please refer to
+\sref{sFrustum}. The new data will be effective for the next frame. The frustum
+of a view overrides the underlying frustum of the segments.
+
+The default Equalizer event handling is using the view API to maintain
+the aspect ratio of destination channels after a window resize. Without
+updating the wall or projection description, the rendering would become
+distorted.
 
@@
 \end{wrapfigure}
-A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
-application process can set the wall or projection parameters at
-runtime. For a description of wall and projection parameters please refer
-to \sref{sFrustum}. The new data will be effective for the next frame.
-
-The frustum of a view overrides the underlying frustum of the
-segments. The frustum of a view does not typically correspond to a
-physical projection, and can therefore be changed at runtime. The
-frustum of a segment does however correspond to a physical projection,
-and is therefore immutable.
-
-The default Equalizer event handling is using the view API to maintain
-the aspect ratio of destination channels after a window resize. Without
-updating the wall or projection description, the rendering would become
-distorted.
-
-When a window is resized, a \textsf{CHANNEL\_RESIZE} event is eventually
-generated. If the corresponding channel has a view,
-\textsf{Channel::processEvent} sends a \textsf{VIEW\_RESIZE} event to
-the application. This event contains the identifier of the view.
-
-The config event is dispatched to \textsf{View::handleEvent}. Using the original
-size and wall or projection description of the view, a new wall or projection is
+When a window is resized, a \textsf{CHANNEL\_RESIZE} event is generated. If the
+corresponding channel has a view, \textsf{Channel::processEvent} sends a
+\textsf{VIEW\_RESIZE} event to the application. This event contains the
+identifier of the view. The config event is dispatched to
+\textsf{View::handleEvent} on the application thread. Using the original size
+and wall or projection description of the view, a new wall or projection is
 computed, keeping the aspect ratio and the height of the frustum constant. This
@@
 
-\subsection{\label{sROI}Region of Interest}
-
-\subsubsection{Motivation}
+\subsubsection{\label{sROI}Region of Interest}
+
+\paragraph{Motivation}
 Regions of interest (ROI) define which part of the framebuffer was updated by
@@
 scalable rendering.
-
-The first optimization concerns the compositing operation. The declared ROI
-allows Equalizer to restrict the readback, compression, network transmission and
-assembly to only the relevant pixels.
 
@@
   }%
-  {\caption{\label{fROIDB}ROI for a two-way (a) and four-way DB (b) compound}}
+  {\caption{\label{fROIDB}ROI for a two-way(a) and four-way DB(b) compound}}
 \end{wrapfigure}
-This is particularly important for DB compounds, which typically cover less
-screen space on each channel as resources are added to the
-decomposition. \fig{fROIDB} illustrates this for a two-way and four-way
-decomposition. Declaring the ROI alleviates the increasing compositing cost as
-resources are added to a DB compound.
+The first optimization concerns the compositing operation. The declared ROI
+allows Equalizer to restrict the readback, compression, network transmission and
+assembly to only the relevant pixels. This is particularly important for DB
+compounds, where less screen space on each channel is covered as resources are
+added to the decomposition. \fig{fROIDB} illustrates this for a two-way and
+four-way decomposition. Declaring the ROI alleviates the increasing compositing
+cost as resources are added to a DB compound.
 
@@
 
-The second optimization is important for load-balanced 2D compounds. The 2D
-load equalizer uses internal timing statistics to form a 2D load grid for
-computing the split of the next frame. Without ROI, an even load within each
-channel has to be assumed. With ROI, an even load within the declared, smaller
-region of interest can be used. This causes a smaller error in the load
-estimation, and therefore a more accurate load prediction.
+The second optimization is important for load-balanced 2D compounds. The 2D load
+equalizer uses internal timing statistics to form a 2D load grid for computing
+the split of the next frame. Without ROI, an even load within each tile has to
+be assumed. With ROI, an even load within the declared, smaller region of
+interest can be used. This causes a smaller error in the load estimation, and
+therefore a more accurate load prediction.
 
@@
 
-\subsubsection{ROI in eqPly}
+\paragraph{ROI in eqPly}
 The application should declare its regions of interest during rendering. This
@@
 graph. Equalizer will track and optimize multiple regions automatically. The
-current implementation of Equalizer merges all regions of a single channel into
-one region for compositing and load-balancing. Later versions may use different,
+current implementation merges all regions of a single channel into one region
+for compositing and load-balancing. Later Equalizer versions may use different,
 more optimal, heuristics based on the application-declared regions.
@@
 
-\subsection{\label{sCompositing}Image Compositing for Scalable Rendering}
-
-Two task methods are responsible for collecting and compositing the
-result image during scalable rendering. Scalable rendering is a use case
-of parallel rendering, when multiple channels are contributing to a single
-view. This requires reading back the pixel data from the source GPU and
-assembling it on the destination GPU.
+\subsubsection{\label{sCompositing}Image Compositing for Scalable Rendering}
+
+Two task methods are responsible for collecting and compositing the result image
+during scalable rendering. Scalable rendering is a use case of parallel
+rendering, where multiple channels are contributing to a single view. This
+requires reading back the pixel data from the source GPU and assembling it on
+the destination GPU.
 
@@
 
-\subsubsection{Compression Plugins}
+\paragraph{Compression Plugins}
 
@@
 
-Please refer to the Equalizer API documentation on the website for the
-full specification for compression plugins. The Equalizer DSO doubles as
-a compression plugin and implements a set of compression engines, which
-can be used as a reference implementation.
+Please refer to the Equalizer API documentation on the website for the full
+specification for compression plugins. The Lunchbox and Equalizer DSOs double as
+compression plugins and implement a set of compression engines, which can be
+used as a reference implementation.
 
@@
 
-\subsubsection{\label{sDirectSend}Parallel Direct Send Compositing}
+\paragraph{\label{sDirectSend}Parallel Direct Send Compositing}
 
@@
 
-In direct send compositing, each rendering channel is also responsible
-for the sort-last composition of one screen-space tile. He receives
-the framebuffer pixels for his tile from all the other channels. The
-size of one tile decreases linearly with the number of source channels,
-which keeps the total amount of pixel data per channel constant.
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
+  {\caption{\label{fDirectSend}Direct Send Compositing}}
+\end{wrapfigure}
+In direct send compositing, each rendering channel is also responsible for the
+sort-last composition of one screen-space tile. It receives the framebuffer
+pixels for its tile from all the other channels. The size of one tile decreases
+linearly with the number of source channels, which keeps the total amount of
+pixel data per channel constant.
 
@@
 presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
-explaining and comparing this algorithm with the binary-swap algorithm.
+explaining and comparing this algorithm to the binary-swap algorithm.
 
@@
 
-Furthermore it has to be possible for the application to implement a read back
-of only the relevant region of interest, that is, the 2D area of the framebuffer
-updated during rendering. This optimization will be fully supported by later
-versions of Equalizer.
-
-\subsubsection{Frame, Frame Data and Images}
+\paragraph{Frame, Frame Data and Images}
 
@@
 \end{wrapfigure}
-Readback and assemble operations on frames and images are designed to be
-asynchronous. They have a start and finish method for both readback and
-assemble to allow the initiation and synchronization of the operation.
-Currently, only synchronous readback and assembly using
-\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
-respective start method of the image. Later versions of Equalizer will
-implement asynchronous pixel transfers.
-
-The offset of input and output frames characterizes the position of the
-frame data relative to the framebuffer, that is, the \textbf{window's}
-lower-left corner. For output frames this is the position of the channel
-relative to the window.
-
 For output frames, the frame data's pixel viewport is the area of the
@@
 
-%\begin{wrapfigure}{r}{.618\textwidth}
+\paragraph{\label{sCompositor}The Compositor}
+
 \begin{figure}[h!t]\center
@@
 \end{figure}
-
-\subsubsection{\label{sCompositor}The Compositor}
-
 The \textsf{Compositor} class gathers a set of static functions which
@@
 
-On a high level, the compositor combines multiple input frames using 2D
-tiling, depth-compositing for polygonal data or sorted, alpha-blended
-compositing for semi-transparent volumetric data. These operations
-composite either directly all images on the GPU, or use a CPU-based
-compositor and then transfer the preintegrated result to the GPU. The
-high-level entry points automatically select the best algorithm. The
-CPU-based compositor uses, based on availability, OpenMP and the
-Paracomp compositing library to accelerate its operation.
+On a high level, the compositor combines multiple input frames using 2D tiling,
+depth-compositing for polygonal data or sorted, alpha-blended compositing for
+semi-transparent volumetric data. These operations composite either directly all
+images on the GPU, or use a CPU-based compositor and then transfer the
+preintegrated result to the GPU. The high-level entry points automatically
+select the best algorithm. The CPU-based compositor uses OpenMP to accelerate
+its operation.
 
@@
 
-\subsubsection{Custom Assembly in eVolve}
+\paragraph{Custom Assembly in eVolve}
 
 The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
-texture-based volume rendering, where the 3D texture is intersected by
-view-aligned slices. The slices are rendered back-to-front and blended
-to produce the final image, as shown in \fig{fSlices}\footnote{Volume
-  Data Set courtesy of: SFB-382 of the German Research Council (DFG)}.
-
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/slices.pdf}
-  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}\vspace{-2ex}
-\end{wrapfigure}
+texture-based volume rendering, where the volume is intersected by view-aligned
+slices. The slices are rendered back-to-front and blended to produce the final
+image, as shown in \fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382
+  of the German Research Council (DFG)}.
+\begin{figure}[h!t]
+  \subfigure[]{\label{fVolResult}
+    \includegraphics[width=.45\textwidth]{images/volResult.png}
+  }\hfil
+  \subfigure[]{\label{fSlices}
+    \includegraphics[width=.45\textwidth]{images/slices.pdf}
+  }%
+  {\caption{Final Result(a) of \fig{fBlendPerspective} using Volume Rendering
+      based on 3D Texture Slicing(b).}}
+\end{figure}
+
 When using 2D (sort-first) or stereo decompositions, no special
@@
 fill rate (less samples per channel) and IO bandwidth for time-dependent data
-(less data per time step and channel). Since the amount of texture memory needed
-for each node decreases linearly, they make it possible to render data sets
-which are not feasible to visualize with any other approach.
+(less data update per time step and channel). Since the amount of texture memory
+needed for each node decreases linearly, sort-last rendering makes it possible
+to render data sets which are not feasible to visualize with any other approach.
 
@@
 
-\begin{wrapfigure}{r}{.382\textwidth}
-  \includegraphics[width=.382\textwidth]{images/volResult.png}
-  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
-\end{wrapfigure}
 For orthographic rendering, determining the compositing order of the
@@
 compositing. \fig{fBlendOrtho} shows this composition for one view.
-
-Finding the correct assembly order for perspective frusta is more
-complex. The perspective distortion invalidates a simple orientation
-criteria like the one used for orthographic frusta. For the view and
-frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
-  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
-1-4-3-2.
-
-To compute the assembly order, \textsf{eVolve} uses the angle between
-the $origin \rightarrow slice$ vector and the near plane, as shown in
-\fig{fBlendPerspective}. When the angle becomes greater than
-90\textdegree, the compositing order of the remaining frames has to be
-changed. The result image of this composition naturally looks the same
-as the volume rendering would when rendered on a single
-channel. \fig{fVolResult} shows the result of the composition from
-\fig{fBlendPerspective}.
-
 \begin{figure}[h!t]
@@
 
+Finding the correct assembly order for perspective frusta is more
+complex. The perspective distortion invalidates a simple orientation
+criteria like the one used for orthographic frusta. For the view and
+frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
+  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
+1-4-3-2.
+
+To compute the assembly order, \textsf{eVolve} uses the angle between
+the $origin \rightarrow slice$ vector and the near plane, as shown in
+\fig{fBlendPerspective}. When the angle becomes greater than
+90\textdegree, the compositing order of the remaining frames has to be
+changed. The result image of this composition naturally looks the same
+as the volume rendering would when rendered on a single
+channel. \fig{fVolResult} shows the result of the composition from
+\fig{fBlendPerspective}.
+
 The assembly algorithm described in this section also works with parallel
@@
 
-\subsection{\label{sFSAA}Subpixel Processing}
+\subsubsection{\label{sFSAA}Subpixel Processing}
 
@@
 
-\subsubsection{Transparent Software Anti-Aliasing}
+\paragraph{Transparent Software Anti-Aliasing}
 
@@
 The default implementation of \textsf{applyFrustum} and \textsf{applyOrtho} in
-\textsf{eq::Channel} apply jittering when a subpixel parameter is set.
+\textsf{eq::Channel} jitter the frustum when a subpixel parameter is set.
 Furthermore, the \textsf{Compositor} uses an \textsf{eq::Accum} buffer to
-accumulate and display the results of a subpixel decomposition in \textsf{assembleFramesUnsorted} and \textsf{assembleFramesSorted}.
-
-\subsubsection{Software Anti-Aliasing in eqPly}
+accumulate and display the results of a subpixel decomposition in
+\textsf{assembleFramesUnsorted} and \textsf{assembleFramesSorted}.
+
+\paragraph{Software Anti-Aliasing in eqPly}
 
@@
 frustum by a subpixel amount. The pixel is split into an 16x16 subpixel grid,
-and a sample is taken randomly within each sub-pixel. The samples are
+and a sample is taken randomly within each subpixel. The samples are
 accumulated and displayed after each step.
@@
 
-{\footnotesize\lstinputlisting[firstline=469, lastline=493]
+{\footnotesize\lstinputlisting[firstline=465, lastline=488]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 is only called on destination channels this operation is done only on the final
-rendering result. When all the steps are done the config will stop rendering new
-frames.
+rendering result after assembly. When all the steps are done the config will
+stop rendering new frames.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=329, lastline=353]
+{\footnotesize\lstinputlisting[firstline=331, lastline=350]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=551, lastline=572]
+{\footnotesize\lstinputlisting[firstline=547, lastline=568]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
 
-The \textsf{FrameData} class holds the information whether the application is in
-idle mode. The \textsf{Config} updates the idle mode information depending on
-the application's state.  Each \textsf{Channel} will perform anti-aliasing when
-no user event requires a redraw.
+The \textsf{FrameData} class holds the applications idle mode. The
+\textsf{Config} updates the idle mode information depending on the application's
+state. Each \textsf{Channel} performs anti-aliasing when no user event requires
+a redraw.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=500, lastline=506]
+{\footnotesize\lstinputlisting[firstline=496, lastline=502]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
@@
 
-{\footnotesize\lstinputlisting[firstline=512, lastline=538]
+{\footnotesize\lstinputlisting[firstline=508, lastline=534]
   {../../../Equalizer/examples/eqPly/channel.cpp}}
 
-\subsubsection{Subpixel Compounds in eqPly}
+\paragraph{Subpixel Compounds in eqPly}
 
@@
 
-When an application is passing an \textsf{eq::Accum} object to the compositor,
-the compositor adds subpixel decomposition result into this accumulation
+When an application passes an \textsf{eq::Accum} object to the compositor, the
+compositor adds subpixel decomposition result into this accumulation
 buffer. Otherwise it uses its own accumulation buffer, which it clears in the
@@
 
-\subsection{\label{sStatistics}Statistics}
-
-\subsubsection{Statistics Gathering}
+\subsubsection{\label{sStatistics}Statistics}
+
+\paragraph{Statistics Gathering}
 
@@
 finished, the events are pushed to the local (app-)node for
-visualization. Furthermore, the server also receives these event which will be
-used to implement the various runtime adjustments by the equalizers.
+visualization. Furthermore, the server also receives these events, which are
+used by the equalizers to implement the various runtime adjustments.
 
@@
 
-\subsubsection{\label{sStatisticsOverlay}Statistics Overlay}
+\paragraph{\label{sStatisticsOverlay}Statistics Overlay}
 
@@
 
-\subsection{\label{sCUDA}GPU Computing with CUDA}
+\subsubsection{\label{sCUDA}GPU Computing with CUDA}
 
@@
 steps to write a distributed GPGPU application using Equalizer are described
-in the following.
+in this section.
 
@@
 
-\subsubsection{CUDA Initialization}
+\paragraph{CUDA Initialization}
 
@@
 
-\subsubsection{CUDA Memory Management}
+\paragraph{CUDA Memory Management}
 
@@
 
-The Collage networking library provides a peer-to-peer communication
+The Collage network library provides a peer-to-peer communication
 infrastructure. It is used by Equalizer to communicate between the application
 node, the server and the render clients. It can also be used by applications to
-implement distributed processing independently or complementary to the
-Equalizer functionality.
+implement distributed processing independently or complementary to the Equalizer
+functionality.
 
@@
 \item[DataIStream] Decodes a buffer of received data into C++ objects and PODs
-  by implementing input stream operators. Performs endian swapping if requested.
-\item[Node/LocalNode] The abstraction of a process in the cluster. Nodes
+  by implementing input stream operators. Performs endian swapping if the
+  endianness differs between the originating and local node.
+\item[ICommand] The other side of OCommand, extending DataIStream.
+\item[Node and LocalNode] The abstraction of a process in the cluster. Nodes
   communicate with each other using connections. A LocalNode listens on various
-  connections and processes requests for a given process. A Node is a proxy for
-  a remote LocalNode.
-\item[Object] Provides object-oriented data distribution of C++ objects between
-  nodes within a session. Objects are registered or mapped on a LocalNode.
-\item[ICommand] The other side of OCommand. Is dispatched to Object or Node to
-  execute a specific task.
+  connections and processes requests for a given process. Received data is
+  wrapped in ICommands and dispatched to command handler methods. A Node is a
+  proxy for a remote LocalNode.
+\item[Object] Provides object-oriented, versioned data distribution of C++
+  objects between nodes within a session. Objects are registered or mapped on a
+  Local\-Node.
 \end{description}
@@
 
-\subsection{Packet Handling}
-
-Nodes and objects communicate using packets derived from \textsf{co::Packet}.
-The basic packet dispatch is implemented in the \textsf{co::Dispatcher} class,
-from which \textsf{co::Node} and \textsf{co::Object} is sub-classed.
-
-The dispatcher allows the registration of a packets with a dispatch queue and an
-invocation method. Each packet has a type and command identifier, which is used
-to identify the receiver, registered queue and method. The method
+\subsection{Command Handling}
+
+Nodes and objects communicate using commands derived from data streams. The
+basic command dispatch is implemented in the \textsf{co::Dispatcher} class, from
+which \textsf{co::Node} and \textsf{co::Object} are sub-classed.
+
+The dispatcher allows the registration of a commands with a dispatch queue and
+an invocation method. Each command has a type and command identifier, which is
+used to identify the receiver, registered queue and method. The method
 \textsf{dispatchCommand} pushes the packet to the registered queue and sets the
-registered command function on the packet command. When the
-commands are dequeued by the processing thread the registered method is executed
-using \textsf{Command::invoke}.
-
-A command function groups the method and \textsf{this} pointer, allowing
-to call a C++ method on a concrete instance. If no queue is registered
-for a certain command, \textsf{dispatchCommand} directly calls the
-registered command function.
-
-This dispatch and invocation functionality is used within Equalizer to
-dispatch commands from the receiver thread to the appropriate node or
-pipe thread, and then to invoke the command when it is processed by
-these threads.
+registered command function on the \textsf{ICommand}. When the commands are
+dequeued by the processing thread the registered method is executed using
+\textsf{ICommand::invoke}.
+
+A command function groups the method and \textsf{this} pointer, allowing to call
+a C++ method on a concrete instance. If no queue is registered for a certain
+command, \textsf{dispatchCommand} directly calls the registered command function
+directly.
+
+This dispatch and invocation functionality is used within Equalizer to dispatch
+commands from the receiver thread to the appropriate node or pipe thread, and
+then to invoke the command when it is processed by these threads.
 
@@
 nodes, e.g., to query connection information to connect to the node. Nodes use
-connections to communicate with each other by sending packets sub-classed from
-\textsf{co::NodePacket}.
+connections to communicate with each other by sending \textsf{co::OCommand}s.
 
@@
 A remote \textsf{Node} can either be connected explicitly by the application or
-due to a connection from the given remote node. The explicit connection can be
-done by programmatically creating a node, adding the necessary
+due to a connection from a remote node. The explicit connection can be done by
+programmatically creating a node, adding the necessary
 \textsf{ConnectionDescription}s and connecting it to the local node. It may also
@@
 Each \textsf{LocalNode} provides a \textsf{co::Zeroconf} communicator, which
-allows node and resource discovery. This requires compiling using a Lunchbox
-library supporting dns\_sd. The service "\_collage.\_tcp" is used to announce
-the presence of a listening \textsf{LocalNode} using the ZeroConf protocol to
-the network, unless the local node has no listening connections. The node
-identifier and all listening connection descriptions are announced using keys
-starting with "co\_". Internally Collage uses this information to connect
-unknown nodes by using the node identifier alone.
+allows node and resource discovery. This requires using a Lunchbox library
+supporting dns\_sd. The service "\_collage.\_tcp" is used to announce the
+presence of a listening \textsf{LocalNode} using the ZeroConf protocol to the
+network, unless the local node has no listening connections. The node identifier
+and all listening connection descriptions are announced using keys starting with
+"co\_". Internally Collage uses this information to connect unknown nodes by
+using the node identifier alone.
 
@@
 \subsubsection{Communication between Nodes}
-
-\fig{fNetNode} shows the communication between two nodes. The nodes in
-the listening state also have a listening connection, which is not shown
-in the Figure for complexity reasons.
-
-When the remote node sends a packet, the listening node receives the
-packet and dispatches it from the receiver thread using the method
-\textsf{dispatchCommand}. The default implementation knows how to
-dispatch packets of type node, session or object. Applications can
-define custom data types for packets, and then have to extend
-\textsf{dispatchCommand} to handle these custom data types.
 
@@
 \end{wrapfigure}
-Node packets are directly dispatched using
-\textsf{Dispat\-cher::\-dis\-patchCommand}. For object packets, the appropriate
-object is found on the local node and \textsf{LocalNode::dis\-patchCommand} will
-dispatch the packet.
-
-If \textsf{dispatchCommand} returns false, the packet will be
+\fig{fNetNode} shows the communication between two nodes. When the remote node
+sends a command, the listening node receives the command and dispatches it from
+the receiver thread using the method \textsf{dispatchCommand}. The default
+implementation knows how to dispatch commands of type node or
+object. Applications can define custom data types for commands, and then have to
+extend \textsf{dispatchCommand} to handle these custom data types.
+
+Node commands are directly dispatched using
+\textsf{Dispat\-cher::\-dis\-patchCommand}. For object commands, the appropriate
+object is found on the local node and \textsf{Object::dis\-patchCommand} is
+called on the instance to dispatch the command.
+
+If \textsf{dispatchCommand} returns false, the command will be
 re-dispatched later. This is used if an object has not been mapped
-locally, and therefore the packet could not be dispatched.
-
-If an application wants to extend communication on the node level, it can either
-define its own datatype for packets, or to define custom node packets. Packets
-with a custom datatype are handled by overriding
-\textsf{LocalNode::dispatchCommand}. Custom node packets can use any command
-greater than \textsf{CMD\_NODE\_CUSTOM} as the packet command. By registering a
-\textsf{co::CommandFunc} for these commands, the \textsf{co::Dispatcher}
-dispatch and invoke mechanism can be used automatically.
+locally, and therefore the command could not be dispatched.
+
+If an application wants to extend communication on the node level, it can define
+its own datatype for commands, define custom node commands or use custom
+commands.
+
+Commands with a custom datatype use a command type greater or equal than
+\textsf{COMMAND\_CUSTOM}. The receiving local node has to override
+\textsf{LocalNode::dis\-patch\-Command} to handle them.
+
+Custom node commands use the command type \textsf{COMMAND\_NODE} and any command
+greater than \textsf{CMD\_NODE\_CUSTOM}. By registering a
+\textsf{co::CommandFunc} on the receiving local node for these commands, the
+\textsf{co::Dispatcher} dispatch and invoke mechanism is used
+automatically.
+
+These two aforementioned mechanisms are very efficient but require a clear,
+linear inheritance from \textsf{co::LocalNode}. In more complex scenarios, e.g.,
+when two libraries share the same local node for communication, custom commands
+can be used. Custom commands are indentified by a unique 128 bit integer, which
+is either generated randomly or based on a hash of a unique URI:
+
+{\footnotesize\lstinputlisting[firstline=25, lastline=26]
+  {../../../Collage/tests/customCommands.cpp}}
+
+A specialized \textsf{Node::send} method transmits a
+\textsf{CustomOCommand}. The application has to register a custom command
+handler on the receiving side, which will be invoked during processing of the
+command on the given thread queue:
+
+{\footnotesize\lstinputlisting[firstline=78, lastline=84]
+  {../../../Collage/tests/customCommands.cpp}}
+
+Like any other output command, \textsf{CustomOCommand}s allow streaming
+additional data to the command:
+
+{\footnotesize\lstinputlisting[firstline=86, lastline=87]
+  {../../../Collage/tests/customCommands.cpp}}
 
@@
 deserialization of the distributed data. Objects are dynamically attached to a
-listening local node, which manages the network communication and packet
+listening local node, which manages the network communication and command
 dispatch between different instances of the same distributed object.
@@
 
+One instance of an object is registered with its local node, which makes this
+object the master instance. Slave instance on the same or other nodes are mapped
+to this master. During mapping they are initialized by transmitting a version of
+the master instance data. During commit, the change delta is pushed from the
+master to all mapped slave objects, using multicast connections when
+available. Slave objects can also commit data to their master instance, which in
+turn may recommit it to all slaves, as described in \sref{sSlaveCommit}.
+Objects can push their instance data to a set of nodes, as described in
+\sref{sObjectPush}.
+
 Distributed objects can be static (immutable) or dynamic. Dynamic objects are
@@
 The entry point for shared data on a render client is the identifier passed by
-the application to \textsf{Config::init}. This identifier typically contains the
-identifier of a static distributed object, and is passed by Equalizer to all
+the application to \textsf{eq::Config::init}. This identifier typically contains
+the identifier of a static distributed object, and is passed by Equalizer to all
 \textsf{configInit} task methods. Normally this initial object is mapped by the
@@
 version is passed with each new frame request.
-
-\begin{figure}[h!t]
-  \subfigure[]{\label{fObjectDist1}
-    \includegraphics[width=.3\textwidth]{images/umlObjectSubclass}}\hfil
-  \subfigure[]{\label{fObjectDist2}
-    \includegraphics[width=.3\textwidth]{images/umlObjectProxy}}\hfil
-  \subfigure[]{\label{fObjectDist3}
-    \includegraphics[width=.3\textwidth]{images/umlObjectMI}}%
-  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
-      Proxies or Multiple Inheritance}}
-\end{figure}
 
@@
 
+\begin{figure}[h!t]
+  \subfigure[]{\label{fObjectDist1}
+    \includegraphics[width=.3\textwidth]{images/umlObjectSubclass}}\hfil
+  \subfigure[]{\label{fObjectDist2}
+    \includegraphics[width=.3\textwidth]{images/umlObjectProxy}}\hfil
+  \subfigure[]{\label{fObjectDist3}
+    \includegraphics[width=.3\textwidth]{images/umlObjectMI}}%
+  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
+      Proxies or Multiple Inheritance}}
+\end{figure}
+
 \subsubsection{Change Handling and Serialization}
@@
 \begin{description}
-  \item[STATIC] The object is not versioned. The instance data is serialized
-    whenever a new slave instance is mapped. The serialization happens in the
-    command thread on the master node, i.e., in parallel to the application
-    threads. Since the object's distributed data is supposed to be static, this
-    should not create any race conditions. No additional data is stored.
-  \item[INSTANCE] The object is versioned, and the instance and delta data is
-    identical, that is, only instance data is serialized. The serialization
-    always happens before \textsf{LocalNode::registerObject} or
+  \item[STATIC] The object is not versioned nor buffered. The instance data is
+    serialized whenever a new slave instance is mapped. The serialization
+    happens in the command thread on the master node, i.e., in parallel to the
+    application threads. Since the object's distributed data is supposed to be
+    static, this should not create any race conditions. No additional data is
+    stored.
+  \item[INSTANCE] The object is versioned and buffered. The instance and delta
+    data are identical, that is, only instance data is serialized. The
+    serialization always happens before \textsf{LocalNode::registerObject} or
     \textsf{Object::commit} returns. Previous instance data is saved to be able
     to map old versions.
-  \item[DELTA] The object is versioned, and the delta data is typically smaller
-    than the instance data. Both the delta and instance data are serialized
-    before \textsf{Object::commit} returns. Previous instance data is saved to
-    be able to map old versions.
-  \item[UNBUFFERED] The object is versioned. No data is stored, and no previous
-    versions can be mapped. The instance data is serialized from the command
-    thread whenever a new slave instance is mapped, i.e., in parallel to
-    application threads. The application has to ensure that this does not create
-    any thread conflicts. The delta data is serialized before
+  \item[DELTA] The object is versioned and buffered. The delta data is typically
+    smaller than the instance data. Both the delta and instance data are
+    serialized before \textsf{Object::commit} returns. The delta data is
+    transmitted to slave instances for synchronization. Previous instance data
+    is saved to be able to map old versions.
+  \item[UNBUFFERED] The object is versioned and unbuffered. No data is stored,
+    and no previous versions can be mapped. The instance data is serialized from
+    the command thread whenever a new slave instance is mapped, i.e., in
+    parallel to application threads. The application has to ensure that this
+    does not create any thread conflicts. The delta data is serialized before
     \textsf{Object::commit} returns. The application may choose to use a
@@
 \textsf{co::Serializable} implements one typical usage pattern for data
-distribution based on \textsf{co::Object}. The \textsf{co::Serializable} data
-distribution is founded on the concept of dirty bits, allowing inheritance with
-data distribution. Dirty bits form a 64-bit mask which marks which parts of the
-object have to be distributed during the next commit. It is easier to use, but
+distribution for \textsf{co::Object}. The \textsf{co::Serializable} data
+distribution is based on the concept of dirty bits, allowing inheritance with
+data distribution. Dirty bits form a 64-bit mask which marks the parts of the
+object to be distributed during the next commit. It is easier to use, but
 imposes one typical way to implement data distribution.
@@
 \begin{description}
-\item[Inherit from \textsf{co::Serializable}:] The base class will
-  provide the dirty bit management and call serialize and deserialize
-  appropriately. By overriding \textsf{get\-Chan\-ge\-Type}, the default
-  versioning strategy might be changed.
-\item[Define new dirty bits:] Define dirty bits for each data item by
-  starting at \textsf{Serializable::DIRTY\_CUSTOM}, shifting this value
-  consecutively for each new dirty bit.
-\item[Implement serialize and deserialize:] For each object-specific
-  dirty bit which is set, stream the corresponding data item to or from
-  the provided stream. Call the parent method first in both
+\item[Inherit from \textsf{co::Serializable}:]
+The base class will provide the dirty bit management and call serialize and
+deserialize appropriately. By optionally overriding
+\textsf{get\-Chan\-ge\-Type}, the default versioning strategy might be changed:
+
+{\footnotesize\lstinputlisting[firstline=35, lastline=46]
+  {../../../Equalizer/examples/eqPly/frameData.h}}
+
+\item[Define new dirty bits:]
+Define dirty bits for each data item by starting at
+\textsf{Serializable::DIRTY\_CUSTOM}, shifting this value consecutively for each
+new dirty bit:
+
+{\footnotesize\lstinputlisting[firstline=124, lastline=131]
+  {../../../Equalizer/examples/eqPly/frameData.h}}
+
+\item[Implement serialize and deserialize:]
+For each object-specific dirty bit which is set, stream the corresponding data
+item to or from the provided stream. Call the parent method first in both
 functions. For application-specific objects, write a (de-)serialization
-  function.
+function:
+
+{\footnotesize\lstinputlisting[firstline=53, lastline=81]
+  {../../../Equalizer/examples/eqPly/frameData.cpp}}
+
 \item[Mark dirty data:] In each 'setter' method, call \textsf{setDirty}
-  with the corresponding dirty bit.
+  with the corresponding dirty bit:
+
+{\footnotesize\lstinputlisting[firstline=214, lastline=218]
+  {../../../Equalizer/examples/eqPly/frameData.cpp}}
 \end{description}
@@
 
+\begin{wrapfigure}{r}{.618\textwidth}
+  \includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
+  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
+\end{wrapfigure}
 When a slave object instance commits data, \textsf{pack} is called on the
@@
 Synchronizing slave data on a master instance does not create a new version. The
-application has to explicitly call \textsf{commit} to pack the changed data,
-which will create a new version send to all slaves.
+application has to explicitly mark the deserialized data dirty and call
+\textsf{commit} to pack and push the changed data to all slaves.
 
@@
 following situations:
-\begin{wrapfigure}{r}{.618\textwidth}
-  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
-  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
-  \vspace{-10ex}
-\end{wrapfigure}
 \begin{itemize}
 \item A slave instance will always receive its own changes when the master
-  commits the slave change.
+  re-commits the slave change.
 \item A slave instance commits data which does not correspond to the last
@@
   result in conflicts.
-\item A master instance applies data to an already changed database.
+\item A master instance applies data to an already changed object.
 \item The master instance commits a new version while a slave instance is
@@
       index for removal.
-    \item Modify a variable by serializing its value (instead of an operation on
-      it) so that the last change will win.
+    \item Modify a variable by serializing its value (instead of the operation
+      modifying it), so that the last change will win.
   \end{itemize}
@@
 
+\subsubsection{\label{sObjectPush}Push Object Distribution}
+
+During normal object mapping, the slave object requests the mapping and thereby
+pulls its instance data from the master object. Push-based data distribution is
+initiated from an object and transmits the data to a set of given nodes. It
+allows more efficient initialization of a group of slave instances when
+multicast is used, since for all nodes only a single instance data is
+transmitted.
+
+The transfer is initiated using \textsf{Object::push}. On all nodes given to the
+push command, \textsf{LocalNode::objectPush} is called once the data is
+received. The push operation prefers multicast connections, that is,
+\textsf{objectPush} is called on all nodes of the multicast groups used, even
+when they are not part of the node list given to the push command.
+
+The push command transmits two application-specific variables: a group
+identifier and an object type.
+
+The group identifier is intended to be used to identify the push operation. For
+example, a set of Equalizer render client nodes requests a scene to be pushed,
+transmitting the scene identifier to the master node. The master node collects
+all requests and then traverses the given scene graph pushing all scene objects
+using this scene identifier as the group identifier. The render client nodes
+will filter the received object pushes against their set of pending requests.
+
+The object identifier is used by the receiving nodes to instantiate the correct
+object types when they need to be created. In the previous example, the object
+identifier would describe the scene graph node type and the render clients would
+use a factory to recreate the scene graph.
+
+The \textsf{LocalNode::objectPush} callback is called from the command thread
+once the instance data is ready. It receives the aforementioned group identifier
+and object type as well as the object identifier and the \textsf{DataIStream}
+containing the instance data. The latter is typically used to call
+\textsf{Object::applyInstanceData} on the newly created object.
+
+The default implementation of \textsf{objectPush} calls registered push
+handlers. For each group identifier, a separate handler function may be
+registered using \textsf{LocalNode::registerPushHandler}.
+
+A push operation does not establish any master-slave object mapping, that is,
+the receiving side will typically use \textsf{LocalNode::mapObject} with
+\textsf{VERSION\_NONE} to establish a slave mapping.
+
 \subsubsection{Communication between Object Instances}
 
-Objects can send command packets to each other by subclassing
-\textsf{co::Ob\-ject\-Packet}, which is derived from \textsf{NodePacket}. The
-object packet extends the node packet by the object identifier and an object
-instance identifier.
+Objects can send commands to each other using \textsf{Ob\-ject\-OCommand}, which
+is derived from \textsf{OCommand}. The object command extends the command by the
+object identifier and an object instance identifier. The object output command
+is created by \textsf{Object::send} and transmitted when it looses its scope.
+
+Multiple instances of the same object (identifier) can be mapped on the same
+node and session simultaneously. Each object instance has a node-unique instance
+identifier. When the instance ID is set to \textsf{EQ\_INSTANCE\_ALL} (the
+default value), a command is dispatch to all object instances registered to the
+session with the correct object ID. Multiple object instances per node are used
+when different threads on the node need to process a different version of the
+same object.
 
@@
 \end{wrapfigure}
-Multiple instances of the same object (identifier) can be mapped on the same
-node and session simultaneously. Each object instance has a node-unique instance
-identifier. When the instance ID is set to \textsf{EQ\_INSTANCE\_ALL} (the
-default value), a command is dispatch to all object instances registered to the
-session with the correct object ID. Multiple object instance per node are used
-when different threads on the node need to process a different version of the
-same object.
-
-When the packet is intended for a specific instance, set the instance identifier
-of the object packet to the corresponding value.
+When the command is intended for a specific instance, set the corresponding
+instance identifier during \textsf{Object::send}.
 
 Distributed objects provide data synchronization functionality, as described
-above. Additionally, applications can send command packets by creating custom
-object packets with a command greater than \textsf{CMD\_OBJECT\_CUSTOM}.
+above. Additionally, applications can send command commands by creating custom
+object commands with a command greater than \textsf{CMD\_OBJECT\_CUSTOM}.
 
@@
 \textsf{eq::View}, are registered, mapped and synchronized by Equalizer. To
-execute the rendering tasks, the server send the task command packet to the
+execute the rendering tasks, the server send the task command command to the
 render clients, which dispatch and execute them in the received order.
@@
 The barrier uses both the data distribution for synchronizing its data,
-as well as custom command packets to implement the barrier logics. A
+as well as custom command commands to implement the barrier logics. A
 barrier is a versioned object. Each version can have a different height,
@@
 application namespace for argument, all Equalizer command line options start
-with \textsf{--eq-}. The following options are recognized:
+with \textsf{--eq-} and all Collage command line options with
+\textsf{--co-}. The following options are recognized:
 
@@
   be used multiple times.
+\item[--eq-gpufilter] applies the given regular expression against
+  \textsf{nodeName:port.device} during autoconfiguration and only uses the
+  matching GPUs.
 \item[--eq-modelunit \textless unitValue\textgreater] is used for scaling the
@@
   \sref{sHWSDUsage}.
-\item[--eq-config-flags \textless multiprocess\textbar multiprocess\_db\textbar
-  ethernet\textbar infiniband\textbar 2D\_horizontal\textbar
-  2D\_vertical\textbar 2D\_tiles\textgreater] is used as input for the
-  autoconfiguration to tweak the configuration generation.
+\item[--eq-config-flags]{\bf \textless multiprocess \textbar\ multiprocess\_db
+  \textbar\ ethernet \textbar\ infiniband \textbar\ 2D\_horizontal
+  \textbar\ 2D\_vertical \textbar\ 2D\_tiles\textgreater} is used as input for
+  the autoconfiguration to tweak the configuration generation.
 \item[--eq-config-prefixes \textless CIDR-prefixes\textgreater] is used as input
@@
   client machines use a different directory layout from the application machine.
-\item[--eq-listen \textless connection\textgreater] configures the
+\item[--co-listen \textless connection\textgreater] configures the
   local node to listen on the given connections. The formation of the connection
@@
   multiple times.
-
+%\item[-v, -vv] Set the default logging level to \textsf{INFO} or \textsf{VERB},
+%  respectively.
 \end{description}
@@
 
-The current Equalizer file format is a one-to-one representation of the
-server's internal data structures. Its purpose is intermediate, that is,
-it will gradually be replaced by automatic resource detection and
-automatic configuration. However the core scalability engine will always
-use a similar structure as currently exposed by the file format.
+The current Equalizer file format is a one-to-one representation of the server's
+internal data structures. Its purpose is intermediate, that is, it will
+gradually be replaced by automatic resource detection and configuration. However
+the core scalability engine will always use a similar structure as currently
+exposed by the file format.
 
 The file format represents an ASCII deserialization of the
-server. Streaming an \textsf{eq::server::Server} to an \textsf{eq::base::Log}
+server. Streaming an \textsf{eq::server::Server} to an \textsf{lunchbox::Log}
 ostream produces a valid configuration file. Likewise, loading a
@@
 
-The file format uses the same syntactical structure as VRML. If your
-editor supports syntax highlighting and formatting for VRML, you can use
-this mode for editing \textsf{.eqc} files.
+The file format uses the same syntactical structure as VRML. If your text editor
+supports syntax highlighting and formatting for VRML, you can use this mode for
+editing \textsf{.eqc} files.
 
@@
   \hline
+  \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_GRAB\_POINTER\\
+  \textbf{Value} & OFF \textbar \ ON\\
+  \textbf{Default} & ON\\
+  \textbf{Details} & Enables grabbing the mouse pointer outside of the window
+  during a drag operation.\\
+  \hline
+\end{tabularx}\\\vfill
+
+\begin{tabularx}{\textwidth}{|l|X|}
+  \hline
   \textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_COLOR\\
@@
 The server section consists of connection description parameters for the
-server listening sockets, and a number of configurations for this
+server listening sockets and a number of configurations for this
 server. Currently only the first configuration is used.
@@
 enforce a SDP connection. RDMA implements native connections over the InfiniBand
-verbs protocol on Linux. RSP provides reliable UDP multicast. PIPE uses a named
-pipe for fast interprocess communication on Windows.
-
-On Linux, RDMA connections provide the fastest implementation for InfiniBand
-fabrics. SDP is slower than RDMA and faster than IP over InfiniBand, and is also
-supported on Windows. Note that you can also use the transparent mode provided
-by most InfiniBand implementations to use SDP with TCP connections by preloading
-the SDP shared library.
+verbs protocol on Linux and Windows. RSP provides reliable multicast over
+UDP. PIPE uses a named pipe for fast interprocess communication on Windows.
+
+RDMA connections provide the fastest implementation for InfiniBand fabrics. SDP
+is slower than RDMA and faster than IP over InfiniBand. Note that you can also
+use the transparent mode provided by most InfiniBand implementations to use SDP
+with TCP connections by preloading the SDP shared library.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=90, lastline=99]
+{\footnotesize\lstinputlisting[firstline=91, lastline=100]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=101, lastline=110]
+{\footnotesize\lstinputlisting[firstline=102, lastline=111]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=112, lastline=130]
+{\footnotesize\lstinputlisting[firstline=113, lastline=131]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-The device identifies the graphics adapter. For the GLX window system
-this is the screen number, i.e., the number after the dot in the DISPLAY
-description (:0.\textbf{1}). The OpenGL output is always restricted by
-glX to the GPU attached to selected screen.
+The device identifies the graphics adapter. For the GLX window system this is
+the screen number, i.e., the number after the dot in the DISPLAY description
+(:0.\textbf{1}). The OpenGL output is restricted by glX to the GPU attached to
+selected screen.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=131, lastline=141]
+{\footnotesize\lstinputlisting[firstline=132, lastline=142]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=143, lastline=163]
+{\footnotesize\lstinputlisting[firstline=144, lastline=166]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=165, lastline=174]
+{\footnotesize\lstinputlisting[firstline=167, lastline=176]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 An observer represents a tracked entity, i.e, one user. It has a name, eye
-separation and focal plane parameters. The name of an observer can be used for
-debugging, it has no influence on the execution of Equalizer. It can be used to
-reference the observer in views, in which case the name should be unique. Not
-all views have to be tracked by an observer.
-
-{\footnotesize\lstinputlisting[firstline=179, lastline=185]
+separation, focal plane parameters and tracking device settings. The name of an
+observer can be used for debugging, it has no influence on the execution of
+Equalizer. It can be used to reference the observer in views, in which case the
+name should be unique. Not all views have to be tracked by an observer.
+The focal plane parameters are described in \sref{sFocusDistance}. The OpenCV
+camera identifies the camere device index to be used for tracking on the
+application node. The VRPN tracker identifies the VRPN device tracker device
+name for the given observer.
+
+{\footnotesize\lstinputlisting[firstline=181, lastline=192]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-A layout represent a set of logical views on one or more canvases. It
+A layout represents a set of logical views on one or more canvases. It
 has a name and child views. The name of a layout can be used for
@@
 
-{\footnotesize\lstinputlisting[firstline=187, lastline=190]
+{\footnotesize\lstinputlisting[firstline=194, lastline=197]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=190, lastline=211]
+{\footnotesize\lstinputlisting[firstline=197, lastline=218]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-The layouts referenced by the canvas can be applied to it by the application at
+The layouts referenced by the canvas can be activated by the application at
 runtime. One layout can be referenced by multiple canvases. The first layout is
@@
 multiple windows. For software swap synchronization, all windows using a swap
-barriers of the same name are synchronized. Hardware swap synchronization is
+barrier of the same name are synchronized. Hardware swap synchronization is
 used when a \textsf{NV\_group} is specified. All windows using the same
@@
 
-{\footnotesize\lstinputlisting[firstline=214, lastline=239]
+{\footnotesize\lstinputlisting[firstline=221, lastline=246]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 
-{\footnotesize\lstinputlisting[firstline=240, lastline=262]
+{\footnotesize\lstinputlisting[firstline=247, lastline=269]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
@@
 output framerate, typically for DPlex compounds. A monitor equalizer
-adjusts the output image zoom to monitor another canvas.
+adjusts the output image zoom to monitor another canvas. A tile equalizer
+automatically sets up tile queues between the destination and all source
+channels.
 
@@
 
-A wall or projection description is used to define the view frustum of
-the compound. The frustum is inherited and typically only defined on the
-topmost compound. The last specified frustum description is used. Sizes
-are specified in meters. \fig{fFrusta} illustrates the frustum
-parameters. Setting a frustum on a compound is discouraged, a proper view and
-segment description should be used instead. View frusta override segment frusta
-which override compound frusta.
+A wall or projection description is used to define the view frustum of the
+compound. The frustum is normally inherited from the view or segment. The
+frustum is inherited and typically only defined on the topmost compound. The
+last specified frustum description is used. Sizes are specified in
+meters. \fig{fFrusta} illustrates the frustum parameters. Setting a frustum on a
+compound is discouraged, a proper view and segment description should be used
+instead. View frusta override segment frusta which override compound frusta.
 
@@
 
-{\footnotesize\lstinputlisting[firstline=264, lastline=337]
+{\footnotesize\lstinputlisting[firstline=271, lastline=358]
+  {../../../equalizergraphics/documents/design/fileFormat.shtml}}
+{\footnotesize\lstinputlisting[firstline=362, lastline=363]
   {../../../equalizergraphics/documents/design/fileFormat.shtml}}
